---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(20260513)
```

# proxymix

<!-- badges: start -->
<!-- badges: end -->

`proxymix` fits multivariate Gaussian-mixture proxies that are **Kullback–Leibler optimal** to user-supplied target densities on $\mathbb{R}^p$. Three regimes from Hoek and Elliott (2024) are unified under one verb:

| Regime | When it applies | Method |
|---|---|---|
| **(i) moment** | $N = 1$ component | Closed-form moment matching |
| **(ii) sample** | i.i.d. samples from the target are available | Classical EM |
| **(iii) kld** | target density `f(x)` can be evaluated but **not** (cheaply) sampled | **KLD-EM with importance sampling** |

Regime (iii) is the reason the package exists. The sample-based mixture packages (`mclust`, `mixtools`, `flexmix`) all assume i.i.d. draws from the target; `proxymix` fits directly against an evaluable (possibly unnormalised) log-density.

**Why not MCMC?** If you can evaluate the unnormalised density you can always run a sampler and then fit a mixture to the draws. The regime-(iii) fit is the shortcut when what you want *is* the compact closed-form object: no chain tuning or convergence diagnostics, a deterministic pipeline given the seed, and a mixture whose marginals, conditionals, moments and samples are then available in closed form through the operator calculus. The trade-off is dimension: the importance sampling that drives regime (iii) loses effective sample size sharply beyond roughly $p = 5$–$10$, and every fit reports its effective sample size so that limit is visible rather than silent. For high-dimensional posteriors, sample with your favourite MCMC and use regime (ii) on the draws.

## Installation

From GitHub:

    # install.packages("remotes")
    remotes::install_github("max578/proxymix")

or locally from the source tree:

    R CMD build proxymix
    R CMD INSTALL proxymix_*.tar.gz

Documentation site: <https://max578.github.io/proxymix/>.

## Quick start

```{r quickstart}
library(proxymix)

## A target you can evaluate but not sample from — a 2D "banana".
banana <- banana_target()

## Fit a 3-component Gaussian mixture proxy via KLD-EM with importance sampling.
fit <- fit_proxymix(banana, N = 3L, regime = "kld",
                    proposal = is_mvt(n_dim = 2L, df = 5),
                    is_size = 2000L, max_iter = 60L, seed = 1L)

print(fit)

## Closed-form operations on the fitted mixture.
gmm_marginalise(fit, keep = 1L)
gmm_conditionalise(fit, given = c(NA, 0.5))
```

## The unified fitting verb

    fit_proxymix(
      target,
      N        = 1L,
      regime   = c("auto", "moment", "sample", "kld"),
      ...
    )

* `target` is an S7 `gmm_target` produced by `gmm_target()` (from a log-density),
  `gmm_target_from_samples()` (from samples), or one of the built-in factories
  (`banana_target()`, `donut_target()`, `mixture_target()`).
* `regime = "auto"` picks the cheapest applicable regime from the target's
  contents.

## Closed-form mixture operators

    dgmm(x, fit)              # density evaluation
    rgmm(n, fit)              # exact sampling
    gmm_marginalise(fit, keep)
    gmm_conditionalise(fit, given)
    gmm_kld(fit_p, fit_q)     # Monte Carlo + variational bounds

## Mapping the optima of an objective

    fit   <- from_objective(f, lower, upper, N = 10L)  # a mixture map of the optima
    gmm_modes(fit)$modes                               # the distinct optima

`from_objective()` treats an objective `f` as the Gibbs measure
`exp(-f / T)` — a regime-(iii) target you can evaluate but not sample — and
returns a closed-form mixture over its low regions, so a multimodal `f` is
recovered as a whole rather than one optimum at a time. `gmm_modes()`
resolves the fitted map into the recovered optima.

## Which function do I need?

| You have | You want | Reach for |
|---|---|---|
| an evaluable (unnormalised) log-density | a compact closed-form proxy | `gmm_target()` then `fit_proxymix(regime = "kld")` |
| i.i.d. samples | a mixture fit | `gmm_target_from_samples()` then `fit_proxymix()` |
| an objective function | a map of its optima | `from_objective()`, `gmm_modes()` |
| a kernel density estimate | a small closed-form surrogate | `from_kde()` |
| a fitted mixture | marginals, conditionals, updates | `gmm_marginalise()`, `gmm_conditionalise()`, `gmm_observe()`, `gmm_affine()` |
| data with holes | multiple imputation | `gmm_impute()` with `mar()` / `mnar()` / `censored()` |
| a time series + state-space model | filtering / stability testing | `gmm_filter()`, `gmm_eos_test()` |
| a fitted mixture | information-theoretic diagnostics | `gmm_entropy()`, `gmm_divergence()`, `gmm_mutual_information()` |
| a fitted mixture + treatment data | causal / decision quantities | `gmm_intervene()`, `fit_uplift()`, `proxy_cate()` |

## Vignettes

The package ships with eleven vignettes:

* **`quickstart`** — one-page tour.
* **`three_regimes`** — a walk-through of regimes (i)–(iii) on toy 2-D targets, including the agreement of (i) and (iii) at $N = 1$.
* **`density_shapes`** — the regime-(iii) demonstration: banana, donut, three-mixture targets fit by importance-sampled KLD-EM.
* **`operator_calculus`** — closed-form pushforward, Bayesian update, aggregation and conditioning on a fitted mixture.
* **`from_kde`** — compressing a kernel density estimate into a Gaussian-mixture proxy.
* **`many_methods`** — one fitted mixture in place of regression, clustering, PCA and ridge regression.
* **`entropy`** — closed-form entropy, divergence and mutual-information diagnostics.
* **`calibration`** — mapping the optima of an objective via its Gibbs measure.
* **`missing_data`** — multiple imputation by conditioning the fitted mixture (missing at random).
* **`missing_data_mnar`** — imputation under value-dependent missingness and censoring, with sensitivity analysis.
* **`end_of_sample`** — testing whether the last few observations of a series are consistent with a fitted state-space model.

## Interactive tutorial

A standalone, single-file HTML primer for `proxymix` is hosted at

> **<https://max578.github.io/proxymix-tutorial/>**

The page is targeted at Adelaide University Mathematics and Statistics undergraduates (Guided depth, default) and postgraduate / PhD readers (Technical depth, via the in-page dial). It runs in the browser with no R install required, and works the same way offline by double-click. Source at [`max578/proxymix-tutorial`](https://github.com/max578/proxymix-tutorial).

## Reference

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate Gaussians.* Stochastic Analysis and Applications. [doi:10.1080/07362994.2024.2372605](https://doi.org/10.1080/07362994.2024.2372605).

## Licence

MIT © Max Moldovan.
