---
title: "Mapping the optima of an objective"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping the optima of an objective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

A calibration problem has landed on the desk of someone who has run
optimisers before and does not trust them here. The loss surface has more
than one basin, and every run from a different starting point lands
somewhere else. The question is: *my objective has more than one minimum,
so how do I get all of them, and some sense of which basins are wide,
instead of whichever one a local optimiser happened to fall into?*

The answer this vignette gives is that the whole set of optima is a density,
and a density is something the package already knows how to fit.

## What

An objective $f(x)$ to be minimised over a box defines an unnormalised
density, the Gibbs measure $\exp(-f(x) / T)$, whose mass concentrates on the
low regions of $f$ as the temperature $T$ falls. That density can be
evaluated point-wise but not directly sampled, which is exactly regime (iii)
of van der Hoek and Elliott (2024).

[`from_objective()`] fits a Gaussian-mixture proxy to $\exp(-f / T)$ by
cooling a short temperature ladder through the same importance-sampled
KLD-EM that [`fit_kld_em()`] runs, warm-starting each step from the previous
fit. Its `lower` and `upper` arguments are the search region: they set where
the optima are sought, and they scale the temperature ladder, the uniform
exploration of the proposal, and the initialisation. The objective may still
be evaluated outside the box during fitting, where non-finite values are
treated as a finite penalty. Setting `minimise = FALSE` concentrates the map
on the maxima of $f$ instead of the minima.

[`gmm_modes()`] reads the distinct optima off the fitted map. It returns the
mode locations, the mixture density at each one, and how many distinct modes
survived deduplication. The density at a mode is the height of the map over
that basin, and ranking the basins by it is the "which basins are wide" half
of the question: a basin the Gibbs measure concentrates on stands higher
than one it barely notices.

[`ess_summary()`] reports the importance-sampling diagnostics that travel
with the fit, so the map can be judged before it is trusted.

The point of the construction is not a single best point but the whole
picture. A multimodal $f$ gives a multimodal Gibbs measure, so the fitted
mixture spreads its components across the basins and recovers the optima
*together*. The result is an ordinary `gmm`, so the closed-form operator
calculus -- marginalisation, conditioning, divergence -- applies to the map
of solutions like any other fitted mixture.

## Do

### A bimodal objective in one dimension

The objective $(\theta^2 - 4)^2$ has two minima, at $\theta = -2$ and
$\theta = 2$. Fit the map and resolve it.

```{r bimodal}
set.seed(20260619)

f <- function(v) (v[1]^2 - 4)^2

fit <- from_objective(f, lower = -5, upper = 5, N = 6L,
                      is_size = 3000L, n_steps = 5L, seed = 1L)

modes <- gmm_modes(fit)
sort(round(modes$modes[, 1], 3))
```

The fitted object is a `gmm_fit`, so the importance-sampling diagnostics are
available alongside the map.

```{r bimodal-ess}
ess_1d <- ess_summary(fit)
c(ess = round(ess_1d$ess, 1), is_size = ess_1d$is_size,
  ess_relative = round(ess_1d$ess_relative, 3))
```

### Four minima at once: Himmelblau

Himmelblau's function has four equal minima. A single-point optimiser would
return one of them and miss the rest. `from_objective()` returns all four in
one fit, provided the number of components `N` is comfortably larger than the
number of optima so that a component can settle on each basin.

```{r himmelblau}
himmelblau <- function(v) {
  x <- v[1]
  y <- v[2]
  (x * x + y - 11)^2 + (x + y * y - 7)^2
}

fit2 <- from_objective(himmelblau, lower = c(-5, -5), upper = c(5, 5),
                       N = 10L, is_size = 4000L, n_steps = 6L, seed = 1L)

found <- gmm_modes(fit2)
found$n
```

Himmelblau's optima are known in closed form, so the recovered map can be
checked against them rather than against itself. Each recovered mode is
matched to its nearest analytic optimum, and the distance between the two is
the error the reader should judge the fit on.

```{r match}
truth <- rbind(c(3, 2), c(-2.805118, 3.131312),
               c(-3.779310, -3.283186), c(3.584428, -1.848126))

pair_dist <- as.matrix(dist(rbind(found$modes, truth)))
n_found <- nrow(found$modes)
pair_dist <- pair_dist[seq_len(n_found), n_found + seq_len(nrow(truth))]
nearest <- apply(pair_dist, 1L, which.min)
gap <- apply(pair_dist, 1L, min)

all_distinct <- length(unique(nearest)) == nrow(truth)
worst_gap <- max(gap)
c(modes_found = found$n, one_per_optimum = all_distinct,
  worst_gap = round(worst_gap, 3))
```

```{r match-table, echo = FALSE}
knitr::kable(
  data.frame(
    recovered_x1 = round(found$modes[, 1], 3),
    recovered_x2 = round(found$modes[, 2], 3),
    analytic_x1 = round(truth[nearest, 1], 3),
    analytic_x2 = round(truth[nearest, 2], 3),
    distance = round(gap, 3),
    mixture_density = round(found$density, 3)
  ),
  col.names = c("Recovered $x_1$", "Recovered $x_2$", "Analytic $x_1$",
                "Analytic $x_2$", "Distance", "Mixture density"),
  caption = paste0(
    "Each recovered mode beside the analytic optimum nearest to it, with ",
    "the distance between them and the mixture density at the mode."
  )
)
```

The figure below puts the recovered modes and the analytic truth on the
log-objective surface, so a basin that was missed would be visible as a
valley with no orange point in it.

```{r himmelblau-map, eval = has_ggplot2, echo = has_ggplot2, fig.height = 5.4, fig.cap = "The recovered modes sit on the four analytic optima of Himmelblau's function, one per basin of the log-objective surface. No basin is left without a recovered mode.", fig.alt = "Log-objective surface of Himmelblau's function as a viridis raster with contour lines, the four recovered modes as orange circles and the analytic optima as black crosses, each pair coinciding."}
xs <- seq(-5, 5, length.out = 140L)
ys <- seq(-5, 5, length.out = 140L)
surface <- expand.grid(x1 = xs, x2 = ys)
surface$log_f <- log1p(
  apply(as.matrix(surface[, c("x1", "x2")]), 1L, himmelblau)
)

mode_df <- data.frame(x1 = found$modes[, 1], x2 = found$modes[, 2])
truth_df <- data.frame(x1 = truth[, 1], x2 = truth[, 2])

ggplot2::ggplot(surface, ggplot2::aes(x1, x2)) +
  ggplot2::geom_raster(ggplot2::aes(fill = log_f), interpolate = TRUE) +
  ggplot2::geom_contour(ggplot2::aes(z = log_f), colour = "white",
                        linewidth = 0.2, alpha = 0.6, bins = 8L) +
  ggplot2::geom_point(data = truth_df, shape = 4, size = 4, stroke = 1.4,
                      colour = "#000000") +
  ggplot2::geom_point(data = mode_df, shape = 21, size = 3, stroke = 1,
                      fill = "#D55E00", colour = "#000000") +
  ggplot2::scale_fill_viridis_c(name = "log(1 + f)", option = "mako",
                                direction = -1) +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::labs(
    title = "One fit, four basins: the map over Himmelblau's minima",
    subtitle = "crosses: analytic optima; filled circles: recovered modes",
    x = expression(x[1]), y = expression(x[2])
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 panel.grid = ggplot2::element_blank())
```

```{r himmelblau-map-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the log-objective surface",
    "figure is skipped.\n")
```

## Read

Both minima of the one-dimensional objective are recovered. The resolved
modes are
`r paste(sprintf("%.3f", sort(modes$modes[, 1])), collapse = " and ")`,
against the analytic $-2$ and $2$, a largest departure of
`r signif(max(abs(abs(modes$modes[, 1]) - 2)), 2)`. The importance sampler
behind that map was healthy, carrying an effective sample size of
`r round(ess_1d$ess)` out of `r ess_1d$is_size` draws, or
`r round(100 * ess_1d$ess_relative)` per cent of the nominal sample.

On Himmelblau, one fit returns `r found$n` distinct modes, and each is
matched to a *different* analytic optimum, so no optimum is claimed twice
and none is missed. The largest distance between a recovered mode and the
optimum it sits on is `r signif(worst_gap, 2)`, on a search box ten units
wide, so each mode locates its basin to well under a fiftieth of the
domain. The map's height over the four basins runs from
`r signif(min(found$density), 3)` to `r signif(max(found$density), 3)`, a
spread of about
`r round(100 * (max(found$density) / min(found$density) - 1))` per cent
between the lowest and the highest peak. Himmelblau's four optima are equal
in objective value, so a perfectly resolved map would put the same height on
each; the spread here is what a finite importance sample and ten components
covering four basins leave behind, and it is the quantity to watch when the
optima are *not* known to be equal.

That answers the question. The optima come out together rather than one at a
time, and the height of the map at each mode ranks the basins by how sharply
the Gibbs measure concentrates on them -- a ranking a single-point optimiser
cannot produce, because it never sees more than one basin.

## Limits

Recovery is most reliable when the number of components `N` is larger than
the number of optima, because a component has to be free to settle on each
basin. Symmetric landscapes, where several optima are exchangeable, need the
most headroom; when in doubt, raise `N`. This vignette used ten components
for four optima and six for two, and does not establish what happens when
the headroom is removed.

The box is a modelling choice and not a formality. It sets where the optima
are sought and it scales the temperature ladder, the exploration of the
proposal, and the initialisation, so an optimum outside the box is not
merely harder to find, it is outside the question being asked. Both examples
here were given boxes that comfortably contain every optimum, which is the
easy case.

Neither example is high-dimensional. Regime (iii) is driven by importance
sampling, whose effective sample size falls sharply as the dimension grows,
and the two objectives fitted here are one- and two-dimensional. The
effective sample size reported alongside every fit is the diagnostic to
watch, and a map fitted at a low effective sample size should not be read as
a reliable inventory of basins.

Finally, the height of the map at a mode is a property of the Gibbs measure
at the final temperature, not of the objective. It is not the objective
value, and it is not a basin mass either: a wide shallow basin can carry more
total mass than a narrow deep one while standing lower at its peak. The
objective itself remains available and should be evaluated at the recovered
modes when what matters is which optimum is lowest rather than which the map
concentrates on.

## What to read next

*Choosing between the three fitting regimes* explains why an objective that
can be evaluated but not sampled lands in regime (iii), and what the other
two regimes do instead. *Fitting a proxy to a density you cannot sample* is
the shortest route into the same fitting verb on a target rather than an
objective. *The closed-form operator calculus on a mixture* covers the
marginalisation, conditioning and divergence operators that apply to the map
of solutions once it has been fitted, since the map is an ordinary mixture.

## References

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
Gaussians.* Stochastic Analysis and Applications.
[doi:10.1080/07362994.2024.2372605](https://doi.org/10.1080/07362994.2024.2372605).

Carreira-Perpinan, M. A. (2000). Mode-finding for mixtures of Gaussian
distributions. *IEEE Transactions on Pattern Analysis and Machine
Intelligence*, 22(11), 1318-1323.

## Reproduce

The ambient seed is 20260619, and both fits additionally pass `seed = 1L`
to `from_objective()`, so the importance draws are reproducible
independently of the ambient random-number state.

```{r session-info}
sessionInfo()
```
