---
title: "Mapping the optima of an objective"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping the optima of an objective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  out.width = "100%"
)
set.seed(20260619)
```

```{r setup}
library(proxymix)
```

# An objective is a density you can evaluate but not sample

An objective $f(x)$ to be minimised over a box defines an unnormalised
density, the Gibbs measure $\exp(-f(x) / T)$, whose mass concentrates on the
low regions of $f$ as the temperature $T$ falls. That density can be
evaluated point-wise but not directly sampled, which is exactly regime (iii)
of van der Hoek and Elliott (2024). `from_objective()` fits a
Gaussian-mixture proxy to $\exp(-f / T)$ by cooling a short temperature
ladder through the same importance-sampled KLD-EM that `fit_kld_em()` runs,
warm-starting each step from the previous fit.

The point of the construction is not a single best point but the whole
picture. A multimodal $f$ gives a multimodal Gibbs measure, so the fitted
mixture spreads its components across the basins and recovers the optima
*together*. The result is an ordinary `gmm`, so the closed-form operator
calculus -- marginalisation, conditioning, divergence -- applies to the map
of solutions. `gmm_modes()` reads the distinct optima off that map.

# A bimodal objective in one dimension

The objective $(\theta^2 - 4)^2$ has two minima, at $\theta = -2$ and
$\theta = 2$. We fit the map and resolve it.

```{r bimodal}
f <- function(v) (v[1]^2 - 4)^2

fit <- from_objective(f, lower = -5, upper = 5, N = 6L,
                      is_size = 3000L, n_steps = 5L, seed = 1L)

modes <- gmm_modes(fit)
sort(round(modes$modes[, 1], 3))
```

Both minima are recovered. The fitted object is a `gmm_fit`, so the
importance-sampling diagnostics are available alongside the map.

```{r bimodal-ess}
ess_summary(fit)
```

# Four minima at once: Himmelblau

Himmelblau's function has four equal minima. A single-point optimiser would
return one of them and miss the rest. `from_objective()` returns all four in
one fit, provided the number of components `N` is comfortably larger than the
number of optima so that a component can settle on each basin.

```{r himmelblau}
himmelblau <- function(v) {
  x <- v[1]
  y <- v[2]
  (x * x + y - 11)^2 + (x + y * y - 7)^2
}

fit2 <- from_objective(himmelblau, lower = c(-5, -5), upper = c(5, 5),
                       N = 10L, is_size = 4000L, n_steps = 6L, seed = 1L)

found <- gmm_modes(fit2)
round(found$modes[order(found$modes[, 1]), ], 3)
```

The four recovered minima sit at $(3, 2)$, $(-2.81, 3.13)$,
$(-3.78, -3.28)$ and $(3.58, -1.85)$ -- the known optima. The figure below
overlays the recovered modes (orange) and the analytic truth (red crosses)
on the log-objective surface.

```{r himmelblau-map, fig.cap = "The recovered modes sit on the four analytic optima of Himmelblau's function, one per basin of the log-objective surface.", fig.alt = "Log-objective surface of Himmelblau's function with the four recovered modes and the analytic optima overlaid."}
truth <- rbind(c(3, 2), c(-2.805118, 3.131312),
               c(-3.779310, -3.283186), c(3.584428, -1.848126))

xs <- seq(-5, 5, length.out = 140)
ys <- seq(-5, 5, length.out = 140)
zz <- outer(xs, ys, Vectorize(function(a, b) himmelblau(c(a, b))))

image(xs, ys, log1p(zz), col = hcl.colors(40, "YlGnBu", rev = TRUE),
      xlab = expression(x[1]), ylab = expression(x[2]),
      main = "Himmelblau: the map over the four minima")
contour(xs, ys, log1p(zz), add = TRUE, col = "grey60",
        nlevels = 8, drawlabels = FALSE)
points(truth[, 1], truth[, 2], pch = 4, col = "red", cex = 2, lwd = 3)
points(found$modes[, 1], found$modes[, 2], pch = 19,
       col = "#D95F0E", cex = 1.4)
legend("topright", bg = "white",
       legend = c("analytic optima", "recovered modes"),
       pch = c(4, 19), col = c("red", "#D95F0E"),
       pt.cex = c(1.4, 1.2), pt.lwd = c(3, 1))
```

The orange points sit on all four red crosses: `from_objective()` located
every basin in one fit rather than converging on whichever optimum the
starting point happened to favour.

# Notes on use

* **Component headroom.** Recovery is most reliable when `N` is larger than
  the number of optima. Symmetric landscapes, where several optima are
  exchangeable, need the most headroom. When in doubt raise `N`.
* **The box is the search region.** `lower` and `upper` set where the optima
  are sought and scale the temperature ladder, the uniform exploration of
  the proposal, and the initialisation. The objective may be evaluated
  outside the box during fitting, where non-finite values are treated as a
  finite penalty.
* **Maximisation.** Set `minimise = FALSE` to concentrate the map on the
  maxima of `f` instead of the minima.
* **The map is a mixture.** Because the return value is a `gmm`, the rest of
  the package applies. The optima can be conditioned, marginalised, or
  compared with `gmm_divergence()` like any other fitted mixture.

# References

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
Gaussians.* Stochastic Analysis and Applications.
[doi:10.1080/07362994.2024.2372605](https://doi.org/10.1080/07362994.2024.2372605).

Carreira-Perpinan, M. A. (2000). Mode-finding for mixtures of Gaussian
distributions. *IEEE Transactions on Pattern Analysis and Machine
Intelligence*, 22(11), 1318-1323.
