---
title: "KLD-EM on three density shapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{KLD-EM on three density shapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
set.seed(20260513)
```

```{r setup}
library(proxymix)
```

This vignette demonstrates regime (iii) across target shapes. The KLD-EM regime fits a Gaussian-mixture proxy to a target whose density we can *evaluate* but cannot easily *sample from*. Existing CRAN packages for Gaussian mixtures — `mclust`, `mixtools`, `flexmix` — all assume i.i.d. samples are available.

We exercise three quite different non-Gaussian target shapes:

* the **banana** — a Rosenbrock-style 2-D bent ridge;
* the **donut** — a rotationally symmetric annulus;
* the **three-mixture** — well-separated 2-D Gaussian components.

For each, we

1. plot the target;
2. fit by `fit_proxymix(regime = "kld")` with a `df = 5` multivariate-t proposal;
3. report the importance-sample effective size, the final IS-estimated KLD, and a Monte-Carlo Hellinger sanity check;
4. overlay the proxy on the target.

## Banana

```{r banana-target, fig.cap = "Banana target."}
banana <- banana_target()

grid_x <- seq(-3.5, 3.5, length.out = 120)
G_b <- expand.grid(x1 = grid_x, x2 = grid_x)
G_b$f <- exp(banana@log_density(as.matrix(G_b)))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(G_b, aes(x1, x2, z = f)) +
    geom_contour_filled(bins = 12L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Banana target",
         x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 11)
}
```

```{r banana-fit}
fit_b <- fit_proxymix(banana, N = 4L, regime = "kld",
                      proposal = is_mvt(n_dim = 2L,
                                        sigma = 4 * diag(2),
                                        df = 5),
                      is_size = 3000L,
                      max_iter = 50L,
                      seed = 1L)
fit_b
```

```{r banana-overlay, fig.cap = "Banana target (filled) with 4-component proxy (dashed)."}
G_b$g <- exp(dgmm(as.matrix(G_b[, c("x1", "x2")]), fit_b, log = TRUE))
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot(G_b, aes(x1, x2)) +
    geom_contour_filled(aes(z = f), bins = 10L, alpha = 0.85) +
    geom_contour(aes(z = g), colour = "white", linetype = "dashed",
                 linewidth = 0.4, bins = 8L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Banana with proxy",
         x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 11)
}
```

## Donut

```{r donut-target, fig.cap = "Donut target (rotationally symmetric annulus)."}
donut <- donut_target()

grid_x <- seq(-4, 4, length.out = 120)
G_d <- expand.grid(x1 = grid_x, x2 = grid_x)
G_d$f <- exp(donut@log_density(as.matrix(G_d)))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot(G_d, aes(x1, x2, z = f)) +
    geom_contour_filled(bins = 12L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Donut target",
         x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 11)
}
```

A donut is genuinely hard for a Gaussian mixture — the modes are *infinitely many* (along the ring). With `N = 6` components, KLD-EM finds a balanced ring-approximation:

```{r donut-fit}
fit_d <- fit_proxymix(donut, N = 6L, regime = "kld",
                      proposal = is_mvt(n_dim = 2L,
                                        sigma = 9 * diag(2),
                                        df = 5),
                      is_size = 3500L,
                      max_iter = 60L,
                      seed = 1L)
fit_d
```

```{r donut-overlay, fig.cap = "Donut target (filled) with 6-component proxy (dashed)."}
G_d$g <- exp(dgmm(as.matrix(G_d[, c("x1", "x2")]), fit_d, log = TRUE))
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot(G_d, aes(x1, x2)) +
    geom_contour_filled(aes(z = f), bins = 10L, alpha = 0.85) +
    geom_contour(aes(z = g), colour = "white", linetype = "dashed",
                 linewidth = 0.4, bins = 8L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Donut with proxy",
         x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 11)
}
```

## Three-mixture

```{r mixture-target, fig.cap = "Three-component mixture target."}
mt <- mixture_target()

grid_x <- seq(-4.5, 4.5, length.out = 120)
G_m <- expand.grid(x1 = grid_x, x2 = grid_x)
G_m$f <- exp(mt@log_density(as.matrix(G_m)))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot(G_m, aes(x1, x2, z = f)) +
    geom_contour_filled(bins = 12L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Three-mixture target",
         x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 11)
}
```

```{r mixture-fit}
fit_m <- fit_proxymix(mt, N = 3L, regime = "kld",
                      proposal = is_mvt(n_dim = 2L,
                                        sigma = 6 * diag(2),
                                        df = 5),
                      is_size = 3000L,
                      max_iter = 50L,
                      seed = 1L)
fit_m
```

```{r mixture-overlay, fig.cap = "Three-mixture target (filled) with 3-component proxy (dashed)."}
G_m$g <- exp(dgmm(as.matrix(G_m[, c("x1", "x2")]), fit_m, log = TRUE))
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot(G_m, aes(x1, x2)) +
    geom_contour_filled(aes(z = f), bins = 10L, alpha = 0.85) +
    geom_contour(aes(z = g), colour = "white", linetype = "dashed",
                 linewidth = 0.4, bins = 8L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Three-mixture with proxy",
         x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 11)
}
```

## Bounded support

The three targets above all live on the whole plane. Some targets are instead **compactly supported** — exactly zero outside a bounded region. The textbook example is the Epanechnikov density \(K(u) = \tfrac{3}{4}(1 - u^2)\) on \([-1, 1]\). No mixture of full-support Gaussians can be exactly zero on an interval, so regime (iii) is the only viable route — and the default \(t\)-proposal is actively unsafe here, since it places importance draws where the target log-density is \(-\infty\), producing non-finite weights.

`proxymix` handles this safely. A target can declare its `support`, and `fit_kld_em()` then selects a *support-matched* `is_uniform()` proposal automatically instead of the crashing default. `epanechnikov_target()` is a built-in compact-support fixture that declares its support:

```{r epan-target}
epan <- epanechnikov_target(n_dim = 1L)
epan
```

Fitting needs no hand-chosen proposal — the declared support drives the choice, announced in a one-line message:

```{r epan-fit}
fit_e <- fit_proxymix(epan, N = 3L, regime = "kld",
                      is_size = 4000L, max_iter = 60L, seed = 1L)
fit_e@diagnostics$proposal_name
c(ess = round(fit_e@diagnostics$ess, 1L),
  support_fraction = fit_e@diagnostics$support_fraction,
  kld_final = round(fit_e@diagnostics$kld_final, 4L))
```

Every importance draw receives a finite weight (the support fraction is 1), and the three-component proxy tracks the inverted-parabola shape closely:

```{r epan-overlay, fig.cap = "Epanechnikov target (solid) with 3-component proxy (dashed)."}
xs <- seq(-1.4, 1.4, length.out = 400L)
df_e <- data.frame(
  x = xs,
  f = exp(epan@log_density(matrix(xs, ncol = 1L))),
  g = exp(dgmm(matrix(xs, ncol = 1L), fit_e, log = TRUE))
)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot(df_e, aes(x)) +
    geom_line(aes(y = f), linewidth = 0.8) +
    geom_line(aes(y = g), linetype = "dashed", linewidth = 0.6,
              colour = "#2c7fb8") +
    labs(title = "Epanechnikov with proxy", y = "density") +
    theme_minimal(base_size = 11)
}
```

## Summary

```{r summary}
summary_tbl <- data.frame(
  target      = c("banana", "donut", "three-mixture"),
  N           = c(gmm_n_components(fit_b),
                  gmm_n_components(fit_d),
                  gmm_n_components(fit_m)),
  is_size     = c(fit_b@diagnostics$is_size,
                  fit_d@diagnostics$is_size,
                  fit_m@diagnostics$is_size),
  ESS         = round(c(fit_b@diagnostics$ess,
                        fit_d@diagnostics$ess,
                        fit_m@diagnostics$ess), 1L),
  kld_final   = round(c(fit_b@diagnostics$kld_final,
                        fit_d@diagnostics$kld_final,
                        fit_m@diagnostics$kld_final), 4L),
  hellinger2  = round(c(
    hellinger_mc(fit_b, n_mc = 2000L, seed = 1L)$h2,
    hellinger_mc(fit_d, n_mc = 2000L, seed = 1L)$h2,
    hellinger_mc(fit_m, n_mc = 2000L, seed = 1L)$h2
  ), 4L)
)
summary_tbl
```

Interpretation:

* **`ESS`** — the effective sample size of the importance-sampling weights, out of `is_size`. ESS well below `is_size` means the proposal is wasteful (heavy weight on few draws); ESS near `is_size` means the proposal is close to the target.
* **`kld_final`** — the importance-sampled estimate of \(\mathrm{KL}(f \Vert g_\theta)\) at the final parameters. *Small is good.*
* **`hellinger2`** — a Monte-Carlo estimate of the squared Hellinger distance \(H^2(f, g)\). Bounded in \([0, 1]\) when both densities are normalised; a small value confirms the proxy reproduces the target's mass *and* shape.

The three-mixture is fit nearly exactly (the proxy lives in the same parametric family as the target). The banana is well-approximated by four Gaussians along the ridge. The donut — a topology a Gaussian mixture cannot reproduce exactly — is approximated by a ring of 6 components, with a non-trivial residual KLD reflecting the structural mismatch.

## Reference

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate Gaussians.* Stochastic Analysis and Applications. <https://doi.org/10.1080/07362994.2024.2372605>.
