---
title: "How well a mixture proxies four awkward shapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How well a mixture proxies four awkward shapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~3 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

Someone with a genuinely awkward target asks the question the routing table
does not answer: *my density is nothing like a Gaussian, and I can only
evaluate it. Where does a mixture proxy actually work, and where does it
stop?* The regime is settled before the question is asked. What is not
settled is whether the answer that comes back deserves to be used.

## What

Four bundled targets stand in for four kinds of awkwardness.
`banana_target()` is a Rosenbrock-style bent ridge, curved but simply
connected. `donut_target()` is a rotationally symmetric annulus, whose
modes form a continuous ring rather than a finite set of points.
`mixture_target()` is three well-separated Gaussian components, which is
awkward only in being multimodal. `epanechnikov_target()` is compactly
supported: exactly zero outside a bounded interval, which no mixture of
full-support Gaussians can reproduce.

`fit_kld_em()` fits all four, through `fit_proxymix(regime = "kld")`. Its
proposal argument is what the fit feeds on, and `proposal_mvt()` supplies
the Student-t default. A target that declares a `support` causes the fit to
select a support-matched `proposal_uniform()` instead, announced in a
message, because a Student-t would place draws where the target's
log-density is minus infinity.

Three diagnostics grade the result. `ess_summary()` and the fit's own
`ess` report how many of the importance draws are effectively carrying the
answer. The `kld_final` diagnostic is the importance-sampled estimate of
\(\mathrm{KL}(f \Vert g_\theta)\) at the fitted parameters, and it comes
with a Monte Carlo standard error. `hellinger_mc()` estimates the squared
Hellinger distance \(H^2(f, g)\) by Monte Carlo, which is bounded in
\([0, 1]\) when both densities are normalised.

`gmm_fit_quality()` carries the refusal contract. When the importance
sample collapses, the fit does not return a mixture that looks ordinary
and is worthless; under `on_low_ess = "abort"` it raises a typed condition
instead.

## Do

### Fitting the three plane-supported shapes

Each target gets a Student-t proposal wide enough to cover it, and a
component count chosen for its shape: four along the banana's ridge, six
around the donut's ring, three for the three-mode mixture.

```{r fits}
fit_b <- fit_proxymix(banana_target(), N = 4L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              sigma = 4 * diag(2),
                                              df = 5),
                      is_size = 3000L, max_iter = 50L, seed = 1L)
fit_d <- fit_proxymix(donut_target(), N = 6L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              sigma = 9 * diag(2),
                                              df = 5),
                      is_size = 3500L, max_iter = 60L, seed = 1L)
fit_m <- fit_proxymix(mixture_target(), N = 3L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              sigma = 6 * diag(2),
                                              df = 5),
                      is_size = 3000L, max_iter = 50L, seed = 1L)
```

```{r overlay-grid}
shape_panel <- function(target, fit, label, lim) {
  gx <- seq(-lim, lim, length.out = 110L)
  base <- expand.grid(x1 = gx, x2 = gx)
  gm <- as.matrix(base)
  data.frame(
    x1 = base$x1, x2 = base$x2,
    target = exp(target@log_density(gm)),
    proxy = dgmm(gm, fit),
    shape = label,
    stringsAsFactors = FALSE
  )
}
overlay_df <- rbind(
  shape_panel(banana_target(), fit_b, "banana, N = 4", 3.5),
  shape_panel(donut_target(), fit_d, "donut, N = 6", 4.0),
  shape_panel(mixture_target(), fit_m, "three-mixture, N = 3", 4.5)
)
overlay_df$shape <- factor(overlay_df$shape,
                           levels = unique(overlay_df$shape))
```

```{r overlay, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.4, fig.cap = "Each target (filled contours) with its fitted proxy overlaid as dashed contours. The banana and the three-mixture are tracked closely; the donut's proxy is a ring of six ellipses approximating a continuous annulus, and the dashed contours break into lobes where the ellipses meet.", fig.alt = "Three contour panels showing the banana, donut and three-mixture densities, each overlaid with dashed contours of its Gaussian-mixture proxy."}
ggplot2::ggplot(overlay_df, ggplot2::aes(x1, x2)) +
  ggplot2::geom_contour_filled(ggplot2::aes(z = target), bins = 10L,
                               alpha = 0.85) +
  ggplot2::geom_contour(ggplot2::aes(z = proxy), colour = "white",
                        linetype = "dashed", linewidth = 0.4, bins = 8L) +
  ggplot2::scale_fill_viridis_d(option = "mako", guide = "none") +
  ggplot2::facet_wrap(~ shape, scales = "free") +
  ggplot2::labs(x = expression(x[1]), y = expression(x[2])) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r overlay-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the three-shape overlay",
    "figure is skipped.\n")
```

### Grading the three fits

The importance-sampled `kld_final` is an estimate and carries Monte Carlo
error, so a second, deterministic measurement is worth having. In two
dimensions the divergence can be integrated directly on a grid, using
nothing the fit produced.

```{r kl-grid}
## KL(f || g) by deterministic grid quadrature -- independent of the
## importance sampler whose estimate it is checking.
kl_grid <- function(target, fit, lim, n_grid = 300L) {
  gx <- seq(-lim, lim, length.out = n_grid)
  gm <- as.matrix(expand.grid(x1 = gx, x2 = gx))
  cell <- (gx[2L] - gx[1L])^2
  log_f <- target@log_density(gm)
  log_g <- dgmm(gm, fit, log = TRUE)
  dens_f <- exp(log_f)
  ok <- is.finite(log_f) & is.finite(log_g) & dens_f > 1e-300
  sum(dens_f[ok] * (log_f[ok] - log_g[ok])) * cell
}
kl_quad <- c(
  banana = kl_grid(banana_target(), fit_b, 6),
  donut = kl_grid(donut_target(), fit_d, 7),
  mixture = kl_grid(mixture_target(), fit_m, 8)
)
```

```{r summary-table, echo = FALSE}
fits <- list(fit_b, fit_d, fit_m)
hell <- lapply(fits, hellinger_mc, n_mc = 2000L, seed = 1L)
summary_tbl <- data.frame(
  Target = c("banana", "donut", "three-mixture"),
  N = vapply(fits, gmm_n_components, integer(1L)),
  `IS draws` = vapply(fits, function(f) f@diagnostics$is_size, numeric(1L)),
  ESS = round(vapply(fits, function(f) f@diagnostics$ess, numeric(1L)), 0L),
  `KL (sampled)` = round(
    vapply(fits, function(f) f@diagnostics$kld_final, numeric(1L)), 4L
  ),
  `Monte Carlo SE` = round(
    vapply(fits, function(f) f@diagnostics$mc_se_kld, numeric(1L)), 4L
  ),
  `KL (quadrature)` = round(unname(kl_quad), 4L),
  `Squared Hellinger` = round(
    vapply(hell, function(h) h$h2, numeric(1L)), 4L
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(
  summary_tbl,
  caption = paste(
    "Fit quality on the three plane-supported shapes. The sampled and",
    "quadrature columns estimate the same Kullback-Leibler divergence,",
    "the first by importance sampling and the second deterministically."
  )
)
```

### How much of the donut's residual is structural

A Gaussian mixture cannot reproduce an annulus exactly, however many
components it is given. Sweeping the component count separates the part of
the residual divergence that more components remove from the part that
they do not.

```{r donut-sweep}
donut_n <- c(3L, 4L, 6L, 10L)
donut_kl <- vapply(donut_n, function(k) {
  f <- fit_proxymix(donut_target(), N = k, regime = "kld",
                    proposal = proposal_mvt(n_dim = 2L,
                                            sigma = 9 * diag(2),
                                            df = 5),
                    is_size = 3500L, max_iter = 60L, seed = 1L)
  kl_grid(donut_target(), f, 7)
}, numeric(1L))
```

```{r donut-sweep-table, echo = FALSE}
knitr::kable(
  data.frame(
    `Components` = donut_n,
    `KL (quadrature)` = round(donut_kl, 4L),
    check.names = FALSE
  ),
  caption = paste(
    "Residual divergence of the donut proxy against the component count,",
    "measured by grid quadrature. The fall from three to six components",
    "is the approximation improving; the flat step from six to ten is the",
    "structural floor a Gaussian mixture cannot cross."
  )
)
```

### Bounded support

The three targets above live on the whole plane. A compactly supported
target is different in kind: the Epanechnikov density
\(K(u) = \tfrac{3}{4}(1 - u^2)\) on \([-1, 1]\) is exactly zero outside the
interval, and the default Student-t proposal is actively unsafe against it
because it places draws where the target's log-density is minus infinity,
producing non-finite weights.

A target may declare its `support`, and the fit then selects a
support-matched `proposal_uniform()` on its own, announced in a message.

```{r epan-target}
epan <- epanechnikov_target(n_dim = 1L)
epan
```

```{r epan-fit}
fit_e <- fit_proxymix(epan, N = 3L, regime = "kld",
                      is_size = 4000L, max_iter = 300L, seed = 1L)
c(proposal = fit_e@diagnostics$proposal_name,
  support_fraction = fit_e@diagnostics$support_fraction,
  iterations = length(kld_trace(fit_e)),
  converged = gmm_fit_quality(fit_e)$converged)
```

```{r epan-grid}
epan_x <- seq(-1.4, 1.4, length.out = 400L)
epan_df <- rbind(
  data.frame(x = epan_x, density = exp(epan@log_density(
    matrix(epan_x, ncol = 1L))), series = "target", stringsAsFactors = FALSE),
  data.frame(x = epan_x, density = dgmm(matrix(epan_x, ncol = 1L), fit_e),
             series = "mixture proxy", stringsAsFactors = FALSE)
)
```

```{r epan-plot, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.4, fig.cap = "The compactly supported Epanechnikov target and its three-component proxy. The proxy tracks the inverted parabola across the interval and leaks a little mass past the boundary at plus and minus one, which is the one thing a mixture of full-support Gaussians can never avoid.", fig.alt = "Line plot of the inverted-parabola Epanechnikov density and its dashed Gaussian-mixture proxy, which extends slightly beyond the support boundary."}
ggplot2::ggplot(epan_df, ggplot2::aes(x, density, colour = series,
                                      linetype = series)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_vline(xintercept = c(-1, 1), colour = "grey60",
                      linewidth = 0.3) +
  ggplot2::scale_colour_manual(
    name = NULL, values = c("target" = "#0072B2",
                            "mixture proxy" = "#D55E00")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL, values = c("target" = "solid", "mixture proxy" = "dashed")
  ) +
  ggplot2::labs(x = expression(x), y = "density") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r epan-plot-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the bounded-support",
    "figure is skipped.\n")
```

### When the proposal misses the target

Every number above rests on the importance sample being informative about
the target. When it is not, the arithmetic still produces a mixture, and
that mixture is worthless. The package treats this as a refusal rather
than a result: a proposal placed far from the target's mass, with
`on_low_ess = "abort"`, raises a typed condition instead of returning.

```{r refusal}
bad <- tryCatch(
  fit_kld_em(
    mixture_target(), N = 3L,
    proposal = proposal_mvn(n_dim = 2L, mean = c(25, 25), cov = diag(2)),
    is_size = 2000L, max_iter = 10L, seed = 1L,
    min_ess = 50, on_low_ess = "abort"
  ),
  error = function(e) e
)
class(bad)[1L]
conditionMessage(bad)
```

## Read

The three plane-supported shapes divide into two groups, and the table says
which. The banana proxy reaches a quadrature divergence of
`r format(signif(kl_quad[["banana"]], 2L))` and the three-mixture
`r format(signif(kl_quad[["mixture"]], 2L))`, from effective sample sizes of
`r round(fit_b@diagnostics$ess)` and `r round(fit_m@diagnostics$ess)` draws
respectively. The donut sits materially higher at
`r format(signif(kl_quad[["donut"]], 2L))`, roughly
`r round(kl_quad[["donut"]] / kl_quad[["banana"]], 1L)` times the banana's,
and that gap is the answer to the question.

The sweep says the gap is structural rather than a matter of budget. Going
from `r donut_n[1L]` to `r donut_n[3L]` components cuts the donut's
divergence from `r format(signif(donut_kl[1L], 3L))` to
`r format(signif(donut_kl[3L], 3L))`, a fall of
`r round(100 * (1 - donut_kl[3L] / donut_kl[1L]))` per cent. Going on to
`r donut_n[4L]` components buys only
`r format(signif(donut_kl[3L] - donut_kl[4L], 2L))` more. Components are
solving the approximation problem right up to the point where they stop,
and what remains is a mixture of finitely many ellipses being asked to be a
continuous ring.

The two estimates of the same divergence disagree in a way worth reading.
The importance-sampled column returns
`r format(signif(fit_d@diagnostics$kld_final, 2L))` for the donut against a
Monte Carlo standard error of
`r format(signif(fit_d@diagnostics$mc_se_kld, 2L))`, while the quadrature
column returns `r format(signif(kl_quad[["donut"]], 2L))`. The sampled
estimate is within about one standard error of zero and of the quadrature
value alike, so on these fits it is not sharp enough to separate a good
proxy from a structurally limited one. The same holds for the squared
Hellinger column, whose entries sit within a few thousandths of zero. The
deterministic measurement is what makes the donut's floor visible.

The compactly supported target is the fourth kind of awkwardness and the
one that never fully resolves. The declared support routed the fit onto a
support-matched proposal automatically, every draw landed inside the
support, and the fit converged after `r length(kld_trace(fit_e))`
iterations, far more than the tens the plane-supported shapes needed. The
figure shows why the iteration works so hard: the proxy must approximate a
hard boundary with densities that are positive everywhere, so it pays at
the edges no matter how long it runs.

So the answer is that a mixture proxy works well on curvature and on
multimodality, degrades gracefully and measurably on a topology it cannot
represent, and on a bounded support converges slowly to a fit that always
leaks a little mass past the boundary. And when the importance sample
carries no information at all, nothing is returned: the refusal above is a
`r class(bad)[1L]` condition, raised in place of a fit whose printed
summary would have looked entirely ordinary. The default is
`on_low_ess = "warn"`, which downgrades the same check to a warning and
lets the caller decide; `"abort"` is what turns it into a contract.

## Limits

The two divergence estimates in the table are measurements, not exact
quantities, and neither is a certificate on its own. The importance-sampled
estimate is a Monte Carlo quantity that can come out slightly negative on a
near-exact fit, as it does for two of the three shapes here, and a
negative estimate of a non-negative divergence means only that the
estimator's error exceeds the quantity being estimated. The squared
Hellinger estimates behave the same way, and `hellinger_mc()` does not
always return a usable standard error alongside them. The grid quadrature
avoids both problems and introduces a different one: it is reliable in two
or three dimensions and becomes intractable well before the dimension at
which importance sampling itself gives out.

The component counts here were chosen by hand, one per shape, to make the
comparison across shapes a comparison of shapes. Choosing the count is a
separate problem, and the donut sweep is the only place this vignette
touches it.

Nothing here establishes that the mixture proxy is the right tool for the
downstream job, only that it reproduces these densities to a measured
accuracy. Whether a divergence of the size found for the donut matters
depends entirely on what is computed from the proxy afterwards, and a
quantity that depends on the tails will be far less forgiving than one that
depends on the bulk.

## What to read next

*Fitting a proxy to a density you cannot sample* is the shorter introduction
to the regime this vignette stress-tests, and reads the fit certificate in
full.

*Choosing between the three fitting regimes* explains why regime (iii) is
the only option for three of the four targets here, and what it costs when
the other regimes are available.

*Reading the entropy of a fitted mixture* takes up the component-count
question the donut sweep only touches, and discovers a count by cooling
rather than by sweeping.

*The closed-form operator calculus on a mixture* is what the proxies fitted
here are for: the exact operations that follow once the shape has been
captured.

## References

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
Gaussians.* Stochastic Analysis and Applications.
<https://doi.org/10.1080/07362994.2024.2372605>.

## Reproduce

Every fit above is seeded with `seed = 1L`, the Monte Carlo Hellinger
estimates with `seed = 1L`, and the grid quadrature is deterministic, so
the whole vignette is reproducible from the package alone.

```{r session-info}
sessionInfo()
```
