---
title: "Testing the last observation for instability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing the last observation for instability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~13 s under rmarkdown::render() with ggplot2 installed,
     dominated by the two size studies; macOS arm64 (Apple silicon),
     R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

An analyst monitoring a series has a model fitted to everything up to
yesterday, and this morning one more observation arrived. The question is
narrow and awkward: *one more observation has just arrived -- is it
consistent with the model I fitted, or has something just changed?* It is
awkward because the suspect window is tiny. With a single new point, or a
handful, the usual structural-break tests (Chow, sup-Wald) are undefined:
the parameters of the post-break regime cannot be estimated from so few
observations. [`gmm_eos_test()`] answers the question in exactly this
regime, following Andrews (2003).

## What

[`gmm_eos_test()`] runs the Gaussian-sum filter implied by the operator
calculus over the series, scores the last `m` one-step filter innovations,
and calibrates that score. It takes the model exactly as
[`gmm_filter()`] does: a single-component [`gmm()`] prior, a `dynamics`
list carrying the state-transition and process-noise matrices, and a
`measurement` list carrying the observation and observation-noise matrices.

The statistic is built from the filter's own output. At each step the filter
returns a one-step predictive mean and an innovation covariance $S_t$, so
the standardised innovation
$z_t = S_t^{-1/2}(y_t - C\hat{x}_{t\mid t-1})$ is standard normal under a
correct, stable, linear-Gaussian model. The statistic sums the last `m`
squared standardised innovations,
$$ \mathrm{EoS}_m = \sum_{t = n-m+1}^{n} z_t^{\top} z_t, $$
which a break inflates.

Two calibrations are offered. `method = "chisq"` refers the statistic to a
$\chi^2$ distribution on `m` times the observation dimension degrees of
freedom, exact when the innovations are Gaussian and the model is given.
`method = "andrews"` is the distribution-free subsampling test of Andrews
(2003): the statistic's rank among the in-sample overlapping `m`-blocks of
the same innovations gives the p-value, which stays calibrated when the
innovations are not Gaussian. The returned object also carries those
in-sample block statistics, so the calibration can be inspected directly.

## Do

```{r seed}
set.seed(20260621)
```

### The model and a stable series

A local-level model: a random walk observed with noise.

```{r model}
prior <- gmm(weights = 1, means = list(0), covariances = list(matrix(10)))
dynamics <- list(A = matrix(1), Q = matrix(0.04))  # the state random walk
measurement <- list(C = matrix(1), R = matrix(1))  # the noisy observation

n <- 120L
level <- cumsum(c(0, rnorm(n - 1L, 0, sqrt(0.04))))
y_stable <- level + rnorm(n, 0, 1)
```

### The same series with a broken final observation

Copy the series and push the final observation away by five standard
deviations of the observation noise.

```{r break}
y_break <- y_stable
y_break[n] <- y_break[n] + 5
```

```{r tests}
test_stable <- gmm_eos_test(
  prior, dynamics, measurement, y_stable, m = 1L, method = "andrews"
)
test_break <- gmm_eos_test(
  prior, dynamics, measurement, y_break, m = 1L, method = "andrews"
)
```

The returned object is a list whose `in_sample_blocks` element holds one
number per in-sample block, so the six scalar fields are tabulated rather
than printed.

```{r tests-kable, echo = FALSE}
eos_row <- function(x) {
  c(
    round(x$statistic, 3L), round(x$p_value, 4L), x$reject,
    x$method, x$m, x$df
  )
}
knitr::kable(
  data.frame(
    field = c("statistic", "p-value", "reject at 0.05", "calibration",
              "window m", "degrees of freedom"),
    stable = eos_row(test_stable),
    broken = eos_row(test_break)
  ),
  col.names = c("Field", "Stable series", "Broken final point"),
  caption = paste(
    "The end-of-sample test on the same series before and after the",
    "final observation is pushed five observation-noise standard",
    "deviations away."
  )
)
```

### What the filter expected

The filter's own path shows why the broken point is surprising. The band
below is the filtered state and two of its standard deviations, taken from
`gmm_filter()` on the stable series, with the broken final observation
marked against it.

```{r filter-path}
filtered <- gmm_filter(
  prior, dynamics, measurement, y_stable, ridge_eps = 0
)
path <- filtered$summary
break_gap <- abs(y_break[n] - path$mean_1[n]) / path$sd_1[n]
```

```{r fig-series, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4, fig.cap = "The local-level series, the latent state that generated it, and the filtered state with two standard deviations either side. The broken final observation sits far outside the band the filter had built from the stable history, which is what inflates the statistic.", fig.alt = "A time series of noisy observations in grey with the latent state and the filtered state overlaid and a shaded uncertainty band, and a single isolated point at the right-hand end sitting well above the band."}
series_df <- data.frame(
  t = seq_len(n),
  observed = y_stable,
  truth = level,
  filtered = path$mean_1,
  lo = path$mean_1 - 2 * path$sd_1,
  hi = path$mean_1 + 2 * path$sd_1
)
break_df <- data.frame(t = n, y = y_break[n])
ggplot2::ggplot(series_df, ggplot2::aes(t)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = lo, ymax = hi), fill = "#0072B2", alpha = 0.18
  ) +
  ggplot2::geom_point(
    ggplot2::aes(y = observed, colour = "observed (stable)"),
    size = 1.1, alpha = 0.7
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = truth, colour = "latent state"), linewidth = 0.8
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered, colour = "filtered state"), linewidth = 0.8
  ) +
  ggplot2::geom_point(
    data = break_df, ggplot2::aes(t, y, colour = "broken final point"),
    size = 2.6
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c(
      "observed (stable)" = "grey60",
      "latent state" = "#009E73",
      "filtered state" = "#0072B2",
      "broken final point" = "#D55E00"
    )
  ) +
  ggplot2::labs(
    x = "time step", y = "observation",
    title = "An end-of-sample break in the last observation"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-series-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the filtered-series figure",
    "is skipped.\n")
```

### How the subsampling p-value is reached

The subsampling calibration compares the end-of-sample statistic with the
in-sample blocks of the same series, so the p-value can be read straight off
the empirical distribution of those blocks.

```{r fig-blocks, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4, fig.cap = "The empirical distribution of the in-sample block statistics for each series, with the end-of-sample statistic marked. The subsampling p-value is the fraction of blocks lying at or beyond the rule, so the stable statistic sits inside the bulk while the broken one clears every block in the series.", fig.alt = "Two panels, each showing a rising step function of the cumulative share of in-sample block statistics against a logarithmic axis, annotated with a vertical rule; in the left panel the rule falls inside the curve, in the right panel it lies far beyond its right-hand end."}
blocks_df <- rbind(
  data.frame(
    block = test_stable$in_sample_blocks, series = "stable series"
  ),
  data.frame(
    block = test_break$in_sample_blocks, series = "broken final point"
  )
)
blocks_df$series <- factor(
  blocks_df$series, levels = c("stable series", "broken final point")
)
rule_df <- data.frame(
  series = factor(
    c("stable series", "broken final point"),
    levels = c("stable series", "broken final point")
  ),
  statistic = c(test_stable$statistic, test_break$statistic)
)
ggplot2::ggplot(blocks_df, ggplot2::aes(block)) +
  ggplot2::stat_ecdf(geom = "step", linewidth = 0.8, colour = "#0072B2") +
  ggplot2::geom_vline(
    data = rule_df, ggplot2::aes(xintercept = statistic),
    colour = "#D55E00", linetype = "dashed", linewidth = 0.8
  ) +
  ggplot2::facet_wrap(~ series) +
  ggplot2::scale_x_log10() +
  ggplot2::labs(
    x = "in-sample block statistic (log scale)",
    y = "cumulative share of blocks",
    title = "The end-of-sample statistic against its own in-sample blocks"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-blocks-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the block-calibration",
    "figure is skipped.\n")
```

### What size the two calibrations actually hold

The chi-square calibration is exact when the model is given. Applied with a
diffuse prior on a short series it need not be, so the size is worth
measuring rather than assuming. The study below simulates from the same
local-level model under the null and records the rejection rate at a nominal
0.05, with the Monte Carlo standard error the replicate count supports.

```{r size-study}
simulate_null <- function(n_obs) {
  lv <- cumsum(c(0, rnorm(n_obs - 1L, 0, sqrt(0.04))))
  lv + rnorm(n_obs, 0, 1)
}
empirical_size <- function(n_obs, method, n_rep) {
  p <- vapply(seq_len(n_rep), function(i1) {
    gmm_eos_test(
      prior, dynamics, measurement, simulate_null(n_obs),
      m = 1L, method = method
    )$p_value
  }, numeric(1L))
  size <- mean(p < 0.05)
  c(size = size, se = sqrt(size * (1 - size) / n_rep), n_rep = n_rep)
} # ends empirical_size, one row of the size study

n_rep_short <- 250L
n_rep_long <- 150L
size_chisq_30 <- empirical_size(30L, "chisq", n_rep_short)
size_chisq_120 <- empirical_size(120L, "chisq", n_rep_long)
size_andrews_30 <- empirical_size(30L, "andrews", n_rep_short)

chisq_sizes <- c(size_chisq_30[["size"]], size_chisq_120[["size"]])
chisq_ses <- c(size_chisq_30[["se"]], size_chisq_120[["se"]])
chisq_worst_se <- max(abs(chisq_sizes - 0.05) / chisq_ses)
```

The subsampling p-value takes only the values $(1 + k) / (2 + n - 2m)$ for
$k = 0, 1, \ldots$, so its attainable size at a nominal level is fixed by
that grid and can be computed exactly, without simulation.

```{r attainable}
attainable_size <- function(n_obs, m = 1L, alpha = 0.05) {
  n_block <- n_obs - 2L * m + 1L
  p_grid <- (1L + seq.int(0L, n_block)) / (1L + n_block)
  mean(p_grid < alpha)
} # ends attainable_size, the discrete ceiling on the subsampling size

floor_30 <- 1 / (30L - 2L + 2L)
floor_120 <- 1 / (120L - 2L + 2L)
attainable_30 <- attainable_size(30L)
attainable_120 <- attainable_size(120L)
andrews_ses_from_ceiling <- abs(size_andrews_30[["size"]] - attainable_30) /
  size_andrews_30[["se"]]
```

```{r size-kable, echo = FALSE}
knitr::kable(
  data.frame(
    calibration = c(
      "chi-square", "chi-square", "subsampling", "subsampling",
      "subsampling"
    ),
    series_length = c("30", "120", "30", "30", "120"),
    quantity = c(
      "simulated size", "simulated size", "simulated size",
      "attainable size (exact)", "attainable size (exact)"
    ),
    value = round(
      c(
        size_chisq_30[["size"]], size_chisq_120[["size"]],
        size_andrews_30[["size"]], attainable_30, attainable_120
      ), 4L
    ),
    std_error = c(
      round(size_chisq_30[["se"]], 4L), round(size_chisq_120[["se"]], 4L),
      round(size_andrews_30[["se"]], 4L), NA_real_, NA_real_
    ),
    replicates = c(
      n_rep_short, n_rep_long, n_rep_short, NA_integer_, NA_integer_
    )
  ),
  col.names = c("Calibration", "n", "Quantity", "Value", "Std. error",
                "Replicates"),
  caption = paste(
    "Rejection rate against a nominal 0.05 under the null, simulated",
    "where the calibration is continuous and computed exactly where the",
    "p-value grid is discrete."
  )
)
```

## Read

On the stable series the statistic is
`r round(test_stable$statistic, 2L)` with a subsampling p-value of
`r round(test_stable$p_value, 3L)`, and the test does not reject: the last
observation sits where the filter expects it. Pushing that same observation
five observation-noise standard deviations away lifts the statistic to
`r round(test_break$statistic, 1L)`, a p-value of
`r round(test_break$p_value, 4L)`, and a rejection. That is the answer to
the opening question in the form the analyst needs it: on this series, the
arriving observation is flagged as inconsistent with the fitted model, and
the same machinery leaves the unbroken series alone.

The first figure shows why. The broken point sits
`r round(break_gap, 1L)` filtered standard deviations from the state the
filter had reconstructed from the preceding
`r n - 1L` observations, well outside the two-standard-deviation band drawn
around it. The second figure shows how that surprise becomes a p-value: for
the stable series the statistic falls inside the bulk of the in-sample
blocks, so a substantial share of blocks lie beyond it, while for the broken
series it clears every one of the
`r length(test_break$in_sample_blocks)` blocks, which drives the p-value to
its floor.

That floor is the first thing the size table makes concrete. At $n = 120$
and $m = 1$ the smallest attainable p-value is
`r round(floor_120, 4L)`, and the broken series attains it exactly. On a
much shorter series the floor bites: at $n = 30$ the smallest attainable
p-value is `r round(floor_30, 3L)`, so the test can reject at the 0.05 level
only when the end-of-sample block is the single largest in the series. The
consequence is that the subsampling test cannot hold a size of 0.05 at these
lengths even in principle. Its attainable size is
`r round(attainable_30, 4L)` at $n = 30$ and `r round(attainable_120, 4L)`
at $n = 120$, and the simulated size at $n = 30$,
`r round(size_andrews_30[["size"]], 3L)` over
`r n_rep_short` replicates with a standard error of
`r round(size_andrews_30[["se"]], 3L)`, sits
`r round(andrews_ses_from_ceiling, 1L)` standard errors from that exact
ceiling. The subsampling calibration is therefore conservative by
construction on short series, not by accident.

The chi-square calibration is continuous and carries no such floor. Its
simulated size is `r round(size_chisq_30[["size"]], 3L)` at $n = 30$
(standard error `r round(size_chisq_30[["se"]], 3L)`, `r n_rep_short`
replicates) and `r round(size_chisq_120[["size"]], 3L)` at $n = 120$
(standard error `r round(size_chisq_120[["se"]], 3L)`, `r n_rep_long`
replicates). Both sit within about
`r round(chisq_worst_se, 1L)`
standard errors of the nominal 0.05, so at these replicate counts the study
resolves no departure from nominal for either length. What it does resolve
is the contrast between the two calibrations: the chi-square test spends its
nominal level, and the subsampling test cannot.

## Limits

The size study above is small on purpose, and its replicate counts bound
what it can settle. A rejection rate estimated from `r n_rep_short`
replicates carries a standard error near
`r round(sqrt(0.05 * 0.95 / n_rep_short), 3L)`, so a departure from nominal
of one or two percentage points is beyond its resolution. The study
establishes the order of the size and the qualitative contrast between the
two calibrations; a claim about the exact size at a given series length
needs a study several orders larger than a vignette can carry, and the
figures here are not that study.

The two calibrations agree when the innovations are Gaussian. They part
company when the observation noise is heavy-tailed: the parametric $\chi^2$
reference is then mis-specified, while the subsampling test, which judges
the end-of-sample block only against the other blocks of the same series,
keeps its size. This vignette simulates Gaussian noise throughout and so
does not exercise that divergence; `method = "andrews"` is the option to
reach for when normality is in doubt, at the cost of the discreteness
documented above. Which calibration is preferable therefore depends on
which failure the analyst minds more: a mis-specified reference
distribution, or a test that cannot spend its nominal level.

The test is defined for a fitted linear-Gaussian state-space model, supplied
exactly as for `gmm_filter()`: a single-component `gmm` prior, a `dynamics`
list with the state-transition and process-noise matrices, and a
`measurement` list with the observation and observation-noise matrices.
Gaussian-sum process or measurement noise is rejected rather than
approximated, because neither calibration is defined for it. The small-`m`
regime (`m = 1, 2, 3`) is the point of the test, and `m` must be smaller
than the series length.

One further caution belongs to the model rather than the test. The band in
the first figure is the filtered state and its standard deviation, which is
what `gmm_filter()` returns; the statistic is built from the one-step
*predictive* innovation, a slightly wider quantity. The figure therefore
shows the geometry of the surprise, not the exact scale the test divides by.
A model whose noise covariances are themselves estimated on the same short
series will understate both, and the test then inherits that optimism: it is
a test of the last observation against the model, not a test of the model.

## What to read next

*The closed-form operator calculus on a mixture* builds the predict-and-update
recursion this test runs internally, and covers `gmm_filter()` and the
mixture reduction that bounds it. *Reading the entropy of a fitted mixture*
covers the diagnostics that describe a fitted mixture once it is in hand.
*Fitting a proxy to a density you cannot sample* is the entry point to the
fitting machinery if the state-space model here is your first contact with
the package.

## References

Andrews, D. W. K. (2003). End-of-Sample Instability Tests. *Econometrica*,
71(6), 1661--1694.

## Reproduce

The vignette sets `set.seed(20260621)` once, before the series is
simulated, and every later result -- the two tests, the filter path and both
size studies -- is a deterministic function of that stream.

```{r session-info}
sessionInfo()
```
