---
title: "Reading the entropy of a fitted mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading the entropy of a fitted mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

An analyst has a fitted mixture on the screen and three questions that the
parameter list does not answer on its own: *how spread out is this mixture,
how far is it from the other fit I ran yesterday, and how many components
does my data actually support?* All three are entropy questions, and on a
Gaussian mixture most of them have closed-form answers. This vignette walks
the entropy surface of a fitted mixture: what is exact, what is estimated,
and what each quantity is safe to conclude.

## What

Six verbs carry the vignette. [`gmm_entropy()`] returns the order-2 Renyi
entropy of a mixture in closed form by default, and a Monte Carlo Shannon
estimate with its standard error and an analytic upper bound under
`order = "shannon"`. [`gmm_divergence()`] returns the symmetric
Cauchy-Schwarz divergence between two mixtures in closed form, and delegates
to the Monte Carlo Kullback-Leibler estimator under `type = "kl"`.

[`gmm_mutual_information()`] measures dependence between two coordinate
blocks of a fitted joint as the Cauchy-Schwarz divergence between the joint
and the product of its marginals, and
[`gmm_conditional_entropy()`] returns the predictive uncertainty of the
target coordinates given the conditioned ones, row by row.

[`gmm_anneal_path()`] cools a deterministic-annealing schedule and records
where the components bifurcate, which both stabilises a fit and discovers a
component count. [`bic_aic()`] reports the integrated completed likelihood
beside the usual criteria, and [`gmm_independence_graph()`] reads the
second-order conditional-independence structure of a fitted mixture.
[`maxent_target()`] runs the argument the other way, from a set of
constraints to the least-committal density consistent with them.

## Do

```{r seed}
set.seed(20260618)
```

### The closed-form spine

The integral of a product of two Gaussian densities is itself a Gaussian
density evaluation: writing $\mathcal{N}(x; m, S)$ for a Gaussian density,
the integral over $x$ of
$\mathcal{N}(x; a, A)\, \mathcal{N}(x; b, B)$ equals
$\mathcal{N}(a; b, A + B)$. Any quantity that is *quadratic* in a mixture --
the integral of $g(x)^2$, or of $p(x)\, q(x)$ -- is therefore a finite sum
of such evaluations, and closed-form. Shannon entropy carries the logarithm
of a sum and has no such closed form, so it is estimated by Monte Carlo.

### Renyi-2 entropy against its analytic value

`gmm_entropy()` returns the closed-form order-2 Renyi entropy by default.
For a single Gaussian the value is exact and matches the analytic form
$H_2(\mathcal{N}(\mu, \Sigma)) =
\tfrac{p}{2}\log(4\pi) + \tfrac{1}{2}\log\det\Sigma$, which is an
independent check on the implementation.

```{r renyi}
g <- gmm(
  weights = c(0.5, 0.5),
  means = list(c(-2, 0), c(2, 0)),
  covariances = list(diag(2), diag(2))
)
h2_mixture <- gmm_entropy(g)

sigma_one <- matrix(c(1, 0.3, 0.3, 1), 2L, 2L)
one <- gmm(weights = 1, means = list(c(0, 0)), covariances = list(sigma_one))
h2_closed <- gmm_entropy(one)
h2_analytic <- 0.5 * (2 * log(4 * pi) +
  as.numeric(determinant(sigma_one, logarithm = TRUE)$modulus))
h2_gap <- abs(h2_closed - h2_analytic)
```

```{r renyi-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "two-component mixture, closed form",
      "single Gaussian, closed form",
      "single Gaussian, analytic formula"
    ),
    value = round(c(h2_mixture, h2_closed, h2_analytic), 6L)
  ),
  col.names = c("Quantity", "Renyi-2 entropy (nats)"),
  caption = "Closed-form order-2 Renyi entropy, and the analytic check."
)
```

### Shannon entropy, bracketed from above

Shannon entropy is returned as a Monte Carlo estimate with its standard
error and an analytic upper bound.

```{r shannon}
sh <- gmm_entropy(g, order = "shannon", n_mc = 5000L, seed = 1L)
sh_slack <- sh$upper_bound - sh$mc
sh_slack_se <- sh_slack / sh$mc_se
```

```{r shannon-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "Monte Carlo estimate", "Monte Carlo standard error",
      "analytic upper bound", "draws"
    ),
    value = c(
      round(sh$mc, 4L), round(sh$mc_se, 4L),
      round(sh$upper_bound, 4L), sh$n_mc
    )
  ),
  col.names = c("Quantity", "Value"),
  caption = "Shannon entropy of the two-component mixture, in nats."
)
```

### Cauchy-Schwarz divergence, and the Kullback-Leibler alternative

`gmm_divergence()` returns the closed-form, symmetric Cauchy-Schwarz
divergence by default. It is non-negative and zero exactly when the two
mixtures are proportional. When the asymmetric Kullback-Leibler divergence
is wanted instead, `type = "kl"` delegates to the package's Monte Carlo
estimator `gmm_kld()`.

```{r divergence}
q <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(diag(2) * 2)
)
d_cs <- gmm_divergence(g, q)
d_self <- gmm_divergence(g, g)
d_kl <- gmm_divergence(g, q, type = "kl", n_mc = 2000L)
d_kl_gap <- d_kl$mc - d_kl$variational
```

```{r divergence-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "Cauchy-Schwarz, g against q", "Cauchy-Schwarz, g against itself",
      "Kullback-Leibler, Monte Carlo", "Kullback-Leibler, Monte Carlo se",
      "Kullback-Leibler, Hershey-Olsen approximation"
    ),
    value = round(
      c(d_cs, d_self, d_kl$mc, d_kl$mc_se, d_kl$variational), 4L
    )
  ),
  col.names = c("Quantity", "Value (nats)"),
  caption = "Two divergences between the same pair of mixtures."
)
```

### Mutual information and predictive entropy

Two further reads come from the same closed-form spine. Mutual information
is non-negative and zero exactly when the blocks are independent;
conditional entropy is the order-2 Renyi entropy of the conditional mixture
`gmm_conditionalise()` returns, evaluated row by row, with `NA` marking the
target coordinate.

```{r mutual-information}
sigma_joint <- matrix(c(1, 0.7, 0.7, 1), 2L, 2L)
joint <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(sigma_joint)
)
mi <- gmm_mutual_information(joint, 1L, 2L)

sigma_free <- diag(2)
independent <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(sigma_free)
)
mi_independent <- gmm_mutual_information(independent, 1L, 2L)

given_grid <- rbind(c(NA, 0), c(NA, 1), c(NA, 2))
h_cond <- gmm_conditional_entropy(joint, given = given_grid)
```

```{r mutual-information-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "mutual information, correlation 0.7",
      "mutual information, independent block",
      "conditional entropy at x2 = 0",
      "conditional entropy at x2 = 1",
      "conditional entropy at x2 = 2"
    ),
    value = round(c(mi, mi_independent, h_cond), 6L)
  ),
  col.names = c("Quantity", "Value (nats)"),
  caption = paste(
    "Dependence between the two coordinates, and the predictive",
    "uncertainty of the first given the second."
  )
)
```

### Cooling the fit: annealing and phase transitions

The same statistical-mechanics framing -- the fitting objective is a
variational free energy $F = \langle E \rangle - T H$ -- yields a robustness
tool. Softening the E-step responsibilities by a temperature $T$, so that
$\gamma_{ik} \propto \pi_k\, \mathcal{N}(x_i;\ \mu_k, \Sigma_k)^{1/T}$, and
cooling $T$ from a high value toward one, turns the fit into a homotopy: at
high $T$ the objective has a single smooth basin, and as $T$ falls the
components bifurcate at critical temperatures. Setting `anneal = TRUE` on
`fit_em_samples()` or `fit_kld_em()` uses this annealed path as a warm start
for the unchanged cold EM loop, which is less sensitive to local optima than
a cold multi-start.

```{r anneal-fit}
x_three <- rbind(
  matrix(rnorm(200L), ncol = 2L) +
    matrix(rep(c(-7, -7), each = 100L), ncol = 2L),
  matrix(rnorm(200L), ncol = 2L) +
    matrix(rep(c(7, -7), each = 100L), ncol = 2L),
  matrix(rnorm(200L), ncol = 2L) +
    matrix(rep(c(0, 8), each = 100L), ncol = 2L)
)
tgt_three <- gmm_target_from_samples(x_three)
fit_annealed <- fit_em_samples(tgt_three, N = 3L, anneal = TRUE, seed = 1L)
fit_annealed@diagnostics$annealed
```

Cooling also discovers the component count. `gmm_anneal_path()` tracks the
number of distinct centroids as the temperature falls; the count occupying
the widest temperature range is the discovered number of components, and
the first bifurcation has the closed-form critical temperature
$T_c = \lambda_{\max}(\Sigma^{-1} C)$, returned as `t_critical_analytic`
for an independent check.

```{r anneal-path}
path <- gmm_anneal_path(x_three, k_max = 6L, n_steps = 60L, seed = 1L)
k_found <- path$k_selected
t_empirical <- path$first_critical_temperature
t_analytic <- path$t_critical_analytic
t_ratio <- t_empirical / t_analytic
```

```{r anneal-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "components discovered", "first critical temperature, empirical",
      "first critical temperature, analytic", "ratio empirical to analytic",
      "cooling steps recorded"
    ),
    value = c(
      k_found, round(t_empirical, 2L), round(t_analytic, 2L),
      round(t_ratio, 3L), nrow(path$path)
    )
  ),
  col.names = c("Quantity", "Value"),
  caption = paste(
    "What the cooling schedule discovered on three well-separated",
    "clusters."
  )
)
```

The table gives two numbers where the cooling gives a whole curve. The
figure below is the curve: the number of distinct centroids against
temperature on the upper panel, the variational free energy on the lower,
with the empirical and analytic first critical temperatures marked.

```{r fig-anneal, eval = has_ggplot2, echo = has_ggplot2, fig.height = 5, fig.cap = "Deterministic annealing on three well-separated clusters. The upper panel is the staircase of distinct centroids as the temperature falls, the lower panel the variational free energy over the same schedule. The dashed rule is the empirical first critical temperature, the dotted rule the closed-form analytic one; the first bifurcation happens between them.", fig.alt = "Two stacked panels against a logarithmic temperature axis running from hot to cold: the upper panel is a rising staircase of the number of distinct centroids, the lower panel a descending free-energy curve, both annotated with two nearly coincident vertical rules marking the first critical temperature."}
anneal_df <- rbind(
  data.frame(
    temperature = path$path$temperature,
    value = path$path$n_effective,
    panel = "distinct centroids"
  ),
  data.frame(
    temperature = path$path$temperature,
    value = path$path$free_energy,
    panel = "variational free energy"
  )
)
ggplot2::ggplot(anneal_df, ggplot2::aes(temperature, value)) +
  ggplot2::geom_vline(
    xintercept = t_analytic, linetype = "dotted",
    colour = "#0072B2", linewidth = 0.7
  ) +
  ggplot2::geom_vline(
    xintercept = t_empirical, linetype = "dashed",
    colour = "#D55E00", linewidth = 0.7
  ) +
  ggplot2::geom_step(linewidth = 0.8, colour = "#000000") +
  ggplot2::facet_wrap(~ panel, ncol = 1L, scales = "free_y") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(
    x = "temperature (log scale, cooling to the left)",
    y = NULL,
    title = "Where the components bifurcate as the fit cools"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-anneal-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the annealing-path figure",
    "is skipped.\n")
```

### Maximum-entropy targets

Entropy also points the other way -- from a set of constraints to the
least-committal density consistent with them. `maxent_target()` builds that
maximum-entropy density: the Gaussian under first- and second-moment
constraints on the full space, the uniform under a support constraint alone,
and a truncated Gaussian under second moments on a box. The bounded cases
declare their support, so regime (iii) fits them under an automatically
selected support-matched proposal.

```{r maxent}
me_gauss <- maxent_target(moments = list(mean = c(0, 0), cov = diag(2)))
me_unif <- maxent_target(support = list(lower = c(0, 0), upper = c(1, 1)))
unif_density <- exp(me_unif@log_density(matrix(c(0.5, 0.5), nrow = 1L)))
```

```{r maxent-kable, echo = FALSE}
knitr::kable(
  data.frame(
    constraint = c("mean and covariance", "unit-square support"),
    family = c(me_gauss@metadata$family, me_unif@metadata$family),
    check = c(NA_real_, round(unif_density, 6L))
  ),
  col.names = c("Constraint supplied", "Maximum-entropy family",
                "Density at the centre"),
  caption = paste(
    "The least-committal density under each constraint set. The unit",
    "square has area one, so its uniform density is one."
  )
)
```

### Selecting a component count with the ICL

`bic_aic()` reports the integrated completed likelihood alongside the BIC
and AIC. The ICL adds to the BIC twice the entropy of the fitted
classification, $\mathrm{ICL} = \mathrm{BIC} + 2 E_N$ with
$E_N = -\sum_{i,k} \gamma_{ik} \log \gamma_{ik} \geq 0$, so it penalises
mixtures whose components overlap and favours well-separated solutions. It
equals the BIC for a single component and is never smaller.

```{r icl}
x_two <- rbind(
  matrix(rnorm(200L, -4), ncol = 2L),
  matrix(rnorm(200L, 4), ncol = 2L)
)
fit_two <- fit_em_samples(gmm_target_from_samples(x_two), N = 2L)
crit <- bic_aic(fit_two)
icl_gap <- crit$icl - crit$bic
```

```{r icl-kable, echo = FALSE}
knitr::kable(
  data.frame(
    criterion = c("BIC", "AIC", "ICL", "classification entropy",
                  "free parameters"),
    value = c(
      formatC(crit$bic, format = "f", digits = 2L),
      formatC(crit$aic, format = "f", digits = 2L),
      formatC(crit$icl, format = "f", digits = 2L),
      formatC(crit$classification_entropy, format = "e", digits = 1L),
      formatC(crit$n_params, format = "d")
    )
  ),
  col.names = c("Criterion", "Value"),
  caption = paste(
    "Information criteria for a two-component fit to two well-separated",
    "clusters."
  )
)
```

### Dependency structure of a mixture

`gmm_independence_graph()` reads the undirected second-order
conditional-independence (Gaussian graphical model) structure of a fitted
mixture: an edge $i - j$ is absent exactly when the partial correlation of
coordinates $i$ and $j$, given all the others, is below a threshold -- the
Markov statement $x_i \perp x_j \mid x_{\mathrm{rest}}$ at second order. The
overall covariance of a mixture is closed-form in its parameters, so no
sampling is needed. A tridiagonal precision matrix is a chain, and the
recovered graph should be that chain.

```{r graph-chain}
omega <- diag(4L)
for (i1 in seq_len(3L)) {
  omega[i1, i1 + 1L] <- -0.5
  omega[i1 + 1L, i1] <- -0.5
} # ends i1, over the off-diagonal band of the chain precision
g_chain <- gmm(
  weights = 1, means = list(rep(0, 4L)), covariances = list(solve(omega))
)
adj_chain <- gmm_independence_graph(g_chain)
edges_chain <- sum(adj_chain) / 2L
```

```{r graph-chain-kable, echo = FALSE}
knitr::kable(
  as.data.frame(adj_chain[, ]),
  caption = paste(
    "Recovered adjacency of a chain precision matrix: an edge between",
    "neighbouring coordinates and nowhere else."
  )
)
```

The distinctive use is **regime (iii)**. Composed with `fit_kld_em()`, the
diagnostic recovers the dependency structure of a target you can only
*evaluate* -- an unnormalised energy density -- where no sample exists to
drive a conventional structure estimator. The continuous field below couples
only neighbouring coordinates, so the recovered graph should again be the
chain.

```{r graph-regime-iii}
energy <- function(x_mat) {
  x_mat <- matrix(x_mat, ncol = 3L)
  rowSums((x_mat^2 - 1)^2) -
    0.7 * (x_mat[, 1L] * x_mat[, 2L] + x_mat[, 2L] * x_mat[, 3L])
}
field <- gmm_target(n_dim = 3L, log_density = function(x_mat) -energy(x_mat))
fit_field <- fit_kld_em(
  field,
  N = 8L,
  proposal = proposal_uniform(3L, -3, 3),
  is_size = 6000L,
  anneal = TRUE,
  seed = 1L,
  support_warn = FALSE
)
adj_field <- gmm_independence_graph(fit_field)
edges_field <- sum(adj_field) / 2L
```

```{r graph-regime-iii-kable, echo = FALSE}
knitr::kable(
  as.data.frame(adj_field[, ]),
  caption = paste(
    "Recovered adjacency of a three-coordinate energy field fitted in",
    "regime (iii), where only the density can be evaluated."
  )
)
```

## Read

The closed-form Renyi-2 entropy of the single Gaussian and its analytic
formula differ by `r formatC(h2_gap, format = "g", digits = 2)` nats, which
confirms that the closed form is the formula it claims to be. The
two-component mixture carries a higher entropy,
`r round(h2_mixture, 3L)` nats against
`r round(h2_closed, 3L)` for the single Gaussian, because its mass is
spread across two separated modes rather than one.

The Shannon estimate is `r round(sh$mc, 3L)` nats with a standard error of
`r round(sh$mc_se, 4L)`, and the analytic bound sits
`r round(sh_slack, 3L)` nats above it, a gap of
`r round(sh_slack_se, 1L)` standard errors. The estimate sitting clear
below the bound is the coherence check the bound is there to provide.

The two divergences answer different questions about the same pair. The
Cauchy-Schwarz divergence is `r round(d_cs, 4L)` nats and is exact; against
itself the mixture scores `r formatC(d_self, format = "g", digits = 2)`,
which is the zero the definition requires. The Monte Carlo
Kullback-Leibler estimate is `r round(d_kl$mc, 3L)` nats with a standard
error of
`r round(d_kl$mc_se, 4L)`, and the deterministic Hershey-Olsen
approximation returned beside it is `r round(d_kl$variational, 3L)`,
`r round(d_kl_gap, 3L)` nats below the estimate, which is the direction that
approximation takes on well-separated components. The Cauchy-Schwarz and
Kullback-Leibler divergences are not on a common scale and should not be
compared to each other, only across pairs of fits under one definition.

Mutual information reads `r round(mi, 4L)` nats between two coordinates
correlated at 0.7, and
`r formatC(mi_independent, format = "g", digits = 2)` between two
independent ones, which is the zero the definition requires. The conditional
entropy is `r round(h_cond[1L], 4L)` nats at every point of the grid,
identical across the three conditioning values because a single-component
joint has a conditional whose spread does not depend on what is conditioned
on; a multi-component joint would vary across the grid.

Cooling discovered `r k_found` components, which is the number of clusters
the data was built from. The first bifurcation was recorded at temperature
`r round(t_empirical, 1L)` against the closed-form
$T_c = \lambda_{\max}(\Sigma^{-1} C) =$ `r round(t_analytic, 1L)`, a ratio
of `r round(t_ratio, 2L)`. The two rules in the figure sit close together
but not on top of each other: the empirical value is the first temperature
on a `r nrow(path$path)`-step grid at which the split was *detected*, so it
lags the temperature at which the split becomes available, and the size of
that lag is set by the grid spacing rather than by the data. The staircase
in the upper panel is the discovery itself -- one centroid held over the
hot end of the schedule, then a jump, then the plateau at `r k_found` that
occupies the widest temperature range and is therefore the count returned.
The lower panel shows the free energy falling smoothly through the same
bifurcations, which is why a cold multi-start has no comparable signal to
follow.

The maximum-entropy constructions return the families the theory names: the
Gaussian under moment constraints, the uniform under a support constraint,
whose density at the centre of the unit square is
`r round(unif_density, 3L)`, the reciprocal of an area of one.

For the two-cluster fit the BIC is `r round(crit$bic, 1L)` and the ICL is
`r round(crit$icl, 1L)`, differing by
`r formatC(icl_gap, format = "e", digits = 1)` because the classification
entropy is `r formatC(crit$classification_entropy, format = "e", digits = 1)`.
Two clusters this well separated leave no ambiguity about which component a
point belongs to, so the ICL has nothing to penalise and coincides with the
BIC to the precision the table reports; on overlapping components the two
criteria separate, and the ICL prefers the smaller count.

Both dependency graphs come back as the chain. The analytic case recovers
`r edges_chain` edges from a tridiagonal precision matrix, and the regime
(iii) case recovers `r edges_field` edges from a three-coordinate energy
field that was never sampled, only evaluated. That is the answer to the
third part of the opening question in its sharpest form: the structure of a
target can be read off a fitted proxy even when no sample of the target
exists.

Taken together, the opening question has three answers. Spread is the
Renyi-2 entropy, exact and cheap. Distance between two fits is the
Cauchy-Schwarz divergence, also exact, with the Kullback-Leibler estimate
available when a downstream consumer requires that definition. The number of
components the data supports is read from the annealing staircase, and
confirmed against the ICL.

## Limits

The Renyi-2 entropy and the Cauchy-Schwarz divergence are exact and cheap,
costing $O(K^2)$ Gaussian evaluations, and are the recommended defaults for
reporting the spread of a fitted mixture and the discrepancy between two
fits. The Shannon entropy and the Kullback-Leibler divergence are the more
familiar quantities but are Monte Carlo estimates, so they carry the
standard error the tables report and should be read against it rather than
to the last digit printed. Reach for them when a downstream consumer
requires those specific definitions, not by default.

The independence graph is a graphical-model diagnostic, not a
causal-discovery method. It returns the undirected Markov skeleton, not edge
directions, and being second order it sees only dependence that enters the
covariance. A pair of coordinates coupled purely through a higher moment, or
through a mixture-gate that leaves the overall covariance untouched, will
appear as an absent edge. The threshold on the partial correlation is a
choice, and a graph read at one threshold is not the graph read at another.

The annealing path discovers a component count on data whose clusters are
well separated, which is the case where every method agrees. The staircase
becomes hard to read when clusters overlap, the widest-plateau rule can then
select a count no criterion would endorse, and the analytic critical
temperature governs only the *first* bifurcation, not the later ones. The
ICL is a criterion, not a test: it ranks candidate counts and offers no
significance statement about the winner.

Everything on this page is read off a mixture that has already been fitted.
Nothing here checks that the mixture is a good proxy for the target it was
fitted to; that is the job of the importance-sampling diagnostics and the
fit-quality certificate, and a confident entropy computed on a degenerate
fit is confidently wrong.

## What to read next

*How well a mixture proxies four awkward shapes* runs regime (iii) across
targets of different geometry and reports the fit-quality diagnostics these
entropy reads assume are healthy. *Choosing between the three fitting
regimes* explains the annealed fit's starting point, the difference between
fitting from samples and fitting from a density you can only evaluate. *The
closed-form operator calculus on a mixture* covers the conditioning
operation that `gmm_conditional_entropy()` calls row by row, and the rest of
the exact algebra on a fitted mixture. *Fitting a proxy to a density you
cannot sample* is the shortest route into the regime (iii) fit used for the
energy-field graph above.

## References

* Biernacki, C., Celeux, G. and Govaert, G. (2000). *Assessing a mixture
  model for clustering with the integrated completed likelihood.* IEEE
  Transactions on Pattern Analysis and Machine Intelligence 22(7),
  719--725. <doi:10.1109/34.865189>.
* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.
* Jaynes, E. T. (1957). *Information theory and statistical mechanics.*
  Physical Review 106(4), 620--630. <doi:10.1103/PhysRev.106.620>.
* Rose, K. (1998). *Deterministic annealing for clustering, compression,
  classification, regression, and related optimization problems.*
  Proceedings of the IEEE 86(11), 2210--2239. <doi:10.1109/5.726788>.

## Reproduce

The vignette sets `set.seed(20260618)` once before the first simulated
dataset, and every verb that draws its own Monte Carlo sample is given an
explicit `seed` argument, so the numbers above are reproducible from this
source alone.

```{r session-info}
sessionInfo()
```
