---
title: "Entropy diagnostics with proxymix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Entropy diagnostics with proxymix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(20260618)
```

```{r setup}
library(proxymix)
```

`proxymix` fits Kullback--Leibler-optimal Gaussian-mixture proxies, so relative entropy is already at the core of the package. This vignette surfaces the entropy structure of a *fitted* mixture as closed-form diagnostics: the quadratic (order-2) Renyi entropy of a mixture, the Cauchy--Schwarz divergence between two mixtures, and a Monte-Carlo Shannon entropy bracketed by an analytic bound.

## The closed-form spine

The integral of a product of two Gaussian densities is itself a Gaussian density evaluation: writing `N(x; m, S)` for a Gaussian density, the integral over `x` of `N(x; a, A) N(x; b, B)` equals `N(a; b, A + B)`. Any quantity that is *quadratic* in a mixture -- the integral of `g(x)^2`, or of `p(x) q(x)` -- is therefore a finite sum of such evaluations, and closed-form. Shannon entropy carries the logarithm of a sum and has no such closed form, so it is estimated by Monte Carlo.

## Renyi-2 entropy

`gmm_entropy()` returns the closed-form order-2 Renyi entropy by default.

```{r}
g <- gmm(weights = c(0.5, 0.5),
         means = list(c(-2, 0), c(2, 0)),
         covariances = list(diag(2), diag(2)))
gmm_entropy(g)
```

For a single Gaussian the value is exact and matches the analytic form `H2(N(mu, S)) = (p / 2) log(4 pi) + (1 / 2) log det(S)`:

```{r}
S <- matrix(c(1, 0.3, 0.3, 1), 2, 2)
one <- gmm(weights = 1, means = list(c(0, 0)), covariances = list(S))
c(closed_form = gmm_entropy(one),
  analytic    = 0.5 * (2 * log(4 * pi) +
                         as.numeric(determinant(S, logarithm = TRUE)$modulus)))
```

## Shannon entropy, bracketed

Shannon entropy is returned as a Monte-Carlo estimate with its standard error and an analytic upper bound that brackets it from above.

```{r}
gmm_entropy(g, order = "shannon", n_mc = 5000L, seed = 1L)
```

The estimate sits at or below the bound up to Monte-Carlo error, which is the coherence check the bound is there to provide.

## Cauchy--Schwarz divergence

`gmm_divergence()` returns the closed-form, symmetric Cauchy--Schwarz divergence by default. It is non-negative and zero exactly when the two mixtures are proportional.

```{r}
q <- gmm(weights = 1, means = list(c(0, 0)), covariances = list(diag(2) * 2))
gmm_divergence(g, q)   # closed-form, symmetric
gmm_divergence(g, g)   # zero
```

When the asymmetric Kullback--Leibler divergence is wanted instead, `type = "kl"` delegates to the package's existing Monte-Carlo estimator `gmm_kld()`.

```{r}
gmm_divergence(g, q, type = "kl", n_mc = 2000L)$mc
```

## Mutual information and predictive entropy

Two further reads come from the same closed-form spine. `gmm_mutual_information()` measures the dependence between two coordinate blocks of a fitted joint as the Cauchy--Schwarz divergence between the joint and the product of the marginals; it is non-negative and zero exactly when the blocks are independent.

```{r}
s <- matrix(c(1, 0.7, 0.7, 1), 2, 2)
joint <- gmm(weights = 1, means = list(c(0, 0)), covariances = list(s))
gmm_mutual_information(joint, 1L, 2L)
```

`gmm_conditional_entropy()` returns the predictive uncertainty of the target coordinates given the conditioned ones -- the order-2 Renyi entropy of the conditional mixture from `gmm_conditionalise()`, evaluated row by row, with `NA` marking the target coordinate.

```{r}
gmm_conditional_entropy(joint, given = rbind(c(NA, 0), c(NA, 1), c(NA, 2)))
```

## When to use which

The Renyi-2 entropy and the Cauchy--Schwarz divergence are exact and cheap (`O(K^2)` Gaussian evaluations), and are the recommended defaults for reporting the spread of a fitted mixture and the discrepancy between two fits. The Shannon entropy and the Kullback--Leibler divergence are the more familiar quantities but are Monte-Carlo estimates; reach for them when a downstream consumer requires those specific definitions.

## Deterministic annealing and phase transitions

The same statistical-mechanics framing -- the fitting objective is a variational free energy $F = \langle E \rangle - T H$ -- yields a robustness tool. Softening the E-step responsibilities by a temperature $T$, so that $\gamma_{ik} \propto \pi_k\, \mathcal{N}(x_i;\ \mu_k, \Sigma_k)^{1/T}$, and cooling $T$ from a high value toward one, turns the fit into a homotopy: at high $T$ the objective has a single smooth basin, and as $T$ falls the components bifurcate at critical temperatures. Setting `anneal = TRUE` on `fit_em_samples()` or `fit_kld_em()` uses this annealed path as a warm-start for the unchanged cold EM loop, which is far less sensitive to local optima than a cold multi-start.

```{r}
x <- rbind(
  matrix(rnorm(200), ncol = 2) + matrix(rep(c(-7, -7), each = 100), ncol = 2),
  matrix(rnorm(200), ncol = 2) + matrix(rep(c( 7, -7), each = 100), ncol = 2),
  matrix(rnorm(200), ncol = 2) + matrix(rep(c( 0,  8), each = 100), ncol = 2)
)
tgt <- gmm_target_from_samples(x)
fit <- fit_em_samples(tgt, N = 3L, anneal = TRUE, seed = 1L)
fit@diagnostics$annealed
```

Cooling also discovers the component count. `gmm_anneal_path()` tracks the number of distinct centroids as the temperature falls; the count occupying the widest temperature range is the discovered number of components, and the first bifurcation has the closed-form critical temperature $T_c = \lambda_{\max}(\Sigma^{-1} C)$, returned as `t_critical_analytic` for an independent check.

```{r}
path <- gmm_anneal_path(x, k_max = 6L, n_steps = 60L, seed = 1L)
path$k_selected
c(empirical = path$first_critical_temperature,
  analytic  = path$t_critical_analytic)
```
