---
title: "Compressing a kernel density estimate into a mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compressing a kernel density estimate into a mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~3 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

An analyst has a kernel density estimate of a two-dimensional quantity and a
pipeline that needs to keep conditioning on one coordinate and integrating
out the other. Every such query against the kernel estimate is a numerical
sweep over `n` kernels, and there is no algebra to compose. The question is:
*I have a kernel density estimate and I need marginals and conditionals in
closed form -- what does compressing it into a K-component mixture cost me?*

## What

A kernel density estimate gives a smooth, non-parametric density from a
sample of `n` points, at three operational costs. Integrating out a
coordinate has no closed form and requires a numerical sweep over the kernel
mixture. Conditioning has no analogue of the Schur complement, so one has to
resample or build a conditional estimate per query. And there is no native
aggregation algebra: pushforwards through linear operators, observation
channel updates and block sums all collapse to Monte Carlo.

[`from_kde()`] compresses the kernel estimate into an `N`-component Gaussian
mixture proxy, with `N` typically much smaller than `n`, by treating the
estimate as a normalised, evaluable target and running regime (iii) KLD-EM
against it. Its `bandwidth` argument takes a rule of thumb by name
(`"silverman"`, `"scott"`) or a numeric value, and `validation_size`
requests a held-out fit quality on an independent importance draw. The proxy
then supports the full closed-form operator set exposed by the rest of the
package -- [`dgmm()`], [`rgmm()`], [`gmm_marginalise()`],
[`gmm_conditionalise()`], and the affine-Gaussian calculus of
[`gmm_affine()`] and [`gmm_observe()`].

[`ess_summary()`] reports the importance-sampling diagnostics of the
compression: the effective sample size, the largest self-normalised weight,
and the held-out validation divergence. [`hellinger_mc()`] gives a Monte
Carlo squared Hellinger distance between the proxy and its target, which is
meaningful here because a kernel estimate is normalised by construction.

This is not new science. It is a practical bridge between a non-parametric
density estimate and a parametric closed-form mixture representation. The
bias inherited from the kernel estimate is reproduced in the proxy, and the
bandwidth controls the bias-variance trade-off.

## Do

### Recovering a known mixture

Sample from a known bimodal mixture and ask `from_kde()` to recover it. The
data here are synthetic: two unit-covariance Gaussians at $(-2, 0)$ and
$(2, 0)$, 150 draws each.

```{r recover-bimodal}
set.seed(20260601)

true_means <- cbind(c(-2, 0), c(2, 0))
true_cov <- diag(2)
x <- rbind(
  mvnfast::rmvn(150L, mu = true_means[, 1L], sigma = true_cov),
  mvnfast::rmvn(150L, mu = true_means[, 2L], sigma = true_cov)
)

fit <- from_kde(
  x, N = 2L,
  bandwidth = "silverman",
  is_size = 2000L, max_iter = 60L, seed = 1L,
  validation_size = 2000L
)
fit
```

The proxy components sit near the true means. Ordering the fitted components
by their first coordinate makes the comparison with the truth direct.

```{r recovered-means}
mu_hat <- vapply(fit@means, function(mu) mu, numeric(2L))
comp_order <- order(mu_hat[1L, ])
mu_hat <- mu_hat[, comp_order, drop = FALSE]
weight_hat <- fit@weights[comp_order]
mean_error <- max(abs(mu_hat - true_means))
```

```{r recovered-means-table, echo = FALSE}
knitr::kable(
  data.frame(
    component = c(1L, 2L),
    fitted_x1 = round(mu_hat[1L, ], 3),
    fitted_x2 = round(mu_hat[2L, ], 3),
    true_x1 = true_means[1L, ],
    true_x2 = true_means[2L, ],
    weight = round(weight_hat, 3)
  ),
  row.names = FALSE,
  col.names = c("Component", "Fitted $x_1$", "Fitted $x_2$", "True $x_1$",
                "True $x_2$", "Weight"),
  caption = paste0(
    "The two recovered component means beside the means of the mixture the ",
    "300 samples were drawn from."
  )
)
```

### Quality of the importance-sampled fit

A healthy regime (iii) fit has an effective sample size close to `is_size`
and a largest weight of a few parts in a thousand. The validation block
reports a held-out divergence on an independent importance draw, which is
the safeguard against overfitting to one particular importance realisation.

```{r fit-quality}
es <- ess_summary(fit)
c(is_size = es$is_size, ess = round(es$ess, 1),
  ess_relative = round(es$ess_relative, 3),
  max_weight = signif(es$max_weight, 3),
  validation_kld = signif(es$validation_kld, 3))
```

### What the compression costs

The proxy and the kernel estimate are both normalised densities on the same
space, so the distance between them can be integrated directly on a grid.
The total variation distance is the natural currency: it is the largest
difference in probability the two densities can assign to any event.

```{r compression-cost}
g1 <- seq(-6, 6, length.out = 160L)
g2 <- seq(-5, 5, length.out = 140L)
grid <- expand.grid(x1 = g1, x2 = g2)
gm <- as.matrix(grid)
cell <- (g1[2L] - g1[1L]) * (g2[2L] - g2[1L])

grid$kde <- exp(fit@target@log_density(gm))
grid$proxy <- dgmm(gm, fit)

total_variation <- 0.5 * sum(abs(grid$kde - grid$proxy)) * cell
hell <- hellinger_mc(fit, n_mc = 3000L, seed = 1L)
c(total_variation = signif(total_variation, 3),
  hellinger_sq = signif(hell$h2, 3), hellinger_se = signif(hell$se, 3))
```

### Bandwidth sensitivity

Smaller bandwidths track the data more tightly, at lower bias and higher
variance; larger bandwidths smooth the proxy. The proxy inherits this
trade-off from the estimate it compresses.

```{r bandwidth-sweep}
bandwidth_grid <- c(0.2, 0.5, 1.0)
fits <- lapply(bandwidth_grid, function(h) {
  from_kde(x, N = 2L, bandwidth = h,
           is_size = 1500L, max_iter = 40L, seed = 1L)
})
```

```{r bandwidth-table, echo = FALSE}
knitr::kable(
  data.frame(
    bandwidth = bandwidth_grid,
    ess = round(vapply(fits, function(f) f@diagnostics$ess, numeric(1L)), 1),
    max_weight = signif(
      vapply(fits, function(f) f@diagnostics$max_weight, numeric(1L)), 3
    ),
    trace_sigma = round(
      vapply(fits, function(f) sum(diag(f@covariances[[1L]])), numeric(1L)), 3
    )
  ),
  row.names = FALSE,
  col.names = c("Bandwidth", "Effective sample size", "Largest weight",
                "Trace of $\\Sigma_1$"),
  caption = paste0(
    "Bandwidth sweep at 1500 importance draws: a wider kernel gives a ",
    "smoother target, an easier importance problem, and a broader first ",
    "component."
  )
)
```

### Composing the compressed proxy

Once compressed, the proxy is an ordinary mixture, so conditioning and
sampling are exact algebra rather than a sweep over 300 kernels.

```{r compose}
slice <- gmm_conditionalise(fit, given = c(NA, 0))
draws <- rgmm(200L, slice)
c(components = gmm_n_components(slice), dimension = gmm_dim(slice),
  draws = nrow(draws))
```

### Visualising the compression

The contour comparison shows where the compression spends its error, which
the summary distances cannot: the two contour sets are read on shared levels,
so any place they part company is a place the proxy has smoothed the
estimate.

```{r visualise, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4.5, fig.cap = "The mixture-proxy log-density contours (orange, dashed) track the kernel estimate's contours (blue, solid) closely across the sampled region, parting company only at the two modes where the two-component proxy smooths the kernel estimate's sampling bumps.", fig.alt = "Contour plot comparing the kernel-density log-density and the Gaussian-mixture proxy log-density on a planar grid, with sample points overlaid; the two contour sets nearly coincide."}
plot_grid <- expand.grid(
  x1 = seq(-5, 5, length.out = 80L),
  x2 = seq(-4, 4, length.out = 60L)
)
pm <- as.matrix(plot_grid)
plot_grid$kde <- fit@target@log_density(pm)
plot_grid$proxy <- log(dgmm(pm, fit))

## Shared contour levels so the two log-densities are directly comparable.
brks <- pretty(range(c(plot_grid$kde, plot_grid$proxy), finite = TRUE), 9L)
sample_df <- data.frame(x1 = x[, 1L], x2 = x[, 2L])

ggplot2::ggplot() +
  ggplot2::geom_point(data = sample_df, ggplot2::aes(x1, x2),
                      colour = "grey60", alpha = 0.25, size = 0.5) +
  ggplot2::geom_contour(
    data = plot_grid,
    ggplot2::aes(x1, x2, z = kde, colour = "Kernel estimate",
                 linetype = "Kernel estimate"),
    breaks = brks, linewidth = 0.45
  ) +
  ggplot2::geom_contour(
    data = plot_grid,
    ggplot2::aes(x1, x2, z = proxy, colour = "Mixture proxy",
                 linetype = "Mixture proxy"),
    breaks = brks, linewidth = 0.55
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("Kernel estimate" = "#0072B2", "Mixture proxy" = "#D55E00")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Kernel estimate" = "solid", "Mixture proxy" = "dashed")
  ) +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::labs(
    title = "Kernel estimate and Gaussian-mixture proxy",
    subtitle = "log-density contours on shared levels",
    x = expression(x[1]), y = expression(x[2])
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 legend.position = "top",
                 panel.grid.minor = ggplot2::element_blank())
```

```{r visualise-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the contour-comparison",
    "figure is skipped.\n")
```

## Read

The `r gmm_n_components(fit)`-component proxy recovers the mixture the 300
samples came from: the fitted means differ from the true means by at most
`r signif(mean_error, 2)`, and the two components carry weights of
`r paste(sprintf("%.3f", sort(fit@weights)), collapse = " and ")` against a
true half and half.

The compression itself was a well-conditioned importance problem. The
effective sample size was `r round(es$ess)` out of `r es$is_size` draws, or
`r round(100 * es$ess_relative)` per cent, and the largest single weight was
`r signif(es$max_weight, 2)`, so no handful of draws carried the fit. On an
independent importance draw the held-out divergence was
`r signif(es$validation_kld, 2)`, against a Monte Carlo standard error of
`r signif(es$mc_se_kld, 2)` on the fitting sample, so the fit does not
depend on the particular importance realisation it was estimated from.

The cost of the compression is the direct answer to the question. Going from
a 300-kernel estimate to a `r gmm_n_components(fit)`-component mixture moves
total variation by `r signif(total_variation, 2)`: the two densities agree on
the probability of any event to within about
`r round(100 * total_variation, 1)` percentage points. The Monte Carlo
estimate of the squared Hellinger
distance is `r signif(hell$h2, 2)`, against a standard error of
`r signif(hell$se, 2)` on that estimate, so it sits within one standard
error of zero and the two densities are not separated by that measure at
this sample size. The figure shows where the remaining
error sits: at the two modes, where a two-component proxy cannot follow the
bumps that 300 finite samples put into the kernel estimate. What is bought
with that is exact conditioning, marginalisation and sampling -- the slice
taken above is a `r gmm_dim(slice)`-dimensional,
`r gmm_n_components(slice)`-component mixture obtained by algebra, not by a
sweep over kernels.

The bandwidth sweep shows the trade-off passing straight through the
compression. A wider kernel gives a smoother target and an easier importance
problem, and the effective sample size rises from
`r round(fits[[1L]]@diagnostics$ess)` at bandwidth
`r sprintf("%.1f", bandwidth_grid[1L])` to
`r round(fits[[3L]]@diagnostics$ess)` at bandwidth
`r sprintf("%.1f", bandwidth_grid[3L])`, out of 1500 draws. The first
component's covariance trace widens over the same sweep, from
`r round(sum(diag(fits[[1L]]@covariances[[1L]])), 2)` to
`r round(sum(diag(fits[[3L]]@covariances[[1L]])), 2)`, which is the proxy
inheriting the estimate's smoothing rather than resisting it.

## Limits

`from_kde()` is well tested for five dimensions or fewer. It warns between
six and ten and refuses above ten. Regime (iii) is driven by importance
sampling, whose effective sample size falls sharply in high dimensions, and
this vignette exercises the two-dimensional case only. For richer ambient
spaces, compose a low-dimensional proxy with the affine-Gaussian operator
calculus rather than fitting in the ambient space.

Only scalar and diagonal bandwidths are supported. A full-matrix bandwidth
effectively encodes a covariance estimate and blurs the distinction between
the kernel estimate and the mixture; where that is what is wanted, fit the
mixture directly with `fit_proxymix(regime = "sample")`.

A kernel estimate is normalised by construction, so the returned `gmm_target`
declares `normalised = TRUE` and `log_normalizer = 0`. That is what makes
`hellinger_mc()` meaningful here and lets the divergence diagnostics report
absolute values rather than shifted ones. A target that is not normalised
does not support the same reading.

The constructor does not choose the bandwidth. Use the conventional rules of
thumb, passed by name, or a cross-validation procedure outside the package,
and pass the chosen bandwidth in. The sweep above shows the consequences of
the choice; it does not make the choice.

Finally, the compression earns its place only in some pipelines. Where the
goal is simply to find a Gaussian-mixture density given samples, the
classical expectation-maximisation regime is more direct, fits in linear
time, and does not pay the regime (iii) importance-sampling tax. Reach for
`from_kde()` when the kernel estimate's smoothing is itself a step in the
pipeline, or when its bias-variance trade-off is part of what is being
validated downstream.

## What to read next

*Choosing between the three fitting regimes* sets out why compressing an
evaluable-but-unsampleable target is regime (iii) work, and what the
classical expectation-maximisation regime would do with the same 300 samples
instead. *The closed-form operator calculus on a mixture* is the payoff of
the compression: the marginalisation, conditioning, pushforward and update
operators that the kernel estimate could not offer. *One mixture, many
methods* places the kernel estimate and the compressed proxy on a single
axis running from one global component to one component per datum.

## References

* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.
* Silverman, B. W. (1986). *Density Estimation for Statistics and Data
  Analysis.* CRC Press.
* Scott, D. W. (1992). *Multivariate Density Estimation: Theory,
  Practice, and Visualization.* Wiley.

## Reproduce

The ambient seed is 20260601, and every `from_kde()` call additionally
passes `seed = 1L`, so the importance draws are reproducible independently of
the ambient random-number state. The Monte Carlo Hellinger distance is drawn
under its own `seed = 1L`.

```{r session-info}
sessionInfo()
```
