---
title: "One mixture, many methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One mixture, many methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. Every fit here is a
     sample-regime EM or a closed-form moment match, so the time is
     dominated by the six figures rather than by the fitting. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

An analyst has fitted a Gaussian mixture over a table of data, and now has a
week of ordinary work in front of her: a regression, a smoother, a
clustering, a principal-components plot, a shrunk coefficient, and a
treatment effect that she suspects varies from unit to unit. Each of those
normally means reaching for a different package. Her question is: *I already
have one fitted mixture over these columns -- which of the tools I would
otherwise reach for does it actually replace, how do I know the answer is
the same one, and what do I give up by using it instead?*

The short answer is that conditioning, marginalisation and
eigendecomposition are all exact on a Gaussian mixture, so six familiar
analyses fall out of one object in closed form, four of them reproducing the
classical answer to machine precision. The rest of this vignette shows each
substitution, checks it against the tool it stands in for, and states the
cost.

## What

A [`gmm`] is a finite Gaussian mixture: weights, means and covariances.
[`gmm_target_from_samples()`] wraps a data matrix as a target and
[`fit_proxymix()`] fits the mixture to it, in the closed-form `"moment"`
regime at one component or the classical expectation-maximisation
`"sample"` regime at more than one.

[`gmm_conditionalise()`] returns the conditional mixture given values on
some coordinates, in closed form by the Schur complement. That single
operator carries the regression, the kernel smoother, the imputation and the
ridge estimator below: a conditional mean read off a joint fit is a
prediction, and which classical estimator it coincides with depends only on
how many components the mixture has and how wide they are.

[`bic_aic()`] scores a component count. The `ridge_eps` argument of the
fitting verbs adds a multiple of the identity to each component covariance,
which is where the penalised-regression section comes from.

The decision family reads one joint fit as a causal object.
[`fit_uplift()`] assembles a mixture over outcome, treatment and covariates
and records the roles and the treatment arms it observed;
[`proxy_cate()`] reads the conditional average treatment effect per unit,
[`proxy_decide()`] turns it into an action under a value and a cost, and
[`proxy_identification_report()`] states what the reading assumes and what
it cannot deliver.

## Do

Each section below keeps the same shape: the **problem**, the **usual
tool**, the **proxymix route**, the code, and the **idea** behind the
substitution. The trade-offs are gathered in *Read*.

```{r shared-helpers}
## The conditional mean of the outcome coordinate, gated across components:
## sum_k pi_k(x) mu_k^{y|x}. This one helper serves the regression, the
## kernel-regression and the ridge sections.
cond_mean <- function(fit, xv) {
  vapply(xv, function(xx) {
    g <- gmm_conditionalise(fit, given = c(NA, xx))
    sum(g@weights * vapply(g@means, function(m) m[1L], numeric(1L)))
  }, numeric(1L))
}
```

```{r sci-notation, include = FALSE}
## Differences at the double-precision floor are typeset as LaTeX powers of
## ten, so that an inline value reads as mathematics rather than as R's
## own exponent notation.
sci <- function(v, digits = 2L) {
  e <- floor(log10(abs(v)))
  paste0("$", signif(v / 10^e, digits), " \\times 10^{", e, "}$")
}
sci_plain <- function(v) formatC(v, format = "e", digits = 1L)
```

### Regression: from `lm()` to a conditioned mixture

**Problem.** Predict an outcome $y$ from a predictor $x$. **Usual tool.**
`lm()` or `glm()`, one global line. **proxymix route.** Fit a mixture over
the joint $(y, x)$, then read $E[y \mid x]$ by conditioning. At $K = 1$ this
*is* least squares; at $K > 1$ the gate bends the line.

```{r reg-fit}
set.seed(20260617)
n <- 400L
x <- runif(n, -3, 3)
y <- 0.3 * x + 1.2 * pmax(x, 0) + rnorm(n, sd = 0.4)  # a bent relationship
dat <- data.frame(y = y, x = x)

joint <- gmm_target_from_samples(cbind(y, x))
fit1 <- fit_proxymix(joint, N = 1L, regime = "moment", ridge_eps = 0)
fit3 <- fit_proxymix(joint, N = 3L, regime = "sample", max_iter = 150L)

## At K = 1 the conditional slope should equal the lm coefficient exactly.
slope_mix <- gmm_conditionalise(fit1, given = c(NA, 1))@means[[1L]] -
  gmm_conditionalise(fit1, given = c(NA, 0))@means[[1L]]
slope_lm <- unname(coef(lm(y ~ x, dat))["x"])
diff_reg <- abs(slope_mix - slope_lm)
```

```{r fig-reg, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "One global line (K = 1, which is ordinary least squares) against a conditioned three-component mixture (K = 3) on data whose true relationship is bent at the origin. The mixture follows the bend; the single line cannot.", fig.alt = "Scatter of y against x with a straight ordinary-least-squares line and a curved mixture conditional mean that follows a bend in the data."}
grid_reg <- data.frame(x = seq(-3, 3, length.out = 200L))
grid_reg$ols <- as.numeric(predict(lm(y ~ x, dat), newdata = grid_reg))
grid_reg$mix <- cond_mean(fit3, grid_reg$x)
ggplot2::ggplot() +
  ggplot2::geom_point(data = dat, ggplot2::aes(x, y), colour = "grey60",
                      alpha = 0.4, size = 0.7) +
  ggplot2::geom_line(data = grid_reg,
                     ggplot2::aes(x, ols, colour = "lm (K = 1)"),
                     linewidth = 0.9) +
  ggplot2::geom_line(data = grid_reg,
                     ggplot2::aes(x, mix, colour = "mixture (K = 3)"),
                     linewidth = 0.9) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("lm (K = 1)" = "#0072B2", "mixture (K = 3)" = "#D55E00")
  ) +
  ggplot2::labs(
    x = "x", y = "y",
    title = "Regression: a global line against a conditioned mixture"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-reg-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the regression figure is",
    "skipped.\n")
```

**Idea.** Each component supplies an affine conditional mean
$\mu_k^{y \mid x}$; the responsibilities $\pi_k(x)$ gate them, so the global
mean $\sum_k \pi_k(x)\, \mu_k^{y \mid x}$ is nonlinear. This is
Gaussian-mixture regression.

### Kernel regression: from Nadaraya-Watson to a conditioned KDE

**Problem.** Predict $y$ from $x$ with no global functional form, a purely
local smoother. **Usual tool.** Nadaraya-Watson kernel regression
(`stats::ksmooth`, package `np`): a kernel-weighted local average of the
responses. **proxymix route.** Place one Gaussian per data point over the
joint $(y, x)$ -- a kernel density estimate -- and read $E[y \mid x]$ by
conditioning. With a Gaussian kernel that conditional mean *is* the
Nadaraya-Watson estimator, exactly. Where `lm()` sat at $K = 1$, one global
line, Nadaraya-Watson sits at $K = n$, one component per datum and fully
local, and `from_kde()` lands anywhere in between: a single axis with least
squares at one end and kernel smoothing at the other.

```{r nw-fit}
h <- 0.4                                     # kernel bandwidth
nw <- function(xq) {
  vapply(xq, function(q) {
    w <- dnorm(q, x, h)                      # Nadaraya-Watson, Gaussian kernel
    sum(w * y) / sum(w)
  }, numeric(1L))
}

## One Gaussian per datum -- a kernel density estimate of the joint (y, x) --
## then condition on x and read the conditional mean.
kde <- gmm(weights = rep(1 / n, n),
           means = lapply(seq_len(n), function(i) c(y[i], x[i])),
           covariances = rep(list(diag(c(h^2, h^2))), n))
xq <- seq(-2.5, 2.5, length.out = 21L)
diff_nw <- max(abs(nw(xq) - cond_mean(kde, xq)))
```

```{r fig-nw, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "One axis, two endpoints: the global line at K = 1 (ordinary least squares) and the fully local kernel smoother at K = n (Nadaraya-Watson), both read off a mixture by the same conditioning operation.", fig.alt = "Scatter of y against x with the straight ordinary-least-squares line and the curved Nadaraya-Watson kernel-regression line."}
grid_nw <- data.frame(x = seq(-3, 3, length.out = 200L))
grid_nw$ols <- as.numeric(predict(lm(y ~ x, dat), newdata = grid_nw))
grid_nw$nw <- nw(grid_nw$x)
ggplot2::ggplot() +
  ggplot2::geom_point(data = dat, ggplot2::aes(x, y), colour = "grey60",
                      alpha = 0.4, size = 0.7) +
  ggplot2::geom_line(data = grid_nw,
                     ggplot2::aes(x, ols, colour = "OLS (K = 1)"),
                     linewidth = 0.9) +
  ggplot2::geom_line(data = grid_nw,
                     ggplot2::aes(x, nw, colour = "kernel (K = n)"),
                     linewidth = 0.9) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("OLS (K = 1)" = "#0072B2", "kernel (K = n)" = "#D55E00")
  ) +
  ggplot2::labs(
    x = "x", y = "y",
    title = "From global line to local kernel, one conditioning step"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-nw-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the kernel-regression",
    "figure is skipped.\n")
```

**Idea.** Conditioning a per-point kernel density estimate gives
responsibilities $\pi_i(x) \propto \mathcal{N}(x \mid x_i, h^2)$ and
per-component means $y_i$, so
$E[y \mid x] = \sum_i \pi_i(x)\, y_i$, the Nadaraya-Watson ratio. Kernel
regression is therefore Gaussian-mixture regression with one component per
datum, a diagonal bandwidth, and only the conditional mean read off. The
bandwidth $h$ plays the role of the component covariance.

### Clustering: from `kmeans()` to the mixture components

**Problem.** Group rows by similarity. **Usual tool.** `kmeans()`, or
`mclust`, which fits exactly this model. **proxymix route.**
`fit_proxymix(N = K, regime = "sample")`: the components *are* the clusters
and the responsibilities are soft assignments.

```{r clust-fit}
set.seed(20260617)
x_clust <- rbind(
  mvnfast::rmvn(150L, c(-2, -1), 0.5 * diag(2)),
  mvnfast::rmvn(150L, c(2, 0), matrix(c(0.6, 0.3, 0.3, 0.4), 2L)),
  mvnfast::rmvn(150L, c(0, 2.5), 0.3 * diag(2))
)
colnames(x_clust) <- c("V1", "V2")
target_clust <- gmm_target_from_samples(x_clust)
fitc <- fit_proxymix(target_clust, N = 3L, regime = "sample",
                     max_iter = 150L)

## Soft assignment = posterior responsibility of each component.
responsibilities <- function(fit, xx) {
  comp <- vapply(seq_len(gmm_n_components(fit)), function(k) {
    fit@weights[k] *
      mvnfast::dmvn(xx, mu = fit@means[[k]], sigma = fit@covariances[[k]])
  }, numeric(nrow(xx)))
  comp / rowSums(comp)
}
resp <- responsibilities(fitc, x_clust)
mean_confidence <- mean(apply(resp, 1L, max))
```

```{r clust-table, echo = FALSE}
knitr::kable(
  head(round(resp, 3L), 4L),
  col.names = paste("component", seq_len(3L)),
  caption = paste0(
    "Soft cluster assignments for the first four rows: the posterior ",
    "responsibility of each of the three components."
  )
)
```

**Idea.** Soft, model-based clustering: a point's assignment to cluster $k$
is its posterior responsibility
$\pi_k\,\mathcal{N}(x \mid \mu_k, \Sigma_k) /
\sum_j \pi_j\,\mathcal{N}(x \mid \mu_j, \Sigma_j)$.

### Principal components: from `prcomp()` to the fitted covariance

**Problem.** Find the directions of greatest variation. **Usual tool.**
`prcomp()`. **proxymix route.** Fit a single Gaussian (`N = 1`,
`regime = "moment"`) and eigendecompose its covariance.

```{r pca}
fit_pca <- fit_proxymix(target_clust, N = 1L, regime = "moment",
                        ridge_eps = 0)
ev <- eigen(fit_pca@covariances[[1L]])$vectors
pr <- prcomp(x_clust)$rotation
## Identical to the prcomp rotation, up to the sign convention of each axis.
diff_pca <- max(abs(abs(ev) - abs(unname(pr))))
```

```{r fig-pca, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "Clustering (colour) and the principal axes (arrows) read off the same data by two different component counts: three components give the clusters, one component gives the covariance whose eigenvectors are the principal axes.", fig.alt = "Three coloured point clusters with two principal-axis arrows drawn from the overall centroid."}
vals <- eigen(fit_pca@covariances[[1L]])$values
mu_pca <- fit_pca@means[[1L]]
axes <- data.frame(
  x = mu_pca[1L], y = mu_pca[2L],
  xend = mu_pca[1L] + 2 * sqrt(vals) * ev[1L, ],
  yend = mu_pca[2L] + 2 * sqrt(vals) * ev[2L, ]
)
pts <- data.frame(x_clust, cluster = factor(max.col(resp)))
ggplot2::ggplot() +
  ggplot2::geom_point(data = pts,
                      ggplot2::aes(V1, V2, colour = cluster),
                      alpha = 0.6, size = 0.9) +
  ggplot2::geom_segment(
    data = axes, ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.2, "cm")), linewidth = 0.8
  ) +
  ggplot2::scale_colour_viridis_d(name = "cluster", end = 0.85) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = expression(x[1]), y = expression(x[2]),
                title = "Clustering and principal axes from one mixture") +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-pca-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the clustering and",
    "principal-axes figure is skipped.\n")
```

**Idea.** Principal-component analysis is the eigendecomposition of the
covariance. The moment-matched single Gaussian carries the sample
covariance exactly, so its eigenvectors are the principal components; with
$K > 1$ the same read gives a local analysis inside each cluster.

### Penalised regression: ridge as a covariance ridge

**Problem.** Stabilise a regression when predictors are collinear. **Usual
tool.** `glmnet` (ridge or lasso), `MASS::lm.ridge`. **proxymix route.** The
`ridge_eps` argument adds $\lambda I$ to the component covariance, which
shrinks the conditional regression: exactly an $L_2$ penalty.

```{r ridge}
lambda <- c(0, 0.5, 2, 8)
slope_pm <- vapply(lambda, function(lam) {
  f <- fit_proxymix(joint, N = 1L, regime = "moment", ridge_eps = lam)
  gmm_conditionalise(f, given = c(NA, 1))@means[[1L]] -
    gmm_conditionalise(f, given = c(NA, 0))@means[[1L]]
}, numeric(1L))
slope_formula <- cov(x, y) / (var(x) + lambda)
diff_ridge <- max(abs(slope_pm - slope_formula))
```

```{r ridge-table, echo = FALSE}
knitr::kable(
  data.frame(lambda = lambda, proxymix = slope_pm,
             ridge_formula = slope_formula),
  digits = 4L,
  col.names = c("Penalty (lambda)", "proxymix slope",
                "Ridge closed form"),
  caption = paste0(
    "The conditional slope under a covariance ridge, against the ridge ",
    "estimator cov(x, y) / (var(x) + lambda)."
  )
)
```

**Idea.** Conditioning gives $\beta = \Sigma_{xx}^{-1}\Sigma_{xy}$. Ridging
$\Sigma_{xx} \to \Sigma_{xx} + \lambda I$ yields
$(\Sigma_{xx} + \lambda I)^{-1}\Sigma_{xy}$, the ridge estimator, so the
slope shrinks toward zero as $\lambda$ grows.

### Treatment effects: from a two-model learner to one joint fit

**Problem.** Estimate a treatment effect that varies from unit to unit, and
then act on it. **Usual tool.** A two-model T-learner, fitting the treated
and control arms separately and subtracting, or a causal forest (`grf`) or
`DoubleML`. **proxymix route.** `fit_uplift()` fits one joint mixture over
(outcome, treatment, covariates), and every decision verb reads that single
fit in closed form. Nothing re-fits per query.

The design below is synthetic, chosen so that the truth is available as an
oracle: the treatment effect is the known linear function
$\tau(x) = 0.5 + x$, so the estimate can be scored rather than admired.

```{r uplift-fit}
set.seed(20260902)
n_up <- 600L
x_up <- rnorm(n_up)
t_up <- rbinom(n_up, 1L, 0.5)
tau_true <- function(v) 0.5 + v
y_up <- 1 + tau_true(x_up) * t_up + rnorm(n_up, sd = 0.5)
dat_up <- data.frame(y = y_up, t = t_up, x = x_up)

model <- fit_uplift(dat_up, "y", "t", "x", N = 2L, regime = "sample",
                    max_iter = 80L, seed = 1L)
model
```

`proxy_cate()` reads the conditional average treatment effect, with a
standard error and an interval, at whatever units are handed to it.

```{r uplift-cate}
grid_up <- data.frame(x = seq(-2, 2, length.out = 41L))
cate <- proxy_cate(model, grid_up)
err_cate <- max(abs(cate$tau - tau_true(grid_up$x)))
cover_cate <- mean(tau_true(grid_up$x) >= cate$ci_lo &
                     tau_true(grid_up$x) <= cate$ci_hi)
```

```{r fig-cate, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "The conditional average treatment effect read off one joint mixture, with its delta-method interval, against the analytic truth that generated the data. The estimate tracks the true effect across the covariate and is shrunk slightly toward zero at both tails.", fig.alt = "Estimated conditional average treatment effect against the covariate, shown as a line with a shaded confidence band, overlaid with a dashed line for the analytic truth; the two nearly coincide and separate slightly at the extremes."}
cate_df <- as.data.frame(cate)
cate_df$x <- grid_up$x
cate_df$truth <- tau_true(grid_up$x)
ggplot2::ggplot(cate_df, ggplot2::aes(x)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi),
                       fill = "#56B4E9", alpha = 0.35) +
  ggplot2::geom_line(ggplot2::aes(y = tau, colour = "proxymix estimate"),
                     linewidth = 0.9) +
  ggplot2::geom_line(ggplot2::aes(y = truth, colour = "analytic truth"),
                     linewidth = 0.9, linetype = "dashed") +
  ggplot2::geom_hline(yintercept = 0, colour = "grey60",
                      linewidth = 0.3) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("proxymix estimate" = "#0072B2",
               "analytic truth" = "#D55E00")
  ) +
  ggplot2::labs(
    x = "covariate x", y = "treatment effect on y",
    title = "A conditional treatment-effect surface from one joint fit"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-cate-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the treatment-effect",
    "figure is skipped.\n")
```

Acting on the estimate is `proxy_decide()`, which turns an effect into a
recommended arm under a value per unit of outcome and a cost of treating.

```{r uplift-decide}
decision <- proxy_decide(model, grid_up, value = 1, cost = 0.5)
switch_x <- grid_up$x[min(which(decision$action == 1L))]
```

```{r uplift-decide-table, echo = FALSE}
sel <- c(1L, 11L, 21L, 31L, 41L)
knitr::kable(
  data.frame(
    x = grid_up$x[sel],
    tau = cate$tau[sel],
    truth = tau_true(grid_up$x[sel]),
    action = decision$action[sel],
    expected_value = decision$expected_value[sel]
  ),
  digits = 3L,
  col.names = c("Covariate x", "Estimated effect", "True effect",
                "Recommended arm", "Expected value"),
  caption = paste0(
    "Five points of the decision surface at a value of 1 per unit of ",
    "outcome and a treatment cost of 0.5, so treating pays when the ",
    "effect exceeds 0.5."
  )
)
```

**A refusal rather than a wrong scale.** The model recorded the treatment
arms it actually saw. Asking for a contrast between arms the data never
contained is refused, with a typed condition, instead of being answered on a
silently rescaled axis.

```{r uplift-refusal}
refusal <- tryCatch(
  proxy_cate(model, grid_up, t1 = 100, t0 = 0),
  error = function(e) e
)
class(refusal)
conditionMessage(refusal)
```

The identification report is the artefact that travels with a decision: it
names the estimand, the assumption it rests on, the overlap of the units
supplied, the magnitude at risk if a latent regime confounds the effect, and
what the model explicitly does not identify.

```{r uplift-report}
proxy_identification_report(model, grid_up)
```

**Idea.** Under ignorability the conditional treatment effect is a contrast
of two conditional means of the same joint mixture, $E[y \mid t_1, x]$ minus
$E[y \mid t_0, x]$, so the operator that produced the regression above
produces the causal contrast as well. What changes is not the algebra but
the assumption the reader is asked to accept, which is why the family ships
a report that states it.

### A few more, in one line each

Imputation is `gmm_conditionalise()` on the observed coordinates, giving a
conditional-mean or, through `rgmm()`, a sampled fill-in. Density
estimation is the native use, and `from_kde()` compresses a kernel estimate
into a closed-form mixture. Kalman filtering is `gmm_observe()`, the mixture
form of the Kalman update.

## Read

Four of the six substitutions reproduce the classical answer to machine
precision, which is the part worth checking rather than assuming.

```{r equality-table, echo = FALSE}
knitr::kable(
  data.frame(
    claim = c(
      "K = 1 conditional slope equals the lm coefficient",
      "Per-point KDE conditional mean equals Nadaraya-Watson",
      "N = 1 eigenvectors equal the prcomp rotation",
      "Ridged conditional slope equals the ridge closed form"
    ),
    difference = sci_plain(c(diff_reg, diff_nw, diff_pca, diff_ridge)),
    stringsAsFactors = FALSE
  ),
  col.names = c("Claim", "Maximum absolute difference"),
  caption = paste0(
    "Each equality claim checked against the tool it stands in for, on ",
    "the data fitted above."
  )
)
```

**Regression.** At one component the conditional slope and the `lm`
coefficient differ by `r sci(diff_reg)`, which is the double-precision
floor: the substitution is an identity, not an approximation. At three
components the same conditioning operation follows the bend in the figure
that the single line cannot. What this buys is nonlinear, heteroscedastic,
even multimodal conditional *densities* rather than only a mean. What it
costs is classical inference: no standard errors or $p$-values come back by
default, and the components are Gaussian.

**Kernel regression.** The conditioned per-point kernel density estimate and
the Nadaraya-Watson smoother agree to `r sci(diff_nw)` over the
`r length(xq)` query points. The same operation also returns the full
conditional density -- predictive variance, quantiles, multimodality --
which the bare Nadaraya-Watson mean does not, and it works in the
evaluate-only regime where there are no points to average and so no literal
kernel estimator exists. The price is that the one-per-datum estimate costs
$O(n)$ per query, which `from_kde()` compresses to $O(K)$, and that the
bandwidth must still be chosen.

**Clustering.** The three components recover the three generated clouds, and
the assignments are nearly hard: the mean largest responsibility across the
`r nrow(x_clust)` rows is `r round(mean_confidence, 3)`, so on
well-separated data the soft assignment is decisive. The gain over
`kmeans()` is ellipsoidal rather than spherical clusters, a soft assignment
that carries its own uncertainty, and a density estimate as a by-product.
The cost is that the component count is not chosen for you, which is what
`bic_aic()` is for, and that `kmeans()` is faster at very large $n$.

**Principal components.** The eigenvectors of the one-component covariance
and the `prcomp` rotation differ by `r sci(diff_pca)` once the sign
convention of each axis is accounted for. At $K > 1$ the same read gives a
local analysis inside each cluster, which `prcomp()` does not offer. For a
single linear projection `prcomp()` remains more direct, since loadings,
scree plots and biplots come with it.

**Ridge.** The slope under a covariance ridge matches the closed-form ridge
estimator to `r sci(diff_ridge)` across all
`r length(lambda)` penalties, and shrinks from
`r round(slope_pm[1], 3)` at no penalty to
`r round(slope_pm[length(lambda)], 3)` at the largest. What is delivered is
$L_2$ shrinkage and numerical stability from the object already in hand.
What is not available is the lasso: $L_2$ shrinkage matches a Gaussian
prior, $L_1$ sparsity a Laplace prior outside the Gaussian family, so this
route does not perform variable selection.

**Treatment effects.** This is the one substitution that is an
approximation rather than an identity, and the figure shows how good an
approximation. Across the covariate range the estimated effect departs from
the analytic truth by at most `r round(err_cate, 3)`, on an effect that runs
from `r round(tau_true(min(grid_up$x)), 1)` to
`r round(tau_true(max(grid_up$x)), 1)`, with the estimate shrunk slightly
toward zero at both tails because two Gaussian components stand in for a
linear interaction. Turned into an action at a value of 1 and a cost of 0.5,
the recommendation switches from control to treatment at
`r round(switch_x, 2)`, against the true break-even at 0. The delta-method
interval covers the truth at `r round(100 * cover_cate)` per cent of the
grid points, short of its nominal 95 per cent, because it prices sampling
variability and not the approximation bias just described.

The refusal is the other half of that section. Asking for a contrast
between treatment values of 100 and 0, on a model fitted to arms 0 and 1,
returns a condition of class `proxymix_treatment_scale_error` and
`orchestra_refusal` rather than a number. The contract is that a decision
verb declines a question the fit cannot support instead of answering it on
an axis the data never contained, which is the difference between an effect
that is wrong and an effect that is absent.

```{r summary-table, echo = FALSE}
knitr::kable(
  data.frame(
    method = c("Regression", "Kernel regression", "Clustering",
               "Principal components", "Ridge", "Treatment effects"),
    usual = c("lm / glm", "ksmooth / np", "kmeans / mclust", "prcomp",
              "glmnet / lm.ridge", "T-learner / grf / DoubleML"),
    route = c("mixture over (y, x) plus gmm_conditionalise()",
              "per-point KDE plus gmm_conditionalise()",
              "fit_proxymix(regime = 'sample')",
              "eigen() of the N = 1 covariance",
              "ridge_eps on the covariance",
              "fit_uplift() plus proxy_cate()"),
    gain = c("nonlinear, heteroscedastic, multimodal conditionals",
             "full conditional density; the evaluate-only regime",
             "soft, ellipsoidal, density-aware clusters",
             "a local analysis per cluster",
             "L2 shrinkage from one object",
             "one fit, plus a stated identification contract"),
    give_up = c("built-in inference; Gaussian components",
                "O(n) per query unless compressed; bandwidth choice",
                "automatic component count; speed at huge n",
                "loadings, scree and biplot conveniences",
                "L1 sparsity and variable selection",
                "exactness; the interval under-covers"),
    stringsAsFactors = FALSE
  ),
  col.names = c("Method", "Usual tool", "proxymix route", "What you gain",
                "What you give up"),
  caption = "Six analyses from one fitted object, and the cost of each."
)
```

The answer to the opening question, then: one fitted mixture genuinely
replaces the linear model, the kernel smoother, the clustering, the
principal-components step and the ridge penalty, four of them exactly, and
it additionally supplies a treatment-effect reading that specialised causal
packages would otherwise provide. What is given up is inference machinery,
the freedom of a non-Gaussian component shape, and, in the causal case,
exactness.

## Limits

Nothing here shows the mixture beating the specialised tool. Four of the
substitutions are identities, so they cannot be better, only more
convenient; the remaining two trade accuracy for having one object serve
several jobs. Use the purpose-built tool where its strengths are what is
needed: high-dimensional sparse problems suit `glmnet`, very large-$n$
clustering suits `kmeans()`, a single linear projection suits `prcomp()`,
and classical inferential guarantees suit `lm()` or `glm()`. The mixture
earns its place when one fitted object should serve several of these at
once, in closed form, and when Gaussian components fit the data.

Every fit on this page is in the sample regime, with samples available and
component counts supplied by hand rather than selected. Choosing $K$ is a
real decision that `bic_aic()` informs and this vignette sidesteps, and a
count much above the number of genuine structures will fit noise into the
conditional law.

The treatment-effect section carries two further limits. First, the
interval under-covers, as the reading above records, because the
delta-method standard error holds the regime gate fixed and prices sampling
variability only; a resampling alternative is available through
`se_method = "mc"` and does not remove the approximation bias, which is what
the shortfall here reflects. Second, with a binary treatment the fitted
mixture separates on the treatment arm itself, so within any one regime the
treatment does not vary and the within-regime treatment slope is zero. This
means `proxy_regime_segments()` returns an effect of zero in every regime on
data of this shape: the effect modification is carried *across* regimes, by
the gate, and is read correctly by `proxy_cate()` but not by the segment
table. A reader wanting interpretable per-segment effects needs a design
where the components separate on the covariate rather than on the arm.

Everything causal here also rests on ignorability, which no goodness-of-fit
statistic can certify, on data generated to satisfy it. The identification
report states the assumption and the magnitude at risk if it fails, and
declines the individual counterfactual law outright; it does not make the
assumption true.

## What to read next

*Choosing between the three fitting regimes* explains what the `"moment"`,
`"sample"` and `"kld"` arguments used throughout this page actually do, and
when each applies.

*Compressing a kernel density estimate into a mixture* is the compression
step that turns the $O(n)$ kernel smoother above into an $O(K)$ one.

*The closed-form operator calculus on a mixture* covers the pushforward and
Kalman-update operators that the one-line extras above only name.

*Imputing missing data with a mixture* takes the conditioning operator used
throughout this page and applies it to holes in the data.

## References

* Athey, S. and Imbens, G. (2016). *Recursive partitioning for
  heterogeneous causal effects.* Proceedings of the National Academy of
  Sciences 113(27), 7353--7360. <doi:10.1073/pnas.1510489113>.
* de Veaux, R. D. (1989). *Mixtures of linear regressions.* Computational
  Statistics & Data Analysis 8(3), 227--245.
* Fraley, C. and Raftery, A. E. (2002). *Model-based clustering,
  discriminant analysis, and density estimation.* Journal of the American
  Statistical Association 97(458), 611--631.
* Jolliffe, I. T. (2002). *Principal Component Analysis*, 2nd ed. Springer.
* Nadaraya, E. A. (1964). *On estimating regression.* Theory of Probability
  and Its Applications 9(1), 141--142.
* Watson, G. S. (1964). *Smooth regression analysis.* Sankhya A 26(4),
  359--372.
* Tipping, M. E. and Bishop, C. M. (1999). *Probabilistic principal
  component analysis.* Journal of the Royal Statistical Society B 61(3),
  611--622.
* Hoerl, A. E. and Kennard, R. W. (1970). *Ridge regression: biased
  estimation for nonorthogonal problems.* Technometrics 12(1), 55--67.
* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.

## Reproduce

Two seeds are used: `20260617` for the regression, clustering and
principal-components data, set in the chunk that builds each, and
`20260902` for the treatment-effect design. `fit_uplift()` is additionally
given `seed = 1L` so its expectation-maximisation start is reproducible
without disturbing the ambient random-number state.

```{r session-info}
sessionInfo()
```
