---
title: "Imputing missing data with a mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputing missing data with a mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~4 s under rmarkdown::render() with ggplot2 and mice
     installed; macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_mice <- requireNamespace("mice", quietly = TRUE)
```

## Why

An analyst has two measurements per unit. The scatter falls into two clear
clusters with almost nothing between them, and a good half of the second
measurement never came back from the field. The question is not whether an
imputer will run. It is: *my data has two clusters and an empty gap between
them, so if I let a standard imputer fill the holes, will it put values in
the gap where I know there is no mass, and will that move the mean I am
going to report?*

This vignette answers both halves on a synthetic two-cluster toy where the
values that went missing are retained, so every imputation can be scored
against the truth it was hiding.

## What

Imputation here is conditioning. If the joint distribution of the data is a
Gaussian mixture, the conditional distribution of a missing entry given the
observed entries of its own row is itself a Gaussian mixture, available in
closed form through the Schur-complement algebra of
[`gmm_conditionalise()`]. Nothing is simulated to get it.

[`gmm_impute()`] fits that mixture to a matrix containing holes and draws `m`
completed datasets from the per-row conditionals. Its `N` argument is the
number of mixture components, chosen by the Bayesian information criterion
when left at its default. A mixture can be multimodal and heteroscedastic,
which is the point: it represents shapes that a single-Gaussian model, or a
linear-Gaussian conditional such as normal regression imputation, cannot,
so the imputations stay faithful to data those models mis-specify.

[`gmm_complete()`] extracts one completed dataset. [`proxy_pool()`] combines
an estimand across the completions, and [`proxy_fmi()`] reports the fraction
of missing information. For a column mean the pooling is closed form -- the
exact large-sample limit of the between-imputation variance -- so no Monte
Carlo noise enters at the pooling step. For any other estimand,
[`as_mids()`] packages the completions as a `mice` object so the joint
mixture imputations flow into `mice::pool()` unchanged: proxymix supplies
the imputation model, `mice` the pooling.

## Do

### A two-cluster toy with the truth retained

Two groups, each a tight cloud, with a positive within-group association
between `x1` and `x2`. This is a synthetic toy: the complete data is
generated first and kept, then part of `x2` is deleted, so the held-out
values are available to score against. The chance of an entry going missing
depends on the observed `x1` and not on the hidden `x2` itself, which is the
missing-at-random setting.

```{r data}
set.seed(20260620)
n <- 600L
lab <- sample(c(-1, 1), n, replace = TRUE)
x1 <- 2 * lab + rnorm(n, 0, 0.6)
x2 <- 2 * lab + 0.5 * (x1 - 2 * lab) + rnorm(n, 0, 0.6)
truth <- cbind(x1 = x1, x2 = x2)

x_holes <- truth
missing <- runif(n) < plogis(0.6 * x1)
x_holes[missing, "x2"] <- NA
frac_missing <- mean(missing)
```

The marginal distribution of `x2` is bimodal, one cluster near $-2$ and one
near $+2$, with little mass between them. Writing the empty middle as the
interval $|x_2| < 1$, the values that were deleted put only a small share of
their mass there.

```{r gap-truth}
in_gap <- function(v) mean(abs(v) < 1)
gap_truth <- in_gap(truth[missing, "x2"])
```

### Impute with a mixture, and with a single Gaussian

```{r impute}
imp <- gmm_impute(x_holes, N = 2L, m = 20L, seed = 1L)
imp
```

A single completed dataset has no holes:

```{r complete}
done <- gmm_complete(imp, 1L)
anyNA(done)
```

Forcing one component reproduces the multivariate-normal assumption, which
is the contrast this vignette is built around.

```{r single}
imp1 <- gmm_impute(x_holes, N = 1L, m = 20L, seed = 1L)
done1 <- gmm_complete(imp1, 1L)
```

```{r gap-table, echo = FALSE}
gap_tbl <- data.frame(
  source = c("held-out truth", "mixture imputation (N = 2)",
             "single-Gaussian imputation (N = 1)"),
  gap_share = c(gap_truth, in_gap(done[missing, "x2"]),
                in_gap(done1[missing, "x2"])),
  stringsAsFactors = FALSE
)
knitr::kable(
  gap_tbl, digits = 3L,
  col.names = c("Source", "Share of values with |x2| < 1"),
  caption = paste0(
    "How much mass each imputer places in the empty middle between the ",
    "two clusters, over the ", sum(missing), " deleted entries."
  )
)
```

### The shape of the imputations

```{r fig-modes, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "Density of the deleted x2 values against the values each imputer supplied in their place. The mixture imputation reproduces both clusters; the single-Gaussian imputation collapses onto one central peak and fills the empty middle, where the truth has almost no mass.", fig.alt = "Three density curves over x2: the held-out truth and the mixture imputation are both bimodal and overlap closely, while the single-Gaussian imputation is unimodal and centred on the gap between the two modes."}
dens_df <- function(v, label) {
  d <- density(v)
  data.frame(x2 = d$x, density = d$y, source = label,
             stringsAsFactors = FALSE)
}
plot_df <- rbind(
  dens_df(truth[missing, "x2"], "held-out truth"),
  dens_df(done[missing, "x2"], "mixture (N = 2)"),
  dens_df(done1[missing, "x2"], "single Gaussian (N = 1)")
)
plot_df$source <- factor(
  plot_df$source,
  levels = c("held-out truth", "mixture (N = 2)",
             "single Gaussian (N = 1)")
)
ggplot2::ggplot(plot_df, ggplot2::aes(x2, density, colour = source)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::annotate("rect", xmin = -1, xmax = 1, ymin = -Inf, ymax = Inf,
                    fill = "grey60", alpha = 0.18) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("held-out truth" = "#000000",
               "mixture (N = 2)" = "#009E73",
               "single Gaussian (N = 1)" = "#D55E00")
  ) +
  ggplot2::labs(
    x = expression(x[2]), y = "density",
    title = "What each imputer put where the data was deleted",
    subtitle = "Shaded band marks the empty middle, |x2| < 1"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-modes-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the imputation-density",
    "figure is skipped.\n")
```

### Pooling a column mean

```{r pool-mean}
pooled <- proxy_pool(imp, "x2")
pooled1 <- proxy_pool(imp1, "x2")
fmi_mix <- proxy_fmi(imp, "x2")
mean_truth <- mean(truth[, "x2"])
err_mix <- abs(pooled$estimate - mean_truth)
err_one <- abs(pooled1$estimate - mean_truth)
```

```{r fixed-notation, include = FALSE}
## Small errors are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
```

```{r pool-table, echo = FALSE}
pool_tbl <- data.frame(
  model = c("complete-data truth", "available cases only",
            "mixture imputation (N = 2)",
            "single-Gaussian imputation (N = 1)"),
  estimate = c(mean(truth[, "x2"]), mean(x_holes[!missing, "x2"]),
               pooled$estimate, pooled1$estimate),
  std_error = c(NA_real_, NA_real_, pooled$std.error, pooled1$std.error),
  conf_low = c(NA_real_, NA_real_, pooled$conf.low, pooled1$conf.low),
  conf_high = c(NA_real_, NA_real_, pooled$conf.high, pooled1$conf.high),
  stringsAsFactors = FALSE
)
pool_tbl$abs_error <- abs(pool_tbl$estimate - mean(truth[, "x2"]))
knitr::kable(
  pool_tbl, digits = 4L,
  col.names = c("Estimate of the mean of x2", "Estimate", "SE",
                "CI lower", "CI upper", "Absolute error"),
  caption = paste0(
    "The mean of x2 recovered four ways, scored against the ",
    "complete-data truth that the deletion hid."
  )
)
```

### Any other estimand, through `mice`

```{r pool-mice, eval = has_mice, echo = has_mice}
mice_fit <- mice::pool(with(as_mids(imp), lm(x2 ~ x1)))
```

```{r pool-mice-table, eval = has_mice, echo = FALSE}
knitr::kable(
  summary(mice_fit), digits = 3L,
  caption = paste0(
    "The regression of x2 on x1, fitted in each of the ", imp@m,
    " completions and pooled by mice::pool()."
  )
)
```

```{r pool-mice-skip, eval = !has_mice, echo = FALSE, results = "asis"}
cat("mice is not installed on this build, so the pooled regression is",
    "skipped; the closed-form column-mean pooling above does not need it.\n")
```

## Read

`r round(100 * frac_missing)` per cent of `x2` was deleted, which is
`r sum(missing)` of the `r n` rows. Over those deleted entries the truth
placed `r round(100 * gap_truth, 1)` per cent of its mass in the empty
middle $|x_2| < 1$. The mixture imputer put
`r round(100 * in_gap(done[missing, "x2"]), 1)` per cent there, close to the
truth. Forcing a single Gaussian put
`r round(100 * in_gap(done1[missing, "x2"]), 1)` per cent there, roughly
`r round(in_gap(done1[missing, "x2"]) / gap_truth, 1)` times as much as the
data ever had: the answer to the opening question is that a single-Gaussian
imputer does fill the gap, and a mixture does not.

The figure shows why. A single Gaussian has one centre and one spread, so
fitted to a bimodal column it places its centre between the two clusters and
imputes values where there was never any true mass. The mixture keeps two
components and reproduces both modes.

That shape error reaches the number an analyst would report. The
complete-data mean of `x2` is `r fixed(mean_truth, 3)`. Using only the rows
that came back gives `r fixed(mean(x_holes[!missing, "x2"]), 3)`, which is
out by `r fixed(abs(mean(x_holes[!missing, "x2"]) - mean_truth), 3)`,
because missingness here is tied to `x1` and `x1` is tied to `x2`. The
mixture imputation recovers the mean to within `r fixed(err_mix, 4)`; the
single-Gaussian imputation is out by `r fixed(err_one, 4)`, about
`r round(err_one / err_mix)` times further. Both imputed intervals cover the
truth, so on this estimand the single Gaussian is inefficient rather than
misleading; the damage it does is to the shape, which the figure makes
visible and the mean does not.

The fraction of missing information is
`r round(unname(fmi_mix), 3)`: although
`r round(100 * frac_missing)` per cent of the column is absent, the
observed `x1` carries enough about `x2` that only about
`r round(100 * unname(fmi_mix))` per cent of the information about the mean
is lost.

## Limits

This vignette covers numeric data that is missing at random: the probability
that an entry is missing may depend on the observed entries but not on the
missing value itself. That assumption is what makes the conditional of the
observed the right law to impute from, and it is not testable from the data
in hand. Value-dependent missingness and censoring at a known bound break
it, and both are supported through the `mechanism` argument rather than
being outside the package.

The comparison here is against a single Gaussian, which is the
multivariate-normal imputation model. It is not a comparison against a
well-specified alternative such as predictive mean matching or a
classification-and-regression-tree imputer, either of which can also
represent two clusters, so nothing here establishes that a mixture beats
every competitor -- only that it beats the Gaussian assumption on data the
Gaussian assumption mis-specifies. A second limit is scale: one column of
one bivariate toy at `n` = `r n`, with the component count supplied rather
than selected, so the result speaks to the shape of the imputation law and
not to behaviour in wide data or with many incomplete columns at once.

The estimand shown in closed form is a column mean. Anything else is pooled
by `mice`, whose rules are the Rubin large-sample rules and inherit their
assumptions. What would change the answer most is the number of components:
at `N` = 1 the imputer is the Gaussian model this vignette argues against,
and a component count chosen well above the number of genuine clusters
would fit noise into the conditional law.

## What to read next

*Missing data that depends on the missing value* takes up the two
mechanisms this vignette assumes away, censoring at a known bound and
missingness whose probability depends on the hidden value itself, and shows
how a sensitivity sweep brackets an answer that the data alone cannot
identify.

*The closed-form operator calculus on a mixture* is the algebra underneath
`gmm_conditionalise()`, and shows what else the same conditioning step buys
once a mixture is in hand.

*One mixture, many methods* places imputation beside the other classical
analyses that fall out of a single fitted mixture.

## References

* Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys.*
  Wiley.
* van Buuren, S. and Groothuis-Oudshoorn, K. (2011). *mice: Multivariate
  imputation by chained equations in R.* Journal of Statistical Software
  45(3), 1--67. <doi:10.18637/jss.v045.i03>.
* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.

## Reproduce

The data-generating seed is `20260620`, set in the chunk that builds the
toy; `gmm_impute()` is given `seed = 1L`, which makes the completions
reproducible without disturbing the ambient random-number state.

```{r session-info}
sessionInfo()
```
