---
title: "Missing data that depends on the missing value"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing data that depends on the missing value}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~18 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. The sensitivity sweep
     is the cost: five slope values at m = 10 completions each. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

An analyst has a column where the large values are the ones that went
missing: an assay that saturates, a yield that was never recorded on the
paddocks that did badly, a follow-up that the worst cases did not attend.
The question is: *the bigger values are exactly the ones I did not get, so
the mean I am about to report is too small -- by how much, and how do I put
a defensible number on that when the data I would need to check is the data
I do not have?*

This vignette answers it on a synthetic toy where the deleted values are
retained, so every estimate can be scored against the complete-data truth
it was hiding.

## What

In proxymix you do not impute a missing entry so much as condition the
fitted mixture on what its being missing tells you. For an entry missing at
random that is nothing beyond the other coordinates, so the imputation law
is the plain mixture conditional. Two common situations carry more
information than that. An entry can be missing because it fell in a known
interval -- below a detection limit, above a ceiling -- which is
*censoring*. Or it can be missing with a probability that depends on its own
unobserved value, larger values dropping out more often, which is *missing
not at random*. Each is a gate $\delta(y)$ multiplying the mixture
conditional before the value is drawn.

[`mar()`], [`censored()`] and [`mnar()`] build those three gates, and
[`gmm_impute()`] takes one through its `mechanism` argument.
[`proxy_pool()`] combines an estimand across the completions.
[`proxy_mnar_sensitivity()`] is the verb for the case that is not
identified: it sweeps the assumed selection slope and reports the pooled
estimate with its interval at each value, so the reader sees the whole
curve rather than one number that hides its own assumption.

```{r gate-table, echo = FALSE}
knitr::kable(
  data.frame(
    Mechanism = c("missing at random", "censored",
                  "missing not at random"),
    Gate = c("constant", "indicator of an interval",
             "selection probability"),
    Constructor = c("mar()", "censored()", "mnar()"),
    Identified = c("yes, given the model", "yes, the bound is known",
                   "no, the slope is assumed"),
    stringsAsFactors = FALSE
  ),
  align = "llll",
  caption = paste0(
    "The three missingness gates, the shape each one multiplies into the ",
    "mixture conditional, and whether the data can identify it."
  )
)
```

## Do

### A value-dependent mechanism biases the ordinary fit

A synthetic two-component mixture in which the second coordinate `y` goes
missing with a probability that grows with `y` itself, so the larger values
are under-represented among the observed. The complete data is generated
first and kept, so the truth is available to score against.

```{r dgp}
set.seed(20260622)
n <- 600L
comp <- sample(1:2, n, replace = TRUE)
mu <- rbind(c(0, 0), c(1.5, 0.5))
chol_r <- chol(matrix(c(1, 0.6, 0.6, 1), 2L))
z_full <- matrix(rnorm(2 * n), n, 2L) %*% chol_r + mu[comp, ]
colnames(z_full) <- c("x1", "y")
truth <- mean(z_full[, 2L])

beta_true <- 0.7
miss <- runif(n) < plogis(-0.5 + beta_true * z_full[, 2L])
dat <- z_full
dat[miss, "y"] <- NA
```

An imputer that assumes the data are missing at random fills the gap from
the conditional of the observed, which is itself short of the large values,
so it stays biased. Supplying the selection slope to `mnar()` lets the fit
account for the mechanism. Both fits use `m` = 10 completions, which is
enough to separate the two mechanisms at this sample size and keeps the
vignette fast.

```{r recover}
m_draws <- 10L
mar_fit <- gmm_impute(dat, N = 2L, m = m_draws, mechanism = mar(),
                      seed = 1L)
mnar_fit <- gmm_impute(dat, N = 2L, m = m_draws,
                       mechanism = mnar("y", beta = beta_true), seed = 1L)
mar_est <- proxy_pool(mar_fit, "y")$estimate
mnar_est <- proxy_pool(mnar_fit, "y")$estimate
```

Analytic pooling is defined for the plain conditional but not for a gated
one, so the gated fits fall back to Rubin's rules and say so; the returned
`method` attribute records which was used.

```{r recover-table, echo = FALSE}
rec_tbl <- data.frame(
  estimator = c("complete-data truth", "available cases only",
                "missing at random", "missing not at random"),
  estimate = c(truth, mean(dat[!miss, "y"]), mar_est, mnar_est),
  stringsAsFactors = FALSE
)
rec_tbl$bias <- rec_tbl$estimate - truth
knitr::kable(
  rec_tbl, digits = 3L,
  col.names = c("Estimator of the mean of y", "Estimate", "Bias"),
  caption = paste0(
    "The mean of y recovered four ways, against the complete-data truth ",
    "the deletion hid. ", sum(miss), " of ", n, " values are missing."
  )
)
```

The slope `beta` is the strength of the dependence on the log-odds scale,
and `beta = 0` is the missing-at-random case. The intercept is not supplied:
it is calibrated to the observed fraction missing.

### The slope is a sensitivity parameter

A missing-not-at-random departure cannot be identified from the observed
data, so a single estimate would overstate what the data support. The
appropriate summary is a curve. `proxy_mnar_sensitivity()` sweeps the slope
and reports the pooled mean with its interval at each value, and the analyst
reads off where a conclusion would change. Five slope values from 0 to 1.2
are enough to show the shape.

```{r sweep}
sweep <- proxy_mnar_sensitivity(dat, "y",
                                beta_grid = seq(0, 1.2, by = 0.3),
                                N = 2L, m = m_draws, seed = 1L)
covers <- sweep$conf.low <= truth & sweep$conf.high >= truth
beta_first <- sweep$beta[min(which(covers))]
```

```{r sweep-table, echo = FALSE}
knitr::kable(
  as.data.frame(sweep)[, c("beta", "estimate", "conf.low", "conf.high")],
  digits = 3L,
  col.names = c("Assumed slope", "Pooled mean", "CI lower", "CI upper"),
  caption = paste0(
    "The pooled mean of y at each assumed selection slope. The ",
    "complete-data truth is ", round(truth, 3), "."
  )
)
```

```{r fig-tornado, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "The pooled mean of y rises with the assumed sensitivity slope, and its interval reaches the complete-data truth near the slope that actually generated the data. Sweeping the slope brackets an answer the observed data cannot identify on its own.", fig.alt = "Pooled mean of y against the assumed sensitivity slope, with a shaded confidence band, a marked point at each swept value, a dashed horizontal line at the complete-data truth, and a vertical rule at the generating slope."}
sweep_df <- as.data.frame(sweep)
ggplot2::ggplot(sweep_df, ggplot2::aes(beta, estimate)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = conf.low, ymax = conf.high),
    fill = "#56B4E9", alpha = 0.3
  ) +
  ggplot2::geom_line(colour = "#0072B2", linewidth = 0.9) +
  ggplot2::geom_point(colour = "#0072B2", size = 2) +
  ggplot2::geom_hline(yintercept = truth, linetype = "dashed",
                      colour = "#000000") +
  ggplot2::geom_vline(xintercept = beta_true, linetype = "dotted",
                      colour = "#D55E00") +
  ggplot2::annotate("text", x = min(sweep_df$beta), y = truth,
                    label = "complete-data truth", hjust = 0, vjust = -0.6,
                    size = 3.2) +
  ggplot2::annotate("text", x = beta_true, y = min(sweep_df$conf.low),
                    label = "generating slope", hjust = 1.05, vjust = 0,
                    size = 3.2, colour = "#D55E00") +
  ggplot2::labs(
    x = "assumed selection slope (log-odds per unit of y)",
    y = "pooled mean of y",
    title = "A sensitivity curve, not a single estimate"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-tornado-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the sensitivity-curve",
    "figure is skipped; the swept values are in the table above.\n")
```

### Censoring is the exact, identifiable case

When the reason for missingness is a known bound, the mechanism is
identified and there is no sensitivity parameter. A detection limit that
hides values of `y` below a threshold is `censored("y", upper = threshold)`,
and proxymix draws from the mixture conditional truncated to the censored
interval rather than substituting a constant such as half the limit.

```{r censor}
thr <- 0.3
cmiss <- z_full[, 2L] < thr
cdat <- z_full
cdat[cmiss, "y"] <- NA
cfit <- gmm_impute(cdat, N = 2L, m = m_draws,
                   mechanism = censored("y", upper = thr), seed = 1L)
cens_est <- proxy_pool(cfit, "y")$estimate
```

```{r censor-table, echo = FALSE}
cens_tbl <- data.frame(
  estimator = c("complete-data truth", "available cases only",
                "half the detection limit", "proxymix, truncated draw"),
  estimate = c(truth, mean(cdat[!cmiss, "y"]),
               mean(ifelse(cmiss, thr / 2, z_full[, 2L])), cens_est),
  stringsAsFactors = FALSE
)
cens_tbl$bias <- cens_tbl$estimate - truth
knitr::kable(
  cens_tbl, digits = 3L,
  col.names = c("Estimator of the mean of y", "Estimate", "Bias"),
  caption = paste0(
    "The mean of y under a detection limit at ", thr, ", which hides ",
    sum(cmiss), " of ", n, " values."
  )
)
```

## Read

`r sum(miss)` of the `r n` values of `y` went missing, and because the
larger ones dropped out preferentially the available cases give
`r round(mean(dat[!miss, "y"]), 3)` against a complete-data truth of
`r round(truth, 3)`, understating it by
`r round(truth - mean(dat[!miss, "y"]), 3)`. Assuming missingness at random
recovers only part of that: the pooled mean is `r round(mar_est, 3)`, still
`r round(truth - mar_est, 3)` low, because the conditional it imputes from
is itself built on the shortened observed values. Supplying the generating
slope to `mnar()` brings the pooled mean to `r round(mnar_est, 3)`, within
`r round(abs(mnar_est - truth), 3)` of the truth. That is the answer to the
first half of the opening question: the bias is real, it is about
`r round(truth - mean(dat[!miss, "y"]), 1)` on this scale, and assuming it
away does not remove it.

The second half is the sensitivity curve. The pooled mean rises
monotonically across the swept slopes, from `r round(sweep$estimate[1], 3)`
at slope 0 to `r round(sweep$estimate[nrow(sweep)], 3)` at slope
`r max(sweep$beta)`, and its interval first reaches the complete-data truth
at an assumed slope of `r beta_first`, against the
`r beta_true` that generated the data. Nothing in the observed data picks
the slope out; sweeping it is what converts an unidentified quantity into a
stated range, and what belongs in a report is the curve itself, with the
reader's own view of a plausible slope applied to it.

Censoring is the case where no sweep is needed. With `r sum(cmiss)` values
hidden below a limit of `r thr`, the available cases give
`r round(mean(cdat[!cmiss, "y"]), 3)`, overstating the truth by
`r round(mean(cdat[!cmiss, "y"]) - truth, 3)`. The common substitution of
half the detection limit gives `r round(mean(ifelse(cmiss, thr / 2,
z_full[, 2L])), 3)`, still `r round(mean(ifelse(cmiss, thr / 2,
z_full[, 2L])) - truth, 3)` high. Drawing from the mixture conditional
truncated to the censored interval gives `r round(cens_est, 3)`, within
`r round(abs(cens_est - truth), 3)` of the truth, because the bound is
known and so the gate is exact rather than assumed.

## Limits

The interval and value-dependent gates act on a single coordinate, and a row
missing that coordinate is assumed to have its other coordinates observed,
which is the detection-limit or single-outcome setting rather than a general
pattern of holes across many columns. A model estimand other than a
coordinate mean is pooled by handing the completions to `mice` through
`as_mids()`, and inherits Rubin's large-sample assumptions.

The sensitivity sweep prices one departure at a time: the selection
probability is a logistic function of the missing coordinate alone, with the
intercept calibrated to the observed missing fraction, so a mechanism that
depends jointly on the missing value and an unobserved third variable is
outside what the sweep brackets. The sweep is also not a test. It cannot
tell an analyst which slope is right, because that is precisely what the
observed data does not carry; it converts one unstated assumption into a
displayed range, and the range is only as wide as the grid the analyst
chose.

The numbers here come from one synthetic realisation at `n` = `r n` with
`m` = `r m_draws` completions, and would move under a different seed. What
would change the answer most is the component count and the assumed slope:
a mixture with too few components cannot represent the two clusters that
make the conditional informative, and a slope far from the truth moves the
pooled mean roughly linearly, which is what the sensitivity table shows.

## What to read next

*Imputing missing data with a mixture* is the companion case this vignette
assumes away, where missingness depends only on what was observed, and it
shows why the mixture conditional beats a single Gaussian on the shape of
the imputations.

*The closed-form operator calculus on a mixture* covers the conditioning
step that every gate multiplies into, and what else it buys.

## References

The selection-model formulation follows Diggle and Kenward (1994) and the
sensitivity-analysis posture follows Little (1993) and the missing-data
literature it anchors. The truncated-Gaussian algebra behind the censored
path is standard, and the closed-form moments are evaluated directly.

* Diggle, P. and Kenward, M. G. (1994). *Informative drop-out in
  longitudinal data analysis.* Journal of the Royal Statistical Society C
  43(1), 49--93. <doi:10.2307/2986113>.
* Little, R. J. A. (1993). *Pattern-mixture models for multivariate
  incomplete data.* Journal of the American Statistical Association
  88(421), 125--134. <doi:10.1080/01621459.1993.10594302>.
* Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys.*
  Wiley.
* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.

## Reproduce

The data-generating seed is `20260622`, set in the chunk that builds the
toy; every call to `gmm_impute()` and `proxy_mnar_sensitivity()` is given
`seed = 1L`, which makes the completions reproducible without disturbing the
ambient random-number state.

```{r session-info}
sessionInfo()
```
