---
title: "The closed-form operator calculus on a mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The closed-form operator calculus on a mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

A modeller has a fitted Gaussian-mixture proxy and a pipeline waiting behind
it: push the latent state through a sensor, fold in a measurement, aggregate
to a coarser grid, condition on what was observed, run the whole thing
forward in time. The question is not whether any of that is possible, but
where the exactness stops: *I have a fitted mixture, so which of the things
I want to do with it next stay exact, and which quietly stop being exact?*
This vignette draws that line, and checks each claimed exactness against an
independent hand computation rather than asserting it.

## What

Four operator families act on a mixture and return a mixture, so they
compose without ever leaving closed form. [`gmm_affine()`] pushes a mixture
through a linear map with additive Gaussian noise, $y = Ax + b + \epsilon$;
[`gmm_aggregate()`] is the same operator specialised to a coarsening matrix.
[`gmm_observe()`] performs the Bayesian update on a noisy linear
observation, the finite-mixture analogue of a Kalman update: it reweights
the components by their marginal evidence and Kalman-updates each one.
[`gmm_marginalise()`] integrates out a coordinate subset, and
[`gmm_conditionalise()`] and [`gmm_missing()`] condition on coordinates
observed exactly, through the Schur-complement algebra.

Two further verbs turn those operators into a recursion.
[`gmm_reduce()`] collapses a mixture to a budget of at most `k_max`
components by a greedy, moment-preserving pairwise merge, so a filter whose
component count would otherwise grow stays bounded. [`gmm_filter()`]
packages the predict, update and reduce loop as a single verb: with Gaussian
noise it is the Kalman filter, and with a Gaussian-sum process or
measurement noise it is the Gaussian-sum filter of Alspach and Sorenson
(1972).

This algebraic dividend is the reason for fitting a *parametric* proxy in
the first place. A non-parametric kernel density estimate or a Monte Carlo
sample bank supports none of it in closed form.

## Do

```{r seed}
set.seed(20260514)
```

### A mixture to operate on

A two-component mixture over a pair of latent variables.

```{r prior}
g_prior <- gmm(
  weights = c(0.6, 0.4),
  means = list(c(-1, 0), c(1.5, 0.5)),
  covariances = list(diag(c(0.6, 0.8)), diag(c(0.7, 0.5)))
)
g_prior
```

### Pushforward through a linear sensor

Suppose the latent state is observed through
$y = A x + \epsilon$ with $\epsilon \sim \mathcal{N}(0, R)$ and a sensor
matrix that reports both coordinates and their sum.

```{r sensor}
A_sensor <- matrix(
  c(1, 0,
    0, 1,
    1, 1),
  nrow = 3L, byrow = TRUE
)
b_sensor <- c(0, 0, 0)
R_sensor <- 0.05 * diag(3)

g_pushed <- gmm_affine(
  g_prior, A_sensor, b_sensor, noise_cov = R_sensor
)
```

The weights are unchanged, the means are $A \mu_k + b$ and the covariances
are $A \Sigma_k A^\top + R$, which the hand computation below reproduces.

```{r sensor-check}
mu_hand <- lapply(g_prior@means, function(mu) {
  as.numeric(A_sensor %*% mu + b_sensor)
})
cov_hand <- lapply(g_prior@covariances, function(s) {
  A_sensor %*% s %*% t(A_sensor) + R_sensor
})
affine_gap <- max(
  abs(unlist(g_pushed@means) - unlist(mu_hand)),
  abs(unlist(g_pushed@covariances) - unlist(cov_hand)),
  abs(g_pushed@weights - g_prior@weights)
)
```

### Bayesian update on a noisy observation

Now observe `y = 0.8` on the first coordinate alone, with noise variance
0.25.

```{r observe}
A_obs <- matrix(c(1, 0), nrow = 1L)
g_post <- gmm_observe(
  g_prior, A = A_obs, y = 0.8, noise_cov = matrix(0.25, 1L, 1L)
)
g_post
```

Component weights have been multiplied by the marginal evidence
$\pi_k \, \mathcal{N}(y;\, A \mu_k,\, S_k)$ and renormalised; component
means and covariances have been Kalman-updated component-wise.

```{r fig-prior-posterior, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4, fig.cap = "The prior mixture and its posterior after one noisy observation of the first coordinate. The posterior is narrower and shifted toward the observed value, and the left-hand component has gained weight because the observation sits closer to it, which is the reweighting the closed-form update performs component by component.", fig.alt = "Two side-by-side density rasters on the same axes; the right-hand posterior panel is visibly more concentrated than the left-hand prior panel and its mass has shifted toward the observed value."}
grid <- expand.grid(
  x = seq(-4, 5, length.out = 80L),
  y = seq(-3, 3, length.out = 60L)
)
gm <- as.matrix(grid)
long <- rbind(
  data.frame(x = grid$x, y = grid$y, d = dgmm(gm, g_prior), part = "Prior"),
  data.frame(
    x = grid$x, y = grid$y, d = dgmm(gm, g_post), part = "Posterior"
  )
)
long$part <- factor(long$part, levels = c("Prior", "Posterior"))
ggplot2::ggplot(long, ggplot2::aes(x, y)) +
  ggplot2::geom_raster(ggplot2::aes(fill = d), interpolate = TRUE) +
  ggplot2::geom_contour(
    ggplot2::aes(z = d), colour = "white", linewidth = 0.2,
    alpha = 0.6, bins = 8L
  ) +
  ggplot2::facet_wrap(~ part) +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::scale_fill_viridis_c(name = "density") +
  ggplot2::labs(
    x = expression(x[1]), y = expression(x[2]),
    title = "Prior and posterior after observing the first coordinate"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold"),
    panel.grid = ggplot2::element_blank()
  )
```

```{r fig-prior-posterior-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the prior-and-posterior",
    "figure is skipped.\n")
```

### The Kalman parity check

For a single-component prior, `gmm_observe()` should collapse exactly to a
Kalman update. The hand recursion below is written out in full so that the
comparison is against arithmetic this package did not produce.

```{r kalman-parity}
g_single <- gmm(
  weights = 1, means = list(c(0, 0)),
  covariances = list(diag(c(1, 2)))
)
g_one_obs <- gmm_observe(
  g_single, A = matrix(c(1, 0), nrow = 1L), y = 0.5,
  noise_cov = matrix(0.5, 1L, 1L)
)

s_prior <- diag(c(1, 2))
h_obs <- matrix(c(1, 0), nrow = 1L)
r_obs <- matrix(0.5, 1L, 1L)
s_innov <- h_obs %*% s_prior %*% t(h_obs) + r_obs
gain <- s_prior %*% t(h_obs) %*% solve(s_innov)
mu_kalman <- as.numeric(gain * 0.5)
cov_kalman <- s_prior - gain %*% h_obs %*% s_prior

kalman_gap <- max(
  abs(mu_kalman - g_one_obs@means[[1L]]),
  abs(cov_kalman - g_one_obs@covariances[[1L]])
)
ridge_default <- 1e-6
kalman_ridges <- kalman_gap / ridge_default
```

### Aggregation through a coarsening matrix

Aggregating a three-coordinate latent into two summaries -- the sum of the
first two coordinates, and the third on its own -- is `gmm_affine()` with a
row-sum matrix, exposed as `gmm_aggregate()`.

```{r aggregate}
g_fine <- gmm(
  weights = c(0.3, 0.4, 0.3),
  means = list(c(0, 0, 0), c(2, 1, -1), c(-1, -1, 2)),
  covariances = list(diag(3), diag(3), diag(3))
)
A_agg <- matrix(
  c(1, 1, 0,
    0, 0, 1),
  nrow = 2L, byrow = TRUE
)
g_coarse <- gmm_aggregate(g_fine, A_agg)
weights_kept <- max(abs(g_coarse@weights - g_fine@weights))
```

```{r aggregate-kable, echo = FALSE}
knitr::kable(
  data.frame(
    component = seq_len(gmm_n_components(g_coarse)),
    weight = round(g_coarse@weights, 3L),
    mean_1 = round(vapply(g_coarse@means, function(m) m[1L], numeric(1L)), 3L),
    mean_2 = round(vapply(g_coarse@means, function(m) m[2L], numeric(1L)), 3L)
  ),
  col.names = c("Component", "Weight", "Mean, sum of x1 and x2", "Mean, x3"),
  caption = paste(
    "The aggregated mixture keeps the component count and the weights of",
    "the fine mixture, with means and covariances mapped through the",
    "coarsening matrix."
  )
)
```

### Conditioning on coordinates observed exactly

If a subset of coordinates is observed with no noise at all, `gmm_missing()`
routes through the Schur-complement path. It is the same operation as
`gmm_conditionalise()` with an index-based signature, and the two agree.

```{r conditioning}
g_cond_index <- gmm_missing(g_prior, observed = 2L, values = 0.5)
g_cond_given <- gmm_conditionalise(g_prior, given = c(NA, 0.5))
cond_gap <- max(
  abs(unlist(g_cond_index@means) - unlist(g_cond_given@means)),
  abs(unlist(g_cond_index@covariances) -
        unlist(g_cond_given@covariances)),
  abs(g_cond_index@weights - g_cond_given@weights)
)
```

### Sequential observations compose

Two observations on disjoint coordinates, applied one after the other,
should equal one observation on the stacked coordinates.

```{r compose}
g_a <- gmm_observe(
  g_prior, A = matrix(c(1, 0), nrow = 1L), y = 0.5,
  noise_cov = matrix(0.25, 1L, 1L)
)
g_ab <- gmm_observe(
  g_a, A = matrix(c(0, 1), nrow = 1L), y = 0.2,
  noise_cov = matrix(0.25, 1L, 1L)
)
g_stack <- gmm_observe(
  g_prior, A = diag(2), y = c(0.5, 0.2), noise_cov = 0.25 * diag(2)
)
compose_gap <- max(
  abs(g_ab@weights - g_stack@weights),
  abs(unlist(g_ab@means) - unlist(g_stack@means)),
  abs(unlist(g_ab@covariances) - unlist(g_stack@covariances))
)
```

### Filtering over time, and the Kalman filter as a special case

The update above is a single step. A *filter* alternates two steps over
time: a **predict** step that pushes the state forward through the dynamics
(`gmm_affine()`, the Chapman-Kolmogorov propagation of a linear stochastic
dynamical system), and an **update** step that folds in each noisy
measurement (`gmm_observe()`). Run on a single-component prior, this
recursion *is* the classical discrete-time Kalman filter; run on a
multi-component prior, it is the **Gaussian-sum filter**, a bank of Kalman
filters carried in parallel.

Take a one-dimensional constant-velocity track, with state
(position, velocity), linear dynamics $A$, process noise $Q$, and a noisy
position sensor $C = [1, 0]$ with noise $R$.

```{r track}
dt <- 1
A_dyn <- matrix(c(1, dt, 0, 1), 2L, 2L, byrow = TRUE)
C_obs <- matrix(c(1, 0), 1L, 2L)
Q_proc <- 0.01 * diag(2)
R_meas <- matrix(0.5, 1L, 1L)

n_steps <- 30L
truth <- matrix(0, n_steps, 2L)
truth[1L, ] <- c(0, 1)
for (k1 in 2:n_steps) {
  truth[k1, ] <- as.numeric(A_dyn %*% truth[k1 - 1L, ]) +
    mvnfast::rmvn(1L, c(0, 0), Q_proc)
} # ends k1, over the simulated state track
y_track <- truth[, 1L] + rnorm(n_steps, 0, sqrt(R_meas[1L, 1L]))
```

The filter is the operator calculus in a loop.

```{r filter-loop}
g_state <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(diag(2))
)
filtered <- numeric(n_steps)
for (k1 in seq_len(n_steps)) {
  if (k1 > 1L) {
    g_state <- gmm_affine(
      g_state, A = A_dyn, b = c(0, 0), noise_cov = Q_proc
    )                                                       # predict
  }
  g_state <- gmm_observe(
    g_state, A = C_obs, y = y_track[k1], noise_cov = R_meas
  )                                                         # update
  filtered[k1] <- g_state@means[[1L]][1L]
} # ends k1, over the filtering recursion
```

Against a hand-coded textbook Kalman filter:

```{r filter-parity}
mu_kf <- c(0, 0)
p_kf <- diag(2)
kf_track <- numeric(n_steps)
for (k1 in seq_len(n_steps)) {
  if (k1 > 1L) {
    mu_kf <- as.numeric(A_dyn %*% mu_kf)
    p_kf <- A_dyn %*% p_kf %*% t(A_dyn) + Q_proc            # predict
  }
  gain_kf <- p_kf %*% t(C_obs) %*%
    solve(C_obs %*% p_kf %*% t(C_obs) + R_meas)             # gain
  mu_kf <- mu_kf + as.numeric(gain_kf %*% (y_track[k1] - C_obs %*% mu_kf))
  p_kf <- (diag(2) - gain_kf %*% C_obs) %*% p_kf
  kf_track[k1] <- mu_kf[1L]
} # ends k1, over the hand-coded Kalman recursion
loop_gap <- max(abs(filtered - kf_track))
```

```{r fig-kalman, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.6, fig.cap = "A constant-velocity track filtered by the single-component operator-calculus recursion, which is the Kalman filter: the noisy position readings, the latent truth, and the filtered estimate tracking it.", fig.alt = "A time series with scattered grey noisy position readings, a smooth latent-truth line, and a filtered-estimate line closely following the truth."}
track_df <- data.frame(
  t = seq_len(n_steps), truth = truth[, 1L], y = y_track,
  filtered = filtered
)
ggplot2::ggplot(track_df, ggplot2::aes(t)) +
  ggplot2::geom_point(
    ggplot2::aes(y = y, colour = "noisy reading"), size = 1.3, alpha = 0.7
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = truth, colour = "latent truth"), linewidth = 0.8
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered, colour = "filtered, one component"),
    linewidth = 0.9
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c(
      "noisy reading" = "grey60",
      "latent truth" = "#0072B2",
      "filtered, one component" = "#D55E00"
    )
  ) +
  ggplot2::labs(
    x = "time step", y = "position",
    title = "Predict and update over time is a Kalman filter"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-kalman-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the Kalman-track figure",
    "is skipped.\n")
```

At more than one component the same two calls run a Kalman filter inside
*every* component and reweight the components by their evidence, which is
the Gaussian-sum filter. It represents beliefs a single Kalman filter cannot
-- multimodal posteriors, several competing hypotheses about where the
target is -- at the cost that each `gmm_observe()` under a Gaussian-mixture
noise or dynamics model multiplies the component count. Collapsing the
mixture back to a fixed budget keeps the filter bounded.

### Bounding the filter through mixture reduction

`gmm_reduce()` collapses a mixture to at most `k_max` components by a
greedy, moment-preserving pairwise merge: at each step the cheapest pair is
replaced by the single Gaussian that preserves their combined weight, mean
and covariance. Two merge costs decide which pair goes first, the
Kullback-Leibler bound of Runnalls (2007) under `cost = "kl"` and a
closed-form Cauchy-Schwarz cost under `cost = "cs"`, built from the same
Gaussian-product identity as `gmm_divergence()`.

Take a six-component mixture that really has three clusters, each split into
two near-identical components, and reduce it to three.

```{r reduce}
g_six <- gmm(
  weights = rep(1 / 6, 6L),
  means = list(
    c(-5, 0), c(-5, 0.15), c(5, 0), c(5.1, -0.1), c(0, 6), c(0.1, 6.1)
  ),
  covariances = rep(list(0.5 * diag(2)), 6L)
)
g_three <- gmm_reduce(g_six, k_max = 3L)

mix_mean <- function(g) Reduce(`+`, Map(`*`, g@weights, g@means))
reduce_shift <- max(abs(mix_mean(g_six) - mix_mean(g_three)))
reduce_divergence <- gmm_divergence(g_six, g_three, type = "cs")
```

```{r reduce-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "components before", "components after",
      "shift in the mixture mean", "Cauchy-Schwarz divergence"
    ),
    value = c(
      formatC(gmm_n_components(g_six), format = "d"),
      formatC(gmm_n_components(g_three), format = "d"),
      formatC(reduce_shift, format = "e", digits = 1L),
      formatC(reduce_divergence, format = "e", digits = 1L)
    )
  ),
  col.names = c("Quantity", "Value"),
  caption = paste(
    "Reducing six components to three by moment-preserving merges,",
    "and what the reduction costs."
  )
)
```

Reducing all the way to a single component returns the moment-matched
Gaussian, so `gmm_reduce()` interpolates between the full mixture and its
single-Gaussian summary. For a smooth, over-parameterised mixture a globally
fitted proxy can beat any sequence of pairwise merges, and `method =
"anneal"` refits a budget-sized proxy by annealed expectation-maximisation
and keeps it when it improves on the merge.

### The bounded filter in one call

The predict, update and reduce loop is packaged as a single verb.
`gmm_filter()` takes the prior belief, the linear-Gaussian `dynamics` and
`measurement`, the observation series, and an optional component cap
`k_max`. With Gaussian noise and `k_max = NULL` it is the Kalman filter;
with a Gaussian-sum process or measurement noise -- a `gmm` in place of a
covariance matrix -- it is the Gaussian-sum filter, bounded by
`gmm_reduce()` after each step.

```{r verb-kalman}
prior_state <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(diag(2))
)
out_verb <- gmm_filter(
  prior_state,
  dynamics = list(A = A_dyn, Q = Q_proc),
  measurement = list(C = C_obs, R = R_meas),
  y = y_track, ridge_eps = 0
)

mu_v <- c(0, 0)
p_v <- diag(2)
kf_verb <- numeric(n_steps)
for (k1 in seq_len(n_steps)) {
  mu_v <- as.numeric(A_dyn %*% mu_v)
  p_v <- A_dyn %*% p_v %*% t(A_dyn) + Q_proc                # predict
  gain_v <- p_v %*% t(C_obs) %*%
    solve(C_obs %*% p_v %*% t(C_obs) + R_meas)              # gain
  mu_v <- mu_v + as.numeric(gain_v %*% (y_track[k1] - C_obs %*% mu_v))
  p_v <- (diag(2) - gain_v %*% C_obs) %*% p_v               # update
  kf_verb[k1] <- mu_v[1L]
} # ends k1, over the predict-then-update reference recursion
verb_gap <- max(abs(out_verb$mean[, 1L] - kf_verb))
```

A heavy-tailed disturbance is awkward for a single Gaussian process noise.
Modelled as a two-component Gaussian sum -- a calm regime and an occasional
larger jump -- the recursion becomes a genuine Gaussian-sum filter. Each
step multiplies the component count by two, so the cap returns it to budget.

```{r verb-gsf}
q_heavy <- gmm(
  weights = c(0.9, 0.1),
  means = list(c(0, 0), c(0, 0)),
  covariances = list(0.01 * diag(2), 0.5 * diag(2))
)
out_gsf <- gmm_filter(
  prior_state,
  dynamics = list(A = A_dyn, Q = q_heavy),
  measurement = list(C = C_obs, R = R_meas),
  y = y_track, k_max = 6L
)
gsf_max_k <- max(out_gsf$summary$n_components)
gsf_rmse <- sqrt(mean((out_gsf$mean[, 1L] - truth[, 1L])^2))
kalman_rmse <- sqrt(mean((out_verb$mean[, 1L] - truth[, 1L])^2))
```

```{r fig-gsf, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.6, fig.cap = "The same track filtered by the Gaussian-sum filter under a two-component heavy-tailed process noise, capped at six components per step. The filtered mean tracks the latent truth while the component count stays at the cap.", fig.alt = "A time series with grey noisy position readings, a latent-truth line, and a Gaussian-sum-filter estimate line following the truth closely."}
gsf_df <- data.frame(
  t = seq_len(n_steps), truth = truth[, 1L], y = y_track,
  filtered = out_gsf$mean[, 1L]
)
ggplot2::ggplot(gsf_df, ggplot2::aes(t)) +
  ggplot2::geom_point(
    ggplot2::aes(y = y, colour = "noisy reading"), size = 1.3, alpha = 0.7
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = truth, colour = "latent truth"), linewidth = 0.8
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered, colour = "Gaussian-sum filter"),
    linewidth = 0.9
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c(
      "noisy reading" = "grey60",
      "latent truth" = "#0072B2",
      "Gaussian-sum filter" = "#D55E00"
    )
  ) +
  ggplot2::labs(
    x = "time step", y = "position",
    title = "A bounded Gaussian-sum filter in one call"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-gsf-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the Gaussian-sum-filter",
    "figure is skipped.\n")
```

```{r parity-kable, echo = FALSE}
knitr::kable(
  data.frame(
    check = c(
      "pushforward against the hand formula",
      "single-component update against a hand Kalman step",
      "index conditioning against value conditioning",
      "two observations against one stacked observation",
      "operator loop against a hand Kalman filter",
      "gmm_filter() against a hand Kalman filter",
      "weights preserved by aggregation"
    ),
    gap = formatC(
      c(
        affine_gap, kalman_gap, cond_gap, compose_gap, loop_gap,
        verb_gap, weights_kept
      ),
      format = "e", digits = 1L
    )
  ),
  col.names = c("Exactness check", "Largest absolute difference"),
  caption = paste(
    "Every exactness claim in this vignette, against an independently",
    "written hand computation. The default numerical-hygiene ridge is",
    "1e-06."
  )
)
```

## Read

Every operator claimed to be exact is exact to the precision the table
reports. The pushforward reproduces $A \mu_k + b$ and
$A \Sigma_k A^\top + R$ to
`r formatC(affine_gap, format = "e", digits = 1L)`, and it leaves the
weights untouched, as aggregation does to
`r formatC(weights_kept, format = "e", digits = 1L)`: a linear map moves
where the components sit, never how much mass each carries.

The single-component update matches a hand-written Kalman step to
`r formatC(kalman_gap, format = "e", digits = 1L)`, which is
`r round(kalman_ridges, 2L)` times the default numerical-hygiene ridge of
`r formatC(ridge_default, format = "e", digits = 0L)` that the operators add
to every covariance to keep it positive-definite. The update is therefore
exact up to that deliberate regularisation, and to nothing else. The two
conditioning signatures agree to
`r formatC(cond_gap, format = "e", digits = 1L)`, so `gmm_missing()` is a
different door onto `gmm_conditionalise()` rather than a different
computation. Two sequential observations on disjoint coordinates agree with
one stacked observation to
`r formatC(compose_gap, format = "e", digits = 1L)`, which is the
associativity that lets a pipeline fold measurements in whatever order they
arrive.

Over time the same picture holds. The explicit predict-and-update loop
agrees with a hand-coded Kalman filter across all `r n_steps` steps to
`r formatC(loop_gap, format = "e", digits = 1L)`, and the packaged
`gmm_filter()` verb agrees with the same reference to
`r formatC(verb_gap, format = "e", digits = 1L)`. The first figure shows the
update contracting the belief toward the observation, and the second shows
the recursion tracking a latent state through noise.

Reduction is the one operator on this page that is designed to lose
something, and the table says how much. Collapsing six components to
`r gmm_n_components(g_three)` moves the mixture mean by
`r formatC(reduce_shift, format = "e", digits = 1L)`, because every merge
preserves the combined weight, mean and covariance by construction; the
Cauchy-Schwarz divergence between the six-component mixture and its
three-component reduction is
`r formatC(reduce_divergence, format = "e", digits = 1L)`, small here only
because the components merged were near-duplicates. On a mixture whose
components genuinely differ, that divergence is where the cost of the budget
shows up.

The bounded filter behaves as advertised. Under a two-component heavy-tailed
process noise the component count would double at every step; capped, it
holds at `r gsf_max_k` across the whole series, and the filtered mean tracks
the latent truth with a root-mean-square error of
`r round(gsf_rmse, 3L)` against `r round(kalman_rmse, 3L)` for the
single-Gaussian Kalman filter on the same readings.

So the line the opening question asked for runs as follows. Pushforward,
aggregation, noisy linear update, marginalisation and exact conditioning all
stay exact, and compose in any order, because each returns a mixture and the
next operator does not know or care how the mixture arose. Reduction is
exact in its moments and approximate in its shape. Everything outside the
affine-Gaussian channel leaves closed form altogether, which is the subject
of the next section.

## Limits

The closed-form algebra rests on two assumptions, and both are load-bearing.
The map must be linear and the noise must be Gaussian. A non-linear sensor
-- a sigmoid observation, a max-pooling aggregator, anything that bends the
coordinate space -- has no closed-form pushforward through a Gaussian
mixture, and must not be silently linearised; a Monte Carlo pushforward is
the correct fallback, and the operators here will not stand in for one. The
same restriction applies to non-Gaussian observation noise. Hierarchical and
random-effect models sit outside the calculus for a third reason: the
operators apply to a *fixed* affine-Gaussian channel, and a prior over the
channel parameters $(A, R)$ themselves requires an integration that is not
closed form in general.

Component growth is the practical limit on the filter. Under Gaussian-sum
process or measurement noise each update multiplies the component count by
the number of noise components, so an uncapped Gaussian-sum filter over a
long series is not a viable object. `gmm_reduce()` is what makes it viable,
and the budget is a modelling choice with consequences: a cap tight enough
to be cheap will merge components that represent genuinely distinct
hypotheses, and the multimodality the Gaussian-sum filter exists to carry is
exactly what a tight cap discards. The divergence reported above is small
because the example merges near-duplicates by design, and it should be
recomputed on any real filtering problem rather than assumed small.

A Gaussian-process latent is the natural comparison and wins on flexibility.
It gives a fully non-parametric, uncertainty-rich representation of the
underlying field, and integration through a linear operator with additive
Gaussian noise is closed form for a Gaussian process too, so on a single
pass there is little to choose between them. What the Gaussian process
lacks is cheap *composition*: repeatedly conditioning, aggregating and
re-mixing its posteriors across configurations carries a growing
kernel-algebra burden, where the mixture calculus stays a finite parameter
list of fixed size. The finite mixture is the less flexible representation
and the more composable one, so the choice turns on how much closed-form
composition the downstream pipeline actually needs.

Finally, everything on this page operates on a mixture that is taken as
given. None of these checks says the mixture is a good proxy for whatever it
was fitted to, and an exact operator applied to a poor proxy propagates the
proxy's error exactly.

## What to read next

*Fitting a proxy to a density you cannot sample* produces the mixtures this
vignette operates on, and reports the fit-quality certificate that says
whether the proxy is worth operating on. *Testing the last observation for
instability* uses the filtering recursion built here as the engine of an
end-of-sample structural-break test. *Reading the entropy of a fitted
mixture* covers the divergence used above to price the reduction, and the
rest of the closed-form diagnostic surface. *One mixture, many methods*
shows the same conditioning operator standing in for regression, kernel
smoothing and principal components.

## References

* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.
* Kalman, R. E. (1960). *A new approach to linear filtering and
  prediction problems.* Transactions of the ASME -- J. Basic Engineering.
* Alspach, D. L. and Sorenson, H. W. (1972). *Nonlinear Bayesian
  estimation using Gaussian sum approximations.* IEEE Transactions on
  Automatic Control 17(4), 439--448.
* Runnalls, A. R. (2007). *Kullback--Leibler approach to Gaussian mixture
  reduction.* IEEE Transactions on Aerospace and Electronic Systems
  43(3), 989--999.
* Murphy, K. P. (2012). *Machine Learning: A Probabilistic Perspective.*
  MIT Press. Ch. 4 (Gaussian models), Ch. 18 (state-space).

## Reproduce

The vignette sets `set.seed(20260514)` once, before the constant-velocity
track is simulated; every operator result above it is deterministic algebra
that draws no random numbers at all.

```{r session-info}
sessionInfo()
```
