---
title: "Affine-Gaussian operator calculus on a Gaussian mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Affine-Gaussian operator calculus on a Gaussian mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
set.seed(20260514)
```

```{r setup}
library(proxymix)
library(ggplot2)
```

# Why an operator calculus

A fitted Gaussian-mixture proxy supports four kinds of downstream
operations in closed form:

* **Pushforward** through a linear sensor or aggregation operator
  (`gmm_affine`, `gmm_aggregate`).
* **Bayesian update** on a noisy linear observation
  (`gmm_observe` — the finite-mixture analogue of a Kalman update).
* **Marginalisation** over a coordinate subset
  (`gmm_marginalise`, shipped since v0.1).
* **Conditioning** on fully observed coordinates
  (`gmm_conditionalise`, `gmm_missing`).

Each operator returns a `gmm`, so the operators compose freely without
ever leaving closed form. This is the algebraic dividend of having
fitted a *parametric* proxy in the first place — a non-parametric KDE
or a Monte Carlo sample bank cannot do this.

# A worked example

Start with a 2D mixture proxy of a pair of latent variables.

```{r}
g_prior <- gmm(
  weights     = c(0.6, 0.4),
  means       = list(c(-1, 0), c(1.5, 0.5)),
  covariances = list(diag(c(0.6, 0.8)), diag(c(0.7, 0.5)))
)
g_prior
```

## Pushforward through a linear sensor

Suppose we observe \(y = A x + \epsilon\) with \(\epsilon \sim
\mathcal{N}(0, R)\) and sensor matrix

```{r}
A <- matrix(c(1, 0,
              0, 1,
              1, 1), nrow = 3L, byrow = TRUE)
b <- c(0, 0, 0)
R <- 0.05 * diag(3)
```

`gmm_affine()` returns the pushed-forward mixture in `R^3`.

```{r}
g_pushed <- gmm_affine(g_prior, A, b, noise_cov = R)
g_pushed
```

The weights are unchanged; the means are \(A \mu_k + b\) and the
covariances are \(A \Sigma_k A^\top + R\). The output is a closed-form
mixture in the sensor space.

## Bayesian update on a noisy linear observation

Suppose now that we observe `y = 0.8` on the *first* coordinate of `x`
alone, with noise variance `R = 0.25`.

```{r}
A_obs <- matrix(c(1, 0), nrow = 1L)
g_post <- gmm_observe(g_prior, A = A_obs, y = 0.8,
                      noise_cov = matrix(0.25, 1L, 1L))
g_post
```

Component weights have been multiplied by the marginal evidence
\(\pi_k \, \mathcal{N}(y;\, A \mu_k,\, S_k)\) and renormalised. Component
means and covariances have been Kalman-updated component-wise.

Visualise the prior and the posterior on a planar grid:

```{r, fig.height = 4}
grid <- expand.grid(x = seq(-4, 5, length.out = 80L),
                    y = seq(-3, 3, length.out = 60L))
gm <- as.matrix(grid)
prior_df <- data.frame(x = grid$x, y = grid$y,
                       d = dgmm(gm, g_prior), part = "prior")
post_df  <- data.frame(x = grid$x, y = grid$y,
                       d = dgmm(gm, g_post),  part = "posterior")
long <- rbind(prior_df, post_df)
long$part <- factor(long$part, levels = c("prior", "posterior"),
                    labels = c("Prior", "Posterior"))
ggplot(long, aes(x, y)) +
  geom_raster(aes(fill = d), interpolate = TRUE) +
  geom_contour(aes(z = d), colour = "white", linewidth = 0.2, alpha = 0.6,
               bins = 8L) +
  facet_wrap(~ part) +
  coord_equal(expand = FALSE) +
  scale_fill_viridis_c(name = "Density") +
  labs(title = expression(paste("Prior vs posterior after observing ",
                                y[1] == 0.8, "  (", sigma == 0.5, ")")),
       x = expression(x[1]), y = expression(x[2])) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = rel(1)),
        strip.text = element_text(face = "bold"),
        panel.grid = element_blank(),
        legend.position = "right")
```

## Kalman parity check

For a single-component prior, `gmm_observe()` collapses exactly to a
Kalman update. Verify:

```{r}
g_single <- gmm(
  weights     = 1,
  means       = list(c(0, 0)),
  covariances = list(diag(c(1, 2)))
)
g_one_obs <- gmm_observe(
  g_single,
  A = matrix(c(1, 0), nrow = 1L),
  y = 0.5,
  noise_cov = matrix(0.5, 1L, 1L)
)

## Hand Kalman:
S      <- diag(c(1, 2))
H      <- matrix(c(1, 0), nrow = 1L)
R_obs  <- matrix(0.5, 1L, 1L)
Sx     <- H %*% S %*% t(H) + R_obs
gain   <- S %*% t(H) %*% solve(Sx)
mu_hat <- as.numeric(gain * 0.5)
S_hat  <- S - gain %*% H %*% S

list(
  mu_diff_max  = max(abs(mu_hat - g_one_obs@means[[1L]])),
  cov_diff_max = max(abs(S_hat  - g_one_obs@covariances[[1L]]))
)
```

The differences are at the level of the numerical-hygiene ridge (1e-6),
so the operator is exact up to numerical tolerance.

## Aggregation through a coarsening matrix

Aggregating a fine-grained latent (3 components) into a coarser one
(sum of the first two coordinates) is a special case of `gmm_affine`
with a row-sum `A`.

```{r}
g_fine <- gmm(
  weights     = c(0.3, 0.4, 0.3),
  means       = list(c(0, 0, 0), c(2, 1, -1), c(-1, -1, 2)),
  covariances = list(diag(3), diag(3), diag(3))
)
A_agg <- matrix(c(1, 1, 0,
                  0, 0, 1), nrow = 2L, byrow = TRUE)
gmm_aggregate(g_fine, A_agg)
```

The aggregated mixture has the same number of components and weights as
the fine one, but with `mu` and `Sigma` mapped through `A`.

## Missing-data conditioning

If a subset of coordinates is *exactly* observed (no noise),
`gmm_missing()` routes through the Schur-complement path:

```{r}
gmm_missing(g_prior, observed = 2L, values = 0.5)
```

This is equivalent to `gmm_conditionalise(g_prior, given = c(NA, 0.5))`
but with a clearer index-based signature for missing-data pipelines.

# Filtering over time — the Kalman filter as a special case

The Bayesian update above is a single step. A *filter* alternates two steps over
time: a **predict** step that pushes the state forward through the dynamics
(`gmm_affine`, the Chapman--Kolmogorov propagation of a linear stochastic
dynamical system), and an **update** step that folds in each noisy measurement
(`gmm_observe`). Run on a single-component prior, this recursion *is* the
classical discrete-time Kalman filter; run on a multi-component prior, it is the
**Gaussian-sum filter** -- a bank of Kalman filters carried in parallel.

Take a one-dimensional constant-velocity track, with state
\((\text{position}, \text{velocity})\), linear dynamics \(A\), process noise
\(Q\), and a noisy position sensor \(C = [1, 0]\) with noise \(R\).

```{r kf-setup}
dt <- 1
A  <- matrix(c(1, dt, 0, 1), 2L, 2L, byrow = TRUE)   # constant velocity
C  <- matrix(c(1, 0), 1L, 2L)                         # observe position only
Q  <- 0.01 * diag(2)                                  # process noise
R  <- matrix(0.5, 1L, 1L)                             # sensor noise
set.seed(42)
n <- 30L
truth <- matrix(0, n, 2L)
truth[1L, ] <- c(0, 1)
for (k in 2:n) {
  truth[k, ] <- as.numeric(A %*% truth[k - 1L, ]) + mvnfast::rmvn(1L, c(0, 0), Q)
}
y <- truth[, 1L] + rnorm(n, 0, sqrt(R[1L, 1L]))       # noisy position readings
```

The filter is the operator calculus in a loop -- `gmm_affine()` to predict,
`gmm_observe()` to update:

```{r kf-loop}
g <- gmm(weights = 1, means = list(c(0, 0)), covariances = list(diag(2)))
filtered <- numeric(n)
for (k in seq_len(n)) {
  if (k > 1L) g <- gmm_affine(g, A = A, b = c(0, 0), noise_cov = Q)  # predict
  g <- gmm_observe(g, A = C, y = y[k], noise_cov = R)                # update
  filtered[k] <- g@means[[1L]][1L]
}
```

Against a hand-coded textbook Kalman filter the two tracks agree to the level of
the numerical-hygiene ridge:

```{r kf-parity}
mu <- c(0, 0); S <- diag(2); kf <- numeric(n)
for (k in seq_len(n)) {
  if (k > 1L) { mu <- as.numeric(A %*% mu); S <- A %*% S %*% t(A) + Q }  # predict
  K  <- S %*% t(C) %*% solve(C %*% S %*% t(C) + R)                       # gain
  mu <- mu + as.numeric(K %*% (y[k] - C %*% mu))
  S  <- (diag(2) - K %*% C) %*% S
  kf[k] <- mu[1L]
}
c(max_abs_diff = max(abs(filtered - kf)))
```

```{r kf-plot, fig.height = 3.4, fig.cap = "A constant-velocity track filtered by the K = 1 operator-calculus recursion (the Kalman filter): noisy readings, the latent truth, and the filtered estimate.", fig.alt = "Time series with scattered noisy position readings, the smooth latent truth line, and the filtered estimate line tracking the truth."}
track <- data.frame(t = seq_len(n), truth = truth[, 1L], y = y,
                    filtered = filtered)
ggplot(track, aes(t)) +
  geom_point(aes(y = y, colour = "noisy reading"), size = 1.3, alpha = 0.7) +
  geom_line(aes(y = truth, colour = "latent truth"), linewidth = 0.8) +
  geom_line(aes(y = filtered, colour = "filtered (K = 1)"), linewidth = 0.9) +
  scale_colour_manual(name = NULL,
                      values = c("noisy reading" = "grey55",
                                 "latent truth" = "#2C7FB8",
                                 "filtered (K = 1)" = "#D95F0E")) +
  labs(title = "Predict + update over time is a Kalman filter",
       x = "time step", y = "position") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

**The mixture generalisation.** At \(K > 1\) the same two calls run a Kalman
filter inside *every* component and reweight the components by their evidence --
the Gaussian-sum filter. It represents beliefs a single Kalman filter cannot
(multimodal posteriors: several competing hypotheses about where the target is),
at the cost that each `gmm_observe` (under a Gaussian-mixture noise or dynamics
model) can multiply the component count. Collapsing the mixture back to a fixed
budget keeps the filter bounded; that collapse is `gmm_reduce()`, next.

# Bounding the filter — mixture reduction

`gmm_reduce()` collapses a mixture to a budget of at most `k_max` components by a
greedy, moment-preserving pairwise merge: at each step the cheapest pair is
replaced by the single Gaussian that preserves their combined weight, mean and
covariance. Two merge costs decide which pair goes first -- the Kullback--Leibler
bound of Runnalls (2007), `cost = "kl"`, and a closed-form Cauchy--Schwarz cost,
`cost = "cs"`, built from the same Gaussian-product identity as
`gmm_divergence()`.

Take a six-component mixture that really has three clusters, each split into two
near-identical components, and reduce it to three:

```{r reduce}
g6 <- gmm(
  weights     = rep(1 / 6, 6),
  means       = list(c(-5, 0), c(-5, 0.15), c(5, 0), c(5.1, -0.1),
                     c(0, 6), c(0.1, 6.1)),
  covariances = rep(list(0.5 * diag(2)), 6)
)
g3 <- gmm_reduce(g6, k_max = 3L)
gmm_n_components(g3)
```

Because every merge is moment-preserving, the reduction leaves the mixture's
global mean and covariance untouched -- and merging near-duplicates barely moves
the mixture, so the divergence from the original is negligible:

```{r reduce-check}
mix_mean <- function(g) Reduce(`+`, Map(`*`, g@weights, g@means))
c(mean_shift = max(abs(mix_mean(g6) - mix_mean(g3))),
  divergence = gmm_divergence(g6, g3, type = "cs"))
```

Reducing all the way to a single component returns the moment-matched Gaussian,
so `gmm_reduce()` interpolates between the full mixture and its single-Gaussian
summary. For a smooth, over-parameterised mixture a globally fitted proxy can
beat any sequence of pairwise merges; `method = "anneal"` refits a budget-sized
proxy by annealed EM and keeps it when it improves on the merge. In a
Gaussian-sum filter, reduction runs after each update, holding the component
count at the budget.

# A bounded filter in one call — `gmm_filter()`

The predict / update / reduce loop is packaged as a single verb. `gmm_filter()`
takes the prior belief, the linear-Gaussian `dynamics` and `measurement`, the
observation series `y`, and an optional component cap `k_max`. With Gaussian
noise and `k_max = NULL` it is the Kalman filter; with a Gaussian-sum process or
measurement noise -- a `gmm` in place of a covariance matrix -- it is the
Gaussian-sum filter, bounded by `gmm_reduce()` after each step.

On the constant-velocity track above, the verb reproduces the explicit Kalman
recursion to machine precision (using the same predict-then-update convention):

```{r gmm-filter-kalman}
prior <- gmm(weights = 1, means = list(c(0, 0)), covariances = list(diag(2)))
out <- gmm_filter(prior,
                  dynamics    = list(A = A, Q = Q),
                  measurement = list(C = C, R = R),
                  y = y, ridge_eps = 0)

mu <- c(0, 0); P <- diag(2); kf2 <- numeric(n)
for (k in seq_len(n)) {
  mu <- as.numeric(A %*% mu); P <- A %*% P %*% t(A) + Q          # predict
  G  <- P %*% t(C) %*% solve(C %*% P %*% t(C) + R)               # gain
  mu <- mu + as.numeric(G %*% (y[k] - C %*% mu))                 # update
  P  <- (diag(2) - G %*% C) %*% P
  kf2[k] <- mu[1L]
}
c(max_abs_diff = max(abs(out$mean[, 1L] - kf2)))
```

A heavy-tailed disturbance is awkward for a single Gaussian process noise.
Modelled as a two-component Gaussian sum -- a calm regime and an occasional
larger jump -- the recursion becomes a genuine Gaussian-sum filter. Each step
multiplies the component count by two, so the cap returns it to budget:

```{r gmm-filter-gsf}
q_heavy <- gmm(weights     = c(0.9, 0.1),
               means       = list(c(0, 0), c(0, 0)),
               covariances = list(0.01 * diag(2), 0.5 * diag(2)))
gsf <- gmm_filter(prior,
                  dynamics    = list(A = A, Q = q_heavy),
                  measurement = list(C = C, R = R),
                  y = y, k_max = 6L)
c(max_components = max(gsf$summary$n_components), steps = nrow(gsf$summary))
```

The component count is held at the cap across the whole series. The `$summary`
data frame records the filtered mean, standard deviation and component count at
each step, and `$filtered` holds the full filtered mixture per step for any
downstream closed-form query.

```{r gmm-filter-plot, fig.height = 3.4, fig.cap = "The same track filtered by the Gaussian-sum filter with a two-component heavy-tailed process noise, capped at six components per step. The filtered mean tracks the latent truth while the component count stays bounded.", fig.alt = "Time series with noisy position readings, the latent truth line, and the Gaussian-sum-filter estimate tracking the truth."}
gsf_track <- data.frame(t = seq_len(n), truth = truth[, 1L], y = y,
                        filtered = gsf$mean[, 1L])
ggplot(gsf_track, aes(t)) +
  geom_point(aes(y = y, colour = "noisy reading"), size = 1.3, alpha = 0.7) +
  geom_line(aes(y = truth, colour = "latent truth"), linewidth = 0.8) +
  geom_line(aes(y = filtered, colour = "Gaussian-sum filter"), linewidth = 0.9) +
  scale_colour_manual(name = NULL,
                      values = c("noisy reading" = "grey55",
                                 "latent truth" = "#2C7FB8",
                                 "Gaussian-sum filter" = "#D95F0E")) +
  labs(title = "A bounded Gaussian-sum filter via gmm_filter()",
       x = "time step", y = "position") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

# Composition rules

The operators commute exactly with `gmm_marginalise()` when applied to
disjoint coordinate subsets, and they associate under sequential
observation:

```{r}
## Two sequential observations on disjoint coordinates
g_a <- gmm_observe(g_prior, A = matrix(c(1, 0), nrow = 1L),
                   y = 0.5, noise_cov = matrix(0.25, 1, 1))
g_ab <- gmm_observe(g_a,    A = matrix(c(0, 1), nrow = 1L),
                    y = 0.2, noise_cov = matrix(0.25, 1, 1))

## Equivalent: one observation on the stacked coordinates
g_stack <- gmm_observe(
  g_prior,
  A         = diag(2),
  y         = c(0.5, 0.2),
  noise_cov = 0.25 * diag(2)
)

list(
  weights_diff = max(abs(g_ab@weights - g_stack@weights)),
  mu1_diff     = max(abs(g_ab@means[[1L]] - g_stack@means[[1L]])),
  cov1_diff    = max(abs(g_ab@covariances[[1L]] - g_stack@covariances[[1L]]))
)
```

The disagreement is at the ridge level — the two routes coincide.

# When to *not* use the operator calculus

* **Non-linear channels.** The closed-form maths uses linearity of `A`
  and Gaussianity of the noise. A non-linear sensor (e.g. a Sigmoid
  observation, a max-pooling aggregator) is *not* closed form and must
  not be silently approximated; use a Monte Carlo pushforward instead.
* **Non-Gaussian observation noise.** Same restriction.
* **Hierarchical / random-effect models.** The operator calculus
  applies to a *fixed* affine-Gaussian channel; a mixture-of-channels
  prior on `(A, R)` requires additional integration that is not closed
  form in general.

# Comparison to a Gaussian-process latent

For a downscaling problem \(g(y) = \mathbb{E}_{x \mid y}[f(x)]\):

* A **Gaussian-process latent** for `f(x)` gives a fully non-parametric,
  uncertainty-rich representation but does not admit closed-form
  aggregation under general `A`. Sampling is closed form; integration
  through a linear operator with additive Gaussian noise *is* closed
  form for a GP, but mixing GP posteriors across aggregation
  configurations is heavier than the operator calculus shown here.
* A **finite Gaussian-mixture latent** (what `proxymix` ships) is less
  flexible but admits the full operator calculus on this page. The
  trade-off is bias against algebraic tractability. The choice depends
  on how much closed-form composition the downstream pipeline needs.

# References

* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.
* Kalman, R. E. (1960). *A new approach to linear filtering and
  prediction problems.* Transactions of the ASME — J. Basic Engineering.
* Alspach, D. L. and Sorenson, H. W. (1972). *Nonlinear Bayesian
  estimation using Gaussian sum approximations.* IEEE Transactions on
  Automatic Control 17(4), 439--448.
* Runnalls, A. R. (2007). *Kullback--Leibler approach to Gaussian mixture
  reduction.* IEEE Transactions on Aerospace and Electronic Systems
  43(3), 989--999.
* Murphy, K. P. (2012). *Machine Learning: A Probabilistic Perspective.*
  MIT Press. Ch. 4 (Gaussian models), Ch. 18 (state-space).
