---
title: "Affine-Gaussian operator calculus on a Gaussian mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Affine-Gaussian operator calculus on a Gaussian mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
set.seed(20260514)
```

```{r setup}
library(proxymix)
library(ggplot2)
```

# Why an operator calculus

A fitted Gaussian-mixture proxy supports four kinds of downstream
operations in closed form:

* **Pushforward** through a linear sensor or aggregation operator
  (`gmm_affine`, `gmm_aggregate`).
* **Bayesian update** on a noisy linear observation
  (`gmm_observe` — the finite-mixture analogue of a Kalman update).
* **Marginalisation** over a coordinate subset
  (`gmm_marginalise`, shipped since v0.1).
* **Conditioning** on fully observed coordinates
  (`gmm_conditionalise`, `gmm_missing`).

Each operator returns a `gmm`, so the operators compose freely without
ever leaving closed form. This is the algebraic dividend of having
fitted a *parametric* proxy in the first place — a non-parametric KDE
or a Monte Carlo sample bank cannot do this.

# A worked example

Start with a 2D mixture proxy of a pair of latent variables.

```{r}
g_prior <- gmm(
  weights     = c(0.6, 0.4),
  means       = list(c(-1, 0), c(1.5, 0.5)),
  covariances = list(diag(c(0.6, 0.8)), diag(c(0.7, 0.5)))
)
g_prior
```

## Pushforward through a linear sensor

Suppose we observe \(y = A x + \epsilon\) with \(\epsilon \sim
\mathcal{N}(0, R)\) and sensor matrix

```{r}
A <- matrix(c(1, 0,
              0, 1,
              1, 1), nrow = 3L, byrow = TRUE)
b <- c(0, 0, 0)
R <- 0.05 * diag(3)
```

`gmm_affine()` returns the pushed-forward mixture in `R^3`.

```{r}
g_pushed <- gmm_affine(g_prior, A, b, noise_cov = R)
g_pushed
```

The weights are unchanged; the means are \(A \mu_k + b\) and the
covariances are \(A \Sigma_k A^\top + R\). The output is a closed-form
mixture in the sensor space.

## Bayesian update on a noisy linear observation

Suppose now that we observe `y = 0.8` on the *first* coordinate of `x`
alone, with noise variance `R = 0.25`.

```{r}
A_obs <- matrix(c(1, 0), nrow = 1L)
g_post <- gmm_observe(g_prior, A = A_obs, y = 0.8,
                      noise_cov = matrix(0.25, 1L, 1L))
g_post
```

Component weights have been multiplied by the marginal evidence
\(\pi_k \, \mathcal{N}(y;\, A \mu_k,\, S_k)\) and renormalised. Component
means and covariances have been Kalman-updated component-wise.

Visualise the prior and the posterior on a planar grid:

```{r, fig.height = 4}
grid <- expand.grid(x = seq(-4, 5, length.out = 80L),
                    y = seq(-3, 3, length.out = 60L))
gm <- as.matrix(grid)
prior_df <- data.frame(x = grid$x, y = grid$y,
                       d = dgmm(gm, g_prior), part = "prior")
post_df  <- data.frame(x = grid$x, y = grid$y,
                       d = dgmm(gm, g_post),  part = "posterior")
long <- rbind(prior_df, post_df)
long$part <- factor(long$part, levels = c("prior", "posterior"),
                    labels = c("Prior", "Posterior"))
ggplot(long, aes(x, y)) +
  geom_raster(aes(fill = d), interpolate = TRUE) +
  geom_contour(aes(z = d), colour = "white", linewidth = 0.2, alpha = 0.6,
               bins = 8L) +
  facet_wrap(~ part) +
  coord_equal(expand = FALSE) +
  scale_fill_viridis_c(name = "Density") +
  labs(title = expression(paste("Prior vs posterior after observing ",
                                y[1] == 0.8, "  (", sigma == 0.5, ")")),
       x = expression(x[1]), y = expression(x[2])) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = rel(1)),
        strip.text = element_text(face = "bold"),
        panel.grid = element_blank(),
        legend.position = "right")
```

## Kalman parity check

For a single-component prior, `gmm_observe()` collapses exactly to a
Kalman update. Verify:

```{r}
g_single <- gmm(
  weights     = 1,
  means       = list(c(0, 0)),
  covariances = list(diag(c(1, 2)))
)
g_one_obs <- gmm_observe(
  g_single,
  A = matrix(c(1, 0), nrow = 1L),
  y = 0.5,
  noise_cov = matrix(0.5, 1L, 1L)
)

## Hand Kalman:
S      <- diag(c(1, 2))
H      <- matrix(c(1, 0), nrow = 1L)
R_obs  <- matrix(0.5, 1L, 1L)
Sx     <- H %*% S %*% t(H) + R_obs
gain   <- S %*% t(H) %*% solve(Sx)
mu_hat <- as.numeric(gain * 0.5)
S_hat  <- S - gain %*% H %*% S

list(
  mu_diff_max  = max(abs(mu_hat - g_one_obs@means[[1L]])),
  cov_diff_max = max(abs(S_hat  - g_one_obs@covariances[[1L]]))
)
```

The differences are at the level of the numerical-hygiene ridge (1e-6),
so the operator is exact up to numerical tolerance.

## Aggregation through a coarsening matrix

Aggregating a fine-grained latent (3 components) into a coarser one
(sum of the first two coordinates) is a special case of `gmm_affine`
with a row-sum `A`.

```{r}
g_fine <- gmm(
  weights     = c(0.3, 0.4, 0.3),
  means       = list(c(0, 0, 0), c(2, 1, -1), c(-1, -1, 2)),
  covariances = list(diag(3), diag(3), diag(3))
)
A_agg <- matrix(c(1, 1, 0,
                  0, 0, 1), nrow = 2L, byrow = TRUE)
gmm_aggregate(g_fine, A_agg)
```

The aggregated mixture has the same number of components and weights as
the fine one, but with `mu` and `Sigma` mapped through `A`.

## Missing-data conditioning

If a subset of coordinates is *exactly* observed (no noise),
`gmm_missing()` routes through the Schur-complement path:

```{r}
gmm_missing(g_prior, observed = 2L, values = 0.5)
```

This is equivalent to `gmm_conditionalise(g_prior, given = c(NA, 0.5))`
but with a clearer index-based signature for missing-data pipelines.

# Composition rules

The operators commute exactly with `gmm_marginalise()` when applied to
disjoint coordinate subsets, and they associate under sequential
observation:

```{r}
## Two sequential observations on disjoint coordinates
g_a <- gmm_observe(g_prior, A = matrix(c(1, 0), nrow = 1L),
                   y = 0.5, noise_cov = matrix(0.25, 1, 1))
g_ab <- gmm_observe(g_a,    A = matrix(c(0, 1), nrow = 1L),
                    y = 0.2, noise_cov = matrix(0.25, 1, 1))

## Equivalent: one observation on the stacked coordinates
g_stack <- gmm_observe(
  g_prior,
  A         = diag(2),
  y         = c(0.5, 0.2),
  noise_cov = 0.25 * diag(2)
)

list(
  weights_diff = max(abs(g_ab@weights - g_stack@weights)),
  mu1_diff     = max(abs(g_ab@means[[1L]] - g_stack@means[[1L]])),
  cov1_diff    = max(abs(g_ab@covariances[[1L]] - g_stack@covariances[[1L]]))
)
```

The disagreement is at the ridge level — the two routes coincide.

# When to *not* use the operator calculus

* **Non-linear channels.** The closed-form maths uses linearity of `A`
  and Gaussianity of the noise. A non-linear sensor (e.g. a Sigmoid
  observation, a max-pooling aggregator) is *not* closed form and must
  not be silently approximated; use a Monte Carlo pushforward instead.
* **Non-Gaussian observation noise.** Same restriction.
* **Hierarchical / random-effect models.** The operator calculus
  applies to a *fixed* affine-Gaussian channel; a mixture-of-channels
  prior on `(A, R)` requires additional integration that is not closed
  form in general.

# Comparison to a Gaussian-process latent

For a downscaling problem \(g(y) = \mathbb{E}_{x \mid y}[f(x)]\):

* A **Gaussian-process latent** for `f(x)` gives a fully non-parametric,
  uncertainty-rich representation but does not admit closed-form
  aggregation under general `A`. Sampling is closed form; integration
  through a linear operator with additive Gaussian noise *is* closed
  form for a GP, but mixing GP posteriors across aggregation
  configurations is heavier than the operator calculus shown here.
* A **finite Gaussian-mixture latent** (what `proxymix` ships) is less
  flexible but admits the full operator calculus on this page. The
  trade-off is bias against algebraic tractability. The choice depends
  on how much closed-form composition the downstream pipeline needs.

# References

* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.
* Kalman, R. E. (1960). *A new approach to linear filtering and
  prediction problems.* Transactions of the ASME — J. Basic Engineering.
* Murphy, K. P. (2012). *Machine Learning: A Probabilistic Perspective.*
  MIT Press. Ch. 4 (Gaussian models), Ch. 18 (state-space).
