---
title: "Compressing a Bayesian posterior you can evaluate but not sample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compressing a Bayesian posterior you can evaluate but not sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~3 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

A modeller has written down a likelihood and a prior, can evaluate their
product at any parameter value, and does not want to spend the afternoon
tuning a sampler to get four numbers out of it. The question is: *I have an
unnormalised log-posterior I can evaluate and no appetite for tuning a
sampler -- can I get the marginal, the tail probability, the evidence and an
error bar out of it?*

This is the workflow the package exists for, end to end: an unnormalised
log-posterior in, and a compact object out that can be marginalised,
conditioned, integrated, and given error bars, with no Markov chain in
between.

## What

[`gmm_target()`] wraps a log-density into the object the fitting verbs
consume. Declaring `normalised = FALSE` records that the normalising
constant is unknown, which is what an unnormalised posterior is.

[`fit_proxymix()`] with `regime = "kld"` runs the importance-sampled KLD-EM
of van der Hoek and Elliott (2024) against that target. Its `proposal`
argument takes the importance proposal, and `adapt = "pmc"` refreshes that
proposal from the current fit as the run proceeds, so the initial proposal
only has to be survivable rather than good. [`proposal_mvt()`] builds a
multivariate Student-t proposal, whose heavy tails are the robust default.

[`gmm_fit_quality()`] returns the certificate that travels with the fit: the
effective sample size of the importance weights, a degeneracy flag, the
support fraction, and a held-out validation gap. It is the part to read
before anything else.

[`gmm_evidence()`] estimates the log marginal likelihood, using the fitted
proxy as its own importance proposal. [`gmm_marginalise()`], [`pgmm()`] and
[`qgmm()`] read marginals, tail probabilities and quantiles off the fitted
mixture in closed form, with no further target evaluations.

[`gmm_fit_ensemble()`] and [`proxy_functional_ci()`] price the fit's own
sampling variability: the ensemble re-weights the fit's cached importance
draws, so a bootstrap interval on any functional of the mixture costs no new
posterior evaluations at all.

## Do

### A real unnormalised posterior

A Bayesian logistic regression of transmission type on weight for the
built-in `mtcars` data, under a flat prior. The log-posterior is the
log-likelihood, known only up to its normalising constant, which is the
model evidence.

```{r posterior}
set.seed(20260705)

y <- mtcars$am
w <- mtcars$wt

log_post <- function(theta) {
  if (is.null(dim(theta))) theta <- matrix(theta, ncol = 2L)
  eta <- outer(rep(1, length(y)), theta[, 1L]) + outer(w, theta[, 2L])
  colSums(y * eta - log1p(exp(eta)))
}

tgt <- gmm_target(
  n_dim = 2L,
  log_density = log_post,
  normalised = FALSE,
  name = "logistic(am ~ wt)"
)
```

### Fit the proxy

A cheap point fit locates the proposal. This is the usual workflow: any
rough guess of location and scale will do, because with `adapt = "pmc"` the
proposal is refreshed from the fit itself.

Population Monte Carlo re-estimates the fitting objective on a refreshed
importance sample at each refresh, so consecutive values of that objective
are not measured against the same draws and the run's monotonicity check
reports the jump. The handler below captures that notice so it can be read
alongside the certificate rather than passing unremarked.

```{r fit}
mle <- stats::glm(am ~ wt, data = mtcars, family = stats::binomial())
q0 <- proposal_mvt(2L, mean = stats::coef(mle),
                   sigma = 9 * stats::vcov(mle), df = 5)

fit_notes <- character(0)
fit <- withCallingHandlers(
  fit_proxymix(tgt, N = 2L, regime = "kld", proposal = q0,
               is_size = 3000L, max_iter = 60L, seed = 1L,
               adapt = "pmc"),
  proxymix_nonmonotone = function(cond) {
    fit_notes <<- c(fit_notes, conditionMessage(cond))
    invokeRestart("muffleWarning")
  }
)

cert <- gmm_fit_quality(fit)
n_refresh <- fit@diagnostics$n_refresh
cat(fit_notes, sep = "\n")
```

```{r fit-quality-table, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c("regime", "converged", "degenerate",
                 "effective sample size", "effective sample size, relative",
                 "smallest per-component effective size", "largest weight",
                 "support fraction", "importance-sampled divergence",
                 "held-out validation gap"),
    value = c(cert$regime, cert$converged, cert$degenerate,
              round(cert$ess, 1), round(cert$ess_relative, 3),
              round(cert$min_component_ess, 1),
              signif(cert$max_weight, 3), cert$support_fraction,
              round(cert$kld_final, 3), signif(cert$validation_gap, 3))
  ),
  row.names = FALSE,
  col.names = c("Certificate entry", "Value"),
  caption = paste0(
    "The fit-quality certificate that travels with the proxy through every ",
    "operation below."
  )
)
```

### The evidence, with an independent check

The fitted proxy doubles as the importance proposal for the model evidence,
the log marginal likelihood under the flat prior. A Laplace approximation
computed directly from the maximum-likelihood fit gives an independent
first-order check that shares none of the importance machinery.

```{r evidence}
ev <- gmm_evidence(fit, n = 4000L, seed = 2L)

## Laplace approximation: log f(theta_hat) + (d/2) log(2 pi)
## - (1/2) log det(-Hessian).
H <- -solve(stats::vcov(mle))
log_z_laplace <- log_post(matrix(stats::coef(mle), nrow = 1L)) +
  log(2 * pi) -
  0.5 * as.numeric(determinant(-H, logarithm = TRUE)$modulus)

evidence_gap <- abs(ev$log_z - log_z_laplace)
c(importance = round(ev$log_z, 3), se = signif(ev$se_log_z, 3),
  laplace = round(log_z_laplace, 3), gap = round(evidence_gap, 3))
```

### Closed-form reads off the proxy

Everything below is exact algebra on the fitted mixture: no chains, no
further target evaluations. The marginal posterior of the slope, the
probability that the slope is negative, and a central 90 per cent interval.

```{r reads}
slope <- gmm_marginalise(fit, keep = 2L)
p_negative <- pgmm(0, slope)
interval <- qgmm(c(0.05, 0.95), slope)
c(p_slope_negative = round(p_negative, 5),
  lower = round(interval[1L], 3), upper = round(interval[2L], 3))
```

### An independent reference for the marginal

The reads above are all functions of the fitted mixture, so on their own
they cannot say whether the mixture is right. In two dimensions the exact
posterior can be integrated on a grid instead, which uses the log-posterior
directly and never touches the fit. Exponentiating against the grid maximum
keeps the sum numerically stable; summing over the intercept axis leaves the
slope marginal, which the trapezoid rule then normalises.

```{r quadrature}
a_grid <- seq(2, 30, length.out = 220L)      # intercept
b_grid <- seq(-11, -1, length.out = 260L)    # slope
quad <- as.matrix(expand.grid(a = a_grid, b = b_grid))

log_dens <- log_post(quad)
dens <- matrix(exp(log_dens - max(log_dens)), nrow = length(a_grid))
da <- a_grid[2L] - a_grid[1L]
db <- b_grid[2L] - b_grid[1L]

marg_quad <- colSums(dens) * da
marg_quad <- marg_quad / (sum(marg_quad) * db)
marg_proxy <- dgmm(matrix(b_grid, ncol = 1L), slope)

curve_gap <- max(abs(marg_quad - marg_proxy))
c(peak_density = signif(max(marg_quad), 3),
  max_curve_gap = signif(curve_gap, 3),
  gap_pct_of_peak = round(100 * curve_gap / max(marg_quad), 1))
```

```{r marginal-figure, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4.2, fig.cap = "The slope marginal read off the fitted mixture in closed form (orange, dashed) against the same marginal obtained by numerical quadrature of the exact log-posterior (blue, solid). The quadrature curve uses the log-posterior directly and never touches the fit, so the agreement is an independent check rather than a self-consistency check.", fig.alt = "Two nearly coincident density curves for the slope parameter, one from the fitted Gaussian-mixture proxy and one from numerical quadrature of the exact posterior."}
marg_df <- rbind(
  data.frame(slope = b_grid, density = marg_quad,
             source = "Quadrature of the exact posterior"),
  data.frame(slope = b_grid, density = marg_proxy,
             source = "Mixture proxy, closed form")
)

ggplot2::ggplot(marg_df,
                ggplot2::aes(slope, density, colour = source,
                             linetype = source)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("Quadrature of the exact posterior" = "#0072B2",
               "Mixture proxy, closed form" = "#D55E00")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Quadrature of the exact posterior" = "solid",
               "Mixture proxy, closed form" = "dashed")
  ) +
  ggplot2::labs(
    title = "The slope marginal, two ways",
    x = expression(paste("slope ", beta[wt], " (log-odds per 1000 lb)")),
    y = "posterior density"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 legend.position = "top")
```

```{r marginal-figure-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the marginal-comparison",
    "figure is skipped.\n")
```

### Error bars on the proxy itself

The numbers above condition on the fitted mixture. The bootstrap ensemble
prices the fit's own sampling variability, at zero new posterior
evaluations, because the replicates re-weight the fit's cached importance
draws.

```{r ensemble}
ens <- gmm_fit_ensemble(fit, B = 80L, seed = 3L)
ci_mean <- proxy_functional_ci(ens, gmm_mean, level = 0.9)
ci_tail <- proxy_functional_ci(
  ens, function(g) pgmm(0, gmm_marginalise(g, keep = 2L)), level = 0.9
)
```

```{r ensemble-table, echo = FALSE}
knitr::kable(
  data.frame(
    functional = c("posterior mean, intercept", "posterior mean, slope",
                   "P(slope < 0)"),
    estimate = signif(c(ci_mean$estimate, ci_tail$estimate), 5),
    conf_low = signif(c(ci_mean$conf.low, ci_tail$conf.low), 5),
    conf_high = signif(c(ci_mean$conf.high, ci_tail$conf.high), 5)
  ),
  row.names = FALSE,
  col.names = c("Functional", "Estimate", "Lower 5%", "Upper 95%"),
  caption = paste0(
    "Bootstrap intervals over ", ens$B, " ensemble replicates, ",
    "obtained by re-weighting the fit's cached importance draws."
  )
)
```

## Read

The certificate is the first thing to read, and it clears the fit. The
effective sample size is `r round(cert$ess)` out of 3000 importance draws,
or `r round(100 * cert$ess_relative)` per cent, with the leanest single
component still carrying `r round(cert$min_component_ess)` effective draws
and no single weight above `r sprintf("%.5f", cert$max_weight)`. The fit is not
flagged degenerate, and the held-out validation gap of
`r signif(cert$validation_gap, 2)` says the fit does not depend on the
particular importance realisation it was estimated from. The proposal was
refreshed `r n_refresh` times over the run, which is what lifted the
effective sample size to that level from a proposal that was only ever a
rough guess. The certificate's divergence entry, `r round(cert$kld_final, 2)`,
is negative because the target is unnormalised: for such a target the
importance-sampled divergence is offset by the unknown $\log Z$, which the
fit records rather than hides, and the entry is read as a relative quantity
across fits of the same target rather than as an absolute divergence.

The evidence comes out at `r round(ev$log_z, 3)`, with a Monte Carlo
standard error of `r signif(ev$se_log_z, 2)`. The Laplace approximation,
which shares none of the importance machinery, gives
`r round(log_z_laplace, 3)`, a gap of `r round(evidence_gap, 3)` on the log
scale. A first-order approximation and an importance estimate agreeing to
within a tenth of a log unit is the check passing.

The closed-form reads answer the substantive question. The posterior
probability that the slope is negative is
`r round(100 * p_negative, 3)` per cent, and the central 90 per cent interval
for the slope runs from `r round(interval[1L], 2)` to
`r round(interval[2L], 2)` log-odds per 1000 lb: heavier cars are close to
certain to be automatics, and the effect size is not small.

The quadrature reference confirms that those reads are reads of the
posterior and not merely of the fit. The two marginal curves differ by at
most `r signif(curve_gap, 2)` in density, against a peak density of
`r signif(max(marg_quad), 3)`, so the largest disagreement anywhere on the
curve is about `r round(100 * curve_gap / max(marg_quad))` per cent of the
peak, and the figure shows it sitting at the mode where a two-component
mixture rounds off the exact posterior's slight asymmetry. The tails, which
are what the interval and the tail probability depend on, track closely.

The ensemble prices the remaining uncertainty in the fit itself. Over
`r ens$B` replicates the posterior mean of the slope is pinned to a 90 per
cent interval of width
`r signif(ci_mean$conf.high[2L] - ci_mean$conf.low[2L], 2)`, and the tail
probability to the interval
`r sprintf("%.5f", ci_tail$conf.low[1L])` to
`r sprintf("%.5f", ci_tail$conf.high[1L])`, both at no further evaluations of
the posterior.

So the answer to the question is yes, on all four counts, and the four
numbers arrive with a certificate saying how far they can be trusted.

## Limits

If the unnormalised posterior can be evaluated then a Markov chain can
always be run against it and a sample-based mixture fitted to the draws. The
direct fit earns its place when what is wanted *is* the compact closed-form
object -- deterministic given the seed, no chain tuning, evidence and error
bars included -- and not otherwise.

The dimension is the binding constraint. The effective sample size of
importance sampling falls sharply beyond roughly five to ten parameters, and
every fit reports it, so the certificate rather than the method's advertising
is what should decide. The posterior here has two parameters, which is the
comfortable end of that range. For high-dimensional posteriors, sample first
and use the classical expectation-maximisation regime on the draws.

The quadrature check above is available only because the posterior is
two-dimensional. It is not part of the workflow and does not scale: at four
or more parameters the grid becomes the expensive object it was introduced
to avoid, and the certificate and the held-out validation gap are what
remain.

The prior here is flat, and the evidence reported is therefore the marginal
likelihood under that flat prior. A proper prior changes the target, the
evidence and the interval, and nothing in this vignette speaks to how the
conclusions move under a different one.

The nearest CRAN neighbour, `AdMit`, adaptively fits a mixture of Student-t
densities to an evaluable kernel for use as a proposal. The Gaussian family
used here is what buys the closed-form operator calculus and the certificate
that travel with the result; a Student-t mixture would fit heavier tails
more naturally and give up that algebra.

## What to read next

*Fitting a proxy to a density you cannot sample* is the shorter first pass
over the same fitting verb, on a target whose shape is known in advance.
*Choosing between the three fitting regimes* explains why an unnormalised
posterior lands in regime (iii) and what the other two regimes need instead.
*The closed-form operator calculus on a mixture* covers the operators that
make the compressed posterior worth having: pushforward, observation update,
marginalisation and conditioning. *Reading the entropy of a fitted mixture*
adds the entropy and divergence diagnostics that summarise a fitted
posterior's spread.

## References

* Cappé, O., Douc, R., Guillin, A., Marin, J.-M. and Robert, C. P.
  (2008). *Adaptive importance sampling in general mixture classes.*
  Statistics and Computing 18, 447--459. <doi:10.1007/s11222-008-9059-x>.
* Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.
* Owen, A. and Zhou, Y. (2000). *Safe and effective importance sampling.*
  Journal of the American Statistical Association 95(449), 135--143.
  <doi:10.1080/01621459.2000.10473909>.
* Rubin, D. B. (1981). *The Bayesian bootstrap.* The Annals of Statistics
  9(1), 130--134. <doi:10.1214/aos/1176345338>.

## Reproduce

The ambient seed is 20260705. Each stochastic step additionally carries its
own seed, so none of them depends on the ambient random-number state: the
fit uses `seed = 1L`, the evidence estimate `seed = 2L`, and the bootstrap
ensemble `seed = 3L`. The quadrature reference is deterministic.

```{r session-info}
sessionInfo()
```
