---
title: "Fitting a proxy to a density you cannot sample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a proxy to a density you cannot sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

An analyst arrives with a function and no sampler. *I can write down a
function that returns the log-density of the thing I care about, and I have
no way to draw from it. Can I get a compact object I am able to sample,
marginalise and condition -- and how would I know the fit is any good?*
Both halves matter. The first is what the package does; the second is what
stops a plausible-looking mixture being trusted when the numerical
machinery behind it has quietly failed.

## What

A `gmm_target()` is the target: a dimension, a `log_density` callable, and
optionally samples. The built-in `banana_target()` used below carries an
exact, normalised log-density and deliberately no samples, so only the
evaluate-only regime applies to it.

`fit_proxymix()` is the single fitting verb. It routes to one of three
regimes on what the target can supply, and the table in *Do* below states
the routing rule. The regime that gives the package its reason to exist is
`"kld"`: importance-sampled KLD-EM, which needs only point-wise evaluation
of the target.

A `proposal_mvt()` is the importance proposal that regime feeds on: a
multivariate Student-t whose heavy tails keep the self-normalised weights
from collapsing onto a handful of draws.

`gmm_fit_quality()` is the fit's certificate: whether the iteration
converged, whether the importance sample degenerated, the effective sample
size, the largest single weight, and the held-out validation gap. It
answers the second half of the question, and it travels with the fit
through every operation downstream.

`dgmm()`, `rgmm()`, `gmm_marginalise()` and `gmm_conditionalise()` are the
closed-form reads off the fitted mixture: density, sampling, integrating a
coordinate out, and fixing one to a value. None of them re-evaluates the
target.

## Do

### The routing rule

```{r regime-table, echo = FALSE}
regime_tbl <- data.frame(
  Regime = c("`\"moment\"`", "`\"sample\"`", "`\"kld\"`"),
  Applies = c(
    "`N == 1` and the target carries samples",
    "`N >= 2` and the target carries samples",
    "the target carries `log_density` only"
  ),
  Method = c(
    "Closed-form moment matching",
    "Classical expectation-maximisation",
    "Importance-sampled KLD-EM"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  regime_tbl,
  caption = paste(
    "How `fit_proxymix()` routes under `regime = \"auto\"`. The three",
    "regimes are those of van der Hoek and Elliott (2024)."
  )
)
```

Only the third row is available for a target that cannot be sampled, and
that is the setting the sample-based mixture packages on CRAN (`mclust`,
`mixtools`, `flexmix`) do not address: all three of them require
independent draws.

### A target with no sampler

The bundled banana target is a non-Gaussian shape in the plane, obtained by
warping an isotropic Gaussian through a quadratic. Its log-density is exact
and normalised. Samples are not attached, so the routing table leaves one
option.

```{r banana}
tgt <- banana_target()
tgt
```

### Fit the proxy

A three-component proxy by KLD-EM against a Student-t proposal with five
degrees of freedom.

```{r fit}
proposal <- proposal_mvt(n_dim = 2L, mean = c(0, 0),
                         sigma = 4 * diag(2), df = 5)
fit <- fit_proxymix(tgt, N = 3L, regime = "kld",
                    proposal = proposal,
                    is_size = 2000L,
                    max_iter = 60L,
                    seed = 1L)
fit
```

### Read the certificate before anything else

```{r certificate}
quality <- gmm_fit_quality(fit)
```

```{r certificate-table, echo = FALSE}
cert_tbl <- data.frame(
  Field = c("regime", "converged", "degenerate", "effective sample size",
            "ESS as a fraction of draws", "smallest per-component ESS",
            "largest single weight", "final KL estimate",
            "held-out validation gap"),
  Value = c(
    quality$regime,
    as.character(quality$converged),
    as.character(quality$degenerate),
    format(round(quality$ess, 1L), nsmall = 1L),
    format(round(quality$ess_relative, 3L), nsmall = 3L),
    format(round(quality$min_component_ess, 1L), nsmall = 1L),
    format(signif(quality$max_weight, 3L)),
    format(signif(quality$kld_final, 3L)),
    format(signif(quality$validation_gap, 3L))
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  cert_tbl,
  caption = paste(
    "The fit certificate returned by `gmm_fit_quality()`. It is the first",
    "thing to read and the last thing to discard."
  )
)
```

### Overlay the proxy on the target

```{r overlay-grid}
grid_x <- seq(-3, 3, length.out = 120L)
grid_g <- expand.grid(x1 = grid_x, x2 = grid_x)
grid_mat <- as.matrix(grid_g)
grid_g$target <- exp(tgt@log_density(grid_mat))
grid_g$proxy <- dgmm(grid_mat, fit)
```

```{r overlay, eval = has_ggplot2, echo = has_ggplot2, fig.cap = sprintf("The banana target (filled contours) with the fitted three-component proxy overlaid as dashed contours. The proxy bends along the ridge rather than covering it with one blob: the fit reached an estimated KL of %s from an effective sample size of %s draws out of %d.", format(signif(fit@diagnostics$kld_final, 2L)), format(round(fit@diagnostics$ess, 0L)), fit@diagnostics$is_size), fig.alt = "Filled contour map of the curved banana density with dashed contours of the three-component Gaussian-mixture proxy following the same ridge."}
ggplot2::ggplot(grid_g, ggplot2::aes(x1, x2)) +
  ggplot2::geom_contour_filled(ggplot2::aes(z = target), bins = 10L,
                               alpha = 0.85) +
  ggplot2::geom_contour(ggplot2::aes(z = proxy), colour = "white",
                        linetype = "dashed", linewidth = 0.45, bins = 8L) +
  ggplot2::scale_fill_viridis_d(option = "mako", guide = "none") +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Target (filled) and fitted proxy (dashed)",
    x = expression(x[1]), y = expression(x[2])
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r overlay-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the target-and-proxy",
    "overlay figure is skipped.\n")
```

### The closed-form reads

Because the fit is a Gaussian mixture, integrating a coordinate out and
fixing one to a value are exact algebra, not further target evaluations.

```{r operations}
gmm_marginalise(fit, keep = 1L)
gmm_conditionalise(fit, given = c(NA, 0.5))
```

Sampling from the proxy is direct, and unlike the target it costs nothing.

```{r sample}
draws <- rgmm(500L, fit)
dim(draws)
```

## Read

The fit answers the first half of the question. The proxy is a
`r gmm_n_components(fit)`-component mixture in
`r gmm_dim(fit)` dimensions, reached in
`r length(kld_trace(fit))` KLD-EM iterations, and it supports sampling,
marginalisation and conditioning with no further access to the target at
all. The overlay shows the three components strung along the ridge rather
than one component sitting across it, which is the whole reason for
fitting a mixture instead of a single Gaussian.

The certificate answers the second half. The iteration converged, and the
fit is not flagged as degenerate. The importance sample retained an
effective size of `r round(fit@diagnostics$ess, 0L)` out of
`r fit@diagnostics$is_size` draws, a retention of
`r round(100 * fit@diagnostics$ess_relative, 1L)` per cent, and the largest
single self-normalised weight is
`r format(signif(fit@diagnostics$max_weight, 2L))`, so no individual draw
is carrying the fit. The smallest per-component effective sample size is
`r round(quality$min_component_ess, 0L)`, so every one of the three
components was estimated from a substantial share of the draws rather than
from a handful. The final KL estimate is
`r format(signif(fit@diagnostics$kld_final, 2L))` with a Monte Carlo
standard error of `r format(signif(fit@diagnostics$mc_se_kld, 2L))`, and the
held-out validation gap is
`r format(signif(quality$validation_gap, 2L))`: the fit does no better on
the draws it was trained on than on an independent set, which is what rules
out the proxy having chased one particular importance sample.

Read the certificate in that order -- converged, degenerate, effective
sample size, largest weight, validation gap -- and a fit that has failed
announces itself before any downstream number is taken seriously.

## Limits

This vignette fits one target in two dimensions and reads the fit that
results. It does not choose the number of components, which is left at
three by hand here; `select_N()` and `bic_aic()` do that job, and a
component count too low for the shape shows up as a residual KL that more
importance draws will not reduce. It also says nothing about how the choice
of proposal changes the answer. The Student-t proposal used above is wide
enough to cover the banana and heavy-tailed enough to keep the weights
spread, and a proposal that misses the target's mass produces a fit whose
certificate is worthless even though the printed mixture looks ordinary.

The effective sample size reported here is a two-dimensional number.
Importance sampling degrades quickly with dimension, and a proxy of the
same nominal quality in five or ten dimensions demands far more draws for
the same retention. The certificate is what makes that degradation visible
rather than silent, and it is reported on every regime-(iii) fit.

## What to read next

*Choosing between the three fitting regimes* takes the routing table above
and works all three regimes end to end on a target whose ground truth is
known, so the cost of choosing the wrong one is visible.

*How well a mixture proxies four awkward shapes* pushes regime (iii) across
a bent ridge, an annulus, a well-separated mixture and a compactly
supported density, and shows the package refusing a fit whose importance
sample has collapsed.

*The closed-form operator calculus on a mixture* takes the marginalisation
and conditioning shown above and adds pushforward through a linear sensor,
the Bayesian update on a noisy observation, and filtering over time.

*Compressing a Bayesian posterior you can evaluate but not sample* is this
workflow applied to an unnormalised log-posterior, with the model evidence
and error bars on the fit itself.

## References

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
Gaussians.* Stochastic Analysis and Applications.
<https://doi.org/10.1080/07362994.2024.2372605>.

## Reproduce

Every fit above is seeded through `fit_proxymix(seed = 1L)`, and the draws
in the sampling chunk come from that same fit, so the whole vignette is a
deterministic function of the package and the shipped target.

```{r session-info}
sessionInfo()
```
