---
title: "Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
set.seed(20260513)
```

```{r setup}
library(proxymix)
```

`proxymix` fits Gaussian-mixture *proxies* to user-supplied target densities. The unified verb is

```r
fit_proxymix(target, N, regime = c("auto", "moment", "sample", "kld"))
```

— and the three fitting regimes come from Hoek and Elliott (2024):

| Regime       | When it applies                                | Method                                |
|--------------|------------------------------------------------|---------------------------------------|
| `"moment"`   | `N == 1` and the target carries samples        | Closed-form moment matching           |
| `"sample"`   | `N >= 2` and the target carries samples        | Classical EM                          |
| `"kld"`      | The target carries `log_density` only          | **Importance-sampled KLD-EM**         |

The `"kld"` regime is the wedge: no other CRAN package fits Gaussian-mixture proxies when you can evaluate `f(x)` but cannot (cheaply) sample from it.

## A target you cannot sample from

We use the bundled "banana" target — a non-Gaussian 2-D shape obtained by warping an isotropic Gaussian through a quadratic. Its log-density is exact and normalised; we deliberately do *not* attach samples, so only regime (iii) applies.

```{r banana}
tgt <- banana_target()
tgt
```

A quick look at the target as a contour grid:

```{r banana-contour, fig.cap = "The banana target density."}
grid_x <- seq(-3, 3, length.out = 120)
grid_y <- seq(-3, 3, length.out = 120)
G <- expand.grid(x1 = grid_x, x2 = grid_y)
G$f <- exp(tgt@log_density(as.matrix(G)))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(G, aes(x1, x2, z = f)) +
    geom_contour_filled(bins = 12L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Banana target", x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 12)
}
```

## Fit a Gaussian-mixture proxy

We fit a 3-component proxy by KLD-EM with importance sampling against a multivariate-Student-t proposal. The fit runs in a fraction of a second.

```{r fit}
proposal <- is_mvt(n_dim = 2L, mean = c(0, 0),
                   sigma = 4 * diag(2), df = 5)

fit <- fit_proxymix(tgt, N = 3L, regime = "kld",
                    proposal = proposal,
                    is_size = 2000L,
                    max_iter = 30L,
                    seed = 1L)
fit
```

Diagnostics:

```{r diagnostics}
data.frame(
  iter = seq_along(kld_trace(fit)),
  kld  = kld_trace(fit)
)
ess_trace(fit)
```

## Overlay the proxy on the target

```{r overlay, fig.cap = "Banana target (filled) with the 3-component proxy overlaid (dashed contours)."}
G$g <- exp(dgmm(as.matrix(G[, c("x1", "x2")]), fit, log = TRUE))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot(G, aes(x1, x2)) +
    geom_contour_filled(aes(z = f), bins = 10L, alpha = 0.8) +
    geom_contour(aes(z = g), colour = "white", linetype = "dashed",
                 linewidth = 0.4, bins = 8L) +
    scale_fill_viridis_d(option = "mako", guide = "none") +
    coord_equal() +
    labs(title = "Target (filled) and proxy (dashed)",
         x = expression(x[1]), y = expression(x[2])) +
    theme_minimal(base_size = 12)
}
```

## Closed-form mixture operations

Because the fit is a Gaussian mixture, marginals and conditionals are closed-form and exact:

```{r ops}
gmm_marginalise(fit, keep = 1L)
gmm_conditionalise(fit, given = c(NA, 0.5))
```

You can sample from the proxy directly:

```{r sample}
x <- rgmm(500L, fit)
head(x)
```

## Where to next

* `vignette("three_regimes")` — what each of regimes (i), (ii), (iii) does, with a head-to-head comparison on a target whose ground truth is known.
* `vignette("density_shapes")` — the wedge demonstration. KLD-EM fit on three different non-Gaussian targets (banana, donut, three-mixture), with KLD / ESS / Hellinger reported.
* `vignette("operator_calculus")` — closed-form pushforward, Bayesian update, aggregation and conditioning on a fitted mixture.
* `vignette("from_kde")` — compressing a kernel density estimate into a Gaussian-mixture proxy.

## Reference

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate Gaussians.* Stochastic Analysis and Applications. <https://doi.org/10.1080/07362994.2024.2372605>.
