---
title: "Choosing between the three fitting regimes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing between the three fitting regimes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render() with ggplot2 installed;
     macOS arm64 (Apple silicon), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why

A modeller with a target in hand and three regimes on offer asks the
practical question: *which of the three does my problem actually need, and
what do I give up if I pick the wrong one?* The routing rule is easy to
state and easy to over-trust, because two of the regimes will happily run
on the same target and return visibly similar mixtures while answering
subtly different questions.

## What

The three regimes are those of van der Hoek and Elliott (2024), and they
differ in what they need to know about the target rather than in what they
return. All three return a `gmm_fit`.

`fit_moment_match()` is regime (i). At `N == 1` the KL-optimal Gaussian
proxy is the one that shares the target's mean and covariance, so the fit
is closed form and needs no iteration at all. It reads moments, which in
practice means it reads the samples the target carries.

`fit_em_samples()` is regime (ii): textbook expectation-maximisation on
independent draws, maximising the likelihood of the samples under an
`N`-component mixture. It needs samples and nothing else.

`fit_kld_em()` is regime (iii): importance-sampled KLD-EM, which minimises
$\mathrm{KL}(f \Vert g_\theta)$ using only point-wise evaluations of the
target's log-density, re-weighted by self-normalised importance weights
against a `proposal_mvt()` proposal. It needs no samples.

`fit_proxymix()` dispatches across all three, and `bic_aic()` and
`select_N()` choose the component count, which none of the three does for
you.

## Do

The bundled three-component mixture target is the right test bench: its
log-density is known exactly, and because the target is itself a Gaussian
mixture its true samples are available too. That is the one setting in
which all three regimes can be run on the same object and compared.

```{r target}
tgt <- mixture_target(with_samples = TRUE, n = 1500L, seed = 1L)
tgt
```

### Regime (i): closed-form moment matching

At `N == 1` the KL-optimal Gaussian proxy has the same mean and covariance
as the target. There is nothing to iterate.

```{r moment}
m_fit <- fit_proxymix(tgt, N = 1L, regime = "moment")
m_fit
```

The recovered moments are the empirical moments of the samples the target
carries, not an approximation to them. The only discrepancy is the
diagonal ridge `ridge_eps` that keeps the covariance positive-definite,
and setting it to zero removes even that.

```{r moment-recover}
m_fit_bare <- fit_proxymix(tgt, N = 1L, regime = "moment", ridge_eps = 0)
moment_gap <- c(
  mean = max(abs(m_fit@means[[1L]] - colMeans(tgt@samples))),
  covariance = max(abs(m_fit@covariances[[1L]] - cov(tgt@samples))),
  covariance_no_ridge = max(abs(m_fit_bare@covariances[[1L]] -
                                  cov(tgt@samples)))
)
signif(moment_gap, 3L)
```

A single Gaussian is a poor description of a three-mode mixture, but it is
the KL-optimal single Gaussian, and it makes a deterministic seed for the
two iterative regimes at `N > 1`.

### Regime (ii): classical EM on samples

When the target carries samples, the textbook algorithm applies:

* **E-step** -- \(\gamma_{nk} = \pi_k \mathcal{N}(x_n \mid \mu_k, \Sigma_k) /
  \sum_j \pi_j \mathcal{N}(x_n \mid \mu_j, \Sigma_j)\).
* **M-step** -- re-weight \(\pi_k, \mu_k, \Sigma_k\) by the
  responsibilities, with a diagonal ridge `epsilon_cov` keeping
  \(\Sigma_k\) positive-definite.
* **Convergence** -- relative change in the empirical log-likelihood below
  `tol`.

```{r em}
s_fit <- fit_proxymix(tgt, N = 3L, regime = "sample",
                      max_iter = 200L, n_starts = 4L)
s_fit
```

### Regime (iii): importance-sampled KLD-EM

Regime (iii) is the reason the package exists: it fits a proxy when the
target's `log_density` is evaluable but no samples are available, which is
the situation for Bayesian posteriors, intractable likelihoods, expensive
simulators, and downscaling targets in spatial statistics. The algorithm
draws `is_size` points from the proposal $q$, forms self-normalised weights
\(W_n \propto f(x_n) / q(x_n)\), and runs EM iterations in which the M-step
minimises \(\mathrm{KL}(f \Vert g_\theta)\) by re-weighting the
responsibilities with $W$.

The proposal is the only free knob beyond the usual mixture configuration,
and a Student-t with five degrees of freedom is a robust default.

```{r kld}
k_fit <- fit_proxymix(tgt, N = 3L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              mean = c(0, 0),
                                              sigma = 6 * diag(2),
                                              df = 5),
                      is_size = 3000L,
                      max_iter = 60L,
                      seed = 1L)
k_fit
```

### The three fits side by side

```{r overlay-grid}
grid_x <- seq(-4.5, 4.5, length.out = 100L)
grid_base <- expand.grid(x1 = grid_x, x2 = grid_x)
grid_mat <- as.matrix(grid_base)
target_d <- exp(tgt@log_density(grid_mat))
panel_of <- function(fit, label) {
  data.frame(
    x1 = grid_base$x1, x2 = grid_base$x2,
    target = target_d,
    proxy = dgmm(grid_mat, fit),
    regime = label,
    stringsAsFactors = FALSE
  )
}
overlay_df <- rbind(
  panel_of(m_fit, "(i) moment, N = 1"),
  panel_of(s_fit, "(ii) sample EM, N = 3"),
  panel_of(k_fit, "(iii) KLD-EM, N = 3")
)
overlay_df$regime <- factor(overlay_df$regime,
                            levels = unique(overlay_df$regime))
```

```{r overlay, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.4, fig.cap = "The three-mixture target (filled contours, identical in all three panels) with each regime's fitted proxy overlaid as dashed contours. Regime (i) is one Gaussian spread across all three modes; regimes (ii) and (iii) both place one component per mode and are visually indistinguishable, despite reading entirely different information about the target.", fig.alt = "Three side-by-side contour panels of the same three-mode target, overlaid with the single-Gaussian moment fit, the sample-EM fit and the KLD-EM fit."}
ggplot2::ggplot(overlay_df, ggplot2::aes(x1, x2)) +
  ggplot2::geom_contour_filled(ggplot2::aes(z = target), bins = 10L,
                               alpha = 0.85) +
  ggplot2::geom_contour(ggplot2::aes(z = proxy), colour = "white",
                        linetype = "dashed", linewidth = 0.4, bins = 5L) +
  ggplot2::scale_fill_viridis_d(option = "mako", guide = "none") +
  ggplot2::facet_wrap(~ regime) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = expression(x[1]), y = expression(x[2])) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r overlay-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the three-panel overlay",
    "figure is skipped.\n")
```

### What each iterative regime is driving down

```{r traces}
trace_df <- rbind(
  data.frame(
    iteration = seq_along(s_fit@diagnostics$loglik_trace),
    value = s_fit@diagnostics$loglik_trace,
    panel = "(ii) sample EM: log-likelihood (up)",
    stringsAsFactors = FALSE
  ),
  data.frame(
    iteration = seq_along(kld_trace(k_fit)),
    value = kld_trace(k_fit),
    panel = "(iii) KLD-EM: estimated KL (down)",
    stringsAsFactors = FALSE
  )
)
```

```{r traces-plot, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.2, fig.cap = "The two iterative regimes optimise different objectives and take different numbers of steps to do it. Sample EM climbs the empirical log-likelihood; KLD-EM drives down the importance-sampled Kullback-Leibler divergence.", fig.alt = "Two panels of iteration traces: a monotone increasing log-likelihood curve for sample EM and a decreasing Kullback-Leibler curve for KLD-EM."}
ggplot2::ggplot(trace_df, ggplot2::aes(iteration, value)) +
  ggplot2::geom_line(colour = "#0072B2", linewidth = 0.8) +
  ggplot2::geom_point(colour = "#0072B2", size = 1.1) +
  ggplot2::facet_wrap(~ panel, scales = "free") +
  ggplot2::labs(x = "iteration", y = "objective") +
  ggplot2::theme_minimal(base_size = 11)
```

```{r traces-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the two-panel iteration",
    "trace figure is skipped.\n")
```

### Regimes (i) and (iii) at N = 1, against an exact oracle

At `N = 1` both regime (i) and regime (iii) target the KL-optimal single
Gaussian, so they should agree. They are compared here on the banana
target, where regime (iii) would be the only option if the samples were
withheld, and both are compared against a third quantity neither of them
produced: the exact mean and covariance of the banana density, obtained by
deterministic quadrature on a fine grid.

```{r n1-fits}
banana <- banana_target(with_samples = TRUE, n = 2000L, seed = 1L)
m_b <- fit_proxymix(banana, N = 1L, regime = "moment")
k_b <- fit_proxymix(banana, N = 1L, regime = "kld",
                    proposal = proposal_mvt(n_dim = 2L,
                                            sigma = 4 * diag(2),
                                            df = 5),
                    is_size = 3000L, max_iter = 50L, seed = 1L)
```

```{r n1-oracle}
## Exact moments of the banana density by grid quadrature -- an oracle
## neither regime used, and independent of both the attached samples and
## the importance sampler.
quad_x <- seq(-8, 8, length.out = 400L)
quad_g <- as.matrix(expand.grid(x1 = quad_x, x2 = quad_x))
quad_cell <- (quad_x[2L] - quad_x[1L])^2
quad_f <- exp(banana@log_density(quad_g))
quad_mass <- sum(quad_f) * quad_cell
quad_mean <- c(sum(quad_f * quad_g[, 1L]),
               sum(quad_f * quad_g[, 2L])) * quad_cell / quad_mass
quad_trace <- sum(quad_f * ((quad_g[, 1L] - quad_mean[1L])^2 +
                              (quad_g[, 2L] - quad_mean[2L])^2)) *
  quad_cell / quad_mass
```

```{r n1-compare}
tr_of <- function(f) sum(diag(f@covariances[[1L]]))
n1 <- list(
  mean_gap = max(abs(m_b@means[[1L]] - k_b@means[[1L]])),
  trace_gap = tr_of(m_b) - tr_of(k_b),
  moment_error = abs(tr_of(m_b) - quad_trace),
  kld_error = abs(tr_of(k_b) - quad_trace),
  sample_error = abs(sum(diag(cov(banana@samples))) - quad_trace)
)
n1$closer <- n1$moment_error / n1$kld_error
signif(unlist(n1), 3L)
```

```{r n1-table, echo = FALSE}
n1_tbl <- data.frame(
  Source = c("regime (i), moment match", "regime (iii), KLD-EM",
             "attached samples", "exact density (quadrature)"),
  Reads = c("the attached samples", "the log-density",
            "the attached samples", "the log-density"),
  `Trace of covariance` = round(
    c(tr_of(m_b), tr_of(k_b), sum(diag(cov(banana@samples))), quad_trace),
    3L
  ),
  `Error against oracle` = round(
    c(tr_of(m_b), tr_of(k_b), sum(diag(cov(banana@samples))), quad_trace) -
      quad_trace,
    3L
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(
  n1_tbl,
  caption = paste(
    "The single-Gaussian proxy at `N = 1`, fitted two ways and scored",
    "against the exact moments of the banana density. The final row is",
    "the oracle, so its error is zero by construction."
  )
)
```

## Read

The moment fit is exact rather than approximate. It reproduces the
empirical mean of the target's samples with a gap of
`r format(signif(moment_gap[["mean"]], 2L))`, and the empirical covariance
with a gap of `r format(signif(moment_gap[["covariance"]], 2L))` -- which is
not a fitting error but the diagonal ridge itself, since removing the ridge
closes that gap to
`r format(signif(moment_gap[["covariance_no_ridge"]], 2L))` as well. Regime
(i) does not approximate the moments; it copies them.

The two iterative regimes land in the same place by different routes.
Sample EM took `r length(s_fit@diagnostics$loglik_trace)` iterations to
converge and KLD-EM took `r length(kld_trace(k_fit))`, and the fitted
weights agree to
`r format(signif(max(abs(sort(s_fit@weights) - sort(k_fit@weights))), 2L))`.
The overlay shows why they look alike here: both put one component on each
mode, which is the correct answer for a target that is itself a
three-component Gaussian mixture. The KLD-EM fit retained an effective
sample size of `r round(k_fit@diagnostics$ess, 0L)` out of
`r k_fit@diagnostics$is_size` importance draws, so it reached that answer
without leaning on a few high-weight points.

The `N = 1` comparison is where the cost of the choice becomes visible, and
it does not favour the regime that looks safer. The two fits differ in
mean by `r format(signif(n1$mean_gap, 2L))` and in trace-covariance by
`r format(signif(n1$trace_gap, 2L))`, which is larger than
importance-sampling noise alone would explain. The oracle column says why.
Regime (i) matches the moments of the `r nrow(banana@samples)` samples the
target happens to carry, and those samples overstate the exact
trace-covariance by `r format(signif(n1$sample_error, 2L))`. Regime (iii)
never sees those samples; it reads the density, and its trace-covariance is
out by only `r format(signif(n1$kld_error, 2L))`, about
`r round(n1$closer, 1L)` times closer to the truth.

So the answer to the opening question is that regime (iii) is not merely
the fallback when samples are missing. When the log-density is available,
it is fitting to the density itself, while regimes (i) and (ii) are fitting
to whatever finite sample was drawn from it. With plentiful samples the
distinction is immaterial; with a modest sample it is the difference
between the two answers above. What you give up by choosing regime (iii)
when samples are plentiful is speed and simplicity: it needs a proposal, it
costs `r k_fit@diagnostics$n_target_evals` target evaluations here, and it
carries an importance-sampling diagnostic that has to be read.

## Limits

This vignette does not choose the number of components. Every fit above
was told `N` by hand, and the three-mixture target was picked precisely
because the right answer is known to be three. On a real target the
component count is the first thing that has to be settled, and `bic_aic()`
and `select_N()` are what settle it; a count set too low shows up as a
residual divergence that more importance draws will not reduce.

The comparison between regimes (ii) and (iii) is flattering to both here
for a reason worth stating. The target is itself a Gaussian mixture with
the same number of components as the proxies, so the model is correctly
specified and there is no approximation error for either regime to trade
against the other. On a target outside the Gaussian-mixture family the two
regimes minimise genuinely different objectives, and the fits separate.

The `N = 1` oracle applies to two dimensions and to a target whose density
integrates to one on the grid used. Grid quadrature of this kind is a
reliable independent check in two or three dimensions and stops being one
well before the dimension at which importance sampling itself becomes the
binding constraint.

## What to read next

*Fitting a proxy to a density you cannot sample* is the shorter route into
regime (iii) alone, and reads the fit certificate that this vignette only
glances at.

*How well a mixture proxies four awkward shapes* takes regime (iii) to
targets outside the Gaussian-mixture family, where the correctly-specified
comfort of this vignette's test bench no longer holds.

*Reading the entropy of a fitted mixture* covers the component-count
question this vignette defers, including the annealed path that discovers a
count rather than being told one.

*One mixture, many methods* shows what a single fitted mixture replaces
once the regime question is settled.

*Mapping the optima of an objective* points regime (iii) at a function that
is not a density at all: an objective read as a Gibbs measure, so that the
fitted mixture returns every basin of a multimodal landscape at once rather
than the one a local optimiser happened to reach.

## References

Hoek, J. van der and Elliott, R. J. (2024). *Mixtures of multivariate
Gaussians.* Stochastic Analysis and Applications.
<https://doi.org/10.1080/07362994.2024.2372605>.

## Reproduce

The target carries `seed = 1L`, both regime-(iii) fits are seeded with
`seed = 1L`, and the quadrature oracle is deterministic, so every number
above is reproducible from the package alone.

```{r session-info}
sessionInfo()
```
