---
title: "Uplift and next-best-action with proxymix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uplift and next-best-action with proxymix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
set.seed(20260617)
```

```{r setup}
library(proxymix)
```

## One fit, the whole decision

A single Gaussian-mixture proxy over `(outcome, treatment, covariates)` yields --
in closed form, from that one fit -- prediction, heterogeneous treatment
effects, the revenue-maximising per-unit action, the off-line value of a
targeting policy, and an audit of what is and is not identified. The engine is
the causal ladder read off a fitted mixture: *seeing* is component
conditioning, *doing* is an `X`-only regime gate, *imagining* is per-component
abduction.

`fit_uplift()` assembles the joint fit and returns an `uplift_model`; every
verb below reads it without re-fitting.

```{r fit}
n <- 4000L
x <- stats::rnorm(n)
t <- stats::rbinom(n, 1L, 0.5)
## A heterogeneous effect: treatment helps more at higher x.
tau_true <- 0.4 + 1.2 * x
y <- 1 + 0.5 * x + tau_true * t + stats::rnorm(n, sd = 0.6)
dat <- data.frame(y = y, t = t, x = x)

model <- fit_uplift(dat, outcome = "y", treatment = "t", covariates = "x",
                    N = 3L, regime = "sample", max_iter = 200L, seed = 1L)
model
```

## Heterogeneous treatment effects

`proxy_cate()` returns the per-unit effect `$\tau(x)$` with a closed-form delta
standard error. At a single component the effect collapses to the ordinary
least-squares treatment coefficient; with several components the mixture
captures the heterogeneity a single line cannot.

```{r cate}
grid <- data.frame(x = seq(-1.5, 1.5, by = 0.5))
ce <- proxy_cate(model, newdata = grid)
ce
```

The estimates track the truth `$0.4 + 1.2x$`:

```{r cate-plot, fig.alt = "Estimated CATE against x with the analytic truth overlaid."}
plot(grid$x, ce$tau, type = "b", pch = 19,
     xlab = "x", ylab = expression(tau(x)),
     main = "Estimated vs true CATE")
lines(grid$x, 0.4 + 1.2 * grid$x, col = "red", lty = 2)
legend("topleft", c("proxymix", "truth"),
       col = c("black", "red"), lty = c(1, 2), pch = c(19, NA), bty = "n")
```

## The next-best action

Under a linear value model, `proxy_decide()` treats a unit when the value of
its effect exceeds the cost. It also reports the action-flip probability -- the
chance the recommendation would reverse under sampling noise.

```{r decide}
proxy_decide(model, grid, value = 1, cost = 0.5)
```

`proxy_policy_value()` scores a whole targeting policy off-line, with no live
experiment. The optimal policy weakly dominates treating everyone or no one.

```{r policy}
nd <- data.frame(x = stats::rnorm(1000))
rbind(
  none    = proxy_policy_value(model, nd, "none",    value = 1, cost = 0.5),
  all     = proxy_policy_value(model, nd, "all",     value = 1, cost = 0.5),
  optimal = proxy_policy_value(model, nd, "optimal", value = 1, cost = 0.5)
)
```

## The honest core: what is identified

The module rides only identified quantities. `proxy_identification_report()`
states the estimand, the assumption it is read under, the overlap rate, and the
confounding gap -- the value at risk if a latent regime confounds treatment and
outcome beyond `X`.

```{r report}
proxy_identification_report(model, nd)
```

The confounding gap is a *sensitivity signal*, not a correction. Under
ignorability the observational and do-operator effects coincide and the gap is
zero; a non-zero gap measures how far the estimate would move under the
latent-confounder reading.

```{r gap}
proxy_confounding_gap(model, grid)
```

### Counterfactuals: the mean, and only the mean

For an observed unit, `gmm_counterfactual()` returns the retrospective
counterfactual outcome of changing its treatment. Only the **mean** is
identified -- the spread of an individual counterfactual depends on an
unidentified cross-world coupling, so the variance and tail accessors refuse to
answer rather than mislead.

```{r counterfactual, error = TRUE}
ru <- proxy_retrospective_uplift(model, observed = dat[1:5, ])
ru

cf <- gmm_counterfactual(model@fit,
                         evidence = c(dat$y[1], dat$t[1], dat$x[1]),
                         do = c(NA, 1, NA), query = 1L)
gmm_cf_mean(cf)        # identified
gmm_cf_variance(cf)    # refused -- not identified
```

## A case the module loses

Episode 4's honesty finding stands: a Gaussian-mixture proxy is not a count
model. On pure over-dispersed count data, a purpose-built Poisson or
negative-binomial regression wins on predictive scoring, and we report the loss
rather than hide it.

```{r loss}
n2 <- 3000L
xc <- stats::rnorm(n2)
tc <- stats::rbinom(n2, 1L, 0.5)
lambda <- exp(0.3 + 0.5 * tc + 0.4 * xc)
yc <- stats::rpois(n2, lambda)
cd <- data.frame(y = yc, t = tc, x = xc)

train <- seq_len(2000L)
test <- setdiff(seq_len(n2), train)

## proxymix on the count outcome (continuous latent scale).
mc <- fit_uplift(cd[train, ], "y", "t", "x", N = 2L, regime = "sample",
                 outcome_type = "count", max_iter = 150L, seed = 2L)
## Predict each test unit under its own observed treatment.
pm_pred <- pmax(vapply(test, function(i) {
  proxy_predict(mc, cd[i, ], t = cd$t[i])$prediction
}, numeric(1L)), 1e-6)

## A purpose-built Poisson GLM.
glm_pois <- stats::glm(y ~ t + x, family = stats::poisson, data = cd[train, ])
glm_pred <- stats::predict(glm_pois, newdata = cd[test, ], type = "response")

## Mean Poisson deviance (lower is better).
dev <- function(yobs, mu) {
  mu <- pmax(mu, 1e-8)
  2 * mean(ifelse(yobs > 0, yobs * log(yobs / mu), 0) - (yobs - mu))
}
data.frame(
  model = c("proxymix (count)", "Poisson GLM"),
  mean_deviance = c(dev(yc[test], pm_pred), dev(yc[test], glm_pred))
)
```

The Poisson model scores better on its own ground. The proxymix reading still
delivers a usable effect estimate and a defensible audit -- but on pure count
signal it is not the tool to reach for, and the report says so.

## How it compares

The decision layer is benchmarked against independent implementations (`grf`
causal forest, S/T/X meta-learners, `DoubleML`) on synthetic ground truth and
two public uplift datasets. The script and full results are in
`system.file("benchmarks", package = "proxymix")`. In brief:

- **Synthetic (known truth).** Lower PEHE than `grf` on all four
  data-generating processes -- on smooth, low-dimensional structure the
  parametric mixture is more efficient than a forest.
- **Hillstrom email.** The best Qini coefficient of the five methods (53.2 vs
  the next-best 50.5), and an average effect matching the debiased `DoubleML`
  estimate to within 1.4%.
- **Scoring speed.** About 6x faster per-unit than `grf::predict` -- the
  vectorised closed-form serving path.
- **Criteo (an honest loss).** On a 12-dimensional, very-low-event-rate target
  proxymix is competitive but **fourth of five** (8.4% below the best), and
  underestimates the average effect. Tree ensembles capture the sparse signal
  better here, and we report it.

The edge is interpretable, auditable, fast closed-form uplift where the mixture
structure matches the data -- not universal accuracy dominance.

## Where this sits

The decision layer graduates the Episode-5 operators `gmm_intervene()` and
`gmm_counterfactual()` to first-class functions and builds the `proxy_*`
verbs on top. It targets the regime-(iii) wedge -- a target you can evaluate
but not sample from -- as much as the classical sampled case: a regime-(iii)
fit feeds the same verbs unchanged, inheriting the importance-sampling
effective-sample-size caveats. See `vignette("three_regimes")` for the regime
hierarchy and `vignette("roadmap")` for what graduates next.
```

