---
title: "Mapping the optima of an objective"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping the optima of an objective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~22 s under rmarkdown::render() with ggplot2, mclust,
     GenSA, DEoptim, cmaes, nloptr and optimx installed, the simulation read
     from its stored results; macOS arm64 (Apple silicon), R 4.6.1, one
     core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The problem

A calibration problem has landed on the desk of someone who has run
optimisers before and does not trust them here. The loss surface has more
than one basin, and every run from a different starting point lands
somewhere else. The question is: *my objective has more than one minimum,
so how do I get all of them, and some sense of how the basins compare,
instead of whichever one a local optimiser happened to fall into?*

The whole set of optima can be treated as a density, and fitting a density
is what the package does.

## Package capabilities

An objective $f(x)$ to be minimised over a box defines an unnormalised
density, the Gibbs measure $\exp(-f(x) / T)$, whose mass concentrates on the
low regions of $f$ as the temperature $T$ falls. That density can be
evaluated point-wise but not directly sampled, which is exactly regime (iii)
of van der Hoek and Elliott (2024).

`from_objective()` fits a Gaussian-mixture proxy to $\exp(-f / T)$ by
cooling a short temperature ladder through the same importance-sampled
KLD-EM that `fit_kld_em()` runs, warm-starting each step from the previous
fit. Its `lower` and `upper` arguments are the search region: they set where
the optima are sought, and they scale the temperature ladder, the uniform
exploration of the proposal, and the initialisation. During fitting the
objective is evaluated only inside the box. Points outside it, and points
where the objective is not finite, receive a finite penalty. Setting `minimise = FALSE` concentrates the map
on the maxima of $f$ instead of the minima.

`gmm_modes()` reads the distinct optima off the fitted map by the
fixed-point mode search of Carreira-Perpiñán (2000), started from every
component mean. It returns the
mode locations, the mixture density at each one, and how many distinct modes
survived deduplication. The density at a mode is the height of the map at
that point: it measures how strongly the Gibbs measure concentrates there. It
is not the objective value, and it is not the mass of the basin.

`ess_summary()` reports the importance-sampling diagnostics that travel
with the fit, so the map can be judged before it is trusted.

The point of the construction is not a single best point but the whole
picture. A multimodal $f$ gives a multimodal Gibbs measure, so the fitted
mixture spreads its components across the basins and recovers the optima
*together*. The result is an ordinary `gmm`, so the closed-form operator
calculus -- marginalisation, conditioning, divergence -- applies to the map
of solutions like any other fitted mixture.

## Addressing the problem

### A bimodal objective in one dimension

The objective $(\theta^2 - 4)^2$ has two minima, at $\theta = -2$ and
$\theta = 2$. Fit the map and resolve it.

```{r bimodal}
set.seed(20260619)

f <- function(v) (v[1]^2 - 4)^2

fit <- from_objective(f, lower = -5, upper = 5, N = 6L,
                      is_size = 3000L, n_steps = 5L, seed = 1L)

modes <- gmm_modes(fit)
sort(round(modes$modes[, 1], 3))
```

The fitted object is a `gmm_fit`, so the importance-sampling diagnostics are
available alongside the map.

```{r bimodal-ess}
ess_1d <- ess_summary(fit)
c(ess = round(ess_1d$ess, 1), is_size = ess_1d$is_size,
  ess_relative = round(ess_1d$ess_relative, 3))
```

### Four minima at once: Himmelblau

Himmelblau's function has four equal minima. A single-point optimiser would
return one of them and miss the rest. `from_objective()` returns all four in
one fit, provided the number of components `N` is comfortably larger than the
number of optima so that a component can settle on each basin.

```{r himmelblau}
himmelblau <- function(v) {
  x <- v[1]
  y <- v[2]
  (x * x + y - 11)^2 + (x + y * y - 7)^2
}

fit2 <- from_objective(himmelblau, lower = c(-5, -5), upper = c(5, 5),
                       N = 10L, is_size = 4000L, n_steps = 6L, seed = 1L)

found <- gmm_modes(fit2)
found$n
```

Himmelblau's optima are known in closed form, so the recovered map can be
checked against them rather than against itself. Each recovered mode is
matched to its nearest analytic optimum, and the distance between the two is
the error the reader should judge the fit on.

```{r match}
truth <- rbind(c(3, 2), c(-2.805118, 3.131312),
               c(-3.779310, -3.283186), c(3.584428, -1.848126))

pair_dist <- as.matrix(dist(rbind(found$modes, truth)))
n_found <- nrow(found$modes)
pair_dist <- pair_dist[seq_len(n_found), n_found + seq_len(nrow(truth))]
nearest <- apply(pair_dist, 1L, which.min)
gap <- apply(pair_dist, 1L, min)

all_distinct <- length(unique(nearest)) == nrow(truth)
worst_gap <- max(gap)
data.frame(modes_found = found$n, one_per_optimum = all_distinct,
           worst_gap = round(worst_gap, 3))
```

```{r fixed-notation, include = FALSE}
## Small numbers are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
## A positive number rounded up to one significant digit, as LaTeX math
## a \times 10^{b}, for "below" bounds.
sci_above <- function(v) {
  e <- floor(log10(v))
  m <- ceiling(v / 10^e - 1e-9)
  if (m == 10) {
    m <- 1
    e <- e + 1
  }
  paste0("$", m, " \\times 10^{", e, "}$")
}
## A stored package version written as the package's DESCRIPTION writes it
## (2.2-8, not 2.2.8), when the installed version is the one stored.
desc_version <- function(pkg) {
  stored <- sim$versions[[pkg]]
  installed <- utils::packageDescription(pkg)$Version
  if (identical(as.character(package_version(installed)), stored)) {
    installed
  } else {
    stored
  }
}
```

```{r match-table, echo = FALSE}
knitr::kable(
  data.frame(
    recovered_x1 = round(found$modes[, 1], 3),
    recovered_x2 = round(found$modes[, 2], 3),
    analytic_x1 = round(truth[nearest, 1], 3),
    analytic_x2 = round(truth[nearest, 2], 3),
    distance = round(gap, 3),
    mixture_density = round(found$density, 3)
  ),
  col.names = c("Recovered $x_1$", "Recovered $x_2$", "Analytic $x_1$",
                "Analytic $x_2$", "Distance", "Mixture density"),
  caption = paste0(
    "Each recovered mode beside the analytic optimum nearest to it, with ",
    "the distance between them and the mixture density at the mode."
  )
)
```

The figure below puts the recovered modes and the analytic truth on the
log-objective surface, so a basin that was missed would be visible as a
valley with no orange point in it.

```{r himmelblau-map-figure, ref.label = "himmelblau-map", eval = has_ggplot2, echo = FALSE, fig.width = 6, fig.height = 4.6, out.width = "75%", fig.pos = "!htbp", fig.cap = "The recovered modes sit on the four analytic optima of Himmelblau's function, one per basin of the log-objective surface. No basin is left without a recovered mode.", fig.alt = "Log-objective surface of Himmelblau's function as a viridis raster with contour lines, the four recovered modes as orange circles and the analytic optima as black crosses, each pair coinciding."}
```

```{r himmelblau-map, eval = FALSE, echo = has_ggplot2}
xs <- seq(-5, 5, length.out = 140L)
ys <- seq(-5, 5, length.out = 140L)
surface <- expand.grid(x1 = xs, x2 = ys)
surface$log_f <- log1p(
  apply(as.matrix(surface[, c("x1", "x2")]), 1L, himmelblau)
)

mode_df <- data.frame(x1 = found$modes[, 1], x2 = found$modes[, 2])
truth_df <- data.frame(x1 = truth[, 1], x2 = truth[, 2])

ggplot2::ggplot(surface, ggplot2::aes(x1, x2)) +
  ggplot2::geom_raster(ggplot2::aes(fill = log_f), interpolate = TRUE) +
  ggplot2::geom_contour(ggplot2::aes(z = log_f), colour = "white",
                        linewidth = 0.2, alpha = 0.6, bins = 8L) +
  ggplot2::geom_point(data = truth_df, shape = 4, size = 4, stroke = 1.4,
                      colour = "#000000") +
  ggplot2::geom_point(data = mode_df, shape = 21, size = 3, stroke = 1,
                      fill = "#D55E00", colour = "#000000") +
  ggplot2::scale_fill_viridis_c(name = "log(1 + f)", option = "mako",
                                direction = -1) +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::labs(
    title = "One fit, four basins: the map over Himmelblau's minima",
    subtitle = "crosses: analytic optima; filled circles: recovered modes",
    x = expression(x[1]), y = expression(x[2])
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 panel.grid = ggplot2::element_blank())
```

```{r himmelblau-map-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the log-objective surface",
    "figure is skipped.\n")
```

## Interpretation

Both minima of the one-dimensional objective are recovered. The resolved
modes are
`r paste0("$", sprintf("%.3f", sort(modes$modes[, 1])), "$", collapse = " and ")`,
against the analytic $-2$ and $2$, a largest departure of
`r signif(max(abs(abs(modes$modes[, 1]) - 2)), 2)`. The importance sampler
behind that map was healthy, carrying an effective sample size of
`r round(ess_1d$ess)` out of `r ess_1d$is_size` draws, or
`r round(100 * ess_1d$ess_relative)` per cent of the nominal sample.

On Himmelblau, one fit returns `r found$n` distinct modes, and each is
matched to a *different* analytic optimum, so no optimum is claimed twice
and none is missed. The largest distance between a recovered mode and the
optimum it sits on is `r signif(worst_gap, 2)`, on a search box ten units
wide, so each mode lies within about 1/`r round(10 / worst_gap)` of the box width
of its optimum. The map's height over the four basins runs from
`r signif(min(found$density), 3)` to `r signif(max(found$density), 3)`, a
spread of about
`r round(100 * (max(found$density) / min(found$density) - 1))` per cent
between the lowest and the highest peak. Himmelblau's four optima are equal
in objective value, so a perfectly resolved map would put the same height on
each; the spread here is what a finite importance sample and ten components
covering four basins leave behind, and it is the quantity to watch when the
optima are *not* known to be equal.

The optima come out together rather than one at a time, and the height of
the map at each mode ranks the modes by how strongly the Gibbs measure
concentrates at them. A point optimiser returns one point. The basins its
search passes through are not returned as a ranked map.

## Limitations

Recovery is most reliable when the number of components `N` is larger than
the number of optima, because a component has to be free to settle on each
basin. Symmetric landscapes, where several optima are exchangeable, need the
most headroom; when in doubt, raise `N`. This vignette used ten components
for four optima and six for two, and does not establish what happens when
the headroom is removed.

The box is a modelling choice and not a formality. It sets where the optima
are sought and it scales the temperature ladder, the exploration of the
proposal, and the initialisation, so an optimum outside the box is not
merely harder to find, it is outside the question being asked. Both examples
here were given boxes that comfortably contain every optimum, which is the
easy case. Outside the box the fit sees only the penalty, so an optimum
beyond its edge is not reported.

The two examples above are one- and two-dimensional. Regime (iii) is driven
by importance sampling, whose effective sample size falls sharply as the
dimension grows, and the numerical illustration below shows how far it falls
on a five-parameter likelihood. The
effective sample size reported alongside every fit is the diagnostic to
watch, and a map fitted at a low effective sample size should not be read as
a reliable inventory of basins.

Finally, the height of the map at a mode is a property of the Gibbs measure
at the final temperature, not of the objective. It is not the objective
value, and it is not the mass of the basin: a wide shallow basin can carry
more total mass than a narrow deep one while standing lower at its mode. The
objective itself remains available and should be evaluated at the recovered
modes when what matters is which optimum is lowest rather than which the map
concentrates on. The numerical illustration that follows puts the map beside
five point optimisers at one evaluation budget, on a real likelihood and on
two test functions.

## Numerical illustration

```{r engines-illustration, include = FALSE}
competitors <- c("GenSA", "DEoptim", "cmaes", "nloptr", "optimx")
run_illustration <- all(vapply(c(competitors, "mclust", "MASS"),
                               requireNamespace, logical(1L),
                               quietly = TRUE))
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("GenSA, DEoptim, cmaes, nloptr, optimx, mclust or MASS is not installed",
    "on this build, so the real-data illustration is skipped.\n")
```

### The likelihood of a two-component mixture on the Old Faithful data

The objective is the negative log-likelihood of a two-component normal
mixture for the 299 waiting times between eruptions of the Old Faithful
geyser in `MASS::geyser` (Azzalini and Bowman, 1990; Venables and Ripley,
2002). The five parameters are the weight of the first component, the two
means and the two standard deviations, sought over the box $[0.05, 0.95]
\times [40, 100]^2 \times [1, 25]^2$. Swapping the two components leaves the
likelihood unchanged, so every minimum has a twin with the labels switched,
and the surface also holds local minima of no practical interest. The map is
fitted beside five point optimisers: GenSA (Xiang et al., 2013), DEoptim
(Mullen et al., 2011), cmaes (Hansen and Ostermeier, 2001),
nloptr's L-BFGS (Liu and Nocedal, 1989) from 20 random starts, and optimx
(Nash and Varadhan, 2011) running L-BFGS-B from 20 random starts. Every
method receives the same budget of 5000 objective evaluations, counted by a
wrapper around the objective, and its version is reported in the Reproduce
section. The reference is the maximum-likelihood fit of mclust (Scrucca et
al., 2016), whose expectation-maximisation maximises this same likelihood
but is not charged to the evaluation budget.

The competitors install from CRAN once and load as usual:

```{r install-competitors, eval = FALSE}
install.packages(c("GenSA", "DEoptim", "cmaes", "nloptr", "optimx",
                   "mclust"))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(GenSA)
library(DEoptim)
library(cmaes)
library(nloptr)
library(optimx)
library(mclust)
```

```{r geyser-objective, eval = run_illustration}
waiting <- MASS::geyser$waiting
lower <- c(0.05, 40, 40, 1, 1)
upper <- c(0.95, 100, 100, 25, 25)

# below the box a standard deviation near zero gives values lower than any
# inside, so the objective is Inf outside; the margin leaves room for the
# finite-difference gradients of the local optimisers at the box edges
margin <- 0.01 * (upper - lower)
geyser_nll <- function(v) {
  if (any(v < lower - margin | v > upper + margin)) return(Inf)
  d <- v[1] * dnorm(waiting, v[2], v[4]) +
    (1 - v[1]) * dnorm(waiting, v[3], v[5])
  -sum(log(d))
}

mle <- Mclust(waiting, G = 2L, modelNames = "V", verbose = FALSE)
nll_ref <- -mle$loglik
```

Which minima the surface holds is settled first, by 400 local descents from
random starts in the box, without any package under comparison. Each descent
below is run by `optim()` and is not charged to any method's budget.

```{r geyser-minima, eval = run_illustration}
set.seed(20260925)
descend <- function(x0) {
  optim(x0, geyser_nll, method = "L-BFGS-B", lower = lower, upper = upper,
        control = list(maxit = 500L))
}
scaled <- function(x) sweep(sweep(x, 2L, lower), 2L, upper - lower, "/")

starts <- matrix(runif(400L * 5L, lower, upper), 400L, byrow = TRUE)
ends <- t(apply(starts, 1L, function(x0) {
  o <- descend(x0)
  c(o$par, o$value)
}))
ends <- ends[order(ends[, 6L]), ]

# distinct end points: more than 0.05 apart, Euclidean, in box-scaled units
keep <- 1L
for (i in seq_len(nrow(ends))[-1L]) {
  gaps <- sqrt(rowSums((scaled(ends[keep, 1:5, drop = FALSE]) -
    matrix(scaled(ends[i, 1:5, drop = FALSE]), length(keep), 5L,
           byrow = TRUE))^2))
  if (all(gaps > 0.05)) keep <- c(keep, i)
}
minima <- ends[keep, , drop = FALSE]
```

```{r geyser-minima-table, eval = run_illustration, echo = FALSE}
minima_tbl <- data.frame(
  basin = paste0("M", seq_len(nrow(minima))),
  weight = round(minima[, 1L], 3), mean_1 = round(minima[, 2L], 2),
  mean_2 = round(minima[, 3L], 2), sd_1 = round(minima[, 4L], 2),
  sd_2 = round(minima[, 5L], 2),
  excess = fixed(minima[, 6L] - nll_ref, 2),
  stringsAsFactors = FALSE
)
knitr::kable(
  minima_tbl, row.names = FALSE, align = c("l", rep("r", 6L)),
  col.names = c("Basin", "Weight", "Mean 1", "Mean 2", "SD 1", "SD 2",
                "Excess over the mclust fit"),
  caption = paste0(
    "The distinct local minima of the negative log-likelihood found by 400 ",
    "local descents, each with its negative log-likelihood minus that of ",
    "the mclust maximum-likelihood fit."
  )
)
```

The descents end in `r nrow(minima)` distinct minima. M1 and M2 are the
maximum-likelihood fit and its label-switched twin, both within
`r fixed(max(abs(minima[1:2, 6L] - nll_ref)), 2)` of the mclust value. The others
lie `r fixed(minima[3L, 6L] - nll_ref, 1)` and more above it: one puts a
narrow component at the lower bound of the scale, the other collapses the
mixture onto one component. A method has found a basin when a local descent
from one of its reported solutions ends within 5% of the box of that
minimum.

```{r geyser-methods, eval = run_illustration}
budget <- 5000L

counted <- function(f) {
  count <- 0L
  list(f = function(v) {
    count <<- count + 1L
    f(v)
  }, count = function() count)
}

n_starts <- 20L
# a numerical gradient costs 2p + 1 calls per iteration
per_start <- budget %/% (n_starts * 11L)

# the ladder ends at temperature 1, where exp(-f) is the likelihood itself
map_quality <- list()
map_modes <- function(f, is_size, n_steps) {
  fit <- from_objective(f, lower, upper, N = 10L, scale = "loglik",
                        is_size = is_size, n_steps = n_steps, seed = 1L)
  map_quality[[as.character(is_size)]] <<- gmm_fit_quality(fit)
  gmm_modes(fit)$modes
}

methods_geyser <- list(
  proxymix = function(f) map_modes(f, 1000L, 3L),
  "proxymix, 26 000" = function(f) map_modes(f, 4000L, 6L),
  GenSA = function(f) {
    matrix(GenSA(fn = f, lower = lower, upper = upper,
                 control = list(max.call = budget, seed = 1L))$par, 1L)
  },
  DEoptim = function(f) {
    out <- DEoptim(f, lower, upper,
                   DEoptim.control(NP = 50L, itermax = 99L, trace = FALSE))
    matrix(out$optim$bestmem, 1L)
  },
  cmaes = function(f) {
    lambda <- 4L + floor(3 * log(5))
    matrix(cma_es((lower + upper) / 2, f, lower = lower, upper = upper,
                  control = list(maxit = budget %/% lambda))$par, 1L)
  },
  nloptr = function(f) {
    x0 <- matrix(runif(n_starts * 5L, lower, upper), n_starts, byrow = TRUE)
    t(apply(x0, 1L, function(x) {
      lbfgs(x, f, lower = lower, upper = upper,
            control = list(maxeval = per_start))$par
    }))
  },
  optimx = function(f) {
    x0 <- matrix(runif(n_starts * 5L, lower, upper), n_starts, byrow = TRUE)
    out <- multistart(x0, f, lower = lower, upper = upper,
                      method = "L-BFGS-B",
                      control = list(maxit = per_start, trace = 0))
    as.matrix(out[, 1:5])
  }
)

basins_of <- function(sol) {
  landed <- t(apply(sol, 1L, function(x) descend(x)$par))
  d <- as.matrix(dist(scaled(rbind(landed, minima[, 1:5]))))
  d <- d[seq_len(nrow(sol)), nrow(sol) + seq_len(nrow(minima)),
         drop = FALSE]
  which(apply(d, 2L, min) < 0.05)
}

# warnings are kept by method: low effective sample size, and any other
low_ess_geyser <- logical(0L)
other_warn_geyser <- 0L
geyser_res <- lapply(names(methods_geyser), function(method) {
  cf <- counted(geyser_nll)
  set.seed(1L)
  low_ess_geyser[method] <<- FALSE
  secs <- system.time(
    sol <- withCallingHandlers(methods_geyser[[method]](cf$f),
                               warning = function(w) {
                                 if (inherits(w, "proxymix_low_ess")) {
                                   low_ess_geyser[method] <<- TRUE
                                 } else {
                                   other_warn_geyser <<- other_warn_geyser + 1L
                                 }
                                 invokeRestart("muffleWarning")
                               })
  )[[3L]]
  data.frame(
    method = method, evaluations = cf$count(), reported = nrow(sol),
    excess = min(apply(sol, 1L, geyser_nll)) - nll_ref,
    basins = paste0("M", basins_of(sol), collapse = ", "),
    secs = secs, stringsAsFactors = FALSE
  )
})
geyser_res <- do.call(rbind, geyser_res)
```

```{r geyser-table, eval = run_illustration, echo = FALSE}
geyser_tbl <- geyser_res
geyser_tbl$excess <- fixed(geyser_tbl$excess, 2)
geyser_tbl$secs <- fixed(geyser_tbl$secs, 2)
knitr::kable(
  geyser_tbl, format = "simple", row.names = FALSE, align = c("l", "r", "r", "r", "l", "r"),
  col.names = c("Method", "Evaluations", "Reported", "Excess over mclust",
                "Basins", "Seconds"),
  caption = paste0(
    "Each method on the Old Faithful likelihood at a budget of ", budget,
    " objective evaluations: the negative log-likelihood of its best ",
    "reported solution minus the mclust fit, and the basins a local ",
    "descent from its reported solutions reaches. proxymix at a nominal ",
    "26 000 evaluations is shown beside the equal-budget rows."
  )
)
g_val <- function(method, what) geyser_res[[what]][geyser_res$method == method]
```

```{r geyser-facts, eval = run_illustration, include = FALSE}
at_mle <- geyser_res$method[abs(geyser_res$excess) < 0.05 &
                              geyser_res$reported == 1L]
one_point <- geyser_res[geyser_res$reported == 1L, ]
map_rows <- c("proxymix", "proxymix, 26 000")
n_low_geyser <- sum(low_ess_geyser[map_rows])
n_degenerate <- sum(vapply(map_quality, function(q) isTRUE(q$degenerate),
                           logical(1L)))
## a narrow component on a repeated waiting time, below the box's scale
## bound, beside one at the sample mean and standard deviation
repeated_wait <- as.numeric(names(which.max(table(waiting))))
nll_formula <- function(v) {
  -sum(log(v[1] * dnorm(waiting, v[2], v[4]) +
             (1 - v[1]) * dnorm(waiting, v[3], v[5])))
}
nll_narrow <- nll_formula(c(0.05, repeated_wait, mean(waiting), 0.01,
                            sd(waiting)))
## how many of the 400 random-start descents end in M1 or M2
d_start <- as.matrix(dist(scaled(rbind(ends[, 1:5], minima[1:2, 1:5]))))
n_start_mle <- sum(apply(d_start[seq_len(nrow(ends)), nrow(ends) + 1:2], 1L,
                         min) < 0.05)
```

At `r budget` evaluations, `r paste(at_mle, collapse = " and ")` reach the
maximum-likelihood value, `r one_point$method[which.max(one_point$excess)]`
ends `r fixed(max(one_point$excess), 2)` above it, and each of the three
reports one basin. The two multi-start methods report both label-switched
twins. They used `r g_val("nloptr", "evaluations")` and
`r g_val("optimx", "evaluations")` evaluations, more than the budget of
`r budget`, although each start was capped at `r per_start` evaluations
(nloptr) or iterations (optimx). The map at the shared budget reports
`r g_val("proxymix", "reported")` modes whose best value is
`r fixed(g_val("proxymix", "excess"), 1)` above the mclust fit. At the last
cooling step the effective sample size is
`r fixed(map_quality[["1000"]]$ess, 1)` of 1000 draws for the fit at the
shared budget and `r fixed(map_quality[["4000"]]$ess, 1)` of 4000 for the
fit at a nominal 26\ 000 evaluations, which used
`r g_val("proxymix, 26 000", "evaluations")`.
`r c("Neither proxymix fit carries", "One of the two proxymix fits carries", "Both proxymix fits carry")[n_degenerate + 1L]`
a degenerate quality certificate, and
`r c("neither fit", "one fit", "both fits")[n_low_geyser + 1L]` raised the
low effective-sample-size warning.
`r if (other_warn_geyser > 0L) paste0("The methods raised ", other_warn_geyser, " other warnings between them.") else ""`
`r if (n_degenerate > 0L) "A map whose last update rests on so few draws does not describe the likelihood surface, so its" else "A map's"`
modes are to be trusted only as far as an independent check confirms them.
Here
the check is a local descent, and from the modes at the shared budget it
reaches `r g_val("proxymix", "basins")`. On this surface the check separates
little: `r n_start_mle` of the `r nrow(ends)` random starts above also
descend to M1 or M2. At 26 000
evaluations the best mode is `r fixed(g_val("proxymix, 26 000", "excess"), 1)`
above the fit and the modes descend to
`r g_val("proxymix, 26 000", "basins")`. The objective returns `Inf` outside
the box for every method,
because the likelihood grows without bound as a standard deviation shrinks
towards zero. A component of weight 0.05 and standard deviation 0.01 on the
`r sum(waiting == repeated_wait)` waiting times of `r repeated_wait` minutes,
beside one at the sample mean and standard deviation, gives a negative
log-likelihood of `r fixed(nll_narrow, 1)`, below the mclust fit's
`r fixed(nll_ref, 1)`. One setting differs from the defaults: the map is
fitted with `scale = "loglik"`, which ends
the temperature ladder at 1, where the Gibbs measure is the likelihood
itself. The default ladder ends far above the scale on which the
log-likelihood varies around its maximum. The
excesses of `r paste0("$", fixed(min(geyser_res$excess), 2), "$")` are the tolerance at which
mclust's expectation-maximisation stops.
This is one run of one objective, so the simulation
below repeats the comparison on two functions whose optima are known.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("calibration_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
sim_value <- function(problem, method, what) {
  s1 <- sim$sim_tab$problem == problem & sim$sim_tab$method %in% method
  sim$sim_tab[[what]][s1]
}
```

```{r rastrigin-minima, include = FALSE}
## local minima of the shifted Rastrigin function on a grid of step 0.01:
## a grid point counts when it is lower than each of its neighbours
grid_x <- seq(-5.12, 5.12, length.out = 1025L)
rast_1d <- function(x, s) (x - s)^2 - 10 * cos(2 * pi * (x - s))
rast_grid <- outer(rast_1d(grid_x, sim$shift[1L]),
                   rast_1d(grid_x, sim$shift[2L]), "+")
n_grid <- length(grid_x)
padded <- matrix(Inf, n_grid + 2L, n_grid + 2L)
padded[seq_len(n_grid) + 1L, seq_len(n_grid) + 1L] <- rast_grid
is_min <- matrix(TRUE, n_grid, n_grid)
for (di in -1:1) {
  for (dj in -1:1) {
    if (di == 0L && dj == 0L) next
    is_min <- is_min &
      rast_grid < padded[seq_len(n_grid) + 1L + di, seq_len(n_grid) + 1L + dj]
  }
}
at_min <- which(is_min, arr.ind = TRUE)
on_edge <- at_min[, 1L] %in% c(1L, n_grid) | at_min[, 2L] %in% c(1L, n_grid)
rast_count <- list(interior = sum(!on_edge), edge = sum(on_edge))
## value of the four minima next to the global one
rast_neighbour <- optimize(function(z) 20 + rast_1d(z, 0) + rast_1d(0, 0),
                           c(0.5, 1.5))$objective
```

The simulation runs each method `r sim$n_rep` times on two functions of two
variables, with a fresh seed each run and the same budget of `r sim$budget`
evaluations. Himmelblau's function (Himmelblau, 1972) has four minima of
equal value on the box $[-5, 5]^2$. The Rastrigin function (Rastrigin, 1974)
on $[-5.12, 5.12]^2$ is shifted so that its global minimum sits at
$(`r sim$shift[1L]`, `r sim$shift[2L]`)$ rather than at the centre of the
box, where a method that starts from the centre would be given it at its
first evaluation. On this box the function has
`r rast_count$interior` local minima inside the box and
`r rast_count$edge` more on its edge. Its target basins are the global
minimum and the four neighbouring local minima about one unit away, each of
value `r fixed(rast_neighbour, 3)`.

The same counter wraps the objective for every method, and it also keeps
the points evaluated. A reported solution finds a basin when it lies within
`r sim$radius` of that optimum, and a run visits a basin when any evaluated
point does, which credits a uniform probe as much as a search. Each method
is scored on the basins its reported solutions find, on the value of its
best reported solution, on the evaluation at which its trajectory had
visited every basin, and on wall time. The two multi-start methods return
their 20 local solutions, the three global optimisers return one point, and
proxymix returns the modes of its map. A row at 26 000 evaluations, the
settings of the Himmelblau example earlier in this vignette, is shown beside
the equal-budget proxymix row. The label 26 000 is the nominal budget of
those settings, and the median run used
`r fixed(sim_value("Himmelblau", "proxymix, 26 000", "evaluations"), 0)`
evaluations on Himmelblau's function and
`r fixed(sim_value("Rastrigin", "proxymix, 26 000", "evaluations"), 0)` on the
Rastrigin function. This chunk is complete and runs as shown,
but it took about `r round(sim$elapsed_secs / 60)` minutes on one core (R
`r sim$versions[["R"]]`, Apple silicon); the tables below are read
from its stored output. For a quick reduced run, change the line
`n_rep <- 100L` to `n_rep <- 5L`.

```{r sim-code, eval = FALSE}
library(proxymix)
library(GenSA)
library(DEoptim)
library(cmaes)
library(nloptr)
library(optimx)

budget <- 5000L  # objective evaluations per run
n_rep <- 100L
radius <- 0.3    # a point within this distance of an optimum is in its basin

himmelblau <- function(v) (v[1]^2 + v[2] - 11)^2 + (v[1] + v[2]^2 - 7)^2
shift <- c(1.7, -2.3)
rastrigin <- function(v) {
  z <- v - shift
  20 + sum(z^2 - 10 * cos(2 * pi * z))
}

problems <- list(
  Himmelblau = list(
    f = himmelblau, lower = c(-5, -5), upper = c(5, 5),
    optima = rbind(c(3, 2), c(-2.805118, 3.131312),
                   c(-3.779310, -3.283186), c(3.584428, -1.848126))
  ),
  Rastrigin = list(
    f = rastrigin, lower = c(-5.12, -5.12), upper = c(5.12, 5.12),
    optima = rbind(shift, shift + c(1, 0), shift - c(1, 0),
                   shift + c(0, 1), shift - c(0, 1))
  )
)

# every method sees the objective through this counter, which keeps the
# first `budget` evaluated points
counted <- function(f) {
  count <- 0L
  trace <- matrix(NA_real_, budget, 2L)
  list(
    f = function(v) {
      count <<- count + 1L
      if (count <= budget) trace[count, ] <<- v
      f(v)
    },
    count = function() count,
    trace = function() trace[seq_len(min(count, budget)), , drop = FALSE]
  )
}

n_starts <- 20L
# a numerical gradient costs 2p + 1 calls per iteration
per_start <- budget %/% (n_starts * 5L)

methods <- list(
  proxymix = function(f, lower, upper, seed) {
    fit <- from_objective(f, lower, upper, N = 10L, is_size = 1000L,
                          n_steps = 3L, seed = seed)
    gmm_modes(fit)$modes
  },
  "proxymix, 26 000" = function(f, lower, upper, seed) {
    fit <- from_objective(f, lower, upper, N = 10L, is_size = 4000L,
                          n_steps = 6L, seed = seed)
    gmm_modes(fit)$modes
  },
  GenSA = function(f, lower, upper, seed) {
    out <- GenSA(fn = f, lower = lower, upper = upper,
                 control = list(max.call = budget, seed = seed))
    matrix(out$par, 1L)
  },
  DEoptim = function(f, lower, upper, seed) {
    out <- DEoptim(f, lower, upper,
                   DEoptim.control(NP = 50L, itermax = 99L, trace = FALSE))
    matrix(out$optim$bestmem, 1L)
  },
  cmaes = function(f, lower, upper, seed) {
    lambda <- 4L + floor(3 * log(length(lower)))
    out <- cma_es((lower + upper) / 2, f, lower = lower, upper = upper,
                  control = list(maxit = budget %/% lambda))
    matrix(out$par, 1L)
  },
  nloptr = function(f, lower, upper, seed) {
    starts <- matrix(runif(n_starts * 2L, lower, upper), n_starts,
                     byrow = TRUE)
    t(apply(starts, 1L, function(x0) {
      lbfgs(x0, f, lower = lower, upper = upper,
            control = list(maxeval = per_start))$par
    }))
  },
  optimx = function(f, lower, upper, seed) {
    starts <- matrix(runif(n_starts * 2L, lower, upper), n_starts,
                     byrow = TRUE)
    out <- multistart(starts, f, lower = lower, upper = upper,
                      method = "L-BFGS-B",
                      control = list(maxit = per_start, trace = 0))
    as.matrix(out[, 1:2])
  }
)

nearest_optimum <- function(points, optima) {
  d <- sqrt(pmax(outer(rowSums(points^2), rowSums(optima^2), "+") -
                   2 * points %*% t(optima), 0))
  list(basin = apply(d, 1L, which.min), gap = apply(d, 1L, min))
}

one_run <- function(problem, method, r) {
  pr <- problems[[problem]]
  cf <- counted(pr$f)
  set.seed(r)
  # flag the run's warnings; they still reach the caller
  low_ess <- FALSE
  other_warnings <- 0L
  secs <- system.time(
    sol <- withCallingHandlers(
      methods[[method]](cf$f, pr$lower, pr$upper, r),
      warning = function(w) {
        if (inherits(w, "proxymix_low_ess")) {
          low_ess <<- TRUE
        } else {
          other_warnings <<- other_warnings + 1L
        }
      }
    )
  )[[3L]]
  n_opt <- nrow(pr$optima)
  # reported solutions: which basins hold one
  rep <- nearest_optimum(sol, pr$optima)
  in_basin <- tabulate(rep$basin[rep$gap < radius], n_opt) > 0L
  # evaluated points: first evaluation inside each basin
  tr <- nearest_optimum(cf$trace(), pr$optima)
  first_visit <- vapply(seq_len(n_opt), function(b) {
    hit <- which(tr$basin == b & tr$gap < radius)
    if (length(hit) == 0L) NA_integer_ else hit[1L]
  }, integer(1L))
  basin <- rep(NA, 5L)
  basin[seq_len(n_opt)] <- in_basin
  data.frame(
    problem = problem, method = method, run = r,
    evaluations = cf$count(), secs = secs,
    n_reported = nrow(sol),
    best_value = min(apply(sol, 1L, pr$f)),
    basins_found = sum(in_basin),
    all_found = all(in_basin),
    global_found = in_basin[1L],
    basin_1 = basin[1L], basin_2 = basin[2L], basin_3 = basin[3L],
    basin_4 = basin[4L], basin_5 = basin[5L],
    visited_all = !anyNA(first_visit),
    # a run that never entered every basin counts as never (Inf)
    evals_to_all = if (anyNA(first_visit)) Inf else max(first_visit),
    low_ess = low_ess, other_warnings = other_warnings
  )
}

res <- do.call(rbind, lapply(names(problems), function(problem) {
  do.call(rbind, lapply(names(methods), function(method) {
    do.call(rbind, lapply(seq_len(n_rep), function(r) {
      one_run(problem, method, r)
    }))
  }))
}))

# rates over runs, and medians of the run-level quantities
by_mean <- aggregate(
  cbind(basin_1, basin_2, basin_3, basin_4, basin_5, all_found, global_found,
        basins_found, visited_all, low_ess) ~ problem + method,
  data = res, FUN = mean, na.action = na.pass
)
by_median <- aggregate(
  cbind(evaluations, best_value, evals_to_all, secs) ~ problem + method,
  data = res, FUN = median, na.action = na.pass
)
sim_tab <- merge(by_mean, by_median, by = c("problem", "method"))
sim_tab
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order <- c("proxymix", "proxymix, 26 000", "GenSA", "DEoptim", "cmaes",
                  "nloptr", "optimx")
tab_sim <- tab_sim[order(tab_sim$problem, match(tab_sim$method, method_order)), ]
basin_cols <- paste0("basin_", 1:5)
basin_tbl <- tab_sim[, c("problem", "method", basin_cols, "all_found")]
basin_tbl[, basin_cols] <- lapply(basin_tbl[, basin_cols], function(v) {
  ifelse(is.na(v), "--", fixed(v, 2))
})
basin_tbl$all_found <- fixed(basin_tbl$all_found, 2)
knitr::kable(
  basin_tbl, format = "simple", row.names = FALSE, align = c("l", "l", rep("r", 6L)),
  col.names = c("Function", "Method", "Basin 1", "Basin 2", "Basin 3",
                "Basin 4", "Basin 5", "All basins"),
  caption = paste0(
    "Fraction of ", sim$n_rep, " runs in which a reported solution lies ",
    "within ", sim$radius, " of each optimum, and of all optima at once. ",
    "Basin 1 of the Rastrigin function is the global minimum. Himmelblau's ",
    "function has four basins, so its Basin 5 cells are empty."
  )
)
```

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.5\textheight\clearpage\fi
```

```{r sim-table-2, echo = FALSE}
score_tbl <- data.frame(
  problem = tab_sim$problem, method = tab_sim$method,
  evaluations = round(tab_sim$evaluations),
  best_value = fixed(tab_sim$best_value, 3),
  visited_all = fixed(tab_sim$visited_all, 2),
  evals_to_all = ifelse(is.finite(tab_sim$evals_to_all),
                        as.character(round(tab_sim$evals_to_all)), "--"),
  secs = fixed(tab_sim$secs, 3),
  stringsAsFactors = FALSE
)
knitr::kable(
  score_tbl, format = "simple", row.names = FALSE, align = c("l", "l", rep("r", 5L)),
  col.names = c("Function", "Method", "Evaluations", "Best value",
                "Visited all", "Evals to visit all", "Seconds"),
  caption = paste0(
    "Over ", sim$n_rep, " runs per method: the median number of objective ",
    "evaluations used, the median of the best reported objective value, ",
    "the fraction of runs whose evaluated points entered every basin, the ",
    "median evaluation at which they had, and the median wall time. The ",
    "minimum of both functions is 0. A run whose evaluated points never ",
    "entered every basin counts as never in the median, so a dash marks a ",
    "method for which half or more of the runs never did."
  )
)
mcse_rate <- sqrt(0.5 * 0.5 / sim$n_rep)
res_h <- sim$res[sim$res$problem == "Himmelblau" &
                   sim$res$method %in% c("GenSA", "DEoptim", "cmaes"), ]
worst_run <- res_h[which.max(res_h$best_value), ]
## the latest run to enter every basin, among the runs that did
latest_visit <- function(problem) {
  s1 <- sim$res$problem == problem & sim$res$method == "proxymix"
  v <- sim$res$evals_to_all[s1]
  max(v[is.finite(v)])
}
never_visit <- function(problem) {
  s1 <- sim$res$problem == problem & sim$res$method == "proxymix"
  sum(!is.finite(sim$res$evals_to_all[s1]))
}
never_phrase <- function(problem) {
  n <- never_visit(problem)
  if (n == 0L) return("")
  paste0(", and ", n, " of its runs never entered every basin")
}
runs_phrase <- function(v) {
  if (all(v == 1)) return("every run")
  paste(paste(fixed(v, 2), collapse = " and "), "of runs")
}
## the uniform probe of 2000 points that begins every from_objective() fit
probe_n <- 2000L
median_visit_h <- sim_value("Himmelblau", "proxymix", "evals_to_all")
## runs flagged for a low effective sample size: the map at the shared
## budget on the Rastrigin function, and any other
s1 <- startsWith(sim$res$method, "proxymix")
s2 <- sim$res$problem == "Rastrigin" & sim$res$method == "proxymix"
low_ess_rast <- sum(sim$res$low_ess[s2])
low_ess_other <- sum(sim$res$low_ess[s1 & !s2])
if (low_ess_rast != sim$n_rep || low_ess_other != 0L) {
  stop("The text on low effective sample size no longer matches the stored ",
       "results.", call. = FALSE)
}
## final effective sample sizes quoted in those warnings
ess_final <- as.numeric(regmatches(
  sim$warnings,
  regexpr("(?<=final cooling step: ESS = )[0-9.]+", sim$warnings, perl = TRUE)
))
runs_decrease <- sum(sim$res$other_warnings[
  startsWith(sim$res$method, "proxymix")] > 0L)
n_all_five <- sum(sim$res$all_found[sim$res$problem == "Rastrigin" &
                                      startsWith(sim$res$method, "proxymix")])
## proxymix slowest per run on both functions at both budgets
map_slowest <- all(vapply(unique(sim$sim_tab$problem), function(p) {
  s1 <- sim$sim_tab$problem == p
  m <- startsWith(sim$sim_tab$method, "proxymix")
  min(sim$sim_tab$secs[s1 & m]) > max(sim$sim_tab$secs[s1 & !m])
}, logical(1L)))
## share of the map's evaluations the multi-start methods used on Himmelblau
multistart_share <- max(sim_value("Himmelblau", c("nloptr", "optimx"),
                                  "evaluations")) /
  sim_value("Himmelblau", "proxymix", "evaluations")
decrease_msg <- grep("decreased by", sim$warnings, value = TRUE)
```

On Himmelblau's function the two multi-start methods report all four basins
in `r fixed(sim_value("Himmelblau", "nloptr", "all_found"), 2)` and
`r fixed(sim_value("Himmelblau", "optimx", "all_found"), 2)` of runs, using
about `r round(sim_value("Himmelblau", "nloptr", "evaluations"))` and
`r round(sim_value("Himmelblau", "optimx", "evaluations"))` of the
`r sim$budget` evaluations, and their best value is at the minimum. GenSA,
DEoptim and cmaes each return one point, so they report one basin per run,
and all three reach a median best value below
`r sci_above(max(sim_value("Himmelblau", c("GenSA", "DEoptim", "cmaes"), "best_value")))`.
The worst single run, of `r worst_run$method`, ended at
`r fixed(worst_run$best_value, 3)`.
The map at `r sim$budget` evaluations reports all four basins in
`r fixed(sim_value("Himmelblau", "proxymix", "all_found"), 2)` of runs and
`r fixed(sim_value("Himmelblau", "proxymix", "basins_found"), 2)` of the four
on average, and its best mode sits at a value of
`r fixed(sim_value("Himmelblau", "proxymix", "best_value"), 2)`, because
three cooling steps of 1000 draws place the components near the basins
without resolving their floors. At 26 000 evaluations the map reports all
four in `r fixed(sim_value("Himmelblau", "proxymix, 26 000", "all_found"), 2)`
of runs with a best value of
`r fixed(sim_value("Himmelblau", "proxymix, 26 000", "best_value"), 2)`.
The evaluated points of GenSA and DEoptim enter all four basins in
`r fixed(sim_value("Himmelblau", "GenSA", "visited_all"), 2)` and
`r fixed(sim_value("Himmelblau", "DEoptim", "visited_all"), 2)` of runs and
those of cmaes in
`r fixed(sim_value("Himmelblau", "cmaes", "visited_all"), 2)`; for proxymix
the figure is `r fixed(sim_value("Himmelblau", "proxymix", "visited_all"), 2)`.
Its median run had entered every basin by evaluation
`r round(median_visit_h)`,
`r if (median_visit_h <= probe_n) "inside" else "after"` the uniform probe
of `r probe_n` points that begins every fit. The latest run to do so
reached evaluation `r latest_visit("Himmelblau")`, and on the Rastrigin
function `r latest_visit("Rastrigin")``r never_phrase("Rastrigin")`.

On the shifted Rastrigin function GenSA and DEoptim find the global basin in
`r runs_phrase(sim_value("Rastrigin", c("GenSA", "DEoptim"), "global_found"))`
and
cmaes, started at the centre of the box with its default step size, in
`r fixed(sim_value("Rastrigin", "cmaes", "global_found"), 2)`, with a best
value of `r fixed(sim_value("Rastrigin", "cmaes", "best_value"), 1)`. The
multi-start methods find it in
`r fixed(sim_value("Rastrigin", "nloptr", "global_found"), 2)` and
`r fixed(sim_value("Rastrigin", "optimx", "global_found"), 2)` of runs: their
20 descents end in whichever local basin they started in, so
they report `r fixed(sim_value("Rastrigin", "nloptr", "basins_found"), 2)`
and `r fixed(sim_value("Rastrigin", "optimx", "basins_found"), 2)` of the
five target basins on average and a best value near
`r fixed(rast_neighbour, 3)`, the floor of the neighbouring basins. The map finds the global basin in
`r fixed(sim_value("Rastrigin", "proxymix", "global_found"), 2)` of runs at
`r sim$budget` evaluations and
`r fixed(sim_value("Rastrigin", "proxymix, 26 000", "global_found"), 2)` at
26 000, reports `r fixed(sim_value("Rastrigin", "proxymix", "basins_found"), 2)`
and `r fixed(sim_value("Rastrigin", "proxymix, 26 000", "basins_found"), 2)`
of the five basins, and all five in `r n_all_five` of its
`r 2L * sim$n_rep` Rastrigin runs at the two settings. The low
effective-sample-size warning was raised in all `r low_ess_rast` Rastrigin
runs of the map at `r sim$budget` evaluations and in none of its other runs.
Each of those fits ended with an effective sample size of
`r fixed(min(ess_final), 1)` to `r fixed(max(ess_final), 1)` of 1000 draws
and was flagged as degenerate, so the map's Rastrigin scores at the shared
budget describe degenerate fits. A warning of a small decrease of the EM
objective was raised in `r runs_decrease` runs. A
rate estimated from `r sim$n_rep` runs carries a Monte Carlo standard error
of up to `r fixed(mcse_rate, 2)`.
`r if (map_slowest) "proxymix was the slowest method per run at either budget." else ""`

In this run the map reported more Himmelblau basins per run than the three
single-point optimisers, which report one point by construction, and fewer
than the two multi-start methods, which used at most
`r round(100 * multistart_share)` per cent of its evaluations. At the shared budget GenSA and DEoptim reached a median
best value below
`r sci_above(max(sim$sim_tab$best_value[sim$sim_tab$method %in% c("GenSA", "DEoptim")]))`
on both functions and cmaes on Himmelblau's function only, while the map's
best mode stayed at
`r fixed(sim_value("Himmelblau", "proxymix", "best_value"), 2)` and
`r fixed(sim_value("Rastrigin", "proxymix", "best_value"), 2)`, and the map
found the Rastrigin global basin in
`r fixed(sim_value("Rastrigin", "proxymix", "global_found"), 2)` of runs.
The comparison covers two functions of two
variables and one likelihood of five, one budget, one radius for basin
membership, and default settings for the competitors, and it does not cover
higher dimensions, noisy objectives or budgets between 5000 and 26 000
evaluations.

## Where next

*Choosing between the three fitting regimes* explains why an objective that
can be evaluated but not sampled lands in regime (iii), and what the other
two regimes do instead. *Fitting a proxy to a density you cannot sample* is
the shortest route into the same fitting verb on a target rather than an
objective. *The closed-form operator calculus on a mixture* covers the
marginalisation, conditioning and divergence operators that apply to the map
of solutions once it has been fitted, since the map is an ordinary mixture.

## References

* Azzalini, A. and Bowman, A. W. (1990). *A look at some data on the Old
  Faithful geyser.* Applied Statistics 39(3), 357--365.
  <doi:10.2307/2347385>.
* Carreira-Perpiñán, M. Á. (2000). *Mode-finding for mixtures of Gaussian
  distributions.* IEEE Transactions on Pattern Analysis and Machine
  Intelligence 22(11), 1318--1323. <doi:10.1109/34.888716>.
* Hansen, N. and Ostermeier, A. (2001). *Completely derandomized
  self-adaptation in evolution strategies.* Evolutionary Computation 9(2),
  159--195. <doi:10.1162/106365601750190398>.
* Himmelblau, D. M. (1972). *Applied Nonlinear Programming.* McGraw-Hill.
* Liu, D. C. and Nocedal, J. (1989). *On the limited memory BFGS method for
  large scale optimization.* Mathematical Programming 45, 503--528.
  <doi:10.1007/BF01589116>.
* Mullen, K. M., Ardia, D., Gil, D. L., Windover, D. and Cline, J. (2011).
  *DEoptim: An R package for global optimization by differential evolution.*
  Journal of Statistical Software 40(6), 1--26.
  <doi:10.18637/jss.v040.i06>.
* Nash, J. C. and Varadhan, R. (2011). *Unifying optimization algorithms to
  aid software system users: optimx for R.* Journal of Statistical Software
  43(9), 1--14. <doi:10.18637/jss.v043.i09>.
* Rastrigin, L. A. (1974). *Systems of Extremal Control.* Nauka.
* Scrucca, L., Fop, M., Murphy, T. B. and Raftery, A. E. (2016). *mclust 5:
  Clustering, classification and density estimation using Gaussian finite
  mixture models.* The R Journal 8(1), 289--317.
  <doi:10.32614/RJ-2016-021>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.
* Venables, W. N. and Ripley, B. D. (2002). *Modern Applied Statistics with
  S.* Fourth edition. Springer.
* Xiang, Y., Gubian, S., Suomela, B. and Hoeng, J. (2013). *Generalized
  simulated annealing for global optimization: the GenSA package.* The R
  Journal 5(1), 13--28. <doi:10.32614/RJ-2013-002>.

## Reproduce

The ambient seed is 20260619. Both fits also pass `seed = 1L` to
`from_objective()`, which fixes their importance draws.

The illustration draws the 400 descent starts after
`set.seed(20260925)` and runs every method after `set.seed(1L)`, with
`seed = 1L` passed to `from_objective()` and to GenSA. Each simulation run
is seeded by its own index. The stored results come from the code shown,
run on `r sim$run_date` with R
`r sim$versions[["R"]]`, proxymix `r desc_version("proxymix")`, GenSA
`r desc_version("GenSA")`, DEoptim `r desc_version("DEoptim")`, cmaes
`r desc_version("cmaes")`, nloptr `r desc_version("nloptr")` and optimx
`r desc_version("optimx")`, and raised `r length(sim$warnings)` warnings:
`r sum(grepl("Effective sample size", sim$warnings))` of low effective
sample size, `r length(decrease_msg)` of the EM objective falling
`r if (length(decrease_msg) > 0L) paste0("by at most ", sci_above(max(as.numeric(regmatches(decrease_msg, regexpr("(?<=decreased by )[0-9.e+-]+", decrease_msg, perl = TRUE))))), " ") else ""`in
a cooling step, and
`r length(sim$warnings) - sum(grepl("Effective sample size", sim$warnings)) - length(decrease_msg)`
others.

```{r session-info, eval = FALSE}
sessionInfo()
```

```{r session-info-print, echo = FALSE}
## printed 68 characters wide; a longer line breaks after its last slash or
## space within the width and continues indented
old_width <- options(width = 68L)
si_vec <- capture.output(sessionInfo())
options(old_width)
wrap_line <- function(s) {
  out <- character(0L)
  while (nchar(s) > 68L) {
    cut <- max(gregexpr("[/ ]", substr(s, 1L, 68L))[[1L]])
    if (cut < 20L) cut <- 68L
    out <- c(out, substr(s, 1L, cut))
    s <- paste0("    ", substr(s, cut + 1L, nchar(s)))
  }
  c(out, s)
}
writeLines(unlist(lapply(si_vec, wrap_line)))
```
