---
title: "How well a mixture proxies four awkward shapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How well a mixture proxies four awkward shapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render() with ggplot2 installed,
     the simulation read from its stored results; macOS arm64 (Apple silicon),
     R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The problem

Someone with a genuinely awkward target asks the question the routing table
does not answer: *my density is nothing like a Gaussian, and I can only
evaluate it. Where does a mixture proxy actually work, and where does it
stop?* The regime is settled before the question is asked. What is not
settled is whether the answer that comes back deserves to be used.

## Package capabilities

Four bundled targets stand in for four kinds of awkwardness.
`banana_target()` is a Rosenbrock-style bent ridge, curved but simply
connected. `donut_target()` is a rotationally symmetric annulus, whose
modes form a continuous ring rather than a finite set of points.
`mixture_target()` is three well-separated Gaussian components, which is
awkward only in being multimodal. `epanechnikov_target()` is compactly
supported: exactly zero outside a bounded interval, which no mixture of
full-support Gaussians can reproduce.

`fit_kld_em()` fits all four, through `fit_proxymix(regime = "kld")`, by
minimising the Kullback-Leibler divergence from the target with the
importance-sampled expectation-maximisation iteration of van der Hoek and
Elliott (2024). Its proposal argument is what the fit feeds on, and `proposal_mvt()` supplies
the Student-t default. A target that declares a `support` causes the fit to
select a support-matched `proposal_uniform()` instead, announced in a
message, because a Student-t would place draws where the target's
log-density is minus infinity.

Four diagnostics grade the result. `ess_summary()` and the fit's own
`ess` report how many of the importance draws are effectively carrying the
answer. The `validation_kld` diagnostic is an importance-sampled estimate of
\(\mathrm{KL}(f \Vert g_\theta)\) at the fitted parameters, taken on a fresh
sample the fit did not use, and `validation_mc_se` is its Monte Carlo
standard error. The `kld_final` diagnostic is the same estimate on the draws
the fit was tuned to, and it reads lower than the held-out value.
`hellinger_mc()` estimates the squared Hellinger distance \(H^2(f, g)\),
which lies in \([0, 1]\) when both densities are normalised. It draws fresh
points from the fit and returns the estimate with its Monte Carlo standard
error.

`gmm_fit_quality()` carries the refusal contract. When the importance
sample collapses, the fit does not return a mixture that looks ordinary
and is worthless; under `on_low_ess = "abort"` it raises a typed condition
instead.

## Addressing the problem

### Fitting the three plane-supported shapes

Each target gets a Student-t proposal wide enough to cover it, and a
component count chosen for its shape: four along the banana's ridge, six
around the donut's ring, three for the three-mode mixture.

```{r fits}
fit_b <- fit_proxymix(banana_target(), N = 4L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              sigma = 4 * diag(2),
                                              df = 5),
                      is_size = 3000L, max_iter = 50L, seed = 1L)
fit_d <- fit_proxymix(donut_target(), N = 6L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              sigma = 9 * diag(2),
                                              df = 5),
                      is_size = 3500L, max_iter = 60L, seed = 1L)
fit_m <- fit_proxymix(mixture_target(), N = 3L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              sigma = 6 * diag(2),
                                              df = 5),
                      is_size = 3000L, max_iter = 50L, seed = 1L)
```

```{r overlay-grid}
shape_panel <- function(target, fit, label, lim) {
  gx <- seq(-lim, lim, length.out = 110L)
  base <- expand.grid(x1 = gx, x2 = gx)
  gm <- as.matrix(base)
  data.frame(
    x1 = base$x1, x2 = base$x2,
    target = exp(target@log_density(gm)),
    proxy = dgmm(gm, fit),
    shape = label,
    stringsAsFactors = FALSE
  )
}
overlay_df <- rbind(
  shape_panel(banana_target(), fit_b, "banana, N = 4", 4.5),
  shape_panel(donut_target(), fit_d, "donut, N = 6", 4.5),
  shape_panel(mixture_target(), fit_m, "three-mixture, N = 3", 4.5)
)
overlay_df$shape <- factor(overlay_df$shape,
                           levels = unique(overlay_df$shape))
```

```{r overlay, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3, fig.pos = "!ht", fig.cap = "Each target (filled contours) with its fitted proxy overlaid as dashed contours. The banana and the three-mixture are tracked closely; the donut's proxy is a ring of six ellipses approximating a continuous annulus, and the dashed contours break into lobes where the ellipses meet.", fig.alt = "Three contour panels showing the banana, donut and three-mixture densities, each overlaid with dashed contours of its Gaussian-mixture proxy."}
ggplot2::ggplot(overlay_df, ggplot2::aes(x1, x2)) +
  ggplot2::geom_contour_filled(ggplot2::aes(z = target), bins = 10L,
                               alpha = 0.85) +
  ggplot2::geom_contour(ggplot2::aes(z = proxy), colour = "white",
                        linetype = "dashed", linewidth = 0.4, bins = 8L) +
  ggplot2::scale_fill_viridis_d(option = "mako", guide = "none") +
  ggplot2::facet_wrap(~ shape) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = expression(x[1]), y = expression(x[2])) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r overlay-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the three-shape overlay",
    "figure is skipped.\n")
```

### Grading the three fits

The held-out `validation_kld` is an importance-sampled estimate and carries
Monte Carlo error, so a second, deterministic measurement is worth having. In two
dimensions the divergence can be integrated directly on a grid, using
nothing the fit produced. The banana's two arms reach far upwards, so its
grid runs from -12 to 12 in each direction.

```{r kl-grid}
## KL(f || g) by deterministic grid quadrature -- independent of the
## importance sampler whose estimate it is checking.
kl_grid <- function(target, fit, lim, n_grid = 300L) {
  gx <- seq(-lim, lim, length.out = n_grid)
  gm <- as.matrix(expand.grid(x1 = gx, x2 = gx))
  cell <- (gx[2L] - gx[1L])^2
  log_f <- target@log_density(gm)
  log_g <- dgmm(gm, fit, log = TRUE)
  dens_f <- exp(log_f)
  ok <- is.finite(log_f) & is.finite(log_g) & dens_f > 1e-300
  sum(dens_f[ok] * (log_f[ok] - log_g[ok])) * cell
}
kl_quad <- c(
  banana = kl_grid(banana_target(), fit_b, 12, n_grid = 601L),
  donut = kl_grid(donut_target(), fit_d, 7),
  mixture = kl_grid(mixture_target(), fit_m, 8)
)
```

```{r kl-grid-mass, include = FALSE}
## the target's mass held by each quadrature grid
grid_mass <- function(target, lim, n_grid = 300L) {
  gx <- seq(-lim, lim, length.out = n_grid)
  gm <- as.matrix(expand.grid(x1 = gx, x2 = gx))
  sum(exp(target@log_density(gm))) * (gx[2L] - gx[1L])^2
}
mass_quad <- c(grid_mass(banana_target(), 12, n_grid = 601L), grid_mass(donut_target(), 7),
               grid_mass(mixture_target(), 8))
```

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.25\textheight\clearpage\fi
```

```{r summary-table, echo = FALSE}
fits <- list(fit_b, fit_d, fit_m)
hell <- lapply(fits, hellinger_mc, n_mc = 2000L, seed = 1L)
kl_val <- vapply(fits, function(f) f@diagnostics$validation_kld, numeric(1L))
kl_val_se <- vapply(fits, function(f) f@diagnostics$validation_mc_se,
                    numeric(1L))
kl_in <- vapply(fits, function(f) f@diagnostics$kld_final, numeric(1L))
n_iter <- vapply(fits, function(f) length(kld_trace(f)), integer(1L))
h2 <- vapply(hell, function(h) h$h2, numeric(1L))
h2_se <- vapply(hell, function(h) h$se, numeric(1L))
## held-out difference between two targets, in standard errors of the difference
kl_pair_z <- function(i, j) {
  abs(kl_val[i] - kl_val[j]) / sqrt(kl_val_se[i]^2 + kl_val_se[j]^2)
}
kl_z_donut <- min(kl_pair_z(2L, 1L), kl_pair_z(2L, 3L))
kl_z_other <- kl_pair_z(1L, 3L)
## the text below states both
stopifnot(kl_z_donut > 2, kl_z_other < 2)
summary_tbl <- data.frame(
  Target = c("banana", "donut", "mixture"),
  N = vapply(fits, gmm_n_components, integer(1L)),
  `IS draws` = vapply(fits, function(f) f@diagnostics$is_size, numeric(1L)),
  ESS = round(vapply(fits, function(f) f@diagnostics$ess, numeric(1L)), 0L),
  Iterations = n_iter,
  `KL (held-out)` = round(kl_val, 4L),
  SE = round(kl_val_se, 4L),
  `KL (quadrature)` = round(unname(kl_quad), 4L),
  # the padded header widens the column so the word fits inside the rule
  "\u00a0\u00a0Hellinger" = round(h2, 4L),
  `H SE` = round(h2_se, 4L),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(
  summary_tbl,
  caption = paste(
    "Fit quality on the three plane-supported shapes. The held-out and",
    "quadrature columns estimate the same Kullback-Leibler divergence,",
    "the first on draws the fit did not use and the second by grid",
    "quadrature. SE is the Monte Carlo standard error of the held-out",
    "value. Hellinger is the squared Hellinger distance, estimated on fresh",
    "draws from the fit, and H SE is its Monte Carlo standard error.",
    "Mixture is the three-component mixture target."
  )
)
```

### How the donut's residual changes with the component count

A finite Gaussian mixture cannot reproduce an annulus exactly. Sweeping the
component count shows how much of the residual divergence more components
remove at a fixed budget of 3,500 importance draws and 60 rounds.

```{r donut-sweep}
donut_n <- c(3L, 4L, 6L, 10L)
donut_kl <- vapply(donut_n, function(k) {
  f <- fit_proxymix(donut_target(), N = k, regime = "kld",
                    proposal = proposal_mvt(n_dim = 2L,
                                            sigma = 9 * diag(2),
                                            df = 5),
                    is_size = 3500L, max_iter = 60L, seed = 1L)
  kl_grid(donut_target(), f, 7)
}, numeric(1L))
```

The 10-component fit in this sweep stops at the 60-round cap. Refitting the
six- and ten-component proxies with 20,000 importance draws, all
other settings unchanged, tests whether the budget or the mixture sets the
residual.

```{r donut-budget}
donut_big <- lapply(c(6L, 10L), function(k) {
  fit_proxymix(donut_target(), N = k, regime = "kld",
               proposal = proposal_mvt(n_dim = 2L,
                                       sigma = 9 * diag(2),
                                       df = 5),
               is_size = 20000L, max_iter = 60L, seed = 1L)
})
donut_kl_big <- vapply(donut_big, function(f) kl_grid(donut_target(), f, 7),
                       numeric(1L))
donut_iter_big <- vapply(donut_big, function(f) length(kld_trace(f)),
                         integer(1L))
fit_d10 <- fit_proxymix(donut_target(), N = 10L, regime = "kld",
                        proposal = proposal_mvt(n_dim = 2L,
                                                sigma = 9 * diag(2),
                                                df = 5),
                        is_size = 3500L, max_iter = 60L, seed = 1L)
donut_iter_small <- length(kld_trace(fit_d10))
data.frame(
  Components = c(6L, 10L),
  `KL, 3,500 draws` = round(donut_kl[3:4], 4L),
  `KL, 20,000 draws` = round(donut_kl_big, 4L),
  `Rounds, 20,000 draws` = donut_iter_big,
  check.names = FALSE
)
```

```{r donut-sweep-table, echo = FALSE}
knitr::kable(
  data.frame(
    `Components` = donut_n,
    `KL (quadrature)` = round(donut_kl, 4L),
    check.names = FALSE
  ),
  caption = paste0(
    "Residual divergence of the donut proxy against the component count, ",
    "measured by grid quadrature. ",
    "From ", donut_n[3L], " to ", donut_n[4L],
    " components the divergence falls by only ",
    format(signif(donut_kl[3L] - donut_kl[4L], 2L)),
    " at 3,500 importance draws."
  )
)
```

### Bounded support

The three targets above live on the whole plane. A compactly supported
target is different in kind: the Epanechnikov density
\(K(u) = \tfrac{3}{4}(1 - u^2)\) on \([-1, 1]\) is exactly zero outside the
interval, and the default Student-t proposal is wasteful against it
because it places draws where the target's log-density is minus infinity.
Those draws receive zero weight and carry no information, and the fit
warns when too many of them fall outside the support.

A target may declare its `support`, and the fit then selects a
support-matched `proposal_uniform()` on its own, announced in a message.

```{r epan-target}
epan <- epanechnikov_target(n_dim = 1L)
epan
```

```{r epan-fit}
fit_e <- fit_proxymix(epan, N = 3L, regime = "kld",
                      is_size = 4000L, max_iter = 300L, seed = 1L)
c(proposal = fit_e@diagnostics$proposal_name,
  support_fraction = fit_e@diagnostics$support_fraction,
  iterations = length(kld_trace(fit_e)),
  converged = gmm_fit_quality(fit_e)$converged)
```

```{r epan-grid}
epan_x <- seq(-1.4, 1.4, length.out = 400L)
epan_df <- rbind(
  data.frame(x = epan_x,
             density = exp(epan@log_density(matrix(epan_x, ncol = 1L))),
             series = "target", stringsAsFactors = FALSE),
  data.frame(x = epan_x, density = dgmm(matrix(epan_x, ncol = 1L), fit_e),
             series = "mixture proxy", stringsAsFactors = FALSE)
)
```

```{r epan-facts, include = FALSE}
proxy_y <- dgmm(matrix(epan_x, ncol = 1L), fit_e)
target_y <- epan_df$density[epan_df$series == "target"]
num2 <- function(v) format(round(v, 2L), nsmall = 2L)
## the point of largest shortfall of the proxy below the target, on each
## side of zero inside the support
short_at <- function(side) {
  s1 <- which(abs(epan_x) < 1 & sign(epan_x) == side)
  s1[which.max(target_y[s1] - proxy_y[s1])]
}
i_side <- c(left = short_at(-1), right = short_at(1))
gap_side <- target_y[i_side] - proxy_y[i_side]
i_big <- i_side[which.max(gap_side)]
i_small <- i_side[which.min(gap_side)]
## the proxy's mass outside the support [-1, 1]
epan_sd <- sqrt(vapply(fit_e@covariances, `[[`, numeric(1L), 1L))
epan_mu <- vapply(fit_e@means, `[[`, numeric(1L), 1L)
epan_leak <- sum(fit_e@weights * (pnorm(-1, epan_mu, epan_sd) +
                                    pnorm(1, epan_mu, epan_sd,
                                          lower.tail = FALSE)))
epan_cap <- paste0(
  "The compactly supported Epanechnikov target and its three-component ",
  "proxy. The proxy follows the inverted parabola only in outline. It ",
  "peaks at ", num2(max(proxy_y)), " against the target's ",
  num2(max(target_y)), ". Its largest shortfall is ",
  names(i_big), " of centre, near $x = ", num2(epan_x[i_big]),
  "$, where it reads ", num2(proxy_y[i_big]), " against ",
  num2(target_y[i_big]), ". ", tools::toTitleCase(names(i_small)),
  " of centre it falls at most ", num2(min(gap_side)),
  " below the target, near $x = ", num2(epan_x[i_small]),
  "$. It leaks mass past the boundary at plus and minus one, as any ",
  "mixture of full-support Gaussians must."
)
```

```{r epan-plot, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.4, fig.cap = epan_cap, fig.alt = "Line plot of the inverted-parabola Epanechnikov density and its dashed Gaussian-mixture proxy, which extends slightly beyond the support boundary."}
ggplot2::ggplot(epan_df, ggplot2::aes(x, density, colour = series,
                                      linetype = series)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_vline(xintercept = c(-1, 1), colour = "grey60",
                      linewidth = 0.3) +
  ggplot2::scale_colour_manual(
    name = NULL, values = c("target" = "#0072B2",
                            "mixture proxy" = "#D55E00")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL, values = c("target" = "solid", "mixture proxy" = "dashed")
  ) +
  ggplot2::labs(x = expression(x), y = "density") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r epan-plot-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the bounded-support",
    "figure is skipped.\n")
```

\FloatBarrier

### When the proposal misses the target

Every number above rests on the importance sample being informative about
the target. When it is not, the arithmetic still produces a mixture, and
that mixture is worthless. The package treats this as a refusal rather
than a result: a proposal placed far from the target's mass, with
`on_low_ess = "abort"`, raises a typed condition instead of returning.

```{r refusal}
bad <- tryCatch(
  fit_kld_em(
    mixture_target(), N = 3L,
    proposal = proposal_mvn(n_dim = 2L, mean = c(25, 25), cov = diag(2)),
    is_size = 2000L, max_iter = 10L, seed = 1L,
    min_ess = 50, on_low_ess = "abort"
  ),
  error = function(e) e
)
class(bad)[1L]
writeLines(strwrap(conditionMessage(bad), width = 70L))
```

## Interpretation

The three plane-supported shapes divide into two groups, and the table says
which. The banana proxy reaches a quadrature divergence of
`r format(signif(kl_quad[["banana"]], 3L))` and the three-mixture
`r format(signif(kl_quad[["mixture"]], 2L))`, from effective sample sizes of
`r round(fit_b@diagnostics$ess)` and `r round(fit_m@diagnostics$ess)` draws
respectively. The donut sits materially higher at
`r format(signif(kl_quad[["donut"]], 2L))`, roughly
`r round(kl_quad[["donut"]] / kl_quad[["banana"]], 1L)` times the banana's.

At 3,500 draws, more components stop closing the gap. Going
from `r donut_n[1L]` to `r donut_n[3L]` components cuts the donut's
divergence from `r format(signif(donut_kl[1L], 3L))` to
`r format(signif(donut_kl[3L], 3L))`, a fall of
`r round(100 * (1 - donut_kl[3L] / donut_kl[1L]))` per cent. Going on to
`r donut_n[4L]` components buys only
`r format(signif(donut_kl[3L] - donut_kl[4L], 2L))` more. That flat step
comes from the fitting budget, not from the mixture. The
`r donut_n[4L]`-component fit used all `r donut_iter_small` of its 60
rounds. With 20,000 importance draws, the same
`r donut_n[4L]` components reach a divergence of
`r format(round(donut_kl_big[2L], 4L), nsmall = 4L)` against
`r format(round(donut_kl[4L], 4L), nsmall = 4L)`, and the fit stops after
`r donut_iter_big[2L]` rounds. Six components reach
`r format(round(donut_kl_big[1L], 4L), nsmall = 4L)` with the larger
sample. A ring needs more components than the other shapes. Placing more
components well needs more importance draws. A finite Gaussian mixture can
approximate any continuous density arbitrarily closely, and with the larger
sample the donut fit comes closer to the ring.

Each quadrature grid holds
`r format(floor(min(mass_quad) * 1e4) / 1e4, nsmall = 4L)` or more of its
target's mass. The two estimates of the same divergence agree within their Monte
Carlo error. The held-out column returns
`r format(signif(kl_val[2L], 2L))` for the donut with a standard error of
`r format(signif(kl_val_se[2L], 2L))`, against a quadrature value of
`r format(signif(kl_quad[["donut"]], 2L))`. Across the three shapes the
held-out and quadrature values differ by at most
`r format(ceiling(10 * max(abs(kl_val - kl_quad) / kl_val_se)) / 10, nsmall = 1L)`
standard errors. The standard errors are
`r round(100 * min(kl_val_se / kl_val))` to
`r round(100 * max(kl_val_se / kl_val))` per cent of the divergences. The
held-out estimate separates the donut from the other two targets by at
least `r format(round(kl_z_donut, 1L), nsmall = 1L)` standard errors of the
difference, but the banana and the mixture differ by only
`r format(round(kl_z_other, 1L), nsmall = 1L)`.

The compactly supported target is the fourth kind of awkwardness and the
one that never fully resolves. The declared support routed the fit onto a
support-matched proposal automatically, every draw landed inside the
support, and the fit converged after `r length(kld_trace(fit_e))`
iterations, against `r min(n_iter)` to `r max(n_iter)` for the
plane-supported shapes in the table. The proxy approximates a hard
boundary with densities that are positive everywhere. Its largest
shortfall in the figure is inside the support, near
$x = `r num2(epan_x[i_big])`$, and the mass it puts outside $[-1, 1]$ is
`r format(round(epan_leak, 4L), nsmall = 4L)`.

So the answer is that a mixture proxy works well on curvature and on
multimodality, degrades measurably on a ring that a handful of ellipses
can only approximate, and on a bounded support converges slowly to a fit that always
leaks a little mass past the boundary. And when the importance sample
carries no information at all, nothing is returned: the refusal above is a
`r paste0(intToUtf8(96L), class(bad)[1L], intToUtf8(96L))` condition, raised in place of a fit whose printed
summary would have looked entirely ordinary. The default is
`on_low_ess = "warn"`, which downgrades the same check to a warning and
lets the caller decide; `"abort"` is what turns it into a contract.

## Limitations

The two divergence estimates in the table are measurements, not exact
quantities, and neither is a certificate on its own. The held-out estimate
is a Monte Carlo quantity whose standard error on these fits is
`r round(100 * min(kl_val_se / kl_val))` to
`r round(100 * max(kl_val_se / kl_val))` per cent of the divergence. The in-sample `kld_final` is computed on the draws
the fit was tuned to and reads lower than the held-out value. Here it reads
`r format(round(kl_in[1L], 4L), nsmall = 4L)` for the banana,
`r format(round(kl_in[2L], 4L), nsmall = 4L)` for the donut and
`r format(round(kl_in[3L], 4L), nsmall = 4L)` for the three-mixture, below
zero for `r sum(kl_in < 0)` of the `r length(kl_in)` shapes, and a negative
estimate of a non-negative divergence means only that the estimator's error
exceeds the quantity being estimated. The squared Hellinger estimates are
drawn afresh from each fit and carry their own Monte Carlo error. The grid
quadrature avoids both problems and introduces a different one: it is reliable in two
or three dimensions and becomes intractable well before the dimension at
which importance sampling itself gives out.

The component counts here were chosen by hand, one per shape, to make the
comparison across shapes a comparison of shapes. Choosing the count is a
separate problem, and the donut sweep is the only place this vignette
touches it.

Nothing here establishes that the mixture proxy is the right tool for the
downstream job, only that it reproduces these densities to a measured
accuracy. Whether a divergence of the size found for the donut matters
depends entirely on what is computed from the proxy afterwards, and a
quantity that depends on the tails will be far less forgiving than one that
depends on the bulk. The numerical illustration that follows puts the same
four shapes to Markov chain Monte Carlo samplers, a Laplace approximation and
a mixture fitted to sampler draws, at one evaluation budget for all.

## Numerical illustration

```{r fixed-notation, include = FALSE}
## Small numbers are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
```

The four targets are evaluable densities with no data behind them, so there
is no empirical dataset to illustrate on, and this section is a simulation.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("density_shapes_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
## the grids are rebuilt from the stored chunk's own shape list, to report
## how much of each target's mass they hold
i1 <- grep("^shapes <- list\\(", sim$code)
i2 <- i1 + which(sim$code[-seq_len(i1)] == ")")[1L]
shapes_sim <- local({
  eval(parse(text = sim$code[i1:i2]))
  shapes
})
target_mass_sim <- vapply(shapes_sim, function(s) {
  gx <- seq(-s$lim + s$h / 2, s$lim - s$h / 2, by = s$h)
  gm <- as.matrix(expand.grid(rep(list(gx), s$target@n_dim)))
  sum(exp(s$target@log_density(gm))) * s$h^s$target@n_dim
}, numeric(1L))
captured_min <- min(sim$res$captured, na.rm = TRUE)
## a lower bound, so the value is rounded down
at_least <- function(v, digits) fixed(floor(v * 10^digits) / 10^digits, digits)
```

The illustration puts the four targets of this vignette to five methods
under one budget of `r sim$budget` target log-density evaluations per
method and replicate, and repeats every fit `r sim$n_rep` times with
different seeds. proxymix chooses its component count by `select_N()` over
one to eight components, and the budget is split evenly across the eight
candidate fits, each of which draws its importance sample and its validation
sample from the proposals used earlier in this vignette. The No-U-Turn
sampler of Stan (Carpenter et al., 2017), through cmdstanr, runs one
chain with 500 warmup iterations and stops at the iteration where its
leapfrog steps reach the budget; the warmup draws are discarded. A first
run finds that iteration, and a second run with the same seed, stopped
there, is the one timed and scored. Each
leapfrog step is counted as one evaluation, although it also computes the
gradient of the log-density, which none of the other methods uses. The
differential-evolution sampler DEzs (ter Braak and Vrugt, 2008) of
BayesianTools (Hartig et al., 2026) runs its three chains for the
budget of iterations, with a uniform prior on a box of half-width 10 around
the origin, or on the support itself for the bounded target, and the first
half of each chain is discarded. mclust (Scrucca et al., 2016) fits a
mixture of one to eight full-covariance components to the NUTS draws that
survived the cut, the two-stage route to a mixture when a sampler is at
hand. The Laplace approximation (Tierney and Kadane, 1986) is a Gaussian at
the mode found by BFGS from a random start, with the covariance from the
numerical Hessian. Integrated nested Laplace approximation is not
applicable, since none of the four targets is the posterior of a
latent-Gaussian model, and it is left out.

mclust and BayesianTools install from CRAN. cmdstanr is not on CRAN and
installs from the Stan R-universe repository, stan-dev.r-universe.dev. It
also needs a CmdStan installation, which `cmdstanr::install_cmdstan()`
provides:

```{r install-competitors, eval = FALSE}
install.packages(c("mclust", "BayesianTools"))
install.packages("cmdstanr",
                 repos = c("https://stan-dev.r-universe.dev",
                           getOption("repos")))
cmdstanr::install_cmdstan()
```

Two scores are taken against references that no method produced. The first
is $\mathrm{KL}(f \Vert g) = \int f \log (f / g)$, with $f$ the target
and $g$ the returned density, by the grid quadrature of the earlier section,
on a grid of cell centres 0.025 units to a side in the plane and 0.0025 on
the line, which is defined for the three methods that
return a density and not for the two samplers. The second is the error in
the mass of one event per shape, the tail $P(x_1 > 2)$ of the banana, the
hole $P(\lVert x \rVert < 1.5)$ of the donut, the tail $P(x_1 > 2.5)$ of
the three-mixture and the mass $P(|x| > 1)$ that the Epanechnikov target
puts outside its support, which is zero. The reference mass is the target's
own, by the same quadrature; a density method's mass is integrated on the
same grid and a sampler's is the share of its retained draws in the event.
Each grid holds at least `r at_least(min(target_mass_sim), 5)` of its target's
mass. On the line the grid covers
$|x| \leq `r shapes_sim$epanechnikov$lim`$, far past the support, so that the mass a density puts outside the support is
counted, and every density a method returned has at least
`r at_least(captured_min, 5)` of its mass on its grid.
Each method also reports the target evaluations it used and its elapsed
time. This chunk is complete and runs as shown, and compiles the four Stan
programs once, but it took about `r round(sim$cpu_secs / 60)` minutes of
processor time on one core (R `r sim$versions[["R"]]`, Apple
silicon); the results below are read from its stored output. A quick
reduced run needs only the line `n_rep <- 60L` changed to a smaller count.

```{r sim-code, eval = FALSE}
library(proxymix)
library(cmdstanr)
library(BayesianTools)
library(mclust)

n_rep <- 60L        # replicates per shape
budget <- 20000L    # target log-density evaluations per method and replicate

# Each shape: its target, the proposal for the mixture fit, the half-width
# and cell width of the quadrature grid, and the event whose mass is scored;
# every event boundary that is a straight line falls on a cell edge, and the
# grid on the line runs well past the support to count the mass that leaks.
shapes <- list(
  banana = list(
    target = banana_target(), lim = 10, h = 0.025,
    lower = c(-10, -10), upper = c(10, 10),
    event = function(x) x[, 1L] > 2, event_name = "P(x1 > 2)",
    proposal = proposal_mvt(2L, sigma = 4 * diag(2), df = 5)
  ),
  donut = list(
    target = donut_target(), lim = 7, h = 0.025,
    lower = c(-10, -10), upper = c(10, 10),
    event = function(x) rowSums(x^2) < 1.5^2, event_name = "P(|x| < 1.5)",
    proposal = proposal_mvt(2L, sigma = 9 * diag(2), df = 5)
  ),
  mixture = list(
    target = mixture_target(), lim = 8, h = 0.025,
    lower = c(-10, -10), upper = c(10, 10),
    event = function(x) x[, 1L] > 2.5, event_name = "P(x1 > 2.5)",
    proposal = proposal_mvt(2L, sigma = 6 * diag(2), df = 5)
  ),
  epanechnikov = list(
    target = epanechnikov_target(), lim = 6, h = 0.0025,
    lower = -1, upper = 1,
    event = function(x) abs(x[, 1L]) > 1, event_name = "P(|x| > 1)",
    proposal = NULL
  )
)

stan_code <- c(
  banana = "parameters { vector[2] x; }
model {
  real z2 = x[2] - 0.5 * (x[1]^2 - 1);
  target += -0.5 * (x[1]^2 + z2^2);
}",
  donut = "parameters { vector[2] x; }
model { target += -square(sqrt(dot_self(x)) - 2.5) / (2 * 0.25); }",
  mixture = "parameters { vector[2] x; }
model {
  target += log_sum_exp([
    log(0.3) + multi_normal_lpdf(x | [-2, -2]', [[0.6, 0], [0, 0.6]]),
    log(0.4) + multi_normal_lpdf(x | [0, 0]', [[0.5, 0.2], [0.2, 0.5]]),
    log(0.3) + multi_normal_lpdf(x | [2, 2]', [[0.4, -0.1], [-0.1, 0.4]])]');
}",
  epanechnikov = "parameters { vector<lower=-1, upper=1>[1] x; }
model { target += log1m(square(x[1])); }"
)
stan_dir <- "density_shapes_stan"
stan_models <- lapply(names(stan_code), function(s) {
  cmdstan_model(write_stan_file(stan_code[[s]], dir = stan_dir, basename = s),
                quiet = TRUE)
})
names(stan_models) <- names(stan_code)

# the grid of cell centres carries the target density, the cell area and
# the event indicator
make_grid <- function(shape) {
  p <- shape$target@n_dim
  gx <- seq(-shape$lim + shape$h / 2, shape$lim - shape$h / 2, by = shape$h)
  gm <- as.matrix(expand.grid(rep(list(gx), p)))
  colnames(gm) <- NULL
  log_f <- shape$target@log_density(gm)
  list(x = gm, log_f = log_f, f = exp(log_f), cell = shape$h^p,
       in_event = shape$event(gm))
}

kl_grid <- function(grid, log_g) {
  ok <- is.finite(grid$log_f) & grid$f > 1e-300
  sum(grid$f[ok] * (grid$log_f[ok] - log_g[ok])) * grid$cell
}
mass_grid <- function(grid, dens) sum(dens[grid$in_event]) * grid$cell

# a copy of the target whose log-density counts its own calls
counted_target <- function(shape) {
  env <- new.env()
  env$n <- 0L
  inner <- shape$target@log_density
  env$target <- gmm_target(
    n_dim = shape$target@n_dim,
    log_density = function(x) {
      if (is.null(dim(x))) x <- matrix(x, nrow = 1L)
      env$n <- env$n + nrow(x)
      inner(x)
    },
    normalised = TRUE, log_normalizer = 0,
    support = shape$target@support, name = shape$target@name
  )
  env
}

# Laplace approximation: a Gaussian at the mode found from a random start
# inside the box; a non-finite log-density is penalised so BFGS stays inside
laplace_fit <- function(target, shape, seed) {
  set.seed(seed)
  start <- runif(length(shape$lower), shape$lower, shape$upper) / 2
  neg_log_f <- function(v) {
    out <- -target@log_density(v)
    if (is.finite(out)) out else 1e10
  }
  opt <- optim(start, neg_log_f, method = "BFGS", hessian = TRUE)
  ev <- eigen(opt$hessian, symmetric = TRUE, only.values = TRUE)$values
  if (min(ev) <= 1e-6 * max(ev)) return(NULL)
  list(mean = opt$par, cov = solve(opt$hessian))
}
dmvn_log <- function(x, mean, cov) {
  r <- chol(cov)
  z <- forwardsolve(t(r), t(x) - mean)
  -0.5 * colSums(z^2) - sum(log(diag(r))) - 0.5 * ncol(x) * log(2 * pi)
}

one_replicate <- function(name, r) {
  shape <- shapes[[name]]
  p <- shape$target@n_dim
  grid <- grids[[name]]
  truth <- mass_grid(grid, grid$f)
  row <- function(method, n_comp, log_g, draws, evals, secs) {
    est <- if (is.null(draws)) mass_grid(grid, exp(log_g))
           else mean(shape$event(draws))
    data.frame(shape = name, rep = r, method = method, n_comp = n_comp,
               kl = if (is.null(log_g)) NA_real_ else kl_grid(grid, log_g),
               mass_error = est - truth, evals = evals, secs = secs,
               divergent = NA_integer_,
               captured = if (is.null(log_g)) NA_real_
                          else sum(exp(log_g)) * grid$cell)
  }

  # proxymix: component count by select_N, the budget split across candidates
  ct <- counted_target(shape)
  # each candidate: is_size fitting draws plus a quarter as many for validation
  is_size <- as.integer(budget / (8 * 1.25))
  t0 <- proc.time()[[3L]]
  sel <- suppressMessages(select_N(
    ct$target, candidates = 1:8, regime = "kld", proposal = shape$proposal,
    is_size = is_size, validation_size = is_size %/% 4L, max_iter = if (name == "epanechnikov") 300L else 60L,
    seed = r
  ))
  secs_px <- proc.time()[[3L]] - t0
  out_px <- row("proxymix", sel$best_n, dgmm(grid$x, sel$best_fit, log = TRUE),
                NULL, ct$n, secs_px)

  # NUTS: one chain, stopped where the leapfrog steps, one target
  # evaluation each, reach the budget; warmup draws discarded. A first run
  # finds that iteration; the second, with the same seed, is timed.
  nuts_run <- function(n_sampling) {
    stan_models[[name]]$sample(
      chains = 1L, iter_warmup = 500L, iter_sampling = n_sampling, seed = r,
      refresh = 0L, show_messages = FALSE, show_exceptions = FALSE,
      save_warmup = TRUE, diagnostics = NULL
    )
  }
  pilot <- nuts_run(8000L)$sampler_diagnostics(inc_warmup = TRUE,
                                                format = "df")
  n_stop <- sum(cumsum(pilot$n_leapfrog__) <= budget)
  t0 <- proc.time()[[3L]]
  fit <- nuts_run(max(n_stop - 500L, 1L))
  secs_nuts <- proc.time()[[3L]] - t0
  diag <- fit$sampler_diagnostics(inc_warmup = TRUE, format = "df")
  keep <- which(cumsum(diag$n_leapfrog__) <= budget)
  d <- fit$draws("x", inc_warmup = TRUE, format = "df")
  draws <- do.call(cbind, lapply(seq_len(p), function(j) {
    d[[paste0("x[", j, "]")]]
  }))
  draws_nuts <- draws[keep[keep > 500L], , drop = FALSE]
  evals_nuts <- sum(diag$n_leapfrog__[keep])
  out_nuts <- row("NUTS", NA_integer_, NULL, draws_nuts, evals_nuts, secs_nuts)
  out_nuts$divergent <- sum(diag$divergent__[keep[keep > 500L]])

  # mclust: a mixture fitted to the NUTS draws, full covariances
  t0 <- proc.time()[[3L]]
  mc <- Mclust(draws_nuts, G = 1:8, modelNames = if (p == 1L) "V" else "VVV",
               verbose = FALSE)
  secs_mc <- proc.time()[[3L]] - t0
  log_g_mc <- dens(grid$x, mc$modelName, parameters = mc$parameters,
                   logarithm = TRUE)
  out_mc <- row("mclust on NUTS", mc$G, log_g_mc, NULL, evals_nuts,
                secs_nuts + secs_mc)

  # DEzs: three chains sharing the budget, the first half of each discarded
  ct <- counted_target(shape)
  setup <- createBayesianSetup(
    likelihood = function(v) ct$target@log_density(v),
    lower = shape$lower, upper = shape$upper
  )
  t0 <- proc.time()[[3L]]
  set.seed(r)
  de <- runMCMC(setup, sampler = "DEzs",
                settings = list(iterations = budget, message = FALSE))
  secs_de <- proc.time()[[3L]] - t0
  draws_de <- getSample(de, start = budget %/% 6L)
  out_de <- row("DEzs", NA_integer_, NULL, draws_de, ct$n, secs_de)

  # Laplace
  ct <- counted_target(shape)
  t0 <- proc.time()[[3L]]
  lap <- laplace_fit(ct$target, shape, r)
  secs_lap <- proc.time()[[3L]] - t0
  out_lap <- if (is.null(lap)) {
    data.frame(shape = name, rep = r, method = "Laplace", n_comp = 1L,
               kl = NA_real_, mass_error = NA_real_, evals = ct$n,
               secs = secs_lap, divergent = NA_integer_,
               captured = NA_real_)
  } else {
    row("Laplace", 1L, dmvn_log(grid$x, lap$mean, lap$cov), NULL, ct$n,
        secs_lap)
  }

  rbind(out_px, out_nuts, out_mc, out_de, out_lap)
}

grids <- lapply(shapes, make_grid)
res <- do.call(rbind, lapply(names(shapes), function(name) {
  do.call(rbind, lapply(seq_len(n_rep), function(r) one_replicate(name, r)))
}))

sim_tab <- aggregate(
  cbind(kl = kl, abs_mass_error = abs(mass_error), evals = evals, secs = secs,
        n_comp = n_comp) ~ shape + method,
  data = res, FUN = mean, na.action = na.pass
)
sim_tab$failed <- aggregate(is.na(mass_error) ~ shape + method, data = res,
                            FUN = sum)[[3L]]
sim_tab$divergent <- aggregate(divergent ~ shape + method, data = res,
                               FUN = sum, na.action = na.pass)[[3L]]
sim_tab
```

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.6\textheight\clearpage\fi
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
shape_order <- c("banana", "donut", "mixture", "epanechnikov")
method_order <- c("proxymix", "NUTS", "mclust on NUTS", "DEzs", "Laplace")
tab_sim <- tab_sim[order(match(tab_sim$shape, shape_order),
                         match(tab_sim$method, method_order)), ]
tab_sim$shape <- c(banana = "banana", donut = "donut",
                   mixture = "mixture",
                   epanechnikov = "Epanechnikov")[tab_sim$shape]
tab_sim$n_comp <- ifelse(is.na(tab_sim$n_comp), "--",
                         format(round(tab_sim$n_comp, 1), nsmall = 1))
tab_sim$n_comp[tab_sim$failed == sim$n_rep] <- "--"
tab_sim$kl <- ifelse(is.na(tab_sim$kl), "--", fixed(tab_sim$kl, 4))
tab_sim$abs_mass_error <- ifelse(is.na(tab_sim$abs_mass_error), "--",
                                 fixed(tab_sim$abs_mass_error, 4))
tab_sim$divergent <- ifelse(is.na(tab_sim$divergent), "--",
                            format(tab_sim$divergent))
## non-breaking spaces keep "mclust on NUTS" on one line of the table
tab_sim$method <- gsub(" ", "\u00a0", tab_sim$method, fixed = TRUE)
knitr::kable(
  tab_sim[, c("shape", "method", "n_comp", "kl", "abs_mass_error", "evals",
              "secs", "failed", "divergent")],
  digits = c(0L, 0L, 0L, 0L, 0L, 0L, 2L, 0L, 0L), row.names = FALSE,
  align = c("l", "l", "r", "r", "r", "r", "r", "r", "r"),
  col.names = c("Shape", "Method", "N", "KL", "Mass error", "Evals",
                "Secs", "Failed", "Div."),
  caption = paste0(
    "Mean number of components, mean KL (the Kullback-Leibler divergence ",
    "$\\mathrm{KL}(f \\Vert g)$, with $f$ the target and $g$ the returned ",
    "density) by grid quadrature, mean absolute ",
    "error of the event mass against the target's mass by the same ",
    "quadrature, mean target evaluations used, mean elapsed seconds and ",
    "the number of replicates that returned no approximation, over ",
    sim$n_rep, " replicates per shape at a budget of ", sim$budget,
    " evaluations. Div. is the total number of divergent transitions over ",
    "the retained NUTS iterations of all replicates. A dash marks a cell ",
    "that does not apply: samplers return no density and so have no ",
    "component count or KL, only NUTS reports divergent ",
    "transitions, and a method that failed in every replicate has no ",
    "component count, KL or mass error. Mixture is the three-component mixture ",
    "target."
  )
)
sim_value <- function(shape, method, what) {
  s1 <- sim$sim_tab$shape == shape & sim$sim_tab$method == method
  sim$sim_tab[[what]][s1]
}
truth_of <- function(shape) sim$event_tab$truth[sim$event_tab$shape == shape]
res <- sim$res
sd_of <- function(shape, method, what) {
  s1 <- res$shape == shape & res$method == method
  sd(res[[what]][s1], na.rm = TRUE)
}
```

```{r sim-facts, echo = FALSE}
mass_sd <- function(shape, method) sd_of(shape, method, "mass_error")
lap_failed <- sim_value("donut", "Laplace", "failed")
nuts_div <- sum(sim$sim_tab$divergent, na.rm = TRUE)
n_comp_px <- vapply(c("banana", "donut", "mixture", "epanechnikov"),
                    sim_value, numeric(1L), method = "proxymix",
                    what = "n_comp")
rng <- function(method, what, digits) {
  v <- sim$sim_tab[[what]][sim$sim_tab$method == method]
  fixed(range(v, na.rm = TRUE), digits)
}
shape_label <- c(banana = "the banana", donut = "the donut",
                 mixture = "the three-mixture",
                 epanechnikov = "the Epanechnikov target")
join_and <- function(v) {
  if (length(v) <= 1L) return(v)
  paste(paste(v[-length(v)], collapse = ", "), "and", v[length(v)])
}
kl_px <- vapply(names(shape_label), sim_value, numeric(1L),
                method = "proxymix", what = "kl")
kl_mc <- vapply(names(shape_label), sim_value, numeric(1L),
                method = "mclust on NUTS", what = "kl")
## proxymix and the two-stage mixture share each replicate's seed, so the
## difference in KL is paired by replicate
kl_pair <- vapply(names(shape_label), function(s) {
  a <- res[res$shape == s & res$method == "proxymix", ]
  b <- res[res$shape == s & res$method == "mclust on NUTS", ]
  d <- a$kl[order(a$rep)] - b$kl[order(b$rep)]
  c(diff = mean(d), se = sd(d) / sqrt(length(d)))
}, numeric(2L))
resolved <- abs(kl_pair["diff", ]) > 2 * kl_pair["se", ]
px_lower <- names(shape_label)[resolved & kl_pair["diff", ] < 0]
mc_lower <- names(shape_label)[resolved & kl_pair["diff", ] > 0]
unresolved <- names(shape_label)[!resolved]
by_txt <- function(v) {
  join_and(paste0(fixed(abs(kl_pair["diff", v]), 4),
                  " (paired standard error ", fixed(kl_pair["se", v], 4), ")"))
}
z_txt <- function(v) {
  join_and(fixed(abs(kl_pair["diff", v]) / kl_pair["se", v], 1))
}
lap_kl <- vapply(names(shape_label), sim_value, numeric(1L),
                 method = "Laplace", what = "kl")
lap_def <- names(shape_label)[!is.na(lap_kl)]
lap_txt <- join_and(paste(fixed(lap_kl[lap_def], 2), "on",
                          shape_label[lap_def]))
pct_higher <- round(100 * (kl_px[mc_lower] / kl_mc[mc_lower] - 1))
budgeted <- c("proxymix", "NUTS", "mclust on NUTS", "DEzs")
method_word <- c(proxymix = "proxymix", NUTS = "NUTS",
                 "mclust on NUTS" = "the two-stage route", DEzs = "DEzs")
## per shape, the two budgeted methods with the lowest mean time, and
## whether their paired difference in seconds exceeds two standard errors
speed <- lapply(names(shape_label), function(s) {
  secs <- vapply(budgeted, sim_value, numeric(1L), shape = s, what = "secs")
  two <- budgeted[order(secs)][1:2]
  a <- res[res$shape == s & res$method == two[1L], ]
  b <- res[res$shape == s & res$method == two[2L], ]
  d <- b$secs[order(b$rep)] - a$secs[order(a$rep)]
  list(first = two[1L], second = two[2L],
       resolved = mean(d) > 2 * sd(d) / sqrt(length(d)))
})
names(speed) <- names(shape_label)
fastest <- vapply(speed, function(v) v$first, character(1L))
fast_resolved <- vapply(speed, function(v) v$resolved, logical(1L))
fast_parts <- vapply(unique(fastest[fast_resolved]), function(m) {
  paste(method_word[[m]], "on",
        join_and(shape_label[names(fastest)[fast_resolved & fastest == m]]))
}, character(1L))
fast_ties <- vapply(names(fastest)[!fast_resolved], function(s) {
  paste0(method_word[[speed[[s]]$first]], " and ",
         method_word[[speed[[s]]$second]], " on ", shape_label[[s]])
}, character(1L))
## parts that contain "and" themselves are joined with a comma before the last
join_parts <- function(v) {
  if (length(v) <= 1L) return(v)
  paste0(paste(v[-length(v)], collapse = ", "), ", and ", v[length(v)])
}
fast_txt <- paste0(
  if (length(fast_parts) > 0L) {
    paste0("By mean elapsed time the fastest budgeted method was ",
           join_parts(fast_parts), ", ahead of the next by more than two ",
           "paired standard errors. ")
  },
  if (length(fast_ties) > 0L) {
    paste0("The two fastest budgeted methods, ", join_parts(fast_ties),
           ", were within two paired standard errors of each other. ")
  },
  if (any(grepl("NUTS", c(fastest, vapply(speed, function(v) v$second,
                                             character(1L)))))) {
    paste0("The NUTS times leave out the untimed first run, of 500 warmup ",
           "and 8000 sampling iterations, that finds where to stop. ")
  }
)
lap_banana <- sim_value("banana", "Laplace", "abs_mass_error")
plane_err <- sim$sim_tab$abs_mass_error[
  sim$sim_tab$method %in% budgeted & sim$sim_tab$shape != "epanechnikov"]
div_shapes <- sim$sim_tab$shape[sim$sim_tab$method == "NUTS" &
                                  sim$sim_tab$divergent > 0]
```

On the divergence, neither route to a mixture is lower at every shape. proxymix
reaches `r fixed(sim_value("banana", "proxymix", "kl"), 4)` on the banana,
`r fixed(sim_value("mixture", "proxymix", "kl"), 4)` on the three-mixture
and `r fixed(sim_value("epanechnikov", "proxymix", "kl"), 4)` on the
Epanechnikov target, against
`r fixed(sim_value("banana", "mclust on NUTS", "kl"), 4)`,
`r fixed(sim_value("mixture", "mclust on NUTS", "kl"), 4)` and
`r fixed(sim_value("epanechnikov", "mclust on NUTS", "kl"), 4)` for the
mixture mclust fits to the NUTS draws. On the donut the two are
`r fixed(sim_value("donut", "proxymix", "kl"), 4)` and
`r fixed(sim_value("donut", "mclust on NUTS", "kl"), 4)`, both above the
other shapes, in line with the component sweep earlier in this vignette,
with `r fixed(n_comp_px[["donut"]], 1)` components chosen on
average by `select_N()`. The Laplace approximation sits far above both
wherever it is defined, at `r lap_txt`. On the three-mixture its one
Gaussian covers one mode. On the donut the Hessian at a point on
the ring is singular along the ring, so the approximation is undefined and
is recorded as failed in `r lap_failed` of `r sim$n_rep` replicates.

On the event mass of the three plane-supported shapes the samplers and the
mixtures have mean absolute errors from `r fixed(min(plane_err), 4)` to
`r fixed(max(plane_err), 4)`. NUTS is out by
`r fixed(sim_value("banana", "NUTS", "abs_mass_error"), 4)` on the banana
tail and `r fixed(sim_value("donut", "NUTS", "abs_mass_error"), 4)` on the
donut hole, DEzs by
`r fixed(sim_value("banana", "DEzs", "abs_mass_error"), 4)` and
`r fixed(sim_value("donut", "DEzs", "abs_mass_error"), 4)`, and proxymix by
`r fixed(sim_value("banana", "proxymix", "abs_mass_error"), 4)` and
`r fixed(sim_value("donut", "proxymix", "abs_mass_error"), 4)`, errors of
the size of the Monte Carlo noise in a few thousand draws. The Laplace
Gaussian gets the banana tail almost exactly, with a mean absolute error of
`r format(signif(lap_banana, 1L), scientific = FALSE)`, because
the first coordinate of the banana is exactly standard normal and the
Gaussian at the mode has that marginal. The bounded
target separates the methods by kind. Both samplers put no draw outside
$[-1, 1]$, and their error is exactly zero, while the mixtures leak
`r fixed(sim_value("epanechnikov", "proxymix", "abs_mass_error"), 4)`
(proxymix) and
`r fixed(sim_value("epanechnikov", "mclust on NUTS", "abs_mass_error"), 4)`
(mclust) of their mass past the boundary and the Laplace Gaussian
`r fixed(sim_value("epanechnikov", "Laplace", "abs_mass_error"), 3)`. The
Laplace approximation also misses the three-mixture tail by
`r fixed(sim_value("mixture", "Laplace", "abs_mass_error"), 3)`.

Every budgeted method used close to its `r sim$budget` evaluations, apart
from DEzs on the bounded target, which evaluated the density
`r fixed(sim_value("epanechnikov", "DEzs", "evals"), 0)` times on average
because BayesianTools `r sim$versions[["BayesianTools"]]` rejects a
proposal outside the prior box before evaluating it. The Laplace
approximation needed `r rng("Laplace", "evals", 0)[1L]` to
`r rng("Laplace", "evals", 0)[2L]` evaluations. In elapsed time per
replicate proxymix took `r rng("proxymix", "secs", 2)[1L]` to
`r rng("proxymix", "secs", 2)[2L]` seconds across the shapes, NUTS
`r rng("NUTS", "secs", 2)[1L]` to `r rng("NUTS", "secs", 2)[2L]`, the
mclust stage on top of NUTS brought the two-stage route to
`r rng("mclust on NUTS", "secs", 2)[1L]` to
`r rng("mclust on NUTS", "secs", 2)[2L]`, and DEzs took
`r rng("DEzs", "secs", 2)[1L]` to `r rng("DEzs", "secs", 2)[2L]`. The Div.
column of the table counts `r nuts_div` divergent transitions across the
retained NUTS iterations, all of them on
`r join_and(shape_label[div_shapes])`.

In this run proxymix reached a lower KL than the two-stage mixture on
`r join_and(shape_label[px_lower])`, by `r by_txt(px_lower)`, and had
event-mass errors of the same order as the samplers on the three
plane-supported shapes. The two-stage mixture reached a lower KL on
`r join_and(shape_label[mc_lower])`,
where the proxymix KL was higher by `r join_and(pct_higher)` per cent
respectively, differences of `r by_txt(mc_lower)`. On
`r join_and(shape_label[unresolved])` the paired difference is
`r z_txt(unresolved)` standard errors, which does not resolve which of the
two is lower at `r sim$n_rep` replicates. `r fast_txt`The samplers
reproduced the bounded support exactly, where every mixture leaked mass
past the boundary, and their estimate of any event's mass is consistent as
the chain lengthens, while a
mixture of at most eight components carries a residual divergence on the
donut at this budget. The comparison covers four shapes in one and
two dimensions at one budget, with component counts capped at eight and
hand-set proposals for the mixture fit, and says nothing about higher
dimensions, other budgets or targets that must be sampled rather than
evaluated.

## Further reading

*Fitting a proxy to a density you cannot sample* is the shorter introduction
to the regime this vignette stress-tests, and reads the fit certificate in
full.

*Choosing between the three fitting regimes* compares regime (iii) with
regimes (i) and (ii), which fit to draws from the target. Every fit here
uses the log-density alone, although each of the four targets can also
supply draws through `with_samples = TRUE`.

*Reading the entropy of a fitted mixture* takes up the component-count
question the donut sweep only touches, and discovers a count by cooling
rather than by sweeping.

*The closed-form operator calculus on a mixture* is what the proxies fitted
here are for: the exact operations that follow once the shape has been
captured.

## References

* Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
  Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A. (2017).
  *Stan: A probabilistic programming language.* Journal of Statistical
  Software 76(1), 1--32. <doi:10.18637/jss.v076.i01>.
* Hartig, F., Minunno, F. and Paul, S. (2026). *BayesianTools:
  General-purpose MCMC and SMC samplers and tools for Bayesian statistics.*
  R package version `r sim$versions[["BayesianTools"]]`.
  <doi:10.32614/CRAN.package.BayesianTools>.
* Scrucca, L., Fop, M., Murphy, T. B. and Raftery, A. E. (2016). *mclust 5:
  Clustering, classification and density estimation using Gaussian finite
  mixture models.* The R Journal 8(1), 289--317.
  <doi:10.32614/RJ-2016-021>.
* ter Braak, C. J. F. and Vrugt, J. A. (2008). *Differential evolution
  Markov chain with snooker updater and fewer chains.* Statistics and
  Computing 18(4), 435--446. <doi:10.1007/s11222-008-9104-9>.
* Tierney, L. and Kadane, J. B. (1986). *Accurate approximations for
  posterior moments and marginal densities.* Journal of the American
  Statistical Association 81(393), 82--86.
  <doi:10.1080/01621459.1986.10478240>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.

## Reproduce

Every fit and Hellinger estimate above is seeded with `seed = 1L`, and the
grid quadrature is deterministic, so
the sections before the numerical illustration are reproducible from the
package alone.

In the simulation each replicate passes its own index as the seed to
`select_N()`, to the Stan sampler and to `set.seed()` before the
differential-evolution run and the Laplace start point. The stored results
come from the code shown, run on `r sim$run_date` with R
`r sim$versions[["R"]]`, proxymix `r sim$versions[["proxymix"]]`, cmdstanr
`r sim$versions[["cmdstanr"]]` on CmdStan `r sim$versions[["CmdStan"]]`,
BayesianTools `r sim$versions[["BayesianTools"]]` and mclust
`r sim$versions[["mclust"]]`, and the run raised `r length(sim$warnings)`
warnings.

```{r session-info}
old_opt <- options(width = 72L)
si_vec <- trimws(capture.output(sessionInfo()), "right")
options(old_opt)
# library paths are single long words, so a long line breaks after a slash
long_vec <- nchar(si_vec) > 72L
si_vec[long_vec] <- gsub("(.{1,68}/)(?=.{8,})", "\\1\n        ",
                         si_vec[long_vec], perl = TRUE)
writeLines(si_vec)
```
