---
title: "Testing the last observation for instability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing the last observation for instability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~30 s under rmarkdown::render() with ggplot2, strucchange,
     changepoint, KFAS and dlm installed, the simulation read from its stored
     results; macOS arm64 (Apple silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r fixed-notation, include = FALSE}
## Small numbers are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
```

## The problem

An analyst monitoring a series has a model fitted to everything up to
yesterday, and this morning one more observation arrived. The question is
narrow and awkward: *one more observation has just arrived -- is it
consistent with the model I fitted, or has something just changed?* It is
awkward because the suspect window is tiny. The breakpoint test of Chow
(1960) and the sup-Wald test of Andrews (1993) need observations on both
sides of the break, and the parameters of the post-break regime cannot be
estimated from a single new point or a handful. Chow's predictive test is
built for this case and assumes normal errors (Andrews, 2003).
`gmm_eos_test()` answers the question in this regime, following Andrews
(2003).

## Package capabilities

`gmm_eos_test()` runs the filter implied by the operator calculus over the
series, which for the single-component prior and Gaussian noise it accepts
is the Kalman filter, scores the last `m` one-step filter innovations,
and calibrates that score. It takes the model exactly as
`gmm_filter()` does: a single-component `gmm()` prior, a `dynamics`
list carrying the state-transition and process-noise matrices, and a
`measurement` list carrying the observation and observation-noise matrices.

The statistic is built from the filter's own output. At each step the filter
returns a one-step predictive mean and an innovation covariance $S_t$, so
the standardised innovation
$z_t = S_t^{-1/2}(y_t - C\hat{x}_{t\mid t-1})$ is standard normal under a
correct, stable, linear-Gaussian model. The statistic sums the last `m`
squared standardised innovations,
$$ \mathrm{EoS}_m = \sum_{t = n-m+1}^{n} z_t^{\top} z_t, $$
which a break inflates.

Two calibrations are offered. `method = "chisq"` refers the statistic to a
$\chi^2$ distribution on `m` times the observation dimension degrees of
freedom, exact when the innovations are Gaussian and the model is given.
`method = "andrews"` is a subsampling calibration after the P-test of
Andrews (2003): the statistic's rank among the in-sample overlapping
`m`-blocks of the same innovations gives the p-value. It differs from
Andrews' test in two ways. Its p-value counts the tested block itself. Its
block statistics are computed at the supplied model, not at estimates that
leave out each block. It does not assume Gaussian innovations, and
its validity is asymptotic, as the series grows with `m` fixed, under
stationarity and ergodicity; at a finite length its size can differ from
nominal, and in a short series it can over-reject. The returned object also carries those
in-sample block statistics, so the calibration can be inspected directly.

## Addressing the problem

```{r seed}
set.seed(20260621)
```

### The model and a stable series

A local-level model: a random walk observed with noise.

```{r model}
prior <- gmm(weights = 1, means = list(0), covariances = list(matrix(10)))
dynamics <- list(A = matrix(1), Q = matrix(0.04))  # the state random walk
measurement <- list(C = matrix(1), R = matrix(1))  # the noisy observation

n <- 120L
level <- cumsum(c(0, rnorm(n - 1L, 0, sqrt(0.04))))
y_stable <- level + rnorm(n, 0, 1)
```

### The same series with a broken final observation

Copy the series and push the final observation away by five standard
deviations of the observation noise.

```{r break}
y_break <- y_stable
y_break[n] <- y_break[n] + 5
```

```{r tests}
test_stable <- gmm_eos_test(
  prior, dynamics, measurement, y_stable, m = 1L, method = "andrews"
)
test_break <- gmm_eos_test(
  prior, dynamics, measurement, y_break, m = 1L, method = "andrews"
)
```

The returned object is a list whose `in_sample_blocks` element holds one
number per in-sample block.

```{r tests-kable, echo = FALSE}
eos_row <- function(x) {
  c(
    round(x$statistic, 3L), round(x$p_value, 4L), x$reject,
    x$method, x$m, if (x$method == "chisq") x$df else "--"
  )
}
knitr::kable(
  data.frame(
    field = c("statistic", "p-value", "reject at 0.05", "calibration",
              "window m", "chi-square degrees of freedom"),
    stable = eos_row(test_stable),
    broken = eos_row(test_break)
  ),
  col.names = c("Field", "Stable series", "Broken final point"),
  caption = paste(
    "The end-of-sample test on the same series before and after the",
    "final observation is pushed five observation-noise standard",
    "deviations away. The subsampling calibration uses no degrees of",
    "freedom, so that row holds a dash."
  )
)
```

### What the filter expected

The filter's own path shows why the broken point is surprising. The band
below is the one-step predictive band: the filter's forecast of each
observation from the ones before it, with two predictive standard
deviations either side, which is the scale the statistic divides by. It is
built from `gmm_filter()` on the stable series, with the broken final
observation marked against it.

```{r filter-path}
filtered <- gmm_filter(
  prior, dynamics, measurement, y_stable, ridge_eps = 0
)
path <- filtered$summary
# one-step predictive mean and sd of y_t, from the filtered state at t - 1
pred_mean <- c(prior@means[[1L]], path$mean_1[-n])
pred_sd <- sqrt(c(prior@covariances[[1L]][1L, 1L], path$sd_1[-n]^2) +
                  dynamics$Q[1L, 1L] + measurement$R[1L, 1L])
outside <- sum(abs(y_stable - pred_mean)[-1L] > 2 * pred_sd[-1L])
```

```{r fig-series, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4, fig.cap = "The local-level series, the latent state that generated it, the filtered state, and the one-step predictive band, two predictive standard deviations either side of the filter's forecast of each observation. The band starts at the second step, after the diffuse prior. The broken final observation sits far outside the band, which is what inflates the statistic.", fig.alt = "A time series of noisy observations in grey with the latent state and the filtered state overlaid and a shaded predictive band around the forecasts, and a single isolated point at the right-hand end sitting well above the band."}
series_df <- data.frame(
  t = seq_len(n),
  observed = y_stable,
  truth = level,
  filtered = path$mean_1,
  lo = pred_mean - 2 * pred_sd,
  hi = pred_mean + 2 * pred_sd
)
break_df <- data.frame(t = n, y = y_break[n])
ggplot2::ggplot(series_df, ggplot2::aes(t)) +
  ggplot2::geom_ribbon(
    data = series_df[-1L, ], ggplot2::aes(ymin = lo, ymax = hi),
    fill = "#0072B2", alpha = 0.18
  ) +
  ggplot2::geom_point(
    ggplot2::aes(y = observed, colour = "observed (stable)"),
    size = 1.1, alpha = 0.7
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = truth, colour = "latent state"), linewidth = 0.8
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered, colour = "filtered state"), linewidth = 0.8
  ) +
  ggplot2::geom_point(
    data = break_df, ggplot2::aes(t, y, colour = "broken final point"),
    size = 2.6
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c(
      "observed (stable)" = "grey60",
      "latent state" = "#009E73",
      "filtered state" = "#0072B2",
      "broken final point" = "#D55E00"
    )
  ) +
  ggplot2::labs(
    x = "time step", y = "observation",
    title = "An end-of-sample break in the last observation"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-series-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the filtered-series figure",
    "is skipped.\n")
```

```{=latex}
\FloatBarrier
```

### How the subsampling p-value is reached

The subsampling calibration compares the end-of-sample statistic with the
in-sample blocks of the same series, so the p-value can be read straight off
the empirical distribution of those blocks.

```{r fig-blocks, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4, fig.cap = "The empirical distribution of the in-sample block statistics for each series, with the end-of-sample statistic marked. The subsampling p-value is $(1 + k) / (n - 2m + 2)$, where $k$ is the number of blocks lying at or beyond the dashed line, so the stable statistic sits inside the bulk while the broken one clears every block in the series.", fig.alt = "Two panels, each showing a rising step function of the cumulative share of in-sample block statistics against a logarithmic axis, annotated with a vertical rule; in the left panel the rule falls inside the curve, in the right panel it lies far beyond its right-hand end."}
blocks_df <- rbind(
  data.frame(
    block = test_stable$in_sample_blocks, series = "stable series"
  ),
  data.frame(
    block = test_break$in_sample_blocks, series = "broken final point"
  )
)
blocks_df$series <- factor(
  blocks_df$series, levels = c("stable series", "broken final point")
)
rule_df <- data.frame(
  series = factor(
    c("stable series", "broken final point"),
    levels = c("stable series", "broken final point")
  ),
  statistic = c(test_stable$statistic, test_break$statistic)
)
ggplot2::ggplot(blocks_df, ggplot2::aes(block)) +
  ggplot2::stat_ecdf(geom = "step", linewidth = 0.8, colour = "#0072B2") +
  ggplot2::geom_vline(
    data = rule_df, ggplot2::aes(xintercept = statistic),
    colour = "#D55E00", linetype = "dashed", linewidth = 0.8
  ) +
  ggplot2::facet_wrap(~ series) +
  ggplot2::scale_x_log10(labels = function(v) {
    format(v, scientific = FALSE, drop0trailing = TRUE, trim = TRUE)
  }) +
  ggplot2::labs(
    x = "in-sample block statistic (log scale)",
    y = "cumulative share of blocks",
    title = "The end-of-sample statistic against its own in-sample blocks"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-blocks-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the block-calibration",
    "figure is skipped.\n")
```

```{=latex}
\FloatBarrier
```

### What size the two calibrations actually hold

```{r prior-weight, include = FALSE}
## weight of the prior mean in the filter's forecast of observation 30:
## the product of one minus the Kalman gain over the first 29 updates
p_pred <- prior@covariances[[1L]][1L, 1L]
prior_weight_30 <- 1
for (i1 in seq_len(29L)) {
  gain <- p_pred / (p_pred + measurement$R[1L, 1L])
  prior_weight_30 <- prior_weight_30 * (1 - gain)
  p_pred <- (1 - gain) * p_pred + dynamics$Q[1L, 1L]
} # ends i1, over the first 29 filter updates
```

The chi-square calibration is exact when the model is given. The diffuse
prior enters the forecast of the last observation with a weight that
shrinks at every step, to about `r fixed(prior_weight_30, 4)` at $n = 30$, so
the chi-square reference is essentially exact here too. The size is still
worth measuring rather than assuming. The study below simulates from the same
local-level model under the null and records the rejection rate at a nominal
0.05, with the Monte Carlo standard error the replicate count supports.

```{r size-study}
simulate_null <- function(n_obs) {
  lv <- cumsum(c(0, rnorm(n_obs - 1L, 0, sqrt(0.04))))
  lv + rnorm(n_obs, 0, 1)
}
empirical_size <- function(n_obs, method, n_rep) {
  p <- vapply(seq_len(n_rep), function(i1) {
    gmm_eos_test(
      prior, dynamics, measurement, simulate_null(n_obs),
      m = 1L, method = method
    )$p_value
  }, numeric(1L))
  size <- mean(p < 0.05)
  c(size = size, se = sqrt(size * (1 - size) / n_rep), n_rep = n_rep)
} # ends empirical_size, one row of the size study

n_rep_short <- 250L
n_rep_long <- 150L
size_chisq_30 <- empirical_size(30L, "chisq", n_rep_short)
size_chisq_120 <- empirical_size(120L, "chisq", n_rep_long)
size_andrews_30 <- empirical_size(30L, "andrews", n_rep_short)

chisq_sizes <- c(size_chisq_30[["size"]], size_chisq_120[["size"]])
chisq_ses <- c(size_chisq_30[["se"]], size_chisq_120[["se"]])
chisq_worst_se <- max(abs(chisq_sizes - 0.05) / chisq_ses)
```

The subsampling p-value takes only the values $(1 + k) / (n - 2m + 2)$ for
$k = 0, 1, \ldots$. If the end block's rank among the blocks is uniform,
the size at a nominal level is the share of that grid below the level, and
it can be computed without simulation. The rank is uniform when the blocks
are exchangeable, which holds at $m = 1$ with the model given, apart from
the first few innovations under the diffuse prior. It does not hold with
estimated parameters, or for the overlapping blocks of $m > 1$.

```{r attainable}
attainable_size <- function(n_obs, m = 1L, alpha = 0.05) {
  n_block <- n_obs - 2L * m + 1L
  p_grid <- (1L + seq.int(0L, n_block)) / (1L + n_block)
  mean(p_grid < alpha)
} # ends attainable_size, the size when the end block's rank is uniform

floor_30 <- 1 / (30L - 2L + 2L)
floor_120 <- 1 / (120L - 2L + 2L)
attainable_30 <- attainable_size(30L)
attainable_120 <- attainable_size(120L)
diff_30 <- size_chisq_30[["size"]] - size_andrews_30[["size"]]
diff_30_se <- sqrt(size_chisq_30[["se"]]^2 + size_andrews_30[["se"]]^2)
andrews_ses_from_grid <- abs(size_andrews_30[["size"]] - attainable_30) /
  size_andrews_30[["se"]]
```

```{=latex}
\par\penalty-100\begingroup\dimen0=16\baselineskip\dimen2=\pagegoal
\advance\dimen2 by -\pagetotal
\ifdim\dimen0>\dimen2 \ifdim\dimen2>0pt \vfil\fi\break\fi\endgroup
```

```{r size-kable, echo = FALSE}
old_opts <- options(knitr.kable.NA = "--")
knitr::kable(
  data.frame(
    calibration = c(
      "chi-square", "chi-square", "subsampling", "subsampling",
      "subsampling"
    ),
    series_length = c("30", "120", "30", "30", "120"),
    quantity = c(
      "simulated size", "simulated size", "simulated size",
      "grid size, uniform rank", "grid size, uniform rank"
    ),
    value = round(
      c(
        size_chisq_30[["size"]], size_chisq_120[["size"]],
        size_andrews_30[["size"]], attainable_30, attainable_120
      ), 4L
    ),
    std_error = c(
      round(size_chisq_30[["se"]], 4L), round(size_chisq_120[["se"]], 4L),
      round(size_andrews_30[["se"]], 4L), NA_real_, NA_real_
    ),
    replicates = c(
      n_rep_short, n_rep_long, n_rep_short, NA_integer_, NA_integer_
    )
  ),
  col.names = c("Calibration", "n", "Quantity", "Value", "Std. error",
                "Replicates"),
  caption = paste(
    "Rejection rate against a nominal 0.05 under the null, simulated for",
    "both calibrations, and for the subsampling calibration also computed",
    "from its p-value grid, which gives its size when the end block's rank",
    "is uniform. The grid values are not simulated, so their standard",
    "error and replicate cells hold a dash."
  )
)
options(old_opts)
```

## Interpretation

On the stable series the statistic is
`r format(round(test_stable$statistic, 3L), nsmall = 3L)` with a subsampling p-value of
`r round(test_stable$p_value, 3L)`, and the test does not reject: the last
observation sits where the filter expects it. Pushing that same observation
five observation-noise standard deviations away lifts the statistic to
`r round(test_break$statistic, 1L)`, a p-value of
`r round(test_break$p_value, 4L)`, and a rejection. That is the answer to
the opening question in the form the analyst needs it: on this series, the
arriving observation is flagged as inconsistent with the fitted model, and
the same machinery leaves the unbroken series alone.

The first figure shows why. The broken point sits
`r fixed(sqrt(test_break$statistic), 1)` predictive standard deviations from
the filter's forecast for $t = `r n`$, the standardised innovation whose
square is the statistic, far outside the band. Of the `r n - 1L` stable
observations from the second step on, `r outside` fall outside the band, or
`r fixed(100 * outside / (n - 1L), 1)` per cent, against the
`r fixed(100 * 2 * pnorm(-2), 1)` per cent a band of two standard
deviations leaves out under the model. The second figure shows how that surprise becomes a p-value: for
the stable series the statistic falls inside the bulk of the in-sample
blocks, so a substantial share of blocks lie beyond it, while for the broken
series it clears every one of the
`r length(test_break$in_sample_blocks)` blocks, which drives the p-value to
its floor.

That floor is the first thing the size table makes concrete. At $n = 120$
and $m = 1$ the smallest attainable p-value is
`r round(floor_120, 4L)`, and the broken series attains it exactly. On a
much shorter series the floor bites: at $n = 30$ the smallest attainable
p-value is `r round(floor_30, 3L)`, so the subsampling calibration can
reject at the 0.05 level only when the end-of-sample block is larger than
every earlier block in the series. With
the model given and $m = 1$, the rank is uniform up to the first few
innovations, so the size is the grid value:
`r round(attainable_30, 4L)` at $n = 30$ and `r round(attainable_120, 4L)`
at $n = 120$, both below 0.05. The simulated size at $n = 30$,
`r round(size_andrews_30[["size"]], 3L)` over
`r n_rep_short` replicates with a standard error of
`r round(size_andrews_30[["se"]], 3L)`, sits
`r round(andrews_ses_from_grid, 1L)` standard errors from the grid value.
With estimated parameters or $m > 1$ the grid value is not the size, and the
simulation in the numerical illustration finds sizes above it.

The chi-square calibration is continuous and carries no such floor. Its
simulated size is `r round(size_chisq_30[["size"]], 3L)` at $n = 30$
(standard error `r round(size_chisq_30[["se"]], 3L)`, `r n_rep_short`
replicates) and `r round(size_chisq_120[["size"]], 3L)` at $n = 120$
(standard error `r round(size_chisq_120[["se"]], 3L)`, `r n_rep_long`
replicates). Both sit within about
`r round(chisq_worst_se, 1L)`
standard errors of the nominal 0.05, so at these replicate counts the study
resolves no departure from nominal for either length. At $n = 30$ the
simulated sizes of the two calibrations differ by `r round(diff_30, 3L)`,
`r round(diff_30 / diff_30_se, 1L)` standard errors of the difference, so the
simulation does not resolve the contrast between them either. The contrast
comes from the grid: with the model given, the subsampling test has a size
of `r round(attainable_30, 4L)` at $n = 30$ and $m = 1$, while the
chi-square calibration is continuous and can spend the full 0.05.

## Limitations

A rejection rate estimated from `r n_rep_short` replicates carries a
standard error near `r round(sqrt(0.05 * 0.95 / n_rep_short), 3L)`, so a
departure from nominal of one or two percentage points is beyond the
resolution of the size study above. It establishes the order of the size,
not the exact size at a given series length.

The two calibrations agree when the innovations are Gaussian. They part
company when the observation noise is heavy-tailed: the parametric $\chi^2$
reference is then mis-specified, while the subsampling test judges the
end-of-sample block only against the other blocks of the same series and
needs no parametric reference. This vignette simulates Gaussian noise throughout and so
does not exercise that divergence; `method = "andrews"` is the option to
reach for when normality is in doubt, at the cost of the discreteness
documented above. Which calibration is preferable therefore depends on
which failure the analyst minds more: a mis-specified reference
distribution, or a test that cannot spend its nominal level. The numerical
illustration that follows places both calibrations beside the
structural-break, changepoint and forecast-interval tools of four other
packages, on the Nile series and in simulation.

The test is defined for a fitted linear-Gaussian state-space model, supplied
exactly as for `gmm_filter()`: a single-component `gmm` prior, a `dynamics`
list with the state-transition and process-noise matrices, and a
`measurement` list with the observation and observation-noise matrices.
Gaussian-sum process or measurement noise is rejected rather than
approximated, because neither calibration is defined for it. The small-`m`
regime (`m = 1, 2, 3`) is the point of the test, and `m` must be smaller
than the series length.

One further caution belongs to the model rather than the test. The band in
the first figure is the one-step predictive band, which the statistic
divides by, not the filtered state's standard deviation that `gmm_filter()`
returns; at $t = `r n`$ the predictive standard deviation is
`r fixed(pred_sd[n] / path$sd_1[n], 1)` times the filtered one. A model whose noise covariances are themselves estimated on the same short
series can understate both, and the test then inherits that optimism: it is
a test of the last observation against the model, not a test of the model.

```{=latex}
\vspace{0pt plus 10\baselineskip}\penalty-200\vspace{0pt plus -10\baselineskip}
```

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_strucchange <- requireNamespace("strucchange", quietly = TRUE)
has_changepoint <- requireNamespace("changepoint", quietly = TRUE)
has_kfas <- requireNamespace("KFAS", quietly = TRUE)
has_dlm <- requireNamespace("dlm", quietly = TRUE)
run_illustration <- has_strucchange && has_changepoint && has_kfas && has_dlm
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("strucchange, changepoint, KFAS or dlm is not installed on this build,",
    "so the real-data illustration is skipped.\n")
```

### The Nile at 1898, with five packages

The annual flow of the Nile at Aswan, shipped with R as `Nile`, falls to a
lower level after 1898 (Cobb, 1978). We fit a local-level model to the years
1871--1898 and ask each method whether the years that follow are consistent
with it: 1899 alone, so $m = 1$, and 1899--1901, so $m = 3$. The documented
change is the reference every method is scored against, and the score is a
p-value at the 0.05 level.

Beside the two calibrations of `gmm_eos_test()` stand six tests: five from
four packages and a t-test from base R. strucchange (Zeileis et al., 2002) supplies a
sup-F test over the breaks whose final segment lies within the last $m$
years, and an OLS-based CUSUM test over the whole sample. changepoint
(Killick and Eckley, 2014) supplies `cpt.mean()` with its defaults, the PELT
search under the MBIC penalty, run on the series scaled to unit variance
over the fitting years because its Normal cost takes the variance as one;
it returns locations rather than a p-value, and a change located within the
last $m$ years counts as a detection. KFAS (Helske, 2017) and dlm (Petris,
2010) each fit their own local level to 1871--1898 by maximum likelihood and
check every tested year against its one-step forecast interval at level
$1 - 0.05 / m$, a Bonferroni correction. It would hold the check's size at
or below 0.05 whatever the window if the fitted parameters were the true
ones; the simulation below measures the size with estimated parameters. The
p-value reported is that of the largest standardised forecast error, times
$m$. The last test is a t-test on the residuals of the forecasts made at
1898 from the local level fitted by `StructTS()`.

The four packages install from CRAN once and load as usual:

```{r install-competitors, eval = FALSE}
install.packages(c("strucchange", "changepoint", "KFAS", "dlm"))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(strucchange)
library(changepoint)
library(KFAS)
library(dlm)
```

The fitting years are 1871--1898, and the local level is estimated on them
alone:

```{r nile-data, eval = run_illustration}
nile <- as.numeric(window(Nile, 1871, 1901))
year <- 1871:1901
n_fit <- 28L
nile_fit <- nile[seq_len(n_fit)]

level_fit <- StructTS(nile_fit, type = "level")
prior_nile <- gmm(weights = 1, means = list(nile_fit[1L]),
                  covariances = list(matrix(1e7)))
dynamics_nile <- list(A = matrix(1), Q = matrix(level_fit$coef[["level"]]))
measurement_nile <- list(C = matrix(1),
                         R = matrix(level_fit$coef[["epsilon"]]))
```

The state-space fits and the t-test use those years alone and are applied
to the series truncated $m$ years later; the strucchange tests and
`cpt.mean()` run on the whole truncated series. A test that is not defined at that window returns `NA`,
which the table below prints as "not defined".

```{r nile-tests, eval = run_illustration}
nile_tests <- function(m) {
  y <- nile[seq_len(n_fit + m)]
  n <- length(y)
  last <- (n_fit + 1L):n

  p_chisq <- gmm_eos_test(prior_nile, dynamics_nile, measurement_nile, y,
                          m = m, method = "chisq")$p_value
  p_andrews <- gmm_eos_test(prior_nile, dynamics_nile, measurement_nile, y,
                            m = m, method = "andrews")$p_value

  # strucchange needs at least two observations after a break in a mean
  p_supf <- if (m >= 2L) {
    unname(sctest(Fstats(y ~ 1, from = n - m, to = n - 2L))$p.value)
  } else NA_real_
  p_cusum <- unname(sctest(efp(y ~ 1, type = "OLS-CUSUM"))$p.value)

  cp <- cpts(cpt.mean(y / sd(nile_fit)))
  found_cpt <- any(cp >= n_fit)

  kfas_fit <- fitSSM(SSModel(nile_fit ~ SSMtrend(1L, Q = NA), H = NA),
                     inits = log(c(var(nile_fit), var(nile_fit))),
                     method = "BFGS")
  kfas_model <- SSModel(y ~ SSMtrend(1L, Q = kfas_fit$model$Q[1L]),
                        H = kfas_fit$model$H[1L])
  kfas_pred <- predict(kfas_model, interval = "prediction", level = 0.95,
                       filtered = TRUE)
  kfas_z <- (y[last] - kfas_pred[last, "fit"]) /
    ((kfas_pred[last, "upr"] - kfas_pred[last, "lwr"]) / (2 * qnorm(0.975)))
  p_kfas <- min(1, m * 2 * pnorm(-max(abs(kfas_z))))

  dlm_build <- function(pars) {
    dlmModPoly(1L, dV = exp(pars[1L]), dW = exp(pars[2L]))
  }
  dlm_fit <- dlmMLE(nile_fit, parm = log(c(var(nile_fit), var(nile_fit))),
                    build = dlm_build)
  dlm_filt <- dlmFilter(y, dlm_build(dlm_fit$par))
  dlm_var <- unlist(dlmSvd2var(dlm_filt$U.R, dlm_filt$D.R)) +
    exp(dlm_fit$par[1L])
  dlm_z <- (y[last] - dlm_filt$f[last]) / sqrt(dlm_var[last])
  p_dlm <- min(1, m * 2 * pnorm(-max(abs(dlm_z))))

  resid <- y[last] - predict(level_fit, n.ahead = m)$pred
  p_t <- if (m > 1L) t.test(resid)$p.value else NA_real_

  list(
    p = c("proxymix chi-square" = p_chisq,
          "proxymix subsampling" = p_andrews,
          "strucchange sup-F" = p_supf,
          "strucchange OLS-CUSUM" = p_cusum,
          "KFAS forecast interval" = p_kfas,
          "dlm forecast interval" = p_dlm,
          "t-test on forecast residuals" = p_t),
    found_cpt = found_cpt
  )
}

nile_1 <- nile_tests(1L)
nile_3 <- nile_tests(3L)
```

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.35\textheight\clearpage\fi
```

```{r nile-table, eval = run_illustration, echo = FALSE}
p_cell <- function(p) {
  ifelse(is.na(p), "not defined",
         paste0(format(round(p, 3L), nsmall = 3L),
                ifelse(p < 0.05, " (reject)", "")))
}
cpt_cell <- function(found) {
  if (found) "change located (reject)" else "no change located"
}
nile_tbl <- data.frame(
  method = c(names(nile_1$p), "changepoint cpt.mean"),
  m1 = c(unname(p_cell(nile_1$p)), cpt_cell(nile_1$found_cpt)),
  m3 = c(unname(p_cell(nile_3$p)), cpt_cell(nile_3$found_cpt)),
  stringsAsFactors = FALSE
)
knitr::kable(
  nile_tbl, align = c("l", "r", "r"), row.names = FALSE,
  col.names = c("Test", "m = 1", "m = 3"),
  caption = paste0(
    "Each test's p-value on the Nile series fitted to 1871--1898, for 1899 ",
    "(m = 1) and for 1899--1901 (m = 3), against the documented change at ",
    "1898; cpt.mean reports a location rather than a p-value."
  )
)
```

```{r nile-facts, eval = run_illustration, include = FALSE}
sd_obs <- sqrt(level_fit$coef[["epsilon"]])
level_1898 <- predict(level_fit, n.ahead = 1L)$pred[[1L]]
drop_1899 <- (level_1898 - nile[n_fit + 1L]) / sd_obs
floor_den_1 <- (n_fit + 1L) - 2L * 1L + 2L
floor_den_3 <- (n_fit + 3L) - 2L * 3L + 2L
```

```{r fig-nile, eval = run_illustration && has_ggplot2, echo = FALSE, fig.height = 4, fig.cap = "The Nile flow for 1871--1901 with the local level fitted to 1871--1898 and its 95% forecast band for the three years that follow. The dashed vertical line separates the fitting years from the tested years. The 1899 observation falls below the band.", fig.alt = "A line of annual Nile flows from 1871 to 1901, a horizontal fitted level through 1898 with a shaded forecast band extending over 1899 to 1901, and the 1899 to 1901 points lying at or below the lower edge of the band."}
fc <- predict(level_fit, n.ahead = 3L)
nile_df <- data.frame(year = year, flow = nile,
                      period = ifelse(year <= 1898, "fitting years",
                                      "tested years"))
band_df <- data.frame(year = 1899:1901, fit = as.numeric(fc$pred),
                      lo = as.numeric(fc$pred - qnorm(0.975) * fc$se),
                      hi = as.numeric(fc$pred + qnorm(0.975) * fc$se))
ggplot2::ggplot(nile_df, ggplot2::aes(year, flow)) +
  ggplot2::geom_ribbon(data = band_df,
                       ggplot2::aes(year, ymin = lo, ymax = hi),
                       inherit.aes = FALSE, fill = "#0072B2", alpha = 0.18) +
  ggplot2::geom_line(colour = "grey60") +
  ggplot2::geom_point(ggplot2::aes(colour = period), size = 2) +
  ggplot2::geom_hline(yintercept = level_1898, colour = "#0072B2",
                      linewidth = 0.8) +
  ggplot2::geom_vline(xintercept = 1898.5, linetype = "dashed",
                      colour = "grey40") +
  ggplot2::scale_colour_manual(name = NULL,
                               values = c("fitting years" = "grey30",
                                          "tested years" = "#D55E00")) +
  ggplot2::labs(x = "year", y = "annual flow",
                title = "The Nile after 1898 against the level fitted before it") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-nile-skip, eval = run_illustration && !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the Nile figure is",
    "skipped.\n")
```

The local level fitted to 1871--1898 has a level variance of
`r if (level_fit$coef[["level"]] == 0) "zero" else fixed(level_fit$coef[["level"]], 1)`, so the model is a constant mean
with an observation standard deviation of `r round(sd_obs)`, and the 1899
flow sits `r fixed(drop_1899, 2)` of those standard deviations below it. At
$m = 1$ the chi-square calibration returns a p-value of
`r fixed(nile_1$p[["proxymix chi-square"]], 3)`, and the KFAS and dlm
interval checks return `r fixed(nile_1$p[["KFAS forecast interval"]], 3)`
and `r fixed(nile_1$p[["dlm forecast interval"]], 3)`: at a single
observation the three are the same test, a one-step forecast error referred
to its standard deviation, and they differ only in the fitted variances. The
subsampling calibration returns
`r fixed(nile_1$p[["proxymix subsampling"]], 3)`, which is its floor of
$1 / (n - 2m + 2) = 1 / `r floor_den_1`$ at this length: the 1899 block is
the largest in the series. The sup-F test is not defined at $m = 1$, because
strucchange's `Fstats()` admits a break in a mean no later than $n - 2$, so
at least two observations must follow it. The t-test is not defined at
$m = 1$ either, because it estimates the variance of the forecast residuals
from the tested years themselves. Chow's predictive test, which takes that
variance from the fitting years, is defined at one observation. The
OLS-CUSUM test, which asks about the whole sample, returns
`r fixed(nile_1$p[["strucchange OLS-CUSUM"]], 3)`. `cpt.mean()` locates
`r if (nile_1$found_cpt) "a change" else "no change"` on the scaled series.
At $m = 3$ the sup-F test is defined and returns
`r fixed(nile_3$p[["strucchange sup-F"]], 3)`, the chi-square calibration
`r fixed(nile_3$p[["proxymix chi-square"]], 3)` and the t-test
`r fixed(nile_3$p[["t-test on forecast residuals"]], 3)`, while the interval
checks, spending their level over three years, return
`r fixed(nile_3$p[["KFAS forecast interval"]], 3)` and
`r fixed(nile_3$p[["dlm forecast interval"]], 3)`. The subsampling p-value
is again at its floor, $1 / `r floor_den_3`$. This is one series and one
documented change, which shows which tests exist at each window but cannot
measure their size or power, so the simulation below does.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("end_of_sample_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
```

The simulation generates `r sim$n_rep` stationary AR(1) series of length
`r sim$n` with coefficient `r sim$phi` and unit innovation variance. For
each series and each window $m \in \{`r paste(sim$m_vec, collapse = ", ")`\}$,
every method is applied twice: to the series as generated, which gives the
size, and to the series with its last $m$ observations raised by
`r sim$delta` innovation standard deviations, which gives the power. The
state-space methods and the t-test are fitted to the first $n - m$
observations; the strucchange tests and `cpt.mean()` run on the whole
series, as described below. The reference is the design itself,
whether or not a shift was applied, and the score is the rejection rate at a
nominal 0.05.

proxymix and the t-test receive the AR(1) fitted by `arima()`, written as a
state-space model with the state observed exactly; the t-test's residuals
are those of the forecasts made at $n - m$. KFAS and dlm each fit their own
AR(1) to the fitting window after its mean is removed, and check the last
$m$ observations against one-step forecast intervals at level $1 - 0.05 /
m$, as on the Nile. strucchange runs the sup-F test on the regression of
$y_t$ on $y_{t-1}$ over the breaks whose final segment lies within the last
$m$ observations, which needs at least three observations after the break,
so it is defined at $m = 3$ and $m = 5$ and not at $m = 1$. Its OLS-CUSUM
test runs over the whole sample. `cpt.mean()` runs on the series scaled to
unit variance over the fitting window, and a change located within the last
$m$ observations counts as a rejection. A rejection rate near 0.05 from
`r sim$n_rep` series carries a Monte Carlo standard error of about
`r fixed(sqrt(0.05 * 0.95 / sim$n_rep), 3)`, and one near 0.5 about
`r fixed(sqrt(0.25 / sim$n_rep), 3)`. This chunk is complete and runs as
shown, but it took about `r round(sim$elapsed_secs / 60)` minutes on one
core (R `r sim$versions[["R"]]`, Apple silicon), so the table that follows it is read from its stored output. For a
quick reduced run, set `n_rep` to a small number such as `20L` on the line
that defines it.

```{r sim-code, eval = FALSE}
library(proxymix)
library(strucchange)
library(changepoint)
library(KFAS)
library(dlm)

n <- 100L        # length of each series
phi <- 0.6       # autoregressive coefficient
delta <- 3       # level shift, in innovation standard deviations
m_vec <- c(1L, 3L, 5L)
n_rep <- 1000L   # series per window

# every test is fitted to the first n - m observations of y and applied to
# its last m; NA marks a test that is not defined at that window
p_values <- function(y, m) {
  n <- length(y)
  y_fit <- y[seq_len(n - m)]
  last <- (n - m + 1L):n

  ar <- arima(y_fit, order = c(1L, 0L, 0L))
  a1 <- coef(ar)[["ar1"]]
  mu <- coef(ar)[["intercept"]]
  s2 <- ar$sigma2
  prior <- gmm(weights = 1, means = list(mu),
               covariances = list(matrix(s2 / (1 - a1^2))))
  dynamics <- list(A = matrix(a1), b = mu * (1 - a1), Q = matrix(s2))
  measurement <- list(C = matrix(1), R = matrix(0))
  p_chisq <- gmm_eos_test(prior, dynamics, measurement, y, m = m,
                          method = "chisq")$p_value
  p_andrews <- gmm_eos_test(prior, dynamics, measurement, y, m = m,
                            method = "andrews")$p_value
  resid <- y[last] - predict(ar, n.ahead = m)$pred
  p_t <- if (m > 1L) t.test(resid)$p.value else NA_real_

  # strucchange needs at least three observations after a break in an AR(1)
  reg <- data.frame(y = y[-1L], y_lag = y[-n])
  n_reg <- nrow(reg)
  p_supf <- if (m >= 3L) {
    fs <- Fstats(y ~ y_lag, data = reg, from = n_reg - m, to = n_reg - 3L)
    unname(sctest(fs)$p.value)
  } else NA_real_
  p_cusum <- unname(sctest(efp(y ~ y_lag, data = reg,
                               type = "OLS-CUSUM"))$p.value)

  cp <- cpts(cpt.mean(y / sd(y_fit)))
  p_cpt <- if (any(cp >= n - m)) 0 else 1

  y_bar <- mean(y_fit)
  kfas_update <- function(pars, model) {
    part <- SSMarima(ar = 0.999 * tanh(pars[1L]), Q = exp(pars[2L]))
    model["T", "arima"] <- part$T
    model["R", "arima"] <- part$R
    model["Q", "arima"] <- part$Q
    model["P1", "arima"] <- part$P1
    model
  }
  kfas_fit <- fitSSM(
    SSModel(I(y_fit - y_bar) ~ -1 + SSMarima(ar = 0.5, Q = 1), H = 0),
    inits = c(0, 0), updatefn = kfas_update, method = "BFGS"
  )
  kfas_par <- kfas_fit$optim.out$par
  kfas_model <- SSModel(
    I(y - y_bar) ~ -1 + SSMarima(ar = 0.999 * tanh(kfas_par[1L]),
                                  Q = exp(kfas_par[2L])),
    H = 0
  )
  kfas_pred <- predict(kfas_model, interval = "prediction", level = 0.95,
                       filtered = TRUE)
  kfas_z <- (y[last] - y_bar - kfas_pred[last, "fit"]) /
    ((kfas_pred[last, "upr"] - kfas_pred[last, "lwr"]) / (2 * qnorm(0.975)))
  p_kfas <- min(1, m * 2 * pnorm(-max(abs(kfas_z))))

  dlm_build <- function(pars) {
    dlmModARMA(ar = 0.999 * tanh(pars[1L]), sigma2 = exp(pars[2L]), dV = 0)
  }
  dlm_fit <- dlmMLE(y_fit - y_bar, parm = c(0, 0), build = dlm_build)
  dlm_filt <- dlmFilter(y - y_bar, dlm_build(dlm_fit$par))
  dlm_var <- unlist(dlmSvd2var(dlm_filt$U.R, dlm_filt$D.R))
  dlm_z <- (y[last] - y_bar - dlm_filt$f[last]) / sqrt(dlm_var[last])
  p_dlm <- min(1, m * 2 * pnorm(-max(abs(dlm_z))))

  c("proxymix chi-square" = p_chisq, "proxymix subsampling" = p_andrews,
    "strucchange sup-F" = p_supf, "strucchange OLS-CUSUM" = p_cusum,
    "changepoint cpt.mean" = p_cpt, "KFAS forecast interval" = p_kfas,
    "dlm forecast interval" = p_dlm, "t-test on forecast residuals" = p_t)
}

one_series <- function(r) {
  set.seed(r)
  y <- as.numeric(arima.sim(list(ar = phi), n = n))
  do.call(rbind, lapply(m_vec, function(m) {
    y_shift <- y
    y_shift[(n - m + 1L):n] <- y_shift[(n - m + 1L):n] + delta
    p_null <- p_values(y, m)
    p_shift <- p_values(y_shift, m)
    data.frame(r = r, m = m, method = rep(names(p_null), 2L),
               scenario = rep(c("no shift", "shift"), each = length(p_null)),
               reject = c(p_null, p_shift) < 0.05,
               stringsAsFactors = FALSE)
  }))
}

res <- do.call(rbind, lapply(seq_len(n_rep), one_series))

sim_tab <- aggregate(reject ~ m + scenario + method, data = res, FUN = mean,
                     na.action = na.pass)
names(sim_tab)[4L] <- "rate"
sim_tab$se <- sqrt(sim_tab$rate * (1 - sim_tab$rate) / n_rep)
sim_tab
```

```{=latex}
\par\penalty-100\begingroup\dimen0=16\baselineskip\dimen2=\pagegoal
\advance\dimen2 by -\pagetotal
\ifdim\dimen0>\dimen2 \ifdim\dimen2>0pt \vfil\fi\break\fi\endgroup
```

```{r sim-table, echo = FALSE}
method_order <- c("proxymix chi-square", "proxymix subsampling",
                  "strucchange sup-F", "strucchange OLS-CUSUM",
                  "changepoint cpt.mean", "KFAS forecast interval",
                  "dlm forecast interval", "t-test on forecast residuals")
sim_value <- function(m, scenario, method, what = "rate") {
  s1 <- sim$sim_tab$m == m & sim$sim_tab$scenario == scenario &
    sim$sim_tab$method == method
  sim$sim_tab[[what]][s1]
}
rate_wide <- function(scenario, m) {
  v <- vapply(method_order, function(k) sim_value(m, scenario, k),
              numeric(1L))
  ifelse(is.na(v), "--", format(round(v, 3L), nsmall = 3L))
}
tab_sim <- data.frame(method = method_order, stringsAsFactors = FALSE)
for (m1 in sim$m_vec) {
  tab_sim[[paste0("size_", m1)]] <- rate_wide("no shift", m1)
  tab_sim[[paste0("power_", m1)]] <- rate_wide("shift", m1)
} # ends m1, one size and one power column per window
knitr::kable(
  tab_sim, align = c("l", rep("r", 2L * length(sim$m_vec))),
  row.names = FALSE,
  col.names = c("Test", as.vector(rbind(paste("Size", sim$m_vec),
                                        paste("Power", sim$m_vec)))),
  caption = paste0(
    "Rejection rate at a nominal 0.05 over ", sim$n_rep,
    " simulated AR(1) series per window: the size, without a shift, and ",
    "the power, with the last m observations raised by ", sim$delta,
    " innovation standard deviations, scored against the design. The ",
    "number in each column heading is the window m. A dash marks a test ",
    "that is not defined at that window."
  )
)
```

```{r sim-facts, echo = FALSE}
size_of <- function(method) {
  vapply(sim$m_vec, function(m) sim_value(m, "no shift", method), numeric(1L))
}
power_of <- function(method) {
  vapply(sim$m_vec, function(m) sim_value(m, "shift", method), numeric(1L))
}
range_text <- function(v, digits = 3L) {
  lo <- fixed(min(v, na.rm = TRUE), digits)
  hi <- fixed(max(v, na.rm = TRUE), digits)
  if (lo == hi) paste(lo, "at every window") else paste(lo, "to", hi)
}
mcse_size <- sqrt(0.05 * 0.95 / sim$n_rep)
size_1 <- vapply(method_order, function(k) sim_value(1L, "no shift", k),
                 numeric(1L))
n_defined_1 <- sum(!is.na(size_1))
# tests at m = 1 whose size is within one Monte Carlo standard error of 0.05
near_nominal_1 <- names(size_1)[!is.na(size_1) &
                                  abs(size_1 - 0.05) <= mcse_size]
stopifnot(identical(near_nominal_1, "proxymix subsampling"))
power_1 <- c(power_of("proxymix chi-square")[1L],
             power_of("KFAS forecast interval")[1L],
             power_of("dlm forecast interval")[1L])
over_se <- (c(size_of("proxymix chi-square"), size_of("KFAS forecast interval"),
              size_of("dlm forecast interval")) - 0.05) / mcse_size
power_5 <- vapply(method_order, function(k) sim_value(5L, "shift", k),
                  numeric(1L))
stopifnot(names(which.max(power_5)) == "t-test on forecast residuals")
t_size_ratio <- size_of("t-test on forecast residuals")[2:3] / 0.05
size_sub <- size_of("proxymix subsampling")
sub_over_se <- (size_sub - 0.05) / mcse_size
stopifnot(all(sub_over_se[-1L] > 1))
power_kfas <- power_of("KFAS forecast interval")
power_dlm <- power_of("dlm forecast interval")
stopifnot(all(power_kfas[-1L] < power_kfas[1L]),
          all(power_dlm[-1L] < power_dlm[1L]))
# a version as its DESCRIPTION writes it, when the installed one is the one stored
desc_version <- function(pkg, stored) {
  if (identical(as.character(packageVersion(pkg)), stored)) {
    utils::packageDescription(pkg)$Version
  } else {
    stored
  }
}
pv <- function(pkg) desc_version(pkg, sim$versions[[pkg]])
count_word <- function(k) {
  c("one", "two", "three", "four", "five", "six", "seven", "eight")[k]
}
attainable_sim <- vapply(sim$m_vec, function(m) attainable_size(sim$n, m),
                         numeric(1L))
```

Of the eight tests, `r count_word(n_defined_1)` are defined at $m = 1$. Four of them
score the one-step forecast error of the last observation: the two
calibrations of `gmm_eos_test()` and the forecast-interval checks of KFAS
and dlm, which at one observation are the same statistic as the chi-square
calibration, with parameters from each package's own fit. The powers of
those three are within `r fixed(max(power_1) - min(power_1), 3)` of each
other, at
`r fixed(power_of("proxymix chi-square")[1L], 3)` for the chi-square
calibration, `r fixed(power_of("KFAS forecast interval")[1L], 3)` for KFAS
and `r fixed(power_of("dlm forecast interval")[1L], 3)` for dlm, and their
sizes are `r fixed(size_of("proxymix chi-square")[1L], 3)`,
`r fixed(size_of("KFAS forecast interval")[1L], 3)` and
`r fixed(size_of("dlm forecast interval")[1L], 3)`, all above the nominal
0.05, because the parameters are estimated on the fitting window and the
tests treat them as known. The subsampling rank does not depend on the
innovation variance, which scales every block alike, but it still uses the
estimated autoregressive coefficient and mean, and its in-sample blocks are
residuals of the fit while the end block is not. It has a size of
`r fixed(size_of("proxymix subsampling")[1L], 3)` and a power of
`r fixed(power_of("proxymix subsampling")[1L], 3)` at $m = 1$. The sup-F test
and the t-test are not defined, the OLS-CUSUM test rejects the shifted
series in `r fixed(power_of("strucchange OLS-CUSUM")[1L], 3)` of cases, and
`cpt.mean()` locates a change in the last observation in
`r fixed(power_of("changepoint cpt.mean")[1L], 3)` of shifted series against
`r fixed(size_of("changepoint cpt.mean")[1L], 3)` of unshifted ones.

Across the three windows the chi-square calibration's size is
`r range_text(size_of("proxymix chi-square"))` and its power
`r range_text(power_of("proxymix chi-square"))`; the subsampling
calibration's size is `r range_text(size_of("proxymix subsampling"))`
and its power `r range_text(power_of("proxymix subsampling"))`. The
interval checks keep their size near its $m = 1$ value, with KFAS at
`r range_text(size_of("KFAS forecast interval"))` and dlm at
`r range_text(size_of("dlm forecast interval"))`. Their power falls from
`r fixed(power_kfas[1L], 3)` at $m = 1$ to `r fixed(power_kfas[2L], 3)` at
$m = 3$ for KFAS and from `r fixed(power_dlm[1L], 3)` to
`r fixed(power_dlm[2L], 3)` for dlm, because each check's level is divided
by $m$ and under an AR(1) only the first shifted observation carries the
whole shift in its one-step forecast error, the later ones carry
$1 - \phi$ of it. The sup-F test at $m = 3$ has one
admissible break and rejects `r fixed(size_of("strucchange sup-F")[2L], 3)`
of unshifted series with strucchange `r pv("strucchange")`,
because its p-value comes from an asymptotic approximation for a break
range that this window places at the edge of the sample; at $m = 5$ its
size is `r fixed(size_of("strucchange sup-F")[3L], 3)` and its power
`r fixed(power_of("strucchange sup-F")[3L], 3)`, against
`r fixed(power_of("proxymix chi-square")[3L], 3)` at a size of
`r fixed(size_of("proxymix chi-square")[3L], 3)` for the chi-square
calibration. The t-test on the forecast residuals rejects
`r fixed(size_of("t-test on forecast residuals")[2L], 3)` and
`r fixed(size_of("t-test on forecast residuals")[3L], 3)` of unshifted
series at $m = 3$ and $m = 5$, because the residuals of forecasts made at
one origin are correlated and the test takes them as independent. The
OLS-CUSUM test rejects unshifted series at a rate of
`r range_text(size_of("strucchange OLS-CUSUM"))` and shifted ones at
`r range_text(power_of("strucchange OLS-CUSUM"))`; a shift confined to the last few observations barely moves a
cumulative sum over the whole sample. `cpt.mean()` locates a change in the
last $m$ observations of unshifted series at a rate of
`r range_text(size_of("changepoint cpt.mean"))` and of shifted ones at
`r range_text(power_of("changepoint cpt.mean"))`. A size estimated from `r sim$n_rep` series has a Monte Carlo
standard error of about `r fixed(mcse_size, 3)`, so the sizes of the
chi-square calibration and of the interval checks sit
`r fixed(min(over_se), 1)` to `r fixed(max(over_se), 1)` standard errors
above 0.05, and the subsampling size at $m = 1$ is not distinguished
from it.

In this run the subsampling calibration of `gmm_eos_test()` was the only
one of the `r count_word(n_defined_1)` tests defined at $m = 1$ to hold a size within a
standard error of nominal there, with a power of
`r fixed(power_of("proxymix subsampling")[1L], 3)`. At $m = 3$ and $m = 5$
its size was `r fixed(size_sub[2L], 3)` and `r fixed(size_sub[3L], 3)`,
`r fixed(sub_over_se[2L], 1)` and `r fixed(sub_over_se[3L], 1)` standard
errors above 0.05. Its grid values at $n = `r sim$n`$ are
`r fixed(attainable_sim[1L], 3)`, `r fixed(attainable_sim[2L], 3)` and
`r fixed(attainable_sim[3L], 3)` at $m = 1$, 3 and 5, which are the size
only when the end block's rank is uniform; with estimated parameters and
overlapping blocks it is not. At $m = 5$ the sup-F
test of strucchange had a size closer to nominal than either calibration,
`r fixed(size_of("strucchange sup-F")[3L], 3)`, and a power of
`r fixed(power_of("strucchange sup-F")[3L], 3)` against
`r fixed(power_of("proxymix chi-square")[3L], 3)` for the chi-square
calibration and `r fixed(power_of("proxymix subsampling")[3L], 3)` for the
subsampling calibration. The t-test had the highest rejection rate under
the shift at $m = 5$,
`r fixed(power_of("t-test on forecast residuals")[3L], 3)`, and
`r fixed(power_of("t-test on forecast residuals")[2L], 3)` at $m = 3$
against `r fixed(power_of("strucchange sup-F")[2L], 3)` for the sup-F test,
but its size was `r fixed(t_size_ratio[1L], 1)` and
`r fixed(t_size_ratio[2L], 1)` times the nominal level at $m = 3$ and
$m = 5$, so these rates are not comparable to the powers of tests that
hold their size. The comparison covers one autoregressive coefficient, one
series length, one shift size and a level shift confined to the tested
window, not the smaller or gradual shifts, the longer windows, or the
non-Gaussian innovations under which the two calibrations part company.

## Where next

*The closed-form operator calculus on a mixture* builds the predict-and-update
recursion this test runs internally, and covers `gmm_filter()` and the
mixture reduction that bounds it. *Reading the entropy of a fitted mixture*
covers the diagnostics that describe a fitted mixture once it is in hand.
*Fitting a proxy to a density you cannot sample* is the entry point to the
fitting machinery if the state-space model here is your first contact with
the package.

## References

* Andrews, D. W. K. (1993). *Tests for parameter instability and
  structural change with unknown change point.* Econometrica 61(4),
  821--856. <doi:10.2307/2951764>.
* Andrews, D. W. K. (2003). *End-of-sample instability tests.* Econometrica
  71(6), 1661--1694. <doi:10.1111/1468-0262.00466>.
* Chow, G. C. (1960). *Tests of equality between sets of coefficients in
  two linear regressions.* Econometrica 28(3), 591--605.
  <doi:10.2307/1910133>.
* Cobb, G. W. (1978). *The problem of the Nile: conditional solution to a
  changepoint problem.* Biometrika 65(2), 243--251.
  <doi:10.1093/biomet/65.2.243>.
* Helske, J. (2017). *KFAS: Exponential family state space models in R.*
  Journal of Statistical Software 78(10), 1--39.
  <doi:10.18637/jss.v078.i10>.
* Killick, R. and Eckley, I. A. (2014). *changepoint: An R package for
  changepoint analysis.* Journal of Statistical Software 58(3), 1--19.
  <doi:10.18637/jss.v058.i03>.
* Petris, G. (2010). *An R package for dynamic linear models.* Journal of
  Statistical Software 36(12), 1--16. <doi:10.18637/jss.v036.i12>.
* Zeileis, A., Leisch, F., Hornik, K. and Kleiber, C. (2002). *strucchange:
  An R package for testing for structural change in linear regression
  models.* Journal of Statistical Software 7(2), 1--38.
  <doi:10.18637/jss.v007.i02>.

## Reproduce

The vignette sets `set.seed(20260621)` once, before the series is
simulated, and every result up to the numerical illustration -- the two
tests, the filter path and both size studies -- is a deterministic function
of that stream.

The Nile illustration draws nothing at random: every fit is a maximum
likelihood estimate from fixed starting values. In the simulation each
series is generated after `set.seed()` with its own index, and both scenarios
at every window share that series. The stored simulation results come from
the code shown, run on `r sim$run_date` with R `r sim$versions[["R"]]`,
proxymix `r pv("proxymix")`, strucchange `r pv("strucchange")`,
changepoint `r pv("changepoint")`, KFAS `r pv("KFAS")` and dlm
`r pv("dlm")`, and the run raised `r length(sim$warnings)`
warnings.

```{r session-info, eval = FALSE}
print(sessionInfo(), locale = FALSE)
```

```{r session-info-print, echo = FALSE}
## printed 68 characters wide; a longer line breaks after its last slash or
## space within the width and continues indented
old_width <- options(width = 68L)
si_vec <- capture.output(print(sessionInfo(), locale = FALSE))
options(old_width)
wrap_line <- function(s) {
  out <- character(0L)
  while (nchar(s) > 68L) {
    cut <- max(gregexpr("[/ ]", substr(s, 1L, 68L))[[1L]])
    if (cut < 20L) cut <- 68L
    out <- c(out, substr(s, 1L, cut))
    s <- paste0("    ", substr(s, cut + 1L, nchar(s)))
  }
  c(out, s)
}
writeLines(unlist(lapply(si_vec, wrap_line)))
```
