---
title: "Reading the entropy of a fitted mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading the entropy of a fitted mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~11 s under rmarkdown::render() with ggplot2, FNN,
     glasso, huge and pcalg installed, the simulation read from its stored
     results; macOS arm64 (Apple silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The problem

A fitted mixture is a list of weights, means and covariance matrices. The
list alone does not answer three common questions: how spread out the
mixture is, how far it is from another fit, and how many components the
data support. All three are questions about entropy, and for a Gaussian
mixture most of them have closed-form answers. This vignette shows which
entropy quantities are exact, which are estimated, and what conclusions
each one supports.

## Package capabilities

The vignette uses eight functions. `gmm_entropy()` returns the order-2 Rényi
entropy of a mixture in closed form by default, and a Monte Carlo Shannon
estimate with its standard error and an analytic upper bound under
`order = "shannon"`. `gmm_divergence()` returns the symmetric
Cauchy-Schwarz divergence between two mixtures in closed form. Under
`type = "kl"` it returns a Monte Carlo estimate of the Kullback-Leibler
divergence $\mathrm{KL}(f \Vert g) = \int f \log (f / g)$, with $f$ the first
mixture and $g$ the second.

`gmm_mutual_information()` measures dependence between two coordinate
blocks of a fitted joint as the Cauchy-Schwarz divergence between the joint
and the product of its marginals, and
`gmm_conditional_entropy()` returns the predictive uncertainty of the
target coordinates given the conditioned ones, row by row, as a Rényi-2
entropy.

`gmm_anneal_path()` cools a fit along a deterministic-annealing schedule
(Rose, 1998) and records where the components split apart. The result gives
both a starting point for a fit and a component count. `bic_aic()` reports the integrated completed likelihood
beside the usual criteria, and `gmm_independence_graph()` reads the
second-order conditional-independence structure of a fitted mixture.
`maxent_target()` works in the opposite direction. It starts from a set of
constraints and builds the density with the largest entropy that meets them
(Jaynes, 1957).

## Addressing the problem

```{r seed}
set.seed(20260618)
```

```{r fixed-notation, include = FALSE}
## Small errors are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
## A stored package version written as the package's DESCRIPTION writes it
## (2.7-12, not 2.7.12), when the installed version is the one stored.
desc_version <- function(pkg) {
  stored <- sim$versions[[pkg]]
  installed <- utils::packageDescription(pkg)$Version
  if (identical(as.character(package_version(installed)), stored)) {
    installed
  } else {
    stored
  }
}
```

### Why some quantities have closed forms

The integral of a product of two Gaussian densities is itself a Gaussian
density evaluation: writing $\mathcal{N}(x; m, S)$ for a Gaussian density,
the integral over $x$ of
$\mathcal{N}(x; a, A)\, \mathcal{N}(x; b, B)$ equals
$\mathcal{N}(a; b, A + B)$. Any quantity that is *quadratic* in a mixture --
the integral of $g(x)^2$, or of $p(x)\, q(x)$ -- is therefore a finite sum
of such evaluations, and closed-form. Shannon entropy carries the logarithm
of a sum and has no such closed form, so it is estimated by Monte Carlo.

### Rényi-2 entropy against its analytic value

`gmm_entropy()` returns the closed-form order-2 Rényi entropy by default.
For a single Gaussian the value is exact and matches the analytic form
$H_2(\mathcal{N}(\mu, \Sigma)) =
\tfrac{p}{2}\log(4\pi) + \tfrac{1}{2}\log\det\Sigma$ in $p$ dimensions,
which is an independent check on the implementation.

```{r renyi}
g <- gmm(
  weights = c(0.5, 0.5),
  means = list(c(-2, 0), c(2, 0)),
  covariances = list(diag(2), diag(2))
)
h2_mixture <- gmm_entropy(g)

sigma_one <- matrix(c(1, 0.3, 0.3, 1), 2L, 2L)
one <- gmm(weights = 1, means = list(c(0, 0)),
           covariances = list(sigma_one))
h2_closed <- gmm_entropy(one)
h2_analytic <- 0.5 * (2 * log(4 * pi) +
  as.numeric(determinant(sigma_one, logarithm = TRUE)$modulus))
h2_gap <- abs(h2_closed - h2_analytic)
```

```{r renyi-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "two-component mixture, closed form",
      "single Gaussian, closed form",
      "single Gaussian, analytic formula"
    ),
    value = round(c(h2_mixture, h2_closed, h2_analytic), 6L)
  ),
  col.names = c("Quantity", "Rényi-2 entropy (nats)"),
  caption = "Closed-form order-2 Rényi entropy, and the analytic check."
)
```

### Shannon entropy, bracketed from above

Shannon entropy is returned as a Monte Carlo estimate with its standard
error and an analytic upper bound.

```{r shannon}
sh <- gmm_entropy(g, order = "shannon", n_mc = 5000L, seed = 1L)
sh_slack <- sh$upper_bound - sh$mc
sh_slack_se <- sh_slack / sh$mc_se
```

```{r shannon-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "Monte Carlo estimate", "Monte Carlo standard error",
      "analytic upper bound", "draws"
    ),
    value = c(
      formatC(c(sh$mc, sh$mc_se, sh$upper_bound), format = "f", digits = 4L),
      formatC(sh$n_mc, format = "d")
    )
  ),
  col.names = c("Quantity", "Value"), align = c("l", "r"),
  caption = "Shannon entropy of the two-component mixture, in nats."
)
```

### Cauchy-Schwarz divergence, and the Kullback-Leibler alternative

`gmm_divergence()` returns the closed-form, symmetric Cauchy-Schwarz
divergence by default. It is non-negative and zero exactly when the two
mixtures are equal. When the asymmetric Kullback-Leibler divergence
is wanted instead, `type = "kl"` delegates to the package's Monte Carlo
estimator `gmm_kld()`.

```{r divergence}
q <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(diag(2) * 2)
)
d_cs <- gmm_divergence(g, q)
d_self <- gmm_divergence(g, g)
# takes no seed argument: the draws come from the stream set.seed() started
d_kl <- gmm_divergence(g, q, type = "kl", n_mc = 2000L)
d_kl_gap <- d_kl$mc - d_kl$variational
```

```{r divergence-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "Cauchy-Schwarz, g against q", "Cauchy-Schwarz, g against itself",
      "Kullback-Leibler, Monte Carlo", "Kullback-Leibler, Monte Carlo SE",
      "Kullback-Leibler, Hershey and Olsen approximation"
    ),
    value = round(
      c(d_cs, d_self, d_kl$mc, d_kl$mc_se, d_kl$variational), 4L
    )
  ),
  col.names = c("Quantity", "Value (nats)"),
  caption = "Two divergences between the same pair of mixtures."
)
```

### Mutual information and predictive entropy

Mutual information and conditional entropy also have closed forms. The
mutual information is non-negative and is zero exactly when the blocks are
independent. The conditional entropy is the order-2 Rényi entropy of the
conditional mixture that `gmm_conditionalise()` returns. It is evaluated
row by row, with `NA` marking the target coordinate.

```{r mutual-information}
sigma_joint <- matrix(c(1, 0.7, 0.7, 1), 2L, 2L)
joint <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(sigma_joint)
)
mi <- gmm_mutual_information(joint, 1L, 2L)

sigma_free <- diag(2)
independent <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(sigma_free)
)
mi_independent <- gmm_mutual_information(independent, 1L, 2L)

given_grid <- rbind(c(NA, 0), c(NA, 1), c(NA, 2))
h_cond <- gmm_conditional_entropy(joint, given = given_grid)
```

```{r mutual-information-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "Cauchy-Schwarz mutual information, correlation 0.7",
      "Cauchy-Schwarz mutual information, independent block",
      "Rényi-2 conditional entropy at $x_2 = 0$",
      "Rényi-2 conditional entropy at $x_2 = 1$",
      "Rényi-2 conditional entropy at $x_2 = 2$"
    ),
    value = round(c(mi, mi_independent, h_cond), 6L)
  ),
  col.names = c("Quantity", "Value (nats)"),
  caption = paste(
    "Dependence between the two coordinates, and the predictive",
    "uncertainty of the first given the second."
  )
)
```

### Cooling the fit: annealing and phase transitions

The objective of the EM algorithm can be written as a free energy
$F = \langle E \rangle - T H$, a quantity borrowed from statistical
mechanics. Varying the temperature $T$ in it gives an annealed fit. The E-step responsibilities are
softened by $T$, so that
$\gamma_{ik} \propto \pi_k\, \mathcal{N}(x_i;\ \mu_k, \Sigma_k)^{1/T}$, and
$T$ is cooled from a high value towards one. At high $T$ the objective has
a single minimum, with every component at the mean of the data. As $T$
falls, the components split apart at critical temperatures. Setting
`anneal = TRUE` on `fit_em_samples()` or `fit_kld_em()` starts the usual EM
fit from the result of this cooling.

```{r anneal-fit}
x_three <- rbind(
  matrix(rnorm(200L), ncol = 2L) +
    matrix(rep(c(-7, -7), each = 100L), ncol = 2L),
  matrix(rnorm(200L), ncol = 2L) +
    matrix(rep(c(7, -7), each = 100L), ncol = 2L),
  matrix(rnorm(200L), ncol = 2L) +
    matrix(rep(c(0, 8), each = 100L), ncol = 2L)
)
tgt_three <- gmm_target_from_samples(x_three)
fit_annealed <- fit_em_samples(tgt_three, N = 3L, anneal = TRUE, seed = 1L)
fit_annealed@diagnostics$annealed
```

Cooling also discovers the component count. `gmm_anneal_path()` tracks the
number of distinct centroids as the temperature falls. The discovered
number of components is the count that spans the widest range of log
temperature, summed over the steps at which it is recorded; on the
geometric cooling schedule this is the count recorded at the most steps. The
first bifurcation has the closed-form critical temperature
$T_c = \lambda_{\max}(\Sigma^{-1} C)$, returned as `t_critical_analytic`
for an independent check. Here $C$ is the covariance of the data and
$\Sigma = \sigma^2 I$ the covariance shared by the annealed components. The
default $\sigma = 1$ makes $T_c$ the largest eigenvalue of $C$.

```{r anneal-path}
path <- gmm_anneal_path(x_three, k_max = 6L, n_steps = 60L, seed = 1L)
k_found <- path$k_selected
t_empirical <- path$first_critical_temperature
t_analytic <- path$t_critical_analytic
t_ratio <- t_empirical / t_analytic
```

```{r anneal-stair, include = FALSE}
## runs of equal centroid count along the cooling schedule, hot to cold
run_len <- rle(path$path$n_effective)
run_end <- cumsum(run_len$lengths)
stair <- data.frame(
  count = run_len$values, steps = run_len$lengths,
  first = path$path$temperature[run_end - run_len$lengths + 1L],
  last = path$path$temperature[run_end]
)
t1 <- function(v) formatC(v, format = "f", digits = 1L)
stretch <- function(i) {
  if (stair$steps[i] == 1L) {
    return(paste(stair$count[i], "at the single step", t1(stair$first[i])))
  }
  paste(stair$count[i], "over the", stair$steps[i], "steps from",
        t1(stair$first[i]), "to", t1(stair$last[i]))
}
n_stair <- nrow(stair)
stair$range <- stair$first - stair$last
widest_linear <- which.max(stair$range)
## with every centroid at the data mean the free energy is the mean energy,
## half the trace of the covariance (divisor n) plus (p / 2) log(2 pi)
n_three <- nrow(x_three)
fe_hot <- 0.5 * sum(diag(cov(x_three))) * (n_three - 1) / n_three +
  ncol(x_three) / 2 * log(2 * pi)
stair_middle <- paste(vapply(seq_len(n_stair)[-c(1L, n_stair)], stretch,
                             character(1L)), collapse = ", then ")
```

```{r anneal-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "components discovered", "first critical temperature, empirical",
      "first critical temperature, analytic", "ratio empirical to analytic",
      "cooling steps recorded"
    ),
    value = c(
      formatC(k_found, format = "d"),
      formatC(c(t_empirical, t_analytic), format = "f", digits = 2L),
      formatC(t_ratio, format = "f", digits = 3L),
      formatC(nrow(path$path), format = "d")
    )
  ),
  col.names = c("Quantity", "Value"), align = c("l", "r"),
  caption = paste(
    "What the cooling schedule discovered on three well-separated",
    "clusters."
  )
)
```

The table gives two numbers where the cooling gives a whole curve. The
figure below is the curve: the number of distinct centroids against
temperature on the upper panel, the free energy on the lower, with the
empirical and analytic first critical temperatures marked. The free energy
recorded at each temperature is
$F(T) = -T\, n^{-1} \sum_i \log \sum_k \pi_k\,
\mathcal{N}(x_i;\ \mu_k, \Sigma)^{1/T}$, with the shared $\Sigma$ above.
It equals $\langle E \rangle - T H$, where
$E_{ik} = -\log \mathcal{N}(x_i;\ \mu_k, \Sigma)$ is averaged over the
soft assignments $\gamma_{ik}$, and
$H = -n^{-1} \sum_{i,k} \gamma_{ik} \log(\gamma_{ik} / \pi_k)$ is the
entropy of the assignments relative to the component weights, zero when
every point is assigned in proportion to the weights.

```{r fig-anneal, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4.4, fig.pos = "!ht", fig.cap = "Deterministic annealing on three well-separated clusters. The upper panel is the staircase of distinct centroids as the temperature falls, the lower panel the deterministic-annealing free energy defined in the text, over the same schedule. The dashed rule is the empirical first critical temperature, where the count first rises above one, and the dotted rule the closed-form analytic one.", fig.alt = "Two stacked panels against a logarithmic temperature axis running from hot to cold: the upper panel is a rising staircase of the number of distinct centroids, the lower panel a free-energy curve that is flat at the hot end and then descends, both annotated with two nearly coincident vertical rules marking the first critical temperature."}
anneal_df <- rbind(
  data.frame(
    temperature = path$path$temperature,
    value = path$path$n_effective,
    panel = "distinct centroids"
  ),
  data.frame(
    temperature = path$path$temperature,
    value = path$path$free_energy,
    panel = "variational free energy"
  )
)
ggplot2::ggplot(anneal_df, ggplot2::aes(temperature, value)) +
  ggplot2::geom_vline(
    xintercept = t_analytic, linetype = "dotted",
    colour = "#0072B2", linewidth = 0.7
  ) +
  ggplot2::geom_vline(
    xintercept = t_empirical, linetype = "dashed",
    colour = "#D55E00", linewidth = 0.7
  ) +
  # a count recorded at one temperature holds until the next, cooler step
  ggplot2::geom_step(direction = "vh", linewidth = 0.8,
                     colour = "#000000") +
  ggplot2::facet_wrap(~ panel, ncol = 1L, scales = "free_y") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(
    x = "temperature (log scale, cooling to the left)",
    y = NULL,
    title = "Where the components bifurcate as the fit cools"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-anneal-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the annealing-path figure",
    "is skipped.\n")
```

### Maximum-entropy targets

Entropy can also be used to build a target. Among all densities that meet a
set of constraints, `maxent_target()` builds the one with the largest
entropy: the Gaussian under first- and second-moment
constraints on the full space, the uniform under a support constraint alone,
and a truncated Gaussian under second moments on a box. The bounded cases
record their support. Regime (iii) of van der Hoek and Elliott (2024) fits
a target from a density that can only be evaluated (`regime = "kld"`). For
these targets it draws its trial points from a distribution chosen
automatically to cover that support.

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.12\textheight\newpage\fi
```

```{r maxent}
me_gauss <- maxent_target(moments = list(mean = c(0, 0), cov = diag(2)))
me_unif <- maxent_target(support = list(lower = c(0, 0), upper = c(1, 1)))
unif_density <- exp(me_unif@log_density(matrix(c(0.5, 0.5), nrow = 1L)))
gauss_density <- exp(me_gauss@log_density(matrix(c(0, 0), nrow = 1L)))
```

```{r maxent-kable, echo = FALSE}
knitr::kable(
  data.frame(
    constraint = c("mean and covariance", "unit-square support"),
    family = c(me_gauss@metadata$family, me_unif@metadata$family),
    check = round(c(gauss_density, unif_density), 6L)
  ),
  col.names = c("Constraint supplied", "Maximum-entropy family",
                "Density at the centre"),
  caption = paste(
    "The least-committal density under each constraint set, at the",
    "centre of its constraint. The standard bivariate Gaussian has density",
    "$1/(2\\pi)$ at its mean; the unit square has area one, so its uniform",
    "density is one."
  )
)
```

### Selecting a component count with the ICL

`bic_aic()` reports the integrated completed likelihood (Biernacki et al.,
2000) alongside the BIC and AIC. The ICL adds to the BIC twice the entropy
of the fitted
classification, $\mathrm{ICL} = \mathrm{BIC} + 2 E_N$ with
$E_N = -\sum_{i,k} \gamma_{ik} \log \gamma_{ik} \geq 0$, so it penalises
mixtures whose components overlap and favours well-separated solutions. It
equals the BIC for a single component and is never smaller.

```{r icl}
x_two <- rbind(
  matrix(rnorm(200L, -4), ncol = 2L),
  matrix(rnorm(200L, 4), ncol = 2L)
)
fit_two <- fit_em_samples(gmm_target_from_samples(x_two), N = 2L, seed = 1L)
crit <- bic_aic(fit_two)
icl_gap <- crit$icl - crit$bic
```

```{r sci-notation, include = FALSE}
## A number as LaTeX math a \times 10^{b}, for values far from one.
sci <- function(v, digits) {
  e <- floor(log10(abs(v)))
  paste0("$", formatC(v / 10^e, format = "f", digits = digits),
         " \\times 10^{", e, "}$")
}
```

```{r icl-kable, echo = FALSE}
knitr::kable(
  data.frame(
    criterion = c("BIC", "AIC", "ICL", "classification entropy",
                  "free parameters"),
    value = c(
      formatC(crit$bic, format = "f", digits = 2L),
      formatC(crit$aic, format = "f", digits = 2L),
      formatC(crit$icl, format = "f", digits = 2L),
      sci(crit$classification_entropy, 1L),
      formatC(crit$n_params, format = "d")
    )
  ),
  col.names = c("Criterion", "Value"), align = c("l", "r"),
  caption = paste(
    "Information criteria for a two-component fit to two well-separated",
    "clusters."
  )
)
```

### Dependency structure of a mixture

`gmm_independence_graph()` reads the partial-correlation (Gaussian
graphical model) structure of the overall covariance of a fitted mixture.
For a mixture fitted to data, it tests the partial correlation of
coordinates $i$ and $j$, given all the others, with Fisher's $z$ at level
0.05, taking the sample size from the fit. The two mixtures in this section
have no sample size, so they are read with `threshold = 0.05`, which draws
an edge $i - j$ when that partial correlation exceeds 0.05 in absolute
value. The overall covariance of a
mixture is closed-form in its parameters, so no sampling is needed. A
tridiagonal precision matrix is a chain, and the recovered graph should be
that chain.

```{r graph-chain}
omega <- diag(4L)
for (i1 in seq_len(3L)) {
  omega[i1, i1 + 1L] <- -0.5
  omega[i1 + 1L, i1] <- -0.5
} # ends i1, over the off-diagonal band of the chain precision
g_chain <- gmm(
  weights = 1, means = list(rep(0, 4L)), covariances = list(solve(omega))
)
adj_chain <- gmm_independence_graph(g_chain, threshold = 0.05)
edges_chain <- sum(adj_chain) / 2L
```

```{r graph-chain-kable, echo = FALSE}
knitr::kable(
  as.data.frame(adj_chain[, ]),
  caption = paste(
    "Recovered adjacency of a chain precision matrix: an edge between",
    "neighbouring coordinates and nowhere else."
  )
)
```

The main use of the graph is with regime (iii). Composed with `fit_kld_em()`, the
diagnostic can recover the dependency structure of a target you can only
*evaluate* -- an unnormalised energy density -- where no sample exists to
drive a conventional structure estimator. Whether it does depends on the
quality of the fit, which the example below shows. The energy of the continuous
field below couples $x_1$ with $x_2$ and $x_2$ with $x_3$ and has no term in
$x_1 x_3$, so $x_1$ and $x_3$ are conditionally independent given $x_2$.

```{r graph-regime-iii}
energy <- function(x_mat) {
  x_mat <- matrix(x_mat, ncol = 3L)
  rowSums((x_mat^2 - 1)^2) -
    0.7 * (x_mat[, 1L] * x_mat[, 2L] + x_mat[, 2L] * x_mat[, 3L])
}
field <- gmm_target(n_dim = 3L,
                    log_density = function(x_mat) -energy(x_mat))

fit_field <- fit_kld_em(
  field,
  N = 8L,
  proposal = proposal_uniform(3L, -3, 3),
  is_size = 6000L,
  anneal = TRUE,
  seed = 1L,
  support_warn = FALSE
)
adj_field <- gmm_independence_graph(fit_field, threshold = 0.05)
edges_field <- sum(adj_field) / 2L
round(attr(adj_field, "pcor"), 3L)
```

Conditional independence does not force a zero partial correlation in
general, so the target's own partial correlations are computed by quadrature
on a grid over $[-3, 3]^3$.

```{r graph-target-pcor}
g_axis <- seq(-3, 3, length.out = 61L)
g_pts <- as.matrix(expand.grid(g_axis, g_axis, g_axis))
g_w <- exp(-energy(g_pts))
g_w <- g_w / sum(g_w)
g_mean <- colSums(g_pts * g_w)
prec_exact <- solve(crossprod(g_pts * sqrt(g_w)) - tcrossprod(g_mean))
pcor_exact <- -prec_exact / tcrossprod(sqrt(diag(prec_exact)))
diag(pcor_exact) <- 1
x_names <- paste0("x", seq_len(3L))
dimnames(pcor_exact) <- list(x_names, x_names)
# mass on grid points more than 2.5 from the origin in any coordinate
outer_mass <- sum(g_w[apply(abs(g_pts), 1L, max) > 2.5])
round(pcor_exact, 4L)
```

The grid points beyond 2.5 in any coordinate carry
`r sci(outer_mass, 0L)` of the grid's mass, so the grid covers the target.
The target's partial correlation of $x_1$ and $x_3$ is
`r fixed(pcor_exact[1L, 3L], 4)` and that of each neighbouring pair
`r fixed(pcor_exact[1L, 2L], 3)`, so the graph of the target at the
threshold of 0.05 is the two-edge chain $x_1 - x_2 - x_3$.

```{r graph-regime-iii-facts, include = FALSE}
q_field <- gmm_fit_quality(fit_field)
pcor_field <- attr(adj_field, "pcor")
## the prose quotes the held-out KL, so the fit must have a validation sample
stopifnot(is.finite(q_field$heldout_kld))
field_flagged <- isTRUE(q_field$degenerate) || isFALSE(q_field$converged) ||
  q_field$heldout_kld > 0.3
field_has_13 <- adj_field[1L, 3L] == 1L
field_is_chain <- edges_field == 2L && !field_has_13
field_edges <- which(adj_field == 1L & upper.tri(adj_field), arr.ind = TRUE)
field_edge_list <- paste0("$x_", field_edges[, 1L], " - x_", field_edges[, 2L],
                          "$")
```

```{r graph-regime-iii-kable, echo = FALSE}
knitr::kable(
  as.data.frame(adj_field[, ]),
  caption = paste(
    "Recovered adjacency of a three-coordinate energy field fitted in",
    "regime (iii), where only the density can be evaluated. The target's",
    "graph is the chain with edges $x_1 - x_2$ and $x_2 - x_3$."
  )
)
```

## Interpretation

The closed-form Rényi-2 entropy of the single Gaussian and its analytic
formula `r if (abs(h2_gap) < 1e-12) "agree to machine precision." else paste("differ by", formatC(h2_gap, format = "g", digits = 2), "nats.")` The
two-component mixture carries a higher entropy,
`r round(h2_mixture, 3L)` nats against
`r round(h2_closed, 3L)` for the single Gaussian, because its mass is
spread across two separated modes rather than one.

The Shannon estimate is `r round(sh$mc, 3L)` nats with a standard error of
`r round(sh$mc_se, 4L)`, and the analytic bound sits
`r round(sh_slack, 3L)` nats above it, a gap of
`r round(sh_slack_se, 1L)` standard errors. The estimate lies below the
bound, as it must.

The two divergences answer different questions about the same pair. The
Cauchy-Schwarz divergence is `r round(d_cs, 4L)` nats and is exact. Against
itself the mixture scores `r formatC(d_self, format = "g", digits = 2)`, as
the definition requires. The Monte Carlo
Kullback-Leibler estimate is `r round(d_kl$mc, 3L)` nats with a standard
error of
`r round(d_kl$mc_se, 4L)`, and the deterministic approximation of Hershey
and Olsen (2007) returned beside it is `r fixed(d_kl$variational, 3)`,
`r round(d_kl_gap, 3L)` nats below the estimate. The Cauchy-Schwarz and
Kullback-Leibler divergences are not on a common scale and should not be
compared to each other, only across pairs of fits under one definition.

The Cauchy-Schwarz mutual information reads `r round(mi, 4L)` nats between
two coordinates correlated at 0.7, and
`r formatC(mi_independent, format = "g", digits = 2)` between two
independent ones, as the definition requires. The Rényi-2
conditional entropy is `r round(h_cond[1L], 4L)` nats at every point of the grid,
identical across the three conditioning values because a single-component
joint has a conditional whose spread does not depend on what is conditioned
on. For a multi-component joint the value can vary across the grid.

Cooling discovered `r k_found` components, which is the number of clusters
the data were built from. The first bifurcation was recorded at temperature
`r round(t_empirical, 1L)` against the closed-form
$T_c = \lambda_{\max}(\Sigma^{-1} C) =$ `r round(t_analytic, 1L)`, a ratio
of `r round(t_ratio, 2L)`. The two rules in the figure sit close together
but not on top of each other: the empirical value is the first temperature
on a `r nrow(path$path)`-step grid at which the split was *detected*, so it
lags the temperature at which the split becomes available. A split is
detected only once two centroids are further apart than the merge tolerance,
a tenth of the square root of the mean coordinate variance of the data, so
the lag depends on that tolerance and on the
data as well as on the grid. The staircase
in the upper panel is the discovery itself. Read from the hot end, it
holds `r if (stair$count[1L] == 1) "one centroid" else paste(stair$count[1L], "centroids")` down to temperature
`r t1(stair$last[1L])`, then `r stair_middle`, and then
`r stair$count[n_stair]` from `r t1(stair$first[n_stair])` to the cold end,
over `r stair$steps[n_stair]` steps.
`r if (stair$count[n_stair] == k_found) paste0("The last stretch occupies the most cooling steps, and so the widest range of log temperature, and gives the count returned, ", k_found, ".") else paste0("The count returned, ", k_found, ", is the one recorded at the most cooling steps, and so over the widest range of log temperature.")`
`r if (stair$count[widest_linear] != k_found) paste0("On a linear temperature scale the widest stretch is instead the ", if (stair$count[widest_linear] == 1) "one-centroid" else paste0(stair$count[widest_linear], "-centroid"), " one, from ", t1(stair$first[widest_linear]), " to ", t1(stair$last[widest_linear]), ".") else ""`
The lower panel is flat above the first critical temperature. There every
centroid sits at the data mean, each point is assigned in proportion to the
weights, $H$ is zero, and $F$ is the mean energy
$\tfrac{1}{2}\,\mathrm{tr}\, C + \tfrac{p}{2} \log(2\pi) =$
`r fixed(fe_hot, 3)`, against `r fixed(path$path$free_energy[1L], 3)`
recorded at the hottest step. Below it the free energy falls as the
centroids separate.

The maximum-entropy constructions return the families the theory names: the
Gaussian under moment constraints, the uniform under a support constraint,
whose density at the centre of the unit square is
`r round(unif_density, 3L)`, the reciprocal of an area of one. The Gaussian
has density `r round(gauss_density, 6L)` at its mean, the $1/(2\pi)$ of a
standard bivariate normal.

For the two-cluster fit the BIC is `r round(crit$bic, 1L)` and the ICL is
`r round(crit$icl, 1L)`, identical to the precision reported. The
classification entropy of `r sci(crit$classification_entropy, 1L)` is
floating-point noise: two clusters this well separated leave no ambiguity
about which component a point belongs to, so the ICL has nothing to
penalise. On overlapping components the two criteria separate, and the ICL
prefers the smaller count.

The analytic case recovers `r edges_chain` edges from a tridiagonal
precision matrix, the chain it was built from. The regime (iii) fit to the
energy field returns `r edges_field` edges,
`r paste(field_edge_list, collapse = ", ")`,
`r if (field_is_chain) "the two-edge chain of the target" else "not the two-edge chain of the target"`.
The partial correlation of $x_1$ and $x_3$ is
`r fixed(pcor_field[1L, 3L], 3)` in the fitted mixture, against
`r fixed(pcor_exact[1L, 3L], 4)` in the target, and the threshold is 0.05.
The fit's quality certificate is
`r if (field_flagged) "flagged, and \x60gmm_independence_graph()\x60 printed a notice saying so" else "not flagged"`.
The fit
`r if (isTRUE(q_field$converged)) "converged" else "did not converge"` and
`r if (isTRUE(q_field$degenerate)) "is" else "is not"` degenerate. Its
KL divergence, estimated on a validation sample of fresh draws that the fit
did not use, is `r fixed(q_field$heldout_kld, 3)` nats. The package flags a
fit that did not converge, is degenerate, or has a held-out KL divergence
above 0.3. Half the weighted variance of the log ratio of target to proxy
over the fitting draws is `r fixed(q_field$kld_approx, 3)` nats. It
approximates the KL divergence only near a good fit. The package uses it for
the flag only when a fit has no validation sample. The
fit's effective sample size is about `r round(q_field$ess)` of
`r fit_field@diagnostics$is_size` draws. The graph read off this fit rests
on that importance sample.

Taken together, the opening question has three answers. Spread is the
Rényi-2 entropy, exact and cheap. Distance between two fits is the
Cauchy-Schwarz divergence, also exact, with the Kullback-Leibler estimate
available when an analysis requires that definition. The number of
components is read from the annealing staircase, which recovered the
`r k_found` clusters the data were built from. The ICL was computed on a
separate two-cluster dataset, not on these data.

## Limitations

The Rényi-2 entropy and the Cauchy-Schwarz divergence are exact and cheap,
costing $O(K^2)$ Gaussian evaluations, and are the recommended defaults for
reporting the spread of a fitted mixture and the discrepancy between two
fits. The Shannon entropy and the Kullback-Leibler divergence are the more
familiar quantities but are Monte Carlo estimates, so they carry the
standard error the tables report and should be read against it rather than
to the last digit printed. Use them when an analysis requires those
specific definitions, not by default.

The independence graph is a graphical-model diagnostic, not a
causal-discovery method. It returns the partial-correlation graph of the
mixture's overall covariance, without edge directions. Being second order,
it sees only dependence that enters the covariance. A pair of coordinates coupled purely through a higher moment, or
through the choice of mixture component in a way that leaves the overall
covariance unchanged, will appear as an absent edge. Fisher's $z$ test assumes that the data rows are
Gaussian. A one-component fit to non-Gaussian data carries that assumption
too, since the variance of $\operatorname{atanh}(r)$ depends on the fourth
moments of the data. The simulation below compares the rate at which the test
recovers a known graph with the rate its level predicts. The level of the test, or the
threshold where no sample size is known, is a choice, and a graph read at
one level is not the graph read at another.

The annealing path discovers a component count on data whose clusters are
well separated, which is the case where every method agrees. The staircase
becomes hard to read when clusters overlap, the widest-plateau rule can then
select a count no criterion would endorse, and the analytic critical
temperature governs only the *first* bifurcation, not the later ones. The
ICL is a criterion, not a test: it ranks the counts that were fitted and offers no
significance statement about the winner.

Everything in this vignette is read off a mixture that has already been
fitted.
Nothing here checks that the mixture is a good proxy for the target it was
fitted to. That is the job of the importance-sampling diagnostics and
`gmm_fit_quality()`. An entropy computed on a degenerate fit can be precise
and still wrong. The checks above compare each quantity with its
own analytic value on mixtures built by hand. The numerical illustration
that follows scores the same quantities, read from mixtures fitted to data,
against quadrature on the true density, and places them beside
nearest-neighbour estimators, mclust's criteria and three graph estimators.

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_fnn <- requireNamespace("FNN", quietly = TRUE)
has_glasso <- requireNamespace("glasso", quietly = TRUE)
has_huge <- requireNamespace("huge", quietly = TRUE)
has_pcalg <- requireNamespace("pcalg", quietly = TRUE)
has_penguins <- exists("penguins", envir = asNamespace("datasets"))
run_illustration <- has_fnn && has_glasso && has_huge && has_pcalg &&
  has_penguins
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("FNN, glasso, huge, pcalg or the penguins data (R 4.5.0 or later) is",
    "not available on this build, so the real-data illustration is",
    "skipped.\n")
```

### Dependence in the Palmer penguins data, read three ways

We read the dependence between the four body measurements of the Palmer
penguins (Gorman et al., 2014), shipped with R as `penguins`
since version 4.5.0 (Horst et al., 2022). The mutual
information between each pair of measurements comes from a mixture fitted to
the whole colony, beside the nearest-neighbour estimator of Kraskov et al.
(2004). The independence graph within each species
comes from a mixture fitted to that species, beside the graphical lasso
(Friedman et al., 2008), the neighbourhood-selection
estimator of Meinshausen and Bühlmann (2006) in huge (Zhao et al., 2012),
and the PC algorithm in pcalg (Kalisch et al., 2012). Real data carry no
reference value for either quantity, so this section places the estimates
side by side and the simulation that follows scores them.

The competitors install from CRAN once and load as usual. pcalg also needs
the graph and RBGL packages, which install from Bioconductor through
BiocManager:

```{r install-competitors, eval = FALSE}
install.packages("BiocManager")
BiocManager::install(c("graph", "RBGL"))
install.packages(c("mclust", "FNN", "glasso", "huge", "pcalg", "cubature"))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(FNN)
library(glasso)
library(huge)
library(pcalg)
```

FNN estimates entropy by the nearest-neighbour rule of Kozachenko and
Leonenko (1987), and its `mutinfo()` implements the first mutual-information
estimator of Kraskov et al. (2004). The simulation below uses the same
estimator written out in R, shown here and checked against `mutinfo()` on
the penguins. The distance to
the $k$-th nearest neighbour in the joint space, in the maximum norm, sets a
radius for each point, and the number of neighbours within that radius in
each marginal gives the estimate. Ten neighbours are used throughout, the
default of `FNN::entropy()`.

```{r ksg-code, eval = run_illustration}
ksg_mi <- function(xa, xb, k) {
  da <- as.matrix(dist(xa, method = "maximum"))
  db <- as.matrix(dist(xb, method = "maximum"))
  dj <- pmax(da, db)
  n <- nrow(dj)
  # the point itself is the first of the sorted distances
  eps <- apply(dj, 1L, function(v) sort(v)[k + 1L])
  n_a <- rowSums(da < eps) - 1L
  n_b <- rowSums(db < eps) - 1L
  digamma(k) + digamma(n) - mean(digamma(n_a + 1L) + digamma(n_b + 1L))
}
```

The birds measured in full are kept, and each measurement is centred and
scaled to unit variance, since the nearest-neighbour estimators depend on
the units and the mixture-based quantities do not. `select_N()` chooses the
number of components by BIC among one to five, and the mutual information
between each pair is read from that fit in two forms: the closed-form
Cauchy-Schwarz form of `gmm_mutual_information()`, and the Shannon form as
the entropy of each marginal less the entropy of the pair, each a Monte Carlo
estimate from `gmm_entropy()` on the marginalised fit.

```{r penguins-data, eval = run_illustration}
cols <- c("bill_len", "bill_dep", "flipper_len", "body_mass")
keep <- complete.cases(penguins[, cols])
peng <- scale(as.matrix(penguins[keep, cols]))
species <- penguins$species[keep]

sel_colony <- select_N(gmm_target_from_samples(peng), candidates = 1:5,
                       seed = 1L)
fit_colony <- sel_colony$best_fit
```

```{r penguins-mi, eval = run_illustration}
h_mc <- function(g) gmm_entropy(g, order = "shannon", n_mc = 5000L,
                                seed = 1L)$mc
pairs <- combn(4L, 2L)
mi_tab <- t(apply(pairs, 2L, function(ij) {
  a <- ij[1L]
  b <- ij[2L]
  c(cs = gmm_mutual_information(fit_colony, a, b),
    shannon = h_mc(gmm_marginalise(fit_colony, a)) +
      h_mc(gmm_marginalise(fit_colony, b)) -
      h_mc(gmm_marginalise(fit_colony, ij)),
    ksg = ksg_mi(peng[, a, drop = FALSE], peng[, b, drop = FALSE], 10L))
}))
```

```{r ksg-check, eval = run_illustration}
ksg_fnn <- apply(pairs, 2L, function(ij) {
  mutinfo(peng[, ij[1L]], peng[, ij[2L]], k = 10L)
})
ksg_gap <- max(abs(mi_tab[, "ksg"] - ksg_fnn))
ksg_gap
```

```{r penguins-mi-table, eval = run_illustration, echo = FALSE}
short <- c(bill_len = "BL", bill_dep = "BD", flipper_len = "FL",
           body_mass = "BM")
mi_df <- data.frame(
  pair = apply(pairs, 2L, function(ij) paste(short[cols[ij]],
                                             collapse = " and ")),
  correlation = cor(peng)[t(pairs)],
  mi_tab,
  stringsAsFactors = FALSE
)
long <- c(bill_len = "bill length", bill_dep = "bill depth",
          flipper_len = "flipper length", body_mass = "body mass")
pair_long <- apply(pairs, 2L, function(ij) paste(long[cols[ij]],
                                                 collapse = " and "))
top_pair <- function(what) pair_long[which.max(mi_df[[what]])]
rank_of <- function(pair, what) {
  rank(-mi_df[[what]])[pair_long == pair]
}
ordinal <- c("first", "second", "third", "fourth", "fifth", "sixth")
top_shannon_text <- if (top_pair("shannon") == top_pair("ksg")) {
  paste0("The most dependent pair on both Shannon readings is ",
         top_pair("shannon"), ".")
} else {
  paste0("The most dependent pair is ", top_pair("shannon"),
         " on the Shannon form read from the mixture and ", top_pair("ksg"),
         " on the nearest-neighbour estimate.")
}
## pairs whose rank differs between the three columns
rank_mat <- cbind(rank(-mi_df$cs), rank(-mi_df$shannon), rank(-mi_df$ksg))
moved <- apply(pairs[, apply(rank_mat, 1L, function(r) {
  length(unique(r)) > 1L
}), drop = FALSE], 2L, function(ij) paste(short[cols[ij]], collapse = "-"))
count_word <- c("one", "two", "three", "four", "five", "six")
moved_text <- if (length(moved) == 0L) {
  "The three columns agree on the order of every pair."
} else if (length(moved) == 1L) {
  paste0("Only ", moved, " changes rank between the three columns.")
} else {
  paste0("Only ", count_word[length(moved)], " pairs change rank between ",
         "the three columns: ",
         paste(moved[-length(moved)], collapse = ", "), " and ",
         moved[length(moved)], ".")
}
knitr::kable(
  mi_df, digits = 3L, row.names = FALSE,
  col.names = c("Pair of measurements", "Correlation",
                "Cauchy-Schwarz, mixture", "Shannon, mixture",
                "Shannon, KSG"),
  caption = paste0(
    "Mutual information in nats between each pair of standardised ",
    "measurements over the ", nrow(peng), " complete birds: two forms ",
    "read from the ", sel_colony$best_n, "-component mixture fitted to ",
    "the colony, and the Kraskov et al. estimate ",
    "with ten neighbours; BL bill length, BD bill depth, FL flipper ",
    "length, BM body mass. No reference value exists for real data."
  )
)
```

Within each species the same four measurements are read for their
second-order conditional-independence structure. The mixture is selected by
BIC among one to three components, and its graph tests each partial
correlation with Fisher's $z$ at level 0.05, the default for a mixture
fitted to data. The graphical lasso has no default
penalty, so its penalty is chosen by BIC on a grid of thirty values; huge
runs its defaults, neighbourhood selection with the rotation information
criterion; the PC algorithm runs Fisher's $z$ test at level 0.01 and its
skeleton is kept.

```{r penguins-graph, eval = run_illustration}
glasso_bic <- function(x) {
  s <- cov(x)
  rhos <- exp(seq(log(1e-3), log(1), length.out = 30L))
  bics <- vapply(rhos, function(rho) {
    wi <- glasso(s, rho)$wi
    edges <- sum(abs(wi[upper.tri(wi)]) > 1e-8)
    loglik <- 0.5 * nrow(x) *
      (determinant(wi)$modulus[1L] - sum(diag(s %*% wi)))
    -2 * loglik + log(nrow(x)) * edges
  }, numeric(1L))
  wi <- glasso(s, rhos[which.min(bics)])$wi
  adj <- (abs(wi) > 1e-8) * 1L
  diag(adj) <- 0L
  adj
}

pc_skeleton <- function(x, alpha) {
  fit <- pc(suffStat = list(C = cor(x), n = nrow(x)),
            indepTest = gaussCItest, alpha = alpha, p = ncol(x))
  adj <- as(fit@graph, "matrix")
  ((adj + t(adj)) > 0) * 1L
}

four_graphs <- function(x, candidates) {
  sel <- select_N(gmm_target_from_samples(x), candidates = candidates,
                  seed = 1L)
  # huge's rotation criterion draws rows at random
  set.seed(1L)
  out <- list(
    proxymix = gmm_independence_graph(sel$best_fit),
    glasso = glasso_bic(x),
    huge = as.matrix(huge.select(huge(x, verbose = FALSE),
                                 verbose = FALSE)$refit),
    pcalg = pc_skeleton(x, alpha = 0.01)
  )
  attr(out, "n_components") <- sel$best_n
  out
}

graphs <- lapply(split(seq_len(nrow(peng)), species), function(rows) {
  four_graphs(peng[rows, ], candidates = 1:3)
})

# components chosen, and the largest partial correlation, within species
n_within <- vapply(graphs, attr, numeric(1L), "n_components")
pcor_within <- vapply(graphs, function(g) {
  p <- attr(g$proxymix, "pcor")
  max(abs(p[upper.tri(p)]))
}, numeric(1L))
data.frame(components = n_within, largest_pcor = round(pcor_within, 3L))
```

```{r penguins-graph-table, eval = run_illustration, echo = FALSE}
edge_string <- function(adj) {
  e <- which(adj == 1L & upper.tri(adj), arr.ind = TRUE)
  if (nrow(e) == 0L) return("none")
  if (nrow(e) == choose(ncol(adj), 2L)) return("all")
  paste(paste0(short[cols[e[, 1L]]], "-", short[cols[e[, 2L]]]),
        collapse = ", ")
}
graph_df <- do.call(rbind, lapply(names(graphs), function(sp) {
  data.frame(
    species = paste0(sp, " (", sum(species == sp), " birds)"),
    method = names(graphs[[sp]]),
    edges = vapply(graphs[[sp]], function(a) sum(a) / 2L, numeric(1L)),
    edge_set = vapply(graphs[[sp]], edge_string, character(1L)),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  graph_df, row.names = FALSE,
  col.names = c("Species", "Method", "Edges", "Edge set"),
  caption = paste0(
    "Undirected independence graph over the four measurements within each ",
    "species, by four methods; BL bill length, BD bill depth, FL flipper ",
    "length, BM body mass. Six edges is the complete graph, listed as all."
  )
)
n_edges <- function(sp, method) graph_df$edges[graph_df$species ==
  paste0(sp, " (", sum(species == sp), " birds)") & graph_df$method == method]
birds <- range(table(species))
```

Given only the four measurements, `select_N()` chose
`r sel_colony$best_n` components for the colony, the number of species.
Every pair carries dependence. `r top_shannon_text` The Cauchy-Schwarz
form puts `r top_pair("cs")` first and `r top_pair("shannon")`
`r ordinal[rank_of(top_pair("shannon"), "cs")]`. `r moved_text`
The Cauchy-Schwarz and Shannon forms are on different scales and are
not to be compared to each other. On the Shannon scale the mixture reads
`r fixed(min(mi_tab[, "ksg"] - mi_tab[, "shannon"]), 2)` to
`r fixed(max(mi_tab[, "ksg"] - mi_tab[, "shannon"]), 2)` nats below the
nearest-neighbour estimate on every pair. The hand-written estimator and
`FNN::mutinfo()`
`r if (ksg_gap < 1e-10) "agree to within $10^{-10}$ nats on all six pairs" else paste("differ by up to", fixed(ksg_gap, 4), "nats over the six pairs")`.
The simulation below scores both forms against a known value.
`r if (all(n_within == 1)) "Within species a single component was selected for all three, so the mixture graph is the Gaussian graphical model, and it returns" else paste0("Within species select_N() chose ", paste(n_within, collapse = ", "), " components for Adelie, Chinstrap and Gentoo, and the mixture graph returns")`
`r n_edges("Adelie", "proxymix")`, `r n_edges("Chinstrap", "proxymix")` and
`r n_edges("Gentoo", "proxymix")` of the six edges for Adelie, Chinstrap and
Gentoo. The graphical lasso returns `r n_edges("Adelie", "glasso")`,
`r n_edges("Chinstrap", "glasso")` and `r n_edges("Gentoo", "glasso")`, and
huge `r n_edges("Adelie", "huge")`, `r n_edges("Chinstrap", "huge")` and
`r n_edges("Gentoo", "huge")`.
The PC algorithm, which removes an edge as soon as the partial correlation
given any one of the conditioning sets it tests is not significant at the 1
per cent level, keeps
`r n_edges("Adelie", "pcalg")`, `r n_edges("Chinstrap", "pcalg")` and
`r n_edges("Gentoo", "pcalg")`. The mixture's graph tests each pair once,
given the other two measurements, at the 5 per cent level, on
`r birds[1L]` to `r birds[2L]` birds. The largest within-species
partial correlation is `r fixed(max(pcor_within), 2)`.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("entropy_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
```

```{r sim-box-mass, include = FALSE}
## the designs and their quadrature box, taken from the stored code
sim_exprs <- parse(text = sim$code)
sim_env <- new.env()
for (e in sim_exprs) {
  if (is.call(e) && identical(e[[1L]], as.name("<-")) && is.name(e[[2L]]) &&
      as.character(e[[2L]]) %in% c("mixture", "overall_box", "designs")) {
    eval(e, sim_env)
  }
}
## union bound: the mass outside the box is at most the sum over
## coordinates of the mass outside each marginal interval
box_missed <- max(vapply(sim_env$designs, function(m) {
  box <- sim_env$overall_box(m)
  sum(vapply(seq_len(m$p), function(j) {
    sum(vapply(seq_along(m$w), function(k) {
      s_jk <- sqrt(m$sig[[k]][j, j])
      m$w[k] * (pnorm(box$lower[j], m$mu[[k]][j], s_jk) +
                  pnorm(box$upper[j], m$mu[[k]][j], s_jk,
                        lower.tail = FALSE))
    }, numeric(1L)))
  }, numeric(1L)))
}, numeric(1L)))
## rounded up to one significant digit, since it is quoted as an upper bound
e_missed <- floor(log10(box_missed))
box_missed <- ceiling(box_missed / 10^e_missed) * 10^e_missed
```

The simulation draws `r sim$n_rep` datasets of `r sim$n` rows from each of
two mixtures whose entropies are known, fits a mixture with the true number
of components, and scores every estimate against adaptive quadrature on the
true density (`cubature::hcubature()`, Genz and Malik, 1980), over a box
reaching seven standard deviations of the mixture either side of its mean in
each coordinate. The box leaves out at most
`r sci(box_missed, 0L)` of the probability mass of either design, a bound
from the tails of the one-dimensional marginals. The first mixture has three
components in two dimensions with unequal covariances, two of them
correlated. The
second has two components in four dimensions, one with a first-order
autoregressive covariance and one with exchangeable correlation, and the
mutual information is between the first two coordinates and the last two.
Four quantities are scored: the Shannon entropy, from `gmm_entropy()` on the
fit with 5000 Monte Carlo draws and from `FNN::entropy()`; the Rényi-2
entropy, closed form on the fit, with no competitor; the Shannon mutual
information, assembled from three `gmm_entropy()` calls on the fit and from
the nearest-neighbour estimator above; and the Cauchy-Schwarz mutual
information, closed form on the fit. The reference for the last is the same
quadrature applied to the three integrals in its definition.

Each dataset also asks how many components it supports. proxymix fits one to
five components and takes the smallest BIC and the smallest ICL from
`bic_aic()`; mclust (Scrucca et al., 2016) takes the largest BIC over its
fourteen covariance models and the largest ICL from `mclustICL()`. The
reference is the number of components the data were drawn from.

The graph design has six coordinates with a chain precision matrix, shared
by two equally weighted components whose means differ by four units along
the first coordinate. Shifting one coordinate leaves the precision of the
overall covariance a chain, so the reference graph is the five-edge chain.
The four graph methods run as in the penguins section, with the mixture
selected among one to four components, and each recovered graph is scored by
its structural Hamming distance from the chain (Tsamardinos et al., 2006),
the number of edges to add or remove.

This chunk is complete and runs as shown, but it took about
`r round(sim$elapsed_secs / 60)` minutes on one core (R
`r sim$versions[["R"]]`, Apple silicon); the results below are
read from its stored output. For a quick reduced run, change the line
`n_rep <- 200L` to a small number such as `n_rep <- 3L`.

```{r sim-code, eval = FALSE}
library(proxymix)
library(mclust)
library(FNN)
library(glasso)
library(huge)
library(pcalg)
library(cubature)

n <- 500L      # rows in each dataset
n_rep <- 200L  # datasets per design
k_nn <- 10L    # neighbours in the nearest-neighbour estimators
k_max <- 5L    # largest component count offered to the selectors

# a mixture as plain lists with its own density and sampler, so the reference
# values and the data do not pass through the package
mixture <- function(w, mu, sig) {
  list(w = w, mu = mu, sig = sig, p = length(mu[[1L]]))
}

dmix <- function(x_mat, m) {
  p <- ncol(x_mat)
  rowSums(vapply(seq_along(m$w), function(k) {
    m$w[k] * exp(-0.5 * mahalanobis(x_mat, m$mu[[k]], m$sig[[k]]) -
                   0.5 * (determinant(m$sig[[k]])$modulus[1L] +
                            p * log(2 * pi)))
  }, numeric(nrow(x_mat))))
}

rmix <- function(n, m) {
  z <- sample.int(length(m$w), n, replace = TRUE, prob = m$w)
  x <- matrix(rnorm(n * m$p), n, m$p)
  for (k in seq_along(m$w)) {
    s1 <- z == k
    x[s1, ] <- x[s1, , drop = FALSE] %*% chol(m$sig[[k]]) +
      matrix(m$mu[[k]], sum(s1), m$p, byrow = TRUE)
  }
  x
}

marginal <- function(m, keep) {
  mixture(m$w, lapply(m$mu, function(v) v[keep]),
          lapply(m$sig, function(s) s[keep, keep, drop = FALSE]))
}

overall_box <- function(m) {
  mean_all <- Reduce(`+`, Map(`*`, m$w, m$mu))
  cov_all <- Reduce(`+`, Map(function(w, mu, s) w * (s + tcrossprod(mu)),
                             m$w, m$mu, m$sig)) - tcrossprod(mean_all)
  half <- 7 * sqrt(diag(cov_all))
  list(lower = mean_all - half, upper = mean_all + half)
}

# an integral of a function of the density over the box, by adaptive
# quadrature; the integrand receives points as columns
box_integral <- function(f, box) {
  hcubature(function(x_mat) matrix(f(t(x_mat)), 1L), box$lower, box$upper,
            tol = 1e-6, vectorInterface = TRUE)$integral
}

shannon_quad <- function(m) {
  box_integral(function(x) {
    d <- dmix(x, m)
    ifelse(d > 0, -d * log(d), 0)
  }, overall_box(m))
}

square_quad <- function(m) box_integral(function(x) dmix(x, m)^2,
                                        overall_box(m))

# Cauchy-Schwarz mutual information from the three integrals in its
# definition, each by quadrature
cs_mi_quad <- function(m, a, b) {
  ma <- marginal(m, a)
  mb <- marginal(m, b)
  cross <- box_integral(function(x) {
    dmix(x, m) * dmix(x[, a, drop = FALSE], ma) *
      dmix(x[, b, drop = FALSE], mb)
  }, overall_box(m))
  0.5 * log(square_quad(m)) + 0.5 * log(square_quad(ma) * square_quad(mb)) -
    log(cross)
}

designs <- list(
  "2-d, three components" = mixture(
    w = c(0.4, 0.35, 0.25),
    mu = list(c(-2, 0), c(2, 1), c(0, 3)),
    sig = list(matrix(c(1, 0.5, 0.5, 1), 2L),
               matrix(c(1.5, -0.6, -0.6, 0.8), 2L),
               diag(c(0.5, 1.2)))
  ),
  "4-d, two components" = mixture(
    w = c(0.6, 0.4),
    mu = list(c(0, 0, 0, 0), c(3, 0, 2, 0)),
    sig = list(0.5^abs(outer(1:4, 1:4, "-")),
               1.5 * (0.3 + 0.7 * diag(4L)))
  )
)
blocks <- list(list(1L, 2L), list(1:2, 3:4))

truth <- do.call(rbind, lapply(seq_along(designs), function(i1) {
  m <- designs[[i1]]
  a <- blocks[[i1]][[1L]]
  b <- blocks[[i1]][[2L]]
  h_joint <- shannon_quad(m)
  data.frame(
    design = names(designs)[i1],
    quantity = c("Shannon entropy", "Renyi-2 entropy",
                 "Shannon mutual information",
                 "Cauchy-Schwarz mutual information"),
    truth = c(h_joint, -log(square_quad(m)),
              shannon_quad(marginal(m, a)) + shannon_quad(marginal(m, b)) -
                h_joint,
              cs_mi_quad(m, a, b)),
    stringsAsFactors = FALSE
  )
}))

# Kraskov, Stoegbauer and Grassberger mutual information, first algorithm
ksg_mi <- function(xa, xb, k) {
  da <- as.matrix(dist(xa, method = "maximum"))
  db <- as.matrix(dist(xb, method = "maximum"))
  dj <- pmax(da, db)
  n <- nrow(dj)
  eps <- apply(dj, 1L, function(v) sort(v)[k + 1L])  # the point itself is v[1]
  n_a <- rowSums(da < eps) - 1L
  n_b <- rowSums(db < eps) - 1L
  digamma(k) + digamma(n) - mean(digamma(n_a + 1L) + digamma(n_b + 1L))
}

one_entropy <- function(i1, r) {
  set.seed(r)
  m <- designs[[i1]]
  a <- blocks[[i1]][[1L]]
  b <- blocks[[i1]][[2L]]
  x <- rmix(n, m)
  tgt <- gmm_target_from_samples(x)
  fit <- fit_em_samples(tgt, N = length(m$w), seed = r)
  h_mc <- function(g) gmm_entropy(g, order = "shannon", n_mc = 5000L,
                                  seed = r)$mc
  est <- data.frame(
    design = names(designs)[i1],
    rep = r,
    quantity = c("Shannon entropy", "Shannon entropy", "Renyi-2 entropy",
                 "Shannon mutual information", "Shannon mutual information",
                 "Cauchy-Schwarz mutual information"),
    method = c("proxymix", "FNN", "proxymix", "proxymix", "KSG", "proxymix"),
    estimate = c(
      h_mc(fit),
      entropy(x, k = k_nn)[k_nn],
      gmm_entropy(fit),
      h_mc(gmm_marginalise(fit, a)) + h_mc(gmm_marginalise(fit, b)) -
        h_mc(fit),
      ksg_mi(x[, a, drop = FALSE], x[, b, drop = FALSE], k_nn),
      gmm_mutual_information(fit, a, b)
    ),
    stringsAsFactors = FALSE
  )
  crit <- lapply(seq_len(k_max), function(k) {
    bic_aic(fit_em_samples(tgt, N = k, seed = r))
  })
  icl_m <- mclustICL(x, G = seq_len(k_max), verbose = FALSE)
  k_sel <- data.frame(
    design = names(designs)[i1],
    rep = r,
    selector = c("proxymix, BIC", "proxymix, ICL", "mclust, BIC",
                 "mclust, ICL"),
    chosen = c(which.min(vapply(crit, `[[`, numeric(1L), "bic")),
               which.min(vapply(crit, `[[`, numeric(1L), "icl")),
               Mclust(x, G = seq_len(k_max), verbose = FALSE)$G,
               which(icl_m == max(icl_m, na.rm = TRUE), arr.ind = TRUE)[1L]),
    stringsAsFactors = FALSE
  )
  list(est = est, k_sel = k_sel)
}

# the graph design: a chain precision shared by two components whose means
# differ along the first coordinate
omega <- diag(6L)
omega[abs(row(omega) - col(omega)) == 1L] <- -0.4
graph_design <- mixture(w = c(0.5, 0.5),
                        mu = list(rep(0, 6L), c(4, rep(0, 5L))),
                        sig = list(solve(omega), solve(omega)))
adj_chain <- (abs(omega) > 0) * 1L
diag(adj_chain) <- 0L

glasso_bic <- function(x) {
  s <- cov(x)
  rhos <- exp(seq(log(1e-3), log(1), length.out = 30L))
  bics <- vapply(rhos, function(rho) {
    wi <- glasso(s, rho)$wi
    edges <- sum(abs(wi[upper.tri(wi)]) > 1e-8)
    loglik <- 0.5 * nrow(x) *
      (determinant(wi)$modulus[1L] - sum(diag(s %*% wi)))
    -2 * loglik + log(nrow(x)) * edges
  }, numeric(1L))
  wi <- glasso(s, rhos[which.min(bics)])$wi
  adj <- (abs(wi) > 1e-8) * 1L
  diag(adj) <- 0L
  adj
}

pc_skeleton <- function(x, alpha) {
  fit <- pc(suffStat = list(C = cor(x), n = nrow(x)),
            indepTest = gaussCItest, alpha = alpha, p = ncol(x))
  adj <- as(fit@graph, "matrix")
  ((adj + t(adj)) > 0) * 1L
}

one_graph <- function(r) {
  set.seed(r)
  x <- rmix(n, graph_design)
  sel <- select_N(gmm_target_from_samples(x), candidates = 1:4, seed = r)
  adj <- list(
    proxymix = gmm_independence_graph(sel$best_fit),
    glasso = glasso_bic(x),
    huge = as.matrix(huge.select(huge(x, verbose = FALSE),
                                 verbose = FALSE)$refit),
    pcalg = pc_skeleton(x, alpha = 0.01)
  )
  data.frame(
    rep = r,
    method = names(adj),
    shd = vapply(adj, function(a) sum(abs(a - adj_chain)) / 2, numeric(1L)),
    edges = vapply(adj, function(a) sum(a) / 2, numeric(1L)),
    n_components = sel$best_n,
    stringsAsFactors = FALSE
  )
}

runs <- lapply(seq_along(designs), function(i1) {
  lapply(seq_len(n_rep), function(r) one_entropy(i1, r))
})
est <- do.call(rbind, lapply(runs, function(d) {
  do.call(rbind, lapply(d, `[[`, "est"))
}))
k_sel <- do.call(rbind, lapply(runs, function(d) {
  do.call(rbind, lapply(d, `[[`, "k_sel"))
}))
graph_res <- do.call(rbind, lapply(seq_len(n_rep), one_graph))

est <- merge(est, truth, by = c("design", "quantity"))
est$error <- est$estimate - est$truth
entropy_tab <- aggregate(
  cbind(bias = error, abs_error = abs(error)) ~ design + quantity + method,
  data = est, FUN = mean
)
entropy_tab$abs_error_se <- aggregate(
  abs(error) ~ design + quantity + method, data = est,
  FUN = function(v) sd(v) / sqrt(length(v))
)[[4L]]

k_sel$true_k <- vapply(designs[k_sel$design], function(m) length(m$w),
                       integer(1L))
k_tab <- aggregate(
  cbind(share_true = chosen == true_k, mean_chosen = chosen) ~
    design + selector,
  data = k_sel, FUN = mean
)

graph_tab <- aggregate(
  cbind(mean_shd = shd, share_exact = shd == 0, mean_edges = edges) ~ method,
  data = graph_res, FUN = mean
)
entropy_tab
k_tab
graph_tab
```

```{r sim-entropy-table, echo = FALSE}
tab_e <- merge(sim$entropy_tab, sim$truth, by = c("design", "quantity"))
quantity_order <- c("Shannon entropy", "Renyi-2 entropy",
                    "Shannon mutual information",
                    "Cauchy-Schwarz mutual information")
method_order <- c("proxymix", "FNN", "KSG")
tab_e <- tab_e[order(tab_e$design, match(tab_e$quantity, quantity_order),
                     match(tab_e$method, method_order)), ]
tab_e_show <- data.frame(
  design = sub(",.*$", "", tab_e$design),
  quantity = sub("Cauchy-Schwarz", "CS", sub("Renyi", "Rényi",
                 sub(" mutual information", " MI", tab_e$quantity))),
  truth = fixed(tab_e$truth, 3),
  method = tab_e$method,
  bias = fixed(tab_e$bias, 3),
  abs_error = fixed(tab_e$abs_error, 3),
  abs_error_se = fixed(tab_e$abs_error_se, 3),
  stringsAsFactors = FALSE
)
knitr::kable(
  tab_e_show, row.names = FALSE, align = c("l", "l", "r", "l", "r", "r", "r"),
  col.names = c("Design", "Quantity", "Quadrature", "Method", "Bias",
                "Mean abs. error", "SE"),
  caption = paste0(
    "Bias and mean absolute error of each estimate, in nats, against ",
    "quadrature on the true density, over ", sim$n_rep, " datasets of ",
    sim$n, " rows per design (2-d, three components; 4-d, two ",
    "components); MI is mutual information, CS Cauchy-Schwarz, and SE the ",
    "Monte Carlo standard error of the mean absolute error."
  )
)
e_value <- function(design, quantity, method, what) {
  s1 <- tab_e$design == design & tab_e$quantity == quantity &
    tab_e$method == method
  tab_e[[what]][s1]
}
d2 <- "2-d, three components"
d4 <- "4-d, two components"
# mixture minus nearest-neighbour absolute error, paired over the datasets
paired_diff <- function(design, quantity, rival) {
  s1 <- sim$est$design == design & sim$est$quantity == quantity
  a <- sim$est[s1 & sim$est$method == "proxymix", ]
  b <- sim$est[s1 & sim$est$method == rival, ]
  d <- abs(a$error[order(a$rep)]) - abs(b$error[order(b$rep)])
  c(diff = mean(d), se = sd(d) / sqrt(length(d)))
}
verdict <- function(design, quantity, rival) {
  pd <- paired_diff(design, quantity, rival)
  z <- pd[["diff"]] / pd[["se"]]
  if (abs(z) < 2) {
    return(paste0("within ", fixed(abs(z), 1), " paired standard errors of ",
                  "the nearest-neighbour estimator's, which does not resolve ",
                  "a difference"))
  }
  paste0(if (z < 0) "smaller" else "larger", " than the nearest-neighbour ",
         "estimator's by ", fixed(abs(pd[["diff"]]), 3), " nats, ",
         fixed(abs(z), 1), " paired standard errors")
}
cs_rel <- vapply(c(d2, d4), function(d) {
  e_value(d, "Cauchy-Schwarz mutual information", "proxymix", "abs_error") /
    e_value(d, "Cauchy-Schwarz mutual information", "proxymix", "truth")
}, numeric(1L))
```

```{r sim-k-table, echo = FALSE}
tab_k <- sim$k_tab
selector_order <- c("proxymix, BIC", "proxymix, ICL", "mclust, BIC",
                    "mclust, ICL")
tab_k <- tab_k[order(tab_k$design, match(tab_k$selector, selector_order)), ]
tab_k_show <- data.frame(
  design = sub(",.*$", "", tab_k$design),
  selector = tab_k$selector,
  share_true = fixed(tab_k$share_true, 2),
  mean_chosen = fixed(tab_k$mean_chosen, 2),
  stringsAsFactors = FALSE
)
knitr::kable(
  tab_k_show, row.names = FALSE, align = c("l", "l", "r", "r"),
  col.names = c("Design", "Selector", "Share choosing the true count",
                "Mean count chosen"),
  caption = paste0(
    "How often each selector chose the number of components the data were ",
    "drawn from (three in the 2-d design, two in the 4-d design), over ",
    sim$n_rep, " datasets per design, with the counts one to five offered."
  )
)
k_value <- function(design, selector, what) {
  tab_k[[what]][tab_k$design == design & tab_k$selector == selector]
}
# share of datasets on which a selector chose two components
share_two <- function(design, selector) {
  s1 <- sim$k_sel$design == design & sim$k_sel$selector == selector
  mean(sim$k_sel$chosen[s1] == 2L)
}
```

```{r sim-graph-table, echo = FALSE}
tab_g <- sim$graph_tab
graph_order <- c("proxymix", "glasso", "huge", "pcalg")
tab_g <- tab_g[order(match(tab_g$method, graph_order)), ]
knitr::kable(
  tab_g, digits = 3L, row.names = FALSE,
  col.names = c("Method", "Mean structural Hamming distance",
                "Share recovering the chain exactly", "Mean edges"),
  caption = paste0(
    "Recovery of the five-edge chain over six coordinates, over ",
    sim$n_rep, " datasets of ", sim$n, " rows, scored by the structural ",
    "Hamming distance from the true graph."
  )
)
g_value <- function(method, what) tab_g[[what]][tab_g$method == method]
# the six-coordinate chain has five edges
method_name <- c(proxymix = "proxymix's test", glasso = "the graphical lasso",
                 huge = "huge", pcalg = "the PC algorithm")
and_list <- function(v) {
  if (length(v) < 2L) return(v)
  paste(paste(v[-length(v)], collapse = ", "), "and", v[length(v)])
}
## mean edge count against five, with its Monte Carlo standard error
edge_stats <- t(vapply(tab_g$method, function(m) {
  e <- sim$graph_res$edges[sim$graph_res$method == m]
  c(mean = mean(e), se = sd(e) / sqrt(length(e)), five = mean(e == 5))
}, numeric(3L)))
z_edges <- (edge_stats[, "mean"] - 5) / edge_stats[, "se"]
z_edges[is.nan(z_edges)] <- 0
more_edges <- method_name[tab_g$method[z_edges > 2]]
fewer_edges <- method_name[tab_g$method[z_edges < -2]]
level_edges <- tab_g$method[abs(z_edges) <= 2]
graph_direction <- if (length(more_edges) > 0L && length(fewer_edges) > 0L) {
  paste0("On average ", and_list(more_edges), " returned more than the ",
         "chain's five edges, and ", and_list(fewer_edges), " fewer.")
} else if (length(more_edges) > 0L) {
  paste0("On average ", and_list(more_edges), " returned more than the ",
         "chain's five edges.")
} else if (length(fewer_edges) > 0L) {
  paste0("On average ", and_list(fewer_edges), " returned fewer than the ",
         "chain's five edges.")
} else {
  ""
}
level_text <- paste(vapply(level_edges, function(m) {
  paste0("For ", method_name[[m]], " the mean of ",
         fixed(edge_stats[m, "mean"], 3), " edges is within two Monte Carlo ",
         "standard errors of five, and exactly five edges were returned in ",
         fixed(edge_stats[m, "five"], 2), " of datasets.")
}, character(1L)), collapse = " ")
graph_direction <- paste(graph_direction, level_text)
## ten absent edges tested at 0.05 each: the chance none is drawn, were the
## tests independent
p_graph <- 6L
n_absent <- choose(p_graph, 2L) - 5L
share_no_false <- 0.95^n_absent
mcse_share <- sqrt(0.5 * 0.5 / sim$n_rep)
```

On the Shannon entropy both estimators read low. In the 2-d design the
mixture's Monte Carlo estimate has a bias of
`r fixed(e_value(d2, "Shannon entropy", "proxymix", "bias"), 3)` nats and a
mean absolute error of
`r fixed(e_value(d2, "Shannon entropy", "proxymix", "abs_error"), 3)`, the
nearest-neighbour estimate a bias of
`r fixed(e_value(d2, "Shannon entropy", "FNN", "bias"), 3)` and an error of
`r fixed(e_value(d2, "Shannon entropy", "FNN", "abs_error"), 3)`; in the
4-d design the biases are
`r fixed(e_value(d4, "Shannon entropy", "proxymix", "bias"), 3)` and
`r fixed(e_value(d4, "Shannon entropy", "FNN", "bias"), 3)` and the errors
`r fixed(e_value(d4, "Shannon entropy", "proxymix", "abs_error"), 3)` and
`r fixed(e_value(d4, "Shannon entropy", "FNN", "abs_error"), 3)`. A mixture
fitted to `r sim$n` points is a little tighter than the mixture that
generated them, and the closed-form Rényi-2 entropy of the fit shows the
same sign, a bias of
`r fixed(e_value(d2, "Renyi-2 entropy", "proxymix", "bias"), 3)` and
`r fixed(e_value(d4, "Renyi-2 entropy", "proxymix", "bias"), 3)` nats with
no Monte Carlo step involved. On the Shannon mutual information the
mixture's estimate, assembled from three Monte Carlo entropies, has a bias
of `r fixed(e_value(d2, "Shannon mutual information", "proxymix", "bias"), 3)`
and
`r fixed(e_value(d4, "Shannon mutual information", "proxymix", "bias"), 3)`
nats and a mean absolute error of
`r fixed(e_value(d2, "Shannon mutual information", "proxymix", "abs_error"), 3)`
and
`r fixed(e_value(d4, "Shannon mutual information", "proxymix", "abs_error"), 3)`;
the nearest-neighbour estimator reads
`r fixed(e_value(d2, "Shannon mutual information", "KSG", "bias"), 3)` high
in two dimensions and
`r fixed(-e_value(d4, "Shannon mutual information", "KSG", "bias"), 3)` low
in four, with errors of
`r fixed(e_value(d2, "Shannon mutual information", "KSG", "abs_error"), 3)`
and
`r fixed(e_value(d4, "Shannon mutual information", "KSG", "abs_error"), 3)`.
The closed-form Cauchy-Schwarz mutual information is within
`r fixed(e_value(d2, "Cauchy-Schwarz mutual information", "proxymix", "abs_error"), 3)`
and
`r fixed(e_value(d4, "Cauchy-Schwarz mutual information", "proxymix", "abs_error"), 3)`
nats of its quadrature value on average, or
`r round(100 * cs_rel[1L])` and `r round(100 * cs_rel[2L])` per cent of
that value.
Neither Shannon estimator reproduces the gap seen on the penguins, where
the mixture read well below the nearest-neighbour estimate on every pair.
Neither design here has both the four coordinates and the three clusters
of the colony. The 2-d design has three components in two coordinates, and
the 4-d design has two components in four.

On the number of components, proxymix's BIC chose the true count in
`r fixed(k_value(d2, "proxymix, BIC", "share_true"), 2)` of the 2-d datasets
and `r fixed(k_value(d4, "proxymix, BIC", "share_true"), 2)` of the 4-d
datasets, and mclust's BIC in
`r fixed(k_value(d2, "mclust, BIC", "share_true"), 2)` and
`r fixed(k_value(d4, "mclust, BIC", "share_true"), 2)`. The ICL prefers
fewer components where components overlap. In the 2-d design, where the
three components overlap most, proxymix's ICL chose the true count in
`r fixed(k_value(d2, "proxymix, ICL", "share_true"), 2)` of datasets and
mclust's in `r fixed(k_value(d2, "mclust, ICL", "share_true"), 2)`, and
chose two components in `r fixed(share_two(d2, "proxymix, ICL"), 2)` and
`r fixed(share_two(d2, "mclust, ICL"), 2)`; in the 4-d design the shares are
`r fixed(k_value(d4, "proxymix, ICL", "share_true"), 2)` and
`r fixed(k_value(d4, "mclust, ICL", "share_true"), 2)`. mclust's ICL is
taken over fourteen covariance models where proxymix fits the full
covariance only, which gives it more ways to separate two overlapping
components before merging them. A share estimated from `r sim$n_rep`
datasets carries a Monte Carlo standard error of up to
`r fixed(mcse_share, 2)`.

On the graph, the PC algorithm recovered the chain exactly in
`r fixed(g_value("pcalg", "share_exact"), 2)` of datasets, with a mean
structural Hamming distance of `r fixed(g_value("pcalg", "mean_shd"), 2)`.
proxymix's test averaged
`r fixed(g_value("proxymix", "mean_shd"), 2)` edges from the chain and
`r fixed(g_value("proxymix", "mean_edges"), 1)` edges in all, huge
`r fixed(g_value("huge", "mean_shd"), 2)` and
`r fixed(g_value("huge", "mean_edges"), 1)`, and the graphical lasso
`r fixed(g_value("glasso", "mean_shd"), 2)` and
`r fixed(g_value("glasso", "mean_edges"), 1)`. `r graph_direction`
proxymix tests each pair given all the other coordinates at level 0.05,
with no correction for the number of pairs.
The PC algorithm removes an edge once any conditioning set it tests leaves
the partial correlation not significant at 1 per cent.
The chain leaves `r n_absent` pairs without an edge. Were their tests
independent, all `r n_absent` would be left out at level 0.05 with
probability $0.95^{`r n_absent`}$, or `r fixed(share_no_false, 2)`, and
proxymix's test recovered the chain in
`r fixed(g_value("proxymix", "share_exact"), 2)` of datasets. Most of its
gap to the PC algorithm is therefore the test level, and the share of datasets
with no false edge in this design was close to what that level predicts. A level of $0.05 / \binom{p}{2}$ for $p$ coordinates,
passed as `alpha`, bounds the chance of any false edge at 0.05 instead. The graphical lasso
and huge choose a penalty by an information criterion.

In this run, in the 2-d design, the mixture's mean absolute error for the
Shannon entropy was `r verdict(d2, "Shannon entropy", "FNN")`, and for the
Shannon mutual information
`r verdict(d2, "Shannon mutual information", "KSG")`. In the 4-d design
its error for the Shannon entropy was
`r verdict(d4, "Shannon entropy", "FNN")`, and for the Shannon mutual
information it was `r verdict(d4, "Shannon mutual information", "KSG")`.
Two mixtures, one sample size, one graph and one test level do not cover
mixtures with many components, higher dimensions, other neighbour counts
for the nearest-neighbour estimators, or the choice of level. In the
simulation, no method was run on a target that could only be evaluated.

## Further reading

*How well a mixture proxies four awkward shapes* runs regime (iii) on
targets of different shapes and reports the fit-quality diagnostics to
check before reading any entropy value. *Choosing between the three fitting
regimes* explains the difference between fitting from samples and fitting
from a density you can only evaluate. *The
closed-form operator calculus on a mixture* covers the conditioning
operation that `gmm_conditional_entropy()` calls row by row, and the rest of
the exact algebra on a fitted mixture. *Fitting a proxy to a density you
cannot sample* introduces the regime (iii) fit behind the energy-field
graph.

## References

* Biernacki, C., Celeux, G. and Govaert, G. (2000). *Assessing a mixture
  model for clustering with the integrated completed likelihood.* IEEE
  Transactions on Pattern Analysis and Machine Intelligence 22(7),
  719--725. <https://doi.org/10.1109/34.865189>.
* Friedman, J., Hastie, T. and Tibshirani, R. (2008). *Sparse inverse
  covariance estimation with the graphical lasso.* Biostatistics 9(3),
  432--441. <https://doi.org/10.1093/biostatistics/kxm045>.
* Genz, A. C. and Malik, A. A. (1980). *Remarks on algorithm 006: An
  adaptive algorithm for numerical integration over an N-dimensional
  rectangular region.* Journal of Computational and Applied Mathematics
  6(4), 295--302. <https://doi.org/10.1016/0771-050X(80)90039-X>.
* Gorman, K. B., Williams, T. D. and Fraser, W. R. (2014). *Ecological
  sexual dimorphism and environmental variability within a community of
  Antarctic penguins (genus Pygoscelis).* PLoS ONE 9(3), e90081.
  <https://doi.org/10.1371/journal.pone.0090081>.
* Hershey, J. R. and Olsen, P. A. (2007). *Approximating the Kullback
  Leibler divergence between Gaussian mixture models.* 2007 IEEE
  International Conference on Acoustics, Speech and Signal Processing
  (ICASSP '07), IV-317--IV-320. <https://doi.org/10.1109/ICASSP.2007.366913>.
* Horst, A. M., Presmanes Hill, A. and Gorman, K. B. (2022). *Palmer
  Archipelago penguins data in the palmerpenguins R package -- an
  alternative to Anderson's irises.* The R Journal 14(1), 244--254.
  <https://doi.org/10.32614/RJ-2022-020>.
* Jaynes, E. T. (1957). *Information theory and statistical mechanics.*
  Physical Review 106(4), 620--630. <https://doi.org/10.1103/PhysRev.106.620>.
* Kalisch, M., Mächler, M., Colombo, D., Maathuis, M. H. and Bühlmann, P.
  (2012). *Causal inference using graphical models with the R package
  pcalg.* Journal of Statistical Software 47(11), 1--26.
  <https://doi.org/10.18637/jss.v047.i11>.
* Kozachenko, L. F. and Leonenko, N. N. (1987). *Sample estimate of the
  entropy of a random vector.* Problems of Information Transmission 23(2),
  95--101.
* Kraskov, A., Stögbauer, H. and Grassberger, P. (2004). *Estimating
  mutual information.* Physical Review E 69(6), 066138.
  <https://doi.org/10.1103/PhysRevE.69.066138>.
* Meinshausen, N. and Bühlmann, P. (2006). *High-dimensional graphs and
  variable selection with the lasso.* The Annals of Statistics 34(3),
  1436--1462. <https://doi.org/10.1214/009053606000000281>.
* Rose, K. (1998). *Deterministic annealing for clustering, compression,
  classification, regression, and related optimization problems.*
  Proceedings of the IEEE 86(11), 2210--2239. <https://doi.org/10.1109/5.726788>.
* Scrucca, L., Fop, M., Murphy, T. B. and Raftery, A. E. (2016). *mclust 5:
  Clustering, classification and density estimation using Gaussian finite
  mixture models.* The R Journal 8(1), 289--317.
  <https://doi.org/10.32614/RJ-2016-021>.
* Tsamardinos, I., Brown, L. E. and Aliferis, C. F. (2006). *The max-min
  hill-climbing Bayesian network structure learning algorithm.* Machine
  Learning 65(1), 31--78. <https://doi.org/10.1007/s10994-006-6889-7>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <https://doi.org/10.1080/07362994.2024.2372605>.
* Zhao, T., Liu, H., Roeder, K., Lafferty, J. and Wasserman, L. (2012).
  *The huge package for high-dimensional undirected graph estimation in R.*
  Journal of Machine Learning Research 13, 1059--1062.

## Reproduce

The vignette sets `set.seed(20260618)` once. Every function call that draws
its own random numbers and has a `seed` argument is given one.
`gmm_divergence()` has no `seed` argument, so its estimate and later draws
depend on chunk order.

The illustration gives `select_N()` and every `gmm_entropy()` call
`seed = 1L`, and computes each species' graphs after `set.seed(1L)`, since
`huge.select()` draws rows at random. Each simulated dataset, its fits and its
Monte Carlo entropies are seeded with the dataset's index. The stored
results come from the code shown, run on `r sim$run_date` with R
`r sim$versions[["R"]]`,
proxymix `r desc_version("proxymix")`, mclust `r desc_version("mclust")`, FNN
`r desc_version("FNN")`, glasso `r desc_version("glasso")`, huge
`r desc_version("huge")`, pcalg `r desc_version("pcalg")` and cubature
`r desc_version("cubature")`, and raised `r length(sim$warnings)`
warnings.

```{r session-info, eval = FALSE}
sessionInfo()
```

```{r session-info-print, echo = FALSE}
## printed 68 characters wide; a longer line breaks after its last slash or
## space within the width and continues indented
old_width <- options(width = 68L)
si_vec <- capture.output(sessionInfo())
options(old_width)
wrap_line <- function(s) {
  out <- character(0L)
  while (nchar(s) > 68L) {
    cut <- max(gregexpr("[/ ]", substr(s, 1L, 68L))[[1L]])
    if (cut < 20L) cut <- 68L
    out <- c(out, substr(s, 1L, cut))
    s <- paste0("    ", substr(s, cut + 1L, nchar(s)))
  }
  c(out, s)
}
writeLines(unlist(lapply(si_vec, wrap_line)))
```
