---
title: "Compressing a kernel density estimate into a mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compressing a kernel density estimate into a mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~25 s under rmarkdown::render() with ggplot2, ks, np,
     mclust, mixtools and KernSmooth installed, the simulation read from
     its stored results; macOS arm64 (Apple silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The problem

An analyst has a kernel density estimate of a two-dimensional quantity and a
pipeline that needs to keep conditioning on one coordinate and integrating
out the other. With a Gaussian kernel the estimate is itself a mixture of
`n` Gaussians, one per point, so every such query has a closed form, but
each one works through all `n` components and returns a mixture of the same
size. The question is: *I have a kernel density estimate and I need many
marginals and conditionals of it -- what does compressing it into a
$K$-component mixture cost me in accuracy, and what does it save per
query?*

## Package capabilities

A kernel density estimate gives a smooth, non-parametric density from a
sample of `n` points. With a Gaussian kernel it is an equal-weight mixture
of `n` Gaussians, so its marginals, its conditionals by the Schur complement
and its pushforwards through affine maps are exact in closed form, each an
`n`-component mixture. Every evaluation, conditional and quantile then works
through all `n` components, and every result stores them.

`from_kde()` compresses the kernel estimate into an `N`-component Gaussian
mixture proxy, with `N` typically much smaller than `n`, by treating the
estimate as a normalised, evaluable target and running the regime (iii) KLD-EM of
van der Hoek and Elliott (2024) against it. Its `bandwidth` argument takes a rule
of thumb by name (`"silverman"` or `"scott"`, after Silverman, 1986, and
Scott, 1992) or a numeric value, and `validation_size`
sets the number of held-out importance draws used to report fit quality
(by default, draws are added until the held-out estimate is precise). The proxy
then supports the same closed-form operator set exposed by the rest of the
package -- `dgmm()`, `rgmm()`, `gmm_marginalise()`,
`gmm_conditionalise()`, and the affine-Gaussian calculus of
`gmm_affine()` and `gmm_observe()` -- on `N` components instead of `n`.

`ess_summary()` reports the importance-sampling diagnostics of the
compression: the effective sample size, the largest self-normalised weight,
and the held-out estimate of the Kullback-Leibler divergence
$\mathrm{KL}(f \Vert g) = \int f \log (f / g)$, with $f$ the kernel estimate
and $g$ the proxy. `hellinger_mc()` estimates the
squared Hellinger distance between the proxy and its target by Monte Carlo.
The distance reported below is computed from fresh draws from the proxy, with
its Monte Carlo standard error.

The proxy targets the kernel estimate, not the density the data came from.
The bandwidth sets the estimate's bias-variance trade-off and the proxy
inherits it, although a proxy with few components can also smooth away some
of the estimate's sampling noise.

## Addressing the problem

### Recovering a known mixture

Sample from a known bimodal mixture and ask `from_kde()` to recover it. The
data here are synthetic: two unit-covariance Gaussians at $(-2, 0)$ and
$(2, 0)$, 150 draws each.

```{r recover-bimodal}
set.seed(20260601)

true_means <- cbind(c(-2, 0), c(2, 0))
true_cov <- diag(2)
x <- rbind(
  mvnfast::rmvn(150L, mu = true_means[, 1L], sigma = true_cov),
  mvnfast::rmvn(150L, mu = true_means[, 2L], sigma = true_cov)
)

fit <- from_kde(
  x, N = 2L,
  bandwidth = "silverman",
  is_size = 2000L, max_iter = 60L, seed = 1L,
  validation_size = 2000L
)
fit
```

The proxy components sit near the true means. Ordering the fitted components
by their first coordinate makes the comparison with the truth direct.

```{r recovered-means}
mu_hat <- vapply(fit@means, function(mu) mu, numeric(2L))
comp_order <- order(mu_hat[1L, ])
mu_hat <- mu_hat[, comp_order, drop = FALSE]
weight_hat <- fit@weights[comp_order]
mean_error <- max(abs(mu_hat - true_means))
```

```{r recovered-means-table, echo = FALSE}
knitr::kable(
  data.frame(
    component = c(1L, 2L),
    fitted_x1 = round(mu_hat[1L, ], 3),
    fitted_x2 = round(mu_hat[2L, ], 3),
    true_x1 = true_means[1L, ],
    true_x2 = true_means[2L, ],
    weight = round(weight_hat, 3)
  ),
  row.names = FALSE,
  col.names = c("Component", "Fitted $x_1$", "Fitted $x_2$", "True $x_1$",
                "True $x_2$", "Weight"),
  caption = paste0(
    "The two recovered component means beside the means of the mixture the ",
    "300 samples were drawn from."
  )
)
```

### Quality of the importance-sampled fit

A healthy regime (iii) fit has an effective sample size close to `is_size`
and a largest weight of a few parts in a thousand. The validation block
reports a held-out divergence on an independent importance draw, which is
the safeguard against overfitting to one particular importance realisation.

```{r fit-quality}
es <- ess_summary(fit)
validation_se <- fit@diagnostics$validation_mc_se
print(data.frame(is_size = es$is_size, ess = round(es$ess, 1),
                 ess_relative = round(es$ess_relative, 3),
                 max_weight = signif(es$max_weight, 3),
                 validation_kld = signif(es$validation_kld, 3),
                 validation_se = signif(validation_se, 2)),
      row.names = FALSE)
```

### What the compression costs

The proxy and the kernel estimate are both normalised densities on the same
space, so the total variation distance between them can be integrated
directly on a grid.
The total variation distance is the natural currency: it is the largest
difference in probability the two densities can assign to any event.

```{r compression-cost}
g1 <- seq(-6, 6, length.out = 160L)
g2 <- seq(-5, 5, length.out = 140L)
grid <- expand.grid(x1 = g1, x2 = g2)
gm <- as.matrix(grid)
cell <- (g1[2L] - g1[1L]) * (g2[2L] - g2[1L])

grid$kde <- exp(fit@target@log_density(gm))
grid$proxy <- dgmm(gm, fit)

total_variation <- 0.5 * sum(abs(grid$kde - grid$proxy)) * cell
mass_grid <- c(kde = sum(grid$kde) * cell, proxy = sum(grid$proxy) * cell)

hell <- hellinger_mc(fit, n_mc = 50000L, seed = 20260926L)
signif(c(total_variation = total_variation, hellinger_sq = hell[["h2"]],
         hellinger_se = hell[["se"]]), 3)
```

```{r fixed-notation, include = FALSE}
## Small errors are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
```

### Bandwidth sensitivity

Smaller bandwidths track the data more tightly, at lower bias and higher
variance; larger bandwidths smooth the proxy. The proxy inherits this
trade-off from the estimate it compresses.

```{r bandwidth-sweep}
bandwidth_grid <- c(0.2, 0.5, 1.0)
fits <- lapply(bandwidth_grid, function(h) {
  from_kde(x, N = 2L, bandwidth = h,
           is_size = 1500L, max_iter = 40L, seed = 1L)
})
## trace of the left-hand component's covariance, component 1 of Table 1
trace_left <- function(f) {
  j <- which.min(vapply(f@means, function(mu) mu[1L], numeric(1L)))
  sum(diag(f@covariances[[j]]))
}
```

```{r bandwidth-table, echo = FALSE}
knitr::kable(
  data.frame(
    bandwidth = bandwidth_grid,
    ess = round(vapply(fits, function(f) f@diagnostics$ess, numeric(1L)), 1),
    max_weight = signif(
      vapply(fits, function(f) f@diagnostics$max_weight, numeric(1L)), 3
    ),
    trace_sigma = round(vapply(fits, trace_left, numeric(1L)), 3)
  ),
  row.names = FALSE,
  col.names = c("Bandwidth", "Effective sample size", "Largest weight",
                "Trace of $\\Sigma_1$"),
  caption = paste0(
    "Bandwidth sweep at 1500 importance draws: a wider kernel gives a ",
    "smoother target, an easier importance problem, and a broader ",
    "component. $\\Sigma_1$ is the covariance of the left-hand component, ",
    "component 1 of Table 1."
  )
)
```

### Composing the compressed proxy

Once compressed, the proxy is an ordinary mixture of two components, so a
conditional or a sample works through two components rather than the 300 of
the kernel estimate.

```{r compose}
slice <- gmm_conditionalise(fit, given = c(NA, 0))
draws <- rgmm(200L, slice)
c(components = gmm_n_components(slice), dimension = gmm_dim(slice),
  draws = nrow(draws))
```

### Visualising the compression

The contour comparison shows where the compression spends its error, which
the summary distances cannot: the two contour sets are read on shared levels,
so any place they part company is a place the proxy has smoothed the
estimate.

```{r visualise, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4, out.width = "70%", fig.pos = "!ht", fig.cap = "Log-density contours of the kernel estimate (blue, solid) and the two-component proxy (orange, dashed) on shared levels. On the innermost level, around the two modes, the two nearly coincide. On the outer, low-density levels, beyond about $|x_2| = 2$, they separate: the kernel estimate's contours bend around the few outlying samples and the proxy's stay smooth.", fig.alt = "Contour plot of the kernel-density log-density and the Gaussian-mixture proxy log-density on a planar grid, with sample points overlaid; the innermost contours around the two modes nearly coincide, and the outer contours separate, the kernel ones wavy and the proxy ones smooth."}
plot_grid <- expand.grid(
  x1 = seq(-5, 5, length.out = 80L),
  x2 = seq(-4, 4, length.out = 60L)
)
pm <- as.matrix(plot_grid)
plot_grid$kde <- fit@target@log_density(pm)
plot_grid$proxy <- log(dgmm(pm, fit))

## Shared contour levels so the two log-densities are directly comparable.
brks <- pretty(range(c(plot_grid$kde, plot_grid$proxy), finite = TRUE), 9L)
sample_df <- data.frame(x1 = x[, 1L], x2 = x[, 2L])

ggplot2::ggplot() +
  ggplot2::geom_point(data = sample_df, ggplot2::aes(x1, x2),
                      colour = "grey60", alpha = 0.25, size = 0.5) +
  ggplot2::geom_contour(
    data = plot_grid,
    ggplot2::aes(x1, x2, z = kde, colour = "Kernel estimate",
                 linetype = "Kernel estimate"),
    breaks = brks, linewidth = 0.45
  ) +
  ggplot2::geom_contour(
    data = plot_grid,
    ggplot2::aes(x1, x2, z = proxy, colour = "Mixture proxy",
                 linetype = "Mixture proxy"),
    breaks = brks, linewidth = 0.55
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("Kernel estimate" = "#0072B2", "Mixture proxy" = "#D55E00")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Kernel estimate" = "solid", "Mixture proxy" = "dashed")
  ) +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::labs(
    title = "Kernel estimate and Gaussian-mixture proxy",
    subtitle = "log-density contours on shared levels",
    x = expression(x[1]), y = expression(x[2])
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 legend.position = "top",
                 panel.grid.minor = ggplot2::element_blank())
```

```{r visualise-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the contour-comparison",
    "figure is skipped.\n")
```

## Interpretation

The `r gmm_n_components(fit)`-component proxy recovers the mixture the 300
samples came from: the fitted means differ from the true means by at most
`r signif(mean_error, 2)`, and the two components carry weights of
`r paste(sprintf("%.3f", sort(fit@weights)), collapse = " and ")` against a
true half and half.

The compression itself was a well-conditioned importance problem. The
effective sample size was `r round(es$ess)` out of `r es$is_size` draws, or
`r round(100 * es$ess_relative)` per cent, and the largest single weight was
`r signif(es$max_weight, 2)`, so no handful of draws carried the fit. On an
independent importance draw of `r es$validation_size`, the held-out
divergence estimate was `r signif(fit@diagnostics$validation_kld, 2)`, with a
Monte Carlo standard error of `r signif(validation_se, 2)`.

The cost of the compression is the direct answer to the question. The
total variation distance between the 300-kernel estimate and the
`r gmm_n_components(fit)`-component mixture is `r fixed(total_variation, 3)`:
the two densities agree on the probability of any event to within about
`r fixed(100 * total_variation, 1)` percentage points. The grid captures
`r fixed(mass_grid[["kde"]], 4)` of the kernel estimate's mass and
`r fixed(mass_grid[["proxy"]], 4)` of the proxy's. The squared Hellinger
distance, estimated by `hellinger_mc()` from
`r format(hell[["n_mc"]], big.mark = ",")` fresh
draws from the proxy, is `r fixed(hell[["h2"]], 4)` with a Monte Carlo standard
error of `r fixed(hell[["se"]], 4)`, about `r round(hell[["h2"]] / hell[["se"]])` standard
errors from zero. The figure shows where the remaining
error sits: around the two modes the contours nearly coincide, and on the
outer, low-density levels the kernel estimate bends around the few outlying
samples while the two-component proxy stays smooth. What is bought
with that is size. The slice taken above is a `r gmm_dim(slice)`-dimensional,
`r gmm_n_components(slice)`-component mixture; the same slice of the kernel
estimate is also exact, and has 300 components.

The bandwidth sweep shows the trade-off passing straight through the
compression. A wider kernel gives a smoother target and an easier importance
problem, and the effective sample size rises from
`r round(fits[[1L]]@diagnostics$ess)` at bandwidth
`r sprintf("%.1f", bandwidth_grid[1L])` to
`r round(fits[[3L]]@diagnostics$ess)` at bandwidth
`r sprintf("%.1f", bandwidth_grid[3L])`, out of 1500 draws. The left-hand
component's covariance trace widens over the same sweep, from
`r round(trace_left(fits[[1L]]), 2)` to
`r round(trace_left(fits[[3L]]), 2)`, which is the proxy
inheriting the estimate's smoothing rather than resisting it.

## Limitations

`from_kde()` fits data of up to five dimensions without comment, warns
between six and ten dimensions and refuses above ten. Regime (iii) is driven by importance
sampling, whose effective sample size falls sharply in high dimensions, and
this vignette exercises the two-dimensional case only. For richer ambient
spaces, compose a low-dimensional proxy with the affine-Gaussian operator
calculus rather than fitting in the ambient space.

Only scalar and diagonal bandwidths are supported. A full-matrix bandwidth
effectively encodes a covariance estimate and blurs the distinction between
the kernel estimate and the mixture; where that is what is wanted, fit the
mixture directly with `fit_proxymix(regime = "sample")`.

A kernel estimate is normalised by construction, so the returned `gmm_target`
is flagged as normalised, with a log normalising constant of zero. That is
what makes a Hellinger distance between the proxy and the estimate
meaningful here and lets the divergence diagnostics report
absolute values rather than shifted ones. A target that is not normalised
does not support the same reading.

The constructor does not choose the bandwidth. Use the conventional rules of
thumb, passed by name, or a cross-validation procedure outside the package,
and pass the chosen bandwidth in. The sweep above shows the consequences of
the choice; it does not make the choice.

Finally, the compression earns its place only in some pipelines. Where the
goal is simply to find a Gaussian-mixture density given samples, the
classical expectation-maximisation regime is more direct, fits in linear
time, and does not pay the regime (iii) importance-sampling tax. Reach for
`from_kde()` when the kernel estimate's smoothing is itself a step in the
pipeline, or when its bias-variance trade-off is part of what is being
validated downstream.
The numerical illustration that follows measures both sides of that
trade-off, the error the compression adds against the kernel estimate and
the time each conditional query then takes, on two real datasets and in
simulation against a kernel estimator, a conditional-density estimator and
two mixture fitters.

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_ks <- requireNamespace("ks", quietly = TRUE)
has_np <- requireNamespace("np", quietly = TRUE)
has_mclust <- requireNamespace("mclust", quietly = TRUE)
has_mixtools <- requireNamespace("mixtools", quietly = TRUE)
has_kernsmooth <- requireNamespace("KernSmooth", quietly = TRUE)
has_penguins <- exists("penguins", envir = asNamespace("datasets"))
run_illustration <- has_ks && has_np && has_mclust && has_mixtools &&
  has_kernsmooth && has_penguins
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("ks, np, mclust, mixtools, KernSmooth or the penguins data (R 4.5.0 or",
    "later) is not available on this build, so the real-data illustration",
    "is skipped.\n")
```

### Conditionals of two real densities, from six packages

The question is what a conditional query costs once the kernel estimate has
been compressed, in error and in time. Two datasets are used: the Old
Faithful geyser data (Azzalini and Bowman, 1990), 272 eruptions with their
duration and the waiting time to the next one, shipped with R as
`faithful`; and the Palmer penguins data (Gorman et al.,
2014), shipped with R as `penguins` since version 4.5.0 (Horst et al.,
2022), of which bill length and flipper length are kept
for the birds with both measured. Both scatters are multimodal, one with two
groups and one with three species.

The reference is the kernel estimate itself, computed by ks (Duong, 2007)
with its diagonal plug-in bandwidth. The same bandwidth is passed to
`from_kde()`, so the proxy compresses exactly the density ks evaluates, and
to `bkde2D()` of KernSmooth (Wand and Jones, 1995), which bins the same
estimate on a grid. np (Hayfield and Racine, 2008) estimates the conditional
density of $y$ given $x$ directly, with its likelihood cross-validated
bandwidths. mclust (Scrucca et al., 2016) and mixtools
(Benaglia et al., 2009) fit Gaussian mixtures to the
raw data by expectation-maximisation, mclust choosing its number of
components by BIC and mixtools at its default of two. The five competitor
packages install from CRAN once and load as usual:

```{r install-competitors, eval = FALSE}
install.packages(c("ks", "np", "mclust", "mixtools", "KernSmooth"))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(mclust)
library(mixtools)
library(np)
options(np.messages = FALSE)
```

At each of 20 values of $x$, the 5th to the 95th percentile of the data,
every method returns the conditional mean and the 90 per cent conditional
quantile of $y$. For a mixture, proxymix's own or one built from the mclust
and mixtools parameters, both come from `gmm_conditionalise()`,
`gmm_mean()` and `qgmm()` in closed form. For the kernel estimates and for
np, the conditional density is evaluated on a grid of 401 values of $y$ and
the two summaries are read off that grid. The score is the mean absolute
difference from the reference over the 20 values, in the units of $y$, and
the time per conditional query is recorded for every method.

```{r illustration-helpers, eval = run_illustration}
as_gmm <- function(w, mu, sigma) {
  gmm(weights = w, means = mu, covariances = sigma)
}

## conditional mean and 90% quantile from a density tabulated on y_grid
grid_summary <- function(y_grid, f) {
  w <- pmax(f, 0) / sum(pmax(f, 0))
  c(mean = sum(w * y_grid),
    q90 = approx(cumsum(w), y_grid, xout = 0.9, ties = "ordered")$y)
}

gmm_summary <- function(g, x0) {
  s <- gmm_conditionalise(g, given = c(x0, NA))
  c(mean = gmm_mean(s), q90 = qgmm(0.9, s))
}

conditional_scores <- function(xy, label) {
  h <- sqrt(diag(ks::Hpi.diag(xy)))
  x_eval <- quantile(xy[, 1L], probs = seq(0.05, 0.95, length.out = 20L),
                     names = FALSE)
  y_grid <- seq(min(xy[, 2L]) - 3 * h[2L], max(xy[, 2L]) + 3 * h[2L],
                length.out = 401L)

  kde_cond <- function(x0) {
    f <- ks::kde(xy, H = diag(h^2), eval.points = cbind(x0, y_grid))
    grid_summary(y_grid, f$estimate)
  }
  reference <- t(vapply(x_eval, kde_cond, numeric(2L)))

  set.seed(20260925)
  fits <- list()
  for (K in c(3L, 5L, 8L)) {
    fits[[paste0("proxymix, K = ", K)]] <-
      from_kde(xy, N = K, bandwidth = h, is_size = 20000L, seed = 1L)
  }
  mc <- Mclust(xy, verbose = FALSE)
  fits[["mclust, G by BIC"]] <- as_gmm(
    mc$parameters$pro,
    lapply(seq_len(mc$G), function(j) mc$parameters$mean[, j]),
    lapply(seq_len(mc$G), function(j) mc$parameters$variance$sigma[, , j])
  )
  ## mvnormalmixEM() prints its iteration count, which is discarded
  invisible(capture.output(mt <- mvnormalmixEM(xy)))
  fits[["mixtools, k = 2"]] <- as_gmm(mt$lambda, mt$mu, mt$sigma)

  cd <- npcdens(npcdensbw(xdat = xy[, 1L], ydat = xy[, 2L]))
  np_cond <- function(x0) {
    f <- predict(cd, newdata = data.frame(xdat = x0, ydat = y_grid))
    grid_summary(y_grid, f)
  }

  bk <- KernSmooth::bkde2D(xy, bandwidth = h, gridsize = c(401L, 401L),
                           range.x = list(range(x_eval), range(y_grid)))
  bk_cond <- function(x0) {
    i <- findInterval(x0, bk$x1, all.inside = TRUE)
    w <- (x0 - bk$x1[i]) / (bk$x1[i + 1L] - bk$x1[i])
    grid_summary(bk$x2, (1 - w) * bk$fhat[i, ] + w * bk$fhat[i + 1L, ])
  }

  queries <- c(
    lapply(fits, function(g) function(x0) gmm_summary(g, x0)),
    list("np, npcdens" = np_cond, "KernSmooth, bkde2D" = bk_cond,
         "ks, kde (reference)" = kde_cond)
  )
  curves <- lapply(names(queries), function(method) {
    run_query <- function() t(vapply(x_eval, queries[[method]], numeric(2L)))
    est <- run_query()
    # median of five timed passes over the evaluation points
    secs <- median(replicate(5L, system.time(run_query())[[3L]]))
    data.frame(data = label, method = method, x = x_eval,
               mean = est[, 1L], q90 = est[, 2L],
               ms_per_query = 1000 * secs / length(x_eval),
               mclust_G = mc$G)
  })
  curves <- do.call(rbind, curves)
  curves$mean_ref <- rep(reference[, 1L], length(queries))
  curves$q90_ref <- rep(reference[, 2L], length(queries))
  curves
}
```

```{r illustration-run, eval = run_illustration}
faithful_xy <- as.matrix(faithful[, c("eruptions", "waiting")])
penguins_xy <- as.matrix(na.omit(penguins[, c("bill_len", "flipper_len")]))
curves <- rbind(conditional_scores(faithful_xy, "faithful"),
                conditional_scores(penguins_xy, "penguins"))
scores <- aggregate(
  cbind(mean_err = abs(mean - mean_ref), q90_err = abs(q90 - q90_ref),
        ms = ms_per_query) ~ data + method,
  data = curves, FUN = mean
)
```

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.3\textheight\clearpage\fi
```

```{r illustration-table, eval = run_illustration, echo = FALSE}
method_order <- unique(curves$method[curves$data == "faithful"])
wide <- merge(
  scores[scores$data == "faithful", c("method", "mean_err", "q90_err", "ms")],
  scores[scores$data == "penguins", c("method", "mean_err", "q90_err", "ms")],
  by = "method", suffixes = c("_f", "_p"), all = TRUE
)
wide$ms <- rowMeans(wide[, c("ms_f", "ms_p")])
wide$ms <- round(wide$ms, 2)
wide <- wide[order(match(wide$method, method_order)), ]
sd_y <- c(faithful = sd(faithful_xy[, 2L]), penguins = sd(penguins_xy[, 2L]))
knitr::kable(
  wide[, c("method", "mean_err_f", "q90_err_f", "mean_err_p", "q90_err_p",
           "ms")],
  digits = c(0L, 2L, 2L, 2L, 2L, 2L), row.names = FALSE,
  col.names = c("Method", "Mean (F)", "90% (F)", "Mean (P)", "90% (P)",
                "ms per query"),
  caption = paste0(
    "Mean absolute error of the conditional mean (Mean) and of the 90% ",
    "conditional quantile (90%) of $y$ given $x$ over 20 values of $x$, ",
    "scored ",
    "against the ks kernel estimate, on faithful (F, waiting time, minutes) ",
    "and on penguins (P, flipper length, mm), and the time per conditional ",
    "query averaged over the two datasets."
  )
)
score_of <- function(data, method_start, what) {
  s1 <- scores$data == data & startsWith(scores$method, method_start)
  scores[[what]][s1]
}
## ratio of the rounded times the table prints
ratio_query <- wide$ms[startsWith(wide$method, "ks")] /
  wide$ms[wide$method == "proxymix, K = 3"]
## signed departures from the reference on faithful, for Figure 2
dev_of <- function(method, what) {
  s1 <- curves$data == "faithful" & curves$method == method
  curves[[what]][s1] - curves[[paste0(what, "_ref")]][s1]
}
x_faithful <- curves$x[curves$data == "faithful" &
                         curves$method == "proxymix, K = 3"]
k3_mean_drop <- -min(dev_of("proxymix, K = 3", "mean"))
k3_drop_at <- x_faithful[which.min(dev_of("proxymix, K = 3", "mean"))]
k8_mean_dev <- max(abs(dev_of("proxymix, K = 8", "mean")))
k8_q90_dev <- max(abs(dev_of("proxymix, K = 8", "q90")))
mc_mean_dev <- max(abs(dev_of("mclust, G by BIC", "mean")))
mc_q90_drop <- -min(dev_of("mclust, G by BIC", "q90"))
```

```{r fig-conditionals, eval = run_illustration && has_ggplot2, echo = run_illustration && has_ggplot2, fig.pos = "!ht", fig.cap = paste0("Conditional mean (left) and 90% conditional quantile (right) of waiting time given eruption duration on the faithful data, from the ks kernel estimate that is the reference, the proxy compressed from it at three and eight components, mclust fitted to the raw data and the np conditional-density estimate. The eight-component proxy stays within ", sprintf("%.1f", ceiling(10 * k8_mean_dev) / 10), " minutes of the reference in the conditional mean and within ", sprintf("%.1f", ceiling(10 * k8_q90_dev) / 10), " minutes in the 90% quantile. In the conditional mean, mclust stays within ", sprintf("%.1f", ceiling(10 * mc_mean_dev) / 10), " minutes of the reference, and the three-component proxy falls up to ", sprintf("%.1f", ceiling(10 * k3_mean_drop) / 10), " minutes below it at a duration of ", sprintf("%.1f", k3_drop_at), " minutes, the upper end of the jump between the two groups. In the 90% quantile, mclust falls up to ", sprintf("%.1f", ceiling(10 * mc_q90_drop) / 10), " minutes below the reference from the jump to the long eruptions."), fig.alt = "Two panels of curves over eruption duration from 1.8 to 4.8 minutes. In each panel the reference kernel estimate, the three- and eight-component proxies, mclust and np all rise from about 55 to about 80 minutes across the middle of the range. In the conditional-mean panel the three-component proxy dips below the reference at the upper end of the jump; in the quantile panel mclust runs below the reference from the jump onwards."}
show <- c("ks, kde (reference)", "proxymix, K = 3", "proxymix, K = 8",
          "mclust, G by BIC", "np, npcdens")
plot_df <- curves[curves$data == "faithful" & curves$method %in% show, ]
plot_df <- rbind(
  data.frame(plot_df[, c("method", "x")], what = "conditional mean",
             y = plot_df$mean),
  data.frame(plot_df[, c("method", "x")], what = "90% conditional quantile",
             y = plot_df$q90)
)
plot_df$method <- factor(plot_df$method, levels = show)
plot_df$what <- factor(plot_df$what,
                       levels = c("conditional mean",
                                  "90% conditional quantile"))
ggplot2::ggplot(plot_df, ggplot2::aes(x, y, colour = method,
                                      linetype = method)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~ what) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("#000000", "#D55E00", "#E69F00", "#0072B2", "#009E73")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("solid", "dashed", "dotted", "solid", "solid")
  ) +
  ggplot2::labs(x = "eruption duration, minutes",
                y = "waiting time, minutes") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-conditionals-skip, eval = run_illustration && !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the conditional-curve",
    "figure is skipped.\n")
```

```{r illustration-facts, eval = run_illustration, include = FALSE}
mclust_G <- vapply(c("faithful", "penguins"), function(d) {
  curves$mclust_G[curves$data == d][1L]
}, integer(1L))
best_proxy <- function(data, what) {
  s1 <- scores$data == data & startsWith(scores$method, "proxymix")
  m <- scores[s1, ]
  m$method[which.min(m[[what]])]
}
proxy_scores <- function(data, what) {
  s1 <- scores$data == data & startsWith(scores$method, "proxymix")
  setNames(scores[[what]][s1], sub("proxymix, ", "", scores$method[s1]))
}
## the proxies a method's error beats, as a phrase
beats_phrase <- function(data, method_start, what) {
  p <- proxy_scores(data, what)
  won <- names(p)[score_of(data, method_start, what) < p]
  won <- won[order(as.integer(sub("K = ", "", won)))]
  if (length(won) == length(p)) return("every proxy")
  if (length(won) == 0L) return("no proxy")
  if (length(won) == 1L) return(paste0("the proxy at $", won, "$"))
  paste0("the proxies at $K$ = ",
         paste(sub("K = ", "", won), collapse = " and "))
}
fitter_err <- function(data, what) {
  c(score_of(data, "mclust", what), score_of(data, "mixtools", what))
}
fitters_further <- all(vapply(c("faithful", "penguins"), function(d) {
  all(vapply(c("mean_err", "q90_err"), function(w) {
    min(fitter_err(d, w)) > max(proxy_scores(d, w))
  }, logical(1L)))
}, logical(1L)))
if (!fitters_further) {
  stop("The text on mclust and mixtools no longer matches Table 3.",
       call. = FALSE)
}
## how np compares with the proxies on one dataset, for both summaries
np_phrase <- function(data) {
  on_mean <- beats_phrase(data, "np", "mean_err")
  on_q90 <- beats_phrase(data, "np", "q90_err")
  if (on_mean == on_q90) {
    return(paste0("closer to the reference than ", on_mean,
                  " for both summaries"))
  }
  paste0("closer to the reference than ", on_mean, " for the mean and than ",
         on_q90, " for the quantile")
}
kernsmooth_err <- max(unlist(lapply(c("faithful", "penguins"), function(d) {
  c(score_of(d, "KernSmooth", "mean_err"), score_of(d, "KernSmooth", "q90_err"))
})))
fastest_query <- wide$method[which.min(wide$ms)]
if (!startsWith(fastest_query, "KernSmooth")) {
  stop("The text on query times no longer matches Table 3.", call. = FALSE)
}
```

Compressing the faithful estimate to three components moves the
conditional mean by `r fixed(score_of("faithful", "proxymix, K = 3", "mean_err"), 2)`
minutes on average and the 90 per cent quantile by
`r fixed(score_of("faithful", "proxymix, K = 3", "q90_err"), 2)` minutes,
against a standard deviation of `r fixed(sd_y[["faithful"]], 1)` minutes in
waiting time. On faithful more components bring the proxy closer to the
reference, to `r fixed(score_of("faithful", "proxymix, K = 8", "mean_err"), 2)`
and `r fixed(score_of("faithful", "proxymix, K = 8", "q90_err"), 2)` minutes
at eight. On penguins they do not: the smallest errors among the proxies are
at $`r sub("proxymix, ", "", best_proxy("penguins", "mean_err"))`$ for the
mean and $`r sub("proxymix, ", "", best_proxy("penguins", "q90_err"))`$ for
the quantile, and all three sit within a millimetre of the reference on average, against a
standard deviation of `r fixed(sd_y[["penguins"]], 1)` mm in flipper length.
mclust chose `r mclust_G[["faithful"]]` components on faithful and
`r mclust_G[["penguins"]]` on penguins. Its conditionals and those of
mixtools at two components are further from the kernel reference than every
proxy's on both datasets, for both summaries: on faithful their 90 per cent
quantiles are off by `r fixed(score_of("faithful", "mclust", "q90_err"), 2)`
and `r fixed(score_of("faithful", "mixtools", "q90_err"), 2)` minutes, against
`r fixed(score_of("faithful", "proxymix, K = 3", "q90_err"), 2)` for the
three-component proxy. That is the expected order: the proxies are fitted to
the kernel estimate and the two mixture fitters to the raw data. np
estimates the conditional density directly. On penguins it is
`r np_phrase("penguins")`. On faithful it is `r np_phrase("faithful")`. The binned
KernSmooth estimate has a mean absolute error of at most
`r fixed(kernsmooth_err, 2)` minutes or millimetres against the reference,
as a second implementation of the same estimator should.

The time column is the other side of the answer, and it measures cost per
query. A conditional of the reference, as computed here, evaluates all
`r nrow(faithful_xy)` or `r nrow(penguins_xy)` kernels at 401 grid points.
The same conditional also has a closed form, a mixture of
`r nrow(faithful_xy)` or `r nrow(penguins_xy)` components, so its cost grows
with $n$ either way; a conditional of the proxy is a Schur complement over
$K$ components and a one-dimensional quantile. Averaged over the two
datasets, a query on the three-component proxy took
`r fixed(wide$ms[wide$method == "proxymix, K = 3"], 2)` ms against
`r fixed(wide$ms[startsWith(wide$method, "ks")], 2)` ms on the reference,
about `r round(ratio_query)` times faster on this build. np's query is
cheaper than the reference, since its bandwidth search is done at fit time,
and dearer than the mixtures. The binned KernSmooth grid is the cheapest of
all, since a conditional is one interpolation between two stored grid rows.
On faithful the figure shows the differences around the jump between the
short and the long eruptions. There the three-component proxy's conditional
mean falls up to `r fixed(ceiling(10 * k3_mean_drop) / 10, 1)` minutes below the reference while
the eight-component proxy stays within `r fixed(ceiling(10 * k8_mean_dev) / 10, 1)` minutes of
it. mclust's conditional mean stays within `r fixed(ceiling(10 * mc_mean_dev) / 10, 1)` minutes
of the reference, and its 90 per cent
quantile falls up to `r fixed(ceiling(10 * mc_q90_drop) / 10, 1)` minutes below the reference
from the jump to the long eruptions. Two datasets are too few to rank the methods, so the
simulation below repeats the comparison on data drawn from a known density.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("from_kde_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
sim_value <- function(method, what) {
  sim$sim_tab[[what]][sim$sim_tab$method == method]
}
s1 <- sim$res$method == "mixtools, k = 3"
mixtools_secs <- c(median = median(sim$res$secs[s1]),
                   max = max(sim$res$secs[s1]))
mclust_G_sim <- table(sim$res$G[sim$res$method == "mclust, G by BIC"])
## paired difference in ISE, proxy minus the kernel estimate, in thousandths
paired_ise <- t(vapply(c(3L, 5L, 8L), function(K) {
  a <- sim$res[sim$res$method == paste0("proxymix, K = ", K), ]
  b <- sim$res[sim$res$method == "ks, kde", ]
  d <- 1000 * (a$ise[order(a$rep)] - b$ise[order(b$rep)])
  c(K = K, mean = mean(d), se = sd(d) / sqrt(length(d)))
}, numeric(3L)))
## the same paired difference for np and KernSmooth
paired_kernel <- function(method) {
  a <- sim$res[sim$res$method == method, ]
  b <- sim$res[sim$res$method == "ks, kde", ]
  d <- 1000 * (a$ise[order(a$rep)] - b$ise[order(b$rep)])
  c(mean = mean(d), se = sd(d) / sqrt(length(d)))
}
np_paired <- paired_kernel("np, npudens")
bk_paired <- paired_kernel("KernSmooth, bkde2D")
## seconds for a reduced run of 5 datasets, at the full run's rate
smoke_secs <- 5 * sim$elapsed_secs / sim$n_rep
mclust_G_mode <- names(mclust_G_sim)[which.max(mclust_G_sim)]
# a version as its DESCRIPTION writes it, when the installed one is the one stored
desc_version <- function(pkg) {
  stored <- sim$versions[[pkg]]
  if (identical(as.character(packageVersion(pkg)), stored)) {
    utils::packageDescription(pkg)$Version
  } else {
    stored
  }
}
```

The simulation draws `r sim$n_rep` datasets of `r sim$n` points from a
three-component mixture in two dimensions, with weights 0.35, 0.35 and 0.30,
means $(-2, 0)$, $(2, 1)$ and $(0, 4)$, and covariances the identity, a
unit-variance matrix with correlation 0.5, and 0.6 times the identity. Each
dataset gets the ks kernel estimate with its diagonal plug-in bandwidth,
compressed by `from_kde()` at $K$ = 3, 5 and 8 with 20 000 importance draws,
as in the real-data section. mclust chooses its number of components by
BIC, mixtools is given the true count of three, np estimates the joint
density with its likelihood cross-validated bandwidths, and KernSmooth bins
the ks estimate. The reference is the true mixture. The score is
the integrated squared error of each density against it, computed by
quadrature on a `r sim$grid_side` by `r sim$grid_side` grid over
$[-6, 6] \times [-4, 8]$, and the seconds each fit took. The run uses `r sim$n_rep` datasets, because each three-component
mixtools fit took a median of `r round(mixtools_secs[["median"]])` seconds and
up to `r round(mixtools_secs[["max"]])`, and the Monte Carlo standard error
of each mean is reported beside it. This chunk is complete and runs as
shown, but it took about `r round(sim$elapsed_secs / 60)` minutes on one core
(R `r sim$versions[["R"]]`, Apple silicon); the results below are read from
its stored output. For a quick reduced run, change the line `n_rep <- 100L` to
`n_rep <- 5L`, which takes about `r fixed(smoke_secs / 60, 1)` minutes.

```{r sim-code, eval = FALSE}
library(proxymix)
library(mclust)
library(mixtools)
library(np)
options(np.messages = FALSE)

n <- 500L       # points in each dataset
n_rep <- 100L   # datasets
K_grid <- c(3L, 5L, 8L)

truth <- list(
  weights = c(0.35, 0.35, 0.30),
  means = list(c(-2, 0), c(2, 1), c(0, 4)),
  covs = list(diag(2), matrix(c(1, 0.5, 0.5, 1), 2L), 0.6 * diag(2))
)
r_truth <- function(n) {
  k <- sample.int(3L, n, replace = TRUE, prob = truth$weights)
  out <- matrix(NA_real_, n, 2L)
  for (j in seq_len(3L)) {
    s <- k == j
    out[s, ] <- mvnfast::rmvn(sum(s), truth$means[[j]], truth$covs[[j]])
  }
  out
}
d_truth <- function(pts) {
  Reduce(`+`, lapply(seq_len(3L), function(j) {
    truth$weights[j] * mvnfast::dmvn(pts, truth$means[[j]], truth$covs[[j]])
  }))
}

# quadrature grid for the integrated squared error
g1 <- seq(-6, 6, length.out = 121L)
g2 <- seq(-4, 8, length.out = 121L)
grid <- as.matrix(expand.grid(x1 = g1, x2 = g2))
cell <- (g1[2L] - g1[1L]) * (g2[2L] - g2[1L])
f_true <- d_truth(grid)
ise <- function(f_hat) sum((f_hat - f_true)^2) * cell

as_gmm <- function(w, mu, sigma) {
  gmm(weights = w, means = mu, covariances = sigma)
}

one_dataset <- function(r) {
  set.seed(r)
  x <- r_truth(n)
  h <- sqrt(diag(ks::Hpi.diag(x)))
  out <- list()
  secs <- system.time(
    kd <- ks::kde(x, H = diag(h^2), eval.points = grid)
  )[[3L]]
  out[["ks, kde"]] <- c(ise = ise(kd$estimate), secs = secs, G = NA)
  for (K in K_grid) {
    secs <- system.time(
      f <- from_kde(x, N = K, bandwidth = h, is_size = 20000L, seed = r)
    )[[3L]]
    out[[paste0("proxymix, K = ", K)]] <-
      c(ise = ise(dgmm(grid, f)), secs = secs, G = K)
  }
  secs <- system.time(mc <- Mclust(x, verbose = FALSE))[[3L]]
  g_mc <- as_gmm(
    mc$parameters$pro,
    lapply(seq_len(mc$G), function(j) mc$parameters$mean[, j]),
    lapply(seq_len(mc$G), function(j) mc$parameters$variance$sigma[, , j])
  )
  out[["mclust, G by BIC"]] <- c(ise = ise(dgmm(grid, g_mc)), secs = secs,
                                 G = mc$G)
  secs <- system.time(mt <- mvnormalmixEM(x, k = 3L))[[3L]]
  out[["mixtools, k = 3"]] <-
    c(ise = ise(dgmm(grid, as_gmm(mt$lambda, mt$mu, mt$sigma))),
      secs = secs, G = 3L)
  secs <- system.time({
    nd <- npudens(npudensbw(x))
    f_np <- predict(nd, newdata = data.frame(grid))
  })[[3L]]
  out[["np, npudens"]] <- c(ise = ise(f_np), secs = secs, G = NA)
  secs <- system.time(
    bk <- KernSmooth::bkde2D(x, bandwidth = h, gridsize = c(121L, 121L),
                             range.x = list(range(g1), range(g2)))
  )[[3L]]
  out[["KernSmooth, bkde2D"]] <- c(ise = ise(as.vector(bk$fhat)),
                                   secs = secs, G = NA)
  data.frame(rep = r, method = names(out), do.call(rbind, out),
             row.names = NULL)
}

res <- do.call(rbind, lapply(seq_len(n_rep), one_dataset))

sim_tab <- aggregate(cbind(ise, secs) ~ method, data = res, FUN = mean)
sim_tab$ise_se <- aggregate(ise ~ method, data = res, FUN = sd)$ise /
  sqrt(n_rep)
sim_tab$G_mode <- aggregate(G ~ method, data = res, na.action = na.pass,
  FUN = function(v) if (all(is.na(v))) NA else as.numeric(names(which.max(table(v))))
)$G
sim_tab
```


The kernel estimate that `from_kde()` compresses has an integrated squared
error of `r fixed(1000 * sim_value("ks, kde", "ise"), 3)` thousandths, and
the three-, five- and eight-component proxies come out at
`r fixed(1000 * sim_value("proxymix, K = 3", "ise"), 3)`,
`r fixed(1000 * sim_value("proxymix, K = 5", "ise"), 3)` and
`r fixed(1000 * sim_value("proxymix, K = 8", "ise"), 3)`, with Monte Carlo
standard errors of about `r fixed(1000 * sim_value("ks, kde", "ise_se"), 2)`.
Paired over the same datasets, the proxy minus the kernel estimate is
`r fixed(paired_ise[1L, "mean"], 2)` thousandths at three components,
`r fixed(paired_ise[2L, "mean"], 2)` at five and
`r fixed(paired_ise[3L, "mean"], 3)` at eight, with standard errors
`r fixed(paired_ise[1L, "se"], 2)`, `r fixed(paired_ise[2L, "se"], 2)` and
`r fixed(paired_ise[3L, "se"], 3)`. At three and five components the
compression lowers the error against the truth, since a proxy with few
components cannot follow the sampling bumps of the kernel estimate; at eight
it adds no detectable error. mclust fitted to the raw data,
which chose `r mclust_G_mode` components in `r max(mclust_G_sim)` of the
`r sim$n_rep` datasets, and mixtools at the true count reach
`r fixed(1000 * sim_value("mclust, G by BIC", "ise"), 3)` and
`r fixed(1000 * sim_value("mixtools, k = 3", "ise"), 3)`, below every proxy,
as expected when the data come from the family being fitted. Paired over
the same datasets, np's error is `r fixed(-np_paired[["mean"]], 3)`
thousandths below that of ks, with a standard error of
`r fixed(np_paired[["se"]], 3)`, or `r fixed(-np_paired[["mean"]] / np_paired[["se"]], 1)`
standard errors. KernSmooth's error is within
`r fixed(abs(bk_paired[["mean"]]), 4)` thousandths of that of ks. mixtools took
`r round(sim_value("mixtools, k = 3", "secs"), 1)` seconds a fit against
`r fixed(sim_value("proxymix, K = 3", "secs"), 2)` to
`r fixed(sim_value("proxymix, K = 8", "secs"), 2)` for the proxies and
`r fixed(sim_value("mclust, G by BIC", "secs"), 2)` for mclust.

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.3\textheight\clearpage\fi
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order_sim <- c("ks, kde", "proxymix, K = 3", "proxymix, K = 5",
                      "proxymix, K = 8", "mclust, G by BIC",
                      "mixtools, k = 3", "np, npudens", "KernSmooth, bkde2D")
tab_sim <- tab_sim[order(match(tab_sim$method, method_order_sim)), ]
tab_sim$ise <- 1000 * tab_sim$ise
tab_sim$ise_se <- 1000 * tab_sim$ise_se
knitr::kable(
  tab_sim[, c("method", "ise", "ise_se", "secs")],
  digits = c(0L, 3L, 3L, 2L), row.names = FALSE,
  col.names = c("Method", "ISE, thousandths", "Monte Carlo SE",
                "Fit, seconds"),
  caption = paste0(
    "Integrated squared error against the true three-component density, ",
    "in thousandths, averaged over ", sim$n_rep, " simulated datasets of ",
    sim$n, " points, with its Monte Carlo standard error, and the mean ",
    "time of one fit."
  )
)
```

In this run proxymix came out ahead of the three kernel estimators on
integrated squared error at $K$ = 3 and 5 and level with ks at $K$ = 8,
and a conditional query on the three-component proxy was about
`r round(ratio_query)` times faster, averaged over the two datasets, than one
on the ks estimate it compresses, evaluated on a grid of 401 values. mclust and mixtools, fitted
directly to the data, had lower integrated squared error than any proxy; np
was closer to the kernel reference than
`r beats_phrase("penguins", "np", "mean_err")` on the penguins conditional
means; and the binned KernSmooth grid answered conditional queries fastest
and closest to the reference. One
well-separated mixture at one sample size and one bandwidth rule were tried;
heavier tails, overlapping components, higher dimensions and bandwidths
chosen by cross-validation are not covered.

## Further reading

*Choosing between the three fitting regimes* sets out why compressing an
evaluable-but-unsampleable target is regime (iii) work, and what the
classical expectation-maximisation regime would do with the same 300 samples
instead. *The closed-form operator calculus on a mixture* covers the
marginalisation, conditioning, pushforward and update operators, at a cost
set by $K$ on the proxy. *One mixture, many
methods* places the kernel estimate and the compressed proxy on a single
axis running from one global component to one component per datum.

## References

* Azzalini, A. and Bowman, A. W. (1990). *A look at some data on the Old
  Faithful geyser.* Applied Statistics 39(3), 357--365.
  <https://doi.org/10.2307/2347385>.
* Benaglia, T., Chauveau, D., Hunter, D. R. and Young, D. S. (2009).
  *mixtools: An R package for analyzing finite mixture models.* Journal of
  Statistical Software 32(6), 1--29. <https://doi.org/10.18637/jss.v032.i06>.
* Duong, T. (2007). *ks: Kernel density estimation and kernel discriminant
  analysis for multivariate data in R.* Journal of Statistical Software
  21(7), 1--16. <https://doi.org/10.18637/jss.v021.i07>.
* Gorman, K. B., Williams, T. D. and Fraser, W. R. (2014). *Ecological
  sexual dimorphism and environmental variability within a community of
  Antarctic penguins (genus Pygoscelis).* PLoS ONE 9(3), e90081.
  <https://doi.org/10.1371/journal.pone.0090081>.
* Hayfield, T. and Racine, J. S. (2008). *Nonparametric econometrics: The
  np package.* Journal of Statistical Software 27(5), 1--32.
  <https://doi.org/10.18637/jss.v027.i05>.
* Horst, A. M., Presmanes Hill, A. and Gorman, K. B. (2022). *Palmer
  Archipelago penguins data in the palmerpenguins R package -- an
  alternative to Anderson's irises.* The R Journal 14(1), 244--254.
  <https://doi.org/10.32614/RJ-2022-020>.
* Scott, D. W. (1992). *Multivariate Density Estimation: Theory,
  Practice, and Visualization.* Wiley.
* Scrucca, L., Fop, M., Murphy, T. B. and Raftery, A. E. (2016). *mclust 5:
  Clustering, classification and density estimation using Gaussian finite
  mixture models.* The R Journal 8(1), 289--317.
  <https://doi.org/10.32614/RJ-2016-021>.
* Silverman, B. W. (1986). *Density Estimation for Statistics and Data
  Analysis.* Chapman and Hall.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <https://doi.org/10.1080/07362994.2024.2372605>.
* Wand, M. P. and Jones, M. C. (1995). *Kernel Smoothing.* Chapman and
  Hall.

## Reproduce

The ambient seed is 20260601, `hellinger_mc()` draws with
`seed = 20260926L`, and every `from_kde()` call passes `seed = 1L`, so the
importance draws do not depend on the ambient random-number state.

The numerical illustration sets `set.seed(20260925)` before the competitor
fits on each dataset, fixing the random starts of np and mixtools. In the
simulation each dataset follows `set.seed()` with its index, also its
`from_kde()` seed. The stored results come from the code shown, run on
`r sim$run_date` with R `r sim$versions[["R"]]`, proxymix
`r desc_version("proxymix")`, ks `r desc_version("ks")`, np
`r desc_version("np")`, mclust `r desc_version("mclust")`, mixtools
`r desc_version("mixtools")` and KernSmooth
`r desc_version("KernSmooth")`, and the run raised
`r length(sim$warnings)` warnings.

```{r session-info, eval = FALSE}
sessionInfo()
```

```{r session-info-print, echo = FALSE}
## printed 68 characters wide; a longer line breaks after its last slash or
## space within the width and continues indented
old_width <- options(width = 68L)
si_vec <- capture.output(sessionInfo())
options(old_width)
wrap_line <- function(s) {
  out <- character(0L)
  while (nchar(s) > 68L) {
    cut <- max(gregexpr("[/ ]", substr(s, 1L, 68L))[[1L]])
    if (cut < 20L) cut <- 68L
    out <- c(out, substr(s, 1L, cut))
    s <- paste0("    ", substr(s, cut + 1L, nchar(s)))
  }
  c(out, s)
}
writeLines(unlist(lapply(si_vec, wrap_line)))
```
