---
title: "One mixture, many methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One mixture, many methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~25 s under rmarkdown::render() with ggplot2, glmnet,
     MASS, grf, tmle, MatchIt, mgcv and sandwich installed, the simulation
     and the tmle reference read from their stored results; macOS arm64
     (Apple silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{=latex}
\providecommand{\pmNeedspace}[1]{\par\penalty-100\ifdim\dimexpr#1\relax>\dimexpr\pagegoal-\pagetotal\relax\newpage\fi}
```

## The problem

An analyst has fitted a Gaussian mixture over a table of data, and now has a
week of ordinary work in front of her: a regression, a smoother, a
clustering, a principal-components plot, a shrunk coefficient, and a
treatment effect that she suspects varies from unit to unit. Each of those
normally means reaching for a different package. Her question is: *I already
have one fitted mixture over these columns -- which of the tools I would
otherwise reach for does it actually replace, how do I know the answer is
the same one, and what do I give up by using it instead?*

The short answer is that conditioning, marginalisation and
eigendecomposition are all exact on a Gaussian mixture, so six familiar
analyses fall out of one family of fitted mixtures in closed form, differing
in the number and width of the components. Four of them reproduce the
classical answer to machine precision, the treatment effect reproduces a
two-model learner to the tolerance of the fit, and the clustering is a
model-based clustering rather than a copy of `kmeans()`. The rest of this
vignette shows each substitution, checks it against the tool it stands in
for, and states the cost.

## Package capabilities

A `gmm` is a finite Gaussian mixture: weights, means and covariances. A
data matrix becomes a target through `gmm_target_from_samples()`, and
`fit_proxymix()` fits the mixture to it, in the closed-form `"moment"`
regime at one component or the classical expectation-maximisation
`"sample"` regime at more than one.

`gmm_conditionalise()` returns the conditional mixture given values on
some coordinates, in closed form by the Schur complement (van der Hoek and
Elliott, 2024). That single
operator carries the regression, the kernel smoother, the imputation and the
ridge estimator below: a conditional mean read off a joint fit is a
prediction, and which classical estimator it coincides with depends only on
how many components the mixture has and how wide they are.

`bic_aic()` scores a component count. The `ridge_eps` argument of the
fitting verbs adds a multiple of the identity to each component covariance,
which is where the penalised-regression section comes from.

The decision family reads one joint fit as a causal object.
`fit_uplift()` assembles a mixture over outcome, treatment and covariates
and records the roles and the treatment arms it observed;
`proxy_cate()` reads the conditional average treatment effect per unit,
`proxy_decide()` turns it into an action under a value and a cost, and
`proxy_identification_report()` states what the reading assumes and what
it cannot deliver.

## Addressing the problem

Each section below keeps the same shape: the **problem**, the **usual
tool**, the **proxymix route**, the code, and the **idea** behind the
substitution. The trade-offs are gathered in *Interpretation*.

```{r shared-helpers}
## The conditional mean of the outcome coordinate, gated across components:
## sum_k pi_k(x) mu_k^{y|x}. This one helper serves the regression, the
## kernel-regression and the ridge sections.
cond_mean <- function(fit, xv) {
  vapply(xv, function(xx) {
    g <- gmm_conditionalise(fit, given = c(NA, xx))
    sum(g@weights * vapply(g@means, function(m) m[1L], numeric(1L)))
  }, numeric(1L))
}
```

```{r sci-notation, include = FALSE}
## Differences at the double-precision floor are typeset as LaTeX powers of
## ten, so that an inline value reads as mathematics rather than as R's
## own exponent notation.
sci <- function(v, digits = 2L) {
  s <- formatC(v, format = "e", digits = digits - 1L)
  paste0("$", sub("e.*", "", s), " \\times 10^{",
         as.integer(sub(".*e", "", s)), "}$")
}
sci_plain <- function(v) formatC(v, format = "e", digits = 1L)
```

### Regression: from `lm()` to a conditioned mixture

**Problem.** Predict an outcome $y$ from a predictor $x$. **Usual tool.**
`lm()` or `glm()`, one global line. **proxymix route.** Fit a mixture over
the joint $(y, x)$, then read $E[y \mid x]$ by conditioning. At $K = 1$ this
*is* least squares; at $K > 1$ the gate bends the line.

```{r reg-fit}
set.seed(20260617)
n <- 400L
x <- runif(n, -3, 3)
y <- 0.3 * x + 1.2 * pmax(x, 0) + rnorm(n, sd = 0.4)  # a bent relationship
dat <- data.frame(y = y, x = x)

joint <- gmm_target_from_samples(cbind(y, x))
fit1 <- fit_proxymix(joint, N = 1L, regime = "moment", ridge_eps = 0)
fit3 <- fit_proxymix(joint, N = 3L, regime = "sample", max_iter = 150L)

## At K = 1 the conditional slope should equal the lm coefficient exactly.
slope_mix <- gmm_conditionalise(fit1, given = c(NA, 1))@means[[1L]] -
  gmm_conditionalise(fit1, given = c(NA, 0))@means[[1L]]
slope_lm <- unname(coef(lm(y ~ x, dat))["x"])
diff_reg <- abs(slope_mix - slope_lm)
```

```{r fig-reg, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "One global line (K = 1, which is ordinary least squares) against a conditioned three-component mixture (K = 3) on data whose true relationship is bent at the origin. The mixture follows the bend; the single line cannot.", fig.alt = "Scatter of y against x with a straight ordinary-least-squares line and a curved mixture conditional mean that follows a bend in the data."}
grid_reg <- data.frame(x = seq(-3, 3, length.out = 200L))
grid_reg$ols <- as.numeric(predict(lm(y ~ x, dat), newdata = grid_reg))
grid_reg$mix <- cond_mean(fit3, grid_reg$x)
ggplot2::ggplot() +
  ggplot2::geom_point(data = dat, ggplot2::aes(x, y), colour = "grey60",
                      alpha = 0.4, size = 0.7) +
  ggplot2::geom_line(data = grid_reg,
                     ggplot2::aes(x, ols, colour = "OLS (K = 1)"),
                     linewidth = 0.9) +
  ggplot2::geom_line(data = grid_reg,
                     ggplot2::aes(x, mix, colour = "mixture (K = 3)"),
                     linewidth = 0.9) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("OLS (K = 1)" = "#0072B2", "mixture (K = 3)" = "#D55E00")
  ) +
  ggplot2::labs(
    x = "x", y = "y",
    title = "Regression: a global line against a conditioned mixture"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-reg-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the regression figure is",
    "skipped.\n")
```

**Idea.** Each component supplies an affine conditional mean
$\mu_k^{y \mid x}$; the responsibilities $\pi_k(x)$ gate them, so the global
mean $\sum_k \pi_k(x)\, \mu_k^{y \mid x}$ is nonlinear. This is
Gaussian-mixture regression. Unlike a mixture of linear regressions
(de Veaux, 1989), whose component weights do not depend on $x$, its gates
do.

### Kernel regression: from Nadaraya-Watson to a conditioned KDE

**Problem.** Predict $y$ from $x$ with no global functional form, a purely
local smoother. **Usual tool.** Nadaraya-Watson kernel regression
(Nadaraya, 1964; Watson, 1964; `stats::ksmooth`, or np, Hayfield and
Racine, 2008): a kernel-weighted local average of the responses. **proxymix route.** Place one Gaussian per data point over the
joint $(y, x)$ -- a kernel density estimate -- and read $E[y \mid x]$ by
conditioning. With a Gaussian kernel that conditional mean *is* the
Nadaraya-Watson estimator, exactly. Where `lm()` sat at $K = 1$, one global
line, Nadaraya-Watson sits at $K = n$, one component per datum and fully
local, and `from_kde()` lands anywhere in between: a single axis with least
squares at one end and kernel smoothing at the other.

```{r nw-fit}
h <- 0.4                                     # kernel bandwidth
nw <- function(xq) {
  vapply(xq, function(q) {
    w <- dnorm(q, x, h)                      # Gaussian kernel weights
    sum(w * y) / sum(w)
  }, numeric(1L))
}

## One Gaussian per datum, a kernel density estimate of the joint (y, x);
## then condition on x and read the conditional mean.
kde <- gmm(weights = rep(1 / n, n),
           means = lapply(seq_len(n), function(i) c(y[i], x[i])),
           covariances = rep(list(diag(c(h^2, h^2))), n))
xq <- seq(-2.5, 2.5, length.out = 21L)
diff_nw <- max(abs(nw(xq) - cond_mean(kde, xq)))
```

```{r fig-nw, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "One axis, two endpoints: the global line at K = 1 (ordinary least squares) and the fully local kernel smoother at K = n (Nadaraya-Watson), both read off a mixture by the same conditioning operation.", fig.alt = "Scatter of y against x with the straight ordinary-least-squares line and the curved Nadaraya-Watson kernel-regression line."}
grid_nw <- data.frame(x = seq(-3, 3, length.out = 200L))
grid_nw$ols <- as.numeric(predict(lm(y ~ x, dat), newdata = grid_nw))
grid_nw$nw <- nw(grid_nw$x)
ggplot2::ggplot() +
  ggplot2::geom_point(data = dat, ggplot2::aes(x, y), colour = "grey60",
                      alpha = 0.4, size = 0.7) +
  ggplot2::geom_line(data = grid_nw,
                     ggplot2::aes(x, ols, colour = "OLS (K = 1)"),
                     linewidth = 0.9) +
  ggplot2::geom_line(data = grid_nw,
                     ggplot2::aes(x, nw, colour = "kernel (K = n)"),
                     linewidth = 0.9) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("OLS (K = 1)" = "#0072B2", "kernel (K = n)" = "#D55E00")
  ) +
  ggplot2::labs(
    x = "x", y = "y",
    title = "From global line to local kernel, one conditioning step"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-nw-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the kernel-regression",
    "figure is skipped.\n")
```

**Idea.** Conditioning a per-point kernel density estimate gives
responsibilities $\pi_i(x) \propto \mathcal{N}(x \mid x_i, h^2)$ and
per-component means $y_i$, so
$E[y \mid x] = \sum_i \pi_i(x)\, y_i$, the Nadaraya-Watson ratio. Kernel
regression is therefore Gaussian-mixture regression with one component per
datum, a diagonal bandwidth, and only the conditional mean read off. The
bandwidth $h$ plays the role of the component covariance.

### Clustering: from `kmeans()` to the mixture components

**Problem.** Group rows by similarity. **Usual tool.** `kmeans()`, or
`mclust`, which fits exactly this model (Fraley and Raftery, 2002).
**proxymix route.**
`fit_proxymix(N = K, regime = "sample")`: the components *are* the clusters
and the responsibilities are soft assignments.

```{r clust-fit}
set.seed(20260617)
x_clust <- rbind(
  mvnfast::rmvn(150L, c(-2, -1), 0.5 * diag(2)),
  mvnfast::rmvn(150L, c(2, 0), matrix(c(0.6, 0.3, 0.3, 0.4), 2L)),
  mvnfast::rmvn(150L, c(0, 2.5), 0.3 * diag(2))
)
colnames(x_clust) <- c("V1", "V2")
target_clust <- gmm_target_from_samples(x_clust)
fitc <- fit_proxymix(target_clust, N = 3L, regime = "sample",
                     max_iter = 150L)

## Soft assignment = posterior responsibility of each component.
responsibilities <- function(fit, xx) {
  comp <- vapply(seq_len(gmm_n_components(fit)), function(k) {
    fit@weights[k] *
      mvnfast::dmvn(xx, mu = fit@means[[k]], sigma = fit@covariances[[k]])
  }, numeric(nrow(xx)))
  comp / rowSums(comp)
}
resp <- responsibilities(fitc, x_clust)
mean_confidence <- mean(apply(resp, 1L, max))
```

```{=latex}
\par\vskip 0pt plus 10\baselineskip\penalty-100\vskip 0pt plus -10\baselineskip
\vskip 10\baselineskip\penalty9999\vskip -10\baselineskip
```

```{r clust-table, echo = FALSE}
knitr::kable(
  formatC(head(resp, 4L), format = "f", digits = 3L),
  col.names = paste("component", seq_len(3L)), align = "r",
  caption = paste0(
    "Posterior responsibilities of the three components for the first four ",
    "rows, rounded to three decimals. Each row is 1 for one component and 0 ",
    "for the others because the three clusters are well separated."
  )
)
```

**Idea.** Soft, model-based clustering: a point's assignment to cluster $k$
is its posterior responsibility
$\pi_k\,\mathcal{N}(x \mid \mu_k, \Sigma_k) /
\sum_j \pi_j\,\mathcal{N}(x \mid \mu_j, \Sigma_j)$.

### Principal components: from `prcomp()` to the fitted covariance

**Problem.** Find the directions of greatest variation. **Usual tool.**
`prcomp()`. **proxymix route.** Fit a single Gaussian (`N = 1`,
`regime = "moment"`) and eigendecompose its covariance.

```{r pca}
fit_pca <- fit_proxymix(target_clust, N = 1L, regime = "moment",
                        ridge_eps = 0)
ev <- eigen(fit_pca@covariances[[1L]])$vectors
pr <- prcomp(x_clust)$rotation
## Identical to the prcomp rotation, up to the sign convention of each axis.
diff_pca <- max(abs(abs(ev) - abs(unname(pr))))
```

```{r fig-pca, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "Clustering (colour) and the principal axes (arrows) read off the same data by two different component counts: three components give the clusters, one component gives the covariance whose eigenvectors are the principal axes.", fig.alt = "Three coloured point clusters with two principal-axis arrows drawn from the overall centroid."}
vals <- eigen(fit_pca@covariances[[1L]])$values
mu_pca <- unname(fit_pca@means[[1L]])
axes <- data.frame(
  x = mu_pca[1L], y = mu_pca[2L],
  xend = mu_pca[1L] + 2 * sqrt(vals) * ev[1L, ],
  yend = mu_pca[2L] + 2 * sqrt(vals) * ev[2L, ]
)
pts <- data.frame(x_clust, cluster = factor(max.col(resp)))
ggplot2::ggplot() +
  ggplot2::geom_point(data = pts,
                      ggplot2::aes(V1, V2, colour = cluster),
                      alpha = 0.6, size = 0.9) +
  ggplot2::geom_segment(
    data = axes, ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.2, "cm")), linewidth = 0.8
  ) +
  ggplot2::scale_colour_viridis_d(name = "cluster", end = 0.85) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = expression(x[1]), y = expression(x[2]),
                title = "Clustering and principal axes from two fits") +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-pca-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the clustering and",
    "principal-axes figure is skipped.\n")
```

**Idea.** Principal-component analysis is the eigendecomposition of the
covariance (Jolliffe, 2002). The moment-matched single Gaussian carries the
sample covariance exactly, so its eigenvectors are the principal components,
as in the maximum-likelihood solution of probabilistic principal component
analysis (Tipping and Bishop, 1999); with
$K > 1$ the same read gives a local analysis inside each cluster.

### Penalised regression: ridge as a covariance ridge

**Problem.** Stabilise a regression when predictors are collinear. **Usual
tool.** `glmnet` (ridge or lasso), `MASS::lm.ridge`. **proxymix route.** The
`ridge_eps` argument adds $\lambda I$ to the component covariance, which
shrinks the conditional regression. The result is exactly the ridge
estimator of Hoerl and Kennard (1970), an $L_2$ penalty, with penalty
$(n - 1)\lambda$ on the centred cross-products, because the covariance
divides them by $n - 1$.

```{r ridge}
lambda <- c(0, 0.5, 2, 8)
slope_pm <- vapply(lambda, function(lam) {
  f <- fit_proxymix(joint, N = 1L, regime = "moment", ridge_eps = lam)
  gmm_conditionalise(f, given = c(NA, 1))@means[[1L]] -
    gmm_conditionalise(f, given = c(NA, 0))@means[[1L]]
}, numeric(1L))
slope_formula <- cov(x, y) / (var(x) + lambda)
diff_ridge <- max(abs(slope_pm - slope_formula))
```

```{r ridge-table, echo = FALSE}
knitr::kable(
  data.frame(lambda = lambda, proxymix = slope_pm,
             ridge_formula = slope_formula),
  digits = 4L,
  col.names = c("Penalty $\\lambda$", "proxymix slope",
                "Ridge closed form"),
  caption = paste0(
    "The conditional slope under a covariance ridge, against the ridge ",
    "estimator $\\mathrm{cov}(x, y) / (\\mathrm{var}(x) + \\lambda)$."
  )
)
```

**Idea.** Conditioning gives $\beta = \Sigma_{xx}^{-1}\Sigma_{xy}$. Ridging
$\Sigma_{xx} \to \Sigma_{xx} + \lambda I$ yields
$(\Sigma_{xx} + \lambda I)^{-1}\Sigma_{xy}$, the ridge estimator, so the
slope shrinks toward zero as $\lambda$ grows.

### Treatment effects: from a two-model learner to one joint fit

**Problem.** Estimate a treatment effect that varies from unit to unit, and
then act on it. **Usual tool.** A two-model T-learner (Künzel et al., 2019),
fitting the treated and control arms separately and subtracting, or a
causal tree or forest (Athey and Imbens, 2016; Athey et al., 2019; `grf`);
for the average effect alone, double machine learning (DoubleML; Bach et
al., 2024). **proxymix route.** `fit_uplift()` fits one joint mixture over
(outcome, treatment, covariates), and every decision verb reads that single
fit in closed form. Nothing re-fits per query.

The design below is synthetic, chosen so that the truth is available as an
oracle: the treatment effect is the known linear function
$\tau(x) = 0.5 + x$, so the estimate can be scored against it.

```{r uplift-fit}
set.seed(20260902)
n_up <- 600L
x_up <- rnorm(n_up)
t_up <- rbinom(n_up, 1L, 0.5)
tau_true <- function(v) 0.5 + v
y_up <- 1 + tau_true(x_up) * t_up + rnorm(n_up, sd = 0.5)
dat_up <- data.frame(y = y_up, t = t_up, x = x_up)

model <- fit_uplift(dat_up, "y", "t", "x", N = 2L, regime = "sample",
                    max_iter = 80L, seed = 1L)
model
```

`proxy_cate()` reads the conditional average treatment effect, with a
standard error and an interval, at whatever units are handed to it.

```{r uplift-cate}
grid_up <- data.frame(x = seq(-2, 2, length.out = 41L))
cate <- proxy_cate(model, grid_up)
err_cate <- max(abs(cate$tau - tau_true(grid_up$x)))

## The two-model learner: one least-squares line per arm, as one
## interacted regression whose t and t:x terms are the per-arm differences.
fit_arms <- lm(y ~ t * x, data = dat_up)
b_arms <- coef(fit_arms)[c("t", "t:x")]
diff_arms <- max(abs(cate$tau - (b_arms[[1L]] + b_arms[[2L]] * grid_up$x)))
A_arms <- cbind(0, 1, 0, grid_up$x)
se_arms <- sqrt(rowSums((A_arms %*% vcov(fit_arms)) * A_arms))
se_ratio <- range(cate$se / se_arms)
## departure of intercept and slope from the truth, in standard errors
z_arms <- (b_arms - c(0.5, 1)) / sqrt(diag(vcov(fit_arms)))[c("t", "t:x")]
arm_of_component <- vapply(model@fit@means, function(m) {
  m[model@roles$treatment]
}, numeric(1L))
```

```{r fig-cate, eval = has_ggplot2, echo = has_ggplot2, fig.cap = paste0("The conditional average treatment effect read off one joint mixture, with its delta-method interval, against the analytic truth that generated the data. The estimate is the straight line of intercept ", sprintf("%.3f", b_arms[[1L]]), " and slope ", sprintf("%.3f", b_arms[[2L]]), ", the difference of the two per-arm least-squares lines; the truth has intercept 0.5 and slope 1."), fig.alt = "Estimated conditional average treatment effect against the covariate, shown as a straight line with a shaded confidence band, overlaid with a dashed line for the analytic truth; the two nearly coincide, the estimate slightly steeper and slightly lower in the middle of the range."}
cate_df <- as.data.frame(cate)
cate_df$x <- grid_up$x
cate_df$truth <- tau_true(grid_up$x)
ggplot2::ggplot(cate_df, ggplot2::aes(x)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ci_lo, ymax = ci_hi),
                       fill = "#56B4E9", alpha = 0.35) +
  ggplot2::geom_line(ggplot2::aes(y = tau, colour = "proxymix estimate"),
                     linewidth = 0.9) +
  ggplot2::geom_line(ggplot2::aes(y = truth, colour = "analytic truth"),
                     linewidth = 0.9, linetype = "dashed") +
  ggplot2::geom_hline(yintercept = 0, colour = "grey60",
                      linewidth = 0.3) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("proxymix estimate" = "#0072B2",
               "analytic truth" = "#D55E00")
  ) +
  ggplot2::labs(
    x = "covariate x", y = "treatment effect on y",
    title = "A conditional treatment-effect surface from one joint fit"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-cate-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the treatment-effect",
    "figure is skipped.\n")
```

Acting on the estimate is `proxy_decide()`, which turns an effect into a
recommended arm under a value per unit of outcome and a cost of treating.

```{r uplift-decide}
decision <- proxy_decide(model, grid_up, value = 1, cost = 0.5)
switch_x <- grid_up$x[min(which(decision$action == 1L))]
```

```{r uplift-decide-table, echo = FALSE}
sel <- c(1L, 11L, 21L, 31L, 41L)
knitr::kable(
  data.frame(
    x = grid_up$x[sel],
    tau = cate$tau[sel],
    truth = tau_true(grid_up$x[sel]),
    action = decision$action[sel],
    expected_value = decision$expected_value[sel]
  ),
  digits = 3L,
  col.names = c("Covariate x", "Estimated effect", "True effect",
                "Recommended arm", "Net value"),
  caption = paste0(
    "Five points of the decision surface at a value of 1 per unit of ",
    "outcome and a treatment cost of 0.5. The net value is that of ",
    "treating: the estimated effect times the value, minus the cost, ",
    "negative where arm 0 is recommended."
  )
)
```

**A refusal rather than a wrong scale.** The model recorded the treatment
arms it actually saw. Asking for a contrast between arms the data never
contained is refused, with a typed condition, instead of being answered on a
silently rescaled axis.

```{r uplift-refusal}
refusal <- tryCatch(
  proxy_cate(model, grid_up, t1 = 100, t0 = 0),
  error = function(e) e
)
class(refusal)
msg_lines <- strsplit(conditionMessage(refusal), "\n")[[1L]]
writeLines(strwrap(msg_lines, width = 70L, exdent = 2L))
```

The identification report is the artefact that travels with a decision: it
names the estimand, the assumption it rests on, the overlap of the units
supplied, the magnitude at risk if a latent regime confounds the effect, and
what the model explicitly does not identify.

```{r uplift-report}
proxy_identification_report(model, grid_up)
```

On these data the confounding gap in the report equals the whole estimated
effect and is not a separate check on it. Each of the two components holds
one treatment arm, and the effect within a component is zero.

**Idea.** Under ignorability the conditional treatment effect is a contrast
of two conditional means of the same joint mixture, $E[y \mid t_1, x]$ minus
$E[y \mid t_0, x]$, so the operator that produced the regression above
produces the causal contrast as well. What changes is not the algebra but
the assumption the reader is asked to accept, which is why the family ships
a report that states it.

### A few more, in one line each

Imputation is `gmm_conditionalise()` on the observed coordinates, giving a
conditional-mean or, through `rgmm()`, a sampled fill-in. Density
estimation is the native use, and `from_kde()` compresses a kernel estimate
into a closed-form mixture. Kalman filtering is `gmm_observe()`, the mixture
form of the Kalman update.

## Interpretation

Four of the six substitutions reproduce the classical answer to machine
precision, and the treatment effect reproduces the two-model learner to the
tolerance of the fit. That is the part worth checking rather than
assuming.

```{r equality-table, echo = FALSE}
knitr::kable(
  data.frame(
    claim = c(
      "$K = 1$ conditional slope equals the lm coefficient",
      "Per-point KDE conditional mean equals Nadaraya-Watson",
      "$N = 1$ eigenvectors equal the prcomp rotation",
      "Ridged conditional slope equals the ridge closed form",
      "Two-component effect equals the per-arm least-squares contrast"
    ),
    difference = sci_plain(c(diff_reg, diff_nw, diff_pca, diff_ridge,
                             diff_arms)),
    stringsAsFactors = FALSE
  ),
  col.names = c("Claim", "Maximum absolute difference"),
  caption = paste0(
    "Each equality claim checked against the tool it stands in for, on ",
    "the data fitted above."
  )
)
```

**Regression.** At one component the conditional slope and the `lm`
coefficient differ by `r sci(diff_reg)`, which is the double-precision
floor: the substitution is an identity, not an approximation. At three
components the same conditioning operation follows the bend in the figure
that the single line cannot. What this buys is nonlinear, heteroscedastic,
even multimodal conditional *densities* rather than only a mean. What it
costs is classical inference: no standard errors or $p$-values come back by
default, and the components are Gaussian.

**Kernel regression.** The conditioned per-point kernel density estimate and
the Nadaraya-Watson smoother agree to `r sci(diff_nw)` over the
`r length(xq)` query points. The same operation also returns the full
conditional density -- predictive variance, quantiles, multimodality --
which the bare Nadaraya-Watson mean does not, and it works in the
evaluate-only regime where there are no points to average and so no literal
kernel estimator exists. The price is that the one-per-datum estimate costs
$O(n)$ per query, which `from_kde()` compresses to $O(K)$, and that the
bandwidth must still be chosen.

**Clustering.** The three components recover the three generated clouds, and
the assignments are nearly hard: the mean largest responsibility across the
`r nrow(x_clust)` rows is `r round(mean_confidence, 3)`, so on
well-separated data the soft assignment is decisive. The gain over
`kmeans()` is ellipsoidal rather than spherical clusters, a soft assignment
that carries its own uncertainty, and a density estimate as a by-product.
The cost is that the component count is not chosen for you, which is what
`bic_aic()` is for, and that `kmeans()` is faster at very large $n$.

**Principal components.** The eigenvectors of the one-component covariance
and the `prcomp` rotation differ by `r sci(diff_pca)` once the sign
convention of each axis is accounted for. At $K > 1$ the same read gives a
local analysis inside each cluster, which `prcomp()` does not offer. For a
single linear projection `prcomp()` remains more direct, since loadings,
scree plots and biplots come with it.

**Ridge.** The slope under a covariance ridge matches the closed-form ridge
estimator to `r sci(diff_ridge)` across all
`r length(lambda)` penalties, and shrinks from
`r round(slope_pm[1], 3)` at no penalty to
`r round(slope_pm[length(lambda)], 3)` at the largest. What is delivered is
$L_2$ shrinkage and numerical stability from the object already in hand.
What is not available is the lasso: $L_2$ shrinkage matches a Gaussian
prior, $L_1$ sparsity a Laplace prior outside the Gaussian family, so this
route does not perform variable selection.

**Treatment effects.** The two fitted components are the two treatment
arms: their treatment coordinates are
`r paste(sort(round(arm_of_component, 3)), collapse = " and ")`, and their
weights `r paste(sprintf("%.3f", sort(model@fit@weights, decreasing = TRUE)), collapse = " and ")`
are the arm shares. Each component carries its own arm's means and
covariance of $(y, x)$, so conditioning on an arm and on $x$ returns that
arm's least-squares line, and the estimated effect is the difference of the
two lines: the two-model learner with a linear model in each arm. The
estimate and the per-arm least-squares contrast differ by at most
`r sci(diff_arms)` over the grid, the convergence tolerance of the
expectation-maximisation fit, and the delta-method standard error is
`r sprintf("%.2f", se_ratio[1L])` to `r sprintf("%.2f", se_ratio[2L])` times
the regression's own. The linear model in each arm is correctly specified
for this design. The estimated line has intercept
`r sprintf("%.3f", b_arms[[1L]])` and slope
`r sprintf("%.3f", b_arms[[2L]])`, against the true 0.5 and 1: the intercept
is `r sprintf("%.1f", abs(z_arms[[1L]]))` standard errors from the truth and
the slope `r sprintf("%.1f", abs(z_arms[[2L]]))`, the sampling error of one
dataset of `r n_up` units. Across the covariate range the estimate departs
from the truth by at most `r round(err_cate, 3)`, on an effect that runs from
`r round(tau_true(min(grid_up$x)), 1)` to
`r round(tau_true(max(grid_up$x)), 1)`. Turned into an action at a value of 1
and a cost of 0.5, the recommendation switches from control to treatment at
`r round(switch_x, 2)`, against the true break-even at 0. The share of grid
points whose interval contains the truth is not a coverage rate: the
`r nrow(grid_up)` points share one dataset and one fitted line, so they miss
or cover together. The simulation in the numerical illustration measures the
coverage of the average effect over independent datasets.

The refusal is the other half of that section. Asking for a contrast
between treatment values of 100 and 0, on a model fitted to arms 0 and 1,
returns a condition of class `proxymix_treatment_scale_error` and
`orchestra_refusal` rather than a number: a decision verb declines a
question the fit cannot support instead of answering it on an axis the data
never contained.

```{r summary-table, echo = FALSE}
summary_method <- c("Regression", "Kernel regression", "Clustering",
                    "Principal components", "Ridge", "Treatment effects")
knitr::kable(
  data.frame(
    method = summary_method,
    usual = c("`lm()`, `glm()`", "`ksmooth()`, np", "`kmeans()`, mclust",
              "`prcomp()`", "glmnet, `lm.ridge()`",
              "T-learner, grf, DoubleML"),
    route = c("mixture over $(y, x)$, then `gmm_conditionalise()`",
              "per-point kernel estimate, then `gmm_conditionalise()`",
              "`fit_proxymix()` in the sample regime",
              "`eigen()` of the one-component covariance",
              "`ridge_eps` on the covariance",
              "`fit_uplift()`, then `proxy_cate()`"),
    stringsAsFactors = FALSE
  ),
  col.names = c("Analysis", "Usual tool", "proxymix route"),
  caption = paste0("Six analyses from one family of fitted mixtures: ",
                   "the tool each replaces.")
)
knitr::kable(
  data.frame(
    method = summary_method,
    gain = c("nonlinear, heteroscedastic, multimodal conditionals",
             "full conditional density; the evaluate-only regime",
             "soft, ellipsoidal, density-aware clusters",
             "a local analysis per cluster",
             "$L_2$ shrinkage from the same fit",
             "one fit, plus a stated identification contract"),
    give_up = c("built-in inference; Gaussian components",
                "$O(n)$ per query unless compressed; bandwidth choice",
                "automatic component count; speed at very large $n$",
                "loadings, scree and biplot conveniences",
                "$L_1$ sparsity and variable selection",
                "per-unit accuracy when an arm needs several components"),
    stringsAsFactors = FALSE
  ),
  col.names = c("Analysis", "What you gain", "What you give up"),
  caption = paste0("Six analyses from one family of fitted mixtures: ",
                   "the gain and the cost of each.")
)
```

The answer to the opening question, then: one family of fitted mixtures
replaces the linear model, the kernel smoother, the principal-components
step and the ridge penalty exactly, reproduces the two-model
treatment-effect learner on this design, and gives a model-based clustering
in place of `kmeans()`. What is given up is inference machinery for the
regression reads and the freedom of a non-Gaussian component shape.

## Limitations

Five of the substitutions reproduce the tool they stand in for, so they
cannot be more accurate than it, only more convenient. The clustering is a
different method from `kmeans()`. The treatment-effect reading reproduces
the two-model learner on the design above, where the component counts are
supplied by hand. In the simulation below, where BIC chooses them, its
per-unit effect is less accurate than the purpose-built learners'. Use the purpose-built tool where its strengths are what is
needed: high-dimensional sparse problems suit `glmnet`, very large-$n$
clustering suits `kmeans()`, a single linear projection suits `prcomp()`,
and classical inferential guarantees suit `lm()` or `glm()`. A single
mixture is the simpler choice when one fitted object should serve several of
these at once, in closed form, and when Gaussian components fit the data. The numerical illustration that follows checks the four identities on real data and places the treatment-effect reading beside four purpose-built estimators, on a job-training study and in simulation.

Every fit on this page has samples to work from. The one-component fits
use the closed-form `"moment"` regime and the others the `"sample"` regime.
Before the numerical illustration the component counts are supplied by
hand; the job-training and simulation fits choose theirs by BIC. Choosing
$K$ is a real decision that `bic_aic()` informs, and a count much above the
number of genuine structures will fit noise into the conditional law.

The treatment-effect section carries two further limits. First, the
delta-method standard error holds the regime gate fixed and prices the
sampling variability of the components only. Here the gate is fixed by the
arm, and the standard error agrees with the regression's; where an arm
needs several components, a resampling standard error that also reflects
the gate is available through `se_method = "mc"`. Second, with a binary treatment the fitted
mixture separates on the treatment arm itself, so within any one regime the
treatment does not vary and the within-regime treatment slope is zero. This
means `proxy_regime_segments()` returns an effect of zero in every regime on
data of this shape: the effect modification is carried *across* regimes, by
the gate, and is read correctly by `proxy_cate()` but not by the segment
table. A reader wanting interpretable per-segment effects needs a design
where the components separate on the covariate rather than on the arm.

Everything causal here also rests on ignorability, which no goodness-of-fit
statistic can certify, on data generated to satisfy it. The identification
report states the assumption and the magnitude at risk if it fails, and
declines the individual counterfactual law outright; it does not make the
assumption true.

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_grf <- requireNamespace("grf", quietly = TRUE)
has_tmle <- requireNamespace("tmle", quietly = TRUE)
has_matchit <- requireNamespace("MatchIt", quietly = TRUE)
has_mgcv <- requireNamespace("mgcv", quietly = TRUE)
has_sandwich <- requireNamespace("sandwich", quietly = TRUE)
has_glmnet <- requireNamespace("glmnet", quietly = TRUE)
has_mass <- requireNamespace("MASS", quietly = TRUE)
has_penguins <- exists("penguins", envir = asNamespace("datasets"))
run_classical <- has_glmnet && has_mass && has_penguins
run_treatment <- has_grf && has_tmle && has_matchit && has_mgcv &&
  has_sandwich
```

```{r illustration-skip, eval = !(run_classical && run_treatment), echo = FALSE, results = "asis"}
cat("One of glmnet, MASS, grf, tmle, MatchIt, mgcv, sandwich or the penguins",
    "data (R 4.5.0 or later) is not available on this build, so part of the",
    "real-data illustration is skipped.\n")
```

```{r fixed-notation, include = FALSE}
## Small values are written as plain decimals rather than in the scientific
## notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
## warnings and messages wrapped to fit the page
options(width = 70L)
## a version as its DESCRIPTION writes it
ver <- function(pkg) utils::packageDescription(pkg)$Version
## a stored version in DESCRIPTION form, when the installed one is the one
## stored
desc_version <- function(pkg, stored) {
  if (identical(as.character(packageVersion(pkg)), stored)) ver(pkg) else stored
}
```

### The classical substitutions on the Palmer penguins

The equalities above were checked on simulated data. Here they are checked
again on the Palmer penguins data (Gorman et al., 2014),
shipped with R as `penguins` since version 4.5.0 (Horst et al.,
2022), against the tool each substitution stands in for:
`lm()`, `ksmooth()`, `kmeans()` and `prcomp()` from R itself,
`MASS::lm.ridge()` (Venables and Ripley, 2002) and `glmnet` (Friedman et al.,
2010). The score is the same for every row, the
largest absolute difference between the mixture read and the tool's
answer, so the table reports agreement and not a ranking.

The two contributed packages install from CRAN once:

```{r install-classical, eval = FALSE}
install.packages(c("MASS", "glmnet"))
```

```{r penguins-data, eval = run_classical}
cols <- c("bill_len", "bill_dep", "flipper_len", "body_mass")
peng <- na.omit(penguins[, c(cols, "species")])
X_peng <- as.matrix(peng[, cols])
n_peng <- nrow(X_peng)
```

The four body measurements of the `r n_peng` birds measured in full are
used, in their own units of millimetres and grams.

Body mass regressed on the three other measurements is the regression and
the ridge case, with the mixture's slopes read as differences of the
conditional mean one unit apart. `lm.ridge()` and `glmnet` both standardise
the predictors before penalising and scale the penalty differently, so
the covariance ridge that reproduces a penalty $\lambda$ is
$\lambda / (n - 1)$ for `lm.ridge()` and $\lambda n / ((n - 1) s_y)$ for
`glmnet`, where $s_y$ is the standard deviation of body mass with divisor
$n$.

```{r penguins-regression, eval = run_classical}
cond_slopes <- function(fit, at) {
  base <- cond_mean_p(fit, at)
  vapply(seq_along(at), function(j) {
    step <- at
    step[j] <- step[j] + 1
    cond_mean_p(fit, step) - base
  }, numeric(1L))
}
## The conditional mean of the first coordinate given all the others.
cond_mean_p <- function(fit, xv) {
  g <- gmm_conditionalise(fit, given = c(NA, xv))
  sum(g@weights * vapply(g@means, function(m) m[1L], numeric(1L)))
}

y_peng <- X_peng[, "body_mass"]
X_pred <- X_peng[, c("bill_len", "bill_dep", "flipper_len")]
joint_peng <- gmm_target_from_samples(cbind(y_peng, X_pred))
fit_ols <- fit_proxymix(joint_peng, N = 1L, regime = "moment",
                        ridge_eps = 0)
slope_ols <- cond_slopes(fit_ols, colMeans(X_pred))
coef_lm <- coef(lm(body_mass ~ bill_len + bill_dep + flipper_len,
                   data = peng))[-1L]
diff_lm <- max(abs(slope_ols - coef_lm))

## Standardised predictors, so the two ridge penalties can be matched.
X_std <- scale(X_pred, center = TRUE, scale = FALSE)
scale_x <- sqrt(colSums(X_std^2) / n_peng)
X_std <- sweep(X_std, 2L, scale_x, "/")
s_y <- sqrt(mean((y_peng - mean(y_peng))^2))
joint_std <- gmm_target_from_samples(cbind(y_peng, X_std))
lambda_pen <- 5
ridge_slopes <- function(eps) {
  f <- fit_proxymix(joint_std, N = 1L, regime = "moment", ridge_eps = eps)
  cond_slopes(f, colMeans(X_std)) / scale_x
}
coef_massridge <- coef(MASS::lm.ridge(
  body_mass ~ bill_len + bill_dep + flipper_len, data = peng,
  lambda = lambda_pen))[-1L]
diff_massridge <- max(abs(ridge_slopes(lambda_pen / (n_peng - 1)) -
                            coef_massridge))
coef_glmnet <- as.numeric(coef(glmnet::glmnet(
  X_pred, y_peng, alpha = 0, lambda = lambda_pen,
  control = list(thresh = 1e-20))))[-1L]
diff_glmnet <- max(abs(
  ridge_slopes(lambda_pen * n_peng / ((n_peng - 1) * s_y)) - coef_glmnet))
```

Body mass smoothed on flipper length is the kernel case. `ksmooth()` with
its `"normal"` kernel and a bandwidth of 10 mm is a Gaussian kernel with
standard deviation $0.25 \times 10 / \Phi^{-1}(0.75)$ mm, cut off at four
standard deviations from the query point, so the untruncated
Nadaraya-Watson estimator with the same kernel is also compared.

```{r penguins-kernel, eval = run_classical}
x_fl <- X_peng[, "flipper_len"]
bw_fl <- 10
h_fl <- 0.25 * bw_fl / qnorm(0.75)
kde_peng <- gmm(
  weights = rep(1 / n_peng, n_peng),
  means = lapply(seq_len(n_peng), function(i) c(y_peng[i], x_fl[i])),
  covariances = rep(list(diag(c(h_fl^2, h_fl^2))), n_peng)
)
xq_fl <- seq(180, 225, by = 5)
mix_fl <- vapply(xq_fl, function(q) cond_mean_p(kde_peng, q), numeric(1L))
ks_fl <- ksmooth(x_fl, y_peng, kernel = "normal", bandwidth = bw_fl,
                 x.points = xq_fl)$y
nw_fl <- vapply(xq_fl, function(q) {
  w <- dnorm(q, x_fl, h_fl)
  sum(w * y_peng) / sum(w)
}, numeric(1L))
diff_ksmooth <- max(abs(ks_fl - mix_fl))
diff_nw_peng <- max(abs(nw_fl - mix_fl))
```

Three clusters and the principal axes come from the four measurements
together. `kmeans()` is run with 20 random starts, the mixture with three
components, and the cluster labels are matched by the permutation that
agrees on most birds before the centres are compared.

```{r penguins-clusters, eval = run_classical}
set.seed(20260925)
km_peng <- kmeans(X_peng, centers = 3L, nstart = 20L)
target_peng <- gmm_target_from_samples(X_peng)
fit_peng3 <- fit_proxymix(target_peng, N = 3L, regime = "sample",
                          max_iter = 200L, seed = 1L)
resp_peng <- responsibilities(fit_peng3, X_peng)
lab_mix <- max.col(resp_peng)
perms <- list(c(1L, 2L, 3L), c(1L, 3L, 2L), c(2L, 1L, 3L),
              c(2L, 3L, 1L), c(3L, 1L, 2L), c(3L, 2L, 1L))
agree_lab <- function(a, b) {
  vapply(perms, function(p) mean(p[a] == b), numeric(1L))
}
agree_km <- agree_lab(km_peng$cluster, lab_mix)
best <- perms[[which.max(agree_km)]]
centres_km <- km_peng$centers[order(best), ]
centres_mix <- do.call(rbind, fit_peng3@means)
diff_kmeans <- max(abs(centres_km - centres_mix))
agree_species_mix <- max(agree_lab(as.integer(peng$species), lab_mix))
agree_species_km <- max(agree_lab(as.integer(peng$species),
                                  km_peng$cluster))

fit_peng1 <- fit_proxymix(target_peng, N = 1L, regime = "moment",
                          ridge_eps = 0)
ev_peng <- eigen(fit_peng1@covariances[[1L]])$vectors
pr_peng <- unname(prcomp(X_peng)$rotation)
diff_prcomp <- max(abs(abs(ev_peng) - abs(pr_peng)))
```

```{r penguins-relative, eval = run_classical, include = FALSE}
## largest difference relative to the largest quantity compared, per tool
rel_diff <- c(
  lm = diff_lm / max(abs(coef_lm)),
  prcomp = diff_prcomp / max(abs(pr_peng)),
  lm_ridge = diff_massridge / max(abs(coef_massridge))
)
```

```{=latex}
\par\vskip 0pt plus 16\baselineskip\penalty-100\vskip 0pt plus -16\baselineskip
\vskip 16\baselineskip\penalty9999\vskip -16\baselineskip
```

```{r penguins-parity-table, eval = run_classical, echo = FALSE}
parity_tbl <- data.frame(
  tool = c(paste0("lm(), R ", getRversion()),
           paste0("ksmooth(), R ", getRversion()),
           "Nadaraya-Watson, untruncated kernel",
           paste0("kmeans(), R ", getRversion()),
           paste0("prcomp(), R ", getRversion()),
           paste0("MASS::lm.ridge(), MASS ", ver("MASS")),
           paste0("glmnet ridge, glmnet ", ver("glmnet"))),
  compared = c("three slopes, g per unit",
               "smoothed mass at 10 flipper lengths, g",
               "smoothed mass at 10 flipper lengths, g",
               "three cluster centres, mm and g",
               "four by four rotation, unit free",
               "three slopes at $\\lambda = 5$, g per unit",
               "three slopes at $\\lambda = 5$, g per unit"),
  difference = sci_plain(c(diff_lm, diff_ksmooth, diff_nw_peng,
                           diff_kmeans, diff_prcomp, diff_massridge,
                           diff_glmnet)),
  stringsAsFactors = FALSE
)
knitr::kable(
  parity_tbl, align = c("l", "l", "r"),
  col.names = c("Usual tool", "What is compared",
                "Maximum absolute difference"),
  caption = paste0(
    "Each mixture read against the tool it stands in for, on the ", n_peng,
    " fully measured penguins: the largest absolute difference over the ",
    "quantities compared."
  )
)
```

The regression, principal-component and `lm.ridge()` reads agree with their
tools to a difference, relative to the largest quantity compared, of at
most `r sci(max(rel_diff))`, about
`r floor(-log10(max(rel_diff)))` significant digits; the `glmnet` read
agrees to the coordinate-descent tolerance, an absolute difference of
`r sci(diff_glmnet)`. The kernel read
differs from `ksmooth()` by `r fixed(diff_ksmooth, 2)` g on masses between
`r round(min(ks_fl))` and `r round(max(ks_fl))` g, all of it from the
cut-off, and agrees with the untruncated estimator to `r sci(diff_nw_peng)`.
Apart from the kernel cut-off, clustering is the one row that is not an
identity. `kmeans()` and the
three-component mixture place `r fixed(100 * max(agree_km), 1)` per cent
of the birds in the same cluster and their matched centres differ by up to
`r round(diff_kmeans)` g in the body-mass coordinate, because `kmeans()`
minimises squared Euclidean distance on the raw columns and body mass in
grams dominates that distance. The mixture's clusters coincide with the
three species for `r round(100 * agree_species_mix, 1)` per cent of the
birds and the `kmeans()` clusters for
`r round(100 * agree_species_km, 1)` per cent; neither method was told the
species.

### The treatment effect on a job-training study

The treatment-effect section above used a synthetic design with the truth
known. Here proxymix and four treatment-effect tools estimate the average
effect of a job-training programme on the treated, on the `lalonde` data
shipped with MatchIt (Ho et al., 2011): the 185 programme
participants and 429 comparison men of Dehejia and Wahba (1999), with
earnings in 1978 as the outcome and age, education, race, marital status,
degree status and earnings in 1974 and 1975 as covariates. No true effect
is known for these data, so the estimates and their standard errors stand
side by side, each with the standard error its own method supplies.

The four competitors are a causal forest (grf; Athey et al.,
2019), targeted maximum likelihood estimation (tmle; Gruber and van
der Laan, 2012), nearest-neighbour propensity-score matching followed by
an outcome regression (MatchIt, then `lm()` with a standard error
clustered on the matched pairs by sandwich; Zeileis et al., 2020), and a T-learner of two generalised
additive models (mgcv; Wood, 2011), one per arm, with the effect read as
the average difference of their predictions over the treated men and a
standard error from the two fitted models. They install from CRAN once,
with gam and dbarts for the learners in tmle's SuperLearner library:

```{r install-treatment, eval = FALSE}
install.packages(c("grf", "tmle", "gam", "dbarts", "MatchIt", "mgcv",
                   "sandwich"))
```

```{r load-treatment, eval = run_treatment, message = FALSE}
library(grf)
library(tmle)
library(MatchIt)
library(mgcv)
```

Race is a three-level factor and enters the mixture as two indicators.

```{r lalonde-data, eval = run_treatment}
lal <- MatchIt::lalonde
lal$black <- as.integer(lal$race == "black")
lal$hispan <- as.integer(lal$race == "hispan")
covs <- c("age", "educ", "black", "hispan", "married", "nodegree",
          "re74", "re75")
treated <- lal[lal$treat == 1L, ]
```

proxymix fits one joint mixture with the component count chosen by BIC,
reads the effect for every treated man and averages. Its standard error
comes from `gmm_fit_ensemble()`, which refits the mixture under bootstrap
reweightings of the rows, and `proxy_functional_ci()`, which turns the
ensemble into a percentile interval for the averaged effect.

```{r lalonde-proxymix, eval = run_treatment, message = FALSE}
model_lal <- fit_uplift(lal, "re78", "treat", covs, seed = 1L)
att_of <- function(g) {
  m <- uplift_model(fit = g, roles = model_lal@roles,
                    assume = model_lal@assume,
                    outcome_type = model_lal@outcome_type,
                    data = model_lal@data, n_train = model_lal@n_train,
                    treatment_levels = model_lal@treatment_levels)
  c(att = mean(proxy_cate(m, treated, se = FALSE)$tau))
}
ens_lal <- gmm_fit_ensemble(model_lal@fit, B = 200L, seed = 2L)
ci_lal <- proxy_functional_ci(ens_lal, att_of, level = 0.95)
se_lal <- sd(vapply(ens_lal$members, att_of, numeric(1L)))
```

The ensemble holds `r length(ens_lal$members)` bootstrap refits. The sweep
over one to four components warned twice that the
expectation-maximisation objective decreased, in the three-component fit,
where the 1974 and 1975 earnings are
`r if (round(100 * mean(lal$re74 == 0)) == round(100 * mean(lal$re75 == 0))) paste0("each zero for about ", round(100 * mean(lal$re74 == 0))) else paste0("zero for ", round(100 * mean(lal$re74 == 0)), " and ", round(100 * mean(lal$re75 == 0)))`
per cent of the men and a component
can collapse onto them; the fit selected by BIC has
`r gmm_n_components(model_lal@fit)` component, and the warnings do not
concern it.

```{r lalonde-competitors, eval = run_treatment}
cf_lal <- causal_forest(lal[, covs], lal$re78, lal$treat,
                        num.threads = 1L, seed = 1L)
att_grf <- average_treatment_effect(cf_lal, target.sample = "treated")

match_lal <- matchit(treat ~ age + educ + race + married + nodegree +
                       re74 + re75, data = lal, method = "nearest")
matched <- match.data(match_lal)
fit_matched <- lm(re78 ~ treat + age + educ + race + married + nodegree +
                    re74 + re75, data = matched, weights = weights)
v_matched <- sandwich::vcovCL(fit_matched, cluster = ~subclass)
att_match <- c(coef(fit_matched)[["treat"]],
               sqrt(v_matched["treat", "treat"]))

f_gam <- re78 ~ s(age) + s(educ) + black + hispan + married + nodegree +
  s(re74) + s(re75)
gam1 <- gam(f_gam, data = lal[lal$treat == 1L, ])
gam0 <- gam(f_gam, data = lal[lal$treat == 0L, ])
L1 <- predict(gam1, treated, type = "lpmatrix")
L0 <- predict(gam0, treated, type = "lpmatrix")
a1 <- colMeans(L1)
a0 <- colMeans(L0)
att_gam <- c(sum(a1 * coef(gam1)) - sum(a0 * coef(gam0)),
             sqrt(drop(t(a1) %*% vcov(gam1) %*% a1 +
                         t(a0) %*% vcov(gam0) %*% a0)))
```

```{r ref-read, echo = FALSE}
ref <- readRDS("many_methods_ref.rds")
if (!identical(ref$code, as.character(knitr::knit_code$get("ref-code")))) {
  stop("The stored tmle results were produced by code other than the ",
       "chunk shown; rerun the reference computation.", call. = FALSE)
}
ref_minutes <- round(ref$elapsed_secs / 60, 1)
```

The causal forest warns that its estimated propensities come close to 0
and 1, which the matching and the mixture also face on these data, and
suggests restricting the estimate to the units with overlap, after
Crump et al. (2009). The forest's estimate is requested with
`target.sample = "treated"`, the effect on the treated, the same estimand
as the other methods, and is not restricted.
tmle runs with its default SuperLearner (Polley et al., 2026) libraries,
which include a
Bayesian additive regression tree learner. This chunk is complete and runs
as shown, but it took about `r ref_minutes` minutes on
one core (R `r ref$versions[["R"]]`, Apple silicon); the results
below are read from its stored output. For a quick run, give `tmle()` the
arguments `Q.SL.library = "SL.glm", g.SL.library = "SL.glm"` on its line,
which fits the linear models alone.

```{r ref-code, eval = FALSE}
library(tmle)
lal <- MatchIt::lalonde
lal$black <- as.integer(lal$race == "black")
lal$hispan <- as.integer(lal$race == "hispan")
covs <- c("age", "educ", "black", "hispan", "married", "nodegree",
          "re74", "re75")
set.seed(20260925)
fit_tmle <- tmle(Y = lal$re78, A = lal$treat, W = lal[, covs])
att_tmle <- c(fit_tmle$estimates$ATT$psi,
              sqrt(fit_tmle$estimates$ATT$var.psi))
```

```{r lalonde-table, eval = run_treatment, echo = FALSE}
att_tbl <- data.frame(
  method = c(paste0("proxymix ", ver("proxymix"), ", one joint mixture"),
             paste0("grf ", ver("grf"), ", causal forest"),
             paste0("tmle ", desc_version("tmle", ref$versions[["tmle"]]),
                    ", default libraries"),
             paste0("MatchIt ", ver("MatchIt"), ", matching then lm()"),
             paste0("mgcv ", ver("mgcv"), " T-learner")),
  att = c(ci_lal$estimate, att_grf[["estimate"]], ref$att_tmle[1L],
          att_match[1L], att_gam[1L]),
  se = c(se_lal, att_grf[["std.err"]], ref$att_tmle[2L], att_match[2L],
         att_gam[2L]),
  stringsAsFactors = FALSE
)
ols_lal <- coef(lm(reformulate(c("treat", covs), "re78"), data = lal))[["treat"]]
att_tbl$lower <- att_tbl$att - qnorm(0.975) * att_tbl$se
att_tbl$upper <- att_tbl$att + qnorm(0.975) * att_tbl$se
att_tbl$lower[1L] <- ci_lal$conf.low
att_tbl$upper[1L] <- ci_lal$conf.high
knitr::kable(
  att_tbl, digits = 0L, align = c("l", "r", "r", "r", "r"),
  col.names = c("Method", "Effect on the treated, USD", "SE",
                "95% lower", "95% upper"),
  caption = paste0(
    "The average effect of the programme on the earnings of the 185 ",
    "treated men, with each method's own standard error: a bootstrap ",
    "standard deviation and percentile interval for proxymix, the ",
    "methods' own standard errors and normal intervals otherwise."
  )
)
```

The five point estimates run from `r round(min(att_tbl$att))` to
`r round(max(att_tbl$att))` USD, with standard errors from
`r round(min(att_tbl$se))` to `r round(max(att_tbl$se))` USD; every
interval overlaps every other, and `r sum(att_tbl$lower > 0)` of the five
`r if (sum(att_tbl$lower > 0) == 1L) "excludes" else "exclude"` zero.
`r if (gmm_n_components(model_lal@fit) == 1L) paste0("With one component selected, the proxymix read is the linear regression adjustment: the treatment coefficient of a least-squares regression of 1978 earnings on the treatment and the covariates over all ", nrow(lal), " men is ", fixed(ols_lal, 2), " USD, against ", fixed(ci_lal$estimate, 2), " for proxymix. ") else ""`The causal forest, which lets the effect vary with
every covariate, is the lowest. One dataset without a known answer cannot
say which is right, so the simulation below scores the same methods
against a known effect.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("many_methods_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
B_sim <- eval(parse(text = grep("^B <- ", sim$code, value = TRUE)))
```

The simulation repeats the treatment-effect analysis on `r sim$n_rep`
datasets of `r sim$n` units with a known heterogeneous effect. Two
covariates $x_1$ and $x_2$ are standard normal. Treatment is assigned with
probability $\mathrm{logit}^{-1}(-0.5 + 0.5 x_1)$, so about two units in five are
treated, the arms differ in $x_1$, and
an unadjusted difference of means is confounded. The outcome is
$y = x_1 + 0.5 x_2^2 + \tau(x)\, t + \varepsilon$ with standard normal
$\varepsilon$ and $\tau(x) = 1 + x_1$, so the effect rises with $x_1$, the
average effect over the population is 1, and the average effect on the
treated is above 1 because treated units have larger $x_1$.

Each method estimates the per-unit effect at every unit of its dataset and
one average effect with a 95 per cent interval. proxymix chooses its
component count by BIC and takes its interval from a
`r B_sim`-member bootstrap ensemble, against
`r length(ens_lal$members)` on the job-training data. The causal forest, tmle and the
T-learner estimate the average effect, and nearest-neighbour matching the
effect on the treated. The matched outcome regression carries
treatment-by-covariate interactions so that it also returns a per-unit
effect. Each average is scored against its value in that dataset: the mean
of the known $\tau(x)$ over the dataset's units, or over its treated units
for matching. tmle runs with a SuperLearner
library of a linear model, a lasso-penalised linear model (`SL.glmnet` at
its default `alpha = 1`, with a cross-validated penalty) and a generalised
additive model, in place of its default library, whose regression-tree
learner took about two minutes per dataset. This chunk is complete and
runs as shown, but it took about `r round(sim$elapsed_secs / 60)` minutes
on one core (R `r sim$versions[["R"]]`, Apple silicon); the
results below are read from its stored output. For a quick run, change the
line `n_rep <- 200L` to `n_rep <- 20L`.

```{r sim-code, eval = FALSE}
library(proxymix)
library(grf)
library(tmle)
library(MatchIt)
library(mgcv)

n <- 1000L     # units in each dataset
n_rep <- 200L
B <- 100L      # bootstrap members behind the proxymix interval

# y = x1 + 0.5 x2^2 + tau(x) t + e, tau(x) = 1 + x1, P(t = 1) = logit^-1(-0.5 + 0.5 x1)
make_data <- function(n) {
  x1 <- rnorm(n)
  x2 <- rnorm(n)
  t <- rbinom(n, 1L, plogis(-0.5 + 0.5 * x1))
  tau <- 1 + x1
  data.frame(y = x1 + 0.5 * x2^2 + tau * t + rnorm(n),
             t = t, x1 = x1, x2 = x2, tau = tau)
}

sl_lib <- c("SL.glm", "SL.glmnet", "SL.gam")
rmse <- function(est, truth) sqrt(mean((est - truth)^2))
row_out <- function(method, estimand, cate, est, se, lower, upper, truth) {
  data.frame(method = method, estimand = estimand, cate_rmse = cate,
             estimate = est, lower = lower, upper = upper, truth = truth)
}

one_dataset <- function(r) {
  set.seed(r)
  d <- make_data(n)
  X <- d[, c("x1", "x2")]
  is_treated <- d$t == 1L
  ate_true <- mean(d$tau)
  att_true <- mean(d$tau[is_treated])
  z <- qnorm(0.975)

  # proxymix: one joint fit, effects by conditioning, percentile bootstrap
  m <- fit_uplift(d, "y", "t", c("x1", "x2"), seed = r)
  ate_of <- function(g) {
    mg <- uplift_model(fit = g, roles = m@roles, assume = m@assume,
                       outcome_type = m@outcome_type, data = m@data,
                       n_train = m@n_train,
                       treatment_levels = m@treatment_levels)
    c(ate = mean(proxy_cate(mg, X, se = FALSE)$tau))
  }
  ens <- gmm_fit_ensemble(m@fit, B = B, seed = r)
  ci <- proxy_functional_ci(ens, ate_of, level = 0.95)
  out_pm <- row_out("proxymix", "ATE",
                    rmse(proxy_cate(m, X, se = FALSE)$tau, d$tau),
                    ci$estimate, NA, ci$conf.low, ci$conf.high, ate_true)

  # causal forest
  cf <- causal_forest(X, d$y, d$t, num.threads = 1L, seed = r)
  ate_cf <- average_treatment_effect(cf)
  out_cf <- row_out("grf", "ATE", rmse(predict(cf)$predictions, d$tau),
                    ate_cf[["estimate"]], ate_cf[["std.err"]],
                    ate_cf[["estimate"]] - z * ate_cf[["std.err"]],
                    ate_cf[["estimate"]] + z * ate_cf[["std.err"]],
                    ate_true)

  # tmle, no per-unit effect
  fit_tmle <- tmle(Y = d$y, A = d$t, W = X, Q.SL.library = sl_lib,
                   g.SL.library = sl_lib)
  est_tmle <- fit_tmle$estimates$ATE
  out_tmle <- row_out("tmle", "ATE", NA, est_tmle$psi, sqrt(est_tmle$var.psi),
                      est_tmle$CI[1L], est_tmle$CI[2L], ate_true)

  # nearest-neighbour matching, then an interacted outcome regression
  mm <- matchit(t ~ x1 + x2, data = d, method = "nearest")
  md <- match.data(mm)
  fit_lm <- lm(y ~ t * (x1 + x2), data = md, weights = weights)
  V <- sandwich::vcovCL(fit_lm, cluster = ~subclass)
  d1 <- d
  d1$t <- 1L
  d0 <- d
  d0$t <- 0L
  M1 <- model.matrix(formula(fit_lm)[-2L], d1)
  M0 <- model.matrix(formula(fit_lm)[-2L], d0)
  cate_lm <- drop((M1 - M0) %*% coef(fit_lm))
  a <- colMeans((M1 - M0)[is_treated, ])
  se_lm <- sqrt(drop(t(a) %*% V %*% a))
  att_lm <- sum(a * coef(fit_lm))
  out_lm <- row_out("MatchIt + lm", "ATT", rmse(cate_lm, d$tau), att_lm,
                    se_lm, att_lm - z * se_lm, att_lm + z * se_lm, att_true)

  # T-learner of two generalised additive models
  g1 <- gam(y ~ s(x1) + s(x2), data = d[is_treated, ])
  g0 <- gam(y ~ s(x1) + s(x2), data = d[!is_treated, ])
  L1 <- predict(g1, X, type = "lpmatrix")
  L0 <- predict(g0, X, type = "lpmatrix")
  cate_gam <- drop(L1 %*% coef(g1) - L0 %*% coef(g0))
  a1 <- colMeans(L1)
  a0 <- colMeans(L0)
  se_gam <- sqrt(drop(t(a1) %*% vcov(g1) %*% a1 + t(a0) %*% vcov(g0) %*% a0))
  ate_gam <- mean(cate_gam)
  out_gam <- row_out("mgcv T-learner", "ATE", rmse(cate_gam, d$tau), ate_gam,
                     se_gam, ate_gam - z * se_gam, ate_gam + z * se_gam,
                     ate_true)

  rbind(out_pm, out_cf, out_tmle, out_lm, out_gam)
}

res <- do.call(rbind, lapply(seq_len(n_rep), one_dataset))

sim_tab <- aggregate(
  cbind(bias = estimate - truth, rmse = (estimate - truth)^2,
        coverage = lower <= truth & truth <= upper,
        width = upper - lower) ~ method + estimand,
  data = res, FUN = mean
)
sim_tab$rmse <- sqrt(sim_tab$rmse)
cate_tab <- aggregate(cate_rmse ~ method, data = res, FUN = mean,
                      na.action = na.pass)
sim_tab$cate_rmse <- cate_tab$cate_rmse[match(sim_tab$method,
                                              cate_tab$method)]
sim_tab
```

```{=latex}
\pmNeedspace{15\baselineskip}
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order <- c("proxymix", "grf", "tmle", "MatchIt + lm",
                  "mgcv T-learner")
tab_sim <- tab_sim[match(method_order, tab_sim$method), ]
old_na <- options(knitr.kable.NA = "--")
knitr::kable(
  tab_sim[, c("method", "estimand", "cate_rmse", "bias", "rmse",
              "coverage", "width")],
  digits = 3L, row.names = FALSE,
  col.names = c("Method", "Estimand", "Per-unit RMSE", "Bias", "RMSE",
                "Coverage", "Width"),
  caption = paste0(
    "Root mean squared error of the per-unit effect over the ", sim$n,
    " units, and bias, root mean squared error, coverage of the nominal ",
    "95% interval and its average width for the average effect, over ",
    sim$n_rep, " simulated datasets, each scored against the mean of the ",
    "true effect over that dataset's units (treated units for the ATT). tmle returns no per-unit effect, so its per-unit ",
    "RMSE is empty."
  )
)
options(old_na)
sim_value <- function(method, what) {
  sim$sim_tab[[what]][sim$sim_tab$method == method]
}
mcse_coverage <- sqrt(0.95 * 0.05 / sim$n_rep)
## coverage shortfall from 0.95 in standard errors at the nominal rate
z_cov <- (0.95 - sim$sim_tab$coverage) / mcse_coverage
names(z_cov) <- sim$sim_tab$method
cov_tl <- sim_value("mgcv T-learner", "coverage")
se_tl <- sqrt(cov_tl * (1 - cov_tl) / sim$n_rep)
bias_method <- c(proxymix = "proxymix", grf = "the causal forest",
                 tmle = "tmle", "MatchIt + lm" = "the matched regression",
                 "mgcv T-learner" = "the T-learner")[[
  sim$sim_tab$method[which.max(abs(sim$sim_tab$bias))]]]
## paired comparisons with proxymix over the same datasets; each method
## has one row per dataset, in dataset order
res_by <- split(sim$res, sim$res$method)
paired <- function(method, what) {
  f <- switch(what,
              width = function(d) d$upper - d$lower,
              sq_err = function(d) (d$estimate - d$truth)^2)
  dd <- f(res_by[[method]]) - f(res_by[["proxymix"]])
  c(mean = mean(dd), se = sd(dd) / sqrt(length(dd)))
}
ate_methods <- c("grf", "tmle", "mgcv T-learner")
z_width <- vapply(c(ate_methods, "MatchIt + lm"), function(m) {
  v <- paired(m, "width")
  abs(v[["mean"]]) / v[["se"]]
}, numeric(1L))
z_sq_err <- vapply(ate_methods, function(m) {
  v <- paired(m, "sq_err")
  abs(v[["mean"]]) / v[["se"]]
}, numeric(1L))
width_order <- sim$sim_tab[order(sim$sim_tab$width), ]
width_label <- c(proxymix = "proxymix", grf = "the causal forest",
                 tmle = "tmle", "MatchIt + lm" = "the matched regression",
                 "mgcv T-learner" = "the T-learner")
width_rank <- paste0(width_label[width_order$method], " (",
                     sprintf("%.3f", width_order$width), ")")
width_phrase <- paste0(paste(head(width_rank, -1L), collapse = ", "),
                       " and ", tail(width_rank, 1L))
```

```{r sim-inspect, echo = FALSE}
## Dataset 1 of the simulation, regenerated from the stored code, to show
## which arm each of the mixture's components belongs to.
eval(parse(text = sim$code[grep("^make_data <- ", sim$code):
                             (grep("^make_data <- ", sim$code) + 7L)]))
set.seed(1L)
d_one <- make_data(sim$n)
m_one <- suppressMessages(fit_uplift(d_one, "y", "t", c("x1", "x2"),
                                     seed = 1L))
arm_one <- vapply(m_one@fit@means, function(m) m[m_one@roles$treatment],
                  numeric(1L))
rmse_one <- sqrt(mean((proxy_cate(m_one, d_one[, c("x1", "x2")],
                                  se = FALSE)$tau - d_one$tau)^2))
rmse_one_stored <- res_by[["proxymix"]]$cate_rmse[1L]
if (abs(rmse_one - rmse_one_stored) > 1e-8) {
  stop("The regenerated dataset 1 does not match the stored run.",
       call. = FALSE)
}
```

On the average effect the five methods are close. Four intervals cover
their target within `r fixed(max(abs(z_cov[names(z_cov) != "mgcv T-learner"])), 1)`
Monte Carlo standard errors of the nominal rate: proxymix in
`r fixed(sim_value("proxymix", "coverage"), 3)` of datasets, the causal
forest and tmle in `r fixed(sim_value("grf", "coverage"), 3)` and
`r fixed(sim_value("tmle", "coverage"), 3)`, and the matched regression in
`r fixed(sim_value("MatchIt + lm", "coverage"), 3)`. That standard error
is about `r fixed(mcse_coverage, 3)` at a true rate of 0.95. The
T-learner's interval covers in `r fixed(cov_tl, 3)` of datasets,
`r fixed(z_cov[["mgcv T-learner"]], 1)` of those standard errors below
0.95, or `r fixed((0.95 - cov_tl) / se_tl, 1)` of its own standard error of
`r fixed(se_tl, 3)`. The largest
absolute bias is `r fixed(max(abs(sim$sim_tab$bias)), 3)`, for
`r bias_method`, and the root mean
squared errors run from `r fixed(min(sim$sim_tab$rmse), 3)` for the
T-learner to `r fixed(max(sim$sim_tab$rmse), 3)` for the matched
regression, whose estimand is the effect on the treated. The proxymix
interval is the second narrowest, at
`r fixed(sim_value("proxymix", "width"), 3)` against
`r fixed(sim_value("mgcv T-learner", "width"), 3)` for the T-learner, whose
model-based standard error leaves its coverage the lowest of the five.

The per-unit effect separates the methods. The T-learner and the
interacted matched regression, whose effect models are both correctly
specified for $\tau(x) = 1 + x_1$, have root mean squared errors of
`r fixed(sim_value("mgcv T-learner", "cate_rmse"), 3)` and
`r fixed(sim_value("MatchIt + lm", "cate_rmse"), 3)`; the causal forest
has `r fixed(sim_value("grf", "cate_rmse"), 3)`; proxymix has
`r fixed(sim_value("proxymix", "cate_rmse"), 3)`, the largest, on an
effect whose standard deviation across units is 1. On the first dataset,
regenerated from the stored code, BIC chose
`r length(arm_one)` components: `r sum(round(arm_one) == 1)` in the treated
arm and `r sum(round(arm_one) == 0)` in the control arm. Within an arm with
several components the gate bends that arm's regression towards the
quadratic baseline in $x_2$, so the per-unit effect, a difference between
the two arms' regressions, carries curvature from one arm only; its root
mean squared error on that dataset is `r fixed(rmse_one, 3)`. Averaged over
the units, much of that error cancels, and the average-effect error of
proxymix sits among the other methods'.

In this run the average-effect intervals ranked by width as `r width_phrase`;
each paired difference in width between proxymix and another method is at
least `r sprintf("%.0f", min(z_width))` standard errors, so the ranking is
resolved, although the matched regression targets the effect on the treated.
On the squared error of the average effect, the paired differences between
proxymix and the causal forest, tmle and the T-learner are at most
`r sprintf("%.1f", max(z_sq_err))` standard errors, so this run does not
clearly separate the four. The T-learner and the matched regression came
out ahead of every other method on the per-unit effect, and proxymix had the
largest per-unit error. The comparison covers one design, one
sample size, two covariates, a linear effect and a quadratic baseline; it
does not cover effects with more structure than a few Gaussian regimes can
carry, many covariates, or the case where the Gaussian components fit the
outcome badly, and it uses a lighter learner library for tmle than its
default.

## Further reading

*Choosing between the three fitting regimes* explains what the `"moment"`
and `"sample"` arguments used throughout this page actually do, when each
applies, and the `"kld"` regime for a density that can only be evaluated.

*Compressing a kernel density estimate into a mixture* is the compression
step that turns the $O(n)$ kernel smoother above into an $O(K)$ one.

*The closed-form operator calculus on a mixture* covers the pushforward and
Kalman-update operators that the one-line extras above only name.

*Imputing missing data with a mixture* takes the conditioning operator used
throughout this page and applies it to holes in the data.

## References

* Athey, S. and Imbens, G. (2016). *Recursive partitioning for
  heterogeneous causal effects.* Proceedings of the National Academy of
  Sciences 113(27), 7353--7360. <https://doi.org/10.1073/pnas.1510489113>.
* Athey, S., Tibshirani, J. and Wager, S. (2019). *Generalized random
  forests.* The Annals of Statistics 47(2), 1148--1178.
  <https://doi.org/10.1214/18-AOS1709>.
* Bach, P., Kurz, M. S., Chernozhukov, V., Spindler, M. and Klaassen, S.
  (2024). *DoubleML: An object-oriented implementation of double machine
  learning in R.* Journal of Statistical Software 108(3), 1--56.
  <https://doi.org/10.18637/jss.v108.i03>.
* Crump, R. K., Hotz, V. J., Imbens, G. W. and Mitnik, O. A. (2009).
  *Dealing with limited overlap in estimation of average treatment
  effects.* Biometrika 96(1), 187--199. <https://doi.org/10.1093/biomet/asn055>.
* Dehejia, R. H. and Wahba, S. (1999). *Causal effects in nonexperimental
  studies: reevaluating the evaluation of training programs.* Journal of
  the American Statistical Association 94(448), 1053--1062.
  <https://doi.org/10.1080/01621459.1999.10473858>.
* de Veaux, R. D. (1989). *Mixtures of linear regressions.* Computational
  Statistics & Data Analysis 8(3), 227--245.
  <https://doi.org/10.1016/0167-9473(89)90043-1>.
* Fraley, C. and Raftery, A. E. (2002). *Model-based clustering,
  discriminant analysis, and density estimation.* Journal of the American
  Statistical Association 97(458), 611--631.
  <https://doi.org/10.1198/016214502760047131>.
* Friedman, J., Hastie, T. and Tibshirani, R. (2010). *Regularization
  paths for generalized linear models via coordinate descent.* Journal of
  Statistical Software 33(1), 1--22. <https://doi.org/10.18637/jss.v033.i01>.
* Gorman, K. B., Williams, T. D. and Fraser, W. R. (2014). *Ecological
  sexual dimorphism and environmental variability within a community of
  Antarctic penguins (genus Pygoscelis).* PLoS ONE 9(3), e90081.
  <https://doi.org/10.1371/journal.pone.0090081>.
* Gruber, S. and van der Laan, M. J. (2012). *tmle: An R package for
  targeted maximum likelihood estimation.* Journal of Statistical Software
  51(13), 1--35. <https://doi.org/10.18637/jss.v051.i13>.
* Hayfield, T. and Racine, J. S. (2008). *Nonparametric econometrics: The
  np package.* Journal of Statistical Software 27(5), 1--32.
  <https://doi.org/10.18637/jss.v027.i05>.
* Ho, D. E., Imai, K., King, G. and Stuart, E. A. (2011). *MatchIt:
  Nonparametric preprocessing for parametric causal inference.* Journal of
  Statistical Software 42(8), 1--28. <https://doi.org/10.18637/jss.v042.i08>.
* Hoerl, A. E. and Kennard, R. W. (1970). *Ridge regression: biased
  estimation for nonorthogonal problems.* Technometrics 12(1), 55--67.
  <https://doi.org/10.1080/00401706.1970.10488634>.
* Horst, A. M., Presmanes Hill, A. and Gorman, K. B. (2022). *Palmer
  Archipelago penguins data in the palmerpenguins R package -- an
  alternative to Anderson's irises.* The R Journal 14(1), 244--254.
  <https://doi.org/10.32614/RJ-2022-020>.
* Jolliffe, I. T. (2002). *Principal Component Analysis*, 2nd ed. Springer.
  <https://doi.org/10.1007/b98835>.
* Künzel, S. R., Sekhon, J. S., Bickel, P. J. and Yu, B. (2019).
  *Metalearners for estimating heterogeneous treatment effects using
  machine learning.* Proceedings of the National Academy of Sciences
  116(10), 4156--4165. <https://doi.org/10.1073/pnas.1804597116>.
* Nadaraya, E. A. (1964). *On estimating regression.* Theory of Probability
  and Its Applications 9(1), 141--142. <https://doi.org/10.1137/1109020>.
* Polley, E., LeDell, E., Kennedy, C. and van der Laan, M. (2026).
  *SuperLearner: Super Learner Prediction.* R package version
  `r ver("SuperLearner")`. <https://doi.org/10.32614/CRAN.package.SuperLearner>.
* Tipping, M. E. and Bishop, C. M. (1999). *Probabilistic principal
  component analysis.* Journal of the Royal Statistical Society B 61(3),
  611--622. <https://doi.org/10.1111/1467-9868.00196>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <https://doi.org/10.1080/07362994.2024.2372605>.
* Venables, W. N. and Ripley, B. D. (2002). *Modern Applied Statistics
  with S*, 4th ed. Springer.
* Watson, G. S. (1964). *Smooth regression analysis.* Sankhyā A 26(4),
  359--372.
* Wood, S. N. (2011). *Fast stable restricted maximum likelihood and
  marginal likelihood estimation of semiparametric generalized linear
  models.* Journal of the Royal Statistical Society B 73(1), 3--36.
  <https://doi.org/10.1111/j.1467-9868.2010.00749.x>.
* Zeileis, A., Köll, S. and Graham, N. (2020). *Various versatile
  variances: An object-oriented implementation of clustered covariances in
  R.* Journal of Statistical Software 95(1), 1--36.
  <https://doi.org/10.18637/jss.v095.i01>.

## Reproduce

Two seeds are used in the first part: `20260617` for the regression,
clustering and principal-components data, set in the chunk that builds
each, and `20260902` for the treatment-effect design. `fit_uplift()` is
additionally given `seed = 1L` so its expectation-maximisation start is
reproducible without disturbing the ambient random-number state.

In the numerical illustration, `kmeans()` runs after `set.seed(20260925)`,
the three-component penguin fit and the job-training fit are given
`seed = 1L`, the bootstrap ensemble `seed = 2L`, and the causal forest
`seed = 1L`. The stored tmle estimate comes from the code shown, run after
`set.seed(20260925)` on `r ref$run_date` with R `r ref$versions[["R"]]`,
tmle `r desc_version("tmle", ref$versions[["tmle"]])` and SuperLearner
`r desc_version("SuperLearner", ref$versions[["SuperLearner"]])`, in about `r ref_minutes` minutes,
and the run raised `r length(ref$warnings)` warnings. In the simulation
each dataset is generated after `set.seed()` with its own index, which also
seeds the mixture fit, its ensemble and the forest. The stored simulation
results come from the code shown, run on `r sim$run_date` with R
`r sim$versions[["R"]]`, proxymix
`r desc_version("proxymix", sim$versions[["proxymix"]])`, grf
`r desc_version("grf", sim$versions[["grf"]])`, tmle
`r desc_version("tmle", sim$versions[["tmle"]])`, SuperLearner
`r desc_version("SuperLearner", sim$versions[["SuperLearner"]])`, MatchIt
`r desc_version("MatchIt", sim$versions[["MatchIt"]])`, mgcv
`r desc_version("mgcv", sim$versions[["mgcv"]])` and sandwich
`r desc_version("sandwich", sim$versions[["sandwich"]])`,
and the run raised `r length(sim$warnings)` warnings.

```{r session-info}
old_opt <- options(width = 72L)
si_vec <- trimws(capture.output(sessionInfo()), "right")
options(old_opt)
# library paths are single long words, so a long line breaks after a slash
long_vec <- nchar(si_vec) > 72L
si_vec[long_vec] <- gsub("(.{1,68}/)(?=.{8,})", "\\1\n        ",
                         si_vec[long_vec], perl = TRUE)
writeLines(si_vec)
```
