---
title: "Imputing missing data with a mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputing missing data with a mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~9 s under rmarkdown::render() with ggplot2, mice and
     Amelia installed, the simulation read from its stored results;
     macOS arm64 (Apple silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_mice <- requireNamespace("mice", quietly = TRUE)
```

## The problem

An analyst has two measurements per unit. The scatter falls into two clear
clusters with almost nothing between them, and a good half of the second
measurement never came back from the field. The question is not whether an
imputer will run. It is: *my data has two clusters and an empty gap between
them, so if I let a standard imputer fill the holes, will it put values in
the gap where I know there is no mass, and will that move the mean I am
going to report?*

This vignette answers both halves on a synthetic two-cluster toy where the
values that went missing are retained, so every imputation can be scored
against the truth it was hiding.

## Package capabilities

Imputation here is conditioning. If the joint distribution of the data is a
Gaussian mixture, the conditional distribution of a missing entry given the
observed entries of its own row is itself a Gaussian mixture, available in
closed form through the Schur-complement algebra of
`gmm_conditionalise()`. Nothing is simulated to get it.

`gmm_impute()` fits that mixture to a matrix containing holes and draws `m`
completed datasets from the per-row conditionals. Its `N` argument is the
number of mixture components, chosen by the Bayesian information criterion
when left at its default. A mixture can be multimodal and can give each
component its own spread, which neither a single-Gaussian model nor a
linear-Gaussian conditional such as normal regression imputation can.

`gmm_complete()` extracts one completed dataset. `proxy_pool()` combines
an estimand across the completions, and `proxy_fmi()` reports the fraction
of missing information. For a column mean the pooling is closed form -- the
limit of the between-imputation variance as the number of completions `m`
grows without bound -- so no Monte Carlo noise enters at the pooling step. For any other estimand,
`as_mids()` packages the completions as a `mice` object so the joint
mixture imputations flow into `mice::pool()` unchanged: proxymix supplies
the imputation model, `mice` the pooling.

## Addressing the problem

### A two-cluster toy with the truth retained

Two groups, each a tight cloud, with a positive within-group association
between `x1` and `x2`. This is a synthetic toy: the complete data is
generated first and kept, then part of `x2` is deleted, so the held-out
values are available to score against. The chance of an entry going missing
depends on the observed `x1` and not on the hidden `x2` itself, which is the
missing-at-random setting.

```{r data}
set.seed(20260620)
n <- 600L
lab <- sample(c(-1, 1), n, replace = TRUE)
x1 <- 2 * lab + rnorm(n, 0, 0.6)
x2 <- 2 * lab + 0.5 * (x1 - 2 * lab) + rnorm(n, 0, 0.6)
truth <- cbind(x1 = x1, x2 = x2)

x_holes <- truth
missing <- runif(n) < plogis(0.6 * x1)
x_holes[missing, "x2"] <- NA
frac_missing <- mean(missing)
```

The marginal distribution of `x2` is bimodal, one cluster near $-2$ and one
near $+2$, with little mass between them. Writing the empty middle as the
interval $|x_2| < 1$, the values that were deleted put only a small share of
their mass there.

```{r gap-truth}
in_gap <- function(v) mean(abs(v) < 1)
gap_truth <- in_gap(truth[missing, "x2"])
```

### Impute with a mixture, and with a single Gaussian

```{r impute}
imp <- gmm_impute(x_holes, N = 2L, m = 20L, seed = 1L)
imp
```

A single completed dataset has no holes:

```{r complete}
done <- gmm_complete(imp, 1L)
anyNA(done)
```

Forcing one component reproduces the multivariate-normal assumption.

```{r single}
imp1 <- gmm_impute(x_holes, N = 1L, m = 20L, seed = 1L)
done1 <- gmm_complete(imp1, 1L)
```

```{r gap-table, echo = FALSE}
gap_tbl <- data.frame(
  source = c("held-out truth", "mixture imputation (N = 2)",
             "single-Gaussian imputation (N = 1)"),
  gap_share = c(gap_truth, in_gap(done[missing, "x2"]),
                in_gap(done1[missing, "x2"])),
  stringsAsFactors = FALSE
)
knitr::kable(
  gap_tbl, digits = 3L,
  col.names = c("Source", "Share of values with $\\lvert x_2 \\rvert < 1$"),
  caption = paste0(
    "How much mass each imputer places in the empty middle between the ",
    "two clusters, over the ", sum(missing), " deleted entries. The two ",
    "imputation rows are read from the first of the ", imp@m,
    " completions of each imputer."
  )
)
```

### The shape of the imputations

```{r fig-modes-facts, include = FALSE}
## peaks and the trough between them, read from the same density()
## estimates the figure draws
dens_shape <- function(v) {
  d <- density(v)
  turn <- diff(sign(diff(d$y)))
  list(peak_x = d$x[which(turn == -2L) + 1L],
       trough_y = d$y[which(turn == 2L) + 1L])
}
shape_truth <- dens_shape(truth[missing, "x2"])
shape_one <- dens_shape(done1[missing, "x2"])
shape_mix <- dens_shape(done[missing, "x2"])
num2 <- function(v) paste0("$", format(round(v, 2L), nsmall = 2L), "$")
num3 <- function(v) format(round(v, 3L), nsmall = 3L)
modes_cap <- paste0(
  "Density of the deleted $x_2$ values against the values each imputer ",
  "supplied in their place. The truth peaks at $x_2$ = ",
  num2(shape_truth$peak_x[1L]), " and ", num2(shape_truth$peak_x[2L]),
  ", the mixture imputation at ", num2(shape_mix$peak_x[1L]), " and ",
  num2(shape_mix$peak_x[2L]), ", and the single-Gaussian imputation at ",
  num2(shape_one$peak_x[1L]), " and ", num2(shape_one$peak_x[2L]),
  ". Between its peaks the single-Gaussian density falls only to ",
  num3(shape_one$trough_y[1L]), " against the truth's ",
  num3(shape_truth$trough_y[1L]), ". It puts ",
  num3(in_gap(done1[missing, "x2"])),
  " of the imputed values in the shaded band, where the truth has ",
  num3(gap_truth), ". Each imputation curve is drawn from the first of the ",
  imp@m, " completions of its imputer."
)
```

```{r fig-modes, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.8, fig.cap = modes_cap, fig.alt = "Three density curves over x2: the held-out truth, the mixture imputation and the single-Gaussian imputation are all bimodal; the single-Gaussian curve has lower peaks set closer to the middle and more density in the shaded gap between the two clusters."}
dens_df <- function(v, label) {
  d <- density(v)
  data.frame(x2 = d$x, density = d$y, source = label,
             stringsAsFactors = FALSE)
}
plot_df <- rbind(
  dens_df(truth[missing, "x2"], "held-out truth"),
  dens_df(done[missing, "x2"], "mixture (N = 2)"),
  dens_df(done1[missing, "x2"], "single Gaussian (N = 1)")
)
plot_df$source <- factor(
  plot_df$source,
  levels = c("held-out truth", "mixture (N = 2)",
             "single Gaussian (N = 1)")
)
ggplot2::ggplot(plot_df, ggplot2::aes(x2, density, colour = source)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::annotate("rect", xmin = -1, xmax = 1, ymin = -Inf, ymax = Inf,
                    fill = "grey60", alpha = 0.18) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("held-out truth" = "#000000",
               "mixture (N = 2)" = "#009E73",
               "single Gaussian (N = 1)" = "#D55E00")
  ) +
  ggplot2::labs(
    x = expression(x[2]), y = "density",
    title = "What each imputer put where the data was deleted",
    subtitle = expression("Shaded band marks the empty middle, " *
                            group("|", x[2], "|") < 1)
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-modes-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the imputation-density",
    "figure is skipped.\n")
```

### Pooling a column mean

```{r pool-mean}
pooled <- proxy_pool(imp, "x2")
pooled1 <- proxy_pool(imp1, "x2")
fmi_mix <- proxy_fmi(imp, "x2")
mean_truth <- mean(truth[, "x2"])
err_mix <- abs(pooled$estimate - mean_truth)
err_one <- abs(pooled1$estimate - mean_truth)
```

```{r fixed-notation, include = FALSE}
## Small errors are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
```

```{r pool-table, echo = FALSE}
pool_tbl <- data.frame(
  model = c("complete-data truth", "available cases only",
            "mixture (N = 2)",
            "single Gaussian (N = 1)"),
  estimate = c(mean(truth[, "x2"]), mean(x_holes[!missing, "x2"]),
               pooled$estimate, pooled1$estimate),
  std_error = c(NA_real_, NA_real_, pooled$std.error, pooled1$std.error),
  conf_low = c(NA_real_, NA_real_, pooled$conf.low, pooled1$conf.low),
  conf_high = c(NA_real_, NA_real_, pooled$conf.high, pooled1$conf.high),
  stringsAsFactors = FALSE
)
pool_tbl$abs_error <- abs(pool_tbl$estimate - mean(truth[, "x2"]))
old_opt <- options(knitr.kable.NA = "--")
pool_out <- knitr::kable(
  pool_tbl, digits = 4L,
  col.names = c("Estimate of the mean of $x_2$", "Estimate", "SE",
                "CI lower", "CI upper", "Absolute error"),
  caption = paste0(
    "The mean of $x_2$ in the complete data and recovered three ways, each ",
    "scored against that complete-data value. Standard errors and ",
    "intervals are shown for the two pooled imputations only. The ",
    "complete-data row is the reference, and the available-cases row is ",
    "the plain mean of the rows that came back."
  )
)
options(old_opt)
pool_out
```

### Any other estimand, through `mice`

```{r pool-mice, eval = has_mice, echo = has_mice}
mice_fit <- mice::pool(with(as_mids(imp), lm(x2 ~ x1)))
```

```{r pool-mice-table, eval = has_mice, echo = FALSE}
mice_tbl <- summary(mice_fit)
mice_tbl$p.value <- ifelse(mice_tbl$p.value < 0.001, "< 0.001",
                           format(round(mice_tbl$p.value, 3L), nsmall = 3L))
knitr::kable(
  mice_tbl, digits = 3L, align = c("l", "r", "r", "r", "r", "r"),
  caption = paste0(
    "The regression of $x_2$ on $x_1$, fitted in each of the ", imp@m,
    " completions and pooled by `mice::pool()`."
  )
)
mice_slope <- mice_tbl[mice_tbl$term == "x1", ]
```

```{r pool-mice-text, eval = has_mice, echo = FALSE, results = "asis"}
cat("Pooled over the mixture completions, the slope of $x_2$ on $x_1$ is ",
    fixed(mice_slope$estimate, 3), " with a standard error of ",
    fixed(mice_slope$std.error, 3), ".\n", sep = "")
```

```{r pool-mice-skip, eval = !has_mice, echo = FALSE, results = "asis"}
cat("mice is not installed on this build, so the pooled regression is",
    "skipped; the closed-form column-mean pooling above does not need it.\n")
```

## Interpretation

```{r gap-all, include = FALSE}
## share in the gap, averaged over every completion of each imputer
gap_over <- function(im) {
  mean(vapply(seq_len(im@m), function(i1) {
    in_gap(gmm_complete(im, i1)[missing, "x2"])
  }, numeric(1L)))
}
gap_all <- c(mixture = gap_over(imp), single = gap_over(imp1))
```

Of the `r n` rows, `r sum(missing)` lost their `x2`, or
`r round(100 * frac_missing)` per cent. Over those deleted entries the truth
placed `r round(100 * gap_truth, 1)` per cent of its mass in the empty
middle $|x_2| < 1$. In the first completion of each imputer, the mixture put
`r round(100 * in_gap(done[missing, "x2"]), 1)` per cent there and the single
Gaussian `r round(100 * in_gap(done1[missing, "x2"]), 1)` per cent. Averaged
over all `r imp@m` completions, the shares are
`r fixed(100 * gap_all[["mixture"]], 2)` per cent for the mixture and
`r fixed(100 * gap_all[["single"]], 2)` per cent for the single Gaussian,
about `r round(gap_all[["single"]] / gap_truth, 1)` times what the data had.
The answer to the opening question is that a single-Gaussian imputer puts
more mass in the gap than the data had. Averaged over its `r imp@m`
completions, the mixture puts
`r if (gap_all[["mixture"]] <= gap_truth) "no more than the data had" else paste0("about as much as the data had, ", round(100 * (gap_all[["mixture"]] - gap_truth), 1), " percentage points more")`,
`r if (in_gap(done[missing, "x2"]) > gap_truth) "although its first completion alone puts more" else "and so does its first completion"`.

```{r gap-why, include = FALSE}
slope_pooled <- coef(lm(x2 ~ x1, data = as.data.frame(x_holes)))[["x1"]]
slope_within <- coef(lm(x2 ~ x1 + factor(lab),
                        data = as.data.frame(truth)))[["x1"]]
```

Each imputed value is drawn given the observed `x1` of its row, and `x1`
is itself bimodal, so the single-Gaussian imputations keep two clusters as
well. The extra mass in the gap comes from the conditional law. A single
Gaussian gives `x2` one straight line in `x1` across both clusters, with a
slope of `r fixed(slope_pooled, 2)` where the slope within each cluster is
`r fixed(slope_within, 2)`, and one residual spread. Its imputed clusters
are therefore wider, and their inner tails reach into the gap. The mixture
has a line and a spread for each cluster.

That shape error reaches the number an analyst would report. The
complete-data mean of `x2` is `r fixed(mean_truth, 3)`. Using only the rows
that came back gives `r fixed(mean(x_holes[!missing, "x2"]), 3)`, which is
out by `r fixed(abs(mean(x_holes[!missing, "x2"]) - mean_truth), 3)`,
because missingness here is tied to `x1` and `x1` is tied to `x2`. The
pooled means of the mixture and the single-Gaussian imputations are out by
`r fixed(err_mix, 4)` and `r fixed(err_one, 4)`, which is
`r fixed(err_mix / pooled$std.error, 2)` and
`r fixed(err_one / pooled1$std.error, 2)` of their standard errors of
`r fixed(pooled$std.error, 4)` and `r fixed(pooled1$std.error, 4)`. On this
dataset both recover the mean to well within their own uncertainty, and the
two errors cannot be ranked. The damage the single Gaussian does is to the
shape, which the figure makes visible and the mean does not.

The fraction of missing information is
`r round(unname(fmi_mix), 3)`: although
`r round(100 * frac_missing)` per cent of the column is absent, the
observed `x1` carries enough about `x2` that only about
`r round(100 * unname(fmi_mix))` per cent of the information about the mean
is lost.

## Limitations

Everything above assumes numeric data that is missing at random: the
probability that an entry is missing may depend on the observed entries but
not on the missing value itself. That assumption is what makes the conditional of the
observed the right law to impute from, and it is not testable from the data
in hand. Value-dependent missingness and censoring at a known bound break
it. The `mechanism` argument of `gmm_impute()` handles censoring, and
handles value-dependent missingness under an assumed selection slope, as the
companion vignette on that case shows.

The comparison in the toy is against a single Gaussian, which is the
multivariate-normal imputation model. It is not a comparison against a
well-specified alternative such as predictive mean matching or a
classification-and-regression-tree imputer, either of which can also
represent two clusters, so nothing here establishes that a mixture beats
every competitor -- only that it beats the Gaussian assumption on data the
Gaussian assumption mis-specifies. A second limit is scale: one column of
one bivariate toy at `n` = `r n`, with the component count supplied rather
than selected, so the result speaks to the shape of the imputation law and
not to behaviour in wide data or with many incomplete columns at once. The
numerical illustration that follows adds predictive mean matching and
bootstrapped multivariate-normal imputation, through mice and Amelia, on
real data and in simulation.

The estimand shown in closed form is a column mean. Anything else is pooled
by `mice`, whose rules are the large-sample rules of Rubin (1987) and
inherit their assumptions. What would change the answer most is the number of components:
at `N` = 1 the imputer is the multivariate-normal model,
and a component count chosen well above the number of genuine clusters
would fit noise into the conditional law.

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_amelia <- requireNamespace("Amelia", quietly = TRUE)
has_penguins <- exists("penguins", envir = asNamespace("datasets"))
run_illustration <- has_mice && has_amelia && has_penguins
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("mice, Amelia or the penguins data (R 4.5.0 or later) is not available",
    "on this build, so the real-data illustration is skipped.\n")
```

### Imputing the Palmer penguins data with three packages

We apply proxymix, mice (van Buuren and Groothuis-Oudshoorn, 2011) and
Amelia (Honaker et al., 2011) to the Palmer penguins data
(Gorman et al., 2014), shipped with R as `penguins` since
version 4.5.0 (Horst et al., 2022). The mice package fills each
incomplete column in turn by predictive mean matching, Amelia draws from a
multivariate-normal model fitted by expectation-maximisation to bootstrap
resamples, and proxymix draws from the fitted mixture.

The two competitors install from CRAN once and load as usual:

```{r install-competitors, eval = FALSE}
install.packages(c("mice", "Amelia"))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(mice)
library(Amelia)
```

The four body measurements of the birds measured in full are kept, and part
of body mass is deleted, with the chance of deletion rising with flipper
length. The complete data stay aside for scoring.

```{r penguins-data, eval = run_illustration}
cols <- c("bill_len", "bill_dep", "flipper_len", "body_mass")
peng <- na.omit(penguins[, cols])

set.seed(20260913)
gone <- runif(nrow(peng)) < plogis((peng$flipper_len - 205) / 10)
peng_holes <- peng
peng_holes$body_mass[gone] <- NA
```

```{r penguins-facts, eval = run_illustration, include = FALSE}
species <- penguins$species[complete.cases(penguins[, cols])]
slope_colony <- coef(lm(body_mass ~ bill_dep, data = peng))[[2L]]
slope_within <- vapply(split(peng, species), function(d) {
  coef(lm(body_mass ~ bill_dep, data = d))[[2L]]
}, numeric(1L))
gone_gentoo <- mean(gone[species == "Gentoo"])
```

That leaves `r nrow(peng)` birds with bill length, bill depth and flipper
length in millimetres and body mass in grams. Deletion removes
`r round(100 * mean(gone))` per cent of the masses, and
`r round(100 * gone_gentoo)` per cent among Gentoo penguins, the heaviest
species, so the values are missing at random given an observed column. The
data carry structure that none of the imputers is told about. The three
species form separate clusters, and across the colony body mass falls by
`r round(abs(slope_colony))` g per millimetre of bill depth, while within
each species it rises by `r round(min(slope_within))` to
`r round(max(slope_within))` g per millimetre.

Each package imputes the same data frame with its default settings and 20
completions:

```{r penguins-impute, eval = run_illustration}
imp_proxymix <- gmm_impute(peng_holes, m = 20L, seed = 1L)
imp_mice <- mice(peng_holes, m = 20L, seed = 1L, printFlag = FALSE)
set.seed(1L)
imp_amelia <- amelia(peng_holes, m = 20L, p2s = 0L)
```

The same analysis then runs on every completed dataset, and one pooling
rule, Rubin's rules as implemented in `mice::pool()`, combines the results
for all three, so the differences come from the imputations alone:

```{r penguins-pool, eval = run_illustration}
completed <- list(
  proxymix = complete(as_mids(imp_proxymix), "all"),
  mice = complete(imp_mice, "all"),
  Amelia = imp_amelia$imputations
)

pool_three <- function(f) {
  t(vapply(completed, function(sets) {
    fits <- lapply(sets, function(d) lm(f, data = d))
    s <- summary(pool(fits), conf.int = TRUE)
    unlist(s[nrow(s), c("estimate", "conf.low", "conf.high")])
  }, numeric(3L)))
}

mean_mass <- pool_three(body_mass ~ 1)
slope_mass <- pool_three(body_mass ~ bill_dep)
```

```{r penguins-table, eval = run_illustration, echo = FALSE}
row_ci <- function(fit) {
  k <- length(coef(fit))
  c(coef(fit)[[k]], confint(fit)[k, ])
}
mean_all <- rbind(row_ci(lm(body_mass ~ 1, data = peng)),
                  row_ci(lm(body_mass ~ 1, data = peng_holes)),
                  mean_mass)
slope_all <- rbind(row_ci(lm(body_mass ~ bill_dep, data = peng)),
                   row_ci(lm(body_mass ~ bill_dep, data = peng_holes)),
                   slope_mass)
with_ci <- function(x, digits) {
  num <- function(v) format(round(v, digits), nsmall = digits)
  paste0(num(x[, 1L]), " (", num(x[, 2L]), " to ", num(x[, 3L]), ")")
}
peng_tbl <- data.frame(
  data_used = c("complete data, before deletion", "available cases only",
                "proxymix", "mice", "Amelia"),
  mean_mass = with_ci(mean_all, 0L),
  slope_mass = with_ci(slope_all, 1L),
  stringsAsFactors = FALSE
)
knitr::kable(
  peng_tbl, align = c("l", "r", "r"),
  col.names = c("Data used", "Mean body mass, g",
                "Slope on bill depth, g per mm"),
  caption = paste0(
    "Mean body mass and its slope on bill depth, each with a 95% interval, ",
    "from the complete data, from the birds whose mass was kept, and from ",
    "each package's 20 completions pooled by Rubin's rules."
  )
)
err_mean <- abs(mean_mass[, "estimate"] - mean_all[1L, 1L])
```

Given only the four measurements, proxymix selected
`r imp_proxymix@n_components` mixture components, the number of species.
Keeping only the birds whose mass survived pulls the mean body mass down from
`r round(mean_all[1L, 1L])` g to `r round(mean_all[2L, 1L])` g, because the
heaviest species lost most of its values. Pooled over the completions,
proxymix, mice and Amelia return means `r round(err_mean[["proxymix"]])`,
`r round(err_mean[["mice"]])` and `r round(err_mean[["Amelia"]])` g from the
complete-data value, and all three intervals contain it. The slope on bill
depth separates them more. The complete data give
`r round(slope_all[1L, 1L], 1)` g per millimetre and the available cases
`r round(slope_all[2L, 1L], 1)`, while proxymix, mice and Amelia give
`r round(slope_mass["proxymix", "estimate"], 1)`,
`r round(slope_mass["mice", "estimate"], 1)` and
`r round(slope_mass["Amelia", "estimate"], 1)`. Amelia moves furthest
towards zero, consistent with a single multivariate-normal model, which gives
bill depth one conditional coefficient across the three species. This is one
deletion of one dataset, which illustrates the imputers but cannot rank them,
so the simulation below repeats the exercise many times.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("missing_data_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
x1_density_b <- function(x) 0.5 * dnorm(x, -1.5) + 0.5 * dnorm(x, 1.5)
share_a <- integrate(function(x) plogis(0.4 + 0.6 * x) * dnorm(x),
                     -Inf, Inf)$value
share_b <- integrate(function(x) plogis(0.4 + 0.6 * x) * x1_density_b(x),
                     -Inf, Inf)$value
```

A single deletion is one draw. The simulation repeats the analysis on
`r sim$n_rep` datasets of `r sim$n` rows in each of two designs whose
population values are known. In design A, $x_1$ is standard normal and
$x_2 = 1.5 x_1 + \varepsilon$ with standard normal $\varepsilon$, one
Gaussian cloud for which a linear-Gaussian imputation model is correctly
specified. In design B, two
equally likely groups centred at $(-1.5, -1.5)$ and $(1.5, 2)$, each with
unit variances, have within-group correlations of 0.6 and -0.3. The mean of
$x_2$ given $x_1$ then bends between the groups, and a single linear-Gaussian
imputation model is mis-specified. In both designs $x_2$ is deleted with
probability $\mathrm{logit}^{-1}(0.4 + 0.6 x_1)$, at random given $x_1$,
which removes `r round(100 * share_a)` and `r round(100 * share_b)` per cent
of $x_2$ on average.

Each imputer makes `r sim$m` completions, and proxymix chooses its number of
components by BIC in every dataset. The mean of $x_2$ and the slope of $x_2$
on $x_1$ are pooled by Rubin's rules as before and scored against their
population values, `r sim$population["A", "mean"]` and
`r sim$population["A", "slope"]` in design A, and
`r sim$population["B", "mean"]` and
`r round(sim$population["B", "slope"], 3)` in design B. The analysis of the
complete data, before deletion, is the reference. This chunk is complete and
runs as shown, but it took about `r round(sim$elapsed_secs / 60)` minutes on
one core (R `r sim$versions[["R"]]`, Apple silicon); the results below are
read from its stored output.

```{r sim-code, eval = FALSE}
library(proxymix)
library(mice)
library(Amelia)

n <- 500L      # rows in each dataset
m <- 20L       # completions per imputer
n_rep <- 500L  # datasets per design

designs <- list(
  A = function(n) {
    x1 <- rnorm(n)
    data.frame(x1 = x1, x2 = 1.5 * x1 + rnorm(n))
  },
  B = function(n) {
    grp <- runif(n) < 0.5
    rho <- ifelse(grp, -0.3, 0.6)
    z1 <- rnorm(n)
    data.frame(
      x1 = ifelse(grp, 1.5, -1.5) + z1,
      x2 = ifelse(grp, 2, -1.5) + rho * z1 + sqrt(1 - rho^2) * rnorm(n)
    )
  }
)

# population mean of x2 and slope of x2 on x1
population <- rbind(
  A = c(mean = 0, slope = 1.5),
  B = c(mean = 0.25, slope = 2.775 / 3.25)
)

# estimate and 95% interval for the last coefficient: the usual interval
# for one dataset, Rubin's rules over several completions
ci_last <- function(fits) {
  if (length(fits) == 1L) {
    k <- length(coef(fits[[1L]]))
    return(c(coef(fits[[1L]])[[k]], confint(fits[[1L]])[k, ]))
  }
  s <- summary(pool(fits), conf.int = TRUE)
  unlist(s[nrow(s), c("estimate", "conf.low", "conf.high")])
}

one_dataset <- function(design, r) {
  set.seed(r)
  full <- designs[[design]](n)
  obs <- full
  obs$x2[runif(n) < plogis(0.4 + 0.6 * full$x1)] <- NA
  sets <- list(
    "complete data" = list(full),
    proxymix = complete(as_mids(gmm_impute(obs, m = m, seed = r)), "all"),
    mice = complete(mice(obs, m = m, seed = r, printFlag = FALSE), "all"),
    Amelia = amelia(obs, m = m, p2s = 0L)$imputations
  )
  out <- lapply(names(sets), function(method) {
    ci <- rbind(
      ci_last(lapply(sets[[method]], function(d) lm(x2 ~ 1, data = d))),
      ci_last(lapply(sets[[method]], function(d) lm(x2 ~ x1, data = d)))
    )
    data.frame(design = design, method = method,
               estimand = c("mean", "slope"),
               estimate = ci[, 1L], lower = ci[, 2L], upper = ci[, 3L])
  })
  do.call(rbind, out)
}

res <- do.call(rbind, lapply(c("A", "B"), function(design) {
  do.call(rbind, lapply(seq_len(n_rep), function(r) one_dataset(design, r)))
}))
res$truth <- population[cbind(res$design, res$estimand)]

sim_tab <- aggregate(
  cbind(bias = estimate - truth, rmse = (estimate - truth)^2,
        coverage = lower <= truth & truth <= upper,
        width = upper - lower) ~ design + estimand + method,
  data = res, FUN = mean
)
sim_tab$rmse <- sqrt(sim_tab$rmse)
sim_tab
```

```{=latex}
\par\ifdim\dimexpr\pagegoal-\pagetotal\relax<0.45\textheight\clearpage\fi
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order <- c("complete data", "proxymix", "mice", "Amelia")
tab_sim <- tab_sim[order(tab_sim$design, tab_sim$estimand,
                         match(tab_sim$method, method_order)), ]
tab_sim$estimand <- ifelse(tab_sim$estimand == "mean", "mean of $x_2$",
                           "slope of $x_2$ on $x_1$")
knitr::kable(
  tab_sim[, c("design", "estimand", "method", "bias", "rmse", "coverage",
              "width")],
  digits = c(0L, 0L, 0L, 4L, 3L, 3L, 3L), row.names = FALSE,
  col.names = c("Design", "Estimand", "Data used", "Bias", "RMSE",
                "Coverage", "Width"),
  caption = paste0(
    "Bias and root mean squared error of the point estimate, coverage of ",
    "the nominal 95% interval and its average width, over ", sim$n_rep,
    " simulated datasets per design, each scored against the population ",
    "value. Design A is one Gaussian cloud, design B two groups."
  )
)
sim_value <- function(design, estimand, method, what) {
  s1 <- sim$sim_tab$design == design & sim$sim_tab$estimand == estimand &
    sim$sim_tab$method == method
  sim$sim_tab[[what]][s1]
}
mcse_coverage <- sqrt(0.95 * 0.05 / sim$n_rep)
gap_se <- function(design, method) {
  (0.95 - sim_value(design, "slope", method, "coverage")) / mcse_coverage
}
own_se <- function(design, method) {
  p <- sim_value(design, "slope", method, "coverage")
  sqrt(p * (1 - p) / sim$n_rep)
}
```

In design A, the one Gaussian cloud, proxymix and Amelia are
indistinguishable. Their intervals cover the population slope in
`r fixed(sim_value("A", "slope", "proxymix", "coverage"), 3)` and
`r fixed(sim_value("A", "slope", "Amelia", "coverage"), 3)` of datasets,
against `r fixed(sim_value("A", "slope", "complete data", "coverage"), 3)`
for the complete data, with biases of
`r fixed(sim_value("A", "slope", "proxymix", "bias"), 4)` and
`r fixed(sim_value("A", "slope", "Amelia", "bias"), 4)` and the same width.
The mice intervals cover the slope in only
`r fixed(sim_value("A", "slope", "mice", "coverage"), 3)` of datasets, with a
bias of `r fixed(sim_value("A", "slope", "mice", "bias"), 4)`. That comes
from predictive mean matching rather than from mice as such. It imputes by
copying observed values, most deletions fall at large $x_1$ where few
observed donors remain, and the copied values cannot reach the upper tail.

In design B, the two groups, the picture changes. The complete-data analysis
covers the slope in
`r fixed(sim_value("B", "slope", "complete data", "coverage"), 3)` of
datasets and proxymix in
`r fixed(sim_value("B", "slope", "proxymix", "coverage"), 3)`, with a bias
of `r fixed(sim_value("B", "slope", "proxymix", "bias"), 4)`, because the
mixture represents the bend between the groups. Amelia's single
multivariate-normal model cannot. Its slope is biased by
`r fixed(sim_value("B", "slope", "Amelia", "bias"), 4)` and its interval
covers the population value in only
`r fixed(sim_value("B", "slope", "Amelia", "coverage"), 3)` of datasets, and
its mean is biased by `r fixed(sim_value("B", "mean", "Amelia", "bias"), 4)`
as well. The mice slope bias is small, at
`r fixed(sim_value("B", "slope", "mice", "bias"), 4)`, but its intervals are
too narrow and cover in
`r fixed(sim_value("B", "slope", "mice", "coverage"), 3)` of datasets. The
proxymix intervals are wider, at
`r fixed(sim_value("B", "slope", "proxymix", "width"), 3)` against
`r fixed(sim_value("B", "slope", "mice", "width"), 3)`, and that width is what
keeps their coverage near nominal. A coverage near 0.95 estimated from
`r sim$n_rep` datasets carries a Monte Carlo standard error of about
`r fixed(mcse_coverage, 2)`, so differences of a hundredth or two can be
noise. Measured against the nominal 0.95 in units of that standard error,
the slope coverage of mice falls `r round(gap_se("A", "mice"))` short in
design A and `r round(gap_se("B", "mice"))` in design B, and that of Amelia
`r round(gap_se("B", "Amelia"))` short in design B. Amelia's own coverage of
`r fixed(sim_value("B", "slope", "Amelia", "coverage"), 3)` carries a
standard error of `r fixed(own_se("B", "Amelia"), 3)`.

The answer to the opening question carries over from the toy to established
imputers. Where the data hold groups that a single Gaussian or a linear
imputation model cannot represent, the mixture keeps both the point estimate
and its interval, and where the data are one Gaussian cloud it performs as
the multivariate-normal model does. Two designs, one missingness mechanism
and one sample size do not map every setting in which imputers differ, and
the penguins were imputed without their species. Supplied as a column,
species would let mice and Amelia model the groups directly.

## Further reading

*Missing data that depends on the missing value* takes up the two
mechanisms assumed away here, censoring at a known bound and
missingness whose probability depends on the hidden value itself, and shows
how to sweep the assumed selection slope, which the observed data inform
only through the assumed shape of the distribution.

*The closed-form operator calculus on a mixture* is the algebra underneath
`gmm_conditionalise()`, and shows what else the same conditioning step buys
once a mixture is in hand.

*One mixture, many methods* places imputation beside the other classical
analyses that fall out of a single fitted mixture.

## References

* Gorman, K. B., Williams, T. D. and Fraser, W. R. (2014). *Ecological
  sexual dimorphism and environmental variability within a community of
  Antarctic penguins (genus Pygoscelis).* PLoS ONE 9(3), e90081.
  <https://doi.org/10.1371/journal.pone.0090081>.
* Honaker, J., King, G. and Blackwell, M. (2011). *Amelia II: A program for
  missing data.* Journal of Statistical Software 45(7), 1--47.
  <https://doi.org/10.18637/jss.v045.i07>.
* Horst, A. M., Presmanes Hill, A. and Gorman, K. B. (2022). *Palmer
  Archipelago penguins data in the palmerpenguins R package -- an
  alternative to Anderson's irises.* The R Journal 14(1), 244--254.
  <https://doi.org/10.32614/RJ-2022-020>.
* Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys.*
  Wiley.
* van Buuren, S. and Groothuis-Oudshoorn, K. (2011). *mice: Multivariate
  imputation by chained equations in R.* Journal of Statistical Software
  45(3), 1--67. <https://doi.org/10.18637/jss.v045.i03>.

## Reproduce

The data-generating seed is `20260620`, set in the chunk that builds the
toy; `gmm_impute()` is given `seed = 1L`, which makes the completions
reproducible without disturbing the ambient random-number state.

The numerical illustration deletes body mass after `set.seed(20260913)` and
gives each imputer the seed 1, through `set.seed(1L)` for Amelia. In the
simulation each dataset is generated after `set.seed()` with its own index.
The stored simulation results come from the code shown, run on
`r sim$run_date` with R `r sim$versions[["R"]]`, proxymix
`r sim$versions[["proxymix"]]`, mice `r sim$versions[["mice"]]` and Amelia
`r sim$versions[["Amelia"]]`, and the run raised `r length(sim$warnings)`
warnings.

```{r session-info, eval = FALSE}
sessionInfo()
```

```{r session-info-print, echo = FALSE}
## printed 68 characters wide; a longer line breaks after its last slash or
## space within the width and continues indented
old_width <- options(width = 68L)
si_vec <- capture.output(sessionInfo())
options(old_width)
wrap_line <- function(s) {
  out <- character(0L)
  while (nchar(s) > 68L) {
    cut <- max(gregexpr("[/ ]", substr(s, 1L, 68L))[[1L]])
    if (cut < 20L) cut <- 68L
    out <- c(out, substr(s, 1L, cut))
    s <- paste0("    ", substr(s, cut + 1L, nchar(s)))
  }
  c(out, s)
}
writeLines(unlist(lapply(si_vec, wrap_line)))
```
