---
title: "Missing data that depends on the missing value"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing data that depends on the missing value}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~45 s under rmarkdown::render() with ggplot2, mice, Amelia,
     AER, survival and sampleSelection installed, the simulation read from
     its stored results; macOS arm64 (Apple silicon), R 4.6.1, one core. The
     two sensitivity sweeps are the cost: five slope values each, at m = 10
     and m = 20 completions. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
options(cli.width = 70)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_mice <- requireNamespace("mice", quietly = TRUE)
```

```{r fixed-notation, include = FALSE}
## Small errors are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
```

## The problem

An analyst has a column where the large values are the ones that went
missing: an assay that saturates, a yield that was never recorded on the
paddocks that did badly, a follow-up that the worst cases did not attend.
The question is: *the bigger values are exactly the ones I did not get, so
the mean I am about to report is too small -- by how much, and how do I put
a defensible number on that when the data I would need to check is the data
I do not have?*

This vignette answers it on a synthetic toy where the deleted values are
retained, so every estimate can be scored against the complete-data truth
it was hiding.

## Package capabilities

In proxymix you do not impute a missing entry so much as condition the
fitted mixture on what its being missing tells you. For an entry missing at
random that is nothing beyond the other coordinates, so the imputation law
is the plain mixture conditional. Two common situations carry more
information than that. An entry can be missing because it fell in a known
interval -- below a detection limit, above a ceiling -- which is
*censoring*. Or it can be missing with a probability that depends on its own
unobserved value, larger values dropping out more often, which is *missing
not at random*; modelling that probability directly is the selection-model
formulation of Diggle and Kenward (1994). Each is a gate $\delta(y)$ multiplying the mixture
conditional before the value is drawn.

`mar()`, `censored()` and `mnar()` build those three gates, and
`gmm_impute()` takes one through its `mechanism` argument.
`proxy_pool()` combines an estimand across the completions.
`proxy_mnar_sensitivity()` is the verb for the case where the selection
slope is assumed: it sweeps the slope and reports, at each value, the pooled
estimate with its interval and the observed-data log-likelihood, so the
reader sees the estimate across the whole range of assumed slopes.

```{r gate-table, echo = FALSE}
knitr::kable(
  data.frame(
    Mechanism = c("missing at random", "censored",
                  "missing not at random"),
    Gate = c("constant", "interval indicator", "selection probability"),
    Constructor = c("mar()", "censored()", "mnar()"),
    Identified = c("yes, given the model", "yes, bound known",
                   "only through the assumed shape"),
    stringsAsFactors = FALSE
  ),
  align = "llll",
  caption = paste0(
    "The three missingness gates, the shape each one multiplies into the ",
    "mixture conditional, and how far the observed data identify it."
  )
)
```

## Addressing the problem

### A value-dependent mechanism biases the ordinary fit

The data are a synthetic two-component mixture in which the second
coordinate `y` goes missing with a probability that grows with `y` itself, so the larger values
are under-represented among the observed. The complete data is generated
first and kept, so the truth is available to score against.

```{r dgp}
set.seed(20260622)
n <- 600L
comp <- sample(1:2, n, replace = TRUE)
mu <- rbind(c(0, 0), c(1.5, 0.5))
chol_r <- chol(matrix(c(1, 0.6, 0.6, 1), 2L))
z_full <- matrix(rnorm(2 * n), n, 2L) %*% chol_r + mu[comp, ]
colnames(z_full) <- c("x1", "y")
truth <- mean(z_full[, 2L])

beta_true <- 0.7
miss <- runif(n) < plogis(-0.5 + beta_true * z_full[, 2L])
dat <- z_full
dat[miss, "y"] <- NA
```

An imputer that assumes the data are missing at random fills the gap from
the conditional of the observed, which is itself short of the large values,
so it stays biased. Supplying the selection slope to `mnar()` lets the fit
account for the mechanism. Both fits use `m` = 10 completions and allow 500
iterations, the limit `proxy_mnar_sensitivity()` uses below.

```{r recover}
m_draws <- 10L
mar_fit <- gmm_impute(dat, N = 2L, m = m_draws, mechanism = mar(),
                      seed = 1L, max_iter = 500L)
mnar_fit <- gmm_impute(dat, N = 2L, m = m_draws,
                       mechanism = mnar("y", beta = beta_true), seed = 1L,
                       max_iter = 500L)
mar_est <- proxy_pool(mar_fit, "y")$estimate
mnar_est <- proxy_pool(mnar_fit, "y")$estimate
```

Analytic pooling is defined for the plain conditional but not for a gated
one, so the gated fits fall back to Rubin's rules (Rubin, 1987) and say so; the returned
`method` attribute records which was used.

```{r recover-table, echo = FALSE}
rec_tbl <- data.frame(
  estimator = c("complete-data truth", "available cases only",
                "missing at random", "missing not at random"),
  estimate = c(truth, mean(dat[!miss, "y"]), mar_est, mnar_est),
  stringsAsFactors = FALSE
)
rec_tbl$bias <- rec_tbl$estimate - truth
knitr::kable(
  rec_tbl, digits = 3L,
  col.names = c("Estimator of the mean of y", "Estimate", "Bias"),
  caption = paste0(
    "The mean of y recovered three ways, against the complete-data truth ",
    "the deletion hid. ", sum(miss), " of ", n, " values are missing."
  )
)
```

The slope `beta` is the strength of the dependence on the log-odds scale,
and `beta = 0` is the missing-at-random case. The intercept is not supplied:
it is calibrated to the observed fraction missing.

### Sweeping the selection slope

Under this selection model the observed data carry information about the
slope, but only through the assumed shape of the distribution of `y`. A
different shape can fit the observed data equally well at another slope. A
single estimate at one slope would therefore rest on that shape, and the
appropriate summary is a curve (Little, 1993). `proxy_mnar_sensitivity()`
sweeps the slope and reports at each value the pooled mean with its
interval, the observed-data log-likelihood of the selection model, and
whether the fit converged. The analyst reads off where a conclusion would
change. Five slope values from 0 to 1.2 are enough to show the shape.

```{r sweep}
sweep <- proxy_mnar_sensitivity(dat, "y",
                                beta_grid = seq(0, 1.2, by = 0.3),
                                N = 2L, m = m_draws, seed = 1L)
covers <- sweep$conf.low <= truth & sweep$conf.high >= truth
beta_first <- sweep$beta[min(which(covers))]
beta_best <- sweep$beta[which.max(sweep$loglik)]
ll_gain <- max(sweep$loglik) - sweep$loglik[1L]
```

```{r sweep-table, echo = FALSE}
sweep_tbl <- as.data.frame(sweep)[, c("beta", "estimate", "conf.low",
                                      "conf.high", "loglik", "converged")]
sweep_tbl$converged <- ifelse(sweep_tbl$converged, "yes", "no")
knitr::kable(
  sweep_tbl,
  digits = c(1L, 3L, 3L, 3L, 1L, 0L),
  col.names = c("Assumed slope", "Pooled mean", "CI lower", "CI upper",
                "Log-likelihood", "Converged"),
  caption = paste0(
    "The pooled mean of y at each assumed selection slope, with the ",
    "observed-data log-likelihood of the selection model and whether the ",
    "mixture fit converged. The complete-data truth is ", round(truth, 3), "."
  )
)
# Monte Carlo standard error of a Rubin-pooled mean: sqrt(B / m)
mc_se0 <- sqrt(sweep$fmi[1L] * sweep$std.error[1L]^2 /
                 (1 + 1 / m_draws) / m_draws)
```

The slope-0 row makes the same missing-at-random assumption as the
missing-at-random row of the previous table, but it is a separate
computation. `proxy_mnar_sensitivity()` refits the mixture through the gated
engine, draws the completions and pools them by Rubin's rules, while the
earlier row pools the plain fit analytically, without drawing. The two
estimates, `r fixed(sweep$estimate[1L], 3)` and `r fixed(mar_est, 3)`, differ
by `r fixed(abs(sweep$estimate[1L] - mar_est), 3)`. With `m` = `r m_draws`
completions, the slope-0 estimate alone has a Monte Carlo standard error of
`r fixed(mc_se0, 3)`.

```{r fig-tornado, eval = has_ggplot2, echo = has_ggplot2, fig.cap = "The pooled mean of y rises with the assumed selection slope, and its interval reaches the complete-data truth near the slope that actually generated the data. The observed data inform the slope only through the assumed shape of y, so the result is the whole curve.", fig.alt = "Pooled mean of y against the assumed selection slope, with a shaded confidence band, a marked point at each swept value, a dashed horizontal line at the complete-data truth, and a vertical rule at the generating slope."}
sweep_df <- as.data.frame(sweep)
ggplot2::ggplot(sweep_df, ggplot2::aes(beta, estimate)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = conf.low, ymax = conf.high),
    fill = "#56B4E9", alpha = 0.3
  ) +
  ggplot2::geom_line(colour = "#0072B2", linewidth = 0.9) +
  ggplot2::geom_point(colour = "#0072B2", size = 2) +
  ggplot2::geom_hline(yintercept = truth, linetype = "dashed",
                      colour = "#000000") +
  ggplot2::geom_vline(xintercept = beta_true, linetype = "dotted",
                      colour = "#D55E00") +
  ggplot2::annotate("text", x = min(sweep_df$beta), y = truth,
                    label = "complete-data truth", hjust = 0, vjust = -0.6,
                    size = 3.2) +
  ggplot2::annotate("text", x = beta_true, y = min(sweep_df$conf.low),
                    label = "generating slope", hjust = 1.05, vjust = 0,
                    size = 3.2, colour = "#D55E00") +
  ggplot2::labs(
    x = "assumed selection slope (log-odds per unit of y)",
    y = "pooled mean of y",
    title = "A sensitivity curve, not a single estimate"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-tornado-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the sensitivity-curve",
    "figure is skipped; the swept values are in the table above.\n")
```

### Censoring is the exact, identifiable case

When the reason for missingness is a known bound, the mechanism is
identified and there is no sensitivity parameter. A detection limit that
hides values of `y` below a threshold is `censored("y", upper = threshold)`,
and proxymix draws from the mixture conditional truncated to the censored
interval rather than substituting a constant such as half the limit.

```{r censor}
thr <- 0.3
cmiss <- z_full[, 2L] < thr
cdat <- z_full
cdat[cmiss, "y"] <- NA
cfit <- gmm_impute(cdat, N = 2L, m = m_draws,
                   mechanism = censored("y", upper = thr), seed = 1L)
cens_est <- proxy_pool(cfit, "y")$estimate
```

```{r censor-table, echo = FALSE}
cens_tbl <- data.frame(
  estimator = c("complete-data truth", "available cases only",
                "half the detection limit", "proxymix, truncated draw"),
  estimate = c(truth, mean(cdat[!cmiss, "y"]),
               mean(ifelse(cmiss, thr / 2, z_full[, 2L])), cens_est),
  stringsAsFactors = FALSE
)
cens_tbl$bias <- cens_tbl$estimate - truth
below_half <- mean(z_full[cmiss, 2L] < thr / 2)
knitr::kable(
  cens_tbl, digits = 3L,
  col.names = c("Estimator of the mean of y", "Estimate", "Bias"),
  caption = paste0(
    "The mean of y under a detection limit at ", thr, ", which hides ",
    sum(cmiss), " of ", n, " values."
  )
)
```

## Interpretation

`r sum(miss)` of the `r n` values of `y` went missing, and because the
larger ones dropped out preferentially the available cases give
`r round(mean(dat[!miss, "y"]), 3)` against a complete-data truth of
`r round(truth, 3)`, understating it by
`r round(truth - mean(dat[!miss, "y"]), 3)`. Assuming missingness at random
recovers only part of that: the pooled mean is `r round(mar_est, 3)`, still
`r round(truth - mar_est, 3)` low, because the conditional it imputes from
is itself built on the shortened observed values. Supplying the generating
slope to `mnar()` brings the pooled mean to `r round(mnar_est, 3)`, within
`r round(abs(mnar_est - truth), 3)` of the truth. That is the answer to the
first half of the opening question: the bias is real, it is about
`r round(truth - mean(dat[!miss, "y"]), 1)` on this scale, and assuming it
away does not remove it.

The second half is the sensitivity curve. The pooled mean rises
monotonically across the swept slopes, from `r round(sweep$estimate[1], 3)`
at slope 0 to `r round(sweep$estimate[nrow(sweep)], 3)` at slope
`r max(sweep$beta)`, and its interval first reaches the complete-data truth
at an assumed slope of `r beta_first`, against the
`r beta_true` that generated the data. The mixture fit
`r if (all(sweep$converged)) "converged at every slope" else paste0("did not converge at slope ", paste(sweep$beta[!sweep$converged], collapse = ", "))`.
The log-likelihood is highest at slope `r beta_best`,
`r fixed(ll_gain, 1)` units above its value at slope 0, so under this model
the observed data lean towards the larger slopes. That lean comes from the
assumed two-component shape of `y`, not from anything the data show about
why values went missing. What belongs in a report is the curve itself, with
the reader's own view of a plausible slope applied to it.

Censoring is the case where no sweep is needed. With `r sum(cmiss)` values
hidden below a limit of `r thr`, the available cases give
`r round(mean(cdat[!cmiss, "y"]), 3)`, overstating the truth by
`r round(mean(cdat[!cmiss, "y"]) - truth, 3)`. The common substitution of
half the detection limit gives `r round(mean(ifelse(cmiss, thr / 2,
z_full[, 2L])), 3)`, still `r round(mean(ifelse(cmiss, thr / 2,
z_full[, 2L])) - truth, 3)` high. Half the limit is a convention for
concentrations, which cannot fall below zero; here `y` is unbounded below,
and `r round(100 * below_half)` per cent of the hidden values lie below
`r thr / 2`, so the substitute sits above most of them. Drawing from the mixture conditional
truncated to the censored interval gives `r round(cens_est, 3)`, within
`r round(abs(cens_est - truth), 3)` of the truth, because the bound is
known and so the gate is exact rather than assumed.

## Limitations

The interval and value-dependent gates act on a single coordinate, and a row
missing that coordinate is assumed to have its other coordinates observed,
which is the detection-limit or single-outcome setting rather than a general
pattern of holes across many columns. A model estimand other than a
coordinate mean is pooled by handing the completions to `mice` through
`as_mids()`, and inherits Rubin's large-sample assumptions.

The sensitivity sweep prices one departure at a time: the selection
probability is a logistic function of the missing coordinate alone, with the
intercept calibrated to the observed missing fraction, so a mechanism that
depends jointly on the missing value and an unobserved third variable is
outside what the sweep brackets. The sweep is also not a test. Its
log-likelihood favours some slopes only because of the assumed shape of
`y`, so it cannot tell an analyst which slope is right. The sweep shows how
the estimate moves with the assumed slope, and the range is only as wide as
the grid the analyst chose.

The numbers here come from one synthetic realisation at `n` = `r n` with
`m` = `r m_draws` completions, and would move under a different seed. What
would change the answer most is the component count and the assumed slope:
a mixture with too few components cannot represent the two clusters that
make the conditional informative, and a slope far from the truth moves the
pooled mean roughly linearly, which is what the sensitivity table shows.
The numerical illustration that follows places both gates beside the
established tools for each case, the Tobit regression for censoring and
delta-adjusted multiple imputation for the sensitivity sweep, on real data
and in simulation.

```{=latex}
\vspace{0pt plus 10\baselineskip}\penalty-200\vspace{0pt plus -10\baselineskip}
```

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_amelia <- requireNamespace("Amelia", quietly = TRUE)
has_aer <- requireNamespace("AER", quietly = TRUE)
has_survival <- requireNamespace("survival", quietly = TRUE)
has_selection <- requireNamespace("sampleSelection", quietly = TRUE)
has_penguins <- exists("penguins", envir = asNamespace("datasets"))
run_illustration <- has_mice && has_amelia && has_aer && has_survival &&
  has_selection && has_penguins
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("mice, Amelia, AER, survival, sampleSelection or the penguins data",
    "(R 4.5.0 or later) is not available on this build, so the real-data",
    "illustration is skipped.\n")
```

### A censored outcome: the Affairs data with three estimators

The censored case has an established estimator, the Tobit regression of
Tobin (1958), which treats an outcome recorded as zero as a latent value at
or below zero and maximises the likelihood of a linear-Gaussian model for
the latent value. `tobit()` in AER (Kleiber and Zeileis, 2008) and
`survreg()` in survival (Therneau and Grambsch, 2000) both fit it, the
first by calling the second. The proxymix route reaches the same model from the other
side: the zeros are set to missing, `censored()` says they lie at or below
zero, and the fitted mixture conditional, truncated to that interval,
supplies completed values that an ordinary regression is then run on and
pooled.

The competitors install from CRAN once and load as usual:

```{r install-competitors, eval = FALSE}
install.packages(c("mice", "Amelia", "AER", "survival", "sampleSelection"))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(mice)
library(Amelia)
library(AER)
library(survival)
library(sampleSelection)
```

The data are the `Affairs` survey of Fair (1978), shipped with AER and the
package's own Tobit example. The number of affairs in the past year is
regressed on age, years married, religiousness, occupation and a
self-rating of the marriage:

```{r affairs-data, eval = run_illustration}
data("Affairs", package = "AER")
cols_aff <- c("affairs", "age", "yearsmarried", "religiousness",
              "occupation", "rating")
aff <- Affairs[, cols_aff]
f_aff <- affairs ~ age + yearsmarried + religiousness + occupation + rating
```

Of the `r nrow(aff)` respondents, `r sum(aff$affairs == 0)` reported no
affairs, which the Tobit model reads as a latent value at or below zero.

```{r affairs-fit, eval = run_illustration}
fit_tobit <- tobit(f_aff, data = aff)
fit_survreg <- survreg(
  Surv(affairs, affairs > 0, type = "left") ~ age + yearsmarried +
    religiousness + occupation + rating,
  data = aff, dist = "gaussian"
)

aff_holes <- aff
aff_holes$affairs[aff$affairs == 0] <- NA
imp_aff <- gmm_impute(aff_holes, N = 1L, m = 20L,
                      mechanism = censored("affairs", upper = 0), seed = 1L)
imp_aff_bic <- gmm_impute(aff_holes, m = 20L,
                          mechanism = censored("affairs", upper = 0),
                          seed = 1L)
pool_aff <- function(imp) {
  fits <- lapply(complete(as_mids(imp), "all"),
                 function(d) lm(f_aff, data = d))
  summary(pool(fits))
}
pooled_aff <- pool_aff(imp_aff)
pooled_aff_bic <- pool_aff(imp_aff_bic)
```

```{r affairs-table, eval = run_illustration, echo = FALSE}
k_aff <- length(coef(fit_tobit))
est_se <- function(est, se, digits = 3L) {
  num <- function(v) format(round(v, digits), nsmall = digits)
  paste0(num(est), " (", num(se), ")")
}
aff_tbl <- data.frame(
  term = c("intercept", "age", "years married", "religiousness",
           "occupation", "rating"),
  tobit = est_se(coef(fit_tobit), sqrt(diag(vcov(fit_tobit)))[seq_len(k_aff)]),
  survreg = est_se(coef(fit_survreg),
                   sqrt(diag(vcov(fit_survreg)))[seq_len(k_aff)]),
  proxymix = est_se(pooled_aff$estimate, pooled_aff$std.error),
  proxymix_bic = est_se(pooled_aff_bic$estimate, pooled_aff_bic$std.error),
  stringsAsFactors = FALSE
)
knitr::kable(
  aff_tbl, align = c("l", "r", "r", "r", "r"),
  col.names = c("Coefficient", "AER tobit", "survival survreg",
                "proxymix, N = 1", paste0("proxymix, N = ",
                                          imp_aff_bic@n_components)),
  caption = paste0(
    "Coefficients of the latent regression of affairs on five covariates, ",
    "with standard errors in brackets, from the Tobit likelihood in two ",
    "packages and from proxymix's truncated imputations pooled by Rubin's ",
    "rules, with the component count fixed at one and selected by BIC."
  )
)
gap_aff <- abs(pooled_aff$estimate - coef(fit_tobit)) /
  sqrt(diag(vcov(fit_tobit)))[seq_len(k_aff)]
slopes_aff <- seq_len(k_aff)[-1L]
towards_zero <- slopes_aff[abs(pooled_aff_bic$estimate[slopes_aff]) <
                             abs(coef(fit_tobit)[slopes_aff])]
# shift from the Tobit estimate in units of the Tobit standard error
shift_aff <- abs(pooled_aff_bic$estimate - coef(fit_tobit)) /
  sqrt(diag(vcov(fit_tobit)))[seq_len(k_aff)]
largest_shift <- towards_zero[which.max(shift_aff[towards_zero])]
name_list <- function(v) {
  if (length(v) == 1L) return(v)
  paste(paste(v[-length(v)], collapse = ", "), "and", v[length(v)])
}
```

The two Tobit fits are the same fit, as the identical columns show. With a
single component, proxymix's model of the latent value given the covariates
is the same linear-Gaussian model, estimated by expectation-maximisation
over the truncated conditional rather than by direct maximisation, and its
pooled coefficients fall within
`r fixed(ceiling(10 * max(gap_aff)) / 10, 1)` Tobit standard errors of the
Tobit estimates.
Left to choose its component count, proxymix selects
`r imp_aff_bic@n_components` components for the joint distribution of the
six columns, all of them coarsely coded and three of them integer scores,
and that fit
`r if (imp_aff_bic@diagnostics$converged) "converged" else "had not converged"`
after `r imp_aff_bic@diagnostics$iterations` iterations, the default limit.
The regression on its completions is then the linear projection of a
conditional that is no longer linear. Against the Tobit estimates, the
coefficients on `r name_list(aff_tbl$term[towards_zero])` move towards zero,
the largest in Tobit standard errors being on
`r aff_tbl$term[largest_shift]`, from
`r fixed(coef(fit_tobit)[largest_shift], 3)` to
`r fixed(pooled_aff_bic$estimate[largest_shift], 3)`, or
`r fixed(shift_aff[largest_shift], 2)` standard errors. On this data the Tobit model is the natural
reading, and the single component is the like-for-like comparison; the
simulation below scores it against the truth.

### Value-dependent deletion: the Palmer penguins data

Established estimators for the missing-not-at-random case exist under a
parametric selection model, such as the two-step estimator of Heckman
(1979). Their answer rests on the assumed shape of the outcome distribution,
so the standard practice is a sensitivity analysis, and the standard tool is
delta adjustment in multiple imputation (van Buuren, 2018, Section 9.2),
available in mice (van Buuren and Groothuis-Oudshoorn, 2011). Delta
adjustment keeps the imputation model fixed and shifts its draws: impute
under the missing-at-random assumption, add a shift $\delta$ to every
imputed value, and repeat over a grid of shifts. The proxymix sweep is a
different kind of sweep. At each slope of the log-odds of being missing in
the missing value, it refits the mixture under the selection model and
imputes from that fit. Two further estimators are run for reference. Amelia (Honaker et al.,
2011) imputes under missing at random and shows what ignoring the mechanism
costs. The Heckman two-step estimator, from sampleSelection (Toomet and
Henningsen, 2008), fits a probit selection equation and corrects the outcome
regression with its inverse Mills ratio.

The data are the Palmer penguins (Gorman et al., 2014), shipped with R as
`penguins` since version 4.5.0 (Horst et al., 2022), with body mass in kilograms. Body mass is deleted with a
probability that rises with body mass itself, at a log-odds slope of 1.5
per kilogram, and the deleted values stay aside for scoring.

```{r penguins-data, eval = run_illustration}
cols <- c("bill_len", "bill_dep", "flipper_len", "body_mass")
peng <- na.omit(penguins[, cols])
peng$body_mass <- peng$body_mass / 1000

set.seed(20260925)
gone <- runif(nrow(peng)) < plogis(1.5 * (peng$body_mass - 4.5))
peng_holes <- peng
peng_holes$body_mass[gone] <- NA
mass_truth <- mean(peng$body_mass)
```

That leaves `r nrow(peng)` birds, of which `r sum(gone)` lose their body
mass. The mean of the retained masses is
`r fixed(mean(peng_holes$body_mass, na.rm = TRUE), 3)` kg against
`r fixed(mass_truth, 3)` kg over all the birds.

The proxymix sweep covers five slopes from 0 to 3 per kilogram, which
bracket the generating slope, at 20 completions each. The mice sweep covers
five shifts from 0 to
0.6 kg, three quarters of a standard deviation of body mass, with
imputation by the normal linear model so that the shift acts on a
model-based draw:

```{r penguins-sweep, eval = run_illustration}
beta_peng <- seq(0, 3, by = 0.75)
delta_peng <- seq(0, 0.6, by = 0.15)

sweep_peng <- proxy_mnar_sensitivity(peng_holes, "body_mass",
                                     beta_grid = beta_peng, m = 20L,
                                     seed = 1L)

pool_mean <- function(fits) {
  s <- summary(pool(fits), conf.int = TRUE)
  unlist(s[1L, c("estimate", "conf.low", "conf.high")])
}
mice_delta <- function(delta) {
  post <- make.post(peng_holes)
  post["body_mass"] <- paste0("imp[[j]][, i] <- imp[[j]][, i] + ", delta)
  imp <- mice(peng_holes, m = 20L, method = "norm", post = post, seed = 1L,
              printFlag = FALSE)
  pool_mean(with(imp, lm(body_mass ~ 1))$analyses)
}
sweep_mice <- t(vapply(delta_peng, mice_delta, numeric(3L)))

set.seed(1L)
imp_amelia <- amelia(peng_holes, m = 20L, p2s = 0L)
mean_amelia <- pool_mean(lapply(imp_amelia$imputations,
                                function(d) lm(body_mass ~ 1, data = d)))

peng_sel <- peng_holes
peng_sel$seen <- !is.na(peng_sel$body_mass)
fit_heckit <- heckit(seen ~ bill_len + bill_dep + flipper_len,
                     body_mass ~ bill_len + bill_dep + flipper_len,
                     data = peng_sel)
# the mean of the outcome equation over every bird, delta-method interval
heckit_mean <- function(fit, data, outcome_cols) {
  x_bar <- c(1, colMeans(data[, outcome_cols]))
  b <- coef(fit)[seq_along(x_bar) + length(x_bar)]
  v <- vcov(fit)[seq_along(x_bar) + length(x_bar),
                 seq_along(x_bar) + length(x_bar)]
  est <- sum(x_bar * b)
  se <- sqrt(as.numeric(t(x_bar) %*% v %*% x_bar))
  c(estimate = est, conf.low = est - qnorm(0.975) * se,
    conf.high = est + qnorm(0.975) * se)
}
mean_heckit <- heckit_mean(fit_heckit, peng_sel, cols[1:3])
```

```{r penguins-values, eval = run_illustration, echo = FALSE}
mean_ci <- function(x) {
  c(mean(x), mean(x) + c(-1, 1) * qt(0.975, length(x) - 1) *
      sd(x) / sqrt(length(x)))
}
peng_rows <- rbind(
  mean_ci(peng$body_mass),
  mean_ci(peng_holes$body_mass[!gone]),
  as.matrix(sweep_peng[, c("estimate", "conf.low", "conf.high")]),
  sweep_mice,
  mean_amelia,
  mean_heckit
)
peng_tbl <- data.frame(
  method = c("complete data", "available cases",
             rep("proxymix, slope", length(beta_peng)),
             rep("mice, shift, kg", length(delta_peng)),
             "Amelia", "heckit"),
  parameter = c("--", "--", format(beta_peng), format(delta_peng), "--", "--"),
  mass = paste0(
    format(round(peng_rows[, 1L], 3), nsmall = 3), " (",
    format(round(peng_rows[, 2L], 3), nsmall = 3), ", ",
    format(round(peng_rows[, 3L], 3), nsmall = 3), ")"),
  error = fixed(peng_rows[, 1L] - mass_truth, 3),
  loglik = c("--", "--", fixed(sweep_peng$loglik, 1),
             rep("--", length(delta_peng) + 2L)),
  stringsAsFactors = FALSE
)
beta_peng_best <- beta_peng[which.max(sweep_peng$loglik)]
covers_peng <- peng_rows[, 2L] <= mass_truth & mass_truth <= peng_rows[, 3L]
names(covers_peng) <- c("complete", "available",
                        paste0("proxymix_", beta_peng),
                        paste0("mice_", delta_peng), "Amelia", "heckit")
gen_row <- 2L + which(beta_peng == 1.5)
## the Heckman fit's error correlation and outcome coefficient on flipper
## length, beside the complete-data regression
heck_rho <- coef(fit_heckit)[["rho"]]
heck_flipper <- coef(fit_heckit)[8L]  # outcome coefficients are 5 to 8
ols_flipper <- coef(lm(body_mass ~ bill_len + bill_dep + flipper_len,
                       data = peng))[["flipper_len"]]
```

Deleting the heavier birds pulls the mean of the available cases down by
`r fixed(mass_truth - peng_rows[2L, 1L], 3)` kg. The proxymix curve at slope
0 and Amelia both impute under missing at random, and both recover most of
that from the three observed measurements, which carry the species
structure: their errors are `r fixed(peng_rows[3L, 1L] - mass_truth, 3)`
and `r fixed(mean_amelia[["estimate"]] - mass_truth, 3)` kg. At the
generating slope of 1.5 the proxymix estimate is
`r fixed(peng_rows[gen_row, 1L] - mass_truth, 3)` kg from the truth. The
mice curve starts lower, at
`r fixed(sweep_mice[1L, 1L] - mass_truth, 3)` kg, and crosses the truth
between shifts of `r delta_peng[max(which(sweep_mice[, 1L] < mass_truth))]`
and `r delta_peng[min(which(sweep_mice[, 1L] > mass_truth))]` kg. Both
curves reach the truth. Every mice shift uses the same imputation model, so
the observed data fit equally well at every shift. The proxymix fit
`r if (all(sweep_peng$converged)) "converged at every slope" else paste0("did not converge at slope ", paste(sweep_peng$beta[!sweep_peng$converged], collapse = ", "))`,
and its log-likelihood is highest at slope `r beta_peng_best`. That lean
comes from the assumed mixture shape of the four measurements, not from the
deletion mechanism. The Heckman estimator's selection equation has no variable that predicts
deletion but not body mass, so it is identified by the shape of the inverse
Mills ratio alone. Here that fit breaks down. Its mean is
`r fixed(mean_heckit[["estimate"]], 2)` kg, with an interval from
`r fixed(mean_heckit[["conf.low"]], 2)` to
`r fixed(mean_heckit[["conf.high"]], 2)` kg, while the lightest bird weighs
`r fixed(min(peng$body_mass), 2)` kg. Its estimated correlation between the
selection and outcome errors is `r fixed(heck_rho, 2)`, outside $[-1, 1]$,
and its coefficient on flipper length is `r fixed(heck_flipper, 3)` kg per
millimetre against `r fixed(ols_flipper, 3)` in the complete data. This is one deletion of one dataset, so the simulation repeats
it.

```{=latex}
\par\penalty-100\begingroup\dimen0=26\baselineskip\dimen2=\pagegoal
\advance\dimen2 by -\pagetotal
\ifdim\dimen0>\dimen2 \ifdim\dimen2>0pt \vfil\fi\break\fi\endgroup
```

```{r penguins-table, eval = run_illustration, echo = FALSE}
knitr::kable(
  peng_tbl, align = c("l", "r", "r", "r", "r"),
  col.names = c("Method", "Value", "Mean body mass, kg",
                "Error", "Log-likelihood"),
  caption = paste0(
    "The mean body mass of ", nrow(peng), " penguins with ", sum(gone),
    " masses deleted with a probability that increases with the mass, ",
    "logistic in the mass minus 4.5 kg, from each estimator, with a 95% ",
    "interval, its error against the mean of the complete data before ",
    "deletion, and for proxymix the observed-data log-likelihood of the ",
    "selection model. Value is the sensitivity parameter, and a dash marks ",
    "an estimator that has none. heckit is the Heckman two-step estimator ",
    "from sampleSelection."
  )
)
```

```{r fig-penguins, eval = run_illustration && has_ggplot2, echo = FALSE, fig.height = 3.8, fig.cap = "The two sensitivity curves for the mean body mass of the penguins, proxymix over the assumed selection slope and mice over the shift added to each imputed value, with 95% bands and the complete-data mean as a dashed line. Each curve crosses the truth somewhere along its grid. The observed data inform the proxymix slope only through the assumed mixture shape, and do not inform the mice shift.", fig.alt = "Two panels. Left, the proxymix pooled mean of body mass against the assumed selection slope from 0 to 3, with a shaded band. Right, the mice delta-adjusted mean against the shift from 0 to 0.6 kg, with a shaded band. A dashed horizontal line marks the complete-data mean in both panels."}
curve_df <- rbind(
  data.frame(method = "proxymix: selection slope, log-odds per kg",
             parameter = beta_peng,
             estimate = sweep_peng$estimate,
             conf.low = sweep_peng$conf.low,
             conf.high = sweep_peng$conf.high),
  data.frame(method = "mice: shift added to each imputed value, kg",
             parameter = delta_peng,
             estimate = sweep_mice[, 1L],
             conf.low = sweep_mice[, 2L],
             conf.high = sweep_mice[, 3L])
)
curve_df$method <- factor(curve_df$method, levels = unique(curve_df$method))
ggplot2::ggplot(curve_df, ggplot2::aes(parameter, estimate)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = conf.low, ymax = conf.high),
                       fill = "#56B4E9", alpha = 0.3) +
  ggplot2::geom_line(colour = "#0072B2", linewidth = 0.9) +
  ggplot2::geom_point(colour = "#0072B2", size = 2) +
  ggplot2::geom_hline(yintercept = mass_truth, linetype = "dashed") +
  ggplot2::facet_wrap(~ method, scales = "free_x") +
  ggplot2::labs(x = "sensitivity parameter", y = "mean body mass, kg",
                title = "Two sensitivity curves for one mean") +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-penguins-skip, eval = run_illustration && !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the sensitivity-curve",
    "figure is skipped; the swept values are in the table above.\n")
```

```{=latex}
\FloatBarrier
```

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("missing_data_mnar_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
y_density <- function(y) 0.5 * dnorm(y, 0) + 0.5 * dnorm(y, 0.5)
share_mnar <- integrate(function(y) plogis(-0.5 + 0.7 * y) * y_density(y),
                        -Inf, Inf)$value
```

The simulation repeats each analysis on `r sim$n_rep` datasets of
`r sim$n` rows in two designs whose population values are known. The
missing-not-at-random design is the toy above at half its size: two equally
likely components centred at $(0, 0)$ and $(1.5, 0.5)$ with unit variances
and correlation 0.6, and $y$ deleted with probability
$\mathrm{logit}^{-1}(-0.5 + 0.7\, y)$, which removes
`r round(100 * share_mnar)` per cent of $y$ on average. The estimand is the
mean of $y$, whose population value is `r sim$truth[["mnar_mean"]]`.
The proxymix sweep covers the slopes `r paste(sim$beta_grid, collapse = ", ")`, so
that the generating slope is on the grid, and the mice sweep covers the shifts
`r paste(sim$delta_grid, collapse = ", ")` in units of $y$, about a fifth
of its standard deviation per step. Amelia, the Heckman two-step estimator
with $x_1$ in both equations, and the available cases are scored beside
them. In the censored design, $x$ is standard normal and
$y^* = a + x + \varepsilon$ with standard normal $\varepsilon$ and $a$ set
so that 30 per cent of $y^*$ falls below zero; the recorded $y$ is
$\max(y^*, 0)$. The estimands are the intercept and slope of the latent
regression, scored for the Tobit likelihood in AER and survival, for
proxymix's truncated imputations with the component count selected by BIC,
and for the regression that takes the recorded zeros at face value.

Each imputer makes `r sim$m` completions, and every interval is a 95 per
cent interval: Rubin's rules for the imputers, Wald for the Tobit fits and
the Heckman mean, and the usual regression or $t$ interval otherwise. The
analysis of the complete data, before deletion or censoring, is the
reference. `proxy_mnar_sensitivity()` warns when a fit does not converge,
and the stored run raised `r length(sim$warnings)` warnings, so
`r if (length(sim$warnings) == 0L) "every sweep fit converged within its 500 iterations" else "some sweep fits did not converge; the Reproduce section lists the warnings"`.
This chunk is complete and runs as shown, but it took about
`r round(sim$elapsed_secs / 60)` minutes on one core (R
`r sim$versions[["R"]]`, Apple silicon), so it is not run when the
vignette builds and the tables below it are read from its stored output. For a quick reduced run, change the line `n_rep <- 300L` to a small
number such as `n_rep <- 3L`.

```{r sim-code, eval = FALSE}
library(proxymix)
library(mice)
library(Amelia)
library(AER)
library(survival)
library(sampleSelection)

n <- 300L      # rows in each dataset
m <- 10L       # completions per imputer
n_rep <- 300L  # datasets per design

beta_grid <- c(0, 0.35, 0.7, 1.05)   # proxymix selection slopes
delta_grid <- c(0, 0.2, 0.4, 0.6)    # mice shifts, in units of y
a_cens <- -sqrt(2) * qnorm(0.3)      # 30% of the latent values fall below 0

truth <- c(mnar_mean = 0.25, cens_intercept = a_cens, cens_slope = 1)

# estimate and 95% interval for every coefficient: the usual interval for
# one dataset, Rubin's rules over several completions
ci_fits <- function(fits) {
  if (length(fits) == 1L) {
    return(cbind(coef(fits[[1L]]), confint(fits[[1L]])))
  }
  s <- summary(pool(fits), conf.int = TRUE)
  as.matrix(s[, c("estimate", "conf.low", "conf.high")])
}

# Wald interval for the regression coefficients of a Tobit fit
ci_wald <- function(fit, k) {
  est <- coef(fit)[seq_len(k)]
  se <- sqrt(diag(vcov(fit)))[seq_len(k)]
  cbind(est, est - qnorm(0.975) * se, est + qnorm(0.975) * se)
}

# the mean of a Heckman outcome equation over every row, delta method
ci_heckit_mean <- function(fit, x_bar) {
  idx <- seq_along(x_bar) + length(x_bar)
  est <- sum(x_bar * coef(fit)[idx])
  v <- as.numeric(t(x_bar) %*% vcov(fit)[idx, idx] %*% x_bar)
  se <- if (is.finite(v) && v >= 0) sqrt(v) else NA_real_
  cbind(est, est - qnorm(0.975) * se, est + qnorm(0.975) * se)
}

rows_of <- function(design, estimand, method, ci) {
  data.frame(design = design, estimand = estimand, method = method,
             estimate = ci[, 1L], lower = ci[, 2L], upper = ci[, 3L])
}

mu <- rbind(c(0, 0), c(1.5, 0.5))
chol_r <- chol(matrix(c(1, 0.6, 0.6, 1), 2L))

one_mnar <- function(r) {
  set.seed(r)
  comp <- sample(1:2, n, replace = TRUE)
  z <- matrix(rnorm(2 * n), n, 2L) %*% chol_r + mu[comp, ]
  full <- data.frame(x1 = z[, 1L], y = z[, 2L])
  obs <- full
  obs$y[runif(n) < plogis(-0.5 + 0.7 * full$y)] <- NA
  seen <- !is.na(obs$y)
  out <- list(
    rows_of("mnar", "mean", "complete data",
            ci_fits(list(lm(y ~ 1, data = full)))),
    rows_of("mnar", "mean", "available cases",
            ci_fits(list(lm(y ~ 1, data = obs[seen, ]))))
  )
  sw <- proxy_mnar_sensitivity(obs, "y", beta_grid = beta_grid, m = m,
                               seed = r)
  out[[3L]] <- rows_of("mnar", "mean", paste0("proxymix, slope ", sw$beta),
                       as.matrix(sw[, c("estimate", "conf.low", "conf.high")]))
  out[[4L]] <- do.call(rbind, lapply(delta_grid, function(delta) {
    post <- make.post(obs)
    post["y"] <- paste0("imp[[j]][, i] <- imp[[j]][, i] + ", delta)
    imp <- mice(obs, m = m, method = "norm", post = post, seed = r,
                printFlag = FALSE)
    rows_of("mnar", "mean", paste0("mice, shift ", delta),
            ci_fits(with(imp, lm(y ~ 1))$analyses))
  }))
  out[[5L]] <- rows_of("mnar", "mean", "Amelia",
                       ci_fits(lapply(amelia(obs, m = m, p2s = 0L)$imputations,
                                      function(d) lm(y ~ 1, data = d))))
  obs$seen <- seen
  fit <- heckit(seen ~ x1, y ~ x1, data = obs)
  out[[6L]] <- rows_of("mnar", "mean", "sampleSelection heckit",
                       ci_heckit_mean(fit, c(1, mean(obs$x1))))
  do.call(rbind, out)
}

one_censored <- function(r) {
  set.seed(r)
  x <- rnorm(n)
  y_star <- a_cens + x + rnorm(n)
  full <- data.frame(x = x, y = y_star)
  obs <- data.frame(x = x, y = pmax(y_star, 0))
  holes <- obs
  holes$y[obs$y == 0] <- NA
  imp <- gmm_impute(holes, m = m, mechanism = censored("y", upper = 0),
                    seed = r)
  cis <- list(
    "complete data" = ci_fits(list(lm(y ~ x, data = full))),
    "zeros at face value" = ci_fits(list(lm(y ~ x, data = obs))),
    "AER tobit" = ci_wald(tobit(y ~ x, data = obs), 2L),
    "survival survreg" = ci_wald(
      survreg(Surv(y, y > 0, type = "left") ~ x, data = obs,
              dist = "gaussian"), 2L),
    proxymix = ci_fits(lapply(complete(as_mids(imp), "all"),
                              function(d) lm(y ~ x, data = d)))
  )
  do.call(rbind, lapply(names(cis), function(method) {
    rows_of("censored", c("intercept", "slope"), method, cis[[method]])
  }))
}

res <- rbind(
  do.call(rbind, lapply(seq_len(n_rep), one_mnar)),
  do.call(rbind, lapply(seq_len(n_rep), one_censored))
)
res$truth <- truth[ifelse(res$design == "mnar", "mnar_mean",
                          paste0("cens_", res$estimand))]
n_undefined <- sum(is.na(res$lower))

sim_tab <- aggregate(
  cbind(bias = estimate - truth, rmse = (estimate - truth)^2,
        coverage = lower <= truth & truth <= upper,
        width = upper - lower) ~ design + estimand + method,
  data = res, FUN = function(v) mean(v, na.rm = TRUE),
  na.action = na.pass
)
sim_tab$rmse <- sqrt(sim_tab$rmse)
sim_tab
```

```{=latex}
\par\penalty-100\begingroup\dimen0=20\baselineskip\dimen2=\pagegoal
\advance\dimen2 by -\pagetotal
\ifdim\dimen0>\dimen2 \ifdim\dimen2>0pt \vfil\fi\break\fi\endgroup
```

```{r sim-tables, echo = FALSE}
tab_sim <- sim$sim_tab
method_mnar <- c("complete data", "available cases",
                 paste0("proxymix, slope ", sim$beta_grid),
                 paste0("mice, shift ", sim$delta_grid),
                 "Amelia", "sampleSelection heckit")
method_cens <- c("complete data", "zeros at face value", "AER tobit",
                 "survival survreg", "proxymix")
tab_mnar <- tab_sim[tab_sim$design == "mnar", ]
tab_mnar <- tab_mnar[match(method_mnar, tab_mnar$method), ]
tab_cens <- tab_sim[tab_sim$design == "censored", ]
tab_cens <- tab_cens[order(match(tab_cens$estimand, c("intercept", "slope")),
                           match(tab_cens$method, method_cens)), ]
knitr::kable(
  tab_mnar[, c("method", "bias", "rmse", "coverage", "width")],
  digits = 3L, row.names = FALSE,
  col.names = c("Method", "Bias", "RMSE", "Coverage", "Width"),
  caption = paste0(
    "The mean of y under value-dependent deletion: bias and root mean ",
    "squared error of the point estimate, coverage of the nominal 95% ",
    "interval and its average width, over ", sim$n_rep, " simulated ",
    "datasets, each scored against the population mean of ",
    sim$truth[["mnar_mean"]], "."
  )
)
```

```{r sim-table-cens, echo = FALSE}
knitr::kable(
  tab_cens[, c("estimand", "method", "bias", "rmse", "coverage", "width")],
  digits = 3L, row.names = FALSE,
  col.names = c("Estimand", "Method", "Bias", "RMSE", "Coverage", "Width"),
  caption = paste0(
    "The latent regression under 30% censoring at zero: bias and root mean ",
    "squared error, coverage of the nominal 95% interval and its average ",
    "width, over ", sim$n_rep, " simulated datasets, each scored against ",
    "the generating intercept of ", round(sim$truth[["cens_intercept"]], 3),
    " and slope of ", sim$truth[["cens_slope"]], "."
  )
)
sim_value <- function(design, estimand, method, what) {
  s1 <- sim$sim_tab$design == design & sim$sim_tab$estimand == estimand &
    sim$sim_tab$method == method
  sim$sim_tab[[what]][s1]
}
mv <- function(method, what) sim_value("mnar", "mean", method, what)
cv <- function(estimand, method, what) {
  sim_value("censored", estimand, method, what)
}
mcse_coverage <- sqrt(0.95 * 0.05 / sim$n_rep)
# a version as its DESCRIPTION writes it, when the installed one is the one stored
desc_version <- function(pkg, stored) {
  if (identical(as.character(packageVersion(pkg)), stored)) {
    utils::packageDescription(pkg)$Version
  } else {
    stored
  }
}
pv <- function(pkg) desc_version(pkg, sim$versions[[pkg]])
cov_mar <- c(mv("proxymix, slope 0", "coverage"),
             mv("mice, shift 0", "coverage"), mv("Amelia", "coverage"))
```

Under value-dependent deletion, every estimator that assumes missing at
random shares one bias. The available cases understate the mean by
`r fixed(-mv("available cases", "bias"), 3)`, and proxymix at slope 0, mice
at shift 0 and Amelia by
`r fixed(-mv("proxymix, slope 0", "bias"), 3)`,
`r fixed(-mv("mice, shift 0", "bias"), 3)` and
`r fixed(-mv("Amelia", "bias"), 3)`, with coverage of
`r fixed(min(cov_mar), 2)` to `r fixed(max(cov_mar), 2)`, because $x_1$ carries only part of
what the deletion depends on. Along each grid the bias falls through zero.
At the generating slope of 0.7, proxymix has a bias of
`r fixed(mv("proxymix, slope 0.7", "bias"), 3)` and covers the population
mean in `r fixed(mv("proxymix, slope 0.7", "coverage"), 3)` of datasets,
against `r fixed(mv("complete data", "coverage"), 3)` for the complete data.
At a shift of 0.4, mice has a bias of
`r fixed(mv("mice, shift 0.4", "bias"), 3)` and coverage of
`r fixed(mv("mice, shift 0.4", "coverage"), 3)`. Both curves reach the
truth somewhere on their grids. What differs is the parameter the analyst
must have a view about: a slope on the log-odds scale, which the observed
data inform only through the assumed mixture shape, or a shift in the units
of $y$, which they do not inform. The Heckman two-step estimator, with $x_1$ in both
equations and no exclusion restriction, has a root mean squared error of
`r round(mv("sampleSelection heckit", "rmse"))` and an average interval
width of `r round(mv("sampleSelection heckit", "width"))`, and in
`r sim$n_undefined` datasets its interval was undefined because the
delta-method variance was not positive; its coverage is over the remaining
datasets. A coverage near 0.95 estimated from `r sim$n_rep` datasets carries
a Monte Carlo standard error of about `r fixed(mcse_coverage, 3)`, so
differences of a hundredth or two can be noise.

In the censored design, the two Tobit fits are again one fit, with slope
bias `r fixed(cv("slope", "AER tobit", "bias"), 3)` and coverage
`r fixed(cv("slope", "AER tobit", "coverage"), 3)`, against
`r fixed(cv("slope", "complete data", "bias"), 3)` and
`r fixed(cv("slope", "complete data", "coverage"), 3)` for the complete
data. Taking the zeros at face value biases the slope by
`r fixed(cv("slope", "zeros at face value", "bias"), 3)` and never covers
it. The truncated imputations of proxymix remove almost all of that bias, to
`r fixed(cv("slope", "proxymix", "bias"), 3)` on the slope and
`r fixed(cv("intercept", "proxymix", "bias"), 3)` on the intercept, but not
all of it, and its intervals cover in
`r fixed(cv("slope", "proxymix", "coverage"), 3)` of datasets although they
are wider than the Tobit intervals,
`r fixed(cv("slope", "proxymix", "width"), 3)` against
`r fixed(cv("slope", "AER tobit", "width"), 3)`. The Tobit likelihood is
the correctly specified maximum-likelihood estimator for this design, and
the imputation route, with its
component count chosen by BIC and its parameters refitted on bootstrap
resamples, pays a small price in both bias and coverage for reaching the
same model from the imputation side.

In this run, proxymix's sweep and mice's delta adjustment did about
equally well in the value-dependent design, each at the grid value nearest
the truth: a bias of `r fixed(mv("proxymix, slope 0.7", "bias"), 3)` and
coverage of `r fixed(mv("proxymix, slope 0.7", "coverage"), 3)` for
proxymix at slope 0.7, against
`r fixed(mv("mice, shift 0.4", "bias"), 3)` and
`r fixed(mv("mice, shift 0.4", "coverage"), 3)` for mice at a shift of 0.4.
The Tobit likelihood in AER and survival did better than proxymix's
imputations in the censored design, with a slope root mean squared error of
`r fixed(cv("slope", "AER tobit", "rmse"), 3)` against
`r fixed(cv("slope", "proxymix", "rmse"), 3)` and coverage of
`r fixed(cv("slope", "AER tobit", "coverage"), 3)` against
`r fixed(cv("slope", "proxymix", "coverage"), 3)`. The comparison covers one mixture shape, one sample
size, a logistic mechanism on one coordinate, censoring at a single known
bound, and grids that both contain a value near the truth; it says nothing
about a mechanism that depends on an unobserved third variable, about
interval censoring, or about how either sweep behaves when the grid does not
reach the truth.

## Further reading

*Imputing missing data with a mixture* is the companion case this vignette
assumes away, where missingness depends only on what was observed, and it
shows how the mixture conditional differs from a single Gaussian in the
shape of the imputations.

*The closed-form operator calculus on a mixture* covers the conditioning
step that every gate multiplies into, and what else it buys.

## References

* Diggle, P. and Kenward, M. G. (1994). *Informative drop-out in
  longitudinal data analysis.* Journal of the Royal Statistical Society C
  43(1), 49--93. <doi:10.2307/2986113>.
* Fair, R. C. (1978). *A theory of extramarital affairs.* Journal of
  Political Economy 86(1), 45--61. <doi:10.1086/260646>.
* Gorman, K. B., Williams, T. D. and Fraser, W. R. (2014). *Ecological
  sexual dimorphism and environmental variability within a community of
  Antarctic penguins (genus Pygoscelis).* PLoS ONE 9(3), e90081.
  <doi:10.1371/journal.pone.0090081>.
* Heckman, J. J. (1979). *Sample selection bias as a specification error.*
  Econometrica 47(1), 153--161. <doi:10.2307/1912352>.
* Honaker, J., King, G. and Blackwell, M. (2011). *Amelia II: A program for
  missing data.* Journal of Statistical Software 45(7), 1--47.
  <doi:10.18637/jss.v045.i07>.
* Horst, A. M., Presmanes Hill, A. and Gorman, K. B. (2022). *Palmer
  Archipelago penguins data in the palmerpenguins R package -- an
  alternative to Anderson's irises.* The R Journal 14(1), 244--254.
  <doi:10.32614/RJ-2022-020>.
* Kleiber, C. and Zeileis, A. (2008). *Applied Econometrics with R.*
  Springer. <doi:10.1007/978-0-387-77318-6>.
* Little, R. J. A. (1993). *Pattern-mixture models for multivariate
  incomplete data.* Journal of the American Statistical Association
  88(421), 125--134. <doi:10.1080/01621459.1993.10594302>.
* Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys.*
  Wiley.
* Therneau, T. M. and Grambsch, P. M. (2000). *Modeling Survival Data:
  Extending the Cox Model.* Springer. <doi:10.1007/978-1-4757-3294-8>.
* Tobin, J. (1958). *Estimation of relationships for limited dependent
  variables.* Econometrica 26(1), 24--36. <doi:10.2307/1907382>.
* Toomet, O. and Henningsen, A. (2008). *Sample selection models in R:
  Package sampleSelection.* Journal of Statistical Software 27(7), 1--23.
  <doi:10.18637/jss.v027.i07>.
* van Buuren, S. (2018). *Flexible Imputation of Missing Data*, second
  edition. Chapman and Hall/CRC. <doi:10.1201/9780429492259>.
* van Buuren, S. and Groothuis-Oudshoorn, K. (2011). *mice: Multivariate
  imputation by chained equations in R.* Journal of Statistical Software
  45(3), 1--67. <doi:10.18637/jss.v045.i03>.

## Reproduce

The data-generating seed is `20260622`, set in the chunk that builds the
toy; every call to `gmm_impute()` and `proxy_mnar_sensitivity()` is given
`seed = 1L`, which makes the completions reproducible without disturbing the
ambient random-number state.

The numerical illustration deletes body mass after `set.seed(20260925)` and
gives each imputer the seed 1, through `set.seed(1L)` for Amelia; the
Affairs fits draw nothing beyond proxymix's seeded completions. In the
simulation each dataset is generated after `set.seed()` with its own index,
which is also the seed given to each imputer. The stored simulation results
come from the code shown, run on `r sim$run_date` with R
`r sim$versions[["R"]]`, proxymix `r pv("proxymix")`, mice `r pv("mice")`,
Amelia `r pv("Amelia")`, AER `r pv("AER")`, survival `r pv("survival")` and
sampleSelection `r pv("sampleSelection")`, and the run raised
`r length(sim$warnings)` warnings.

```{r session-info}
sessionInfo()
```
