---
title: "The closed-form operator calculus on a mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The closed-form operator calculus on a mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~21 s under rmarkdown::render() with ggplot2, dlm and
     KFAS installed, the simulation read from its stored results;
     macOS arm64 (Apple silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r inline-numbers, include = FALSE}
# in the PDF, keep a signed number or an exponent on one line
keep_whole <- function(x) {
  if (knitr::is_latex_output()) paste0("\\mbox{", x, "}") else x
}
inline_default <- knitr::knit_hooks$get("inline")
knitr::knit_hooks$set(inline = function(x) {
  out <- inline_default(x)
  if (is.character(out) && length(out) == 1L &&
        grepl("^-?[0-9.]+(e[-+][0-9]+)?$", out) && grepl("-", out)) {
    out <- keep_whole(out)
  }
  out
})
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The problem

A modeller has a fitted Gaussian-mixture proxy and a pipeline waiting behind
it: push the latent state through a sensor, fold in a measurement, aggregate
to a coarser grid, condition on what was observed, run the whole thing
forward in time. The question is not whether any of that is possible, but
where the exactness stops: *I have a fitted mixture, so which of the things
I want to do with it next stay exact, and which quietly stop being exact?*
This vignette draws that line, and checks each claimed exactness against an
independent hand computation rather than asserting it.

## Package capabilities

Four operator families act on a mixture and return a mixture, so they
compose without ever leaving closed form. `gmm_affine()` pushes a mixture
through a linear map with additive Gaussian noise, $y = Ax + b + \epsilon$;
`gmm_aggregate()` is the same operator specialised to a coarsening matrix.
`gmm_observe()` performs the Bayesian update on a noisy linear
observation, the finite-mixture analogue of a Kalman (1960) update: it
reweights the components by their marginal evidence and Kalman-updates each
one. `gmm_marginalise()` integrates out a coordinate subset, and
`gmm_conditionalise()` and `gmm_missing()` condition on coordinates
observed exactly, through the Schur-complement algebra of the multivariate
normal (Murphy, 2012, ch. 4).

Two further verbs turn those operators into a recursion.
`gmm_reduce()` collapses a mixture to a budget of at most `k_max`
components by a greedy, moment-preserving pairwise merge, so a filter whose
component count would otherwise grow stays bounded. `gmm_filter()`
packages the predict, update and reduce loop as a single verb: with Gaussian
noise it is the Kalman filter, and with a Gaussian-sum process or
measurement noise it is the Gaussian-sum filter of Alspach and Sorenson
(1972).

This algebraic dividend is the reason for fitting a *parametric* proxy in
the first place. A non-parametric kernel density estimate or a Monte Carlo
sample bank supports none of it in closed form. The mixture approximation
of van der Hoek and Elliott (2024) is motivated by the same use, in
non-linear Kalman-type filtering.

## Addressing the problem

```{r seed}
set.seed(20260514)
```

### A mixture to operate on

A two-component mixture over a pair of latent variables.

```{r prior}
g_prior <- gmm(
  weights = c(0.6, 0.4),
  means = list(c(-1, 0), c(1.5, 0.5)),
  covariances = list(diag(c(0.6, 0.8)), diag(c(0.7, 0.5)))
)
g_prior
```

### Pushforward through a linear sensor

Suppose the latent state is observed through
$y = A x + \epsilon$ with $\epsilon \sim \mathcal{N}(0, R)$ and a sensor
matrix that reports both coordinates and their sum.

```{r sensor}
A_sensor <- matrix(
  c(1, 0,
    0, 1,
    1, 1),
  nrow = 3L, byrow = TRUE
)
b_sensor <- c(0, 0, 0)
R_sensor <- 0.05 * diag(3)

g_pushed <- gmm_affine(
  g_prior, A_sensor, b_sensor, noise_cov = R_sensor
)
```

The weights are unchanged, the means are $A \mu_k + b$ and the covariances
are $A \Sigma_k A^\top + R$, which the hand computation below reproduces.

```{r sensor-check}
mu_hand <- lapply(g_prior@means, function(mu) {
  as.numeric(A_sensor %*% mu + b_sensor)
})
cov_hand <- lapply(g_prior@covariances, function(s) {
  A_sensor %*% s %*% t(A_sensor) + R_sensor
})
affine_gap <- max(
  abs(unlist(g_pushed@means) - unlist(mu_hand)),
  abs(unlist(g_pushed@covariances) - unlist(cov_hand)),
  abs(g_pushed@weights - g_prior@weights)
)
```

### Bayesian update on a noisy observation

Now observe `y = 0.8` on the first coordinate alone, with noise variance
0.25.

```{r observe}
A_obs <- matrix(c(1, 0), nrow = 1L)
g_post <- gmm_observe(
  g_prior, A = A_obs, y = 0.8, noise_cov = matrix(0.25, 1L, 1L)
)
g_post
```

Component weights have been multiplied by the marginal evidence
$\pi_k \, \mathcal{N}(y;\, A \mu_k,\, S_k)$ and renormalised; component
means and covariances have been Kalman-updated component-wise.

```{r fig-prior-posterior, eval = has_ggplot2, echo = has_ggplot2, fig.height = 4, fig.cap = sprintf("The prior mixture and its posterior after observing 0.8 on the first coordinate. The right-hand component, centred at $x_1 = %s$, sits closer to the observation, and its weight rises from %s to %s while the left-hand component's falls from %s to %s. Within each component the $x_1$ variance shrinks and the $x_1$ mean moves toward the observed value; the covariances are diagonal, so the $x_2$ mean and variance of each component are unchanged.", format(g_prior@means[[2L]][1L]), format(g_prior@weights[2L]), format(round(g_post@weights[2L], 3L)), format(g_prior@weights[1L]), format(round(g_post@weights[1L], 3L))), fig.alt = "Two side-by-side density rasters on the same axes; the right-hand posterior panel is visibly more concentrated than the left-hand prior panel and its mass has shifted toward the observed value."}
grid <- expand.grid(
  x = seq(-4, 5, length.out = 80L),
  y = seq(-3, 3, length.out = 60L)
)
gm <- as.matrix(grid)
long <- rbind(
  data.frame(x = grid$x, y = grid$y, d = dgmm(gm, g_prior), part = "Prior"),
  data.frame(
    x = grid$x, y = grid$y, d = dgmm(gm, g_post), part = "Posterior"
  )
)
long$part <- factor(long$part, levels = c("Prior", "Posterior"))
ggplot2::ggplot(long, ggplot2::aes(x, y)) +
  ggplot2::geom_raster(ggplot2::aes(fill = d), interpolate = TRUE) +
  ggplot2::geom_contour(
    ggplot2::aes(z = d), colour = "white", linewidth = 0.2,
    alpha = 0.6, bins = 8L
  ) +
  ggplot2::facet_wrap(~ part) +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::scale_fill_viridis_c(name = "density") +
  ggplot2::labs(
    x = expression(x[1]), y = expression(x[2]),
    title = "Prior and posterior after observing the first coordinate"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold"),
    panel.grid = ggplot2::element_blank()
  )
```

```{r fig-prior-posterior-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the prior-and-posterior",
    "figure is skipped.\n")
```

### The Kalman parity check

For a single-component prior, `gmm_observe()` should collapse exactly to a
Kalman update. The hand recursion below is written out in full so that the
comparison is against arithmetic this package did not produce.

```{r kalman-parity}
g_single <- gmm(
  weights = 1, means = list(c(0, 0)),
  covariances = list(diag(c(1, 2)))
)
g_one_obs <- gmm_observe(
  g_single, A = matrix(c(1, 0), nrow = 1L), y = 0.5,
  noise_cov = matrix(0.5, 1L, 1L)
)

s_prior <- diag(c(1, 2))
h_obs <- matrix(c(1, 0), nrow = 1L)
r_obs <- matrix(0.5, 1L, 1L)
s_innov <- h_obs %*% s_prior %*% t(h_obs) + r_obs
gain <- s_prior %*% t(h_obs) %*% solve(s_innov)
mu_kalman <- as.numeric(gain * 0.5)
cov_kalman <- s_prior - gain %*% h_obs %*% s_prior

kalman_gap <- max(
  abs(mu_kalman - g_one_obs@means[[1L]]),
  abs(cov_kalman - g_one_obs@covariances[[1L]])
)
ridge_default <- 1e-6
```

### Aggregation through a coarsening matrix

Aggregating a three-coordinate latent into two summaries -- the sum of the
first two coordinates, and the third on its own -- is `gmm_affine()` with a
row-sum matrix, exposed as `gmm_aggregate()`.

```{r aggregate}
g_fine <- gmm(
  weights = c(0.3, 0.4, 0.3),
  means = list(c(0, 0, 0), c(2, 1, -1), c(-1, -1, 2)),
  covariances = list(diag(3), diag(3), diag(3))
)
A_agg <- matrix(
  c(1, 1, 0,
    0, 0, 1),
  nrow = 2L, byrow = TRUE
)
g_coarse <- gmm_aggregate(g_fine, A_agg)
weights_kept <- max(abs(g_coarse@weights - g_fine@weights))
```

```{r aggregate-kable, echo = FALSE}
knitr::kable(
  data.frame(
    component = seq_len(gmm_n_components(g_coarse)),
    weight = round(g_coarse@weights, 3L),
    mean_1 = round(vapply(g_coarse@means, function(m) m[1L], numeric(1L)), 3L),
    mean_2 = round(vapply(g_coarse@means, function(m) m[2L], numeric(1L)), 3L)
  ),
  col.names = c("Component", "Weight", "Mean of $x_1 + x_2$", "Mean of $x_3$"),
  caption = paste(
    "The aggregated mixture keeps the component count and the weights of",
    "the fine mixture; the table shows the component means mapped through",
    "the coarsening matrix."
  )
)
```

### Conditioning on coordinates observed exactly

If a subset of coordinates is observed with no noise at all, `gmm_missing()`
routes through the Schur-complement path. It is the same operation as
`gmm_conditionalise()` with an index-based signature, and the two agree.

```{r conditioning}
g_cond_index <- gmm_missing(g_prior, observed = 2L, values = 0.5)
g_cond_given <- gmm_conditionalise(g_prior, given = c(NA, 0.5))
cond_gap <- max(
  abs(unlist(g_cond_index@means) - unlist(g_cond_given@means)),
  abs(unlist(g_cond_index@covariances) -
        unlist(g_cond_given@covariances)),
  abs(g_cond_index@weights - g_cond_given@weights)
)
```

### Sequential observations compose

Two observations on disjoint coordinates, applied one after the other,
should equal one observation on the stacked coordinates.

```{r compose}
g_a <- gmm_observe(
  g_prior, A = matrix(c(1, 0), nrow = 1L), y = 0.5,
  noise_cov = matrix(0.25, 1L, 1L)
)
g_ab <- gmm_observe(
  g_a, A = matrix(c(0, 1), nrow = 1L), y = 0.2,
  noise_cov = matrix(0.25, 1L, 1L)
)
g_stack <- gmm_observe(
  g_prior, A = diag(2), y = c(0.5, 0.2), noise_cov = 0.25 * diag(2)
)
compose_gap <- max(
  abs(g_ab@weights - g_stack@weights),
  abs(unlist(g_ab@means) - unlist(g_stack@means)),
  abs(unlist(g_ab@covariances) - unlist(g_stack@covariances))
)
```

### Filtering over time, and the Kalman filter as a special case

The update above is a single step. A *filter* alternates two steps over
time: a **predict** step that pushes the state forward through the dynamics
(`gmm_affine()`, the Chapman-Kolmogorov propagation of a linear stochastic
dynamical system), and an **update** step that folds in each noisy
measurement (`gmm_observe()`). Run on a single-component prior, this
recursion *is* the classical discrete-time Kalman filter; run on a
multi-component prior, it is the **Gaussian-sum filter**, a bank of Kalman
filters carried in parallel.

Take a one-dimensional constant-velocity track, with state
(position, velocity), linear dynamics $A$, process noise $Q$, and a noisy
position sensor $C = [1, 0]$ with noise $R$.

```{r track}
dt <- 1
A_dyn <- matrix(c(1, dt, 0, 1), 2L, 2L, byrow = TRUE)
C_obs <- matrix(c(1, 0), 1L, 2L)
Q_proc <- 0.01 * diag(2)
R_meas <- matrix(0.5, 1L, 1L)

n_steps <- 30L
truth <- matrix(0, n_steps, 2L)
truth[1L, ] <- c(0, 1)
for (k1 in 2:n_steps) {
  truth[k1, ] <- as.numeric(A_dyn %*% truth[k1 - 1L, ]) +
    mvnfast::rmvn(1L, c(0, 0), Q_proc)
} # ends k1, over the simulated state track
y_track <- truth[, 1L] + rnorm(n_steps, 0, sqrt(R_meas[1L, 1L]))
```

The filter is the operator calculus in a loop.

```{r filter-loop}
g_state <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(diag(2))
)
filtered <- numeric(n_steps)
for (k1 in seq_len(n_steps)) {
  if (k1 > 1L) {
    g_state <- gmm_affine(
      g_state, A = A_dyn, b = c(0, 0), noise_cov = Q_proc
    )                                                       # predict
  }
  g_state <- gmm_observe(
    g_state, A = C_obs, y = y_track[k1], noise_cov = R_meas
  )                                                         # update
  filtered[k1] <- g_state@means[[1L]][1L]
} # ends k1, over the filtering recursion
```

Against a hand-coded textbook Kalman filter:

```{r filter-parity}
mu_kf <- c(0, 0)
p_kf <- diag(2)
kf_track <- numeric(n_steps)
for (k1 in seq_len(n_steps)) {
  if (k1 > 1L) {
    mu_kf <- as.numeric(A_dyn %*% mu_kf)
    p_kf <- A_dyn %*% p_kf %*% t(A_dyn) + Q_proc            # predict
  }
  gain_kf <- p_kf %*% t(C_obs) %*%
    solve(C_obs %*% p_kf %*% t(C_obs) + R_meas)             # gain
  mu_kf <- mu_kf + as.numeric(gain_kf %*% (y_track[k1] - C_obs %*% mu_kf))
  p_kf <- (diag(2) - gain_kf %*% C_obs) %*% p_kf
  kf_track[k1] <- mu_kf[1L]
} # ends k1, over the hand-coded Kalman recursion
loop_gap <- max(abs(filtered - kf_track))
```

`gmm_affine()` and `gmm_observe()` add a ridge of $10^{-6}$ to the diagonal
of every covariance they return, to keep it positive-definite. The same loop
with `ridge_eps = 0` in both calls leaves the ridge out.

```{r filter-loop-noridge}
g_state <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(diag(2))
)
filtered_noridge <- numeric(n_steps)
for (k1 in seq_len(n_steps)) {
  if (k1 > 1L) {
    g_state <- gmm_affine(
      g_state, A = A_dyn, b = c(0, 0), noise_cov = Q_proc, ridge_eps = 0
    )
  }
  g_state <- gmm_observe(
    g_state, A = C_obs, y = y_track[k1], noise_cov = R_meas, ridge_eps = 0
  )
  filtered_noridge[k1] <- g_state@means[[1L]][1L]
} # ends k1, over the filtering recursion without the ridge
loop_gap_noridge <- max(abs(filtered_noridge - kf_track))
```

```{r fig-kalman, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.6, fig.cap = "A constant-velocity track filtered by the single-component operator-calculus recursion, which is the Kalman filter: the noisy position readings, the latent truth, and the filtered estimate tracking it.", fig.alt = "A time series with scattered grey noisy position readings, a smooth latent-truth line, and a filtered-estimate line closely following the truth."}
track_df <- data.frame(
  t = seq_len(n_steps), truth = truth[, 1L], y = y_track,
  filtered = filtered
)
ggplot2::ggplot(track_df, ggplot2::aes(t)) +
  ggplot2::geom_point(
    ggplot2::aes(y = y, colour = "noisy reading"), size = 1.3, alpha = 0.7
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = truth, colour = "latent truth"), linewidth = 0.8
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered, colour = "filtered, one component"),
    linewidth = 0.9
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c(
      "noisy reading" = "grey60",
      "latent truth" = "#0072B2",
      "filtered, one component" = "#D55E00"
    )
  ) +
  ggplot2::labs(
    x = "time step", y = "position",
    title = "Predict and update over time is a Kalman filter"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-kalman-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the Kalman-track figure",
    "is skipped.\n")
```

At more than one component the same two calls run a Kalman filter inside
*every* component and reweight the components by their evidence, which is
the Gaussian-sum filter. It represents beliefs a single Kalman filter cannot
-- multimodal posteriors, several competing hypotheses about where the
target is. The cost is component growth under noise that is itself a
Gaussian mixture: a mixture process noise multiplies the component count in
the predict step, where `gmm_affine()` runs once per noise component, and a
mixture measurement noise multiplies it in the update step, where
`gmm_observe()` runs once per noise component. Collapsing the mixture back
to a fixed budget keeps the filter bounded.

### Bounding the filter through mixture reduction

`gmm_reduce()` collapses a mixture to at most `k_max` components by a
greedy, moment-preserving pairwise merge: at each step the cheapest pair is
replaced by the single Gaussian that preserves their combined weight, mean
and covariance. Two merge costs decide which pair goes first, the
Kullback-Leibler bound of Runnalls (2007) under `cost = "kl"` and a
closed-form Cauchy-Schwarz cost under `cost = "cs"`, built from the same
Gaussian-product identity as `gmm_divergence()`.

Take a six-component mixture that really has three clusters, each split into
two near-identical components, and reduce it to three.

```{r reduce}
g_six <- gmm(
  weights = rep(1 / 6, 6L),
  means = list(
    c(-5, 0), c(-5, 0.15), c(5, 0), c(5.1, -0.1), c(0, 6), c(0.1, 6.1)
  ),
  covariances = rep(list(0.5 * diag(2)), 6L)
)
g_three <- gmm_reduce(g_six, k_max = 3L)

mix_mean <- function(g) Reduce(`+`, Map(`*`, g@weights, g@means))
reduce_shift <- max(abs(mix_mean(g_six) - mix_mean(g_three)))
reduce_divergence <- gmm_divergence(g_six, g_three, type = "cs")
```

```{r number-words, include = FALSE}
n_word <- function(k) {
  c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine",
    "ten")[k]
}
```

```{r reduce-kable, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c(
      "components before", "components after",
      "shift in the mixture mean", "Cauchy-Schwarz divergence"
    ),
    value = c(
      formatC(gmm_n_components(g_six), format = "d"),
      formatC(gmm_n_components(g_three), format = "d"),
      formatC(reduce_shift, format = "e", digits = 1L),
      formatC(reduce_divergence, format = "e", digits = 1L)
    )
  ),
  col.names = c("Quantity", "Value"),
  caption = paste(
    "Reducing six components to three by moment-preserving merges,",
    "and what the reduction costs."
  )
)
```

Reducing all the way to a single component returns the moment-matched
Gaussian, so `gmm_reduce()` interpolates between the full mixture and its
single-Gaussian summary. For a smooth, over-parameterised mixture a globally
fitted proxy can beat any sequence of pairwise merges, and `method =
"anneal"` refits a budget-sized proxy by annealed expectation-maximisation
and keeps it when it improves on the merge.

### The bounded filter in one call

The predict, update and reduce loop is packaged as a single verb.
`gmm_filter()` takes the prior belief, the linear-Gaussian `dynamics` and
`measurement`, the observation series, and an optional component cap
`k_max`. With Gaussian noise and `k_max = NULL` it is the Kalman filter;
with a Gaussian-sum process or measurement noise -- a `gmm` in place of a
covariance matrix -- it is the Gaussian-sum filter, bounded by
`gmm_reduce()` after each step.

```{r verb-kalman}
prior_state <- gmm(
  weights = 1, means = list(c(0, 0)), covariances = list(diag(2))
)
out_verb <- gmm_filter(
  prior_state,
  dynamics = list(A = A_dyn, Q = Q_proc),
  measurement = list(C = C_obs, R = R_meas),
  y = y_track, ridge_eps = 0
)

mu_v <- c(0, 0)
p_v <- diag(2)
kf_verb <- numeric(n_steps)
for (k1 in seq_len(n_steps)) {
  mu_v <- as.numeric(A_dyn %*% mu_v)
  p_v <- A_dyn %*% p_v %*% t(A_dyn) + Q_proc                # predict
  gain_v <- p_v %*% t(C_obs) %*%
    solve(C_obs %*% p_v %*% t(C_obs) + R_meas)              # gain
  mu_v <- mu_v + as.numeric(gain_v %*% (y_track[k1] - C_obs %*% mu_v))
  p_v <- (diag(2) - gain_v %*% C_obs) %*% p_v               # update
  kf_verb[k1] <- mu_v[1L]
} # ends k1, over the predict-then-update reference recursion
verb_gap <- max(abs(out_verb$mean[, 1L] - kf_verb))
```

A heavy-tailed disturbance is awkward for a single Gaussian process noise.
Modelled as a two-component Gaussian sum -- a calm regime and an occasional
larger jump -- the recursion becomes a genuine Gaussian-sum filter. Each
step multiplies the component count by two, so the cap returns it to budget.
The track above was simulated with Gaussian process noise, so this
heavy-tailed noise model is misspecified for it.

```{r verb-gsf}
q_heavy <- gmm(
  weights = c(0.9, 0.1),
  means = list(c(0, 0), c(0, 0)),
  covariances = list(0.01 * diag(2), 0.5 * diag(2))
)
out_gsf <- gmm_filter(
  prior_state,
  dynamics = list(A = A_dyn, Q = q_heavy),
  measurement = list(C = C_obs, R = R_meas),
  y = y_track, k_max = 6L
)
gsf_max_k <- max(out_gsf$summary$n_components)
gsf_rmse <- sqrt(mean((out_gsf$mean[, 1L] - truth[, 1L])^2))
kalman_rmse <- sqrt(mean((out_verb$mean[, 1L] - truth[, 1L])^2))
```

```{r gsf-counts, include = FALSE}
gsf_k <- out_gsf$summary$n_components
gsf_first_cap <- which(gsf_k == gsf_max_k)[1L]
gsf_early_vec <- paste0(gsf_k[seq_len(gsf_first_cap - 1L)], " after step ",
                        seq_len(gsf_first_cap - 1L))
gsf_early <- paste(gsf_early_vec, collapse = ", ")
gsf_counts <- paste0(gsf_early, " and ", gsf_max_k, ", the cap, from step ",
                     gsf_first_cap, " to step ", n_steps)
```

```{r fig-gsf, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.6, fig.cap = sprintf("The same track filtered by the Gaussian-sum filter under a two-component heavy-tailed process noise, capped at six components per step. The filtered mean tracks the latent truth; the component count is %s.", gsf_counts), fig.alt = "A time series with grey noisy position readings, a latent-truth line, and a Gaussian-sum-filter estimate line following the truth closely."}
gsf_df <- data.frame(
  t = seq_len(n_steps), truth = truth[, 1L], y = y_track,
  filtered = out_gsf$mean[, 1L]
)
ggplot2::ggplot(gsf_df, ggplot2::aes(t)) +
  ggplot2::geom_point(
    ggplot2::aes(y = y, colour = "noisy reading"), size = 1.3, alpha = 0.7
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = truth, colour = "latent truth"), linewidth = 0.8
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered, colour = "Gaussian-sum filter"),
    linewidth = 0.9
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c(
      "noisy reading" = "grey60",
      "latent truth" = "#0072B2",
      "Gaussian-sum filter" = "#D55E00"
    )
  ) +
  ggplot2::labs(
    x = "time step", y = "position",
    title = "A bounded Gaussian-sum filter in one call"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-gsf-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the Gaussian-sum-filter",
    "figure is skipped.\n")
```

```{r parity-kable, echo = FALSE}
knitr::kable(
  data.frame(
    check = c(
      "pushforward against the hand formula",
      "single-component update against a hand Kalman step",
      "index conditioning against value conditioning",
      "two observations against one stacked observation",
      "operator loop against a hand Kalman filter",
      "operator loop, `ridge_eps = 0`, against a hand Kalman filter",
      "`gmm_filter()`, `ridge_eps = 0`, against a hand Kalman filter",
      "weights preserved by aggregation"
    ),
    gap = formatC(
      c(
        affine_gap, kalman_gap, cond_gap, compose_gap, loop_gap,
        loop_gap_noridge, verb_gap, weights_kept
      ),
      format = "e", digits = 1L
    )
  ),
  col.names = c("Exactness check", "Largest absolute difference"),
  caption = paste(
    "The exactness checks of this section, each against an independently",
    "written hand computation; the Nile comparisons follow in the numerical",
    "illustration. `gmm_affine()` and `gmm_observe()` add a",
    "ridge of $10^{-6}$ to each covariance unless `ridge_eps = 0`."
  )
)
```

## Interpretation

Every operator claimed to be exact matches its hand computation, up to the
ridge of $10^{-`r round(-log10(ridge_default))`}$ that
`gmm_affine()` and `gmm_observe()` add to each covariance by default. The
pushforward reproduces $A \mu_k + b$ and $A \Sigma_k A^\top + R$ to
`r formatC(affine_gap, format = "e", digits = 1L)`, the size of that ridge,
and it leaves the weights untouched, as aggregation does to
`r formatC(weights_kept, format = "e", digits = 1L)`: a linear map moves
where the components sit, never how much mass each carries.

The single-component update matches a hand-written Kalman step to
`r formatC(kalman_gap, format = "e", digits = 1L)`, again the size of the
ridge. The two conditioning signatures agree to
`r formatC(cond_gap, format = "e", digits = 1L)`, so `gmm_missing()` and
`gmm_conditionalise()` perform the same computation. Two sequential
observations on disjoint coordinates agree with one stacked observation to
`r formatC(compose_gap, format = "e", digits = 1L)`, the size of the one
extra ridge the second call adds. Folding the two measurements in one at a
time gives the same belief as folding them in together.

Over time the ridge accumulates. The explicit predict-and-update loop makes
`r 2L * n_steps - 1L` operator calls over `r n_steps` steps, each adding the
ridge, and agrees with a hand-coded Kalman filter to
`r formatC(loop_gap, format = "e", digits = 1L)`. With `ridge_eps = 0` the
same loop agrees to `r formatC(loop_gap_noridge, format = "e", digits = 1L)`,
and the packaged `gmm_filter()` verb, called with `ridge_eps = 0`, agrees
with the same reference to
`r formatC(verb_gap, format = "e", digits = 1L)`. The first figure shows the
update contracting the belief toward the observation, and the second shows
the recursion tracking a latent state through noise.

Reduction is the one operator on this page that is designed to lose
something, and the table says how much. Collapsing `r n_word(gmm_n_components(g_six))` components to
`r n_word(gmm_n_components(g_three))` moves the mixture mean by
`r formatC(reduce_shift, format = "e", digits = 1L)`, because every merge
preserves the combined weight, mean and covariance by construction; the
Cauchy-Schwarz divergence between the six-component mixture and its
three-component reduction is
`r formatC(reduce_divergence, format = "e", digits = 1L)`, small here only
because the components merged were near-duplicates. On a mixture whose
components genuinely differ, that divergence is where the cost of the budget
shows up.

Under a two-component heavy-tailed process noise the component count
doubles at every step until it reaches the cap: from a one-component prior
it is `r gsf_counts`. The filtered mean tracks the latent truth with a
root-mean-square error of `r round(gsf_rmse, 3L)` against
`r round(kalman_rmse, 3L)` for the single-Gaussian Kalman filter on the
same readings. That is one realisation of `r n_steps` steps under a
misspecified noise model, so it does not rank the two filters.

So the line the opening question asked for runs as follows. Pushforward,
aggregation, noisy linear update and exact conditioning each matched its
hand computation, and any sequence of them stays in closed form, because
each returns a mixture and the next operator does not know or care how the
mixture arose. Reduction is
exact in its moments and approximate in its shape. Everything outside the
affine-Gaussian channel leaves closed form altogether, which is the subject
of the next section.

## Limitations

The closed-form algebra rests on two assumptions, and both are load-bearing.
The map must be linear and the noise must be Gaussian. A non-linear sensor
-- a sigmoid observation, a max-pooling aggregator, anything that bends the
coordinate space -- has no closed-form pushforward through a Gaussian
mixture, and must not be silently linearised; a Monte Carlo pushforward is
the correct fallback, and the operators here will not stand in for one. The
same restriction applies to non-Gaussian observation noise, with one
exception: noise that is itself a finite Gaussian mixture stays inside the
calculus, and the numerical illustration that follows tests the capped
filter under such a noise against a particle filter, beside two Kalman
filters and a Student-t filter. Hierarchical and
random-effect models sit outside the calculus for a third reason: the
operators apply to a *fixed* affine-Gaussian channel, and a prior over the
channel parameters $(A, R)$ themselves requires an integration that is not
closed form in general.

Component growth is the practical limit on the filter. Under Gaussian-sum
process or measurement noise each predict or update step with mixture noise
multiplies the component count by the number of noise components, so an uncapped Gaussian-sum filter over a
long series is not a viable object. `gmm_reduce()` is what makes it viable,
and the budget is a modelling choice with consequences: a cap tight enough
to be cheap will merge components that represent genuinely distinct
hypotheses, and the multimodality the Gaussian-sum filter exists to carry is
exactly what a tight cap discards. The divergence reported above is small
because the example merges near-duplicates by design, and it should be
recomputed on any real filtering problem rather than assumed small.

A Gaussian-process latent is the natural comparison and wins on flexibility.
It gives a fully non-parametric, uncertainty-rich representation of the
underlying field, and integration through a linear operator with additive
Gaussian noise is closed form for a Gaussian process too, so on a single
pass there is little to choose between them. What the Gaussian process
lacks is cheap *composition*: repeatedly conditioning, aggregating and
re-mixing its posteriors across configurations carries a growing
kernel-algebra burden, where the mixture calculus stays a finite parameter
list of fixed size. The finite mixture is the less flexible representation
and the more composable one, so the choice turns on how much closed-form
composition the downstream pipeline actually needs.

Finally, everything on this page operates on a mixture that is taken as
given. None of these checks says the mixture is a good proxy for whatever it
was fitted to, and an exact operator applied to a poor proxy propagates the
proxy's error exactly.

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_dlm <- requireNamespace("dlm", quietly = TRUE)
has_kfas <- requireNamespace("KFAS", quietly = TRUE)
run_illustration <- has_dlm && has_kfas
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("dlm or KFAS is not available on this build, so the Nile illustration",
    "is skipped.\n")
```

```{r fixed-notation, include = FALSE}
## Small numbers are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
sci <- function(v) formatC(v, format = "e", digits = 1L)
```

### The Nile at one component, beside two Kalman filters

`gmm_filter()`, dlm (Petris, 2010) and KFAS (Helske, 2017) are applied to
the annual flow of the Nile at Aswan from 1871 to 1970 (Cobb, 1978; Durbin and
Koopman, 2012), shipped with R as `Nile`. The model is the local level: the
level follows a random walk with variance $Q$ and each year's flow is the
level plus Gaussian noise of variance $H$. The two variances are estimated
once by maximum likelihood through `dlm::dlmMLE()` and handed to all three
filters, together with one initial belief, a level of 1000 with variance
$10^5$. KFAS places its initial belief on the level of the first year rather
than on the level before it, so it receives the initial variance plus $Q$.

The competitors install from CRAN once and load as usual:

```{r install-competitors, eval = FALSE}
install.packages(c("dlm", "KFAS", "nimble", "nimbleSMC"))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(dlm)
library(KFAS)
```

```{r nile-fit, eval = run_illustration}
y_nile <- as.numeric(datasets::Nile)
n_nile <- length(y_nile)
build_level <- function(p) dlmModPoly(1L, dV = exp(p[1L]), dW = exp(p[2L]))
fit_nile <- dlmMLE(y_nile, parm = c(log(15000), log(1500)),
                   build = build_level)
h_nile <- exp(fit_nile$par[1L])   # observation noise variance
q_nile <- exp(fit_nile$par[2L])   # variance of the level increments
m0_nile <- 1000
c0_nile <- 1e5
```

```{r nile-three, eval = run_illustration}
mod_dlm <- dlmModPoly(1L, dV = h_nile, dW = q_nile, m0 = m0_nile,
                      C0 = c0_nile)
f_dlm <- dlmFilter(y_nile, mod_dlm)
mean_dlm <- as.numeric(f_dlm$m[-1L])
var_dlm <- unlist(dlmSvd2var(f_dlm$U.C, f_dlm$D.C))[-1L]
ll_dlm <- -dlmLL(y_nile, mod_dlm) - n_nile / 2 * log(2 * pi)

mod_kfas <- SSModel(y_nile ~ SSMtrend(1L, Q = q_nile, a1 = m0_nile,
                                      P1 = c0_nile + q_nile), H = h_nile)
f_kfas <- KFS(mod_kfas, filtering = "state", smoothing = "none")
mean_kfas <- as.numeric(f_kfas$att)
var_kfas <- as.numeric(f_kfas$Ptt)
ll_kfas <- as.numeric(logLik(mod_kfas))

prior_nile <- gmm(weights = 1, means = list(m0_nile),
                  covariances = list(matrix(c0_nile)))
f_pm <- gmm_filter(
  prior_nile,
  dynamics = list(A = matrix(1), Q = matrix(q_nile)),
  measurement = list(C = matrix(1), R = matrix(h_nile)),
  y = y_nile, ridge_eps = 0
)
mean_pm <- f_pm$mean[, 1L]
var_pm <- f_pm$summary$sd_1^2
ll_pm <- sum(f_pm$summary$log_evidence)
```

`dlmLL()` returns the negative log-likelihood without the
$\tfrac{n}{2} \log 2\pi$ term, which the code above adds back so that the
three log-likelihoods are on one scale.

```{r nile-parity-table, eval = run_illustration, echo = FALSE}
gap <- function(a, b) max(abs(a - b))
parity_tbl <- data.frame(
  pair = c("proxymix against dlm", "proxymix against KFAS",
           "dlm against KFAS"),
  mean_gap = sci(c(gap(mean_pm, mean_dlm), gap(mean_pm, mean_kfas),
                   gap(mean_dlm, mean_kfas))),
  var_gap = sci(c(gap(var_pm, var_dlm), gap(var_pm, var_kfas),
                  gap(var_dlm, var_kfas))),
  ll_gap = sci(c(abs(ll_pm - ll_dlm), abs(ll_pm - ll_kfas),
                 abs(ll_dlm - ll_kfas))),
  stringsAsFactors = FALSE
)
knitr::kable(
  parity_tbl, align = c("l", "r", "r", "r"),
  col.names = c("Pair of filters", "Filtered means", "Filtered variances",
                "Log-likelihood"),
  caption = paste0(
    "Largest absolute difference over the ", n_nile, " filtered means and ",
    "filtered variances of the Nile local-level model, and the absolute ",
    "difference in log-likelihood, between each pair of filters given the ",
    "same two variances and the same initial belief."
  )
)
var_gap_pm <- max(gap(var_pm, var_dlm), gap(var_pm, var_kfas))
rel_var_gap <- max(pmax(abs(var_pm - var_dlm), abs(var_pm - var_kfas)) /
                     var_pm)
ll_three <- fixed(c(ll_pm, ll_dlm, ll_kfas), 2)
```

The estimated variances are $H$ = `r fixed(h_nile, 0)` and $Q$ =
`r fixed(q_nile, 0)`. At one component `gmm_filter()` returns the Kalman
filter: its filtered means agree with dlm and KFAS to
`r sci(max(gap(mean_pm, mean_dlm), gap(mean_pm, mean_kfas)))` and its
filtered variances, which settle at `r round(var_pm[n_nile])`, to
`r sci(var_gap_pm)`, a relative difference of at most
`r sci(rel_var_gap)`. The three log-likelihoods are
`r if (length(unique(ll_three)) == 1L) paste("all", keep_whole(ll_three[1L])) else paste(keep_whole(ll_three), collapse = ", ")`.

### The same series with an outlier component in the noise

The flow falls by about a quarter after 1898 and never recovers, and a
Gaussian filter treats the first low reading like any other. The second run
keeps $Q$ and the noise variance $H$ but splits the noise into two
components: a core of weight 0.9 and variance $H / 1.9$, and an outlier
component of weight 0.1 with ten times that variance, so the total variance
is unchanged and only the shape of the noise differs. The exact
Gaussian-sum filter would carry $2^{100}$ components by the last year, so
the run is capped at four components per step. The reference is a bootstrap
particle filter (Gordon et al., 1993) under the same
two-component noise, written out in full below and run with 10 000
particles.

```{r nile-mixture, eval = run_illustration}
r_mix_nile <- gmm(
  weights = c(0.9, 0.1), means = list(0, 0),
  covariances = list(matrix(h_nile / 1.9), matrix(10 * h_nile / 1.9))
)
f_mix <- gmm_filter(
  prior_nile,
  dynamics = list(A = matrix(1), Q = matrix(q_nile)),
  measurement = list(C = matrix(1), R = r_mix_nile),
  y = y_nile, k_max = 4L
)
mean_mix <- f_mix$mean[, 1L]
ll_mix <- sum(f_mix$summary$log_evidence)
```

```{r nile-pf, eval = run_illustration}
bootstrap_filter <- function(y, n_particles, m0, c0, q, r_sd, r_w) {
  n <- length(y)
  particles <- rnorm(n_particles, m0, sqrt(c0))
  filtered <- numeric(n)
  log_lik <- 0
  for (t in seq_len(n)) {
    particles <- particles + rnorm(n_particles, 0, sqrt(q))
    lik <- r_w[1L] * dnorm(y[t], particles, r_sd[1L]) +
      r_w[2L] * dnorm(y[t], particles, r_sd[2L])
    log_lik <- log_lik + log(mean(lik))
    filtered[t] <- sum(lik * particles) / sum(lik)
    particles <- particles[sample.int(n_particles, n_particles,
                                      replace = TRUE, prob = lik)]
  }
  list(mean = filtered, log_lik = log_lik)
}

set.seed(20260925)
pf_nile <- bootstrap_filter(
  y_nile, 10000L, m0_nile, c0_nile, q_nile,
  r_sd = sqrt(c(h_nile / 1.9, 10 * h_nile / 1.9)), r_w = c(0.9, 0.1)
)
```

```{r fig-nile, eval = run_illustration && has_ggplot2, echo = run_illustration && has_ggplot2, fig.height = 3.8, fig.cap = "The annual flow of the Nile with the filtered level under Gaussian noise, which is the Kalman filter, and under the two-component noise of the same variance, which is the capped Gaussian-sum filter. The particle filter under the two-component noise is drawn dotted and lies on the Gaussian-sum line.", fig.alt = "A time series of 100 annual Nile flow readings as grey points, with three filtered-level lines: the Gaussian filter, the four-component mixture filter, and a dotted particle-filter line that coincides with the mixture filter."}
nile_df <- data.frame(
  year = as.numeric(time(datasets::Nile)), flow = y_nile,
  gaussian = mean_pm, mixture = mean_mix, particle = pf_nile$mean
)
ggplot2::ggplot(nile_df, ggplot2::aes(year)) +
  ggplot2::geom_point(ggplot2::aes(y = flow, colour = "annual flow"),
                      size = 1.3, alpha = 0.7) +
  ggplot2::geom_line(ggplot2::aes(y = gaussian, colour = "Gaussian noise"),
                     linewidth = 0.8) +
  ggplot2::geom_line(
    ggplot2::aes(y = mixture, colour = "two-component noise"),
    linewidth = 0.9
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = particle, colour = "particle filter"),
    linewidth = 0.7, linetype = "dotted"
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("annual flow" = "grey60", "Gaussian noise" = "#0072B2",
               "two-component noise" = "#D55E00",
               "particle filter" = "#000000")
  ) +
  ggplot2::labs(x = "year", y = expression("flow, " * 10^8 * " m"^3),
                title = "The Nile under Gaussian and two-component noise") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-nile-skip, eval = run_illustration && !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the Nile figure is",
    "skipped.\n")
```

```{r nile-mixture-table, eval = run_illustration, echo = FALSE}
mix_tbl <- data.frame(
  filter = c("Gaussian noise (dlm, KFAS, proxymix at one component)",
             "two-component noise, proxymix capped at four components",
             "two-component noise, particle filter, 10 000 particles"),
  ll = fixed(c(ll_pm, ll_mix, pf_nile$log_lik), 2),
  max_gap = c(fixed(c(gap(mean_pm, pf_nile$mean),
                      gap(mean_mix, pf_nile$mean)), 1), "--"),
  mean_gap = c(fixed(c(mean(abs(mean_pm - pf_nile$mean)),
                       mean(abs(mean_mix - pf_nile$mean))), 1), "--"),
  stringsAsFactors = FALSE
)
knitr::kable(
  mix_tbl, align = c("l", "r", "r", "r"),
  col.names = c("Filter", "Log-likelihood", "Largest gap", "Mean gap"),
  caption = paste0(
    "Log-likelihood of the ", n_nile, " Nile readings under each filter, ",
    "and the largest and mean absolute gap between its filtered level and ",
    "the particle filter's, the reference, over the ", n_nile, " years. ",
    "The particle filter has no gap cells because it is the reference."
  )
)
i_1899 <- which(time(datasets::Nile) == 1899)
step_gauss <- mean_pm[i_1899] - mean_pm[i_1899 - 1L]
step_mix <- mean_mix[i_1899] - mean_mix[i_1899 - 1L]
```

```{r nile-read, eval = run_illustration, include = FALSE}
# differences of the log-likelihoods as the table rounds them
ll_mix_over_gauss <- round(ll_mix, 2L) - round(ll_pm, 2L)
mix_pf_ll_gap <- round(ll_mix, 2L) - round(pf_nile$log_lik, 2L)
```

The two-component noise lowers the log-likelihood of the series by
`r fixed(-ll_mix_over_gauss, 2)` against the Gaussian noise of the same
variance, so the Nile readings do not favour the split, and the capped
filter's log-likelihood is within `r fixed(abs(mix_pf_ll_gap), 2)` of the
particle filter's estimate under the same noise. Its filtered level stays
within
`r fixed(gap(mean_mix, pf_nile$mean), 1)` of the particle filter's over the
`r n_nile` years, on readings that range from `r min(y_nile)` to
`r max(y_nile)`, while the Gaussian filter's level differs from it by up to
`r fixed(gap(mean_pm, pf_nile$mean), 1)`. At 1899, the first reading after
the drop, the Gaussian filter lowers its level by `r fixed(abs(step_gauss), 1)` and
the mixture filter by `r fixed(abs(step_mix), 1)`, because the mixture
filter gives part of that reading to the outlier component. The series has
no true level to score against, so the simulation below supplies one.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("operator_calculus_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
```

The simulation repeats the exercise on `r sim$n_rep` series of
`r sim$n_t` steps whose level is known. The level is a random walk with
increment variance `r sim$q_state`, started from a belief of mean 0 and
variance `r sim$c0`, and each reading is the level plus noise drawn from a
core component of standard deviation `r sim$r_sd[1L]` with probability
`r sim$r_w[1L]`, or from an outlier component of standard deviation
`r sim$r_sd[2L]` with probability `r sim$r_w[2L]`. Six filters are given the
true increment variance and starting belief. proxymix runs under the
true two-component noise, capped at
four components per step. dlm and KFAS run the Kalman filter under Gaussian
noise of the mixture's variance, `r sim$r_var`. The bootstrap particle
filter of the previous section runs under the two-component noise with
10 000 particles and is the reference. nimble (de Valpine et al., 2017)
runs its bootstrap filter through nimbleSMC (Michaud et al., 2021) with
10 000 particles, once
under the two-component noise and once under a Student-t noise with 4
degrees of freedom whose scale, `r fixed(sim$t_scale, 3)`, is the maximum
likelihood fit to the two-component noise; the second is the robust
competitor. nimble's filter is set to resample at every step, as the
hand-written one does, so that its stored samples are the filtered
distribution at every step.

Each filter is scored by the root mean squared error of its filtered mean
against the true level over the `r sim$n_t` steps, by its log-likelihood of
the series, and by its running time.

This chunk is complete and runs as shown, but it took about
`r round(sim$elapsed_secs / 60)` minutes on one core (R
`r sim$versions[["R"]]`, Apple silicon); the results below are
read from its stored output. For a quick reduced run, change the line
`n_rep <- 300L` to `n_rep <- 3L`. The per-series times in the table below
add up to about `r round(sum(sim$sim_tab$secs) * sim$n_rep / 60)` minutes;
the rest is compiling nimble's two models and time the process spent
waiting rather than computing.

```{r sim-code, eval = FALSE}
library(proxymix)
library(dlm)
library(KFAS)
library(nimble)
library(nimbleSMC)

n_t <- 200L           # steps per series
n_particles <- 10000L
n_rep <- 300L         # series

q_state <- 0.25       # variance of the level increments
r_sd <- c(1, 5)       # noise standard deviations, core and outlier
r_w <- c(0.9, 0.1)    # their weights
r_var <- sum(r_w * r_sd^2)
m0 <- 0
c0 <- 10

# scale of a Student-t noise with 4 degrees of freedom, fitted by maximum
# likelihood to the two-component noise
t_df <- 4
neg_ll_t <- function(s) {
  log(s) - integrate(function(v) {
    (r_w[1L] * dnorm(v, 0, r_sd[1L]) + r_w[2L] * dnorm(v, 0, r_sd[2L])) *
      dt(v / s, t_df, log = TRUE)
  }, -60, 60)$value
}
t_scale <- optimize(neg_ll_t, c(0.3, 5))$minimum

bootstrap_filter <- function(y, n_particles, m0, c0, q, r_sd, r_w) {
  n <- length(y)
  particles <- rnorm(n_particles, m0, sqrt(c0))
  filtered <- numeric(n)
  log_lik <- 0
  for (t in seq_len(n)) {
    particles <- particles + rnorm(n_particles, 0, sqrt(q))
    lik <- r_w[1L] * dnorm(y[t], particles, r_sd[1L]) +
      r_w[2L] * dnorm(y[t], particles, r_sd[2L])
    log_lik <- log_lik + log(mean(lik))
    filtered[t] <- sum(lik * particles) / sum(lik)
    particles <- particles[sample.int(n_particles, n_particles,
                                      replace = TRUE, prob = lik)]
  }
  list(mean = filtered, log_lik = log_lik)
}

# the two-component noise as a nimble distribution
dmixnorm <- nimbleFunction(
  run = function(x = double(0), mean = double(0), sd1 = double(0),
                 sd2 = double(0), w = double(0),
                 log = integer(0, default = 0)) {
    returnType(double(0))
    d <- w * dnorm(x, mean, sd1) + (1 - w) * dnorm(x, mean, sd2)
    if (log) return(log(d)) else return(d)
  }
)
rmixnorm <- nimbleFunction(
  run = function(n = integer(0), mean = double(0), sd1 = double(0),
                 sd2 = double(0), w = double(0)) {
    returnType(double(0))
    if (runif(1) < w) return(rnorm(1, mean, sd1))
    return(rnorm(1, mean, sd2))
  }
)
registerDistributions(list(
  dmixnorm = list(BUGSdist = "dmixnorm(mean, sd1, sd2, w)",
                  types = c("value = double(0)"))
))

code_mix <- nimbleCode({
  x[1] ~ dnorm(m0, var = c0 + q_state)
  y[1] ~ dmixnorm(x[1], sd1, sd2, w1)
  for (t in 2:n_t) {
    x[t] ~ dnorm(x[t - 1], var = q_state)
    y[t] ~ dmixnorm(x[t], sd1, sd2, w1)
  }
})
code_t <- nimbleCode({
  x[1] ~ dnorm(m0, var = c0 + q_state)
  y[1] ~ dt(mu = x[1], sigma = t_scale, df = t_df)
  for (t in 2:n_t) {
    x[t] ~ dnorm(x[t - 1], var = q_state)
    y[t] ~ dt(mu = x[t], sigma = t_scale, df = t_df)
  }
})

# one compiled model and filter per noise model; the data are replaced
# in the compiled model for each series
build_filter <- function(code) {
  consts <- list(n_t = n_t, m0 = m0, c0 = c0, q_state = q_state,
                 sd1 = r_sd[1L], sd2 = r_sd[2L], w1 = r_w[1L],
                 t_df = t_df, t_scale = t_scale)
  model <- nimbleModel(code, data = list(y = rep(0, n_t)),
                       constants = consts, inits = list(x = rep(0, n_t)))
  cmodel <- compileNimble(model)
  filt <- buildBootstrapFilter(model, nodes = "x",
                               control = list(saveAll = TRUE, thresh = 1))
  list(model = cmodel, filter = compileNimble(filt, project = model))
}
nimble_mix <- build_filter(code_mix)
nimble_t <- build_filter(code_t)

run_nimble <- function(nb, y) {
  nb$model$y <- y
  log_lik <- nb$filter$run(n_particles)
  list(mean = colMeans(as.matrix(nb$filter$mvEWSamples, "x")),
       log_lik = log_lik)
}

prior_sim <- gmm(weights = 1, means = list(m0), covariances = list(matrix(c0)))
r_mix_sim <- gmm(weights = r_w, means = list(0, 0),
                 covariances = list(matrix(r_sd[1L]^2), matrix(r_sd[2L]^2)))
mod_dlm_sim <- dlmModPoly(1L, dV = r_var, dW = q_state, m0 = m0, C0 = c0)

one_series <- function(r) {
  set.seed(r)
  x <- m0 + sqrt(c0) * rnorm(1L) + cumsum(rnorm(n_t, 0, sqrt(q_state)))
  outlier <- runif(n_t) < r_w[2L]
  y <- x + rnorm(n_t, 0, ifelse(outlier, r_sd[2L], r_sd[1L]))

  timed <- function(expr) {
    secs <- system.time(out <- expr)[["elapsed"]]
    c(out, secs = secs)
  }
  fits <- list(
    "proxymix, two-component noise" = timed({
      f <- gmm_filter(prior_sim,
                      dynamics = list(A = matrix(1), Q = matrix(q_state)),
                      measurement = list(C = matrix(1), R = r_mix_sim),
                      y = y, k_max = 4L)
      list(mean = f$mean[, 1L], log_lik = sum(f$summary$log_evidence))
    }),
    "dlm, Gaussian noise" = timed({
      f <- dlmFilter(y, mod_dlm_sim)
      list(mean = as.numeric(f$m[-1L]),
           log_lik = -dlmLL(y, mod_dlm_sim) - n_t / 2 * log(2 * pi))
    }),
    "KFAS, Gaussian noise" = timed({
      mod <- SSModel(y ~ SSMtrend(1L, Q = q_state, a1 = m0,
                                  P1 = c0 + q_state), H = r_var)
      f <- KFS(mod, filtering = "state", smoothing = "none")
      list(mean = as.numeric(f$att), log_lik = as.numeric(logLik(mod)))
    }),
    "particle filter, two-component noise" = timed(
      bootstrap_filter(y, n_particles, m0, c0, q_state, r_sd, r_w)
    ),
    "nimble, two-component noise" = timed(run_nimble(nimble_mix, y)),
    "nimble, Student-t noise" = timed(run_nimble(nimble_t, y))
  )
  data.frame(
    series = r, method = names(fits),
    rmse = vapply(fits, function(f) sqrt(mean((f$mean - x)^2)), numeric(1L)),
    log_lik = vapply(fits, function(f) f$log_lik, numeric(1L)),
    secs = vapply(fits, function(f) f[["secs"]], numeric(1L)),
    row.names = NULL
  )
}

res <- do.call(rbind, lapply(seq_len(n_rep), one_series))

sim_tab <- aggregate(cbind(rmse, log_lik, secs) ~ method, data = res,
                     FUN = mean)
sim_tab$rmse_se <- aggregate(rmse ~ method, data = res,
                             FUN = function(v) sd(v) / sqrt(length(v)))$rmse
sim_tab
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order <- c("proxymix, two-component noise", "dlm, Gaussian noise",
                  "KFAS, Gaussian noise",
                  "particle filter, two-component noise",
                  "nimble, two-component noise", "nimble, Student-t noise")
tab_sim <- tab_sim[match(method_order, tab_sim$method), ]
knitr::kable(
  data.frame(
    method = tab_sim$method,
    rmse = fixed(tab_sim$rmse, 4),
    rmse_se = fixed(tab_sim$rmse_se, 4),
    log_lik = fixed(tab_sim$log_lik, 2),
    secs = fixed(tab_sim$secs, 3),
    stringsAsFactors = FALSE
  ),
  align = c("l", "r", "r", "r", "r"), row.names = FALSE,
  col.names = c("Filter", "RMSE", "SE of mean RMSE", "Log-likelihood",
                "Seconds"),
  caption = paste0(
    "Root mean squared error of the filtered mean against the true level, ",
    "computed per series and averaged over the series; the standard error ",
    "of that average; the mean log-likelihood of the series; and the mean ",
    "running time per series, over ", sim$n_rep, " simulated series of ",
    sim$n_t, " steps."
  )
)
sim_value <- function(method, what) {
  sim$sim_tab[[what]][sim$sim_tab$method == method]
}
rmse_pf <- sim_value("particle filter, two-component noise", "rmse")
rmse_pm <- sim_value("proxymix, two-component noise", "rmse")
rmse_dlm <- sim_value("dlm, Gaussian noise", "rmse")
rmse_kfas <- sim_value("KFAS, Gaussian noise", "rmse")
rmse_nmix <- sim_value("nimble, two-component noise", "rmse")
rmse_t <- sim_value("nimble, Student-t noise", "rmse")
ll_pf <- sim_value("particle filter, two-component noise", "log_lik")
ll_pm_sim <- sim_value("proxymix, two-component noise", "log_lik")
ll_gauss_sim <- sim_value("dlm, Gaussian noise", "log_lik")
ll_t_sim <- sim_value("nimble, Student-t noise", "log_lik")
secs_pm <- sim_value("proxymix, two-component noise", "secs")
secs_pf <- sim_value("particle filter, two-component noise", "secs")
secs_dlm <- sim_value("dlm, Gaussian noise", "secs")
if (is.null(sim$res)) {
  stop("The stored results carry no per-series results; rerun the ",
       "simulation.", call. = FALSE)
}
# the filters share each series, so differences in RMSE are paired by series:
# mean difference m1 minus m2, its standard error, and the share of series
# on which m2 has the lower RMSE
paired <- function(m1, m2) {
  a <- sim$res[sim$res$method == m1, ]
  b <- sim$res[sim$res$method == m2, ]
  d <- a$rmse[order(a$series)] - b$rmse[order(b$series)]
  c(diff = mean(d), se = sd(d) / sqrt(length(d)), share = mean(d > 0))
}
pm <- "proxymix, two-component noise"
pm_dlm <- paired("dlm, Gaussian noise", pm)
pm_t <- paired("nimble, Student-t noise", pm)
pm_pf <- paired(pm, "particle filter, two-component noise")
z_of <- function(p) p[["diff"]] / p[["se"]]
# a version as its DESCRIPTION writes it, when the installed one is the one stored
desc_version <- function(pkg) {
  stored <- sim$versions[[pkg]]
  if (identical(as.character(packageVersion(pkg)), stored)) {
    utils::packageDescription(pkg)$Version
  } else {
    stored
  }
}
```

Under the two-component noise the particle filter reaches a root mean
squared error of `r fixed(rmse_pf, 4)`, and the capped Gaussian-sum filter
`r fixed(rmse_pm, 4)`, with nimble's bootstrap filter under the same noise
at `r fixed(rmse_nmix, 4)`; the Monte Carlo standard error of each is about
`r fixed(sim_value("particle filter, two-component noise", "rmse_se"), 4)`.
The two Kalman filters, dlm and KFAS, give the same filtered means, and
`r if (fixed(rmse_dlm, 4) == fixed(rmse_kfas, 4)) paste("both reach an RMSE of", fixed(rmse_dlm, 4)) else paste("reach RMSEs of", fixed(rmse_dlm, 4), "and", fixed(rmse_kfas, 4))`,
so the Gaussian model of
the noise costs `r round(100 * (rmse_dlm / rmse_pf - 1))` per cent in error
on these series. The Student-t filter sits between, at `r fixed(rmse_t, 4)`.
The mean log-likelihood of the capped filter, `r fixed(ll_pm_sim, 2)`, is
within `r fixed(abs(round(ll_pm_sim, 2L) - round(ll_pf, 2L)), 2)` of the
particle filter's `r fixed(ll_pf, 2)`, against `r fixed(ll_gauss_sim, 2)` under the Gaussian
noise and `r fixed(ll_t_sim, 2)` under the Student-t noise. The capped filter
took `r fixed(secs_pm, 3)` s per series, the particle filter
`r fixed(secs_pf, 3)` s and dlm `r fixed(secs_dlm, 3)` s.

In this run proxymix under the two-component noise came out ahead of the
two Gaussian filters, with a mean RMSE lower by
`r fixed(pm_dlm[["diff"]], 3)`, about `r round(z_of(pm_dlm))` paired
standard errors of the difference, and lower on
`r round(pm_dlm[["share"]] * sim$n_rep)` of the `r sim$n_rep` series. Its mean RMSE
differed from the particle filter's by `r fixed(abs(pm_pf[["diff"]]), 4)`
(paired standard error `r fixed(pm_pf[["se"]], 4)`), and it was identical
to dlm and KFAS on the Nile at one component. Against the Student-t filter
its mean RMSE was `r fixed(abs(pm_t[["diff"]]), 4)`
`r if (pm_t[["diff"]] > 0) "lower" else "higher"`,
`r fixed(abs(z_of(pm_t)), 1)` paired standard errors of the difference,
which `r if (abs(z_of(pm_t)) < 2) "does not resolve the two" else "resolves the difference"`,
and it was lower on `r round(pm_t[["share"]] * sim$n_rep)` of the
`r sim$n_rep` series.
dlm and KFAS ran two orders of magnitude faster than any filter under the
two-component noise, the hand-written particle filter ran about
`r fixed(secs_pm / secs_pf, 1)` times faster than the capped filter, and on the Nile the Gaussian noise fitted the readings better than a
two-component noise of the same variance; the particle filters are the
reference proxymix is measured against rather than a competitor it beats. The comparison covers one scalar local-level model at one series
length. Every filter was given the true increment variance and starting
belief. The mixture filters were given the true two-component noise, dlm
and KFAS a Gaussian noise of the mixture's variance, and the Student-t
filter a scale fitted to the mixture. Noise with more components, a
higher-dimensional state, a process-noise mixture and estimated rather than
known parameters were not tried.

## Further reading

*Fitting a proxy to a density you cannot sample* produces the mixtures this
vignette operates on, and reports the fit-quality certificate that says
whether the proxy is worth operating on. *Testing the last observation for
instability* uses the filtering recursion built here as the engine of an
end-of-sample structural-break test. *Reading the entropy of a fitted
mixture* covers the divergence used above to price the reduction, and the
rest of the closed-form diagnostic surface. *One mixture, many methods*
shows the same conditioning operator standing in for regression, kernel
smoothing and principal components.

## References

* Alspach, D. L. and Sorenson, H. W. (1972). *Nonlinear Bayesian
  estimation using Gaussian sum approximations.* IEEE Transactions on
  Automatic Control 17(4), 439--448. <https://doi.org/10.1109/TAC.1972.1100034>.
* Cobb, G. W. (1978). *The problem of the Nile: Conditional solution to a
  changepoint problem.* Biometrika 65(2), 243--251.
  <https://doi.org/10.1093/biomet/65.2.243>.
* de Valpine, P., Turek, D., Paciorek, C. J., Anderson-Bergman, C.,
  Temple Lang, D. and Bodik, R. (2017). *Programming with models: Writing
  statistical algorithms for general model structures with NIMBLE.*
  Journal of Computational and Graphical Statistics 26(2), 403--413.
  <https://doi.org/10.1080/10618600.2016.1172487>.
* Durbin, J. and Koopman, S. J. (2012). *Time Series Analysis by State
  Space Methods*, 2nd edition. Oxford University Press.
  <https://doi.org/10.1093/acprof:oso/9780199641178.001.0001>.
* Gordon, N. J., Salmond, D. J. and Smith, A. F. M. (1993). *Novel
  approach to nonlinear/non-Gaussian Bayesian state estimation.* IEE
  Proceedings F, Radar and Signal Processing 140(2), 107--113.
  <https://doi.org/10.1049/ip-f-2.1993.0015>.
* Helske, J. (2017). *KFAS: Exponential family state space models in R.*
  Journal of Statistical Software 78(10), 1--39.
  <https://doi.org/10.18637/jss.v078.i10>.
* Kalman, R. E. (1960). *A new approach to linear filtering and
  prediction problems.* Journal of Basic Engineering 82(1), 35--45.
  <https://doi.org/10.1115/1.3662552>.
* Michaud, N., de Valpine, P., Turek, D., Paciorek, C. J. and Nguyen, D.
  (2021). *Sequential Monte Carlo methods in the nimble and nimbleSMC R
  packages.* Journal of Statistical Software 100(3), 1--39.
  <https://doi.org/10.18637/jss.v100.i03>.
* Murphy, K. P. (2012). *Machine Learning: A Probabilistic Perspective.*
  MIT Press. Ch. 4 (Gaussian models).
* Petris, G. (2010). *An R package for dynamic linear models.* Journal of
  Statistical Software 36(12), 1--16. <https://doi.org/10.18637/jss.v036.i12>.
* Runnalls, A. R. (2007). *Kullback--Leibler approach to Gaussian mixture
  reduction.* IEEE Transactions on Aerospace and Electronic Systems
  43(3), 989--999. <https://doi.org/10.1109/TAES.2007.4383588>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <https://doi.org/10.1080/07362994.2024.2372605>.

## Reproduce

The vignette sets `set.seed(20260514)` once, at the start of *Addressing
the problem*. The only draws that follow it are the simulated
constant-velocity track; every other operator result is deterministic
algebra that draws no random numbers at all.

The numerical illustration draws random numbers in one place, the particle
filter on the Nile, after `set.seed(20260925)`. In the simulation each
series is generated after `set.seed()` with its own index, and the particle
filters that follow draw from the same stream. The stored simulation results
come from the code shown, run on `r sim$run_date` with R
`r sim$versions[["R"]]`, proxymix `r desc_version("proxymix")`, dlm
`r desc_version("dlm")`, KFAS `r desc_version("KFAS")`, nimble
`r desc_version("nimble")` and nimbleSMC `r desc_version("nimbleSMC")`,
and the run raised `r length(sim$warnings)` warnings.

```{r session-info}
old_opt <- options(width = 72L)
si_vec <- trimws(capture.output(sessionInfo()), "right")
options(old_opt)
# library paths are single long words, so a long line breaks after a slash
long_vec <- nchar(si_vec) > 72L
si_vec[long_vec] <- gsub("(.{1,68}/)(?=.{8,})", "\\1\n        ",
                         si_vec[long_vec], perl = TRUE)
writeLines(si_vec)
```
