---
title: "Compressing a Bayesian posterior you can evaluate but not sample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compressing a Bayesian posterior you can evaluate but not sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~20 s under rmarkdown::render() with ggplot2, AER,
     LearnBayes, INLA and BayesianTools installed, the NUTS reference and
     the simulation read from their stored results; macOS arm64 (Apple
     silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r inline-numbers, include = FALSE}
# in the PDF, keep a signed number or an exponent on one line
keep_whole <- function(x) {
  if (knitr::is_latex_output()) paste0("\\mbox{", x, "}") else x
}
inline_default <- knitr::knit_hooks$get("inline")
knitr::knit_hooks$set(inline = function(x) {
  out <- inline_default(x)
  if (is.character(out) && length(out) == 1L &&
        grepl("^-?[0-9.]+(e[-+][0-9]+)?$", out) && grepl("-", out)) {
    out <- keep_whole(out)
  }
  out
})
# a version as its DESCRIPTION writes it, when the installed one is the one stored
desc_version <- function(pkg, stored = as.character(packageVersion(pkg))) {
  if (requireNamespace(pkg, quietly = TRUE) &&
        identical(as.character(packageVersion(pkg)), stored)) {
    utils::packageDescription(pkg)$Version
  } else {
    stored
  }
}
# a small positive number as a power of ten, typeset as math
power_ten <- function(x, digits = 2L) {
  e <- floor(log10(x))
  paste0("$", signif(x / 10^e, digits), " \\times 10^{", e, "}$")
}
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The problem

A modeller has written down a likelihood and a prior, can evaluate their
product at any parameter value, and does not want to spend the afternoon
tuning a sampler to get four numbers out of it. The question is: *I have an
unnormalised log-posterior I can evaluate and no appetite for tuning a
sampler -- can I get the marginal, the tail probability, the evidence and an
error bar out of it?*

The workflow below takes an unnormalised log-posterior and returns a
compact object that can be marginalised, conditioned, integrated and given
error bars, with no Markov chain in between.

## Package capabilities

`gmm_target()` wraps a log-density into the object the fitting verbs
consume. Declaring `normalised = FALSE` records that the normalising
constant is unknown, which is what an unnormalised posterior is.

`fit_proxymix()` with `regime = "kld"` runs the importance-sampled KLD-EM
of van der Hoek and Elliott (2024) against that target. Its `proposal`
argument takes the importance proposal, and `adapt = "pmc"` refreshes that
proposal from the current fit as the run proceeds, by population Monte Carlo
(Cappé et al., 2008). Each refreshed proposal keeps a share of its mass on
the initial one as a defensive component (Owen and Zhou, 2000), so the
initial proposal only has to be survivable rather than good. `proposal_mvt()` builds a
multivariate Student-t proposal, whose heavy tails are the robust default.

`gmm_fit_quality()` returns the certificate that travels with the fit: the
effective sample size of the importance weights, a degeneracy flag, the
support fraction, and a held-out validation gap. It is the part to read
before anything else.

`gmm_evidence()` estimates the log marginal likelihood, using the fitted
proxy as its own importance proposal. `gmm_marginalise()`, `pgmm()` and
`qgmm()` read marginals, tail probabilities and quantiles off the fitted
mixture in closed form, with no further target evaluations.

`gmm_fit_ensemble()` and `proxy_functional_ci()` price the fit's own
sampling variability: the ensemble re-weights the fit's cached importance
draws with Bayesian-bootstrap weights (Rubin, 1981), so a bootstrap interval on any functional of the mixture costs no new
posterior evaluations at all.

## Addressing the problem

### A real unnormalised posterior

A Bayesian logistic regression of transmission type on weight for the
built-in `mtcars` data, under a flat prior. The log-posterior is the
log-likelihood, known only up to its normalising constant, the integral of
the likelihood over the two coefficients. A flat prior is improper, so this
constant depends on the arbitrary scale given to the prior, and it cannot
be used as a model evidence to compare models.

```{r posterior}
set.seed(20260705)

y <- mtcars$am
w <- mtcars$wt

log_post <- function(theta) {
  if (is.null(dim(theta))) theta <- matrix(theta, ncol = 2L)
  eta <- outer(rep(1, length(y)), theta[, 1L]) + outer(w, theta[, 2L])
  colSums(y * eta - log1p(exp(eta)))
}

tgt <- gmm_target(
  n_dim = 2L,
  log_density = log_post,
  normalised = FALSE,
  name = "logistic(am ~ wt)"
)
```

### Fit the proxy

A cheap point fit locates the proposal. This is the usual workflow: any
rough guess of location and scale will do, because with `adapt = "pmc"` the
proposal is refreshed from the fit itself.

Population Monte Carlo re-estimates the fitting objective on a fresh
importance sample at each refresh. In this run the objective rises between
refreshes and falls at some of them, and the monotonicity check reports the
largest fall as a notice. The handler below prints the notice's first line;
its second line, a hint about ridge regularisation, is written for runs
without refreshes.

```{r fit}
mle <- stats::glm(am ~ wt, data = mtcars, family = stats::binomial())
q0 <- proposal_mvt(2L, mean = stats::coef(mle),
                   sigma = 9 * stats::vcov(mle), df = 5)

fit_notes <- character(0)
fit <- withCallingHandlers(
  fit_proxymix(tgt, N = 2L, regime = "kld", proposal = q0,
               is_size = 3000L, max_iter = 60L, seed = 1L,
               adapt = "pmc"),
  proxymix_nonmonotone = function(cond) {
    fit_notes <<- c(fit_notes, conditionMessage(cond))
    invokeRestart("muffleWarning")
  }
)

cert <- gmm_fit_quality(fit)
n_refresh <- fit@diagnostics$n_refresh
cat(sub("\n.*", "", fit_notes), sep = "\n")
```

```{r fit-quality-table, echo = FALSE}
knitr::kable(
  data.frame(
    quantity = c("regime", "converged", "degenerate",
                 "effective sample size", "effective sample size, relative",
                 "smallest per-component effective size", "largest weight",
                 "support fraction", "importance-sampled divergence, in-sample",
                 "held-out validation gap"),
    value = c(cert$regime, cert$converged, cert$degenerate,
              round(cert$ess, 1), round(cert$ess_relative, 3),
              round(cert$min_component_ess, 1),
              signif(cert$max_weight, 3), cert$support_fraction,
              round(cert$kld_final, 3), signif(cert$validation_gap, 3))
  ),
  row.names = FALSE,
  col.names = c("Certificate entry", "Value"),
  caption = paste0(
    "The fit-quality certificate that travels with the proxy through every ",
    "operation below."
  )
)
```

### The evidence, with an independent check

The fitted proxy doubles as the importance proposal for the normalising
constant, which `gmm_evidence()` reports as the log evidence. A Laplace approximation
(Tierney and Kadane, 1986) computed directly from the maximum-likelihood
fit gives an independent
first-order check that shares none of the importance machinery.

```{r evidence}
ev <- gmm_evidence(fit, n = 4000L, seed = 2L)

## Laplace approximation: log f(theta_hat) + (d/2) log(2 pi)
## - (1/2) log det(-Hessian).
H <- -solve(stats::vcov(mle))
log_z_laplace <- log_post(matrix(stats::coef(mle), nrow = 1L)) +
  log(2 * pi) -
  0.5 * as.numeric(determinant(-H, logarithm = TRUE)$modulus)

evidence_gap <- abs(ev$log_z - log_z_laplace)
c(importance = round(ev$log_z, 3), se = signif(ev$se_log_z, 3),
  laplace = round(log_z_laplace, 3), gap = round(evidence_gap, 3))
```

### Closed-form reads off the proxy

Everything below is exact algebra on the fitted mixture: no chains, no
further target evaluations. The marginal posterior of the slope, the
probability that the slope is negative, and a central 90 per cent interval.

```{r reads}
slope <- gmm_marginalise(fit, keep = 2L)
p_negative <- pgmm(0, slope)
p_positive <- pgmm(0, slope, lower.tail = FALSE)
interval <- qgmm(c(0.05, 0.95), slope)
c(p_slope_negative = round(p_negative, 5),
  p_slope_positive = signif(p_positive, 2),
  lower = round(interval[1L], 3), upper = round(interval[2L], 3))
```

### An independent reference for the marginal

The reads above are all functions of the fitted mixture, so on their own
they cannot say whether the mixture is right. In two dimensions the exact
posterior can be integrated on a grid instead, which uses the log-posterior
directly and never touches the fit. Exponentiating against the grid maximum
keeps the sum numerically stable. A rectangle (Riemann) sum over the grid
gives the log evidence, and summing over the intercept axis leaves the slope
marginal, which the same rectangle sum then normalises. The intercept and the
slope are strongly correlated, so the grid has to reach far along the ridge;
a much wider, coarser grid measures the share of the posterior mass that the
working grid leaves out.

```{r quadrature}
a_grid <- seq(-2, 70, length.out = 600L)     # intercept
b_grid <- seq(-21, 0, length.out = 400L)     # slope
quad <- as.matrix(expand.grid(a = a_grid, b = b_grid))

log_dens <- log_post(quad)
dens <- matrix(exp(log_dens - max(log_dens)), nrow = length(a_grid))
da <- a_grid[2L] - a_grid[1L]
db <- b_grid[2L] - b_grid[1L]
log_z_grid <- log(sum(dens) * da * db) + max(log_dens)

# the same posterior on a much wider, coarser grid, one slope at a time
a_wide <- seq(-20, 150, length.out = 700L)
b_wide <- seq(-50, 10, length.out = 700L)
log_wide <- vapply(b_wide, function(b) log_post(cbind(a_wide, b)),
                   numeric(length(a_wide)))
dens_wide <- exp(log_wide - max(log_wide))
in_grid <- outer(a_wide >= min(a_grid) & a_wide <= max(a_grid),
                 b_wide >= min(b_grid) & b_wide <= max(b_grid))
mass_outside <- sum(dens_wide[!in_grid]) / sum(dens_wide)

marg_quad <- colSums(dens) * da
marg_quad <- marg_quad / (sum(marg_quad) * db)
marg_proxy <- dgmm(matrix(b_grid, ncol = 1L), slope)

curve_gap <- max(abs(marg_quad - marg_proxy))
c(mass_outside = signif(mass_outside, 2))
c(log_z = round(log_z_grid, 3), peak_density = signif(max(marg_quad), 3))
c(max_curve_gap = signif(curve_gap, 3),
  gap_pct_of_peak = round(100 * curve_gap / max(marg_quad), 1))
```

```{r marginal-figure, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.8, fig.cap = "The slope marginal read off the fitted mixture in closed form (orange, dashed) against the same marginal obtained by numerical quadrature of the exact log-posterior (blue, solid). The quadrature curve uses the log-posterior directly and never touches the fit, so the comparison is an independent check rather than a self-consistency check. Both curves are drawn where the quadrature density exceeds a thousandth of its peak.", fig.alt = "Two close density curves for the slope parameter, one from the fitted Gaussian-mixture proxy and one from numerical quadrature of the exact posterior; the proxy curve is slightly higher at the mode and over the inner part of the left tail."}
# plotted where the quadrature density is above a thousandth of its peak
shown <- marg_quad > 1e-3 * max(marg_quad)
marg_df <- rbind(
  data.frame(slope = b_grid[shown], density = marg_quad[shown],
             source = "Quadrature of the exact posterior"),
  data.frame(slope = b_grid[shown], density = marg_proxy[shown],
             source = "Mixture proxy, closed form")
)

ggplot2::ggplot(marg_df,
                ggplot2::aes(slope, density, colour = source,
                             linetype = source)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("Quadrature of the exact posterior" = "#0072B2",
               "Mixture proxy, closed form" = "#D55E00")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Quadrature of the exact posterior" = "solid",
               "Mixture proxy, closed form" = "dashed")
  ) +
  ggplot2::labs(
    title = "The slope marginal, two ways",
    x = expression(paste("slope ", beta[wt], " (log-odds per 1000 lb)")),
    y = "posterior density"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 legend.position = "top")
```

The working grid stops at a slope of zero, so the probability that the
slope is positive is integrated on a finer grid of its own, against the
same normalising constant.

```{r quadrature-tail}
a_tail <- seq(-10 + 0.01, 10, by = 0.02)     # cell centres, intercept
b_tail <- seq(0.005, 3, by = 0.01)           # cell centres, slope > 0
log_tail <- vapply(b_tail, function(b) log_post(cbind(a_tail, b)),
                   numeric(length(a_tail)))
p_positive_quad <- sum(exp(log_tail - log_z_grid)) * 0.02 * 0.01
# the log-density on the grid's outer edges, relative to its peak
edge_drop <- max(log_tail[c(1L, nrow(log_tail)), ],
                 log_tail[, ncol(log_tail)]) - max(log_tail)
c(quadrature = signif(p_positive_quad, 2),
  proxy = signif(p_positive, 2),
  ratio = signif(p_positive / p_positive_quad, 2))
```

```{r quadrature-interval, include = FALSE}
# the cumulative sum reaches each cell's upper edge
cdf_quad <- cumsum(marg_quad) * db
interval_quad <- stats::approx(cdf_quad, b_grid + db / 2,
                               xout = c(0.05, 0.95), ties = mean)$y
gap_at <- b_grid[which.max(abs(marg_quad - marg_proxy))]
```

```{r marginal-figure-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the marginal-comparison",
    "figure is skipped.\n")
```

### Error bars on the proxy itself

The numbers above condition on the fitted mixture. The bootstrap ensemble
prices the fit's own sampling variability, at zero new posterior
evaluations, because the replicates re-weight the fit's cached importance
draws.

```{r ensemble}
ens <- gmm_fit_ensemble(fit, B = 80L, seed = 3L)
ci_mean <- proxy_functional_ci(ens, gmm_mean, level = 0.9)
```

The same interval for the probability of a negative slope comes with a
warning, which the next chunk keeps and prints.

```{r ensemble-tail}
# the message is laid out at this width, so it fits the page
old_opt <- options(cli.width = 72L)
tail_message <- NULL
ci_tail <- withCallingHandlers(
  proxy_functional_ci(
    ens, function(g) pgmm(0, gmm_marginalise(g, keep = 2L)), level = 0.9
  ),
  proxymix_tail_functional = function(w) {
    tail_message <<- conditionMessage(w)
    invokeRestart("muffleWarning")
  }
)
options(old_opt)
writeLines(tail_message)
```

```{r ensemble-tail-check, include = FALSE}
# the sentences on the tail warning below hold only if the warning is raised
if (is.null(tail_message) || 1 - ci_tail$estimate >= 1e-3) {
  stop("The tail-probability warning this section describes was not ",
       "raised, or the probability is not within 1e-3 of one.",
       call. = FALSE)
}
```

The warning is raised because this probability lies within $10^{-3}$ of
one.

```{r ensemble-table, echo = FALSE}
# three decimals for the two means, five for the probability
ens_fmt <- function(v) {
  vapply(seq_along(v), function(i1) {
    formatC(v[i1], format = "f", digits = c(3L, 3L, 5L)[i1])
  }, character(1L))
}
knitr::kable(
  data.frame(
    functional = c("posterior mean, intercept", "posterior mean, slope",
                   "P(slope < 0)"),
    estimate = ens_fmt(c(ci_mean$estimate, ci_tail$estimate)),
    conf_low = ens_fmt(c(ci_mean$conf.low, ci_tail$conf.low)),
    conf_high = ens_fmt(c(ci_mean$conf.high, ci_tail$conf.high))
  ),
  align = c("l", "r", "r", "r"),
  row.names = FALSE,
  col.names = c("Functional", "Estimate", "Lower 5%", "Upper 95%"),
  caption = paste0(
    "Bootstrap intervals over ", ens$B, " ensemble replicates."
  )
)
```

## Interpretation

The certificate is the first thing to read, and it clears the fit. The
effective sample size is `r round(cert$ess)` out of 3000 importance draws,
or `r round(100 * cert$ess_relative)` per cent, with the leanest single
component still carrying `r round(cert$min_component_ess)` effective draws
and no single weight above `r sprintf("%.5f", cert$max_weight)`. The fit is not
flagged degenerate. The held-out validation gap of
`r signif(cert$validation_gap, 2)` is about
`r round(abs(cert$validation_gap) / fit@diagnostics$validation_mc_se)` Monte Carlo
standard errors, detectable but small next to the divergence itself. The proposal was
refreshed `r n_refresh` times over the run. The certificate's in-sample divergence entry,
`r round(cert$kld_final, 3)`, is negative because the target is
unnormalised: for such a target the importance-sampled divergence is offset
by the unknown $\log Z$, and the entry is read as a relative quantity across
fits of the same target rather than as an absolute divergence. It is
computed on the draws the fit was tuned to and reads lower than the
held-out value, `r sprintf("%.3f", fit@diagnostics$validation_kld)` with a Monte
Carlo standard error of `r signif(fit@diagnostics$validation_mc_se, 1)`.

The log normalising constant comes out at `r round(ev$log_z, 3)`, with a
Monte Carlo standard error of `r signif(ev$se_log_z, 2)`. Grid quadrature
of the same posterior, above, gives `r round(log_z_grid, 3)`, a gap of
`r signif(abs(ev$log_z - log_z_grid), 2)` or about
`r format(round(abs(ev$log_z - log_z_grid) / ev$se_log_z, 1), nsmall = 1L)`
standard errors. The
Laplace approximation, which shares none of the importance machinery, gives
`r round(log_z_laplace, 3)`, a gap of `r round(evidence_gap, 3)` on the log
scale, or about `r round(evidence_gap / ev$se_log_z)` standard errors: it
agrees with the other two to first order.

The closed-form reads answer the substantive question. The posterior
probability that the slope is negative is
`r round(100 * p_negative, 3)` per cent, and the central 90 per cent interval
for the slope runs from `r round(interval[1L], 2)` to
`r round(interval[2L], 2)` log-odds per 1000 lb: heavier cars are almost
certainly less likely to be manual, and the effect size is not small. The
proxy puts `r power_ten(p_positive)` on a positive slope, and quadrature of
the exact posterior gives `r power_ten(p_positive_quad)`, so the proxy's
value is too large by a factor of about
`r signif(p_positive / p_positive_quad, 1)`. On the edges of the tail grid
the density is below $10^{-`r -ceiling(edge_drop / log(10))`}$ of its peak, so the
grid holds that mass.

The quadrature reference checks those reads against the posterior itself.
The working grid leaves out `r power_ten(mass_outside)` of the posterior
mass, as measured on the wider grid. The two marginal curves differ by at most `r signif(curve_gap, 2)` in
density, against a peak density of `r signif(max(marg_quad), 3)`, or about
`r round(100 * curve_gap / max(marg_quad))` per cent of the peak. The
largest gap is on the right shoulder, at a slope of
`r round(gap_at, 1)`, where the mixture sits
`r if (marg_proxy[b_grid == gap_at] < marg_quad[b_grid == gap_at]) "below" else "above"`
the exact curve. The mixture is also higher at the mode, lower on
the left shoulder, higher over the inner part of the left tail and slightly
lower over its far end. The upper end of the 90 per cent
interval agrees with quadrature, `r round(interval[2L], 2)` against
`r round(interval_quad[2L], 2)`, and the lower end sits further out,
`r round(interval[1L], 2)` against `r round(interval_quad[1L], 2)`.

The ensemble prices the variability of the fit under re-weighting of its
importance draws, at no further evaluations of the posterior. Over
`r ens$B` replicates the posterior mean of the slope has a 90 per cent
interval of width
`r signif(ci_mean$conf.high[2L] - ci_mean$conf.low[2L], 2)`, and the
probability of a negative slope the interval
`r sprintf("%.5f", ci_tail$conf.low[1L])` to
`r sprintf("%.5f", ci_tail$conf.high[1L])`. That interval
`r if (ci_tail$conf.low[1L] <= 1 - p_positive_quad && 1 - p_positive_quad <= ci_tail$conf.high[1L]) "contains" else "excludes"`
the quadrature value of
`r format(1 - p_positive_quad, digits = 8L)`. It covers the variability of
the importance-sampling fit only, not the error of the Gaussian family in
the far tail, which here is about
`r round(log10(p_positive / p_positive_quad))` orders of magnitude on the
probability of a positive slope. `proxy_functional_ci()` warns, with class
`proxymix_tail_functional`, when the functional is a probability whose value
or complement is below $10^{-3}$, as it did above.

So the answer to the question is yes for the marginal, the normalising
constant and the central interval, each checked against quadrature above.
For a probability this far in the tail the proxy is not accurate. The
certificate and the ensemble interval do not show this, and the tail
warning is the only sign of it.

## Limitations

If the unnormalised posterior can be evaluated then a Markov chain can
always be run against it and a sample-based mixture fitted to the draws. The
direct fit is the better route when what is wanted *is* the compact
closed-form object -- deterministic given the seed, no chain tuning,
evidence and error bars included -- and not otherwise.

The dimension is the binding constraint. The effective sample size of
importance sampling falls sharply beyond roughly five to ten parameters, and
every fit reports it in its certificate. The posterior here has two parameters, which is the
comfortable end of that range. For high-dimensional posteriors, sample first
and use the classical expectation-maximisation regime on the draws.

The quadrature check above is available only because the posterior is
two-dimensional. It is not part of the workflow and does not scale: at four
or more parameters the grid becomes the expensive object it was introduced
to avoid, and the certificate and the held-out validation gap are what
remain.

The prior here is flat and improper, so the normalising constant reported
as the evidence is the integral of the likelihood, and it changes with the
scale an improper prior is given; it is not a model evidence for comparing
models. A proper prior changes the target, the evidence and the interval,
and nothing in this vignette speaks to how the conclusions move under a
different one.

The nearest CRAN neighbour, `AdMit` (Ardia et al., 2009), adaptively fits a mixture of Student-t
densities to an evaluable kernel for use as a proposal. The Gaussian family
used here is what buys the closed-form operator calculus and the certificate
that travel with the result; a Student-t mixture would fit heavier tails
more naturally and give up that algebra. The numerical illustration that
follows puts the fit beside a Hamiltonian Monte Carlo sampler, a Laplace
approximation, INLA and a differential-evolution sampler, on a
five-parameter posterior under a proper prior and in simulation, with a long
sampler run as the reference.

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_aer <- requireNamespace("AER", quietly = TRUE)
has_learnbayes <- requireNamespace("LearnBayes", quietly = TRUE)
has_inla <- requireNamespace("INLA", quietly = TRUE)
has_bt <- requireNamespace("BayesianTools", quietly = TRUE)
run_illustration <- has_aer && has_learnbayes && has_inla && has_bt
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("AER, LearnBayes, INLA or BayesianTools is not available on this build,",
    "so the real-data illustration is skipped.\n")
```

### The Affairs probit posterior, five ways

Fair's (1978) survey of 601 people, shipped as `Affairs` in AER (Kleiber
and Zeileis, 2008), records whether each respondent reported an
extramarital affair in the past year, together with age, years married, a
self-rating of religiousness from 1 to 5 and a self-rating of the marriage
from 1 to 5. The model is a probit regression of any affair on those four
covariates, five coefficients with the intercept, under independent normal
priors of mean 0 and standard deviation 10 on each. The reference is a NUTS
run of 40 000 draws of Stan (Carpenter et al., 2017) through cmdstanr
(Gabry et al., 2025), with bridge
sampling on those draws for the evidence (Gronau et al., 2020). Against it stand a NUTS run of 4000 draws, also with bridge sampling,
the Laplace approximation of `LearnBayes::laplace()` (Albert, 2009), INLA
(Rue et al., 2009), the differential-evolution sampler DEzs (ter Braak and
Vrugt, 2008) of BayesianTools (Hartig et al., 2026), and proxymix. Every method
returns the posterior mean, standard deviation, central 95 per cent
interval and $P(\beta_j > 0)$ for each coefficient, and every method but
DEzs an estimate of the log evidence.

One scoring rule covers all of them. The error in a mean, a standard
deviation or an interval endpoint is its absolute distance from the
reference in units of the reference posterior standard deviation of the
same coefficient, averaged over the five coefficients; the error in
$P(\beta_j > 0)$ is the absolute difference in probability, averaged the
same way; the error in the log evidence is absolute. Wall time is the time
each method took to produce those numbers on one core, and the settings
column counts the arguments in the code shown that steer the computation
rather than define the model.

The competitors install once and load as usual; cmdstanr and INLA are
distributed from their own repositories rather than from CRAN, and cmdstanr
also installs CmdStan itself:

```{r install-competitors, eval = FALSE}
install.packages(c("AER", "LearnBayes", "bridgesampling", "BayesianTools"))
install.packages("cmdstanr",
                 repos = c("https://stan-dev.r-universe.dev",
                           getOption("repos")))
install.packages("INLA",
                 repos = c("https://inla.r-inla-download.org/R/stable",
                           getOption("repos")))
cmdstanr::install_cmdstan()
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(LearnBayes)
library(INLA)
library(BayesianTools)
```

The data and the log posterior, in a form that takes one parameter vector
and in a form that takes a matrix of them, one per row:

```{r affairs-data, eval = run_illustration}
data("Affairs", package = "AER")
prior_sd <- 10
y_aff <- as.integer(Affairs$affairs > 0)
X_aff <- model.matrix(~ age + yearsmarried + religiousness + rating, Affairs)
p_aff <- ncol(X_aff)
data_aff <- list(n = nrow(X_aff), p = p_aff, X = X_aff, y = y_aff,
                 prior_sd = prior_sd)

log_lik_probit <- function(theta, data) {
  eta <- data$X %*% theta
  sum(data$y * pnorm(eta, log.p = TRUE) +
        (1 - data$y) * pnorm(-eta, log.p = TRUE))
}
log_post_probit <- function(theta, data) {
  log_lik_probit(theta, data) + sum(dnorm(theta, 0, data$prior_sd, log = TRUE))
}
log_post_probit_rows <- function(theta, data) {
  if (is.null(dim(theta))) theta <- matrix(theta, ncol = data$p)
  apply(theta, 1L, log_post_probit, data = data)
}
```

```{r ref-read, eval = run_illustration, echo = FALSE}
ref <- readRDS("posterior_proxy_ref.rds")
ref_code <- as.character(c(knitr::knit_code$get("affairs-data"),
                           knitr::knit_code$get("ref-code")))
if (!identical(ref$code, ref_code)) {
  stop("The stored NUTS reference was produced by code other than the ",
       "chunks shown; rerun the reference.", call. = FALSE)
}
ref_aff <- ref$long$summary
```

The two NUTS runs are the reference and the first competitor. The Stan
program is written out by the code itself, and `cmdstanr` compiles it. This
chunk is complete and runs as shown, but it took about
`r if (run_illustration) round(ref$elapsed_secs / 60) else ""` minutes on
one core (R `r if (run_illustration) ref$versions[["R"]] else ""`,
Apple silicon); the results below are read from its stored output. Changing
`10000L` in the last lines to `1000L` runs a reduced version in under a
minute.

```{r ref-code, eval = FALSE}
library(cmdstanr)
library(bridgesampling)

stan_file <- file.path(tempdir(), "probit.stan")
writeLines(c(
  "data {",
  "  int<lower=1> n;",
  "  int<lower=1> p;",
  "  matrix[n, p] X;",
  "  array[n] int<lower=0, upper=1> y;",
  "  real<lower=0> prior_sd;",
  "}",
  "parameters {",
  "  vector[p] beta;",
  "}",
  "model {",
  "  vector[n] eta = X * beta;",
  "  beta ~ normal(0, prior_sd);",
  "  for (i in 1:n) {",
  "    target += std_normal_lcdf(y[i] == 1 ? eta[i] : -eta[i]);",
  "  }",
  "}"
), stan_file)
model <- cmdstan_model(stan_file)

# posterior mean, SD, central 95% interval and P(beta > 0) per coefficient
summarise_draws <- function(draws) {
  cbind(mean = colMeans(draws),
        sd = apply(draws, 2L, sd),
        lower = apply(draws, 2L, quantile, probs = 0.025, names = FALSE),
        upper = apply(draws, 2L, quantile, probs = 0.975, names = FALSE),
        p_pos = colMeans(draws > 0))
}

# four sequential chains of 1000 warm-up and iter_sampling draws, then
# bridge sampling on the draws for the log evidence
run_nuts <- function(data, iter_sampling, seed) {
  secs_sample <- system.time(
    fit <- model$sample(data, chains = 4L, parallel_chains = 1L,
                        iter_warmup = 1000L, iter_sampling = iter_sampling,
                        seed = seed, refresh = 0L, show_messages = FALSE,
                        show_exceptions = FALSE)
  )[["elapsed"]]
  draws <- fit$draws("beta", format = "matrix")
  draws <- matrix(draws, ncol = data$p,
                  dimnames = list(NULL, colnames(draws)))
  bounds <- setNames(rep(-Inf, data$p), colnames(draws))
  secs_bridge <- system.time(
    bridge <- bridge_sampler(draws, log_posterior = log_post_probit,
                             data = data, lb = bounds, ub = -bounds,
                             silent = TRUE)
  )[["elapsed"]]
  diag <- fit$summary("beta")
  list(summary = summarise_draws(draws), log_z = bridge$logml,
       secs_sample = secs_sample, secs_bridge = secs_bridge,
       ess_min = min(diag$ess_bulk), rhat_max = max(diag$rhat),
       draws = draws[seq(1L, nrow(draws), by = 10L), ])
}

ref <- list(long = run_nuts(data_aff, 10000L, 1L),
            short = run_nuts(data_aff, 1000L, 1L + 100000L))
```

Each of the other four methods runs when the vignette builds, on the same
data and the same log posterior. proxymix uses the recipe of the two-parameter
example above, a Student-t proposal placed at the maximum-likelihood fit and
refreshed by population Monte Carlo, and the same starting point is given
to the Laplace approximation. INLA is given the prior as a precision on
every fixed effect, and its evidence is the marginal likelihood it reports.
DEzs is given the log-likelihood and the normal prior as a density with a
sampler, and its first 1000 iterations per chain are discarded.
`summarise_draws()` is the function of the reference chunk above, and
`summarise_gmm()` and `summarise_normal()` read the same five quantities off
a mixture and off a normal approximation.

```{r affairs-summarise-draws, eval = run_illustration, echo = FALSE}
summarise_draws <- function(draws) {
  cbind(mean = colMeans(draws),
        sd = apply(draws, 2L, sd),
        lower = apply(draws, 2L, quantile, probs = 0.025, names = FALSE),
        upper = apply(draws, 2L, quantile, probs = 0.975, names = FALSE),
        p_pos = colMeans(draws > 0))
}
```

```{r affairs-fit, eval = run_illustration}
summarise_gmm <- function(g) {
  marg <- lapply(seq_len(gmm_dim(g)),
                 function(j) gmm_marginalise(g, keep = j))
  cbind(mean = gmm_mean(g),
        sd = sqrt(diag(gmm_cov(g))),
        lower = vapply(marg, function(m) qgmm(0.025, m), numeric(1L)),
        upper = vapply(marg, function(m) qgmm(0.975, m), numeric(1L)),
        p_pos = vapply(marg, function(m) pgmm(0, m, lower.tail = FALSE),
                       numeric(1L)))
}
summarise_normal <- function(mean, sd) {
  z <- qnorm(0.975)
  cbind(mean = mean, sd = sd, lower = mean - z * sd, upper = mean + z * sd,
        p_pos = pnorm(mean / sd))
}

mle_aff <- glm(y_aff ~ X_aff - 1, family = binomial(link = "probit"))

## proxymix: fit, closed-form reads, evidence
aff_notes <- 0L
secs_px <- system.time({
  tgt_aff <- gmm_target(
    n_dim = p_aff,
    log_density = function(theta) log_post_probit_rows(theta, data_aff),
    normalised = FALSE, name = "probit(any affair)"
  )
  q0_aff <- proposal_mvt(p_aff, mean = coef(mle_aff),
                         sigma = 9 * vcov(mle_aff), df = 5)
  fit_aff <- withCallingHandlers(
    fit_proxymix(tgt_aff, N = 2L, regime = "kld", proposal = q0_aff,
                 is_size = 3000L, max_iter = 60L, seed = 1L, adapt = "pmc"),
    proxymix_nonmonotone = function(cond) {
      aff_notes <<- aff_notes + 1L
      invokeRestart("muffleWarning")
    }
  )
  summ_px <- summarise_gmm(fit_aff)
  ev_aff <- gmm_evidence(fit_aff, n = 4000L, seed = 2L)
})[["elapsed"]]
cert_aff <- gmm_fit_quality(fit_aff)

## proxymix: error bars on the posterior means, from the ensemble
secs_ens <- system.time({
  ens_aff <- gmm_fit_ensemble(fit_aff, B = 80L, seed = 3L)
  ci_aff <- proxy_functional_ci(ens_aff, gmm_mean, level = 0.95)
})[["elapsed"]]

## Laplace approximation
secs_lap <- system.time(
  lap_aff <- laplace(log_post_probit, coef(mle_aff), data_aff)
)[["elapsed"]]
summ_lap <- summarise_normal(lap_aff$mode, sqrt(diag(lap_aff$var)))

## INLA
df_aff <- data.frame(y = y_aff, X_aff[, -1L])
secs_inla <- system.time(
  inla_aff <- inla(y ~ age + yearsmarried + religiousness + rating,
                   family = "binomial", Ntrials = 1, data = df_aff,
                   control.family = list(
                     control.link = list(model = "probit")
                   ),
                   control.fixed = list(prec = 1 / prior_sd^2,
                                        prec.intercept = 1 / prior_sd^2),
                   num.threads = "1:1")
)[["elapsed"]]
summ_inla <- cbind(
  mean = inla_aff$summary.fixed$mean,
  sd = inla_aff$summary.fixed$sd,
  lower = inla_aff$summary.fixed[["0.025quant"]],
  upper = inla_aff$summary.fixed[["0.975quant"]],
  p_pos = 1 - vapply(inla_aff$marginals.fixed,
                     function(m) inla.pmarginal(0, m), numeric(1L))
)

## BayesianTools, DEzs
setup_aff <- createBayesianSetup(
  likelihood = function(theta) log_lik_probit(theta, data_aff),
  prior = createPrior(
    density = function(theta) sum(dnorm(theta, 0, prior_sd, log = TRUE)),
    sampler = function(n = 1L) matrix(rnorm(n * p_aff, 0, prior_sd), n),
    lower = rep(-5 * prior_sd, p_aff), upper = rep(5 * prior_sd, p_aff)
  )
)
set.seed(4L)
secs_bt <- system.time(
  bt_aff <- runMCMC(setup_aff, sampler = "DEzs",
                    settings = list(iterations = 15000L, message = FALSE))
)[["elapsed"]]
summ_bt <- summarise_draws(getSample(bt_aff, start = 1000L))
```

```{r fixed-notation, include = FALSE}
## Small errors are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
method_label <- c("NUTS, 4000 draws" = "the 4000-draw NUTS run",
                  proxymix = "proxymix",
                  Laplace = "the Laplace approximation",
                  INLA = "INLA", DEzs = "DEzs")
# a list in the order of one column, each entry with its value
ordered_text <- function(methods, values, digits, unit = "") {
  o <- order(values)
  items <- paste0(method_label[methods[o]], " (",
                  trimws(fixed(values[o], digits)), unit, ")")
  paste0(paste(items[-length(items)], collapse = ", "), " and ",
         items[length(items)])
}
n_word <- function(k) {
  c("none", "one", "two", "three", "four", "five", "six", "seven", "eight",
    "nine", "ten")[k + 1L]
}
```

```{r affairs-table, eval = run_illustration, echo = FALSE}
score_against <- function(summ, log_z, ref, ref_log_z) {
  sd_ref <- ref[, "sd"]
  c(mean = mean(abs(summ[, "mean"] - ref[, "mean"]) / sd_ref),
    sd = mean(abs(summ[, "sd"] - ref[, "sd"]) / sd_ref),
    endpoints = mean(c(abs(summ[, "lower"] - ref[, "lower"]),
                       abs(summ[, "upper"] - ref[, "upper"])) / sd_ref),
    p_pos = mean(abs(summ[, "p_pos"] - ref[, "p_pos"])),
    log_z = abs(log_z - ref_log_z))
}
aff_scores <- rbind(
  "NUTS, 4000 draws" = score_against(ref$short$summary, ref$short$log_z,
                                     ref_aff, ref$long$log_z),
  proxymix = score_against(summ_px, ev_aff$log_z, ref_aff, ref$long$log_z),
  Laplace = score_against(summ_lap, lap_aff$int, ref_aff, ref$long$log_z),
  INLA = score_against(summ_inla, inla_aff$mlik[1L], ref_aff,
                       ref$long$log_z),
  DEzs = score_against(summ_bt, NA_real_, ref_aff, ref$long$log_z)
)
aff_tbl <- data.frame(
  method = rownames(aff_scores),
  aff_scores,
  secs = c(ref$short$secs_sample + ref$short$secs_bridge, secs_px, secs_lap,
           secs_inla, secs_bt),
  settings = c(3L, 5L, 1L, 0L, 2L),
  stringsAsFactors = FALSE
)
old_na <- options(knitr.kable.NA = "--")
knitr::kable(
  aff_tbl, digits = c(0L, 3L, 3L, 3L, 4L, 4L, 2L, 0L), row.names = FALSE,
  col.names = c("Method", "Mean", "SD", "Endpoints", "$P(\\beta_j > 0)$",
                "$\\log Z$", "Seconds", "Settings"),
  caption = paste0(
    "Error of each method against the 40 000-draw NUTS reference on the ",
    "Affairs probit posterior: mean, SD and interval endpoints in units of ",
    "the reference posterior SD, averaged over the five coefficients; ",
    "$P(\\beta_j > 0)$ in probability; log evidence $\\log Z$ against bridge ",
    "sampling on the reference draws. Seconds on one core; the NUTS time includes its ",
    "bridge sampling. DEzs returns no evidence estimate, so its log-evidence ",
    "cell is empty."
  )
)
options(old_na)
covered_aff <- ci_aff$conf.low <= ref_aff[, "mean"] &
  ref_aff[, "mean"] <= ci_aff$conf.high
ref_se_mean <- ref_aff[, "sd"] / sqrt(ref$long$ess_min)
# standard error of a difference in a mean between the two independent runs,
# in reference standard deviations
se_diff_nuts <- sqrt(1 / ref$long$ess_min + 1 / ref$short$ess_min)
```

```{r affairs-figure, eval = run_illustration && has_ggplot2, echo = FALSE, fig.height = 4.2, fig.cap = "The marginal posterior of the marriage-rating coefficient in the Affairs probit model: a density estimate from every tenth draw of the 40 000-draw NUTS reference (black, solid), the proxymix marginal read off the fitted mixture in closed form (orange, dashed) and the Laplace approximation (blue, dotted).", fig.alt = "Three nearly coincident density curves for the rating coefficient, from the NUTS reference draws, the closed-form mixture marginal and the Laplace normal approximation."}
j <- p_aff
grid <- seq(ref_aff[j, "mean"] - 4 * ref_aff[j, "sd"],
            ref_aff[j, "mean"] + 4 * ref_aff[j, "sd"], length.out = 300L)
d_ref <- density(ref$long$draws[, j], from = min(grid), to = max(grid),
                 n = 300L)
fig_df <- rbind(
  data.frame(beta = d_ref$x, density = d_ref$y, source = "NUTS reference"),
  data.frame(beta = grid,
             density = dgmm(matrix(grid, ncol = 1L),
                            gmm_marginalise(fit_aff, keep = j)),
             source = "proxymix, closed form"),
  data.frame(beta = grid,
             density = dnorm(grid, lap_aff$mode[j], sqrt(lap_aff$var[j, j])),
             source = "Laplace")
)
fig_df$source <- factor(fig_df$source,
                        levels = c("NUTS reference", "proxymix, closed form",
                                   "Laplace"))
ggplot2::ggplot(fig_df, ggplot2::aes(beta, density, colour = source,
                                     linetype = source)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_colour_manual(
    name = NULL,
    values = c("NUTS reference" = "#000000",
               "proxymix, closed form" = "#D55E00",
               "Laplace" = "#0072B2")
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("NUTS reference" = "solid", "proxymix, closed form" = "dashed",
               "Laplace" = "dotted")
  ) +
  ggplot2::labs(
    title = "The marriage-rating coefficient, three ways",
    x = expression(paste("coefficient ", beta[rating], " (probit scale)")),
    y = "posterior density"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 legend.position = "top")
```

```{r affairs-prose, eval = run_illustration, echo = FALSE}
aff_value <- function(method, what) aff_tbl[aff_tbl$method == method, what]
time_order <- ordered_text(aff_tbl$method, aff_tbl$secs, 2L, " s")
```

The reference run's smallest bulk effective sample size over the five
coefficients is `r format(round(ref$long$ess_min), big.mark = ",")`, so
its posterior means carry a Monte Carlo standard error of at most about
`r fixed(max(ref_se_mean / ref_aff[, "sd"]), 3)` reference standard
deviations. The 4000-draw NUTS run is independent of the reference, with a
smallest bulk effective sample size of
`r format(round(ref$short$ess_min), big.mark = ",")`, so a difference in a
mean between the two runs has a standard error of about
`r fixed(se_diff_nuts, 3)` reference standard deviations. Its means differ
from the reference by `r fixed(aff_value("NUTS, 4000 draws", "mean"), 3)`
on average, and bridge sampling on its draws is
`r fixed(aff_value("NUTS, 4000 draws", "log_z"), 4)` off the reference
evidence. proxymix's means are
`r fixed(aff_value("proxymix", "mean"), 3)` reference standard deviations
from the reference and its endpoints
`r fixed(aff_value("proxymix", "endpoints"), 3)`, its evidence is
`r fixed(aff_value("proxymix", "log_z"), 4)` off the bridge-sampling value,
and its certificate reports a relative effective sample size of
`r fixed(cert_aff$ess_relative, 2)` with `r aff_notes` monotonicity
notice`r if (aff_notes != 1L) "s"` over the run. INLA's means are
`r fixed(aff_value("INLA", "mean"), 3)` off and its endpoints
`r fixed(aff_value("INLA", "endpoints"), 3)`, the closest of the
approximations. The Laplace approximation is
`r fixed(aff_value("Laplace", "mean"), 3)` off on the means, because it
reports the mode, and `r fixed(aff_value("Laplace", "log_z"), 4)` off on
the evidence. DEzs (BayesianTools `r desc_version("BayesianTools")`),
run for 15 000 iterations from prior draws with the first 1000 per chain
discarded, returns standard deviations
`r fixed(aff_value("DEzs", "sd"), 3)` reference standard deviations off on
this posterior, whose covariates are on their raw scales. proxymix's 95 per
cent ensemble intervals on the posterior means contain the reference mean
for `r n_word(sum(covered_aff))` of the `r n_word(p_aff)` coefficients, at a further
`r fixed(secs_ens, 1)` s. On wall time the order, fastest first, is
`r time_order`.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("posterior_proxy_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
```

One posterior is one draw. The simulation repeats the comparison on
`r sim$n_rep` datasets of `r sim$n` observations for a logistic regression
with `r sim$p_vec[1L]` and with `r sim$p_vec[2L]` coefficients, the
intercept included. Covariates are standard normal, the population
coefficients are the first $p$ of `r paste(sim$beta_pop, collapse = ", ")`,
where $p$ is the number of coefficients, and the prior is normal with mean
0 and standard deviation `r sim$prior_sd` on every coefficient. For each
dataset the reference is a fresh NUTS run of 40 000 draws with bridge
sampling for the evidence, and the same five methods are scored against it
by the rule of the Affairs table. Two further quantities are recorded for
proxymix: whether its 95 per cent ensemble interval on each posterior mean
contains the reference mean, and the time of that ensemble. This chunk is
complete and runs as shown, but it took about
`r round(sim$elapsed_secs / 60)` minutes on one core (R
`r sim$versions[["R"]]`, Apple silicon); the results below are
read from its stored output. Changing `n_rep <- 50L` to `n_rep <- 3L` runs
a reduced version in a few minutes.

```{r sim-code, eval = FALSE}
library(proxymix)
library(cmdstanr)
library(LearnBayes)
library(INLA)
library(bridgesampling)
library(BayesianTools)

n <- 200L                # observations in each dataset
n_rep <- 50L
p_vec <- c(2L, 5L)       # coefficients, intercept included
prior_sd <- 10
beta_pop <- c(0.5, 1.5, -1, 0.75, 0)   # the first p entries are used

stan_file <- file.path(tempdir(), "logistic.stan")
writeLines(c(
  "data {",
  "  int<lower=1> n;",
  "  int<lower=1> p;",
  "  matrix[n, p] X;",
  "  array[n] int<lower=0, upper=1> y;",
  "  real<lower=0> prior_sd;",
  "}",
  "parameters {",
  "  vector[p] beta;",
  "}",
  "model {",
  "  beta ~ normal(0, prior_sd);",
  "  y ~ bernoulli_logit(X * beta);",
  "}"
), stan_file)
model <- cmdstan_model(stan_file)

log_lik_one <- function(theta, data) {
  eta <- data$X %*% theta
  sum(data$y * eta - log1p(exp(eta)))
}
log_post_one <- function(theta, data) {
  log_lik_one(theta, data) + sum(dnorm(theta, 0, data$prior_sd, log = TRUE))
}
log_post_rows <- function(theta, data) {
  if (is.null(dim(theta))) theta <- matrix(theta, ncol = data$p)
  apply(theta, 1L, log_post_one, data = data)
}

# posterior mean, SD, central 95% interval and P(beta > 0) per coefficient
summarise_draws <- function(draws) {
  cbind(mean = colMeans(draws),
        sd = apply(draws, 2L, sd),
        lower = apply(draws, 2L, quantile, probs = 0.025, names = FALSE),
        upper = apply(draws, 2L, quantile, probs = 0.975, names = FALSE),
        p_pos = colMeans(draws > 0))
}
summarise_gmm <- function(g) {
  marg <- lapply(seq_len(gmm_dim(g)), function(j) gmm_marginalise(g, keep = j))
  cbind(mean = gmm_mean(g),
        sd = sqrt(diag(gmm_cov(g))),
        lower = vapply(marg, function(m) qgmm(0.025, m), numeric(1L)),
        upper = vapply(marg, function(m) qgmm(0.975, m), numeric(1L)),
        p_pos = vapply(marg, function(m) pgmm(0, m, lower.tail = FALSE),
                       numeric(1L)))
}
summarise_normal <- function(mean, sd) {
  z <- qnorm(0.975)
  cbind(mean = mean, sd = sd, lower = mean - z * sd, upper = mean + z * sd,
        p_pos = pnorm(mean / sd))
}

# four sequential chains of 1000 warm-up and iter_sampling draws, then
# bridge sampling on the draws for the log evidence
run_nuts <- function(data, iter_sampling, seed) {
  secs_sample <- system.time(
    fit <- model$sample(data, chains = 4L, parallel_chains = 1L,
                        iter_warmup = 1000L, iter_sampling = iter_sampling,
                        seed = seed, refresh = 0L, show_messages = FALSE,
                        show_exceptions = FALSE)
  )[["elapsed"]]
  draws <- fit$draws("beta", format = "matrix")
  draws <- matrix(draws, ncol = data$p,
                  dimnames = list(NULL, colnames(draws)))
  bounds <- setNames(rep(-Inf, data$p), colnames(draws))
  secs_bridge <- system.time(
    bridge <- bridge_sampler(draws, log_posterior = log_post_one,
                             data = data, lb = bounds, ub = -bounds,
                             silent = TRUE)
  )[["elapsed"]]
  list(summary = summarise_draws(draws), log_z = bridge$logml,
       secs = secs_sample + secs_bridge,
       ess_min = min(fit$summary("beta")$ess_bulk))
}

# errors against the reference, one row per coefficient
score <- function(p, r, method, summ, log_z, secs, ref, ref_log_z) {
  sd_ref <- ref[, "sd"]
  data.frame(
    p = p, rep = r, method = method, coef = seq_len(p),
    err_mean = abs(summ[, "mean"] - ref[, "mean"]) / sd_ref,
    err_sd = abs(summ[, "sd"] - ref[, "sd"]) / sd_ref,
    err_lower = abs(summ[, "lower"] - ref[, "lower"]) / sd_ref,
    err_upper = abs(summ[, "upper"] - ref[, "upper"]) / sd_ref,
    err_p_pos = abs(summ[, "p_pos"] - ref[, "p_pos"]),
    err_log_z = abs(log_z - ref_log_z),
    secs = secs, covered = NA, secs_ensemble = NA_real_,
    ess_relative = NA_real_, ref_ess_min = NA_real_, n_notes = NA_integer_
  )
}

one_dataset <- function(p, r) {
  set.seed(r)
  X <- cbind(1, matrix(rnorm(n * (p - 1L)), n))
  y <- rbinom(n, 1L, plogis(X %*% beta_pop[seq_len(p)]))
  data <- list(n = n, p = p, X = X, y = y, prior_sd = prior_sd)

  long <- run_nuts(data, 10000L, r)
  ref <- long$summary
  ref_log_z <- long$log_z

  short <- run_nuts(data, 1000L, r + 100000L)
  out_nuts <- score(p, r, "NUTS, 4000 draws", short$summary, short$log_z,
                    short$secs, ref, ref_log_z)
  out_nuts$ref_ess_min <- long$ess_min

  mle <- glm(y ~ X - 1, family = binomial())
  notes <- 0L    # monotonicity notices from this fit
  secs_px <- system.time({
    tgt <- gmm_target(n_dim = p,
                      log_density = function(theta) log_post_rows(theta, data),
                      normalised = FALSE)
    q0 <- proposal_mvt(p, mean = coef(mle), sigma = 9 * vcov(mle), df = 5)
    fit <- withCallingHandlers(
      fit_proxymix(tgt, N = 2L, regime = "kld", proposal = q0,
                   is_size = 3000L, max_iter = 60L, seed = r, adapt = "pmc"),
      proxymix_nonmonotone = function(cond) {
        notes <<- notes + 1L
        invokeRestart("muffleWarning")
      }
    )
    summ_px <- summarise_gmm(fit)
    ev <- gmm_evidence(fit, n = 4000L, seed = r)
  })[["elapsed"]]
  secs_ens <- system.time({
    ens <- gmm_fit_ensemble(fit, B = 80L, seed = r)
    ci <- proxy_functional_ci(ens, gmm_mean, level = 0.95)
  })[["elapsed"]]
  out_px <- score(p, r, "proxymix", summ_px, ev$log_z, secs_px, ref,
                  ref_log_z)
  out_px$covered <- ci$conf.low <= ref[, "mean"] & ref[, "mean"] <= ci$conf.high
  out_px$secs_ensemble <- secs_ens
  out_px$ess_relative <- gmm_fit_quality(fit)$ess_relative
  out_px$n_notes <- notes

  secs_lap <- system.time(
    lap <- laplace(log_post_one, coef(mle), data)
  )[["elapsed"]]
  out_lap <- score(p, r, "Laplace",
                   summarise_normal(lap$mode, sqrt(diag(lap$var))), lap$int,
                   secs_lap, ref, ref_log_z)

  df <- data.frame(y = y, X[, -1L, drop = FALSE])
  names(df) <- c("y", paste0("x", seq_len(p - 1L)))
  secs_inla <- system.time(
    fi <- inla(reformulate(names(df)[-1L], "y"), family = "binomial",
               Ntrials = 1, data = df,
               control.fixed = list(prec = 1 / prior_sd^2,
                                    prec.intercept = 1 / prior_sd^2),
               num.threads = "1:1")
  )[["elapsed"]]
  summ_inla <- cbind(
    mean = fi$summary.fixed$mean, sd = fi$summary.fixed$sd,
    lower = fi$summary.fixed[["0.025quant"]],
    upper = fi$summary.fixed[["0.975quant"]],
    p_pos = 1 - vapply(fi$marginals.fixed,
                       function(m) inla.pmarginal(0, m), numeric(1L))
  )
  out_inla <- score(p, r, "INLA", summ_inla, fi$mlik[1L], secs_inla, ref,
                    ref_log_z)

  setup <- createBayesianSetup(
    likelihood = function(theta) log_lik_one(theta, data),
    prior = createPrior(
      density = function(theta) sum(dnorm(theta, 0, prior_sd, log = TRUE)),
      sampler = function(n = 1L) matrix(rnorm(n * p, 0, prior_sd), n)
    )
  )
  set.seed(r)
  secs_bt <- system.time(
    bt <- runMCMC(setup, sampler = "DEzs",
                  settings = list(iterations = 15000L, message = FALSE))
  )[["elapsed"]]
  out_bt <- score(p, r, "DEzs", summarise_draws(getSample(bt, start = 1000L)),
                  NA_real_, secs_bt, ref, ref_log_z)

  rbind(out_nuts, out_px, out_lap, out_inla, out_bt)
}

res <- do.call(rbind, lapply(p_vec, function(p) {
  do.call(rbind, lapply(seq_len(n_rep), function(r) one_dataset(p, r)))
}))

sim_tab <- aggregate(
  cbind(err_mean, err_sd, err_lower, err_upper, err_p_pos, err_log_z,
        secs) ~ p + method,
  data = res, FUN = mean, na.action = na.pass
)
is_px <- res$method == "proxymix"
coverage <- tapply(res$covered[is_px], res$p[is_px], mean)
secs_ensemble <- tapply(res$secs_ensemble[is_px], res$p[is_px], mean)
ess_relative <- tapply(res$ess_relative[is_px], res$p[is_px], mean)
ref_ess_min <- min(res$ref_ess_min, na.rm = TRUE)
notes_per_fit <- res$n_notes[is_px & res$coef == 1L]
n_notes <- sum(notes_per_fit)
# errors per dataset, averaged over coefficients, for paired comparisons
attr(sim_tab, "per_dataset") <- aggregate(
  cbind(err_mean, err_sd, err_lower, err_upper, err_p_pos, err_log_z) ~
    p + rep + method,
  data = res, FUN = mean, na.action = na.pass
)
sim_tab
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order <- c("NUTS, 4000 draws", "proxymix", "Laplace", "INLA", "DEzs")
tab_sim <- tab_sim[order(tab_sim$p, match(tab_sim$method, method_order)), ]
tab_sim$endpoints <- (tab_sim$err_lower + tab_sim$err_upper) / 2
tab_print <- tab_sim[, c("p", "method", "err_mean", "err_sd", "endpoints",
                         "err_p_pos", "err_log_z", "secs")]
# two significant figures with trailing zeros kept, as in 1.0
tab_print$secs <- sub("\\.$", "", trimws(formatC(signif(tab_print$secs, 2L),
                                                format = "fg", digits = 2L,
                                                flag = "#")))
old_na <- options(knitr.kable.NA = "--")
knitr::kable(
  tab_print, align = c("r", "l", "r", "r", "r", "r", "r", "r"),
  digits = c(0L, 0L, 3L, 3L, 3L, 4L, 4L, 0L), row.names = FALSE,
  col.names = c("$p$", "Method", "Mean", "SD", "Endpoints",
                "$P(\\beta_j > 0)$",
                "$\\log Z$", "Seconds"),
  caption = paste0(
    "Mean error of each method against a 40 000-draw NUTS reference over ",
    sim$n_rep, " simulated logistic-regression datasets per $p$: mean, ",
    "SD and interval endpoints in units of the reference posterior SD, ",
    "averaged over the coefficients; $P(\\beta_j > 0)$ in probability; log ",
    "evidence $\\log Z$ against bridge sampling on the reference draws. ",
    "Seconds per ",
    "dataset on one core, to two significant figures; the NUTS time ",
    "includes its bridge sampling. DEzs returns no evidence estimate, so ",
    "its log-evidence cells are empty."
  )
)
options(old_na)
sim_value <- function(p, method, what) {
  tab_sim[[what]][tab_sim$p == p & tab_sim$method == method]
}
coverage_vec <- sim$coverage[as.character(sim$p_vec)]
mcse_coverage <- sqrt(coverage_vec * (1 - coverage_vec) /
                        (sim$n_rep * sim$p_vec))
sim_order <- function(p, what, digits) {
  s1 <- tab_sim$p == p
  ordered_text(tab_sim$method[s1], tab_sim[[what]][s1], digits)
}
secs_p <- function(p, method) sim_value(p, method, "secs")
# the range of monotonicity notices per fit, in words
count_word <- function(k) if (k <= 10L) n_word(k) else format(k)
notes_range <- range(sim$notes_per_fit)
notes_txt <- if (notes_range[1L] == notes_range[2L]) {
  paste(count_word(notes_range[1L]), "per fit")
} else {
  paste("from", count_word(notes_range[1L]), "to",
        count_word(notes_range[2L]), "per fit")
}
# two methods on the same datasets: mean difference in dataset-level error
# and its paired standard error
per_ds <- attr(sim$sim_tab, "per_dataset")
per_ds$endpoints <- (per_ds$err_lower + per_ds$err_upper) / 2
paired_diff <- function(p, m1, m2, what) {
  a <- per_ds[per_ds$p == p & per_ds$method == m1, ]
  b <- per_ds[per_ds$p == p & per_ds$method == m2, ]
  d <- a[[what]][order(a$rep)] - b[[what]][order(b$rep)]
  c(diff = mean(d), se = sd(d) / sqrt(length(d)))
}
# the closest method on one error, the next closest, and their paired
# difference in standard errors
lead <- function(p, what) {
  s1 <- tab_sim$p == p & !is.na(tab_sim[[what]])
  m <- tab_sim$method[s1][order(tab_sim[[what]][s1])[1:2]]
  d <- paired_diff(p, m[2L], m[1L], what)
  list(first = m[1L], second = m[2L], z = d[["diff"]] / d[["se"]])
}
lead_list <- lapply(c("err_mean", "endpoints", "err_p_pos", "err_log_z"),
                    function(what) lapply(sim$p_vec, lead, what = what))
names(lead_list) <- c("mean", "endpoints", "p_pos", "log_z")
lead_first <- function(what) {
  unique(vapply(lead_list[[what]], function(l) l$first, character(1L)))
}
lead_z <- function(what) vapply(lead_list[[what]], function(l) l$z,
                                numeric(1L))
px_nuts_mean <- lapply(sim$p_vec, paired_diff, m1 = "proxymix",
                       m2 = "NUTS, 4000 draws", what = "err_mean")
px_nuts_z <- vapply(px_nuts_mean, function(d) d[["diff"]] / d[["se"]],
                    numeric(1L))
# how many times faster the Laplace approximation is than the next method
lap_ratio <- vapply(sim$p_vec, function(p) {
  s1 <- tab_sim$p == p & tab_sim$method != "Laplace"
  min(tab_sim$secs[s1]) / secs_p(p, "Laplace")
}, numeric(1L))
```

The times in this table were measured while other jobs shared the
machine: the run took `r round(sim$elapsed_secs / 60)` minutes of wall time
for `r round(sim$cpu_secs / 60)` minutes of processor time. They compare the
methods within one value of $p$, not across the two values.

The reference's smallest bulk effective sample size over every dataset was
`r format(round(sim$ref_ess_min), big.mark = ",")`. At that size the
reference mean is itself off by about
`r fixed(sqrt(2 / pi) / sqrt(sim$ref_ess_min), 3)` posterior standard
deviations on average, and mean errors of that size are the noise of the
reference. On the posterior means INLA is closest at both sizes, at
`r fixed(sim_value(2L, "INLA", "err_mean"), 3)` and
`r fixed(sim_value(5L, "INLA", "err_mean"), 3)`, which is at that noise
floor, with the 4000-draw NUTS
run at `r fixed(sim_value(2L, "NUTS, 4000 draws", "err_mean"), 3)` and
`r fixed(sim_value(5L, "NUTS, 4000 draws", "err_mean"), 3)` and proxymix at
`r fixed(sim_value(2L, "proxymix", "err_mean"), 3)` and
`r fixed(sim_value(5L, "proxymix", "err_mean"), 3)`. The paired
difference between proxymix and the 4000-draw NUTS run on the means is
`r fixed(abs(px_nuts_z[1L]), 1)` and `r fixed(abs(px_nuts_z[2L]), 1)`
standard errors, which
`r if (all(abs(px_nuts_z) < 2)) "does not resolve the two" else "resolves the difference"`.
The Laplace approximation reports the mode, and its means are
`r fixed(sim_value(2L, "Laplace", "err_mean"), 3)` and
`r fixed(sim_value(5L, "Laplace", "err_mean"), 3)` off. On the interval
endpoints the order changes. Closest first, at `r sim$p_vec[1L]`
coefficients it is `r sim_order(sim$p_vec[1L], "endpoints", 3L)`; at
`r sim$p_vec[2L]` it is `r sim_order(sim$p_vec[2L], "endpoints", 3L)`.
INLA's upper endpoints
are further off than its lower ones, at
`r fixed(sim_value(2L, "INLA", "err_upper"), 3)` and
`r fixed(sim_value(5L, "INLA", "err_upper"), 3)` against
`r fixed(sim_value(2L, "INLA", "err_lower"), 3)` and
`r fixed(sim_value(5L, "INLA", "err_lower"), 3)`. DEzs, at
`r fixed(sim_value(5L, "DEzs", "err_mean"), 3)` on the means and
`r fixed(sim_value(5L, "DEzs", "endpoints"), 3)` on the endpoints at
`r sim$p_vec[2L]` coefficients, is the least precise of the samplers for
its run length. On the evidence proxymix is
`r fixed(sim_value(2L, "proxymix", "err_log_z"), 4)` and
`r fixed(sim_value(5L, "proxymix", "err_log_z"), 4)` off the reference,
bridge sampling on the 4000 NUTS draws
`r fixed(sim_value(2L, "NUTS, 4000 draws", "err_log_z"), 4)` and
`r fixed(sim_value(5L, "NUTS, 4000 draws", "err_log_z"), 4)`, the Laplace
approximation `r fixed(sim_value(2L, "Laplace", "err_log_z"), 4)` and
`r fixed(sim_value(5L, "Laplace", "err_log_z"), 4)`, and INLA
`r fixed(sim_value(2L, "INLA", "err_log_z"), 4)` and
`r fixed(sim_value(5L, "INLA", "err_log_z"), 4)`. proxymix's 95 per cent
ensemble intervals on the posterior means contain the reference mean in
`r fixed(sim$coverage[["2"]], 2)` of cases at `r sim$p_vec[1L]` coefficients
and `r fixed(sim$coverage[["5"]], 2)` at `r sim$p_vec[2L]`, over
`r sim$n_rep * sim$p_vec[1L]` and `r sim$n_rep * sim$p_vec[2L]` intervals,
with Monte Carlo standard errors of about `r fixed(mcse_coverage[1L], 3)`
and `r fixed(mcse_coverage[2L], 3)`, larger if the intervals within one
dataset are correlated; the ensemble took a further
`r fixed(sim$secs_ensemble[["5"]], 1)` s per dataset at `r sim$p_vec[2L]`
coefficients. The fit's relative effective sample size averaged
`r fixed(sim$ess_relative[["2"]], 2)` and `r fixed(sim$ess_relative[["5"]], 2)`,
and the run raised `r sim$n_notes` monotonicity notices from population
Monte Carlo over the `r sim$n_rep * length(sim$p_vec)` fits,
`r notes_txt`.
The simulation code counts these notices; the two-parameter example above
prints its notice.

In this run `r method_label[lead_first("endpoints")]` came out ahead on
the interval endpoints and on the log evidence at both sizes, by at least
`r floor(min(lead_z("endpoints"), lead_z("log_z")))` paired standard errors
of the difference from the next closest method, and its ensemble intervals
on the posterior means covered the reference at close to their nominal
rate at five coefficients and below it at two.
`r method_label[lead_first("mean")]` came out ahead on the posterior means
at both sizes, by at least `r floor(min(lead_z("mean")))` paired standard
errors. It was also closest on the tail probabilities, by
`r fixed(lead_z("p_pos")[1L], 1)` paired standard errors at
`r sim$p_vec[1L]` coefficients and `r fixed(lead_z("p_pos")[2L], 1)` at
`r sim$p_vec[2L]`; `r c("neither is", "only the first is", "only the second is", "both are")[1L + sum(c(1L, 2L) * (lead_z("p_pos") > 2))]`
resolved. The Laplace
approximation was the fastest, `r round(min(lap_ratio))` to
`r round(max(lap_ratio))` times faster than the next method, with an
evidence error close to INLA's. The comparison covers logistic regression
with standard normal covariates at one sample size and two dimensions under
a single weak prior. It does not cover skewed or multimodal posteriors,
separated data, or dimensions above five, the lower end of the five to ten
parameters beyond which the Limitations section places the
importance-sampling floor.

## Further reading

*Fitting a proxy to a density you cannot sample* is the shorter first pass
over the same fitting verb, on a target whose shape is known in advance.
*Choosing between the three fitting regimes* explains why an unnormalised
posterior lands in regime (iii) and what the other two regimes need instead.
*The closed-form operator calculus on a mixture* covers the operators that
make the compressed posterior worth having: pushforward, observation update,
marginalisation and conditioning. *Reading the entropy of a fitted mixture*
adds the entropy and divergence diagnostics that summarise a fitted
posterior's spread.

## References

* Albert, J. (2009). *Bayesian Computation with R.* Second edition.
  Springer. <doi:10.1007/978-0-387-92298-0>.
* Ardia, D., Hoogerheide, L. F. and van Dijk, H. K. (2009). *Adaptive
  mixture of Student-t distributions as a flexible candidate distribution
  for efficient simulation: The R package AdMit.* Journal of Statistical
  Software 29(3), 1--32. <doi:10.18637/jss.v029.i03>.
* Cappé, O., Douc, R., Guillin, A., Marin, J.-M. and Robert, C. P.
  (2008). *Adaptive importance sampling in general mixture classes.*
  Statistics and Computing 18, 447--459. <doi:10.1007/s11222-008-9059-x>.
* Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
  Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A. (2017).
  *Stan: A probabilistic programming language.* Journal of Statistical
  Software 76(1), 1--32. <doi:10.18637/jss.v076.i01>.
* Fair, R. C. (1978). *A theory of extramarital affairs.* Journal of
  Political Economy 86(1), 45--61. <doi:10.1086/260629>.
* Gabry, J., Češnovar, R., Johnson, A. and Bronder, S. (2025). *cmdstanr:
  R interface to CmdStan.* R package version 0.9.0.
  <https://mc-stan.org/cmdstanr/>.
* Gronau, Q. F., Singmann, H. and Wagenmakers, E.-J. (2020).
  *bridgesampling: An R package for estimating normalizing constants.*
  Journal of Statistical Software 92(10), 1--29.
  <doi:10.18637/jss.v092.i10>.
* Hartig, F., Minunno, F. and Paul, S. (2026). *BayesianTools:
  General-purpose MCMC and SMC samplers and tools for Bayesian statistics.*
  R package version 0.1.9. <doi:10.32614/CRAN.package.BayesianTools>.
* Kleiber, C. and Zeileis, A. (2008). *Applied Econometrics with R.*
  Springer. <doi:10.1007/978-0-387-77318-6>.
* Owen, A. and Zhou, Y. (2000). *Safe and effective importance sampling.*
  Journal of the American Statistical Association 95(449), 135--143.
  <doi:10.1080/01621459.2000.10473909>.
* Rubin, D. B. (1981). *The Bayesian bootstrap.* The Annals of Statistics
  9(1), 130--134. <doi:10.1214/aos/1176345338>.
* Rue, H., Martino, S. and Chopin, N. (2009). *Approximate Bayesian
  inference for latent Gaussian models by using integrated nested Laplace
  approximations.* Journal of the Royal Statistical Society: Series B
  71(2), 319--392. <doi:10.1111/j.1467-9868.2008.00700.x>.
* ter Braak, C. J. F. and Vrugt, J. A. (2008). *Differential Evolution
  Markov Chain with snooker updater and fewer chains.* Statistics and
  Computing 18, 435--446. <doi:10.1007/s11222-008-9104-9>.
* Tierney, L. and Kadane, J. B. (1986). *Accurate approximations for
  posterior moments and marginal densities.* Journal of the American
  Statistical Association 81(393), 82--86.
  <doi:10.1080/01621459.1986.10478240>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.

## Reproduce

The ambient seed is 20260705. Each stochastic step additionally carries its
own seed, so none of them depends on the ambient random-number state: the
fit uses `seed = 1L`, the evidence estimate `seed = 2L`, and the bootstrap
ensemble `seed = 3L`. The quadrature reference is deterministic.

In the Affairs illustration the proxymix fit, evidence and ensemble carry
seeds 1, 2 and 3 again, DEzs runs after `set.seed(4L)`, the reference NUTS
run uses `seed = 1L` and the 4000-draw run `seed = 1L + 100000L`, so the
two runs are independent. The stored reference comes from the chunks shown, run on
`r if (run_illustration) ref$run_date else "the date recorded in the stored file"`
with cmdstanr
`r if (run_illustration) desc_version("cmdstanr", ref$versions[["cmdstanr"]]) else ""`,
CmdStan `r if (run_illustration) ref$versions[["CmdStan"]] else ""` and
bridgesampling
`r if (run_illustration) desc_version("bridgesampling", ref$versions[["bridgesampling"]]) else ""`;
the long run took
`r if (run_illustration) round(ref$long$secs_sample) else ""` s and its
largest split R-hat was
`r if (run_illustration) fixed(ref$long$rhat_max, 3) else ""`. In the
simulation each dataset is generated after `set.seed()` with its own
index, which also seeds the reference NUTS run, the proxymix fit, its
evidence and ensemble, and DEzs; the 4000-draw NUTS run takes the index
plus 100000 as its seed. The stored simulation results come from the code shown,
run on `r sim$run_date` with R `r sim$versions[["R"]]`, proxymix
`r desc_version("proxymix", sim$versions[["proxymix"]])`, cmdstanr
`r desc_version("cmdstanr", sim$versions[["cmdstanr"]])`, CmdStan
`r sim$versions[["CmdStan"]]`, LearnBayes
`r desc_version("LearnBayes", sim$versions[["LearnBayes"]])`, INLA
`r desc_version("INLA", sim$versions[["INLA"]])`, bridgesampling
`r desc_version("bridgesampling", sim$versions[["bridgesampling"]])` and
BayesianTools
`r desc_version("BayesianTools", sim$versions[["BayesianTools"]])`, and the run raised
`r length(sim$warnings)` warnings.

```{r session-info}
old_opt <- options(width = 72L)
si_vec <- trimws(capture.output(sessionInfo()), "right")
options(old_opt)
# library paths are single long words, so a long line breaks after a slash
long_vec <- nchar(si_vec) > 72L
si_vec[long_vec] <- gsub("(.{1,68}/)(?=.{8,})", "\\1\n        ",
                         si_vec[long_vec], perl = TRUE)
writeLines(si_vec)
```
