---
title: "Fitting a proxy to a density you cannot sample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a proxy to a density you cannot sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render() with ggplot2 installed, the
     simulation read from its stored results;
     macOS arm64 (Apple silicon), R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
# the figure below is drawn only when ggplot2 is installed
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r fixed-notation, include = FALSE}
## Small errors are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
```

```{=latex}
\providecommand{\pmNeedspace}[1]{\par\penalty-100\ifdim\dimexpr#1\relax>\dimexpr\pagegoal-\pagetotal\relax\newpage\fi}
```

## The problem

An analyst arrives with a function and no sampler. *I can write down a
function that returns the log-density of the thing I care about, and I have
no way to draw from it. Can I get a compact object I am able to sample,
marginalise and condition -- and how would I know the fit is any good?*
Both halves matter. The first is what the package does; the second is what
stops a plausible-looking mixture being trusted when the numerical
machinery behind it has quietly failed.

## Package capabilities

A `gmm_target()` is the target: a dimension, a `log_density` callable, and
optionally samples. The built-in `banana_target()` used below carries an
exact, normalised log-density and deliberately no samples, so only the
evaluate-only regime applies to it.

`fit_proxymix()` is the single fitting verb. It routes to one of three
regimes on what the target can supply, and the table under *Addressing the
problem* states the routing rule. The regime for a target that can only be
evaluated is `"kld"`: importance-sampled KLD-EM, which needs only
point-wise evaluation of the target.

A `proposal_mvt()` is the importance proposal that regime feeds on: a
multivariate Student-t whose heavy tails keep the self-normalised weights
from collapsing onto a handful of draws.

`gmm_fit_quality()` is the fit's certificate: whether the iteration
converged, whether the importance sample degenerated, the effective sample
size, the largest single weight, and the held-out validation gap. It
answers the second half of the question, and it travels with the fit
through every operation downstream.

`dgmm()`, `rgmm()`, `gmm_marginalise()` and `gmm_conditionalise()` are the
closed-form reads off the fitted mixture: density, sampling, integrating a
coordinate out, and fixing one to a value. None of them re-evaluates the
target.

## Addressing the problem

### The routing rule

```{r regime-table, echo = FALSE}
regime_tbl <- data.frame(
  Regime = c("`\"moment\"`", "`\"sample\"`", "`\"kld\"`"),
  Applies = c(
    "`N == 1` and the target carries samples or moments",
    "`N >= 2` and the target carries samples",
    "the target carries `log_density` only"
  ),
  Method = c(
    "Closed-form moment matching",
    "Classical expectation-maximisation",
    "Importance-sampled KLD-EM"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  regime_tbl,
  caption = paste(
    "How `fit_proxymix()` routes under `regime = \"auto\"`. The three",
    "regimes are those of van der Hoek and Elliott (2024)."
  )
)
```

Only the third row is available for a target that cannot be sampled, and
that is the setting the sample-based mixture packages on CRAN (`mclust`,
Scrucca et al., 2016; `mixtools`, Benaglia et al., 2009; `flexmix`, Grün
and Leisch, 2008) do not address: all three of them require independent
draws.

### A target with no sampler

The bundled banana target is a non-Gaussian shape in the plane, obtained by
warping an isotropic Gaussian through a quadratic. Its log-density is exact
and normalised. Samples are not attached, so the routing table leaves one
option.

```{r banana}
tgt <- banana_target()
tgt
```

### Fit the proxy

A three-component proxy by KLD-EM, with a Student-t proposal on five
degrees of freedom.

```{r fit}
proposal <- proposal_mvt(n_dim = 2L, mean = c(0, 0),
                         sigma = 4 * diag(2), df = 5)
fit <- fit_proxymix(tgt, N = 3L, regime = "kld",
                    proposal = proposal,
                    is_size = 2000L,
                    max_iter = 60L,
                    seed = 1L)
fit
```

```{=latex}
\pmNeedspace{20\baselineskip}
```

### Read the certificate before anything else

```{r certificate}
quality <- gmm_fit_quality(fit)
```

```{r certificate-table, echo = FALSE}
cert_tbl <- data.frame(
  Field = c("regime", "converged", "degenerate", "effective sample size",
            "ESS as a fraction of draws", "smallest per-component ESS",
            "largest single weight", "KL estimate, in-sample",
            "KL estimate, held-out (standard error)",
            "validation gap, held-out minus in-sample"),
  Value = c(
    quality$regime,
    as.character(quality$converged),
    as.character(quality$degenerate),
    format(round(quality$ess, 1L), nsmall = 1L),
    format(round(quality$ess_relative, 3L), nsmall = 3L),
    format(round(quality$min_component_ess, 1L), nsmall = 1L),
    format(signif(quality$max_weight, 3L)),
    format(signif(quality$kld_final, 3L)),
    paste0(fixed(fit@diagnostics$validation_kld, 4), " (",
           fixed(fit@diagnostics$validation_mc_se, 4), ")"),
    format(signif(quality$validation_gap, 3L))
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  cert_tbl,
  caption = paste(
    "The fit certificate and the held-out KL estimate."
  )
)
```

The in-sample KL estimate is computed on the draws the fit was tuned to,
so it reads lower than the held-out estimate, which the fit computes on
`r format(fit@diagnostics$validation_size, big.mark = ",")` fresh draws.

### Overlay the proxy on the target

```{r overlay-grid}
grid_x <- seq(-3, 3, length.out = 120L)
grid_g <- expand.grid(x1 = grid_x, x2 = grid_x)
grid_mat <- as.matrix(grid_g)
grid_g$target <- exp(tgt@log_density(grid_mat))
grid_g$proxy <- dgmm(grid_mat, fit)
```

```{r overlay, eval = has_ggplot2, echo = has_ggplot2, fig.cap = sprintf("The banana target (filled contours) with the fitted three-component proxy overlaid as dashed contours. The proxy contours bend along the ridge. Its held-out KL estimate is %s (Monte Carlo standard error %s), from an effective sample size of %s draws out of %d.", fixed(fit@diagnostics$validation_kld, 4), fixed(fit@diagnostics$validation_mc_se, 4), format(round(fit@diagnostics$ess, 0L)), fit@diagnostics$is_size), fig.alt = "Filled contour map of the curved banana density with dashed contours of the three-component Gaussian-mixture proxy following the same ridge."}
ggplot2::ggplot(grid_g, ggplot2::aes(x1, x2)) +
  ggplot2::geom_contour_filled(ggplot2::aes(z = target), bins = 10L,
                               alpha = 0.85) +
  ggplot2::geom_contour(ggplot2::aes(z = proxy), colour = "white",
                        linetype = "dashed", linewidth = 0.45, bins = 8L) +
  ggplot2::scale_fill_viridis_d(option = "mako", guide = "none") +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Target (filled) and fitted proxy (dashed)",
    x = expression(x[1]), y = expression(x[2])
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r overlay-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the target-and-proxy",
    "overlay figure is skipped.\n")
```

### The closed-form reads

Because the fit is a Gaussian mixture, integrating a coordinate out and
fixing one to a value are exact algebra, not further target evaluations.

```{r operations}
marg <- gmm_marginalise(fit, keep = 1L)
cond <- gmm_conditionalise(fit, given = c(NA, 0.5))
# weight, mean and standard deviation of each component of a 1-D mixture
components_1d <- function(g) {
  data.frame(weight = gmm_weights(g),
             mean = vapply(gmm_means(g), `[[`, numeric(1L), 1L),
             sd = sqrt(vapply(gmm_covariances(g), `[[`, numeric(1L), 1L)))
}
round(components_1d(marg), 4L)
round(components_1d(cond), 4L)
```

The first table is the proxy's density of $x_1$; the second is its density
of $x_1$ given $x_2 = 0.5$.

Draws from the proxy are exact and cheap, while the target has no sampler.

```{r sample}
set.seed(1L)
draws <- rgmm(500L, fit)
dim(draws)
```

## Interpretation

The fit answers the first half of the question. The proxy is a
`r gmm_n_components(fit)`-component mixture in
`r gmm_dim(fit)` dimensions, reached in
`r length(kld_trace(fit))` KLD-EM iterations, and it supports sampling,
marginalisation and conditioning with no further access to the target at
all. In the overlay the proxy's contours bend along the ridge, which the
elliptical contours of a single Gaussian cannot do, and that is the reason
for fitting a mixture instead of a single Gaussian.

The certificate answers the second half. The iteration converged, and the
fit is not flagged as degenerate. The importance sample retained an
effective size of `r round(fit@diagnostics$ess, 0L)` out of
`r format(fit@diagnostics$is_size, big.mark = ",")` draws, a retention of
`r round(100 * fit@diagnostics$ess_relative, 1L)` per cent, and the largest
single self-normalised weight is
`r format(signif(fit@diagnostics$max_weight, 2L))`, so no individual draw
is carrying the fit. The smallest per-component effective sample size is
`r round(quality$min_component_ess, 0L)`, so every one of the three
components was estimated from a substantial share of the draws rather than
from a handful. The held-out KL estimate is
`r fixed(fit@diagnostics$validation_kld, 4)` with a Monte Carlo standard
error of `r fixed(fit@diagnostics$validation_mc_se, 4)`. The in-sample
estimate of `r fixed(quality$kld_final, 4)` is lower by the validation gap
of `r fixed(quality$validation_gap, 4)`, which is about
`r round(abs(quality$validation_gap) / fit@diagnostics$validation_mc_se)`
standard errors of the held-out estimate.

Read the certificate in that order -- converged, degenerate, effective
sample size, largest weight, validation gap -- and a fit that has failed
announces itself before any downstream number is taken seriously.

## Limitations

This vignette fits one target in two dimensions and reads the fit that
results. It does not choose the number of components, which is left at
three by hand here; `select_N()` and `bic_aic()` do that job, and a
component count too low for the shape shows up as a residual KL that more
importance draws will not reduce. It also says nothing about how the choice
of proposal changes the answer. The Student-t proposal used above is wide
enough to cover the banana and heavy-tailed enough to keep the weights
spread, and a proposal that misses the target's mass produces a fit whose
certificate is worthless even though the printed mixture looks ordinary.

The effective sample size reported here is a two-dimensional number.
Importance sampling degrades quickly with dimension, and a proxy of the
same nominal quality in five or ten dimensions demands far more draws for
the same retention. The certificate is what makes that degradation visible
rather than silent, and it is reported on every fit in the `"kld"` regime.

Nothing above compares the proxy with the other ways of getting at a
density that can be evaluated but not sampled. The numerical illustration
that follows places it beside a Laplace approximation, a Markov chain Monte
Carlo run and a mixture fitted to that run's draws, all scored against the
same quadrature of the target.


## Numerical illustration

There is no dataset in this vignette: a target that can only be evaluated
has no sample by definition, so the illustration is a simulation.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("quickstart_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
```

```{r fit-quadrature, include = FALSE}
# the three-component fit from above, scored on the simulation's grid
quad_grid <- as.matrix(expand.grid(x1 = seq(-5 + sim$h / 2, 5, by = sim$h),
                                   x2 = seq(-5 + sim$h / 2, 12, by = sim$h)))
quad_log_f <- tgt@log_density(quad_grid)
kl_fit_quad <- sum(exp(quad_log_f) *
                     (quad_log_f - dgmm(quad_grid, fit, log = TRUE))) * sim$h^2
quad_mass <- sum(exp(quad_log_f)) * sim$h^2
```

The simulation fits the banana target `r sim$n_rep` times by each of five
routes and scores every fit against the same reference: a midpoint-rule
quadrature of the exact target density on a grid of step `r sim$h` over
$[-5, 5] \times [-5, 12]$, whose cells sum to a mass of
`r fixed(quad_mass, 6)`. Two things are scored. The first is
the Kullback-Leibler divergence
$\mathrm{KL}(f \,\|\, g) = \int f \log (f / g)$, with $f$ the target and
$g$ the fitted density, computed on the grid, which only the
routes that return a density can be given. The second is the mass in the
tail $P(x_1 > 2)$, which the quadrature puts at
`r fixed(sim$tail_ref, 5)` against the exact value
$1 - \Phi(2) = `r fixed(1 - pnorm(2), 5)`$, and which a sampler answers by
counting its draws. Each route's
wall time is recorded, and so is the number of values the user had to set
in the code shown, other than a seed.

The five routes are proxymix with the number of components chosen by
`select_N()` at its defaults; a Laplace approximation, the single Gaussian
whose mean is the mode and whose covariance is the inverse curvature there
(Tierney and Kadane, 1986); the no-U-turn sampler (NUTS; Hoffman and
Gelman, 2014) of Stan through cmdstanr at its defaults, four chains of 1000
warm-up and 1000 sampling iterations run in sequence (Carpenter et al.,
2017); mclust fitted to those NUTS draws (Scrucca et al., 2016), with
`G = 1:6`, the candidate set `select_N()` uses, and the unconstrained
covariance family, which is the family proxymix fits; and the
differential-evolution sampler DEzs (ter Braak and Vrugt, 2008) of
BayesianTools (Hartig et al., 2026) at its default of 10,000 iterations,
which version `r sim$versions[["BayesianTools"]]` spreads over three
internal chains, given a box prior wide enough to hold all of the target's
mass, with the first 999 iterations of each chain discarded. The Laplace route is one call to
`optim()` and a matrix inverse, and its density is evaluated with
`dmvnorm()` from mclust. The Stan program is compiled once, in the run that produced the stored
results, and not when this vignette renders.

mclust and BayesianTools install from CRAN. cmdstanr is not on CRAN and
installs from the Stan R-universe repository, stan-dev.r-universe.dev. It
also needs a CmdStan installation, which `cmdstanr::install_cmdstan()`
provides:

```{r install-competitors, eval = FALSE}
install.packages(c("mclust", "BayesianTools"))
install.packages("cmdstanr",
                 repos = c("https://stan-dev.r-universe.dev",
                           getOption("repos")))
```

This chunk is complete and runs as shown, but it took about
`r round(sim$elapsed_secs / 60)` minutes on one core (R
`r sim$versions[["R"]]`, Apple silicon); the results below are read from
its stored output. Changing `n_rep <- 200L` to `n_rep <- 10L` gives a
reduced run of about `r round(10 * sim$elapsed_secs / sim$n_rep)`
seconds plus the compilation of the Stan program.

```{r sim-code, eval = FALSE}
library(proxymix)
library(cmdstanr)
library(mclust)
library(BayesianTools)

n_rep <- 200L  # replicates

tgt <- banana_target()

# midpoint-rule quadrature; the target's mass outside the box is below 1e-5
h <- 0.05
grid <- as.matrix(expand.grid(x1 = seq(-5 + h / 2, 5, by = h),
                              x2 = seq(-5 + h / 2, 12, by = h)))
log_f <- tgt@log_density(grid)
f_grid <- exp(log_f)
tail_ref <- sum(f_grid[grid[, 1L] > 2]) * h^2
kl_of <- function(log_g) sum(f_grid * (log_f - log_g)) * h^2

# mass above x1 = 2 under a Gaussian mixture, from its x1 marginal
tail_of <- function(w, mean1, sd1) {
  sum(w * pnorm(2, mean1, sd1, lower.tail = FALSE))
}

# the same target as a Stan program, compiled once for the whole run
stan_file <- file.path(tempdir(), "banana.stan")
writeLines(c(
  "parameters {",
  "  vector[2] x;",
  "}",
  "model {",
  "  target += -0.5 * (square(x[1])",
  "                    + square(x[2] - 0.5 * (square(x[1]) - 1)));",
  "}"), stan_file)
banana_stan <- cmdstan_model(stan_file)

box <- createBayesianSetup(likelihood = function(x) tgt@log_density(x),
                           lower = c(-6, -6), upper = c(6, 14))

# values set by the user in this code, other than seeds
tuning <- c(proxymix = 0L, Laplace = 2L, NUTS = 0L,
            "NUTS + mclust" = 2L, DEzs = 3L)

timed <- function(expr) {
  t0 <- proc.time()[[3L]]
  value <- expr
  list(value = value, secs = proc.time()[[3L]] - t0)
}

one_replicate <- function(r) {
  set.seed(r)
  px <- timed(select_N(tgt, seed = r))
  fit <- px$value$best_fit
  fit_1 <- gmm_marginalise(fit, keep = 1L)

  la <- timed(optim(c(0, 0), function(x) -tgt@log_density(x),
                    method = "BFGS", hessian = TRUE))
  la_mean <- la$value$par
  la_cov <- solve(la$value$hessian)

  nuts <- timed(banana_stan$sample(seed = r, refresh = 0L,
                                   show_messages = FALSE))
  draws <- nuts$value$draws("x", format = "matrix")
  n_divergent <- sum(nuts$value$diagnostic_summary(
    "divergences", quiet = TRUE)$num_divergent)

  mc <- timed(Mclust(draws, G = 1:6, modelNames = "VVV", verbose = FALSE))
  mc_par <- mc$value$parameters

  de <- timed(runMCMC(box, sampler = "DEzs",
                      settings = list(message = FALSE)))
  de_draws <- getSample(de$value, start = 1000L)

  data.frame(
    rep = r,
    method = names(tuning),
    kl = c(kl_of(dgmm(grid, fit, log = TRUE)),
           kl_of(dmvnorm(grid, la_mean, la_cov, log = TRUE)),
           NA,
           kl_of(dens(grid, mc$value$modelName, parameters = mc_par,
                      logarithm = TRUE)),
           NA),
    tail = c(tail_of(gmm_weights(fit_1),
                     vapply(gmm_means(fit_1), `[[`, numeric(1L), 1L),
                     sqrt(vapply(gmm_covariances(fit_1), `[[`,
                                 numeric(1L), 1L))),
             pnorm(2, la_mean[1L], sqrt(la_cov[1L, 1L]), lower.tail = FALSE),
             mean(draws[, 1L] > 2),
             tail_of(mc_par$pro, mc_par$mean[1L, ],
                     sqrt(mc_par$variance$sigma[1L, 1L, ])),
             mean(de_draws[, 1L] > 2)),
    secs = c(px$secs, la$secs, nuts$secs, nuts$secs + mc$secs, de$secs),
    n_components = c(px$value$best_n, 1L, NA, mc$value$G, NA),
    divergent = c(NA, NA, n_divergent, NA, NA)
  )
}

res <- do.call(rbind, lapply(seq_len(n_rep), one_replicate))
res$tail_error <- res$tail - tail_ref

sim_tab <- do.call(rbind, lapply(split(res, res$method), function(d) {
  data.frame(method = d$method[1L],
             kl_mean = mean(d$kl), kl_sd = sd(d$kl),
             tail_bias = mean(d$tail_error),
             tail_rmse = sqrt(mean(d$tail_error^2)),
             # delta-method standard error of the root mean squared error
             tail_rmse_se = sd(d$tail_error^2) / sqrt(nrow(d)) /
               (2 * sqrt(mean(d$tail_error^2))),
             secs = median(d$secs),
             tuning = tuning[[d$method[1L]]],
             divergent_share = mean(d$divergent > 0),
             divergent_max = max(d$divergent))
}))
sim_tab
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order <- names(sim$tuning)
tab_sim <- tab_sim[match(method_order, tab_sim$method), ]
tab_sim$kl <- ifelse(is.na(tab_sim$kl_mean), "--",
                     paste0(fixed(tab_sim$kl_mean, 4), " (",
                            fixed(tab_sim$kl_sd, 4), ")"))
tab_sim$tail_bias <- fixed(tab_sim$tail_bias, 4)
tab_sim$tail_rmse <- fixed(tab_sim$tail_rmse, 4)
tab_sim$secs <- fixed(tab_sim$secs, 2)
knitr::kable(
  tab_sim[, c("method", "kl", "tail_bias", "tail_rmse", "secs", "tuning")],
  row.names = FALSE,
  align = c("l", "r", "r", "r", "r", "r"),
  col.names = c("Route", "KL, mean (sd)", "Tail bias", "Tail RMSE",
                "Seconds", "Values set"),
  caption = paste0(
    "Kullback-Leibler divergence of the fitted density from the target, ",
    "bias and root mean squared error of the estimated mass above ",
    "$x_1 = 2$, median wall time and the number of values the user sets, ",
    "over ", sim$n_rep, " fits of the banana target, each scored against ",
    "the grid quadrature of the exact target. NUTS and DEzs return draws ",
    "and no density, so their KL cells are empty. A value set is an ",
    "argument other than the target and a seed that can change the ",
    "result: the Laplace starting point and optimiser, the mclust ",
    "candidate range and covariance family, and the DEzs prior bounds ",
    "and burn-in."
  )
)
sim_value <- function(method, what) {
  sim$sim_tab[[what]][sim$sim_tab$method == method]
}
# unpaired standard error of a difference in mean divergence
kl_gap_se <- function(a, b) {
  sqrt((sim_value(a, "kl_sd")^2 + sim_value(b, "kl_sd")^2) / sim$n_rep)
}
kl_gap <- sim_value("NUTS + mclust", "kl_mean") - sim_value("proxymix", "kl_mean")
# the smallest gap in tail error from proxymix, in standard errors of the gap
rmse_z <- vapply(c("NUTS", "NUTS + mclust", "DEzs"), function(m) {
  (sim_value(m, "tail_rmse") - sim_value("proxymix", "tail_rmse")) /
    sqrt(sim_value(m, "tail_rmse_se")^2 +
           sim_value("proxymix", "tail_rmse_se")^2)
}, numeric(1L))
# each sampling route's bias in standard errors, the spread taken from bias
# and RMSE; the Laplace route is deterministic and has no standard error
bias_z <- with(sim$sim_tab, abs(tail_bias) /
                 sqrt(pmax(tail_rmse^2 - tail_bias^2, 0) / sim$n_rep))
bias_z <- bias_z[sim$sim_tab$method != "Laplace"]
n_chosen <- table(factor(sim$n_components[["proxymix"]], levels = 1:6))
n_mode <- names(which.max(n_chosen))
share_mode <- n_chosen[[n_mode]] / sim$n_rep
```

Of the three routes that return a density, proxymix has the smallest
divergence from the target, `r fixed(sim_value("proxymix", "kl_mean"), 4)`
nats on average against
`r fixed(sim_value("NUTS + mclust", "kl_mean"), 4)` for the mixture fitted
to the NUTS draws, a gap of `r fixed(kl_gap, 4)` with an unpaired standard
error of `r fixed(kl_gap_se("proxymix", "NUTS + mclust"), 4)`, and the Laplace approximation is far behind at
`r fixed(sim_value("Laplace", "kl_mean"), 3)`, because one Gaussian cannot
bend along the ridge. On the same grid the three-component fit from
*Addressing the problem* has a divergence of `r fixed(kl_fit_quad, 4)`.
Its held-out estimate of `r fixed(fit@diagnostics$validation_kld, 4)` is
`r if (abs(kl_fit_quad - fit@diagnostics$validation_kld) < 2 * fit@diagnostics$validation_mc_se) "within" else "outside"`
two Monte Carlo standard errors of that value. Its in-sample estimate of
`r fixed(fit@diagnostics$kld_final, 4)` is lower than the quadrature value
by a factor of `r fixed(kl_fit_quad / fit@diagnostics$kld_final, 1)`.
`select_N()` chose `r n_mode` components in
`r round(100 * share_mode)` per cent of the replicates. On the tail mass
no sampling route's bias is more than
`r if (max(bias_z) < 2) "two" else ceiling(max(bias_z))` of its standard
errors from zero, and the routes differ in spread. The
proxy's root mean squared error is
`r fixed(sim_value("proxymix", "tail_rmse"), 4)`, below the
`r fixed(sim_value("NUTS", "tail_rmse"), 4)` of NUTS counting its 4000
draws, the `r fixed(sim_value("NUTS + mclust", "tail_rmse"), 4)` of the
two-stage mixture and the `r fixed(sim_value("DEzs", "tail_rmse"), 4)` of
DEzs, and the smallest of these gaps is
`r fixed(min(rmse_z), 1)` standard errors of the difference, from the
delta-method standard error of each root mean squared error. The Laplace approximation gets this tail right, at an error of
`r fixed(sim_value("Laplace", "tail_rmse"), 5)`, the size of the
quadrature's own difference from $1 - \Phi(2)$, because the mode and
curvature of the banana in $x_1$ are those of the standard normal that is
its exact marginal. cmdstanr reported divergent transitions in
`r round(sim_value("NUTS", "divergent_share") * sim$n_rep)` of the
`r sim$n_rep` NUTS runs, at most
`r sim_value("NUTS", "divergent_max")` of the 4000 transitions in any one
run, as a printed message rather than an R warning, and the draws were kept
as they were. A
proxymix fit took a median of `r fixed(sim_value("proxymix", "secs"), 2)`
seconds, a NUTS run `r fixed(sim_value("NUTS", "secs"), 2)`, the two-stage
mixture `r fixed(sim_value("NUTS + mclust", "secs"), 2)` and DEzs
`r fixed(sim_value("DEzs", "secs"), 2)`; the machine was running other jobs
during the run, so these times compare with each other and not with an
idle core. The proxymix route set no values beyond the target; the Laplace
route needed a starting point and an optimiser, NUTS needed the target
rewritten as a Stan program, the two-stage route added the candidate range
and covariance family for mclust, and DEzs needed the prior box and a
burn-in.

In this run proxymix gave the smallest divergence and, after the Laplace
approximation, the smallest tail error, from one call with no sampler. The
Laplace approximation was exact for this
tail and the fastest. NUTS needed no values set beyond the target written
as a Stan program, and DEzs needed three. The comparison is one
two-dimensional target with a known normaliser and a single tail
probability; it says nothing about higher dimensions, multimodal targets,
unnormalised posteriors, or tails further out than $x_1 = 2$, and none of
the samplers was tuned beyond its defaults.

## Further reading

*Choosing between the three fitting regimes* takes the routing table above
and works all three regimes end to end on a target whose ground truth is
known, so the cost of choosing the wrong one is visible.

*How well a mixture proxies four awkward shapes* pushes the `"kld"` regime across
a bent ridge, an annulus, a well-separated mixture and a compactly
supported density, and shows the package refusing a fit whose importance
sample has collapsed.

*The closed-form operator calculus on a mixture* takes the marginalisation
and conditioning shown above and adds pushforward through a linear sensor,
the Bayesian update on a noisy observation, and filtering over time.

*Compressing a Bayesian posterior you can evaluate but not sample* is this
workflow applied to an unnormalised log-posterior, with the model evidence
and error bars on the fit itself.


## References

* Benaglia, T., Chauveau, D., Hunter, D. R. and Young, D. S. (2009).
  *mixtools: An R package for analyzing finite mixture models.* Journal of
  Statistical Software 32(6), 1--29. <doi:10.18637/jss.v032.i06>.
* Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
  Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A. (2017).
  *Stan: A probabilistic programming language.* Journal of Statistical
  Software 76(1), 1--32. <doi:10.18637/jss.v076.i01>.
* Grün, B. and Leisch, F. (2008). *FlexMix version 2: Finite mixtures
  with concomitant variables and varying and constant parameters.* Journal
  of Statistical Software 28(4), 1--35. <doi:10.18637/jss.v028.i04>.
* Hartig, F., Minunno, F. and Paul, S. (2026). *BayesianTools:
  General-purpose MCMC and SMC samplers and tools for Bayesian statistics.*
  R package version `r sim$versions[["BayesianTools"]]`.
  <doi:10.32614/CRAN.package.BayesianTools>.
* Hoffman, M. D. and Gelman, A. (2014). *The No-U-Turn sampler: Adaptively
  setting path lengths in Hamiltonian Monte Carlo.* Journal of Machine
  Learning Research 15(47), 1593--1623.
  <https://jmlr.org/papers/v15/hoffman14a.html>.
* Scrucca, L., Fop, M., Murphy, T. B. and Raftery, A. E. (2016). *mclust 5:
  Clustering, classification and density estimation using Gaussian finite
  mixture models.* The R Journal 8(1), 289--317.
  <doi:10.32614/RJ-2016-021>.
* ter Braak, C. J. F. and Vrugt, J. A. (2008). *Differential evolution
  Markov chain with snooker updater and fewer chains.* Statistics and
  Computing 18(4), 435--446. <doi:10.1007/s11222-008-9104-9>.
* Tierney, L. and Kadane, J. B. (1986). *Accurate approximations for
  posterior moments and marginal densities.* Journal of the American
  Statistical Association 81(393), 82--86.
  <doi:10.1080/01621459.1986.10478240>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  <doi:10.1080/07362994.2024.2372605>.

## Reproduce

The fit above is seeded through `fit_proxymix(seed = 1L)`, and the
sampling chunk draws from it after `set.seed(1L)`, so every number above the
numerical illustration is reproducible from the package and the shipped
target.

In the simulation each replicate starts with `set.seed()` on its own index,
which the same index also passes to `select_N()` and to the Stan sampler.
The stored simulation results come from the code shown, run on
`r sim$run_date` with R `r sim$versions[["R"]]`, proxymix
`r sim$versions[["proxymix"]]`, cmdstanr `r sim$versions[["cmdstanr"]]`
over CmdStan `r sim$versions[["CmdStan"]]`, mclust
`r sim$versions[["mclust"]]` and BayesianTools
`r sim$versions[["BayesianTools"]]`, and the run raised
`r length(sim$warnings)` R warnings.

```{r session-info}
old_opt <- options(width = 72L)
si_vec <- trimws(capture.output(sessionInfo()), "right")
options(old_opt)
# library paths are single long words, so a long line breaks after a slash
long_vec <- nchar(si_vec) > 72L
si_vec[long_vec] <- gsub("(.{1,68}/)(?=.{8,})", "\\1\n        ",
                         si_vec[long_vec], perl = TRUE)
writeLines(si_vec)
```
