---
title: "Choosing between the three fitting regimes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing between the three fitting regimes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~20 s under rmarkdown::render() with ggplot2, mclust,
     mixtools, flexmix and ks installed, the simulation and the NUTS
     reference read from their stored results; macOS arm64 (Apple silicon),
     R 4.6.1, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
```

```{r library}
library(proxymix)
```

```{r engines}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r fixed-notation, include = FALSE}
## Small numbers are written as plain decimals rather than in the
## scientific notation knitr's inline hook would otherwise typeset.
fixed <- function(v, digits) {
  format(round(v, digits), nsmall = digits, scientific = FALSE)
}
## A small number as a power of ten in LaTeX, to two significant digits.
power_ten <- function(v) {
  if (v == 0) return("$0$")
  e <- floor(log10(abs(v)))
  m <- signif(v / 10^e, 2L)
  if (m == 1) sprintf("$10^{%d}$", e) else sprintf("$%s \\times 10^{%d}$", format(m), e)
}
```

## The problem

A modeller with a target in hand and three regimes on offer asks the
practical question: *which of the three does my problem actually need, and
what do I give up if I pick the wrong one?* The routing rule is easy to
state and easy to over-trust, because two of the regimes will happily run
on the same target and return visibly similar mixtures while answering
subtly different questions.

## Package capabilities

The three regimes are those of van der Hoek and Elliott (2024), and they
differ in what they need to know about the target rather than in what they
return. All three return a `gmm_fit`.

`fit_moment_match()` is regime (i). At `N == 1` the KL-optimal Gaussian
proxy is the one that shares the target's mean and covariance, so the fit
is closed form and needs no iteration at all. It reads moments, which in
practice means it reads the samples the target carries.

`fit_em_samples()` is regime (ii): textbook expectation-maximisation on
independent draws, maximising the likelihood of the samples under an
`N`-component mixture. It needs samples and nothing else.

`fit_kld_em()` is regime (iii): importance-sampled KLD-EM, which minimises
$\mathrm{KL}(f \Vert g_\theta)$ using only point-wise evaluations of the
target's log-density, re-weighted by self-normalised importance weights
against a `proposal_mvt()` proposal. It needs no samples.

`fit_proxymix()` dispatches across all three, and `bic_aic()` and
`select_N()` choose the component count, which none of the three does for
you.

## Addressing the problem

The bundled three-component mixture target is the right test bench: its
log-density is known exactly, and because the target is itself a Gaussian
mixture its true samples are available too. That is the one setting in
which all three regimes can be run on the same object and compared.

```{r target}
tgt <- mixture_target(with_samples = TRUE, n = 1500L, seed = 1L)
tgt
```

### Regime (i): closed-form moment matching

At `N == 1` the KL-optimal Gaussian proxy has the same mean and covariance
as the target. There is nothing to iterate.

```{r moment}
m_fit <- fit_proxymix(tgt, N = 1L, regime = "moment")
m_fit
```

The recovered moments are the empirical moments of the samples the target
carries, not an approximation to them. The only discrepancy is the
diagonal ridge `ridge_eps` that keeps the covariance positive-definite,
and setting it to zero removes even that.

```{r moment-recover}
m_fit_bare <- fit_proxymix(tgt, N = 1L, regime = "moment", ridge_eps = 0)
moment_gap <- c(
  mean = max(abs(m_fit@means[[1L]] - colMeans(tgt@samples))),
  covariance = max(abs(m_fit@covariances[[1L]] - cov(tgt@samples))),
  covariance_no_ridge = max(abs(m_fit_bare@covariances[[1L]] -
                                  cov(tgt@samples)))
)
signif(moment_gap, 3L)
```

A single Gaussian is a poor description of a three-mode mixture, but it is
the KL-optimal single Gaussian, and it makes a deterministic seed for the
two iterative regimes at `N > 1`.

### Regime (ii): classical EM on samples

When the target carries samples, the textbook algorithm applies:

* **E-step** -- \(\gamma_{nk} = \pi_k \mathcal{N}(x_n \mid \mu_k, \Sigma_k) /
  \sum_j \pi_j \mathcal{N}(x_n \mid \mu_j, \Sigma_j)\).
* **M-step** -- re-weight \(\pi_k, \mu_k, \Sigma_k\) by the
  responsibilities, with a diagonal ridge `ridge_eps` keeping
  \(\Sigma_k\) positive-definite.
* **Convergence** -- relative change in the empirical log-likelihood below
  `tol`.

```{r em}
s_fit <- fit_proxymix(tgt, N = 3L, regime = "sample",
                      max_iter = 200L, n_starts = 4L, seed = 1L)
s_fit
```

### Regime (iii): importance-sampled KLD-EM

Regime (iii) fits a proxy when the target's `log_density` is evaluable
but no samples are available, which is the situation for Bayesian
posteriors and for downscaling targets in spatial statistics. The algorithm
draws `is_size` points from the proposal $q$, forms self-normalised weights
\(W_n \propto f(x_n) / q(x_n)\), and runs EM iterations in which the M-step
minimises \(\mathrm{KL}(f \Vert g_\theta)\) by re-weighting the
responsibilities with $W$.

The proposal is the only free knob beyond the usual mixture configuration,
and a Student-t with five degrees of freedom is a robust default.

```{r kld}
k_fit <- fit_proxymix(tgt, N = 3L, regime = "kld",
                      proposal = proposal_mvt(n_dim = 2L,
                                              mean = c(0, 0),
                                              sigma = 6 * diag(2),
                                              df = 5),
                      is_size = 3000L,
                      max_iter = 60L,
                      seed = 1L)
k_fit
```

### The three fits side by side

```{r overlay-grid}
grid_x <- seq(-4.5, 4.5, length.out = 100L)
grid_base <- expand.grid(x1 = grid_x, x2 = grid_x)
grid_mat <- as.matrix(grid_base)
target_d <- exp(tgt@log_density(grid_mat))
panel_of <- function(fit, label) {
  data.frame(
    x1 = grid_base$x1, x2 = grid_base$x2,
    target = target_d,
    proxy = dgmm(grid_mat, fit),
    regime = label,
    stringsAsFactors = FALSE
  )
}
overlay_df <- rbind(
  panel_of(m_fit, "(i) moment, N = 1"),
  panel_of(s_fit, "(ii) sample EM, N = 3"),
  panel_of(k_fit, "(iii) KLD-EM, N = 3")
)
overlay_df$regime <- factor(overlay_df$regime,
                            levels = unique(overlay_df$regime))
```

```{r overlay, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.4, fig.cap = "The three-mixture target (filled contours, identical in all three panels) with each regime's fitted proxy overlaid as dashed contours. Regime (i) is one Gaussian spread across all three modes; regimes (ii) and (iii) both place one component per mode and are visually indistinguishable, despite reading entirely different information about the target.", fig.alt = "Three side-by-side contour panels of the same three-mode target, overlaid with the single-Gaussian moment fit, the sample-EM fit and the KLD-EM fit."}
ggplot2::ggplot(overlay_df, ggplot2::aes(x1, x2)) +
  ggplot2::geom_contour_filled(ggplot2::aes(z = target), bins = 10L,
                               alpha = 0.85) +
  ggplot2::geom_contour(ggplot2::aes(z = proxy), colour = "white",
                        linetype = "dashed", linewidth = 0.4, bins = 5L) +
  ggplot2::scale_fill_viridis_d(option = "mako", guide = "none") +
  ggplot2::facet_wrap(~ regime) +
  ggplot2::coord_equal() +
  ggplot2::labs(x = expression(x[1]), y = expression(x[2])) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r overlay-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the three-panel overlay",
    "figure is skipped.\n")
```

### What each iterative regime optimises

```{r traces}
trace_df <- rbind(
  data.frame(
    iteration = seq_along(s_fit@diagnostics$loglik_trace),
    value = s_fit@diagnostics$loglik_trace,
    panel = "(ii) sample EM: log-likelihood (up)",
    stringsAsFactors = FALSE
  ),
  data.frame(
    iteration = seq_along(kld_trace(k_fit)),
    value = kld_trace(k_fit),
    panel = "(iii) KLD-EM: in-sample KL estimate (down)",
    stringsAsFactors = FALSE
  )
)
```

```{r traces-plot, eval = has_ggplot2, echo = has_ggplot2, fig.height = 3.2, fig.cap = "The two iterative regimes optimise different objectives and take different numbers of steps to do it. Sample EM climbs the empirical log-likelihood; KLD-EM drives down the importance-sampled Kullback-Leibler divergence, estimated on the draws the fit is tuned to, which can fall below zero.", fig.alt = "Two panels of iteration traces: a monotone increasing log-likelihood curve for sample EM and a decreasing Kullback-Leibler curve for KLD-EM."}
ggplot2::ggplot(trace_df, ggplot2::aes(iteration, value)) +
  ggplot2::geom_line(colour = "#0072B2", linewidth = 0.8) +
  ggplot2::geom_point(colour = "#0072B2", size = 1.1) +
  ggplot2::facet_wrap(~ panel, scales = "free") +
  ggplot2::labs(x = "iteration", y = "objective") +
  ggplot2::theme_minimal(base_size = 11)
```

```{r traces-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the two-panel iteration",
    "trace figure is skipped.\n")
```

```{r kld-quadrature, include = FALSE}
# KL(target || k_fit) by quadrature; the grid holds all but 1e-7 of the mass
quad_step <- 0.06
kq_x <- seq(-6, 6, by = quad_step)
kq_grid <- as.matrix(expand.grid(x1 = kq_x, x2 = kq_x))
kq_log_f <- tgt@log_density(kq_grid)
k_fit_kl_quad <- sum(exp(kq_log_f) *
                       (kq_log_f - dgmm(kq_grid, k_fit, log = TRUE))) *
  quad_step^2
kq_mass <- sum(exp(kq_log_f)) * quad_step^2
```

### Regimes (i) and (iii) at N = 1, against the exact answer

At `N = 1` both regime (i) and regime (iii) target the KL-optimal single
Gaussian, so they should agree. They are compared here on the banana
target, where regime (iii) would be the only option if the samples were
withheld, and both are compared against a third quantity neither of them
produced: the exact mean and covariance of the banana density. The banana
is $x_1 = z_1$, $x_2 = z_2 + b\,(z_1^2 - 1)$ with $z_1$ and $z_2$
independent standard normal and $b = 1/2$. Its mean is $(0, 0)$, its two
coordinates are uncorrelated, and because $\mathrm{Var}(z_1^2) = 2$ the
trace of its covariance is $1 + (1 + 2b^2) = 5/2$.

```{r n1-fits}
banana <- banana_target(with_samples = TRUE, n = 2000L, seed = 1L)
m_b <- fit_proxymix(banana, N = 1L, regime = "moment")
k_b <- fit_proxymix(banana, N = 1L, regime = "kld",
                    proposal = proposal_mvt(n_dim = 2L,
                                            sigma = 4 * diag(2),
                                            df = 5),
                    is_size = 3000L, max_iter = 50L, seed = 1L)
```

```{r n1-exact}
# the warp of banana_target(): x2 = z2 + b * (z1^2 - 1)
b <- 0.5
exact_mean <- c(0, 0)
exact_trace <- 1 + (1 + 2 * b^2)
```

```{r n1-grid, include = FALSE}
# a grid quadrature of the same moments, for the Limitations section
quad_x <- seq(-8, 8, length.out = 400L)
quad_g <- as.matrix(expand.grid(x1 = quad_x, x2 = quad_x))
quad_cell <- (quad_x[2L] - quad_x[1L])^2
quad_f <- exp(banana@log_density(quad_g))
quad_mass <- sum(quad_f) * quad_cell
quad_trace <- sum(quad_f * (quad_g[, 1L]^2 + quad_g[, 2L]^2)) *
  quad_cell / quad_mass
```

```{r n1-compare}
tr_of <- function(f) sum(diag(f@covariances[[1L]]))
n1 <- list(
  mean_gap = max(abs(m_b@means[[1L]] - k_b@means[[1L]])),
  trace_gap = tr_of(m_b) - tr_of(k_b),
  moment_error = tr_of(m_b) - exact_trace,
  kld_error = tr_of(k_b) - exact_trace,
  sample_error = sum(diag(cov(banana@samples))) - exact_trace
)
n1$closer <- abs(n1$moment_error) / abs(n1$kld_error)
data.frame(value = signif(unlist(n1), 3L))
```

```{r n1-table, echo = FALSE}
n1_tbl <- data.frame(
  Source = c("regime (i), moment match", "regime (iii), KLD-EM",
             "attached samples", "exact moments (closed form)"),
  Reads = c("the attached samples", "the log-density",
            "the attached samples", "the definition of the target"),
  `Trace of covariance` = round(
    c(tr_of(m_b), tr_of(k_b), sum(diag(cov(banana@samples))), exact_trace),
    3L
  ),
  `Error against exact value` = round(
    c(tr_of(m_b), tr_of(k_b), sum(diag(cov(banana@samples))), exact_trace) -
      exact_trace,
    3L
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(
  n1_tbl,
  caption = paste(
    "The single-Gaussian proxy at `N = 1`, fitted two ways and scored",
    "against the exact moments of the banana density. The error is the",
    "trace minus the exact trace, so a negative error is an understatement.",
    "The final row is the exact value, so its error is zero by construction."
  )
)
```

The attached samples and the importance draws are one realisation each.
Repeating both over fresh draws shows how far each trace moves by chance.

```{r n1-spread}
sample_traces <- vapply(seq_len(500L), function(s) {
  sum(diag(cov(banana_target(with_samples = TRUE, n = 2000L,
                             seed = s)@samples)))
}, numeric(1L))
kld_traces <- vapply(seq_len(100L), function(s) {
  tr_of(fit_proxymix(banana, N = 1L, regime = "kld",
                     proposal = proposal_mvt(n_dim = 2L,
                                             sigma = 4 * diag(2),
                                             df = 5),
                     is_size = 3000L, max_iter = 50L, seed = s))
}, numeric(1L))
spread <- data.frame(
  replicates = c(length(sample_traces), length(kld_traces)),
  mean = c(mean(sample_traces), mean(kld_traces)),
  sd = c(sd(sample_traces), sd(kld_traces)),
  row.names = c("trace of 2000 samples", "regime (iii) trace")
)
round(spread, 3L)
```

## Interpretation

The moment fit is exact rather than approximate. It reproduces the
empirical mean of the target's samples with a gap of
`r power_ten(moment_gap[["mean"]])`, and the empirical covariance
with a gap of `r power_ten(moment_gap[["covariance"]])` -- which is
not a fitting error but the diagonal ridge itself, since removing the ridge
closes that gap to
`r power_ten(moment_gap[["covariance_no_ridge"]])` as well. Regime
(i) does not approximate the moments; it copies them.

The two iterative regimes land in the same place by different routes.
Sample EM took `r length(s_fit@diagnostics$loglik_trace)` iterations to
converge and KLD-EM took `r length(kld_trace(k_fit))`, and the fitted
weights agree to
`r format(signif(max(abs(sort(s_fit@weights) - sort(k_fit@weights))), 2L))`.
The overlay shows why they look alike here: both put one component on each
mode, which is the correct answer for a target that is itself a
three-component Gaussian mixture. The KLD-EM fit retained an effective
sample size of `r round(k_fit@diagnostics$ess, 0L)` out of
`r format(k_fit@diagnostics$is_size, big.mark = ",")` importance draws, so it reached that answer
without leaning on a few high-weight points.

The KLD-EM trace ends at
`r fixed(k_fit@diagnostics$kld_final, 4)`, below zero, although a
divergence cannot be negative. The trace is estimated on the importance
draws the fit was tuned to, so it reads low. On
`r format(k_fit@diagnostics$validation_size, big.mark = ",")` held-out draws the estimate is
`r fixed(k_fit@diagnostics$validation_kld, 4)` with a Monte Carlo standard
error of `r fixed(k_fit@diagnostics$validation_mc_se, 4)`, and quadrature of
the exact target density on a grid over $[-6, 6]^2$, which leaves out
`r power_ten(1 - kq_mass)` of the target's mass, gives
`r fixed(k_fit_kl_quad, 4)`.

The `N = 1` comparison is where the cost of the choice becomes visible.
The two fits differ in mean by `r format(signif(n1$mean_gap, 2L))` and in
trace-covariance by `r fixed(n1$trace_gap, 3)`.
Regime (i) matches the moments of the `r nrow(banana@samples)` samples the
target happens to carry, and those samples
`r if (n1$sample_error > 0) "overstate" else "understate"` the exact
trace-covariance of `r exact_trace` by `r fixed(abs(n1$sample_error), 3)`.
Regime (iii) never sees those samples; it reads the density, and its
trace-covariance is `r fixed(abs(n1$kld_error), 3)`
`r if (n1$kld_error < 0) "below" else "above"` the exact value, about
`r round(n1$closer, 1L)` times closer in this realisation. Both errors are
single draws. Over `r length(sample_traces)` fresh samples of
`r nrow(banana@samples)` points the sample trace has a standard deviation
of `r fixed(sd(sample_traces), 3)`, so this sample's error is
`r fixed(abs(n1$sample_error) / sd(sample_traces), 1)` standard deviations.
Over `r length(kld_traces)` importance seeds the regime-(iii) trace has a
standard deviation of `r fixed(sd(kld_traces), 3)` and a mean
`r fixed(abs(mean(kld_traces) - exact_trace), 3)`
`r if (mean(kld_traces) < exact_trace) "below" else "above"` the exact
value, `r if (abs(mean(kld_traces) - exact_trace) > 2 * sd(kld_traces) / sqrt(length(kld_traces))) "more" else "less"`
than two standard errors of that mean.

So regime (iii) is not merely the fallback when samples are missing. When
the log-density is available, it is fitting to the density itself, while
regimes (i) and (ii) are fitting to whatever finite sample was drawn from
it. With plentiful samples the distinction is immaterial. With
`r nrow(banana@samples)` samples the sampling error of the trace is of the
same order as the importance-sampling error of regime (iii) at
`r format(k_b@diagnostics$is_size, big.mark = ",")` draws, so which fit lands closer in a single
run is partly chance. What you give up by choosing regime (iii)
when samples are plentiful is speed and simplicity: it needs a proposal, it
costs `r format(k_fit@diagnostics$n_target_evals, big.mark = ",")` target evaluations here
(`r format(k_fit@diagnostics$is_size, big.mark = ",")` importance draws and
`r format(k_fit@diagnostics$validation_size, big.mark = ",")` held-out draws), and it
carries an importance-sampling diagnostic that has to be read.

## Limitations

This vignette does not choose the number of components. Every fit above
was told `N` by hand, and the three-mixture target was picked precisely
because the right answer is known to be three. On a real target the
component count is the first thing that has to be settled, and `bic_aic()`
and `select_N()` are what settle it; a count set too low shows up as a
residual divergence that more importance draws will not reduce.

The comparison between regimes (ii) and (iii) is flattering to both here
for a reason worth stating. The target is itself a Gaussian mixture with
the same number of components as the proxies, so the model is correctly
specified and there is no approximation error for either regime to trade
against the other. On a target outside the Gaussian-mixture family the two
regimes minimise genuinely different objectives, and the fits separate.

The exact `N = 1` answer is closed form only because the banana is a warped
Gaussian. A grid quadrature of the same moments on $[-8, 8]^2$, with
`r length(quad_x)` points a side, leaves out `r power_ten(1 - quad_mass)`
of the mass and puts the trace at `r fixed(quad_trace, 3)`,
`r fixed(exact_trace - quad_trace, 3)` below the exact value, because the
mass it leaves out lies far out in the $x_2$ tail, where it weighs heavily
on the variance. Grid quadrature is an independent check in two or three
dimensions once its captured mass has been checked, and it stops being one
well before the dimension at which importance sampling itself becomes the
binding constraint.

Regime (ii) is compared above only with the package's own regimes. The
numerical illustration that follows puts it beside three established
mixture packages on real data and in simulation, and puts regime (iii)
beside a Markov chain Monte Carlo route to the same density.

## Numerical illustration

```{r engines-illustration, include = FALSE}
has_mclust <- requireNamespace("mclust", quietly = TRUE)
has_mixtools <- requireNamespace("mixtools", quietly = TRUE)
has_flexmix <- requireNamespace("flexmix", quietly = TRUE)
has_ks <- requireNamespace("ks", quietly = TRUE)
run_illustration <- has_mclust && has_mixtools && has_flexmix && has_ks
```

```{r illustration-skip, eval = !run_illustration, echo = FALSE, results = "asis"}
cat("mclust, mixtools, flexmix or ks is not installed on this build, so",
    "the real-data illustration is skipped.\n")
```

### Fitting the Old Faithful eruptions with four packages

The question is which fitted two-component mixture predicts eruptions it
has not seen, and whether regime (iii) reaches the same answer from a
density alone. The data are the `r nrow(datasets::faithful)` eruptions of
the Old Faithful geyser shipped with R as `faithful` (Azzalini and Bowman,
1990), each an eruption length in minutes and the waiting time to the next
eruption in minutes. The scatter has two clusters. Half of the eruptions
are held out, every mixture is fitted to the other half, and each is scored
by the mean log-likelihood it assigns to the held-out half. mclust (Scrucca
et al., 2016), mixtools (Benaglia et al., 2009) and flexmix (Leisch, 2004;
Grün and Leisch, 2008) fit the mixture by expectation-maximisation on the
training half, as regime (ii) does.

Regime (iii) is given no samples. Its target is the kernel density estimate
of the training half from ks (Duong, 2007), read as a function that can be
evaluated at any point, and its competitor on that target is the route an
analyst without proxymix would take: draw from the density with the
No-U-Turn sampler in Stan (Carpenter et al., 2017) through cmdstanr, then
fit a two-component mixture to the draws with mclust. The kernel density
estimate itself is scored on the held-out half as the reference for both.

mclust, mixtools, flexmix and ks install from CRAN. cmdstanr is not on
CRAN and installs from the Stan R-universe repository,
stan-dev.r-universe.dev. It also needs a CmdStan installation, which
`cmdstanr::install_cmdstan()` provides:

```{r install-competitors, eval = FALSE}
install.packages(c("mclust", "mixtools", "flexmix", "ks"))
install.packages("cmdstanr",
                 repos = c("https://stan-dev.r-universe.dev",
                           getOption("repos")))
```

```{r load-competitors, eval = run_illustration, message = FALSE}
library(mclust)
library(mixtools)
library(flexmix)
library(ks)
```

```{r faithful-split, eval = run_illustration}
faithful_mat <- as.matrix(datasets::faithful)
set.seed(20260925)
i_train <- sort(sample.int(nrow(faithful_mat), nrow(faithful_mat) / 2L))
train <- faithful_mat[i_train, ]
test <- faithful_mat[-i_train, ]
```

Each fit below is converted to a `gmm` object, so one function,
`dgmm()`, scores every mixture on the held-out half. mclust chooses its
covariance structure by BIC from its full list; flexmix is run through
`stepFlexmix()` with five random starts, the number proxymix uses by
default, where `flexmix()` itself makes one start from a random
assignment and `stepFlexmix()` defaults to three; mixtools runs with its
defaults.

```{r regime-ii-fits, eval = run_illustration, results = "hide"}
as_gmm_mclust <- function(fit) {
  gmm(
    weights = fit$parameters$pro,
    means = lapply(seq_len(fit$G), function(k) fit$parameters$mean[, k]),
    covariances = lapply(seq_len(fit$G), function(k) {
      fit$parameters$variance$sigma[, , k]
    })
  )
}
as_gmm_mixtools <- function(fit) {
  gmm(weights = fit$lambda, means = fit$mu, covariances = fit$sigma)
}
as_gmm_flexmix <- function(fit) {
  comps <- lapply(fit@components, function(cc) cc[[1L]]@parameters)
  gmm(
    weights = prior(fit),
    means = lapply(comps, function(p) unname(p$center)),
    covariances = lapply(comps, function(p) unname(p$cov))
  )
}
timed <- function(expr) {
  secs <- system.time(g <- expr)[["elapsed"]]
  list(g = g, secs = secs)
}

train_df <- data.frame(eruptions = train[, 1L], waiting = train[, 2L])
sample_target <- gmm_target_from_samples(train)
set.seed(1L)
fits <- list(
  "proxymix, regime (i), N = 1" =
    timed(fit_proxymix(sample_target, N = 1L, regime = "moment")),
  "proxymix, regime (ii)" =
    timed(fit_proxymix(sample_target, N = 2L, regime = "sample")),
  mclust = timed(as_gmm_mclust(Mclust(train, G = 2L, verbose = FALSE))),
  mixtools = timed(as_gmm_mixtools(mvnormalmixEM(train, k = 2L,
                                                 verb = FALSE))),
  flexmix = timed(as_gmm_flexmix(stepFlexmix(
    cbind(eruptions, waiting) ~ 1, data = train_df, k = 2L, nrep = 5L,
    model = FLXMCmvnorm(diagonal = FALSE), verbose = FALSE
  )))
)
```

The kernel density estimate uses the ks plug-in bandwidth matrix. Its log
density is written out as a log-sum-exp over the training points, so it
stays finite far from the data, and it is checked against the estimate
that `kde()` from ks computes directly at the data points before being
wrapped as a target with no samples.

```{r kde-target, eval = run_illustration}
kde_fit <- kde(train)
H_inv <- solve(kde_fit$H)
log_det_H <- determinant(kde_fit$H)$modulus[[1L]]
centre_q <- rowSums((train %*% H_inv) * train)
log_kde <- function(x) {
  if (is.null(dim(x))) x <- matrix(x, nrow = 1L)
  q <- outer(rowSums((x %*% H_inv) * x), centre_q, "+") -
    2 * x %*% H_inv %*% t(train)
  lk <- -0.5 * q - log(2 * pi) - 0.5 * log_det_H
  m <- apply(lk, 1L, max)
  m + log(rowSums(exp(lk - m))) - log(nrow(train))
}
kde_direct <- kde(train, H = kde_fit$H, eval.points = faithful_mat)$estimate
kde_check <- max(abs(exp(log_kde(faithful_mat)) - kde_direct))
kde_target <- gmm_target(n_dim = 2L, log_density = log_kde,
                         normalised = TRUE, log_normalizer = 0,
                         name = "kde of the training half")

fits[["proxymix, regime (iii), on the KDE"]] <- timed(fit_proxymix(
  kde_target, N = 2L, regime = "kld",
  proposal = proposal_mvt(n_dim = 2L, mean = colMeans(train),
                          sigma = 2 * cov(train), df = 5),
  is_size = 4000L, seed = 1L
))
k_faithful <- fits[["proxymix, regime (iii), on the KDE"]]$g
```

```{r ref-read, echo = FALSE}
ref <- readRDS("three_regimes_ref.rds")
if (!identical(ref$code, as.character(knitr::knit_code$get("ref-code")))) {
  stop("The stored sampling results were produced by code other than ",
       "the chunk shown; rerun the reference computation.", call. = FALSE)
}
```

Four chains of the No-U-Turn sampler draw from the same kernel density
estimate, and mclust fits a two-component mixture to the pooled draws. The
sampler reads the Stan program below from the file `kde_target.stan` in a
folder `three_regimes_stan` under the working directory.

```stan
data {
  int<lower=1> K;                 // number of kernel centres
  int<lower=1> D;                 // dimension
  array[K] vector[D] centre;      // kernel centres, the training data
  cov_matrix[D] H;                // kernel bandwidth matrix
}
transformed data {
  matrix[D, D] L = cholesky_decompose(H);
}
parameters {
  vector[D] x;
}
model {
  vector[K] lp;
  for (k in 1:K) lp[k] = multi_normal_cholesky_lpdf(x | centre[k], L);
  target += log_sum_exp(lp) - log(K);
}
```

This chunk is complete and runs as shown, but it took about
`r format(round(ref$elapsed_secs / 30) / 2)` minutes on one core (R
`r ref$versions[["R"]]`, Apple silicon); the results below are read from
its stored output. That time covers the whole chunk: compiling the Stan
program, the kernel density estimate, the four chains and the mclust fit.
For a quick run, change `iter_sampling = 25000L` to
`iter_sampling = 1000L`.

```{r ref-code, eval = FALSE}
library(cmdstanr)
library(ks)
library(mclust)

faithful_mat <- as.matrix(datasets::faithful)
set.seed(20260925)
i_train <- sort(sample.int(nrow(faithful_mat), nrow(faithful_mat) / 2L))
train <- faithful_mat[i_train, ]
kde_fit <- kde(train)

model <- cmdstan_model(file.path("three_regimes_stan", "kde_target.stan"),
                       dir = tempdir())
nuts <- model$sample(
  data = list(K = nrow(train), D = 2L, centre = train, H = kde_fit$H),
  chains = 4L, parallel_chains = 1L,
  iter_warmup = 2000L, iter_sampling = 25000L,
  seed = 1L, refresh = 0L
)
draws <- nuts$draws("x", format = "matrix")
nuts_summary <- as.data.frame(nuts$summary("x"))
nuts_divergent <- sum(nuts$diagnostic_summary()$num_divergent)
two_stage <- Mclust(draws, G = 2L, verbose = FALSE)
two_stage_par <- two_stage$parameters[c("pro", "mean", "variance")]
```

```{r ref-use, eval = run_illustration, echo = FALSE}
if (!identical(ref$i_train, i_train)) {
  stop("The stored sampling results used a different training half.",
       call. = FALSE)
}
fits[["NUTS draws from the KDE, then mclust"]] <- list(
  g = as_gmm_mclust(list(G = 2L, parameters = ref$two_stage_par)),
  secs = ref$elapsed_secs
)
nuts_rhat <- max(ref$nuts_summary$rhat)
nuts_ess <- min(ref$nuts_summary$ess_bulk)
```

```{r faithful-table, eval = run_illustration, echo = FALSE}
held_out <- function(g) mean(dgmm(test, g, log = TRUE))
faithful_tbl <- data.frame(
  method = c(names(fits), "kernel density estimate (ks)"),
  reads = c(rep("the training half", 5L), "the KDE as a density",
            "the KDE as a density", "the training half"),
  loglik = c(vapply(fits, function(f) held_out(f$g), numeric(1L)),
             mean(log_kde(test))),
  secs = c(vapply(fits, function(f) f$secs, numeric(1L)), NA_real_),
  stringsAsFactors = FALSE
)
faithful_show <- faithful_tbl
faithful_show$loglik <- fixed(faithful_show$loglik, 3)
is_nuts <- faithful_show$method == "NUTS draws from the KDE, then mclust"
faithful_show$secs <- ifelse(is_nuts, fixed(faithful_show$secs, 0),
                             fixed(faithful_show$secs, 2))
faithful_show$secs[is.na(faithful_tbl$secs)] <- "--"
knitr::kable(
  faithful_show, row.names = FALSE, align = c("l", "l", "r", "r"),
  # the padded header widens the column so the word fits inside the rule
  col.names = c("Method", "Reads", "Held-out mean log-likelihood",
                "\u00a0\u00a0\u00a0Seconds"),
  caption = paste0(
    "Mean log-likelihood per eruption assigned to the ", nrow(test),
    " held-out eruptions by each fitted mixture, and the seconds each ",
    "fit took. The seconds of the NUTS row cover the whole reference ",
    "chunk (compiling the Stan program, the kernel density estimate, the ",
    "four chains and the mclust fit); the other rows time the fit alone. ",
    "The kernel density estimate is the target of the two ",
    "density-reading rows and their reference, and is not timed."
  )
)
ll_of <- function(method) {
  faithful_tbl$loglik[faithful_tbl$method == method]
}
secs_of <- function(method) {
  faithful_tbl$secs[faithful_tbl$method == method]
}
ll_ii <- ll_of("proxymix, regime (ii)")
ll_iii <- ll_of("proxymix, regime (iii), on the KDE")
ll_kde <- ll_of("kernel density estimate (ks)")
ll_nuts <- ll_of("NUTS draws from the KDE, then mclust")
ll_two <- c(ll_ii, ll_of("mclust"), ll_of("mixtools"), ll_of("flexmix"))
```

```{r fig-faithful, eval = run_illustration && has_ggplot2, echo = run_illustration && has_ggplot2, fig.height = 4.2, fig.cap = "The held-out half of the Old Faithful eruptions with the two-component mixture fitted by regime (ii) to the training half (solid contours) and by regime (iii) to the kernel density estimate of the training half (dashed contours), both drawn at the same density levels. Both routes place one component on each cluster. The regime-(iii) mixture is wider and lower, because the kernel density estimate it fitted adds the kernel bandwidth to the spread of each cluster, so fewer of its contours reach the upper levels.", fig.alt = "Scatter of held-out eruption length against waiting time in two clusters, overlaid with two sets of density contours that both enclose the two clusters."}
plot_x <- seq(1.2, 5.8, length.out = 120L)
plot_y <- seq(35, 105, length.out = 120L)
plot_grid <- as.matrix(expand.grid(eruptions = plot_x, waiting = plot_y))
contour_df <- rbind(
  data.frame(plot_grid, density = dgmm(plot_grid, fits[[2L]]$g),
             route = "regime (ii), training half"),
  data.frame(plot_grid, density = dgmm(plot_grid, k_faithful),
             route = "regime (iii), KDE")
)
test_df <- data.frame(eruptions = test[, 1L], waiting = test[, 2L])
ggplot2::ggplot() +
  ggplot2::geom_point(data = test_df,
                      ggplot2::aes(eruptions, waiting),
                      colour = "grey45", size = 1.2) +
  ggplot2::geom_contour(data = contour_df,
                        ggplot2::aes(eruptions, waiting, z = density,
                                     colour = route, linetype = route),
                        bins = 6L, linewidth = 0.6) +
  ggplot2::scale_colour_manual(name = NULL,
                               values = c("#0072B2", "#D55E00")) +
  ggplot2::scale_linetype_manual(name = NULL,
                                 values = c("solid", "dashed")) +
  ggplot2::labs(x = "eruption length, minutes",
                y = "waiting time to the next eruption, minutes") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r fig-faithful-skip, eval = run_illustration && !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the Old Faithful",
    "figure is skipped.\n")
```

```{r faithful-facts, eval = run_illustration, include = FALSE}
mclust_model <- Mclust(train, G = 2L, verbose = FALSE)$modelName
weight_ii <- max(fits[[2L]]$g@weights)
weight_iii <- max(k_faithful@weights)
kde_peak <- max(kde_direct)
```

The four regime-(ii) fits are within
`r fixed(ceiling(1000 * (max(ll_two) - min(ll_two))) / 1000, 3)` of one another on the held-out
half: proxymix gives `r fixed(ll_ii, 3)`, mclust `r fixed(ll_of("mclust"), 3)`
with its `r mclust_model` covariance structure, mixtools
`r fixed(ll_of("mixtools"), 3)` and flexmix `r fixed(ll_of("flexmix"), 3)`.
The fits took `r fixed(secs_of("proxymix, regime (ii)"), 2)`,
`r fixed(secs_of("mclust"), 2)`, `r fixed(secs_of("mixtools"), 2)` and
`r fixed(secs_of("flexmix"), 2)` s in that order. The single Gaussian of regime (i) scores
`r fixed(ll_of("proxymix, regime (i), N = 1"), 3)`, the cost of one
component for two clusters.

Regime (iii) never saw the training points directly. From the kernel density estimate
alone it reaches `r fixed(ll_iii, 3)`, beside the estimate's own
`r fixed(ll_kde, 3)`, in `r fixed(secs_of("proxymix, regime (iii), on the KDE"), 2)`
s with an effective sample size of
`r round(k_faithful@diagnostics$ess)` of `r format(k_faithful@diagnostics$is_size, big.mark = ",")`
importance draws. Its larger component carries weight
`r fixed(weight_iii, 3)`, against `r fixed(weight_ii, 3)` for the regime-(ii)
fit to the same half. The sampling route reaches `r fixed(ll_nuts, 3)` from
`r format(ref$n_draws, big.mark = ",")` draws in
`r round(ref$elapsed_secs)` s for the whole reference chunk, with `r ref$nuts_divergent` divergent
transitions, a largest R-hat of `r fixed(nuts_rhat, 2)` over the two
coordinates and a smallest bulk effective sample size of
`r format(round(nuts_ess), big.mark = ",")`. The density is checked: the
largest absolute difference between the density from the log-sum-exp form
and the estimate `kde()` computes at the `r nrow(faithful_mat)` eruptions
is `r power_ten(kde_check)`, where the largest density at those points is
`r fixed(kde_peak, 4)`. This
is one split of one dataset, so the simulation below repeats the regime-(ii)
comparison where the truth is known.

### A simulation benchmark

```{r sim-read, echo = FALSE}
sim <- readRDS("three_regimes_sim.rds")
if (!identical(sim$code, as.character(knitr::knit_code$get("sim-code")))) {
  stop("The stored simulation results were produced by code other than ",
       "the chunk shown; rerun the simulation.", call. = FALSE)
}
```

One split is one draw. The simulation repeats the fit on `r sim$n_rep`
datasets at each of two sample sizes, `r sim$n_sizes[1L]` and
`r sim$n_sizes[2L]`, drawn from the three-component mixture of the
`mixture_target()` bench above, whose weights are 0.3, 0.4 and 0.3 and
whose means sit at $(-2, -2)$, $(0, 0)$ and $(2, 2)$. Every method is told
$N = 3$. proxymix regime (ii), mclust and flexmix fit the sample. proxymix
regime (iii) is given the mixture's log density and no sample, with an
importance-sampling seed of its own per dataset, so its row shows what the
density route returns while the sample route sees `r sim$n_sizes[1L]` or
`r sim$n_sizes[2L]` points. mixtools is left out of the simulation: one fit
of a `r sim$n_sizes[2L]`-point dataset with three components, timed at the
end of the chunk, took `r round(sim$mixtools_secs)` s, so
`r 2L * sim$n_rep` datasets would take hours.

Two scores are computed against quantities no method produced. The first
is $\mathrm{KL}(f \Vert g) = \int f \log (f / g)$, with $f$ the true
mixture and $g$ the fitted mixture, by quadrature on a grid of `r format(length(seq(-6, 6, by = 0.06))^2, big.mark = ",")`
points over $[-6, 6]^2$, which leaves out `r power_ten(1 - kq_mass)` of the
true mixture's mass. The second is the adjusted Rand index
(Hubert and Arabie, 1985) between the component that generated each point
and the component each fitted mixture assigns it to, computed by one rule
for every method, the largest posterior responsibility. The share of fits
whose divergence exceeds 0.1 is reported as the misfit rate.
This chunk is complete and runs as shown, but it took about
`r round(sim$elapsed_secs / 60)` minutes on one core (R
`r sim$versions[["R"]]`, Apple silicon); the results below are read from
its stored output. For a quick run, change
`n_rep <- 300L` to `n_rep <- 3L`.

```{r sim-code, eval = FALSE}
library(proxymix)
library(mclust)
library(flexmix)

n_rep <- 300L  # datasets per sample size
n_sizes <- c(200L, 1000L)

# the mixture behind mixture_target(): three components in two dimensions
w_true <- c(0.3, 0.4, 0.3)
mu_true <- list(c(-2, -2), c(0, 0), c(2, 2))
S_true <- list(matrix(c(0.6, 0.0, 0.0, 0.6), 2L, 2L),
               matrix(c(0.5, 0.2, 0.2, 0.5), 2L, 2L),
               matrix(c(0.4, -0.1, -0.1, 0.4), 2L, 2L))
truth <- gmm(weights = w_true, means = mu_true, covariances = S_true)
density_only <- mixture_target()

# KL(truth || g) by quadrature; the grid holds all but 1e-7 of the mass
grid_x <- seq(-6, 6, by = 0.06)
grid <- as.matrix(expand.grid(x1 = grid_x, x2 = grid_x))
cell <- (grid_x[2L] - grid_x[1L])^2
log_f <- dgmm(grid, truth, log = TRUE)
f_grid <- exp(log_f)
kl_quad <- function(g) {
  sum(f_grid * (log_f - dgmm(grid, g, log = TRUE))) * cell
}

# component with the largest posterior responsibility for each row
hard_assign <- function(x, g) {
  lp <- vapply(seq_along(g@weights), function(k) {
    one <- gmm(weights = 1, means = g@means[k], covariances = g@covariances[k])
    log(g@weights[k]) + dgmm(x, one, log = TRUE)
  }, numeric(nrow(x)))
  max.col(lp)
}

as_gmm_mclust <- function(fit) {
  gmm(
    weights = fit$parameters$pro,
    means = lapply(seq_len(fit$G), function(k) fit$parameters$mean[, k]),
    covariances = lapply(seq_len(fit$G), function(k) {
      fit$parameters$variance$sigma[, , k]
    })
  )
}
as_gmm_flexmix <- function(fit) {
  comps <- lapply(fit@components, function(cc) cc[[1L]]@parameters)
  gmm(
    weights = prior(fit),
    means = lapply(comps, function(p) unname(p$center)),
    covariances = lapply(comps, function(p) unname(p$cov))
  )
}
timed <- function(expr) {
  secs <- system.time(g <- expr)[["elapsed"]]
  list(g = g, secs = secs)
}

draw_dataset <- function(n, r) {
  set.seed(r)
  z <- sample.int(3L, n, replace = TRUE, prob = w_true)
  x <- matrix(0, n, 2L)
  for (k in seq_len(3L)) {
    one <- gmm(weights = 1, means = mu_true[k], covariances = S_true[k])
    x[z == k, ] <- rgmm(sum(z == k), one)
  }
  list(x = x, z = z)
}

one_dataset <- function(n, r) {
  d <- draw_dataset(n, r)
  x <- d$x
  z <- d$z
  df <- data.frame(x1 = x[, 1L], x2 = x[, 2L])
  runs <- list(
    "proxymix, regime (ii)" = timed(fit_proxymix(
      gmm_target_from_samples(x), N = 3L, regime = "sample"
    )),
    "proxymix, regime (iii)" = timed(fit_proxymix(
      density_only, N = 3L, regime = "kld",
      proposal = proposal_mvt(n_dim = 2L, sigma = 6 * diag(2L), df = 5),
      is_size = 3000L, seed = n + r
    )),
    mclust = timed(as_gmm_mclust(Mclust(x, G = 3L, verbose = FALSE))),
    flexmix = timed(as_gmm_flexmix(stepFlexmix(
      cbind(x1, x2) ~ 1, data = df, k = 3L, nrep = 5L,
      model = FLXMCmvnorm(diagonal = FALSE), verbose = FALSE
    )))
  )
  data.frame(
    n = n, rep = r, method = names(runs),
    kl = vapply(runs, function(f) kl_quad(f$g), numeric(1L)),
    ari = vapply(runs, function(f) {
      adjustedRandIndex(hard_assign(x, f$g), z)
    }, numeric(1L)),
    secs = vapply(runs, function(f) f$secs, numeric(1L))
  )
}

res <- do.call(rbind, lapply(n_sizes, function(n) {
  do.call(rbind, lapply(seq_len(n_rep), function(r) one_dataset(n, r)))
}))

sim_tab <- aggregate(
  cbind(kl = kl, misfit = kl > 0.1, ari = ari, secs = secs) ~ n + method,
  data = res, FUN = mean
)
sim_tab

# one mixtools fit at the larger size, timed to justify leaving it out
mixtools_secs <- system.time(
  mixtools::mvnormalmixEM(draw_dataset(n_sizes[2L], 1L)$x, k = 3L,
                          verb = FALSE)
)[["elapsed"]]
mixtools_secs
```

```{r sim-table, echo = FALSE}
tab_sim <- sim$sim_tab
method_order <- c("proxymix, regime (ii)", "mclust", "flexmix",
                  "proxymix, regime (iii)")
tab_sim <- tab_sim[order(tab_sim$n, match(tab_sim$method, method_order)), ]
knitr::kable(
  tab_sim[, c("n", "method", "kl", "misfit", "ari", "secs")],
  digits = c(0L, 0L, 4L, 3L, 3L, 3L), row.names = FALSE,
  # the padded header widens the column so the word fits inside the rule
  col.names = c("n", "Method", "KL to the truth", "Misfit rate",
                "Adjusted Rand index", "\u00a0\u00a0\u00a0Seconds"),
  caption = paste0(
    "Mean Kullback-Leibler divergence $\\mathrm{KL}(f \\Vert g)$, with $f$ ",
    "the true mixture and $g$ the fitted one, by quadrature, share of ",
    "fits with divergence above 0.1, mean adjusted Rand index against the ",
    "generating components, and mean seconds per fit, over ", sim$n_rep,
    " simulated datasets at each sample size."
  )
)
sim_value <- function(n, method, what) {
  s1 <- sim$sim_tab$n == n & sim$sim_tab$method == method
  sim$sim_tab[[what]][s1]
}
n_small <- sim$n_sizes[1L]
n_large <- sim$n_sizes[2L]
mcse_rate <- sqrt(0.5 * 0.5 / sim$n_rep)
# mean and standard error of the per-dataset difference a - b
paired <- function(n, a, b, what) {
  x <- sim$res[sim$res$n == n & sim$res$method == a, ]
  y <- sim$res[sim$res$n == n & sim$res$method == b, ]
  d <- x[[what]][order(x$rep)] - y[[what]][order(y$rep)]
  c(diff = mean(d), se = sd(d) / sqrt(length(d)))
}
resolved <- function(p) abs(p[["diff"]]) > 2 * p[["se"]]
ii <- "proxymix, regime (ii)"
iii <- "proxymix, regime (iii)"
kl_large <- paired(n_large, ii, "mclust", "kl")
ari_large <- paired(n_large, ii, "mclust", "ari")
kl_small <- paired(n_small, ii, "mclust", "kl")
ari_small <- paired(n_small, ii, "mclust", "ari")
kl_iii <- rbind(paired(n_small, ii, iii, "kl"), paired(n_large, ii, iii, "kl"))
# misfits at the smaller size, and the exact McNemar test on the discordant pairs
misfit_of <- function(method) {
  x <- sim$res[sim$res$n == n_small & sim$res$method == method, ]
  x$kl[order(x$rep)] > 0.1
}
mis_ii <- misfit_of(ii)
mis_mc <- misfit_of("mclust")
mcnemar_p <- binom.test(sum(mis_ii & !mis_mc),
                        sum(xor(mis_ii, mis_mc)))$p.value
# a version as its DESCRIPTION writes it, when the installed one is the one stored
desc_version <- function(pkg, stored) {
  if (identical(as.character(packageVersion(pkg)), stored)) {
    utils::packageDescription(pkg)$Version
  } else {
    stored
  }
}
flexmix_version <- desc_version("flexmix", sim$versions[["flexmix"]])
```

At `r n_large` points neither proxymix regime (ii) nor mclust exceeds the
divergence threshold. Their divergences are
`r fixed(sim_value(n_large, ii, "kl"), 4)` and
`r fixed(sim_value(n_large, "mclust", "kl"), 4)`, a paired difference,
proxymix minus mclust over the same datasets, of
`r fixed(kl_large[["diff"]], 4)` with a standard error of
`r fixed(kl_large[["se"]], 4)`. Their adjusted Rand indices are
`r fixed(sim_value(n_large, ii, "ari"), 3)` and
`r fixed(sim_value(n_large, "mclust", "ari"), 3)`, a paired difference of
`r fixed(ari_large[["diff"]], 4)` (standard error
`r fixed(ari_large[["se"]], 4)`). At `r n_small` points their divergences
rise to `r fixed(sim_value(n_small, ii, "kl"), 4)` and
`r fixed(sim_value(n_small, "mclust", "kl"), 4)`, a paired difference of
`r fixed(kl_small[["diff"]], 4)` (standard error
`r fixed(kl_small[["se"]], 4)`), and their adjusted Rand indices differ by
`r fixed(ari_small[["diff"]], 4)` (standard error
`r fixed(ari_small[["se"]], 4)`), which
`r if (resolved(ari_small)) "is" else "is not"` resolved. The divergence
exceeds 0.1 in `r sum(mis_ii)` of the `r sim$n_rep` datasets for proxymix
and `r sum(mis_mc)` for mclust; `r sum(mis_ii & !mis_mc)` datasets are
misfits for proxymix alone and `r sum(mis_mc & !mis_ii)` for mclust alone,
for an exact McNemar test p-value of `r fixed(mcnemar_p, 2)`. flexmix
`r flexmix_version` with five random starts exceeds it in
`r fixed(sim_value(n_small, "flexmix", "misfit"), 3)` of the
`r n_small`-point datasets and
`r fixed(sim_value(n_large, "flexmix", "misfit"), 3)` of the
`r n_large`-point ones, which lifts its mean divergence to
`r fixed(sim_value(n_small, "flexmix", "kl"), 4)` and
`r fixed(sim_value(n_large, "flexmix", "kl"), 4)` and lowers its adjusted
Rand index to `r fixed(sim_value(n_small, "flexmix", "ari"), 3)` and
`r fixed(sim_value(n_large, "flexmix", "ari"), 3)`. A rate estimated from
`r sim$n_rep` datasets carries a Monte Carlo standard error of at most
`r fixed(mcse_rate, 2)`.

The regime-(iii) row does not move with the sample size, because it reads
the density and not the sample. Its divergence is
`r fixed(sim_value(n_small, "proxymix, regime (iii)", "kl"), 4)` and
`r fixed(sim_value(n_large, "proxymix, regime (iii)", "kl"), 4)` at the two
sizes, `r round(sim_value(n_small, "proxymix, regime (ii)", "kl") / sim_value(n_small, "proxymix, regime (iii)", "kl"), 1)`
times below the sample route at `r n_small` points and
`r round(sim_value(n_large, "proxymix, regime (ii)", "kl") / sim_value(n_large, "proxymix, regime (iii)", "kl"), 1)`
times below it at `r n_large`; the paired differences, regime (ii) minus
regime (iii), are `r fixed(kl_iii[1L, "diff"], 4)` and
`r fixed(kl_iii[2L, "diff"], 4)`, with standard errors of
`r fixed(kl_iii[1L, "se"], 4)` and `r fixed(kl_iii[2L, "se"], 4)`. Its
assignments of the sample it never fitted reach an adjusted Rand index of
`r fixed(sim_value(n_large, "proxymix, regime (iii)", "ari"), 3)` at
`r n_large` points. Per fit at `r n_large` points, proxymix regime (ii)
took `r fixed(sim_value(n_large, "proxymix, regime (ii)", "secs"), 3)` s,
regime (iii)
`r fixed(sim_value(n_large, "proxymix, regime (iii)", "secs"), 3)` s,
mclust `r fixed(sim_value(n_large, "mclust", "secs"), 3)` s and flexmix
with five starts `r fixed(sim_value(n_large, "flexmix", "secs"), 3)` s.

In this run proxymix regime (ii) had
`r if (resolved(kl_small) && resolved(kl_large) && kl_small[["diff"]] < 0 && kl_large[["diff"]] < 0) "a lower mean divergence than mclust at both sizes, by more than two paired standard errors" else "a mean divergence that the paired comparison does not separate from mclust's at one or both sizes"`,
and regime (iii) returned a closer mixture than any sample route because
it read the density the samples came from. At `r n_small` points mclust
exceeded the divergence threshold in fewer datasets than proxymix regime
(ii), `r sum(mis_mc)` against `r sum(mis_ii)` with a McNemar p-value of
`r fixed(mcnemar_p, 2)`, and mclust chose its covariance structure as it
went, which no other method did. The comparison covers one mixture in two
dimensions with the component count supplied, and neither a mis-specified
target, a chosen component count nor a higher dimension was tried.

## Further reading

*Fitting a proxy to a density you cannot sample* is the shorter route into
regime (iii) alone, and reads the fit certificate that this vignette only
glances at.

*How well a mixture proxies four awkward shapes* takes regime (iii) to
targets outside the Gaussian-mixture family, where the correctly-specified
comfort of this vignette's test bench no longer holds.

*Reading the entropy of a fitted mixture* covers the component-count
question this vignette defers, including the annealed path that discovers a
count rather than being told one.

*One mixture, many methods* shows what a single fitted mixture replaces
once the regime question is settled.

*Mapping the optima of an objective* points regime (iii) at a function that
is not a density at all: an objective read as a Gibbs measure, so that the
fitted mixture returns every basin of a multimodal landscape at once rather
than the one a local optimiser happened to reach.

## References

* Azzalini, A. and Bowman, A. W. (1990). *A look at some data on the Old
  Faithful geyser.* Journal of the Royal Statistical Society, Series C
  (Applied Statistics) 39(3), 357--365.\
  <https://doi.org/10.2307/2347385>.
* Benaglia, T., Chauveau, D., Hunter, D. R. and Young, D. S. (2009).
  *mixtools: An R package for analyzing finite mixture models.* Journal of
  Statistical Software 32(6), 1--29.\
  <https://doi.org/10.18637/jss.v032.i06>.
* Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
  Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A. (2017).
  *Stan: A probabilistic programming language.* Journal of Statistical
  Software 76(1), 1--32.\
  <https://doi.org/10.18637/jss.v076.i01>.
* Duong, T. (2007). *ks: Kernel density estimation and kernel discriminant
  analysis for multivariate data in R.* Journal of Statistical Software
  21(7), 1--16.\
  <https://doi.org/10.18637/jss.v021.i07>.
* Grün, B. and Leisch, F. (2008). *FlexMix version 2: Finite mixtures
  with concomitant variables and varying and constant parameters.* Journal
  of Statistical Software 28(4), 1--35.\
  <https://doi.org/10.18637/jss.v028.i04>.
* Hubert, L. and Arabie, P. (1985). *Comparing partitions.* Journal of
  Classification 2(1), 193--218.\
  <https://doi.org/10.1007/BF01908075>.
* Leisch, F. (2004). *FlexMix: A general framework for finite mixture
  models and latent class regression in R.* Journal of Statistical Software
  11(8), 1--18.\
  <https://doi.org/10.18637/jss.v011.i08>.
* Scrucca, L., Fop, M., Murphy, T. B. and Raftery, A. E. (2016). *mclust
  5: Clustering, classification and density estimation using Gaussian
  finite mixture models.* The R Journal 8(1), 289--317.\
  <https://doi.org/10.32614/RJ-2016-021>.
* van der Hoek, J. and Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.\
  <https://doi.org/10.1080/07362994.2024.2372605>.

## Reproduce

The target carries `seed = 1L`, the regime-(ii) fit and both regime-(iii)
fits on the bench are seeded with `seed = 1L`, the exact moments are closed
form, and the replicate traces use seeds 1 to `r length(sample_traces)` for
the samples and 1 to `r length(kld_traces)` for the importance draws, so
every number in the first five sections is reproducible from the package
alone.

The numerical illustration splits the eruptions after `set.seed(20260925)`,
runs the four regime-(ii) fits after `set.seed(1L)`, and gives the
regime-(iii) fit `seed = 1L`. The stored sampling results come from the
code shown, run on `r ref$run_date` with R `r ref$versions[["R"]]`,
cmdstanr `r ref$versions[["cmdstanr"]]`, CmdStan
`r ref$versions[["CmdStan"]]`, ks `r ref$versions[["ks"]]` and mclust
`r ref$versions[["mclust"]]`, and that run raised
`r length(ref$warnings)` warnings. In the simulation each dataset is
generated after `set.seed()` with its own index, and the regime-(iii)
importance sampler is seeded with that index plus the sample size. The
stored simulation results come from the code shown, run on `r sim$run_date` with R `r sim$versions[["R"]]`,
proxymix `r sim$versions[["proxymix"]]`, mclust
`r sim$versions[["mclust"]]` and flexmix `r flexmix_version`,
and that run raised `r length(sim$warnings)` warnings. The mixtools
version used above is `r format(packageVersion("mixtools"))`.

```{r session-info}
old_opt <- options(width = 72L)
si_vec <- trimws(capture.output(sessionInfo()), "right")
options(old_opt)
# library paths are single long words, so a long line breaks after a slash
long_vec <- nchar(si_vec) > 72L
si_vec[long_vec] <- gsub("(.{1,68}/)(?=.{8,})", "\\1\n        ",
                         si_vec[long_vec], perl = TRUE)
writeLines(si_vec)
```
